### 2024-06-20 doc/manuscript.Rmd
manuscript set up docs

---
title: ''
output: 
    bookdown::word_document2: 
      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    #pdf_document: 
  #officedown::rdocx_document
bibliography: growth-precip-thailand.bib
csl: new-phytologist.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Display items

## Tables

## Figures

# Read in files

```

**Title:** Vulnerability of tree growth to precipitation in Thailand

**Authors**  
```{r eval = TRUE, echo=FALSE, warning=FALSE, eval=F}
library(flextable)
set_flextable_defaults(font.family = "Arial", font.size = 9)
team<- read.csv("display/authors.csv", stringsAsFactors = FALSE, check.names = FALSE)

ft <- flextable(team)
ft <- line_spacing(ft, space = 1, part = "all")
ft <- autofit(ft)
ft <- width(ft, 1, width = 1.75)
ft <- width(ft, 2, width = 1.75)
ft <- width(ft, 3, width = 3)
ft

```

\newpage

# Abstract {-}

\newpage

# Introduction {-}

Tree woody growth key is a key component of aboveground productivity and affects the global carbon carbon cycle, but the drivers of its variation are poorly understood.
- Woody growth contributes x% of the aboveground woody productivity
- Long-lasting effects on the terrestrial carbon pool
- Intrinsic and extrinsic factors affecting variation.

# Methods {-}

ForestGEO data [@anderson-teixeira_ctfsforestgeo_2015]

***Sites and data*** -

# Results {-}

# Discussion {-}

\newpage
# References  {-}

### 2024-06-20 doc/supplementary_information.Rmd
manuscript set up docs

---
title: "Supplementary Information: Vulnerability of tree growth"
author: 
date: 
output: 
  pdf_document:
    fig_caption: yes
    toc: true # table of content true
    toc_depth: 2  
bibliography: growth-precip-thailand.bib
suppress-bibliography: false
csl: methods-in-ecology-and-evolution.csl
header-includes:  \usepackage{float}
                  \usepackage{caption}
                  \captionsetup[table]{font=footnotesize}
                  \captionsetup[figure]{font=footnotesize}
                  \captionsetup[figure]{labelformat=empty}
                  \captionsetup[table]{labelformat=empty}
                  \usepackage{pdflscape}
                  \newcommand{\blandscape}{\begin{landscape}}
                  \newcommand{\elandscape}{\end{landscape}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Appendices

#SI table numbers

#SI table captions

#SI figure numbers

#SI figure captions

#https://github.com/EcoClimLab/Operations/blob/master/Data_Management-Scientific_Workflow/manuscript_prep_with_Rmd.md
```

```{r read files, echo=F}
# Read in files

```
\newpage

### 2024-06-20 scripts/explore_analysis.Rmd
climate data and exploratory analysis script

---
title: "Climate Sensitivity of Tropical Tree Growth"
author: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Climate Sensitivity of Tropical Tree Growth

## HKK climate

```{r echo = FALSE, message=F, warning=F}

#load libraries
library(tidyverse)
library(lubridate)
library(httr)
library(janitor)

#reading the GitHub PAT (stored privately and locally)
pat<-as.character(read.table("git_pat.txt")[1,1])

#remotes:::download("DendroByBand.RData",
#                   "https://raw.githubusercontent.com/forestgeo/HKK-dendrobands/main/data/modified_data/DendroByBand.RData",
#                   auth_token = pat)

load("DendroByBand.RData")

dendrobyband<-merge(dendrobyband, trees, by="Tag")

hkk_OBSID$year<-year(hkk_OBSID$Date2)
hkk_OBSID$month<-month(hkk_OBSID$Date2)
hkk_OBSID$day<-day(hkk_OBSID$Date2)

#correcting two sets of wrong dates

hkk_OBSID$year[hkk_OBSID$year==2209]<-2009
hkk_OBSID$year[hkk_OBSID$month==1 & hkk_OBSID$Cno==5]<-2011

dendrobyband<-merge(dendrobyband, hkk_OBSID, by=c("OBSID", "Cno", "Date2", "Tag"))

dendrobyband$Date2<-as.Date(paste0(dendrobyband$year,"-", dendrobyband$month, "-", dendrobyband$day))

#this can be refined if needed
wet_census<-c(1, 3, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24, 26, 28)

#getting climate data for all

req <- GET("https://api.github.com/repos/forestgeo/Climate/git/trees/master?recursive=1")
stop_for_status(req)
filelist <- unlist(lapply(content(req)$tree, "[", "path"), use.names = F)
path<-grep("Climate_Data/CRU/CRU_v4_06", filelist, value = TRUE, fixed = TRUE)
datacsvs<-grep("pre|tmx", path, value = T)

#datacsvs<-datacsvs[!grepl("BLANK", datacsvs)]
datacsvs<-paste0("https://github.com/forestgeo/Climate/raw/master/", datacsvs)

start_date<-min(hkk_OBSID$year,na.rm=T)
end_date<-max(hkk_OBSID$year,na.rm=T)

climate_data_orig<-bind_rows(datacsvs %>%
                               read_csv(na = c("", "NA", "N/A", "NULL"), show_col_types = FALSE))%>%
  clean_names() %>%
  filter(sites_sitename %in% c("HKK",
                               "Huai Kha Khaeng")) %>%
  select(-c(sites_sitename, n_obs)) %>%
  rename(year=yr,
         month=mo)%>%
  filter(start_date-1 <= year & year <= end_date) # limiting to specified date range but extending to a year before

census_int<-dendrobyband%>%
  group_by(Cno) %>%
  dplyr::summarise(start=min(Date2, na.rm = T),
                   end=max(Date2, na.rm=T))%>%
  dplyr::mutate(type=ifelse(Cno %in% wet_census, "wet", "dry"))

pre_yr_census<-ggplot()+
  geom_area(data=climate_data_orig[climate_data_orig$variable=="pre",], aes(x=as.Date(paste0(year, "-", month, "-", "01"), format = "%Y-%m-%d"), y=value), fill="blue", alpha=0.5)+
  geom_line(data=climate_data_orig[climate_data_orig$variable=="tmx",], aes(x=as.Date(paste0(year, "-", month, "-", "01"), format = "%Y-%m-%d"), y=value*5), col="red")+
  geom_vline(data=census_int[census_int$type=="wet",], aes(xintercept=start), col="darkgreen")+
  geom_vline(data=census_int[census_int$type=="wet",], aes(xintercept=end), col="darkgreen")+
  geom_vline(data=census_int[census_int$type=="dry",], aes(xintercept=start), col="brown4")+
  geom_vline(data=census_int[census_int$type=="dry",], aes(xintercept=end), col="brown4")+
  scale_y_continuous(    # Features of the first axis
    name = "monthly rainfall (mm)",

    # Add a second axis and specify its features
    sec.axis = sec_axis(~.*0.2, name="maximum temperature (Celsius °)"))+
  xlab("Date")+theme_bw()+
  ggtitle("HKK dendroband censuses and CRU climate data")+
  #scale_x_continuous(n.breaks=15)+
  theme(axis.text.x = element_text(angle=45))

pre_yr_census

```
```{r echo=F}

#remove all data with decimal errors, misplaced decimal

dendro.data<-dendrobyband %>%
  dplyr::rename(sp=SPCODE.UPDATE)%>%
  mutate(year=year(Date2)) %>%
  #filter(year!=2209)%>% #quick fix to remove bad dates - there are 30 entries
  #filter(start_date <= year & year <= end_date) %>%
  filter(Cno %in% wet_census)%>%
  filter(is.na(DBH1_flag), DBH1_error=="none",
         win.decimal.error=="none", is.na(win.flag))%>%
  group_by(year, sp, Tag) %>%
  dplyr::summarise(calcDBH=calcDBH[which(n.dendro==min(n.dendro))],
                   Date2=Date2[which(n.dendro==min(n.dendro))],
                   Cii=max(Cii, na.rm=T))

dendro.data_all<-dendro.data%>%
  group_by(Tag) %>% #grouping by treeID to calculate increment
  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
         inc=calcDBH-calcDBH_min1,
         inc_annual=(inc/as.numeric(tdif))*365,
         inc.time=year-lag(year),
         #removing increment value if the previous measure was before the last year
         inc_annual=ifelse(inc.time==1, inc_annual, NA),
         #has.out=ifelse(1 %in% out, 1, 0),
         treeID=paste0("HKK.", Tag))%>%
  filter(!is.na(inc))%>%
  ungroup()%>%
  filter(!is.na(inc_annual))%>%
  ungroup()%>%
  #filter(start_date <= year & year <= end_date) %>%
  group_by(sp, treeID, year)%>%
  dplyr::summarise(
    diainc = mean(inc_annual, na.rm=T),
    Cii=max(Cii, na.rm=T))

#first fitting a line through the time series
obs.model.dbh<-nlme::lme(calcDBH~year, random=~1|Tag, data=dendro.data,
                         na.action=na.omit,
                         method="ML", control=(msMaxIter=100)) #had to increase the number of iterations

#standardised residuals calculated https://stat.ethz.ch/R-manual/R-devel/library/nlme/html/residuals.lme.html

std.res.dbh<-residuals(obs.model.dbh, type="pearson")
hist(std.res.dbh)
dendro.data$std.res<-rep(NA, nrow(dendro.data))
dendro.data$std.res[!is.na(dendro.data$calcDBH)]<-std.res.dbh
#defining all observation with abs(residual) >3 as an outlier
dendro.data$out<-factor(ifelse(abs(dendro.data$std.res)<=3, 0, 1))

#find the trees which have outlier observations
outs.Tag<-unique(dendro.data$Tag[dendro.data$out==1])

#plot trees that have outliers
ggplot(dendro.data[dendro.data$Tag %in%outs.Tag,],
       aes(x = year, y = calcDBH, col=factor(out), group = Tag))+
  geom_line(alpha=0.5)+geom_point()+
  scale_color_manual(values=c("grey20", "red", "grey"))

dendro.data.noouts<-dendro.data %>%
  mutate(dbh3=ifelse(abs(std.res)<=3, calcDBH, NA)) %>% #create an outlier removed column
  group_by(Tag) %>% #grouping by treeID to calculate increment
  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
         inc=calcDBH-calcDBH_min1,
         inc_annual=(inc/as.numeric(tdif))*365,
         inc.time=year-lag(year),
         #removing increment value if the previous measure was before the last year
         #inc=ifelse(inc.time==1, inc, NA),
         inc_annual=ifelse(inc.time==1, inc_annual, NA),
         has.out=ifelse(1 %in% out, 1, 0),
         treeID=paste0("HKK.", Tag))%>%
  filter(!is.na(inc))%>%
  ungroup()%>%
  filter(!is.na(inc_annual))%>%
  ungroup()%>%
  #filter(start_date <= year & year <= end_date) %>%
  group_by(sp, treeID, year)%>%
  dplyr::summarise(
    diainc = mean(inc_annual, na.rm=T),
    calcDBH_min1 = mean(calcDBH_min1, na.rm=T),
    dbh=mean(calcDBH, na.rm=T),
    out = ifelse(1 %in% out, 1, 0),
    has.out = ifelse(1 %in% has.out, 1, 0), 
    cii=max(Cii, na.rm=T))%>%
  filter(out==0)%>%
  ungroup() %>%
  dplyr::mutate(
    diainc.scaled = scale(diainc),
    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))%>%
  filter(out.inc==0)

```

## Raw diameter increments of all species

```{r echo=FALSE}

library(ggplot2)

growth_yr_withouts<-ggplot(dendro.data_all, aes(x=year, y=diainc, group=year))+
  geom_violin()+
  ylab("Annualised diameter increment (cm)")+theme_bw()

growth_yr_withouts
```

## Removing outliers based (more than 3 SD)

```{r echo=FALSE}

growth_yr_noouts<-ggplot(dendro.data.noouts, aes(x=year, y=diainc, group=year))+ geom_violin()+ ylab("Annualised diameter increment (cm)")+theme_bw()

growth_yr_noouts
```

## Tree ring species

- Four species with tree ring chronologies in HKK

```{r ring trees, echo=F}
#These are species that have been cored near HKK and reported in Zuidema et al., Anderson-Teixeira et al., etc
ring.sp<-c("AFZEXY", "MELIAZ", "TOONCI", "CHUKTA", "NEOLOB")

trees.with.ring<-trees[trees$SPCODE.UPDATE %in% ring.sp,]

dendro.ring<-dendro.data.noouts %>%
  filter(sp %in% ring.sp)

hkk_sp_ring<-trees.with.ring %>%
  group_by(SPCODE.UPDATE)%>%
  dplyr::summarise(n.tree=length(SPCODE.UPDATE))
knitr::kable(hkk_sp_ring, format='html', table.attr="style='width:50%;'")

```

```{r maps of species with tree ring records, echo=F}
library(ggplot2)

pat<-as.character(read.table("git_pat.txt")[1,1])

remotes:::download("stem1.RData",
                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem1.rdata",
                   auth_token = pat)
remotes:::download("stem2.RData",
                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem2.rdata",
                   auth_token = pat)
remotes:::download("stem3.RData",
                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem3.rdata",
                   auth_token = pat)
remotes:::download("stem4.RData",
                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem4.rdata",
                   auth_token = pat)

load("stem1.RData")
load("stem2.RData")
load("stem3.RData")
load("stem4.RData")

hkk_census<-bind_rows(hkk.stem1, hkk.stem2, hkk.stem3, hkk.stem4)

hkk_census <- hkk_census %>%
  filter(status=="A")

hkk_census_ring_sp<-hkk_census %>%
  filter(CensusID==3)%>%
  filter(sp %in% ring.sp)

hkk_census_sp <- hkk_census %>%
  filter(CensusID==4)%>%
  #group_by(sp, tag, CensusID) %>%
  #dplyr::summarise(tot.agb=sum(agb, na.rm=T)) %>%
  group_by(sp, tag)%>%
  dplyr::summarise(tot.agb=sum(agb, na.rm=T)) %>%
  group_by(sp) %>%
  dplyr::summarise(n.tree=length(unique(tag)),
                   sum.agb=sum(tot.agb))%>%
  ungroup()%>%
  dplyr::mutate(pct.agb=sum.agb*100/sum(sum.agb, na.rm=T))%>%
  arrange(-pct.agb)%>%
  dplyr::mutate(pct.agb.cum=cumsum(pct.agb))

sp_top15<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA",
            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA",
            "GARCSP", "CROTRO", "GARUPI", "NEOLOB", "PHOEPA")

#pct agb of the top 15 dendro species
sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% sp_top15])

ring.sp.map.census<-ggplot(hkk_census_ring_sp, aes(x=gx, y=gy, size=as.numeric(dbh/10), color=factor(sp)))+
  geom_point(alpha=0.5)+
   scale_size(range = c(.1, 10), name="DBH (cm)")+
  guides(color=guide_legend(title="Species"))+
  #scale_colour_gradient(low="orange", high="purple") +
  theme_minimal()+
  coord_fixed()

ring.sp.map.census

ring.sp.map<-ggplot(trees.with.ring, aes(x=X, y=Y, size=as.numeric(DBH.04), color=factor(SPCODE.UPDATE)))+
  geom_point(alpha=0.5)+
   scale_size(range = c(.1, 10), name="DBH (cm)")+
  guides(color=guide_legend(title="Species"))+
  #scale_colour_gradient(low="orange", high="purple") +
  theme_minimal()+
  coord_fixed()

ring.sp.map
```
Species with tree rings contribute `r round(sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% ring.sp]),2)` % aboveground biomass in 2004

## Dendrobands on ring species

```{r echo=F}

growth_sp<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
                                              fill=factor(sp),
                                              color=factor(sp)))+ 
  geom_violin(position = "dodge")+ 
  geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #facet_grid(.~year)+
  #geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
  geom_hline(yintercept = median(dendro.ring$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
      theme(axis.text.x = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  geom_text(aes(label=paste0("median=\n", round(median(dendro.ring$diainc, na.rm=T),2), " cm"), 
                x=0.6, y=median(dendro.ring$diainc, na.rm=T)+0.15), col="black", size=3)+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))

growth_sp

growth_sp_box<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
                                              #color=factor(sp),
                                              fill=factor(sp)))+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #facet_grid(.~year)+
  geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
  geom_hline(yintercept = median(dendro.ring$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
      theme(axis.text.x = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))

#growth_sp_box

growth_yr_sp<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
                                              #color=factor(sp),
                                              fill=factor(sp)))+ 
  #geom_violin(position = "dodge")+ 
  geom_boxplot(position="dodge")+
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  facet_grid(.~year)+
  geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
  theme(axis.text.x = element_blank())+
  guides(color="none")

growth_yr_sp

```

## Detrending by tree with smooth splines

```{r detrending, echo=F}

#trying for one tree

dendro.ring<-dendro.ring %>%
  group_by(sp, treeID) %>%
  dplyr::mutate(n.obs=length(treeID))%>%
  filter(n.obs>=4)%>%
  ungroup()%>%
  group_by(sp, treeID)%>%
  dplyr::mutate(inc.resid=residuals(smooth.spline(x=calcDBH_min1, y=diainc)),
                inc.spline=fitted(smooth.spline(x=calcDBH_min1, y=diainc)))

growth_sp_detrend<-ggplot(dendro.ring, aes(x=factor(sp), 
                                           y=inc.resid, 
                                           #color=factor(sp),
                                           fill=factor(sp)))+ 
  #geom_violin()+ 
  geom_boxplot()+ 
  ylab("Residuals of annualised diameter increment")+
  theme_bw()+
  theme(axis.text.x = element_text(angle=45))
growth_sp_detrend

growth_yr_detrend<-ggplot(dendro.ring, aes(x=year, y=inc.resid, group=year))+ geom_boxplot()+ 
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  theme(axis.text.x = element_text(angle=45))
growth_yr_detrend

```

## Detrended plots by species

```{r echo=F}

growth_yr_sp_detrend<-ggplot(dendro.ring, aes(x=factor(sp), y=inc.resid,
                                              #group=c(year, sp),
                                              #group=sp,
                                              #color=factor(sp),
                                              fill=factor(sp)))+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  geom_hline(yintercept=0)+
  facet_grid(.~year)+
  theme(axis.text.x = element_blank())+
  guides(color="none")

growth_yr_sp_detrend

```

```{r echo=F}

spag_all<-ggplot(dendro.ring, aes(x=year, y=inc.spline, group=treeID)) +
  geom_line()+ggtitle("fitted values (smooth spline)")

spag_all

spag_resid<-ggplot(dendro.ring, aes(x=year, y=inc.resid, group=treeID)) +
  geom_line()+ggtitle("residuals smooth spline")

spag_resid
```
## Detrending all species

```{r detrending all, echo=F}

#trying for one tree

dendro.data.noouts<-dendro.data.noouts %>%
  group_by(sp, treeID) %>%
  dplyr::mutate(n.obs=length(treeID))%>%
  filter(n.obs>=4)%>%
  ungroup()%>%
  filter(treeID!="HKK.92562")%>% #removing an issue with spline for now
  group_by(sp, treeID)%>%
  dplyr::mutate(inc.resid=residuals(smooth.spline(x=calcDBH_min1, y=diainc)),
                inc.spline=fitted(smooth.spline(x=calcDBH_min1, y=diainc)))
```

For `r length(unique(dendro.data.noouts$sp))` species with dendrobands, all individual trees detrended for DBH with smooth splines

```{r echo=F}

growth_yr_detrend_all<-ggplot(dendro.data.noouts, aes(x=year, y=inc.resid, group=year))+ 
  #geom_violin()+ 
  geom_boxplot()+ 
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  theme(axis.text.x = element_text(angle=45))
growth_yr_detrend_all

```

## top 10 species

```{r echo=F}

dendro.by.sp<-dendro.data.noouts %>%
  group_by (sp, year) %>%
  dplyr::summarise(n.inds=length(sp),
    mean.inc = mean(inc.resid, na.rm=T),
                   median.inc = median(inc.resid, na.rm=T),
                   sd.inc = sd(inc.resid, na.rm=T),
    lower=mean.inc-sd.inc,
    upper=mean.inc+sd.inc)%>%
  filter(n.inds>=10)%>%
  arrange(year, -median.inc)

sp_top5<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA")

sp_top10<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA", 
            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA")

dendro.top5<-dendro.by.sp %>%
  filter(sp %in% sp_top5)

dendro.top5.dat<-dendro.data.noouts %>%
  filter(sp %in% sp_top5)

dendro.top10.dat<-dendro.data.noouts %>%
  filter(sp %in% sp_top10)

hkk_census_top10_sp<-hkk_census %>%
  filter(CensusID==3)%>%
  filter(sp %in% sp_top10)

#pct agb of the top 15 dendro species
sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% sp_top15])

top10sp.map.census<-ggplot(hkk_census_top10_sp, aes(x=gx, y=gy, 
                                                    #size=as.numeric(dbh/10),
                                                    color=factor(sp, levels=sp_top10)))+
  geom_point(size=1)+
   scale_size(range = c(.1, 10), name="dbh")+
  guides(color=guide_legend(title="Species"))+
  scale_color_viridis_d()+
  ggtitle("top 10 species in 2004")+
  #scale_colour_gradient(low="orange", high="purple") +
  theme_minimal()+
  coord_fixed()

top10sp.map.census

trees.top10<-trees[trees$SPCODE.UPDATE %in% sp_top10,]

top10sp.map.dendro<-ggplot(trees.top10, aes(x=X, y=Y, 
                                                    #size=as.numeric(dbh/10), 
                                                    color=factor(SPCODE.UPDATE, levels=sp_top10)))+
  geom_point(size=1)+
   scale_size(range = c(.1, 10), name="dbh")+
  guides(color=guide_legend(title="Species"))+
  scale_color_viridis_d()+
  ggtitle("top 10 species in 2004")+
  #scale_colour_gradient(low="orange", high="purple") +
  theme_minimal()+
  coord_fixed()

top10sp.map.dendro

```

## Growth of most abundant species

```{r echo=F}

growth_sp5<-ggplot(dendro.top5.dat, aes(x=factor(sp, levels=sp_top5), y=diainc,
                                              fill=factor(sp, levels=sp_top5),
                                              color=factor(sp, levels=sp_top5)))+ 
  geom_violin(position = "dodge")+ 
    geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top5.dat$diainc, na.rm=T))+
  geom_hline(yintercept = median(dendro.top5.dat$diainc, na.rm=T), lty=2)+
  geom_text(aes(label=paste0("median=\n", round(median(dendro.top5.dat$diainc, na.rm=T),2), " cm"), 
                x=0.6, y=median(dendro.top5.dat$diainc, na.rm=T)+0.15), col="black", size=3)+
  geom_hline(yintercept = 0, lty=3)+
      theme(axis.text.x = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))+
  xlab("")

growth_sp5

growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(sp, levels=sp_top10), y=diainc,
                                              fill=factor(sp, levels=sp_top10),
                                              color=factor(sp, levels=sp_top10)))+ 
  geom_violin(position = "dodge")+ 
    geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_text(aes(label=paste0("median=\n", round(median(dendro.top10.dat$diainc, na.rm=T),2), " cm"), 
                x=0.6, y=median(dendro.top10.dat$diainc, na.rm=T)+0.15), col="black", size=3)+
  geom_hline(yintercept = 0, lty=3)+
      theme(axis.text.x = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))+
  xlab("")

growth_sp10

growth_yr_sp<-ggplot(dendro.top5, aes(x=year, y=mean.inc, 
                                      color=sp))+ 
  geom_point()+
  geom_errorbar(aes(x=year, ymin=lower, ymax=upper))+
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  theme(axis.text.x = element_text(angle=45))+
  facet_grid(sp~.)+
  guides(color="none")+
  geom_hline(yintercept = 0)+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  geom_vline(xintercept=c(2010, 2016), lty=2)
growth_yr_sp

#png("plots/growth_by_sp.png", height=8, width=4, units="in", res=300)
#growth_yr_sp
#dev.off()

```

## Size classes

```{r echo=F}

size_top10<-ggplot(dendro.top10.dat, aes(x=dbh))+
  geom_density()+
  geom_vline(aes(xintercept=35))+
  geom_vline(aes(xintercept=75))

size_top10

#trying three categories
# <50, 50-100, >100

dendro.top10.dat<-dendro.top10.dat %>%
  dplyr::mutate(size=ifelse(dbh<35, "< 35 cm dbh",
                       ifelse(dbh<70, "35-75 cm dbh", ">75 cm dbh")))

growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(sp, levels=sp_top10), y=diainc,
                                              #color=factor(sp),
                                          #group=factor(size, levels=c("small", "medium", "large")),
                                              fill=factor(sp, levels=sp_top10)))+ 
  #geom_violin(position = "dodge")+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
  facet_grid(factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh"))~.)+
      theme(axis.text.y = element_blank())+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))+coord_flip()

growth_sp10

growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh")), y=diainc,
                                              #color=factor(sp),
                                          #group=factor(size, levels=c("small", "medium", "large")),
                                              fill=factor(sp, levels=sp_top10)))+

  geom_violin(position = "dodge")+ 
  geom_boxplot(position = "dodge", col="black", fill="white", width=0.1)+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
  #facet_grid(factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh"))~.)+
      theme(axis.text.x = element_text(angle=45))+
  xlab("")+
  facet_grid(.~factor(sp, levels=sp_top10))+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))#+coord_flip()

growth_sp10

```

## Canopy illumination

```{r echo=F}

cii_top10<-ggplot(dendro.top10.dat, aes(x=cii))+
  geom_histogram()

cii_top10

#trying three categories
# <50, 50-100, >100

#dendro.top10.dat<-dendro.top10.dat %>%
#  dplyr::mutate(size=ifelse(dbh<50, "small",
#                       ifelse(dbh<100, "medium", "large")))

growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
                        aes(x=factor(sp, levels=sp_top10), y=diainc,
                                              #color=factor(sp),
                                          #group=factor(size, levels=c("small", "medium", "large")),
                                              fill=factor(sp, levels=sp_top10)))+ 
  #geom_violin(position = "dodge")+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
  facet_grid(cii~.)+
      theme(axis.text.y = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))+coord_flip()

growth_sp10_cii

growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
                        aes(
                          #x=factor(sp, levels=sp_top10), 
                          x=factor(cii), 
                            y=diainc,
                                              #color=factor(sp),
                                          #group=factor(size, levels=c("small", "medium", "large")),
                                              fill=factor(sp, levels=sp_top10)))+ 
  #geom_violin(position = "dodge")+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
  facet_grid(.~factor(sp, levels = sp_top10))+
      theme(axis.text.y = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  #coord_flip()+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))
growth_sp10_cii

```
## spatial patterns

```{r spatial patterns, echo=F}

trees.top10 <- trees.top10 %>%
  dplyr::mutate(treeID=paste0("HKK.", Tag),
                plotID=paste0(Col, "_", Row))

dendro.top10.dat<-merge(dendro.top10.dat, 
                        dplyr::select(trees.top10, c("Col", "Row", "treeID", "X", "Y", "plotID")), 
                        by="treeID", all.x=T)

dendro.top10.dat<-dendro.top10.dat %>%
  dplyr::mutate(plotID=paste0(Col, "_", Row))

census_spatial<-hkk_census %>%
  group_by(quadrat) %>%
  dplyr::summarise(sp.no=length(unique(sp)),
                   sp.top.10 = length(unique(sp %in% sp_top10)))%>%
  dplyr::mutate(Col=as.numeric(str_sub(quadrat, 1, 2)),
                Row=as.numeric(str_sub(quadrat, 3, 4)))

tiles<-ggplot(census_spatial, aes(x=Col, y=Row, fill=sp.no))+
  geom_tile()+
  #facet_wrap(~year)+
  coord_fixed()+
  #xlim(c(0,49))+ylim(c(0, 24))+
  theme_bw()

tiles

tiles<-ggplot(census_spatial, aes(x=Col, y=Row, fill=sp.top.10))+
  geom_tile()+
  #facet_wrap(~year)+
  coord_fixed()+
  #xlim(c(0,49))+ylim(c(0, 24))+
  theme_bw()

tiles

spatial_summary<- dendro.top10.dat %>%
  group_by(Col, Row, plotID, sp, year) %>%
  dplyr::summarise(inc.sum=sum(inc.resid, na.rm=T)) %>%
  group_by(Col, Row, plotID, year) %>%
  dplyr::summarise(inc.diff=var(inc.sum, na.rm=T))

tiles<-ggplot(spatial_summary, aes(x=Col, y=Row, fill=log(inc.diff, 10)))+
  geom_tile()+
  facet_wrap(~year)+coord_fixed()+
  xlim(c(0,49))+ylim(c(0, 24))+
  theme_bw()

tiles

```

## evergreen/dec

```{r evg/dec, echo=F}

sp_top15<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA", 
            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA",
            "GARCSP", "CROTRO", "GARUPI", "NEOLOB", "PHOEPA")

#these 15 species had the most dendrobands and together made up 51.22% of the agb in 2009

#below values eyeballed from Williams et al. 2008 Fig 1.
sp_top15_phenoscore<-c(2.5, 0.8, 1.3, 4, 0.7, 
                       1, 1.25, 1.27, 0.2, 0.3,
                       0.4, 0.65, 3.95, 0.15, 0.25)

#no info on Alphonsea ventricosa in the paper. 
#Apparently evergreen, so used 1 as value - middle of evergreen range

pheno<-as.data.frame(cbind(sp_top15, sp_top15_phenoscore))
colnames(pheno)<-c("sp", "phenoscore")

dendro.top10.dat<-merge(dendro.top10.dat, pheno, by="sp")

#categorising as evg and deciduous
dendro.top10.dat$cat<-ifelse(dendro.top10.dat$phenoscore<=1, "evg", "dec")

growth_yr_cat_detrend<-ggplot(dendro_top15, aes(x=factor(cat), y=inc.resid,
                                              #group=c(year, sp),
                                              #group=sp,
                                              fill=factor(cat)))+ 
  geom_violin(position = "dodge")+ 
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  facet_grid(.~year)+
  geom_hline(yintercept = 0)+
  scale_fill_manual(values=c("sienna4", "darkgreen"))+
  theme(axis.text.x = element_blank())

growth_yr_cat_detrend

growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
                        aes(
                          #x=factor(sp, levels=sp_top10), 
                          x=factor(cii), 
                            y=diainc,
                                              #color=factor(sp),
                                          #group=factor(size, levels=c("small", "medium", "large")),
                                              fill=factor(sp, levels=sp_top10)))+ 
  #geom_violin(position = "dodge")+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
  facet_grid(cat~factor(sp, levels = sp_top10))+
      theme(axis.text.y = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  #coord_flip()+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))
growth_sp10_cii

```

## Repeat all steps for dry censuses

```{r echo=F}

#remove all data with decimal errors, misplaced decimal

`%nin%` <- Negate(`%in%`)

dendro.data<-dendrobyband %>%
  dplyr::rename(sp=SPCODE.UPDATE)%>%
  mutate(year=year(Date2)) %>%
  #filter(year!=2209)%>% #quick fix to remove bad dates - there are 30 entries
  #filter(start_date <= year & year <= end_date) %>%
  filter(Cno %nin% wet_census)%>%
  filter(is.na(DBH1_flag), DBH1_error=="none",
         win.decimal.error=="none", is.na(win.flag))%>%
  group_by(year, sp, Tag) %>%
  dplyr::summarise(calcDBH=calcDBH[which(n.dendro==min(n.dendro))],
                   Date2=Date2[which(n.dendro==min(n.dendro))],
                   Cii=max(Cii, na.rm=T))

dendro.data_all<-dendro.data%>%
  group_by(Tag) %>% #grouping by treeID to calculate increment
  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
         inc=calcDBH-calcDBH_min1,
         inc_annual=(inc/as.numeric(tdif))*365,
         inc.time=year-lag(year),
         #removing increment value if the previous measure was before the last year
         inc_annual=ifelse(inc.time==1, inc_annual, NA),
         #has.out=ifelse(1 %in% out, 1, 0),
         treeID=paste0("HKK.", Tag))%>%
  filter(!is.na(inc))%>%
  ungroup()%>%
  filter(!is.na(inc_annual))%>%
  ungroup()%>%
  #filter(start_date <= year & year <= end_date) %>%
  group_by(sp, treeID, year)%>%
  dplyr::summarise(
    diainc = mean(inc_annual, na.rm=T),
    Cii=max(Cii, na.rm=T))

#first fitting a line through the time series
obs.model.dbh<-nlme::lme(calcDBH~year, random=~1|Tag, data=dendro.data,
                         na.action=na.omit,
                         method="ML", control=(msMaxIter=100)) #had to increase the number of iterations

#standardised residuals calculated https://stat.ethz.ch/R-manual/R-devel/library/nlme/html/residuals.lme.html

std.res.dbh<-residuals(obs.model.dbh, type="pearson")
hist(std.res.dbh)
dendro.data$std.res<-rep(NA, nrow(dendro.data))
dendro.data$std.res[!is.na(dendro.data$calcDBH)]<-std.res.dbh
#defining all observation with abs(residual) >3 as an outlier
dendro.data$out<-factor(ifelse(abs(dendro.data$std.res)<=3, 0, 1))

#find the trees which have outlier observations
outs.Tag<-unique(dendro.data$Tag[dendro.data$out==1])

#plot trees that have outliers
ggplot(dendro.data[dendro.data$Tag %in%outs.Tag,],
       aes(x = year, y = calcDBH, col=factor(out), group = Tag))+
  geom_line(alpha=0.5)+geom_point()+
  scale_color_manual(values=c("grey20", "red", "grey"))

dendro.data.noouts<-dendro.data %>%
  mutate(dbh3=ifelse(abs(std.res)<=3, calcDBH, NA)) %>% #create an outlier removed column
  group_by(Tag) %>% #grouping by treeID to calculate increment
  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
         inc=calcDBH-calcDBH_min1,
         inc_annual=(inc/as.numeric(tdif))*365,
         inc.time=year-lag(year),
         #removing increment value if the previous measure was before the last year
         #inc=ifelse(inc.time==1, inc, NA),
         inc_annual=ifelse(inc.time==1, inc_annual, NA),
         has.out=ifelse(1 %in% out, 1, 0),
         treeID=paste0("HKK.", Tag))%>%
  filter(!is.na(inc))%>%
  ungroup()%>%
  filter(!is.na(inc_annual))%>%
  ungroup()%>%
  #filter(start_date <= year & year <= end_date) %>%
  group_by(sp, treeID, year)%>%
  dplyr::summarise(
    diainc = mean(inc_annual, na.rm=T),
    calcDBH_min1 = mean(calcDBH_min1, na.rm=T),
    dbh=mean(calcDBH, na.rm=T),
    out = ifelse(1 %in% out, 1, 0),
    has.out = ifelse(1 %in% has.out, 1, 0), 
    cii=max(Cii, na.rm=T))%>%
  filter(out==0)%>%
  ungroup() %>%
  dplyr::mutate(
    diainc.scaled = scale(diainc),
    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))%>%
  filter(out.inc==0)

```

## Raw diameter increments of all species

```{r echo=FALSE}

library(ggplot2)

growth_yr_withouts<-ggplot(dendro.data_all, aes(x=year, y=diainc, group=year))+
  geom_violin()+
  ylab("Annualised diameter increment (cm)")+theme_bw()

growth_yr_withouts
```

## Removing outliers based (more than 3 SD)

```{r echo=FALSE}

growth_yr_noouts<-ggplot(dendro.data.noouts, aes(x=year, y=diainc, group=year))+ geom_violin()+ ylab("Annualised diameter increment (cm)")+theme_bw()

growth_yr_noouts
```

## Tree ring species

- Four species with tree ring chronologies in HKK

```{r ring trees, echo=F}
#These are species that have been cored near HKK and reported in Zuidema et al., Anderson-Teixeira et al., etc
ring.sp<-c("AFZEXY", "MELIAZ", "TOONCI", "CHUKTA", "NEOLOB")

trees.with.ring<-trees[trees$SPCODE.UPDATE %in% ring.sp,]

dendro.ring<-dendro.data.noouts %>%
  filter(sp %in% ring.sp)

hkk_sp_ring<-trees.with.ring %>%
  group_by(SPCODE.UPDATE)%>%
  dplyr::summarise(n.tree=length(SPCODE.UPDATE))
knitr::kable(hkk_sp_ring, format='html', table.attr="style='width:50%;'")

```

```{r maps of species with tree ring records, echo=F}
library(ggplot2)

pat<-as.character(read.table("git_pat.txt")[1,1])

# remotes:::download("stem1.RData",
#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem1.rdata",
#                    auth_token = pat)
# remotes:::download("stem2.RData",
#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem2.rdata",
#                    auth_token = pat)
# remotes:::download("stem3.RData",
#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem3.rdata",
#                    auth_token = pat)
# remotes:::download("stem4.RData",
#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem4.rdata",
#                    auth_token = pat)
#

load("stem1.RData")
load("stem2.RData")
load("stem3.RData")
load("stem4.RData")

hkk_census<-bind_rows(hkk.stem1, hkk.stem2, hkk.stem3, hkk.stem4)

hkk_census <- hkk_census %>%
  filter(status=="A")

hkk_census_ring_sp<-hkk_census %>%
  filter(CensusID==3)%>%
  filter(sp %in% ring.sp)

hkk_census_sp <- hkk_census %>%
  filter(CensusID==4)%>%
  #group_by(sp, tag, CensusID) %>%
  #dplyr::summarise(tot.agb=sum(agb, na.rm=T)) %>%
  group_by(sp, tag)%>%
  dplyr::summarise(tot.agb=sum(agb, na.rm=T)) %>%
  group_by(sp) %>%
  dplyr::summarise(n.tree=length(unique(tag)),
                   sum.agb=sum(tot.agb))%>%
  ungroup()%>%
  dplyr::mutate(pct.agb=sum.agb*100/sum(sum.agb, na.rm=T))%>%
  arrange(-pct.agb)%>%
  dplyr::mutate(pct.agb.cum=cumsum(pct.agb))

sp_top15<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA",
            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA",
            "GARCSP", "CROTRO", "GARUPI", "NEOLOB", "PHOEPA")

#pct agb of the top 15 dendro species
sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% sp_top15])

ring.sp.map.census<-ggplot(hkk_census_ring_sp, aes(x=gx, y=gy, size=as.numeric(dbh/10), color=factor(sp)))+
  geom_point(alpha=0.5)+
   scale_size(range = c(.1, 10), name="DBH (cm)")+
  guides(color=guide_legend(title="Species"))+
  #scale_colour_gradient(low="orange", high="purple") +
  theme_minimal()+
  coord_fixed()

ring.sp.map.census

ring.sp.map<-ggplot(trees.with.ring, aes(x=X, y=Y, size=as.numeric(DBH.04), color=factor(SPCODE.UPDATE)))+
  geom_point(alpha=0.5)+
   scale_size(range = c(.1, 10), name="DBH (cm)")+
  guides(color=guide_legend(title="Species"))+
  #scale_colour_gradient(low="orange", high="purple") +
  theme_minimal()+
  coord_fixed()

ring.sp.map
```
Species with tree rings contribute `r round(sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% ring.sp]),2)` % aboveground biomass in 2004

## Dendrobands on ring species

```{r echo=F}

growth_sp<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
                                              fill=factor(sp),
                                              color=factor(sp)))+ 
  geom_violin(position = "dodge")+ 
  geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #facet_grid(.~year)+
  #geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
  geom_hline(yintercept = median(dendro.ring$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
      theme(axis.text.x = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  geom_text(aes(label=paste0("median=\n", round(median(dendro.ring$diainc, na.rm=T),2), " cm"), 
                x=0.6, y=median(dendro.ring$diainc, na.rm=T)+0.15), col="black", size=3)+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))

growth_sp

growth_sp_box<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
                                              #color=factor(sp),
                                              fill=factor(sp)))+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #facet_grid(.~year)+
  geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
  geom_hline(yintercept = median(dendro.ring$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
      theme(axis.text.x = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))

#growth_sp_box

growth_yr_sp<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
                                              #color=factor(sp),
                                              fill=factor(sp)))+ 
  #geom_violin(position = "dodge")+ 
  geom_boxplot(position="dodge")+
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  facet_grid(.~year)+
  geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
  theme(axis.text.x = element_blank())+
  guides(color="none")

growth_yr_sp

```

## Detrending by tree with smooth splines

```{r detrending, echo=F}

#trying for one tree

dendro.ring<-dendro.ring %>%
  group_by(sp, treeID) %>%
  dplyr::mutate(n.obs=length(treeID))%>%
  filter(n.obs>=4)%>%
  ungroup()%>%
  group_by(sp, treeID)%>%
  dplyr::mutate(inc.resid=residuals(smooth.spline(x=calcDBH_min1, y=diainc)),
                inc.spline=fitted(smooth.spline(x=calcDBH_min1, y=diainc)))

growth_sp_detrend<-ggplot(dendro.ring, aes(x=factor(sp), 
                                           y=inc.resid, 
                                           #color=factor(sp),
                                           fill=factor(sp)))+ 
  #geom_violin()+ 
  geom_boxplot()+ 
  ylab("Residuals of annualised diameter increment")+
  theme_bw()+
  theme(axis.text.x = element_text(angle=45))
growth_sp_detrend

growth_yr_detrend<-ggplot(dendro.ring, aes(x=year, y=inc.resid, group=year))+ geom_boxplot()+ 
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  theme(axis.text.x = element_text(angle=45))
growth_yr_detrend

```

## Detrended plots by species

```{r echo=F}

growth_yr_sp_detrend<-ggplot(dendro.ring, aes(x=factor(sp), y=inc.resid,
                                              #group=c(year, sp),
                                              #group=sp,
                                              #color=factor(sp),
                                              fill=factor(sp)))+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  geom_hline(yintercept=0)+
  facet_grid(.~year)+
  theme(axis.text.x = element_blank())+
  guides(color="none")

growth_yr_sp_detrend

```

```{r echo=F}

spag_all<-ggplot(dendro.ring, aes(x=year, y=inc.spline, group=treeID)) +
  geom_line()+ggtitle("fitted values (smooth spline)")

spag_all

spag_resid<-ggplot(dendro.ring, aes(x=year, y=inc.resid, group=treeID)) +
  geom_line()+ggtitle("residuals smooth spline")

spag_resid
```
## Detrending all species

```{r detrending all, echo=F}

#trying for one tree

dendro.data.noouts<-dendro.data.noouts %>%
  group_by(sp, treeID) %>%
  dplyr::mutate(n.obs=length(treeID))%>%
  filter(n.obs>=4)%>%
  ungroup()%>%
  filter(treeID!="HKK.92562")%>% #removing an issue with spline for now
  group_by(sp, treeID)%>%
  dplyr::mutate(inc.resid=residuals(smooth.spline(x=calcDBH_min1, y=diainc)),
                inc.spline=fitted(smooth.spline(x=calcDBH_min1, y=diainc)))
```

For `r length(unique(dendro.data.noouts$sp))` species with dendrobands, all individual trees detrended for DBH with smooth splines

```{r echo=F}
growth_yr_detrend_all<-ggplot(dendro.data.noouts, aes(x=year, y=inc.resid, group=year))+ 
  #geom_violin()+ 
  geom_boxplot()+ 
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  theme(axis.text.x = element_text(angle=45))
growth_yr_detrend_all

```

## top 10 species

```{r echo=F}

dendro.by.sp<-dendro.data.noouts %>%
  group_by (sp, year) %>%
  dplyr::summarise(n.inds=length(sp),
    mean.inc = mean(inc.resid, na.rm=T),
                   median.inc = median(inc.resid, na.rm=T),
                   sd.inc = sd(inc.resid, na.rm=T),
    lower=mean.inc-sd.inc,
    upper=mean.inc+sd.inc)%>%
  filter(n.inds>=10)%>%
  arrange(year, -median.inc)

sp_top5<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA")

sp_top10<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA", 
            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA")

dendro.top5<-dendro.by.sp %>%
  filter(sp %in% sp_top5)

dendro.top5.dat<-dendro.data.noouts %>%
  filter(sp %in% sp_top5)

dendro.top10.dat<-dendro.data.noouts %>%
  filter(sp %in% sp_top10)

hkk_census_top10_sp<-hkk_census %>%
  filter(CensusID==3)%>%
  filter(sp %in% sp_top10)

#pct agb of the top 15 dendro species
sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% sp_top15])

top10sp.map.census<-ggplot(hkk_census_top10_sp, aes(x=gx, y=gy, 
                                                    #size=as.numeric(dbh/10),
                                                    color=factor(sp, levels=sp_top10)))+
  geom_point(size=1)+
   scale_size(range = c(.1, 10), name="dbh")+
  guides(color=guide_legend(title="Species"))+
  scale_color_viridis_d()+
  ggtitle("top 10 species in 2004")+
  #scale_colour_gradient(low="orange", high="purple") +
  theme_minimal()+
  coord_fixed()

top10sp.map.census

trees.top10<-trees[trees$SPCODE.UPDATE %in% sp_top10,]

top10sp.map.dendro<-ggplot(trees.top10, aes(x=X, y=Y, 
                                                    #size=as.numeric(dbh/10), 
                                                    color=factor(SPCODE.UPDATE, levels=sp_top10)))+
  geom_point(size=1)+
   scale_size(range = c(.1, 10), name="dbh")+
  guides(color=guide_legend(title="Species"))+
  scale_color_viridis_d()+
  ggtitle("top 10 species in 2004")+
  #scale_colour_gradient(low="orange", high="purple") +
  theme_minimal()+
  coord_fixed()

top10sp.map.dendro

```

## Growth of most abundant species

```{r echo=F}

growth_sp5<-ggplot(dendro.top5.dat, aes(x=factor(sp, levels=sp_top5), y=diainc,
                                              fill=factor(sp, levels=sp_top5),
                                              color=factor(sp, levels=sp_top5)))+ 
  geom_violin(position = "dodge")+ 
    geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top5.dat$diainc, na.rm=T))+
  geom_hline(yintercept = median(dendro.top5.dat$diainc, na.rm=T), lty=2)+
  geom_text(aes(label=paste0("median=\n", round(median(dendro.top5.dat$diainc, na.rm=T),2), " cm"), 
                x=0.6, y=median(dendro.top5.dat$diainc, na.rm=T)+0.15), col="black", size=3)+
  geom_hline(yintercept = 0, lty=3)+
      theme(axis.text.x = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))+
  xlab("")

growth_sp5

growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(sp, levels=sp_top10), y=diainc,
                                              fill=factor(sp, levels=sp_top10),
                                              color=factor(sp, levels=sp_top10)))+ 
  geom_violin(position = "dodge")+ 
    geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_text(aes(label=paste0("median=\n", round(median(dendro.top10.dat$diainc, na.rm=T),2), " cm"), 
                x=0.6, y=median(dendro.top10.dat$diainc, na.rm=T)+0.15), col="black", size=3)+
  geom_hline(yintercept = 0, lty=3)+
      theme(axis.text.x = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))+
  xlab("")

growth_sp10

growth_yr_sp<-ggplot(dendro.top5, aes(x=year, y=mean.inc, 
                                      color=sp))+ 
  geom_point()+
  geom_errorbar(aes(x=year, ymin=lower, ymax=upper))+
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  theme(axis.text.x = element_text(angle=45))+
  facet_grid(sp~.)+
  guides(color="none")+
  geom_hline(yintercept = 0)+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  geom_vline(xintercept=c(2010, 2016), lty=2)
growth_yr_sp

#png("plots/growth_by_sp.png", height=8, width=4, units="in", res=300)
#growth_yr_sp
#dev.off()

```

## Size classes

```{r echo=F}

size_top10<-ggplot(dendro.top10.dat, aes(x=dbh))+
  geom_density()+
  geom_vline(aes(xintercept=35))+
  geom_vline(aes(xintercept=75))

size_top10

#trying three categories
# <50, 50-100, >100

dendro.top10.dat<-dendro.top10.dat %>%
  dplyr::mutate(size=ifelse(dbh<35, "< 35 cm dbh",
                       ifelse(dbh<70, "35-75 cm dbh", ">75 cm dbh")))

growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(sp, levels=sp_top10), y=diainc,
                                              #color=factor(sp),
                                          #group=factor(size, levels=c("small", "medium", "large")),
                                              fill=factor(sp, levels=sp_top10)))+ 
  #geom_violin(position = "dodge")+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
  facet_grid(factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh"))~.)+
      theme(axis.text.y = element_blank())+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))+coord_flip()

growth_sp10

growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh")), y=diainc,
                                              #color=factor(sp),
                                          #group=factor(size, levels=c("small", "medium", "large")),
                                              fill=factor(sp, levels=sp_top10)))+

  geom_violin(position = "dodge")+ 
  geom_boxplot(position = "dodge", col="black", fill="white", width=0.1)+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
  #facet_grid(factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh"))~.)+
      theme(axis.text.x = element_text(angle=45))+
  xlab("")+
  facet_grid(.~factor(sp, levels=sp_top10))+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  ggtitle("HKK dendroband data from 2008 to 2023")+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))#+coord_flip()

growth_sp10

```

## Canopy illumination

```{r echo=F}

cii_top10<-ggplot(dendro.top10.dat, aes(x=cii))+
  geom_histogram()

cii_top10

#trying three categories
# <50, 50-100, >100

#dendro.top10.dat<-dendro.top10.dat %>%
#  dplyr::mutate(size=ifelse(dbh<50, "small",
#                       ifelse(dbh<100, "medium", "large")))

growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
                        aes(x=factor(sp, levels=sp_top10), y=diainc,
                                              #color=factor(sp),
                                          #group=factor(size, levels=c("small", "medium", "large")),
                                              fill=factor(sp, levels=sp_top10)))+ 
  #geom_violin(position = "dodge")+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
  facet_grid(cii~.)+
      theme(axis.text.y = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))+coord_flip()

growth_sp10_cii

growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
                        aes(
                          #x=factor(sp, levels=sp_top10), 
                          x=factor(cii), 
                            y=diainc,
                                              #color=factor(sp),
                                          #group=factor(size, levels=c("small", "medium", "large")),
                                              fill=factor(sp, levels=sp_top10)))+ 
  #geom_violin(position = "dodge")+ 
  geom_boxplot(position = "dodge")+ 
  ylab("Annualised diameter increment (cm)")+theme_bw()+
  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
  geom_hline(yintercept = 0, lty=3)+
  facet_grid(.~factor(sp, levels = sp_top10))+
      theme(axis.text.y = element_blank())+
  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
  scale_fill_viridis_d()+
  scale_color_viridis_d()+
  #coord_flip()+
  guides(fill=guide_legend(title="species"),
         color=guide_legend(title="species"))
growth_sp10_cii

```

## Top 15 species split by leaf habit

```{r evg/dec, echo=F}

sp_top15<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA", 
            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA",
            "GARCSP", "CROTRO", "GARUPI", "NEOLOB", "PHOEPA")

#these 15 species had the most dendrobands and together made up 51.22% of the agb in 2009

#below values eyeballed from Williams et al. 2008 Fig 1.
sp_top15_phenoscore<-c(2.5, 0.8, 1.3, 4, 0.7, 
                       1, 1.25, 1.27, 0.2, 0.3,
                       0.4, 0.65, 3.95, 0.15, 0.25)

#no info on Alphonsea ventricosa in the paper. 
#Apparently evergreen, so used 1 as value - middle of evergreen range

pheno<-as.data.frame(cbind(sp_top15, sp_top15_phenoscore))
colnames(pheno)<-c("sp", "phenoscore")

dendro_top15<-dendro.data.noouts[dendro.data.noouts$sp %in% sp_top15,]
dendro_top15<-merge(dendro_top15, pheno, by="sp")

#categorising as evg and deciduous
dendro_top15$cat<-ifelse(dendro_top15$phenoscore<=2, "evg", "dec")

growth_yr_cat_detrend<-ggplot(dendro_top15, aes(x=factor(cat), y=inc.resid,
                                              #group=c(year, sp),
                                              #group=sp,
                                              fill=factor(cat)))+ 
  geom_violin(position = "dodge")+ 
  ylab("Residuals of annualised diameter increment")+theme_bw()+
  facet_grid(.~year)+
  geom_hline(yintercept = 0)+
  scale_fill_manual(values=c("sienna4", "darkgreen"))+
  theme(axis.text.x = element_blank())

growth_yr_cat_detrend

```

- identified and removed growth release
- hand pick high quality time series (10-20 per species maybe)
- try VH's model - temporal autocorr
- mean and median growth 
- by month plots - time difference
- different rates

### 2024-06-20 doc/manuscript.Rmd
draft hypotheses
@teixeirak - draft of hypotheses to test for the tree growth vulnerability to drought paper

@@
 
+Here we use a 14-year record of dendrometer band measurements in seasonally dry forest in Thailand to test the vulnerability of tropical tree growth to drought.
+We expect variation in the absolute growth during drought years and drought resistance (growth in drought year/growth in a previous year with normal rainfall) across trees.
+We hypothesise that growth during drought is affected by:
+- **habitat** : water availability is a key driver of tropical tree growth [@wagner_water_2012].
+We hypothesise that drought resistance higher in trees growing in low-lying areas because of closer access to ground water reserves, and lower in upland trees.
+- **leaf habit** : drought resistance is higher in deciduous species because of leaf strategies that minimise water loss during months of high vulnerability.
+Deciduous and evergreen species have differential sensitivity to drought [@desouza_drought_2020].
+- **tree size** : larger trees face higher risk of mortality during drought, but potentially suffer less growth vulnerability.  
+- **exposure** : trees with higher exposure because of their canopy position have lower drought resistance than trees with lower exposure due to the direct effects of temperature and vapour pressure deficit.
+- **competition** : trees in denser stands have lower drought resistance than trees in sparser stands because of more intense competition for groundwater.
+Differences in rooting depths could add complexities to this effect, however, we do not have direct measures that could test this effect.
+- **exposure x size** :  While larger trees are likely to have uniform high exposure, smaller trees can have high or low canopy position based on stand characteristics (stand density, presence of a canopy gap etc.) Drought resistance of smaller trees with high exposure is expected to be lower than that of trees with low exposure because of the direct influence of temperature and light that could lead to cavitation.
+(BCI light x size interaction in @rüger_growth_2011)
+- **leaf habit x exposure** : species canopy strategies along with their leaf habit could exacerbate or counteract drought vulnerability [@rahman_speciesspecific_2019].
+Under high exposure, deciduous species are potentially more drought resistant than evergreen species, while the pattern could be less clear under low exposure.
+- **leaf habit x habitat** : tropical evergreen and deciduous species have different habitat preferences [@kunert_leaf_2021] that could affect their drought resistance.   
+- **habitat x exposure** : trees with high exposure in upland habitats are expected to have lowest drought resistance because of the compounding influence of abiotic stressors.
+
 # Methods {-}

### 2024-06-22 doc/manuscript.Rmd
add some into notes
@krishnaanujan , FYI.
If you give me access to a shared Zotero library I can add all these refs.

@@
 
+- Large trees tend to undergo greater growth declines during drought compared to smaller trees [e.g., @bennett_larger_2015; @mcgregor_tree_2021; @anderson-teixeira_joint_2022]. 
+- It remains poorly understood the extent to which this is shaped by tree size itself, crown exposure, water access, and traits that tend to covary with size (e.g., decidiousness) -- all of which interact to shape drought resistance
+
+- There is theory and evidence that tree size itself matters.
+- Theoretically, we expect that greater height makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium, as described by Darcy's law [@mcdowell_darcy_2015; @fernández-de-uña_role_inpress].
+- Height... [see refs in @vinod_thermal_2023; @olson_plant_2018; @couvreurWaterTransportTall2018].
+- As tree size increases, leaves exert lower control over hydraulic resistance [@wolfe_leaves_2023].
+- @chen_hydraulic_2022
+
+- There is theory and evidence that crown exposure matters [ @vinod_thermal_2023;  @refs_in_ vinod_thermal_2023; @scharnweber_confessions_2019].
+- Microclimate buffering leads to cooler, moister understory air [@vinod_thermal_2023].
+- Soils under closed canopies would also be cooler during hot times of the year [@lembrechts_global_2022].
+- Reduced evaporative demand would also make them moister, and this might be added to by hydraulic redistribution.
+- Trees with exposed crowns suffered significant crown dieback at greater rates in the 2012-16 CA drought [@ma_tree_2023].
+
+- Water access....
+- Larger trees have larger root systems, but do not necessarily access deeper water [@ref_from_Panama]
+- Even when trees are accessing deeper water, this does not mean that they're in better shape during drought.
+Rather, trees that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@rutuja]
+- indeed, there is evidence that trees near streams undergo greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta] during drought
+- rather than necessarily helping during drought, water access will shape the size the size and traits of species living in habitat, with stream habitats tending to have larger trees [@ref] and more evergreen trees
+
+- Many hydraulic-related traits vary with tree height [ @vinod_thermal_2023], including the frequency of dry season deciduous leaf loss -- both within species and at the community level [@condit_ref; @meakem].
+- (discuss drought strategies - avoidance / adaptation )
+
 Here we use a 14-year record of dendrometer band measurements in seasonally dry forest in Thailand to test the vulnerability of tropical tree growth to drought.
@@
 Deciduous and evergreen species have differential sensitivity to drought [@desouza_drought_2020].
-- **tree size** : larger trees face higher risk of mortality during drought, but potentially suffer less growth vulnerability.  
+- **tree size** : larger trees face higher risk of mortality during drought, as well as greater growth reductions [e.g., @bennett_larger_2015; @mcgregor_tree_2021; @anderson-teixeira_joint_2022].  
 - **exposure** : trees with higher exposure because of their canopy position have lower drought resistance than trees with lower exposure due to the direct effects of temperature and vapour pressure deficit.

### 2024-06-25 doc/manuscript.Rmd
Update manuscript.Rmd
@krishnaanujan - a couple other additions that I forgot to save/push

@@
 - indeed, there is evidence that trees near streams undergo greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta] during drought
-- rather than necessarily helping during drought, water access will shape the size the size and traits of species living in habitat, with stream habitats tending to have larger trees [@ref] and more evergreen trees
+- rather than necessarily helping during drought, water access will shape the size the size and traits of species living in habitat, with stream habitats tending to have larger trees [@ref] and more evergreen trees, and also their average growth rate
 
@@
 - **habitat** : water availability is a key driver of tropical tree growth [@wagner_water_2012]. 
-We hypothesise that drought resistance higher in trees growing in low-lying areas because of closer access to ground water reserves, and lower in upland trees.
 - **leaf habit** : drought resistance is higher in deciduous species because of leaf strategies that minimise water loss during months of high vulnerability.

### 2024-06-25 doc/manuscript.Rmd
conflicts resolved manuscript

--- 2024-06-22 doc/manuscript.Rmd
@@
 
+<!---KAT framing---->
+
+-High Crown exposure makes trees more vulnerable to drought, but drought deciduous habit or perennial water access allows them to escape this  
+-Disentangling the effects of tree size, crown exposure, water access, and deciduousness on drought sensitivity
+
 Here we use a 14-year record of dendrometer band measurements in seasonally dry forest in Thailand to test the vulnerability of tropical tree growth to drought.
@@
 We hypothesise that drought resistance higher in trees growing in low-lying areas because of closer access to ground water reserves, and lower in upland trees.
+
+<!----KAT - wetter habitats have higher growth reductions - because used to more water---->
+
 - **leaf habit** : drought resistance is higher in deciduous species because of leaf strategies that minimise water loss during months of high vulnerability.
@@
 
+***Models***
+
 # Results {-}

### 2024-06-25 doc/manuscript.Rmd

--- 2024-06-25 doc/manuscript.Rmd
@@
 We hypothesise that growth during drought is affected by:
+
+<!----KAT - wetter habitats have higher growth reductions - because used to more water---->
+
 - **habitat** : water availability is a key driver of tropical tree growth [@wagner_water_2012]. 

### 2024-06-25 scripts/explore_analysis.Rmd
data summary plots

@@
 
-## Climate Sensitivity of Tropical Tree Growth
-
 ## HKK climate
@@
 #reading the GitHub PAT (stored privately and locally)
-pat<-as.character(read.table("git_pat.txt")[1,1])
-
-#remotes:::download("DendroByBand.RData",
+pat<-as.character(read.table("../data/HKK-dendro/git_pat.txt")[1,1])
+
+# remotes:::download("DendroByBand.RData",
 #                    "https://raw.githubusercontent.com/forestgeo/HKK-dendrobands/main/data/modified_data/DendroByBand.RData",
@@
 
-load("DendroByBand.RData")
+load("../data/HKK-dendro/DendroByBand.RData")
 
@@
 
-req <- GET("https://api.github.com/repos/forestgeo/Climate/git/trees/master?recursive=1")
-stop_for_status(req)
-filelist <- unlist(lapply(content(req)$tree, "[", "path"), use.names = F)
-path<-grep("Climate_Data/CRU/CRU_v4_06", filelist, value = TRUE, fixed = TRUE)
-datacsvs<-grep("pre|tmx", path, value = T)
-
-#datacsvs<-datacsvs[!grepl("BLANK", datacsvs)]
-datacsvs<-paste0("https://github.com/forestgeo/Climate/raw/master/", datacsvs)
-
-start_date<-min(hkk_OBSID$year,na.rm=T)
-end_date<-max(hkk_OBSID$year,na.rm=T)
-
-climate_data_orig<-bind_rows(datacsvs %>%
-                               read_csv(na = c("", "NA", "N/A", "NULL"), show_col_types = FALSE))%>%
-  clean_names() %>%
-  filter(sites_sitename %in% c("HKK",
-                               "Huai Kha Khaeng")) %>%
-  select(-c(sites_sitename, n_obs)) %>%
-  rename(year=yr,
-         month=mo)%>%
-  filter(start_date-1 <= year & year <= end_date) # limiting to specified date range but extending to a year before
+#CHIRPS data ----
+
+# CHIRPS data GEE code here - https://code.earthengine.google.com/894597bd093906fb0ff82320f29c0842
+
+chirps_hkk<-read.csv("../data/climate/CHIRPS_from_GEE.csv")
+colnames(chirps_hkk)[1]<-"time"
+
+#making dates manageable
+chirps_hkk<- chirps_hkk %>%
+  dplyr::mutate(time=as.Date(time, format = "%b %d, %Y"),
+                year=year(time),
+                month=month(time))
 
@@
 pre_yr_census<-ggplot()+
-  geom_area(data=climate_data_orig[climate_data_orig$variable=="pre",], aes(x=as.Date(paste0(year, "-", month, "-", "01"), format = "%Y-%m-%d"), y=value), fill="blue", alpha=0.5)+
-  geom_line(data=climate_data_orig[climate_data_orig$variable=="tmx",], aes(x=as.Date(paste0(year, "-", month, "-", "01"), format = "%Y-%m-%d"), y=value*5), col="red")+
+  geom_area(data=chirps_hkk, aes(x=time, y=precipitation), fill="blue", alpha=0.5)+
+  #geom_line(data=climate_data_orig[climate_data_orig$variable=="tmx",], aes(x=as.Date(paste0(year, "-", month, "-", "01"), format = "%Y-%m-%d"), y=value*5), col="red")+
   geom_vline(data=census_int[census_int$type=="wet",], aes(xintercept=start), col="darkgreen")+
@@
   geom_vline(data=census_int[census_int$type=="dry",], aes(xintercept=end), col="brown4")+
-  scale_y_continuous(    # Features of the first axis
-    name = "monthly rainfall (mm)",
-
-    # Add a second axis and specify its features
-    sec.axis = sec_axis(~.*0.2, name="maximum temperature (Celsius °)"))+
+  # scale_y_continuous(    # Features of the first axis
+  #   name = "monthly rainfall (mm)",
+  #   
+  #   # Add a second axis and specify its features
+  #   sec.axis = sec_axis(~.*0.2, name="maximum temperature (Celsius °)"))+
   xlab("Date")+theme_bw()+
-  ggtitle("HKK dendroband censuses and CRU climate data")+
+  ggtitle("HKK dendroband censuses and CHIRPS rainfall data")+
   #scale_x_continuous(n.breaks=15)+
@@
 pre_yr_census
+
+#rainfall per year
+
+#-----------------Decisions on climate data-----------------------=
+# Summarising climate data to relevant time window for SCBI
+# As per issue #42 and KAT pub https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/gcb.15934
+# HKK best climate variables are:
+# PPT - previous Sep to current June (it is actually precipitation day freq, but using sum now)
+# Tmax - current Apr to Oct
+#---------------------------------------------------------------=
+
+#applying these decisions and rescaling
+
+#mean annual precip from 1981 to 2023 is 1698.4
+
+#code - https://code.earthengine.google.com/b0eda787f2e1ba578bbf50a8a9d04843
+
+climate_data_yr<-chirps_hkk %>%
+  group_by(year) %>%
+  summarise(
+    tot_precip=sum(precipitation),
+    pre_SeptoDec=sum(precipitation[month>=9]), #Sep to Dec precip for each year
+    pre_JantoJun=sum(precipitation[month<=6]) #Jan to June precip for each year
+    #tmx_window=sum(value[variable == "tmx"& month %in% 4:10])
+)%>% #sum of potential evapotranspiration for current May to July
+  mutate(
+    pre_window=pre_JantoJun+lag(pre_SeptoDec), #adding prev year's June to Dec and current year's Jan to June
+    tot_pre_diff=tot_precip-1698.4
+    )
+
+#standard deviation is 254.06
+
+precip_plot<-ggplot(climate_data_yr)+
+  geom_bar(aes(x=year, y=tot_pre_diff), stat="identity")+
+  ylab("difference from 40-year mean (mm)")+
+  geom_hline(yintercept = c(254.06, -254.06), lty=2)+
+  theme_bw()
+precip_plot
+
+# library
+library(ggridges)
+library(ggplot2)
+library(viridis)
+library(hrbrthemes)
+
+chirps_hkk$year_fac<-as.character(chirps_hkk$year)
+chirps_hkk$year_fac<-factor(chirps_hkk$year)
+
+# Plot
+ggplot(chirps_hkk, aes(x = precipitation, y = year_fac, 
+                       fill = stat(x))) +
+  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
+  scale_fill_viridis(name = "rainfall", option = "C") +
+  labs(title = 'Rainfall in HKK') +
+  #theme_ipsum() +
+  theme_bw()+
+    theme(
+      legend.position="none",
+      panel.spacing = unit(0.1, "lines"),
+      strip.text.x = element_text(size = 8)
+    )
 
@@
 
+# species_all<-dendro.data.noouts %>%
+#   group_by(sp, treeID) %>%
+#   dplyr::summarise(n.obs=length(treeID))%>%
+#   ungroup() %>%
+#   group_by(sp) %>%
+#   dplyr::summarise(n.tree=length(treeID),
+#                    n.obs.tot=sum(n.obs))%>%
+#   ungroup()%>%
+#   arrange(-n.tree)
+# 
+# write.csv(species_all, "../data/HKK-dendro/species.csv", row.names = F)
+
 ```
@@
 
-```{r echo=FALSE}
+```{r echo=FALSE, eval=F}
 
@@
 
-## Tree ring species
-
-- Four species with tree ring chronologies in HKK
-
-```{r ring trees, echo=F}
-#These are species that have been cored near HKK and reported in Zuidema et al., Anderson-Teixeira et al., etc
-ring.sp<-c("AFZEXY", "MELIAZ", "TOONCI", "CHUKTA", "NEOLOB")
-
-trees.with.ring<-trees[trees$SPCODE.UPDATE %in% ring.sp,]
-
-dendro.ring<-dendro.data.noouts %>%
-  filter(sp %in% ring.sp)
-
-hkk_sp_ring<-trees.with.ring %>%
-  group_by(SPCODE.UPDATE)%>%
-  dplyr::summarise(n.tree=length(SPCODE.UPDATE))
-knitr::kable(hkk_sp_ring, format='html', table.attr="style='width:50%;'")
-
-```
-
-```{r maps of species with tree ring records, echo=F}
+```{r maps of species with tree ring records, echo=F, eval=F}
 library(ggplot2)
@@
 
-ring.sp.map.census<-ggplot(hkk_census_ring_sp, aes(x=gx, y=gy, size=as.numeric(dbh/10), color=factor(sp)))+
-  geom_point(alpha=0.5)+
-   scale_size(range = c(.1, 10), name="DBH (cm)")+
-  guides(color=guide_legend(title="Species"))+
-  #scale_colour_gradient(low="orange", high="purple") +
-  theme_minimal()+
-  coord_fixed()
-
-ring.sp.map.census
-
-ring.sp.map<-ggplot(trees.with.ring, aes(x=X, y=Y, size=as.numeric(DBH.04), color=factor(SPCODE.UPDATE)))+
-  geom_point(alpha=0.5)+
-   scale_size(range = c(.1, 10), name="DBH (cm)")+
-  guides(color=guide_legend(title="Species"))+
-  #scale_colour_gradient(low="orange", high="purple") +
-  theme_minimal()+
-  coord_fixed()
-
-ring.sp.map
-```
-Species with tree rings contribute `r round(sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% ring.sp]),2)` % aboveground biomass in 2004
-
-## Dendrobands on ring species
-
-```{r echo=F}
-
-growth_sp<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
-                                              fill=factor(sp),
-                                              color=factor(sp)))+ 
-  geom_violin(position = "dodge")+ 
-  geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #facet_grid(.~year)+
-  #geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
-  geom_hline(yintercept = median(dendro.ring$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-      theme(axis.text.x = element_blank())+
-  ggtitle("HKK dendroband data from 2008 to 2023")+
-  geom_text(aes(label=paste0("median=\n", round(median(dendro.ring$diainc, na.rm=T),2), " cm"), 
-                x=0.6, y=median(dendro.ring$diainc, na.rm=T)+0.15), col="black", size=3)+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))
-
-growth_sp
-
-growth_sp_box<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
-                                              #color=factor(sp),
-                                              fill=factor(sp)))+ 
-  geom_boxplot(position = "dodge")+ 
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #facet_grid(.~year)+
-  geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
-  geom_hline(yintercept = median(dendro.ring$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-      theme(axis.text.x = element_blank())+
-  ggtitle("HKK dendroband data from 2008 to 2023")+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))
-
-#growth_sp_box
-
-growth_yr_sp<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
-                                              #color=factor(sp),
-                                              fill=factor(sp)))+ 
-  #geom_violin(position = "dodge")+ 
-  geom_boxplot(position="dodge")+
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  facet_grid(.~year)+
-  geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
-  theme(axis.text.x = element_blank())+
-  guides(color="none")
-
-growth_yr_sp
-
-```
-
-## Detrending by tree with smooth splines
-
-```{r detrending, echo=F}
-
-#trying for one tree
-
-dendro.ring<-dendro.ring %>%
-  group_by(sp, treeID) %>%
-  dplyr::mutate(n.obs=length(treeID))%>%
-  filter(n.obs>=4)%>%
-  ungroup()%>%
-  group_by(sp, treeID)%>%
-  dplyr::mutate(inc.resid=residuals(smooth.spline(x=calcDBH_min1, y=diainc)),
-                inc.spline=fitted(smooth.spline(x=calcDBH_min1, y=diainc)))
-
-growth_sp_detrend<-ggplot(dendro.ring, aes(x=factor(sp), 
-                                           y=inc.resid, 
-                                           #color=factor(sp),
-                                           fill=factor(sp)))+ 
-  #geom_violin()+ 
-  geom_boxplot()+ 
-  ylab("Residuals of annualised diameter increment")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle=45))
-growth_sp_detrend
-
-growth_yr_detrend<-ggplot(dendro.ring, aes(x=year, y=inc.resid, group=year))+ geom_boxplot()+ 
-  ylab("Residuals of annualised diameter increment")+theme_bw()+
-  theme(axis.text.x = element_text(angle=45))
-growth_yr_detrend
-
-```
-
-## Detrended plots by species
-
-```{r echo=F}
-
-growth_yr_sp_detrend<-ggplot(dendro.ring, aes(x=factor(sp), y=inc.resid,
-                                              #group=c(year, sp),
-                                              #group=sp,
-                                              #color=factor(sp),
-                                              fill=factor(sp)))+ 
-  geom_boxplot(position = "dodge")+ 
-  ylab("Residuals of annualised diameter increment")+theme_bw()+
-  geom_hline(yintercept=0)+
-  facet_grid(.~year)+
-  theme(axis.text.x = element_blank())+
-  guides(color="none")
-
-growth_yr_sp_detrend
-
-```
-
-```{r echo=F}
-
-spag_all<-ggplot(dendro.ring, aes(x=year, y=inc.spline, group=treeID)) +
-  geom_line()+ggtitle("fitted values (smooth spline)")
-
-spag_all
-
-spag_resid<-ggplot(dendro.ring, aes(x=year, y=inc.resid, group=treeID)) +
-  geom_line()+ggtitle("residuals smooth spline")
-
-spag_resid
-```
+```
+
 ## Detrending all species
@@
 ```{r detrending all, echo=F}
-
-#trying for one tree
 
@@
 top10sp.map.dendro
+
+```
+## Fastest and slowest in 2015 and 2016
+
+```{r echo=F}
+
+dendro.data.noouts <- dendro.data.noouts %>%
+  group_by(year) %>% 
+  dplyr::mutate(growth_quantile = ntile(diainc, 4),
+                resid_quantile = ntile (inc.resid, 4))
+
+growth_2015 <- dendro.data.noouts %>%
+  filter(year==2015)
+
+quantiles_2015<-quantile(growth_2015$diainc, probs=c(0.25, 0.5, 0.75, 1))
+
+growth_2015_plot<-ggplot(growth_2015, aes(x=diainc)) +
+  geom_density(fill="brown", col="brown", alpha=0.5) +
+  geom_vline(xintercept=quantiles_2015, col="brown")+ theme_bw()
+
+growth_2014 <- dendro.data.noouts %>%
+  filter(year==2014)
+
+quantiles_2014<-quantile(growth_2014$diainc, probs=c(0.25, 0.5, 0.75, 1))
+
+growth_2016 <- dendro.data.noouts %>%
+  filter(year==2016)
+
+quantiles_2016<-quantile(growth_2016$diainc, probs=c(0.25, 0.5, 0.75, 1))
+
+growth_2016_plot<-ggplot(growth_2016, aes(x=diainc)) +
+  geom_density(fill="brown", col="brown", alpha=0.5) +
+  geom_vline(xintercept=quantiles_2016, col="brown")+ theme_bw()
+
+growth_2017 <- dendro.data.noouts %>%
+  filter(year==2017)
+
+quantiles_2017<-quantile(growth_2017$diainc, probs=c(0.25, 0.5, 0.75, 1))
+
+growth_2017_plot<-ggplot(growth_2017, aes(x=diainc)) +
+  geom_density(fill="brown", col="brown", alpha=0.5) +
+  geom_vline(xintercept=quantiles_2016, col="brown")+ theme_bw()
+
+growth_2014_2015_plot<-ggplot() +
+  geom_density(data = growth_2014, aes(x=diainc), fill="forestgreen", col="forestgreen", alpha=0.5) +
+  geom_vline(xintercept=quantiles_2014, col="forestgreen")+
+  geom_density(data= growth_2015, aes (x=diainc), fill="brown", col="brown", alpha=0.5) +
+  geom_text(aes(x=0.75, y=1, label="2014"), col="forestgreen", size=3)+
+  geom_text(aes(x=0.75, y=1.5, label="2015"), col="brown", size=3)+
+  geom_vline(xintercept=quantiles_2015, col="brown")+ 
+  xlab("diameter increment (cm)") + ylab("density")+
+  theme_bw()
+
+dendro.data.noouts$year_fac<-as.character(dendro.data.noouts$year)
+dendro.data.noouts$year_fac<-factor(dendro.data.noouts$year_fac, levels=as.character(rev(2009:2023)))
+
+growth_all_years<-ggplot(dendro.data.noouts, aes(x = diainc, y = year_fac, 
+                       fill = stat(x))) +
+  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
+  scale_fill_viridis(name = "growth", option = "H", direction=-1) +
+  labs(title = 'tree growth in HKK (cm)', y="year") +
+  #theme_ipsum() +
+  geom_vline(xintercept=0, lty=2)+
+  theme_bw()+
+    theme(
+      legend.position="none",
+      panel.spacing = unit(0.1, "lines"),
+      strip.text.x = element_text(size = 8)
+    )
+
+growth_all_years
+
+png("../doc/display/growth_2014_2015.png", width=4, height=4, units="in", res=300)
+growth_2014_2015_plot
+dev.off()
+
+png("../doc/display/growth_allyrs.png", width=4, height=4, units="in", res=300)
+growth_all_years
+dev.off()
+
+```
+
+##Distribution of data with size, exposure and deciduousness
+
+```{r echo=F}
+
+#2014 data distribution with size, exposure and deciduousness
+
+png("../doc/display/dbh_cii.png", width=4, height=4, units="in", res=300)
+ggplot(growth_2014, aes(x = dbh, y = factor(cii), 
+                       fill = stat(x))) +
+  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
+  scale_fill_viridis(name = "growth", option = "C", direction=-1) +
+  labs(title = 'DBH in 2014 (cm)', y="canopy illumination") +
+  #theme_ipsum() +
+  theme_bw()+
+    theme(
+      legend.position="none",
+      panel.spacing = unit(0.1, "lines"),
+      strip.text.x = element_text(size = 8)
+    )
+
+length(unique(growth_2014$sp))
+
+png("../doc/display/growth_cii.png", width=4, height=4, units="in", res=300)
+ggplot(growth_2014, aes(x = diainc, y = factor(cii), 
+                       fill = stat(x))) +
+  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
+  scale_fill_viridis(name = "growth", option = "C", direction=-1) +
+  labs(title = 'growth in 2014 (cm)', y="canopy illumination") +
+  #theme_ipsum() +
+  theme_bw()+
+    theme(
+      legend.position="none",
+      panel.spacing = unit(0.1, "lines"),
+      strip.text.x = element_text(size = 8)
+    )
+dev.off()
 

### 2024-06-27 scripts/explore_analysis.Rmd
data summaries; resistance for 2010 and 2014

@@
 ---
-title: "Climate Sensitivity of Tropical Tree Growth"
+title: "Drought vulnerability of tree growth"
 author: ""
@@
 #this can be refined if needed
-wet_census<-c(1, 3, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24, 26, 28)
+#wet_census<-c(1, 3, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24, 26, 28)
 
@@
   group_by(Cno) %>%
-  dplyr::summarise(start=min(Date2, na.rm = T),
-                   end=max(Date2, na.rm=T))%>%
-  dplyr::mutate(type=ifelse(Cno %in% wet_census, "wet", "dry"))
+  dplyr::summarise(start=as.Date(min(Date2, na.rm = T)),
+                   end=as.Date(max(Date2, na.rm=T)))#%>%
+  #dplyr::mutate(type=ifelse(Cno %in% wet_census, "wet", "dry"))
 
@@
   #geom_line(data=climate_data_orig[climate_data_orig$variable=="tmx",], aes(x=as.Date(paste0(year, "-", month, "-", "01"), format = "%Y-%m-%d"), y=value*5), col="red")+
-  geom_vline(data=census_int[census_int$type=="wet",], aes(xintercept=start), col="darkgreen")+
-  geom_vline(data=census_int[census_int$type=="wet",], aes(xintercept=end), col="darkgreen")+
-  geom_vline(data=census_int[census_int$type=="dry",], aes(xintercept=start), col="brown4")+
-  geom_vline(data=census_int[census_int$type=="dry",], aes(xintercept=end), col="brown4")+
+  geom_rect(data=census_int, aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf), 
+              fill = "grey10", alpha = 0.5) +
+  geom_text(data=census_int, aes(x=start, y=500, 
+                                 label= paste0("C", 1:29), angle=90), size=2)+
   # scale_y_continuous(    # Features of the first axis
@@
   #   sec.axis = sec_axis(~.*0.2, name="maximum temperature (Celsius °)"))+
-  xlab("Date")+theme_bw()+
+  xlab("")+theme_bw()+ ylab("total monthly precipitation (mm)")+
   ggtitle("HKK dendroband censuses and CHIRPS rainfall data")+
@@
 
+png("../doc/display/census_precip.png", width=6, height=3, res=300, units="in") 
 pre_yr_census
+dev.off()
 
@@
   #filter(start_date <= year & year <= end_date) %>%
-  filter(Cno %in% wet_census)%>%
+  #filter(Cno %in% wet_census)%>%
+  filter(is.na(DBH1_flag), DBH1_error=="none",
+         win.decimal.error=="none", is.na(win.flag))%>%
+  group_by(Cno, sp, Tag) %>%
+  dplyr::summarise(calcDBH=calcDBH[which(n.dendro==min(n.dendro))],
+                   Date2=Date2[which(n.dendro==min(n.dendro))],
+                   Cii=max(Cii, na.rm=T))
+
+#first fitting a line through the time series
+obs.model.dbh<-nlme::lme(calcDBH~Cno, random=~1|Tag, data=dendro.data,
+                         na.action=na.omit,
+                         method="ML", control=(msMaxIter=100)) #had to increase the number of iterations
+
+#standardised residuals calculated https://stat.ethz.ch/R-manual/R-devel/library/nlme/html/residuals.lme.html
+
+std.res.dbh<-residuals(obs.model.dbh, type="pearson")
+hist(std.res.dbh)
+dendro.data$std.res<-rep(NA, nrow(dendro.data))
+dendro.data$std.res[!is.na(dendro.data$calcDBH)]<-std.res.dbh
+#defining all observation with abs(residual) >3 as an outlier
+dendro.data$out<-factor(ifelse(abs(dendro.data$std.res)<=3, 0, 1))
+
+#find the trees which have outlier observations
+outs.Tag<-unique(dendro.data$Tag[dendro.data$out==1])
+
+#plot trees that have outliers
+ggplot(dendro.data[dendro.data$Tag %in%outs.Tag,],
+       aes(x = Cno, y = calcDBH, col=factor(out), group = Tag))+
+  geom_line(alpha=0.5)+geom_point()+
+  scale_color_manual(values=c("grey20", "red", "grey"))
+
+dendro.data.noouts<-dendro.data %>%
+  filter(abs(std.res)<=3)%>%
+  select(-c(std.res, out))
+  #mutate(dbh3=ifelse(abs(std.res)<=3, calcDBH, NA)) #%>% #create an outlier removed column
+
+```
+```{r calculating increments for specific years}
+
+dendro_2010<- dendro.data.noouts %>%
+  filter(Cno %in% c(1, 3, 5))%>%
+  group_by(Tag) %>% #grouping by treeID to calculate increment
+  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
+         cii_min1=dplyr::lag(Cii, n=1, default=NA, order_by=Tag),
+         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
+         inc=calcDBH-calcDBH_min1,
+         inc_annual=(inc/as.numeric(tdif))*365,
+         inc.time=year(Date2)-lag(year(Date2)),
+         #removing increment value if the previous measure was before the last year
+         #inc=ifelse(inc.time==1, inc, NA),
+         inc_annual=ifelse(inc.time==1, inc_annual, NA),
+         treeID=paste0("HKK.", Tag))%>%
+  filter(!is.na(inc))%>%
+  filter(!is.na(inc_annual))%>%
+  ungroup()%>%
+  dplyr::mutate(
+    diainc.scaled = scale(inc_annual),
+    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))#%>%
+  #filter(out.inc==0)
+
+quantiles_2009<-quantile(dendro_2010$inc_annual[dendro_2010$Cno==3], 
+                         probs=c(0.25, 0.5, 0.75, 1))
+
+quantiles_2010<-quantile(dendro_2010$inc_annual[dendro_2010$Cno==5], 
+                         probs=c(0.25, 0.5, 0.75, 1))
+
+dendro_2010_plot<-ggplot(data = dendro_2010, aes(x=inc_annual, group=factor(Cno), fill=factor(Cno), col=factor(Cno))) +
+  geom_density(alpha=0.5) +
+  scale_color_manual(values=c("forestgreen", "brown"))+
+  scale_fill_manual(values=c("forestgreen", "brown"))+
+  geom_vline(xintercept=quantiles_2009, col="forestgreen")+
+  geom_text(aes(x=0.75, y=1, label="2009"), col="forestgreen", size=3)+
+  geom_text(aes(x=0.75, y=1.5, label="2010"), col="brown", size=3)+
+  geom_vline(xintercept=quantiles_2010, col="brown")+ 
+  xlab("diameter increment (cm)") + ylab("density")+
+  guides(col="none", fill="none")+
+  theme_bw()
+
+dendro_2015<- dendro.data.noouts %>%
+  filter(Cno %in% c(11, 13, 15))%>%
+  group_by(Tag) %>% #grouping by treeID to calculate increment
+  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
+         cii_min1=dplyr::lag(Cii, n=1, default=NA, order_by=Tag),
+         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
+         inc=calcDBH-calcDBH_min1,
+         inc_annual=(inc/as.numeric(tdif))*365,
+         inc.time=year(Date2)-lag(year(Date2)),
+         #removing increment value if the previous measure was before the last year
+         #inc=ifelse(inc.time==1, inc, NA),
+         inc_annual=ifelse(inc.time==1, inc_annual, NA),
+         treeID=paste0("HKK.", Tag))%>%
+  filter(!is.na(inc))%>%
+  filter(!is.na(inc_annual))%>%
+  ungroup()%>%
+  dplyr::mutate(
+    diainc.scaled = scale(inc_annual),
+    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))%>%
+  filter(out.inc==0)
+
+quantiles_2014<-quantile(dendro_2015$inc_annual[dendro_2015$Cno==13], 
+                         probs=c(0.25, 0.5, 0.75, 1))
+
+quantiles_2015<-quantile(dendro_2015$inc_annual[dendro_2015$Cno==15], 
+                         probs=c(0.25, 0.5, 0.75, 1))
+
+dendro_2015_plot<-ggplot(data = dendro_2015, aes(x=inc_annual, group=factor(Cno), fill=factor(Cno), col=factor(Cno))) +
+  geom_density(alpha=0.5) +
+  scale_color_manual(values=c("forestgreen", "brown"))+
+  scale_fill_manual(values=c("forestgreen", "brown"))+
+  geom_vline(xintercept=quantiles_2014, col="forestgreen")+
+  geom_text(aes(x=0.75, y=1, label="2014"), col="forestgreen", size=3)+
+  geom_text(aes(x=0.75, y=1.5, label="2015"), col="brown", size=3)+
+  geom_vline(xintercept=quantiles_2015, col="brown")+ 
+  xlab("diameter increment (cm)") + ylab("density")+
+  guides(col="none", fill="none")+
+  theme_bw()
+
+library(gridExtra)
+png("../doc/display/increments_raw.png", width=8, height=4, units="in", res=300)
+grid.arrange(dendro_2010_plot, dendro_2015_plot, nrow=1)
+dev.off()
+
+```
+```{r drought resistance}
+
+dec_williams<-read.csv("../data/HKK-dendro/species.csv")
+
+dec_williams<-dec_williams %>%
+  filter(n.tree >= 5)
+
+resistance_2010 <- dendro_2010 %>%
+  group_by(Tag) %>% #grouping by treeID to calculate increment
+  mutate(inc_min1=dplyr::lag(inc_annual, n=1, default=NA, order_by=Tag),
+         resistance.div = inc_annual/inc_min1,
+         resistance.dif = inc_annual-inc_min1)%>%
+  filter(!is.na(inc_min1), !is.infinite(resistance.div))
+
+resistance_2010<-merge(resistance_2010, dec_williams, by="sp", all.x=T)
+
+resistance_2010_div<-ggplot(data = resistance_2010, aes(x=resistance.div)) +
+  geom_density(alpha=0.5) +
+  theme_bw()
+
+resistance_2010_dif<-ggplot(data = resistance_2010, aes(x=resistance.dif)) +
+  geom_density(alpha=0.5) +
+  theme_bw()
+
+resistance_2015 <- dendro_2015 %>%
+  group_by(Tag) %>% #grouping by treeID to calculate increment
+  mutate(inc_min1=dplyr::lag(inc_annual, n=1, default=NA, order_by=Tag),
+         resistance.div = inc_annual/inc_min1,
+         resistance.dif = inc_annual-inc_min1)%>%
+  filter(!is.na(inc_min1), !is.infinite(resistance.div))
+
+resistance_2015<-merge(resistance_2015, dec_williams, by="sp", all.x=T)
+
+resistance_2015_div<-ggplot(data = resistance_2015, aes(x=resistance.div)) +
+  geom_density(alpha=0.5) +
+  theme_bw()
+
+resistance_2015_dif<-ggplot(data = resistance_2015, aes(x=resistance.dif)) +
+  geom_density(alpha=0.5) +
+  theme_bw()
+
+png("../doc/display/resistance_dif.png", width=8, height=4, units="in", res=300)
+grid.arrange(resistance_2010_dif, resistance_2015_dif, nrow=1)
+dev.off()
+
+```
+```{r basic model}
+
+model_2010<-glm(resistance.dif ~ calcDBH_min1 + cii_min1 + williams_dec+
+                  calcDBH_min1*cii_min1 + 
+                  calcDBH_min1*williams_dec+
+                  cii_min1*williams_dec,
+                data=resistance_2010)
+
+summary(model_2010)
+
+model_2015<-glm(resistance.dif ~ calcDBH_min1 + cii_min1 + williams_dec+
+                  calcDBH_min1*cii_min1 + 
+                  calcDBH_min1*williams_dec+
+                  cii_min1*williams_dec,
+                data=resistance_2015)
+
+summary(model_2015)
+
+resistance_all<-rbind(resistance_2010, resistance_2015)
+resistance_all$yr<-c(rep(2010, nrow(resistance_2010)), rep(2015, nrow(resistance_2015)))
+
+resistance_model<-resistance_all %>%
+  select(resistance.dif, calcDBH_min1, cii_min1, williams_dec, yr)
+
+resistance_model<-resistance_model[complete.cases(resistance_model),]
+
+library(lme4)
+library(nlme)
+model_all<-lme(resistance.dif ~ calcDBH_min1 + cii_min1 + williams_dec+
+                  calcDBH_min1*cii_min1 + 
+                  calcDBH_min1*williams_dec+
+                  cii_min1*williams_dec,
+                random=~1|yr,
+                data=resistance_model)
+
+model_all
+anova(model_all)
+
+library(nlme)
+library(ggplot2)
+
+# Create a new data frame for predictions
+new_data <- expand.grid(
+  calcDBH_min1 = seq(min(resistance_model$calcDBH_min1), max(resistance_model$calcDBH_min1), length.out = 100),
+  cii_min1 = seq(min(resistance_model$cii_min1), max(resistance_model$cii_min1), length.out = 100),
+  williams_dec = seq(min(resistance_model$williams_dec), max(resistance_model$williams_dec), length.out = 100)
+)
+
+# Generate predictions
+new_data$predicted <- predict(model_all, newdata = new_data, level = 0)
+
+# Create the plot
+ggplot(resistance_model, aes(x = williams_dec, y = resistance.dif)) +
+  geom_point() +
+  geom_line(data = new_data[new_data$cii_min1==2,], aes(y = predicted)) +
+  theme_minimal() +
+  labs(x = "Deciduousness", y = "Drought resistance", title = "Predictions from LME Model")
+
+```
+
+```{r maps of species with tree ring records, echo=F, eval=F}
+library(ggplot2)
+
+pat<-as.character(read.table("git_pat.txt")[1,1])
+
+remotes:::download("stem1.RData",
+                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem1.rdata",
+                   auth_token = pat)
+remotes:::download("stem2.RData",
+                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem2.rdata",
+                   auth_token = pat)
+remotes:::download("stem3.RData",
+                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem3.rdata",
+                   auth_token = pat)
+remotes:::download("stem4.RData",
+                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem4.rdata",
+                   auth_token = pat)
+
+load("stem1.RData")
+load("stem2.RData")
+load("stem3.RData")
+load("stem4.RData")
+
+hkk_census<-bind_rows(hkk.stem1, hkk.stem2, hkk.stem3, hkk.stem4)
+
+hkk_census <- hkk_census %>%
+  filter(status=="A")
+
+hkk_census_ring_sp<-hkk_census %>%
+  filter(CensusID==3)%>%
+  filter(sp %in% ring.sp)
+
+hkk_census_sp <- hkk_census %>%
+  filter(CensusID==4)%>%
+  #group_by(sp, tag, CensusID) %>%
+  #dplyr::summarise(tot.agb=sum(agb, na.rm=T)) %>%
+  group_by(sp, tag)%>%
+  dplyr::summarise(tot.agb=sum(agb, na.rm=T)) %>%
+  group_by(sp) %>%
+  dplyr::summarise(n.tree=length(unique(tag)),
+                   sum.agb=sum(tot.agb))%>%
+  ungroup()%>%
+  dplyr::mutate(pct.agb=sum.agb*100/sum(sum.agb, na.rm=T))%>%
+  arrange(-pct.agb)%>%
+  dplyr::mutate(pct.agb.cum=cumsum(pct.agb))
+
+sp_top15<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA",
+            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA",
+            "GARCSP", "CROTRO", "GARUPI", "NEOLOB", "PHOEPA")
+
+#pct agb of the top 15 dendro species
+sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% sp_top15])
+
+```
+
+## Fastest and slowest in 2015 and 2016
+
+```{r echo=F}
+
+dendro.data.noouts <- dendro.data.noouts %>%
+  group_by(year) %>% 
+  dplyr::mutate(growth_quantile = ntile(diainc, 4),
+                resid_quantile = ntile (inc.resid, 4))
+
+growth_2015 <- dendro.data.noouts %>%
+  filter(year==2015)
+
+quantiles_2015<-quantile(growth_2015$diainc, probs=c(0.25, 0.5, 0.75, 1))
+
+growth_2015_plot<-ggplot(growth_2015, aes(x=diainc)) +
+  geom_density(fill="brown", col="brown", alpha=0.5) +
+  geom_vline(xintercept=quantiles_2015, col="brown")+ theme_bw()
+
+growth_2014 <- dendro.data.noouts %>%
+  filter(Cno==11)
+
+quantiles_2014<-quantile(growth_2014$diainc, probs=c(0.25, 0.5, 0.75, 1))
+
+growth_2016 <- dendro.data.noouts %>%
+  filter(year==2016)
+
+quantiles_2016<-quantile(growth_2016$diainc, probs=c(0.25, 0.5, 0.75, 1))
+
+growth_2016_plot<-ggplot(growth_2016, aes(x=diainc)) +
+  geom_density(fill="brown", col="brown", alpha=0.5) +
+  geom_vline(xintercept=quantiles_2016, col="brown")+ theme_bw()
+
+growth_2017 <- dendro.data.noouts %>%
+  filter(year==2017)
+
+quantiles_2017<-quantile(growth_2017$diainc, probs=c(0.25, 0.5, 0.75, 1))
+
+growth_2017_plot<-ggplot(growth_2017, aes(x=diainc)) +
+  geom_density(fill="brown", col="brown", alpha=0.5) +
+  geom_vline(xintercept=quantiles_2016, col="brown")+ theme_bw()
+
+growth_2014_2015_plot<-ggplot() +
+  geom_density(data = growth_2014, aes(x=diainc), fill="forestgreen", col="forestgreen", alpha=0.5) +
+  geom_vline(xintercept=quantiles_2014, col="forestgreen")+
+  geom_density(data= growth_2015, aes (x=diainc), fill="brown", col="brown", alpha=0.5) +
+  geom_text(aes(x=0.75, y=1, label="2014"), col="forestgreen", size=3)+
+  geom_text(aes(x=0.75, y=1.5, label="2015"), col="brown", size=3)+
+  geom_vline(xintercept=quantiles_2015, col="brown")+ 
+  xlab("diameter increment (cm)") + ylab("density")+
+  theme_bw()
+
+dendro.data.noouts$year_fac<-as.character(dendro.data.noouts$year)
+dendro.data.noouts$year_fac<-factor(dendro.data.noouts$year_fac, levels=as.character(rev(2009:2023)))
+
+growth_all_years<-ggplot(dendro.data.noouts, aes(x = diainc, y = year_fac, 
+                       fill = stat(x))) +
+  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
+  scale_fill_viridis(name = "growth", option = "H", direction=-1) +
+  labs(title = 'tree growth in HKK (cm)', y="year") +
+  #theme_ipsum() +
+  geom_vline(xintercept=0, lty=2)+
+  theme_bw()+
+    theme(
+      legend.position="none",
+      panel.spacing = unit(0.1, "lines"),
+      strip.text.x = element_text(size = 8)
+    )
+
+growth_all_years
+
+png("../doc/display/growth_2014_2015.png", width=4, height=4, units="in", res=300)
+growth_2014_2015_plot
+dev.off()
+
+png("../doc/display/growth_allyrs.png", width=4, height=4, units="in", res=300)
+growth_all_years
+dev.off()
+
+```
+
+##Distribution of data with size, exposure and deciduousness
+
+```{r echo=F}
+
+#2014 data distribution with size, exposure and deciduousness
+
+png("../doc/display/dbh_cii.png", width=4, height=4, units="in", res=300)
+ggplot(growth_2014, aes(x = dbh, y = factor(cii), 
+                       fill = stat(x))) +
+  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
+  scale_fill_viridis(name = "growth", option = "C", direction=-1) +
+  labs(title = 'DBH in 2014 (cm)', y="canopy illumination") +
+  #theme_ipsum() +
+  theme_bw()+
+    theme(
+      legend.position="none",
+      panel.spacing = unit(0.1, "lines"),
+      strip.text.x = element_text(size = 8)
+    )
+dev.off()
+
+length(unique(growth_2014$sp))
+
+png("../doc/display/growth_cii.png", width=4, height=4, units="in", res=300)
+ggplot(growth_2014, aes(x = diainc, y = factor(cii), 
+                       fill = stat(x))) +
+  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
+  scale_fill_viridis(name = "growth", option = "C", direction=-1) +
+  labs(title = 'growth in 2014 (cm)', y="canopy illumination") +
+  #theme_ipsum() +
+  theme_bw()+
+    theme(
+      legend.position="none",
+      panel.spacing = unit(0.1, "lines"),
+      strip.text.x = element_text(size = 8)
+    )
+dev.off()
+
+dec_williams<-read.csv("../data/HKK-dendro/species.csv")
+
+dec_williams<-dec_williams %>%
+  filter(n.tree >= 5)
+
+trial<-merge(growth_2014, dec_williams, by="sp")
+
+png("../doc/display/dbh_cii.png", width=4, height=4, units="in", res=300)
+ggplot(trial, aes(x = williams_dec, y = factor(Cii), 
+                       fill = stat(x))) +
+  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
+  scale_fill_viridis(name = "growth", option = "C", direction=-1) +
+  labs(title = 'deciduousness', y="canopy illumination") +
+  #theme_ipsum() +
+  theme_bw()+
+    theme(
+      legend.position="none",
+      panel.spacing = unit(0.1, "lines"),
+      strip.text.x = element_text(size = 8)
+    )
+dev.off()
+
+```
+
+## Size classes
+
+```{r echo=F}
+
+size_top10<-ggplot(dendro.top10.dat, aes(x=dbh))+
+  geom_density()+
+  geom_vline(aes(xintercept=35))+
+  geom_vline(aes(xintercept=75))
+
+size_top10
+
+#trying three categories
+# <50, 50-100, >100
+
+dendro.top10.dat<-dendro.top10.dat %>%
+  dplyr::mutate(size=ifelse(dbh<35, "< 35 cm dbh",
+                       ifelse(dbh<70, "35-75 cm dbh", ">75 cm dbh")))
+
+growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(sp, levels=sp_top10), y=diainc,
+                                              #color=factor(sp),
+                                          #group=factor(size, levels=c("small", "medium", "large")),
+                                              fill=factor(sp, levels=sp_top10)))+ 
+  #geom_violin(position = "dodge")+ 
+  geom_boxplot(position = "dodge")+ 
+  ylab("Annualised diameter increment (cm)")+theme_bw()+
+  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
+  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
+  geom_hline(yintercept = 0, lty=3)+
+  facet_grid(factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh"))~.)+
+      theme(axis.text.y = element_blank())+
+  scale_fill_viridis_d()+
+  scale_color_viridis_d()+
+  ggtitle("HKK dendroband data from 2008 to 2023")+
+  guides(fill=guide_legend(title="species"),
+         color=guide_legend(title="species"))+coord_flip()
+
+growth_sp10
+
+growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh")), y=diainc,
+                                              #color=factor(sp),
+                                          #group=factor(size, levels=c("small", "medium", "large")),
+                                              fill=factor(sp, levels=sp_top10)))+
+
+  geom_violin(position = "dodge")+ 
+  geom_boxplot(position = "dodge", col="black", fill="white", width=0.1)+ 
+  ylab("Annualised diameter increment (cm)")+theme_bw()+
+  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
+  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
+  geom_hline(yintercept = 0, lty=3)+
+  #facet_grid(factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh"))~.)+
+      theme(axis.text.x = element_text(angle=45))+
+  xlab("")+
+  facet_grid(.~factor(sp, levels=sp_top10))+
+  scale_fill_viridis_d()+
+  scale_color_viridis_d()+
+  ggtitle("HKK dendroband data from 2008 to 2023")+
+  guides(fill=guide_legend(title="species"),
+         color=guide_legend(title="species"))#+coord_flip()
+
+growth_sp10
+
+```
+
+## Canopy illumination
+
+```{r echo=F}
+
+cii_top10<-ggplot(dendro.top10.dat, aes(x=cii))+
+  geom_histogram()
+
+cii_top10
+
+#trying three categories
+# <50, 50-100, >100
+
+#dendro.top10.dat<-dendro.top10.dat %>%
+#  dplyr::mutate(size=ifelse(dbh<50, "small",
+#                       ifelse(dbh<100, "medium", "large")))
+
+growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
+                        aes(x=factor(sp, levels=sp_top10), y=diainc,
+                                              #color=factor(sp),
+                                          #group=factor(size, levels=c("small", "medium", "large")),
+                                              fill=factor(sp, levels=sp_top10)))+ 
+  #geom_violin(position = "dodge")+ 
+  geom_boxplot(position = "dodge")+ 
+  ylab("Annualised diameter increment (cm)")+theme_bw()+
+  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
+  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
+  geom_hline(yintercept = 0, lty=3)+
+  facet_grid(cii~.)+
+      theme(axis.text.y = element_blank())+
+  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
+  scale_fill_viridis_d()+
+  scale_color_viridis_d()+
+  guides(fill=guide_legend(title="species"),
+         color=guide_legend(title="species"))+coord_flip()
+
+growth_sp10_cii
+
+growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
+                        aes(
+                          #x=factor(sp, levels=sp_top10), 
+                          x=factor(cii), 
+                            y=diainc,
+                                              #color=factor(sp),
+                                          #group=factor(size, levels=c("small", "medium", "large")),
+                                              fill=factor(sp, levels=sp_top10)))+ 
+  #geom_violin(position = "dodge")+ 
+  geom_boxplot(position = "dodge")+ 
+  ylab("Annualised diameter increment (cm)")+theme_bw()+
+  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
+  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
+  geom_hline(yintercept = 0, lty=3)+
+  facet_grid(.~factor(sp, levels = sp_top10))+
+      theme(axis.text.y = element_blank())+
+  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
+  scale_fill_viridis_d()+
+  scale_color_viridis_d()+
+  #coord_flip()+
+  guides(fill=guide_legend(title="species"),
+         color=guide_legend(title="species"))
+growth_sp10_cii
+
+```
+## spatial patterns
+
+```{r spatial patterns, echo=F}
+
+trees.top10 <- trees.top10 %>%
+  dplyr::mutate(treeID=paste0("HKK.", Tag),
+                plotID=paste0(Col, "_", Row))
+
+dendro.top10.dat<-merge(dendro.top10.dat, 
+                        dplyr::select(trees.top10, c("Col", "Row", "treeID", "X", "Y", "plotID")), 
+                        by="treeID", all.x=T)
+
+dendro.top10.dat<-dendro.top10.dat %>%
+  dplyr::mutate(plotID=paste0(Col, "_", Row))
+
+census_spatial<-hkk_census %>%
+  group_by(quadrat) %>%
+  dplyr::summarise(sp.no=length(unique(sp)),
+                   sp.top.10 = length(unique(sp %in% sp_top10)))%>%
+  dplyr::mutate(Col=as.numeric(str_sub(quadrat, 1, 2)),
+                Row=as.numeric(str_sub(quadrat, 3, 4)))
+
+tiles<-ggplot(census_spatial, aes(x=Col, y=Row, fill=sp.no))+
+  geom_tile()+
+  #facet_wrap(~year)+
+  coord_fixed()+
+  #xlim(c(0,49))+ylim(c(0, 24))+
+  theme_bw()
+
+tiles
+
+tiles<-ggplot(census_spatial, aes(x=Col, y=Row, fill=sp.top.10))+
+  geom_tile()+
+  #facet_wrap(~year)+
+  coord_fixed()+
+  #xlim(c(0,49))+ylim(c(0, 24))+
+  theme_bw()
+
+tiles
+
+spatial_summary<- dendro.top10.dat %>%
+  group_by(Col, Row, plotID, sp, year) %>%
+  dplyr::summarise(inc.sum=sum(inc.resid, na.rm=T)) %>%
+  group_by(Col, Row, plotID, year) %>%
+  dplyr::summarise(inc.diff=var(inc.sum, na.rm=T))
+
+tiles<-ggplot(spatial_summary, aes(x=Col, y=Row, fill=log(inc.diff, 10)))+
+  geom_tile()+
+  facet_wrap(~year)+coord_fixed()+
+  xlim(c(0,49))+ylim(c(0, 24))+
+  theme_bw()
+
+tiles
+
+```
+
+## evergreen/dec
+
+```{r evg/dec, echo=F}
+
+sp_top15<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA", 
+            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA",
+            "GARCSP", "CROTRO", "GARUPI", "NEOLOB", "PHOEPA")
+
+#these 15 species had the most dendrobands and together made up 51.22% of the agb in 2009
+
+#below values eyeballed from Williams et al. 2008 Fig 1.
+sp_top15_phenoscore<-c(2.5, 0.8, 1.3, 4, 0.7, 
+                       1, 1.25, 1.27, 0.2, 0.3,
+                       0.4, 0.65, 3.95, 0.15, 0.25)
+
+#no info on Alphonsea ventricosa in the paper. 
+#Apparently evergreen, so used 1 as value - middle of evergreen range
+
+pheno<-as.data.frame(cbind(sp_top15, sp_top15_phenoscore))
+colnames(pheno)<-c("sp", "phenoscore")
+
+dendro.top10.dat<-merge(dendro.top10.dat, pheno, by="sp")
+
+#categorising as evg and deciduous
+dendro.top10.dat$cat<-ifelse(dendro.top10.dat$phenoscore<=1, "evg", "dec")
+
+growth_yr_cat_detrend<-ggplot(dendro_top15, aes(x=factor(cat), y=inc.resid,
+                                              #group=c(year, sp),
+                                              #group=sp,
+                                              fill=factor(cat)))+ 
+  geom_violin(position = "dodge")+ 
+  ylab("Residuals of annualised diameter increment")+theme_bw()+
+  facet_grid(.~year)+
+  geom_hline(yintercept = 0)+
+  scale_fill_manual(values=c("sienna4", "darkgreen"))+
+  theme(axis.text.x = element_blank())
+
+growth_yr_cat_detrend
+
+growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
+                        aes(
+                          #x=factor(sp, levels=sp_top10), 
+                          x=factor(cii), 
+                            y=diainc,
+                                              #color=factor(sp),
+                                          #group=factor(size, levels=c("small", "medium", "large")),
+                                              fill=factor(sp, levels=sp_top10)))+ 
+  #geom_violin(position = "dodge")+ 
+  geom_boxplot(position = "dodge")+ 
+  ylab("Annualised diameter increment (cm)")+theme_bw()+
+  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
+  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
+  geom_hline(yintercept = 0, lty=3)+
+  facet_grid(cat~factor(sp, levels = sp_top10))+
+      theme(axis.text.y = element_blank())+
+  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
+  scale_fill_viridis_d()+
+  scale_color_viridis_d()+
+  #coord_flip()+
+  guides(fill=guide_legend(title="species"),
+         color=guide_legend(title="species"))
+growth_sp10_cii
+
+```
+
+## Repeat all steps for dry censuses
+
+```{r echo=F}
+
+#remove all data with decimal errors, misplaced decimal
+
+`%nin%` <- Negate(`%in%`)
+
+dendro.data<-dendrobyband %>%
+  dplyr::rename(sp=SPCODE.UPDATE)%>%
+  mutate(year=year(Date2)) %>%
+  #filter(year!=2209)%>% #quick fix to remove bad dates - there are 30 entries
+  #filter(start_date <= year & year <= end_date) %>%
+  filter(Cno %nin% wet_census)%>%
   filter(is.na(DBH1_flag), DBH1_error=="none",
@@
 
-# species_all<-dendro.data.noouts %>%
-#   group_by(sp, treeID) %>%
-#   dplyr::summarise(n.obs=length(treeID))%>%
-#   ungroup() %>%
-#   group_by(sp) %>%
-#   dplyr::summarise(n.tree=length(treeID),
-#                    n.obs.tot=sum(n.obs))%>%
-#   ungroup()%>%
-#   arrange(-n.tree)
-# 
-# write.csv(species_all, "../data/HKK-dendro/species.csv", row.names = F)
-
 ```
@@
 
-```{r echo=FALSE, eval=F}
+```{r echo=FALSE}
 
@@
 
-```{r maps of species with tree ring records, echo=F, eval=F}
+## Tree ring species
+
+- Four species with tree ring chronologies in HKK
+
+```{r ring trees, echo=F}
+#These are species that have been cored near HKK and reported in Zuidema et al., Anderson-Teixeira et al., etc
+ring.sp<-c("AFZEXY", "MELIAZ", "TOONCI", "CHUKTA", "NEOLOB")
+
+trees.with.ring<-trees[trees$SPCODE.UPDATE %in% ring.sp,]
+
+dendro.ring<-dendro.data.noouts %>%
+  filter(sp %in% ring.sp)
+
+hkk_sp_ring<-trees.with.ring %>%
+  group_by(SPCODE.UPDATE)%>%
+  dplyr::summarise(n.tree=length(SPCODE.UPDATE))
+knitr::kable(hkk_sp_ring, format='html', table.attr="style='width:50%;'")
+
+```
+
+```{r maps of species with tree ring records, echo=F}
 library(ggplot2)
@@
 
-remotes:::download("stem1.RData",
-                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem1.rdata",
-                   auth_token = pat)
-remotes:::download("stem2.RData",
-                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem2.rdata",
-                   auth_token = pat)
-remotes:::download("stem3.RData",
-                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem3.rdata",
-                   auth_token = pat)
-remotes:::download("stem4.RData",
-                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem4.rdata",
-                   auth_token = pat)
+# remotes:::download("stem1.RData",
+#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem1.rdata",
+#                    auth_token = pat)
+# remotes:::download("stem2.RData",
+#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem2.rdata",
+#                    auth_token = pat)
+# remotes:::download("stem3.RData",
+#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem3.rdata",
+#                    auth_token = pat)
+# remotes:::download("stem4.RData",
+#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem4.rdata",
+#                    auth_token = pat)
+#
 
@@
 
-```
-
+ring.sp.map.census<-ggplot(hkk_census_ring_sp, aes(x=gx, y=gy, size=as.numeric(dbh/10), color=factor(sp)))+
+  geom_point(alpha=0.5)+
+   scale_size(range = c(.1, 10), name="DBH (cm)")+
+  guides(color=guide_legend(title="Species"))+
+  #scale_colour_gradient(low="orange", high="purple") +
+  theme_minimal()+
+  coord_fixed()
+
+ring.sp.map.census
+
+ring.sp.map<-ggplot(trees.with.ring, aes(x=X, y=Y, size=as.numeric(DBH.04), color=factor(SPCODE.UPDATE)))+
+  geom_point(alpha=0.5)+
+   scale_size(range = c(.1, 10), name="DBH (cm)")+
+  guides(color=guide_legend(title="Species"))+
+  #scale_colour_gradient(low="orange", high="purple") +
+  theme_minimal()+
+  coord_fixed()
+
+ring.sp.map
+```
+Species with tree rings contribute `r round(sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% ring.sp]),2)` % aboveground biomass in 2004
+
+## Dendrobands on ring species
+
+```{r echo=F}
+
+growth_sp<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
+                                              fill=factor(sp),
+                                              color=factor(sp)))+ 
+  geom_violin(position = "dodge")+ 
+  geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
+  ylab("Annualised diameter increment (cm)")+theme_bw()+
+  #facet_grid(.~year)+
+  #geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
+  geom_hline(yintercept = median(dendro.ring$diainc, na.rm=T), lty=2)+
+  geom_hline(yintercept = 0, lty=3)+
+      theme(axis.text.x = element_blank())+
+  ggtitle("HKK dendroband data from 2008 to 2023")+
+  geom_text(aes(label=paste0("median=\n", round(median(dendro.ring$diainc, na.rm=T),2), " cm"), 
+                x=0.6, y=median(dendro.ring$diainc, na.rm=T)+0.15), col="black", size=3)+
+  guides(fill=guide_legend(title="species"),
+         color=guide_legend(title="species"))
+
+growth_sp
+
+growth_sp_box<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
+                                              #color=factor(sp),
+                                              fill=factor(sp)))+ 
+  geom_boxplot(position = "dodge")+ 
+  ylab("Annualised diameter increment (cm)")+theme_bw()+
+  #facet_grid(.~year)+
+  geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
+  geom_hline(yintercept = median(dendro.ring$diainc, na.rm=T), lty=2)+
+  geom_hline(yintercept = 0, lty=3)+
+      theme(axis.text.x = element_blank())+
+  ggtitle("HKK dendroband data from 2008 to 2023")+
+  guides(fill=guide_legend(title="species"),
+         color=guide_legend(title="species"))
+
+#growth_sp_box
+
+growth_yr_sp<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
+                                              #color=factor(sp),
+                                              fill=factor(sp)))+ 
+  #geom_violin(position = "dodge")+ 
+  geom_boxplot(position="dodge")+
+  ylab("Annualised diameter increment (cm)")+theme_bw()+
+  facet_grid(.~year)+
+  geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
+  theme(axis.text.x = element_blank())+
+  guides(color="none")
+
+growth_yr_sp
+
+```
+
+## Detrending by tree with smooth splines
+
+```{r detrending, echo=F}
+
+#trying for one tree
+
+dendro.ring<-dendro.ring %>%
+  group_by(sp, treeID) %>%
+  dplyr::mutate(n.obs=length(treeID))%>%
+  filter(n.obs>=4)%>%
+  ungroup()%>%
+  group_by(sp, treeID)%>%
+  dplyr::mutate(inc.resid=residuals(smooth.spline(x=calcDBH_min1, y=diainc)),
+                inc.spline=fitted(smooth.spline(x=calcDBH_min1, y=diainc)))
+
+growth_sp_detrend<-ggplot(dendro.ring, aes(x=factor(sp), 
+                                           y=inc.resid, 
+                                           #color=factor(sp),
+                                           fill=factor(sp)))+ 
+  #geom_violin()+ 
+  geom_boxplot()+ 
+  ylab("Residuals of annualised diameter increment")+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle=45))
+growth_sp_detrend
+
+growth_yr_detrend<-ggplot(dendro.ring, aes(x=year, y=inc.resid, group=year))+ geom_boxplot()+ 
+  ylab("Residuals of annualised diameter increment")+theme_bw()+
+  theme(axis.text.x = element_text(angle=45))
+growth_yr_detrend
+
+```
+
+## Detrended plots by species
+
+```{r echo=F}
+
+growth_yr_sp_detrend<-ggplot(dendro.ring, aes(x=factor(sp), y=inc.resid,
+                                              #group=c(year, sp),
+                                              #group=sp,
+                                              #color=factor(sp),
+                                              fill=factor(sp)))+ 
+  geom_boxplot(position = "dodge")+ 
+  ylab("Residuals of annualised diameter increment")+theme_bw()+
+  geom_hline(yintercept=0)+
+  facet_grid(.~year)+
+  theme(axis.text.x = element_blank())+
+  guides(color="none")
+
+growth_yr_sp_detrend
+
+```
+
+```{r echo=F}
+
+spag_all<-ggplot(dendro.ring, aes(x=year, y=inc.spline, group=treeID)) +
+  geom_line()+ggtitle("fitted values (smooth spline)")
+
+spag_all
+
+spag_resid<-ggplot(dendro.ring, aes(x=year, y=inc.resid, group=treeID)) +
+  geom_line()+ggtitle("residuals smooth spline")
+
+spag_resid
+```
 ## Detrending all species
@@
 ```{r detrending all, echo=F}
+
+#trying for one tree
 
@@
 ```{r echo=F}
-
 growth_yr_detrend_all<-ggplot(dendro.data.noouts, aes(x=year, y=inc.resid, group=year))+ 
@@
 top10sp.map.dendro
-
-```
-## Fastest and slowest in 2015 and 2016
-
-```{r echo=F}
-
-dendro.data.noouts <- dendro.data.noouts %>%
-  group_by(year) %>% 
-  dplyr::mutate(growth_quantile = ntile(diainc, 4),
-                resid_quantile = ntile (inc.resid, 4))
-
-growth_2015 <- dendro.data.noouts %>%
-  filter(year==2015)
-
-quantiles_2015<-quantile(growth_2015$diainc, probs=c(0.25, 0.5, 0.75, 1))
-
-growth_2015_plot<-ggplot(growth_2015, aes(x=diainc)) +
-  geom_density(fill="brown", col="brown", alpha=0.5) +
-  geom_vline(xintercept=quantiles_2015, col="brown")+ theme_bw()
-
-growth_2014 <- dendro.data.noouts %>%
-  filter(year==2014)
-
-quantiles_2014<-quantile(growth_2014$diainc, probs=c(0.25, 0.5, 0.75, 1))
-
-growth_2016 <- dendro.data.noouts %>%
-  filter(year==2016)
-
-quantiles_2016<-quantile(growth_2016$diainc, probs=c(0.25, 0.5, 0.75, 1))
-
-growth_2016_plot<-ggplot(growth_2016, aes(x=diainc)) +
-  geom_density(fill="brown", col="brown", alpha=0.5) +
-  geom_vline(xintercept=quantiles_2016, col="brown")+ theme_bw()
-
-growth_2017 <- dendro.data.noouts %>%
-  filter(year==2017)
-
-quantiles_2017<-quantile(growth_2017$diainc, probs=c(0.25, 0.5, 0.75, 1))
-
-growth_2017_plot<-ggplot(growth_2017, aes(x=diainc)) +
-  geom_density(fill="brown", col="brown", alpha=0.5) +
-  geom_vline(xintercept=quantiles_2016, col="brown")+ theme_bw()
-
-growth_2014_2015_plot<-ggplot() +
-  geom_density(data = growth_2014, aes(x=diainc), fill="forestgreen", col="forestgreen", alpha=0.5) +
-  geom_vline(xintercept=quantiles_2014, col="forestgreen")+
-  geom_density(data= growth_2015, aes (x=diainc), fill="brown", col="brown", alpha=0.5) +
-  geom_text(aes(x=0.75, y=1, label="2014"), col="forestgreen", size=3)+
-  geom_text(aes(x=0.75, y=1.5, label="2015"), col="brown", size=3)+
-  geom_vline(xintercept=quantiles_2015, col="brown")+ 
-  xlab("diameter increment (cm)") + ylab("density")+
-  theme_bw()
-
-dendro.data.noouts$year_fac<-as.character(dendro.data.noouts$year)
-dendro.data.noouts$year_fac<-factor(dendro.data.noouts$year_fac, levels=as.character(rev(2009:2023)))
-
-growth_all_years<-ggplot(dendro.data.noouts, aes(x = diainc, y = year_fac, 
-                       fill = stat(x))) +
-  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
-  scale_fill_viridis(name = "growth", option = "H", direction=-1) +
-  labs(title = 'tree growth in HKK (cm)', y="year") +
-  #theme_ipsum() +
-  geom_vline(xintercept=0, lty=2)+
-  theme_bw()+
-    theme(
-      legend.position="none",
-      panel.spacing = unit(0.1, "lines"),
-      strip.text.x = element_text(size = 8)
-    )
-
-growth_all_years
-
-png("../doc/display/growth_2014_2015.png", width=4, height=4, units="in", res=300)
-growth_2014_2015_plot
-dev.off()
-
-png("../doc/display/growth_allyrs.png", width=4, height=4, units="in", res=300)
-growth_all_years
-dev.off()
-
-```
-
-##Distribution of data with size, exposure and deciduousness
-
-```{r echo=F}
-
-#2014 data distribution with size, exposure and deciduousness
-
-png("../doc/display/dbh_cii.png", width=4, height=4, units="in", res=300)
-ggplot(growth_2014, aes(x = dbh, y = factor(cii), 
-                       fill = stat(x))) +
-  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
-  scale_fill_viridis(name = "growth", option = "C", direction=-1) +
-  labs(title = 'DBH in 2014 (cm)', y="canopy illumination") +
-  #theme_ipsum() +
-  theme_bw()+
-    theme(
-      legend.position="none",
-      panel.spacing = unit(0.1, "lines"),
-      strip.text.x = element_text(size = 8)
-    )
-
-length(unique(growth_2014$sp))
-
-png("../doc/display/growth_cii.png", width=4, height=4, units="in", res=300)
-ggplot(growth_2014, aes(x = diainc, y = factor(cii), 
-                       fill = stat(x))) +
-  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
-  scale_fill_viridis(name = "growth", option = "C", direction=-1) +
-  labs(title = 'growth in 2014 (cm)', y="canopy illumination") +
-  #theme_ipsum() +
-  theme_bw()+
-    theme(
-      legend.position="none",
-      panel.spacing = unit(0.1, "lines"),
-      strip.text.x = element_text(size = 8)
-    )
-dev.off()
 
@@
 ```
-## spatial patterns
-
-```{r spatial patterns, echo=F}
-
-trees.top10 <- trees.top10 %>%
-  dplyr::mutate(treeID=paste0("HKK.", Tag),
-                plotID=paste0(Col, "_", Row))
-
-dendro.top10.dat<-merge(dendro.top10.dat, 
-                        dplyr::select(trees.top10, c("Col", "Row", "treeID", "X", "Y", "plotID")), 
-                        by="treeID", all.x=T)
-
-dendro.top10.dat<-dendro.top10.dat %>%
-  dplyr::mutate(plotID=paste0(Col, "_", Row))
-
-census_spatial<-hkk_census %>%
-  group_by(quadrat) %>%
-  dplyr::summarise(sp.no=length(unique(sp)),
-                   sp.top.10 = length(unique(sp %in% sp_top10)))%>%
-  dplyr::mutate(Col=as.numeric(str_sub(quadrat, 1, 2)),
-                Row=as.numeric(str_sub(quadrat, 3, 4)))
-
-tiles<-ggplot(census_spatial, aes(x=Col, y=Row, fill=sp.no))+
-  geom_tile()+
-  #facet_wrap(~year)+
-  coord_fixed()+
-  #xlim(c(0,49))+ylim(c(0, 24))+
-  theme_bw()
-
-tiles
-
-tiles<-ggplot(census_spatial, aes(x=Col, y=Row, fill=sp.top.10))+
-  geom_tile()+
-  #facet_wrap(~year)+
-  coord_fixed()+
-  #xlim(c(0,49))+ylim(c(0, 24))+
-  theme_bw()
-
-tiles
-
-spatial_summary<- dendro.top10.dat %>%
-  group_by(Col, Row, plotID, sp, year) %>%
-  dplyr::summarise(inc.sum=sum(inc.resid, na.rm=T)) %>%
-  group_by(Col, Row, plotID, year) %>%
-  dplyr::summarise(inc.diff=var(inc.sum, na.rm=T))
-
-tiles<-ggplot(spatial_summary, aes(x=Col, y=Row, fill=log(inc.diff, 10)))+
-  geom_tile()+
-  facet_wrap(~year)+coord_fixed()+
-  xlim(c(0,49))+ylim(c(0, 24))+
-  theme_bw()
-
-tiles
-
-```
-
-## evergreen/dec
+
+## Top 15 species split by leaf habit
 
@@
 
-dendro.top10.dat<-merge(dendro.top10.dat, pheno, by="sp")
+dendro_top15<-dendro.data.noouts[dendro.data.noouts$sp %in% sp_top15,]
+dendro_top15<-merge(dendro_top15, pheno, by="sp")
 
 #categorising as evg and deciduous
-dendro.top10.dat$cat<-ifelse(dendro.top10.dat$phenoscore<=1, "evg", "dec")
+dendro_top15$cat<-ifelse(dendro_top15$phenoscore<=2, "evg", "dec")
 
@@
 
-growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
-                        aes(
-                          #x=factor(sp, levels=sp_top10), 
-                          x=factor(cii), 
-                            y=diainc,
-                                              #color=factor(sp),
-                                          #group=factor(size, levels=c("small", "medium", "large")),
-                                              fill=factor(sp, levels=sp_top10)))+ 
-  #geom_violin(position = "dodge")+ 
-  geom_boxplot(position = "dodge")+ 
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
-  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-  facet_grid(cat~factor(sp, levels = sp_top10))+
-      theme(axis.text.y = element_blank())+
-  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  #coord_flip()+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))
-growth_sp10_cii
-
-```
-
-## Repeat all steps for dry censuses
-
-```{r echo=F}
-
-#remove all data with decimal errors, misplaced decimal
-
-`%nin%` <- Negate(`%in%`)
-
-dendro.data<-dendrobyband %>%
-  dplyr::rename(sp=SPCODE.UPDATE)%>%
-  mutate(year=year(Date2)) %>%
-  #filter(year!=2209)%>% #quick fix to remove bad dates - there are 30 entries
-  #filter(start_date <= year & year <= end_date) %>%
-  filter(Cno %nin% wet_census)%>%
-  filter(is.na(DBH1_flag), DBH1_error=="none",
-         win.decimal.error=="none", is.na(win.flag))%>%
-  group_by(year, sp, Tag) %>%
-  dplyr::summarise(calcDBH=calcDBH[which(n.dendro==min(n.dendro))],
-                   Date2=Date2[which(n.dendro==min(n.dendro))],
-                   Cii=max(Cii, na.rm=T))
-
-dendro.data_all<-dendro.data%>%
-  group_by(Tag) %>% #grouping by treeID to calculate increment
-  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
-         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
-         inc=calcDBH-calcDBH_min1,
-         inc_annual=(inc/as.numeric(tdif))*365,
-         inc.time=year-lag(year),
-         #removing increment value if the previous measure was before the last year
-         inc_annual=ifelse(inc.time==1, inc_annual, NA),
-         #has.out=ifelse(1 %in% out, 1, 0),
-         treeID=paste0("HKK.", Tag))%>%
-  filter(!is.na(inc))%>%
-  ungroup()%>%
-  filter(!is.na(inc_annual))%>%
-  ungroup()%>%
-  #filter(start_date <= year & year <= end_date) %>%
-  group_by(sp, treeID, year)%>%
-  dplyr::summarise(
-    diainc = mean(inc_annual, na.rm=T),
-    Cii=max(Cii, na.rm=T))
-
-#first fitting a line through the time series
-obs.model.dbh<-nlme::lme(calcDBH~year, random=~1|Tag, data=dendro.data,
-                         na.action=na.omit,
-                         method="ML", control=(msMaxIter=100)) #had to increase the number of iterations
-
-#standardised residuals calculated https://stat.ethz.ch/R-manual/R-devel/library/nlme/html/residuals.lme.html
-
-std.res.dbh<-residuals(obs.model.dbh, type="pearson")
-hist(std.res.dbh)
-dendro.data$std.res<-rep(NA, nrow(dendro.data))
-dendro.data$std.res[!is.na(dendro.data$calcDBH)]<-std.res.dbh
-#defining all observation with abs(residual) >3 as an outlier
-dendro.data$out<-factor(ifelse(abs(dendro.data$std.res)<=3, 0, 1))
-
-#find the trees which have outlier observations
-outs.Tag<-unique(dendro.data$Tag[dendro.data$out==1])
-
-#plot trees that have outliers
-ggplot(dendro.data[dendro.data$Tag %in%outs.Tag,],
-       aes(x = year, y = calcDBH, col=factor(out), group = Tag))+
-  geom_line(alpha=0.5)+geom_point()+
-  scale_color_manual(values=c("grey20", "red", "grey"))
-
-dendro.data.noouts<-dendro.data %>%
-  mutate(dbh3=ifelse(abs(std.res)<=3, calcDBH, NA)) %>% #create an outlier removed column
-  group_by(Tag) %>% #grouping by treeID to calculate increment
-  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
-         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
-         inc=calcDBH-calcDBH_min1,
-         inc_annual=(inc/as.numeric(tdif))*365,
-         inc.time=year-lag(year),
-         #removing increment value if the previous measure was before the last year
-         #inc=ifelse(inc.time==1, inc, NA),
-         inc_annual=ifelse(inc.time==1, inc_annual, NA),
-         has.out=ifelse(1 %in% out, 1, 0),
-         treeID=paste0("HKK.", Tag))%>%
-  filter(!is.na(inc))%>%
-  ungroup()%>%
-  filter(!is.na(inc_annual))%>%
-  ungroup()%>%
-  #filter(start_date <= year & year <= end_date) %>%
-  group_by(sp, treeID, year)%>%
-  dplyr::summarise(
-    diainc = mean(inc_annual, na.rm=T),
-    calcDBH_min1 = mean(calcDBH_min1, na.rm=T),
-    dbh=mean(calcDBH, na.rm=T),
-    out = ifelse(1 %in% out, 1, 0),
-    has.out = ifelse(1 %in% has.out, 1, 0), 
-    cii=max(Cii, na.rm=T))%>%
-  filter(out==0)%>%
-  ungroup() %>%
-  dplyr::mutate(
-    diainc.scaled = scale(diainc),
-    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))%>%
-  filter(out.inc==0)
-
-```
-
-## Raw diameter increments of all species
-
-```{r echo=FALSE}
-
-library(ggplot2)
-
-growth_yr_withouts<-ggplot(dendro.data_all, aes(x=year, y=diainc, group=year))+
-  geom_violin()+
-  ylab("Annualised diameter increment (cm)")+theme_bw()
-
-growth_yr_withouts
-```
-
-## Removing outliers based (more than 3 SD)
-
-```{r echo=FALSE}
-
-growth_yr_noouts<-ggplot(dendro.data.noouts, aes(x=year, y=diainc, group=year))+ geom_violin()+ ylab("Annualised diameter increment (cm)")+theme_bw()
-
-growth_yr_noouts
-```
-
-## Tree ring species
-
-- Four species with tree ring chronologies in HKK
-
-```{r ring trees, echo=F}
-#These are species that have been cored near HKK and reported in Zuidema et al., Anderson-Teixeira et al., etc
-ring.sp<-c("AFZEXY", "MELIAZ", "TOONCI", "CHUKTA", "NEOLOB")
-
-trees.with.ring<-trees[trees$SPCODE.UPDATE %in% ring.sp,]
-
-dendro.ring<-dendro.data.noouts %>%
-  filter(sp %in% ring.sp)
-
-hkk_sp_ring<-trees.with.ring %>%
-  group_by(SPCODE.UPDATE)%>%
-  dplyr::summarise(n.tree=length(SPCODE.UPDATE))
-knitr::kable(hkk_sp_ring, format='html', table.attr="style='width:50%;'")
-
-```
-
-```{r maps of species with tree ring records, echo=F}
-library(ggplot2)
-
-pat<-as.character(read.table("git_pat.txt")[1,1])
-
-# remotes:::download("stem1.RData",
-#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem1.rdata",
-#                    auth_token = pat)
-# remotes:::download("stem2.RData",
-#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem2.rdata",
-#                    auth_token = pat)
-# remotes:::download("stem3.RData",
-#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem3.rdata",
-#                    auth_token = pat)
-# remotes:::download("stem4.RData",
-#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem4.rdata",
-#                    auth_token = pat)
-#
-
-load("stem1.RData")
-load("stem2.RData")
-load("stem3.RData")
-load("stem4.RData")
-
-hkk_census<-bind_rows(hkk.stem1, hkk.stem2, hkk.stem3, hkk.stem4)
-
-hkk_census <- hkk_census %>%
-  filter(status=="A")
-
-hkk_census_ring_sp<-hkk_census %>%
-  filter(CensusID==3)%>%
-  filter(sp %in% ring.sp)
-
-hkk_census_sp <- hkk_census %>%
-  filter(CensusID==4)%>%
-  #group_by(sp, tag, CensusID) %>%
-  #dplyr::summarise(tot.agb=sum(agb, na.rm=T)) %>%
-  group_by(sp, tag)%>%
-  dplyr::summarise(tot.agb=sum(agb, na.rm=T)) %>%
-  group_by(sp) %>%
-  dplyr::summarise(n.tree=length(unique(tag)),
-                   sum.agb=sum(tot.agb))%>%
-  ungroup()%>%
-  dplyr::mutate(pct.agb=sum.agb*100/sum(sum.agb, na.rm=T))%>%
-  arrange(-pct.agb)%>%
-  dplyr::mutate(pct.agb.cum=cumsum(pct.agb))
-
-sp_top15<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA",
-            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA",
-            "GARCSP", "CROTRO", "GARUPI", "NEOLOB", "PHOEPA")
-
-#pct agb of the top 15 dendro species
-sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% sp_top15])
-
-ring.sp.map.census<-ggplot(hkk_census_ring_sp, aes(x=gx, y=gy, size=as.numeric(dbh/10), color=factor(sp)))+
-  geom_point(alpha=0.5)+
-   scale_size(range = c(.1, 10), name="DBH (cm)")+
-  guides(color=guide_legend(title="Species"))+
-  #scale_colour_gradient(low="orange", high="purple") +
-  theme_minimal()+
-  coord_fixed()
-
-ring.sp.map.census
-
-ring.sp.map<-ggplot(trees.with.ring, aes(x=X, y=Y, size=as.numeric(DBH.04), color=factor(SPCODE.UPDATE)))+
-  geom_point(alpha=0.5)+
-   scale_size(range = c(.1, 10), name="DBH (cm)")+
-  guides(color=guide_legend(title="Species"))+
-  #scale_colour_gradient(low="orange", high="purple") +
-  theme_minimal()+
-  coord_fixed()
-
-ring.sp.map
-```
-Species with tree rings contribute `r round(sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% ring.sp]),2)` % aboveground biomass in 2004
-
-## Dendrobands on ring species
-
-```{r echo=F}
-
-growth_sp<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
-                                              fill=factor(sp),
-                                              color=factor(sp)))+ 
-  geom_violin(position = "dodge")+ 
-  geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #facet_grid(.~year)+
-  #geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
-  geom_hline(yintercept = median(dendro.ring$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-      theme(axis.text.x = element_blank())+
-  ggtitle("HKK dendroband data from 2008 to 2023")+
-  geom_text(aes(label=paste0("median=\n", round(median(dendro.ring$diainc, na.rm=T),2), " cm"), 
-                x=0.6, y=median(dendro.ring$diainc, na.rm=T)+0.15), col="black", size=3)+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))
-
-growth_sp
-
-growth_sp_box<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
-                                              #color=factor(sp),
-                                              fill=factor(sp)))+ 
-  geom_boxplot(position = "dodge")+ 
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #facet_grid(.~year)+
-  geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
-  geom_hline(yintercept = median(dendro.ring$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-      theme(axis.text.x = element_blank())+
-  ggtitle("HKK dendroband data from 2008 to 2023")+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))
-
-#growth_sp_box
-
-growth_yr_sp<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
-                                              #color=factor(sp),
-                                              fill=factor(sp)))+ 
-  #geom_violin(position = "dodge")+ 
-  geom_boxplot(position="dodge")+
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  facet_grid(.~year)+
-  geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
-  theme(axis.text.x = element_blank())+
-  guides(color="none")
-
-growth_yr_sp
-
-```
-
-## Detrending by tree with smooth splines
-
-```{r detrending, echo=F}
-
-#trying for one tree
-
-dendro.ring<-dendro.ring %>%
-  group_by(sp, treeID) %>%
-  dplyr::mutate(n.obs=length(treeID))%>%
-  filter(n.obs>=4)%>%
-  ungroup()%>%
-  group_by(sp, treeID)%>%
-  dplyr::mutate(inc.resid=residuals(smooth.spline(x=calcDBH_min1, y=diainc)),
-                inc.spline=fitted(smooth.spline(x=calcDBH_min1, y=diainc)))
-
-growth_sp_detrend<-ggplot(dendro.ring, aes(x=factor(sp), 
-                                           y=inc.resid, 
-                                           #color=factor(sp),
-                                           fill=factor(sp)))+ 
-  #geom_violin()+ 
-  geom_boxplot()+ 
-  ylab("Residuals of annualised diameter increment")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle=45))
-growth_sp_detrend
-
-growth_yr_detrend<-ggplot(dendro.ring, aes(x=year, y=inc.resid, group=year))+ geom_boxplot()+ 
-  ylab("Residuals of annualised diameter increment")+theme_bw()+
-  theme(axis.text.x = element_text(angle=45))
-growth_yr_detrend
-
-```
-
-## Detrended plots by species
-
-```{r echo=F}
-
-growth_yr_sp_detrend<-ggplot(dendro.ring, aes(x=factor(sp), y=inc.resid,
-                                              #group=c(year, sp),
-                                              #group=sp,
-                                              #color=factor(sp),
-                                              fill=factor(sp)))+ 
-  geom_boxplot(position = "dodge")+ 
-  ylab("Residuals of annualised diameter increment")+theme_bw()+
-  geom_hline(yintercept=0)+
-  facet_grid(.~year)+
-  theme(axis.text.x = element_blank())+
-  guides(color="none")
-
-growth_yr_sp_detrend
-
-```
-
-```{r echo=F}
-
-spag_all<-ggplot(dendro.ring, aes(x=year, y=inc.spline, group=treeID)) +
-  geom_line()+ggtitle("fitted values (smooth spline)")
-
-spag_all
-
-spag_resid<-ggplot(dendro.ring, aes(x=year, y=inc.resid, group=treeID)) +
-  geom_line()+ggtitle("residuals smooth spline")
-
-spag_resid
-```
-## Detrending all species
-
-```{r detrending all, echo=F}
-
-#trying for one tree
-
-dendro.data.noouts<-dendro.data.noouts %>%
-  group_by(sp, treeID) %>%
-  dplyr::mutate(n.obs=length(treeID))%>%
-  filter(n.obs>=4)%>%
-  ungroup()%>%
-  filter(treeID!="HKK.92562")%>% #removing an issue with spline for now
-  group_by(sp, treeID)%>%
-  dplyr::mutate(inc.resid=residuals(smooth.spline(x=calcDBH_min1, y=diainc)),
-                inc.spline=fitted(smooth.spline(x=calcDBH_min1, y=diainc)))
-```
-
-For `r length(unique(dendro.data.noouts$sp))` species with dendrobands, all individual trees detrended for DBH with smooth splines
-
-```{r echo=F}
-growth_yr_detrend_all<-ggplot(dendro.data.noouts, aes(x=year, y=inc.resid, group=year))+ 
-  #geom_violin()+ 
-  geom_boxplot()+ 
-  ylab("Residuals of annualised diameter increment")+theme_bw()+
-  theme(axis.text.x = element_text(angle=45))
-growth_yr_detrend_all
-
-```
-
-## top 10 species
-
-```{r echo=F}
-
-dendro.by.sp<-dendro.data.noouts %>%
-  group_by (sp, year) %>%
-  dplyr::summarise(n.inds=length(sp),
-    mean.inc = mean(inc.resid, na.rm=T),
-                   median.inc = median(inc.resid, na.rm=T),
-                   sd.inc = sd(inc.resid, na.rm=T),
-    lower=mean.inc-sd.inc,
-    upper=mean.inc+sd.inc)%>%
-  filter(n.inds>=10)%>%
-  arrange(year, -median.inc)
-
-sp_top5<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA")
-
-sp_top10<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA", 
-            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA")
-
-dendro.top5<-dendro.by.sp %>%
-  filter(sp %in% sp_top5)
-
-dendro.top5.dat<-dendro.data.noouts %>%
-  filter(sp %in% sp_top5)
-
-dendro.top10.dat<-dendro.data.noouts %>%
-  filter(sp %in% sp_top10)
-
-hkk_census_top10_sp<-hkk_census %>%
-  filter(CensusID==3)%>%
-  filter(sp %in% sp_top10)
-
-#pct agb of the top 15 dendro species
-sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% sp_top15])
-
-top10sp.map.census<-ggplot(hkk_census_top10_sp, aes(x=gx, y=gy, 
-                                                    #size=as.numeric(dbh/10),
-                                                    color=factor(sp, levels=sp_top10)))+
-  geom_point(size=1)+
-   scale_size(range = c(.1, 10), name="dbh")+
-  guides(color=guide_legend(title="Species"))+
-  scale_color_viridis_d()+
-  ggtitle("top 10 species in 2004")+
-  #scale_colour_gradient(low="orange", high="purple") +
-  theme_minimal()+
-  coord_fixed()
-
-top10sp.map.census
-
-trees.top10<-trees[trees$SPCODE.UPDATE %in% sp_top10,]
-
-top10sp.map.dendro<-ggplot(trees.top10, aes(x=X, y=Y, 
-                                                    #size=as.numeric(dbh/10), 
-                                                    color=factor(SPCODE.UPDATE, levels=sp_top10)))+
-  geom_point(size=1)+
-   scale_size(range = c(.1, 10), name="dbh")+
-  guides(color=guide_legend(title="Species"))+
-  scale_color_viridis_d()+
-  ggtitle("top 10 species in 2004")+
-  #scale_colour_gradient(low="orange", high="purple") +
-  theme_minimal()+
-  coord_fixed()
-
-top10sp.map.dendro
-
-```
-
-## Growth of most abundant species
-
-```{r echo=F}
-
-growth_sp5<-ggplot(dendro.top5.dat, aes(x=factor(sp, levels=sp_top5), y=diainc,
-                                              fill=factor(sp, levels=sp_top5),
-                                              color=factor(sp, levels=sp_top5)))+ 
-  geom_violin(position = "dodge")+ 
-    geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
-ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #geom_hline(yintercept = mean(dendro.top5.dat$diainc, na.rm=T))+
-  geom_hline(yintercept = median(dendro.top5.dat$diainc, na.rm=T), lty=2)+
-  geom_text(aes(label=paste0("median=\n", round(median(dendro.top5.dat$diainc, na.rm=T),2), " cm"), 
-                x=0.6, y=median(dendro.top5.dat$diainc, na.rm=T)+0.15), col="black", size=3)+
-  geom_hline(yintercept = 0, lty=3)+
-      theme(axis.text.x = element_blank())+
-  ggtitle("HKK dendroband data from 2008 to 2023")+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))+
-  xlab("")
-
-growth_sp5
-
-growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(sp, levels=sp_top10), y=diainc,
-                                              fill=factor(sp, levels=sp_top10),
-                                              color=factor(sp, levels=sp_top10)))+ 
-  geom_violin(position = "dodge")+ 
-    geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
-ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
-  geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
-  geom_text(aes(label=paste0("median=\n", round(median(dendro.top10.dat$diainc, na.rm=T),2), " cm"), 
-                x=0.6, y=median(dendro.top10.dat$diainc, na.rm=T)+0.15), col="black", size=3)+
-  geom_hline(yintercept = 0, lty=3)+
-      theme(axis.text.x = element_blank())+
-  ggtitle("HKK dendroband data from 2008 to 2023")+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))+
-  xlab("")
-
-growth_sp10
-
-growth_yr_sp<-ggplot(dendro.top5, aes(x=year, y=mean.inc, 
-                                      color=sp))+ 
-  geom_point()+
-  geom_errorbar(aes(x=year, ymin=lower, ymax=upper))+
-  ylab("Residuals of annualised diameter increment")+theme_bw()+
-  theme(axis.text.x = element_text(angle=45))+
-  facet_grid(sp~.)+
-  guides(color="none")+
-  geom_hline(yintercept = 0)+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  geom_vline(xintercept=c(2010, 2016), lty=2)
-growth_yr_sp
-
-#png("plots/growth_by_sp.png", height=8, width=4, units="in", res=300)
-#growth_yr_sp
-#dev.off()
-
-```
-
-## Size classes
-
-```{r echo=F}
-
-size_top10<-ggplot(dendro.top10.dat, aes(x=dbh))+
-  geom_density()+
-  geom_vline(aes(xintercept=35))+
-  geom_vline(aes(xintercept=75))
-
-size_top10
-
-#trying three categories
-# <50, 50-100, >100
-
-dendro.top10.dat<-dendro.top10.dat %>%
-  dplyr::mutate(size=ifelse(dbh<35, "< 35 cm dbh",
-                       ifelse(dbh<70, "35-75 cm dbh", ">75 cm dbh")))
-
-growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(sp, levels=sp_top10), y=diainc,
-                                              #color=factor(sp),
-                                          #group=factor(size, levels=c("small", "medium", "large")),
-                                              fill=factor(sp, levels=sp_top10)))+ 
-  #geom_violin(position = "dodge")+ 
-  geom_boxplot(position = "dodge")+ 
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
-  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-  facet_grid(factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh"))~.)+
-      theme(axis.text.y = element_blank())+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  ggtitle("HKK dendroband data from 2008 to 2023")+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))+coord_flip()
-
-growth_sp10
-
-growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh")), y=diainc,
-                                              #color=factor(sp),
-                                          #group=factor(size, levels=c("small", "medium", "large")),
-                                              fill=factor(sp, levels=sp_top10)))+
-
-  geom_violin(position = "dodge")+ 
-  geom_boxplot(position = "dodge", col="black", fill="white", width=0.1)+ 
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
-  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-  #facet_grid(factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh"))~.)+
-      theme(axis.text.x = element_text(angle=45))+
-  xlab("")+
-  facet_grid(.~factor(sp, levels=sp_top10))+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  ggtitle("HKK dendroband data from 2008 to 2023")+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))#+coord_flip()
-
-growth_sp10
-
-```
-
-## Canopy illumination
-
-```{r echo=F}
-
-cii_top10<-ggplot(dendro.top10.dat, aes(x=cii))+
-  geom_histogram()
-
-cii_top10
-
-#trying three categories
-# <50, 50-100, >100
-
-#dendro.top10.dat<-dendro.top10.dat %>%
-#  dplyr::mutate(size=ifelse(dbh<50, "small",
-#                       ifelse(dbh<100, "medium", "large")))
-
-growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
-                        aes(x=factor(sp, levels=sp_top10), y=diainc,
-                                              #color=factor(sp),
-                                          #group=factor(size, levels=c("small", "medium", "large")),
-                                              fill=factor(sp, levels=sp_top10)))+ 
-  #geom_violin(position = "dodge")+ 
-  geom_boxplot(position = "dodge")+ 
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
-  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-  facet_grid(cii~.)+
-      theme(axis.text.y = element_blank())+
-  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))+coord_flip()
-
-growth_sp10_cii
-
-growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
-                        aes(
-                          #x=factor(sp, levels=sp_top10), 
-                          x=factor(cii), 
-                            y=diainc,
-                                              #color=factor(sp),
-                                          #group=factor(size, levels=c("small", "medium", "large")),
-                                              fill=factor(sp, levels=sp_top10)))+ 
-  #geom_violin(position = "dodge")+ 
-  geom_boxplot(position = "dodge")+ 
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
-  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-  facet_grid(.~factor(sp, levels = sp_top10))+
-      theme(axis.text.y = element_blank())+
-  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  #coord_flip()+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))
-growth_sp10_cii
-
-```
-
-## Top 15 species split by leaf habit
-
-```{r evg/dec, echo=F}
-
-sp_top15<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA", 
-            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA",
-            "GARCSP", "CROTRO", "GARUPI", "NEOLOB", "PHOEPA")
-
-#these 15 species had the most dendrobands and together made up 51.22% of the agb in 2009
-
-#below values eyeballed from Williams et al. 2008 Fig 1.
-sp_top15_phenoscore<-c(2.5, 0.8, 1.3, 4, 0.7, 
-                       1, 1.25, 1.27, 0.2, 0.3,
-                       0.4, 0.65, 3.95, 0.15, 0.25)
-
-#no info on Alphonsea ventricosa in the paper. 
-#Apparently evergreen, so used 1 as value - middle of evergreen range
-
-pheno<-as.data.frame(cbind(sp_top15, sp_top15_phenoscore))
-colnames(pheno)<-c("sp", "phenoscore")
-
-dendro_top15<-dendro.data.noouts[dendro.data.noouts$sp %in% sp_top15,]
-dendro_top15<-merge(dendro_top15, pheno, by="sp")
-
-#categorising as evg and deciduous
-dendro_top15$cat<-ifelse(dendro_top15$phenoscore<=2, "evg", "dec")
-
-growth_yr_cat_detrend<-ggplot(dendro_top15, aes(x=factor(cat), y=inc.resid,
-                                              #group=c(year, sp),
-                                              #group=sp,
-                                              fill=factor(cat)))+ 
-  geom_violin(position = "dodge")+ 
-  ylab("Residuals of annualised diameter increment")+theme_bw()+
-  facet_grid(.~year)+
-  geom_hline(yintercept = 0)+
-  scale_fill_manual(values=c("sienna4", "darkgreen"))+
-  theme(axis.text.x = element_blank())
-
-growth_yr_cat_detrend
-
 ```

### 2024-07-02 scripts/explore_analysis.Rmd
updated figures and explore analysis

@@
 
+spei_hkk<-read.csv("../data/climate/SPEI_HKK_from_GEE.csv")
+colnames(spei_hkk)[1]<-"time"
+
+#making dates manageable
+spei_hkk<- spei_hkk %>%
+  dplyr::mutate(time=as.Date(time, format = "%b %d, %Y"),
+                year=year(time),
+                month=month(time))
+
 pre_yr_census<-ggplot()+
-  geom_area(data=chirps_hkk, aes(x=time, y=precipitation), fill="blue", alpha=0.5)+
-  #geom_line(data=climate_data_orig[climate_data_orig$variable=="tmx",], aes(x=as.Date(paste0(year, "-", month, "-", "01"), format = "%Y-%m-%d"), y=value*5), col="red")+
+  #geom_area(data=chirps_hkk, aes(x=time, y=precipitation), fill="blue", alpha=0.5)+
+  geom_line(data=chirps_hkk, aes(x=time, y=precipitation), col="blue", alpha=0.5)+
+  geom_line(data=spei_hkk, aes(x=time, y=(100*SPEI_01_month-250)), col="red")+
   geom_rect(data=census_int, aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf), 
@@
                                  label= paste0("C", 1:29), angle=90), size=2)+
-  # scale_y_continuous(    # Features of the first axis
-  #   name = "monthly rainfall (mm)",
-  #   
-  #   # Add a second axis and specify its features
-  #   sec.axis = sec_axis(~.*0.2, name="maximum temperature (Celsius °)"))+
-  xlab("")+theme_bw()+ ylab("total monthly precipitation (mm)")+
-  ggtitle("HKK dendroband censuses and CHIRPS rainfall data")+
+   scale_y_continuous(    # Features of the first axis
+     name = "monthly rainfall (mm)",
+
+     # Add a second axis and specify its features
+     sec.axis = sec_axis(~(.+250)*0.01, name="SPEI"))+
+  xlab("")+theme_bw()+ ylab("total monthly precipitation (mm) from CHIRPS")+
+  ggtitle("HKK dendroband censuses and climate")+
   #scale_x_continuous(n.breaks=15)+
@@
          inc_annual=(inc/as.numeric(tdif))*365,
-         inc.time=year(Date2)-lag(year(Date2)),
+         inc.time=Cno-lag(Cno),
          #removing increment value if the previous measure was before the last year
          #inc=ifelse(inc.time==1, inc, NA),
-         inc_annual=ifelse(inc.time==1, inc_annual, NA),
+         inc_annual=ifelse(inc.time==2, inc_annual, NA),
          treeID=paste0("HKK.", Tag))%>%
@@
     diainc.scaled = scale(inc_annual),
-    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))#%>%
-  #filter(out.inc==0)
+    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))%>%
+  filter(out.inc==0)
 
@@
 
+dendro_2019<- dendro.data.noouts %>%
+  filter(Cno %in% c(19, 21, 23))%>%
+  group_by(Tag) %>% #grouping by treeID to calculate increment
+  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
+         cii_min1=dplyr::lag(Cii, n=1, default=NA, order_by=Tag),
+         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
+         inc=calcDBH-calcDBH_min1,
+         inc_annual=(inc/as.numeric(tdif))*365,
+         inc.time=Cno-lag(Cno),
+         #removing increment value if the previous measure was before the last year
+         #inc=ifelse(inc.time==1, inc, NA),
+         inc_annual=ifelse(inc.time==2, inc_annual, NA),
+         treeID=paste0("HKK.", Tag))%>%
+  filter(!is.na(inc))%>%
+  filter(!is.na(inc_annual))%>%
+  ungroup()%>%
+  dplyr::mutate(
+    diainc.scaled = scale(inc_annual),
+    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))%>%
+  filter(out.inc==0)
+
+quantiles_2018<-quantile(dendro_2019$inc_annual[dendro_2019$Cno==21], 
+                         probs=c(0.25, 0.5, 0.75, 1))
+
+quantiles_2019<-quantile(dendro_2019$inc_annual[dendro_2019$Cno==23], 
+                         probs=c(0.25, 0.5, 0.75, 1))
+
+dendro_2019_plot<-ggplot(data = dendro_2019, aes(x=inc_annual, group=factor(Cno), fill=factor(Cno), col=factor(Cno))) +
+  geom_density(alpha=0.5) +
+  scale_color_manual(values=c("forestgreen", "brown"))+
+  scale_fill_manual(values=c("forestgreen", "brown"))+
+  geom_vline(xintercept=quantiles_2018, col="forestgreen")+
+  geom_text(aes(x=0.75, y=1, label="2018"), col="forestgreen", size=3)+
+  geom_text(aes(x=0.75, y=1.5, label="2019"), col="brown", size=3)+
+  geom_vline(xintercept=quantiles_2019, col="brown")+ 
+  xlab("diameter increment (cm)") + ylab("density")+
+  guides(col="none", fill="none")+
+  theme_bw()
+
 library(gridExtra)
 png("../doc/display/increments_raw.png", width=8, height=4, units="in", res=300)
-grid.arrange(dendro_2010_plot, dendro_2015_plot, nrow=1)
+grid.arrange(dendro_2010_plot, dendro_2015_plot, dendro_2019_plot, nrow=1)
 dev.off()
@@
 ```
+
+```{r long-term growth rates}
+
+library(ggplot2)
+
+pat<-as.character(read.table("../data/HKK-dendro/git_pat.txt")[1,1])
+
+remotes:::download("../data/HKK-dendro/stem1.RData",
+                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem1.rdata",
+                   auth_token = pat)
+remotes:::download("../data/HKK-dendro/stem2.RData",
+                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem2.rdata",
+                   auth_token = pat)
+remotes:::download("../data/HKK-dendro/stem3.RData",
+                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem3.rdata",
+                   auth_token = pat)
+remotes:::download("../data/HKK-dendro/stem4.RData",
+                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem4.rdata",
+                   auth_token = pat)
+
+load("../data/HKK-dendro/stem1.RData")
+load("../data/HKK-dendro/stem2.RData")
+load("../data/HKK-dendro/stem3.RData")
+load("../data/HKK-dendro/stem4.RData")
+
+hkk_census<-bind_rows(hkk.stem1, hkk.stem2, hkk.stem3, hkk.stem4)
+
+hkk_census <- hkk_census %>%
+  filter(status=="A")%>%
+  filter(tag %in% dendro.data.noouts$Tag)%>%
+  group_by(StemTag)%>%
+  dplyr::mutate(dbhmin1 = dplyr::lag(dbh, n=1, default=NA, order_by=StemTag),
+                inc_annual = (dbh-dbhmin1)/5)%>%
+  filter(!is.na(dbhmin1))%>%
+  dplyr::summarise(avg_inc = mean(inc_annual, na.rm=T),
+                   sd_inc = sd(inc_annual, na.rm=T))
+
+inc_census<-ggplot(data = hkk_census, aes(x=avg_inc)) +
+  geom_density(alpha=0.5) +
+  xlab("mean annual increment (cm)")+
+  theme_bw()
+
+png("../doc/display/census_increment.png", width=4, height=4, units="in", res=300)
+inc_census
+dev.off()
+
+```
+
 ```{r drought resistance}
@@
 
+#2010 drought--------
+
 resistance_2010 <- dendro_2010 %>%
@@
          resistance.div = inc_annual/inc_min1,
-         resistance.dif = inc_annual-inc_min1)%>%
+         resistance.dif = inc_annual-inc_min1,
+         resistance.prop = (inc_annual-inc_min1)/inc_min1)%>%
   filter(!is.na(inc_min1), !is.infinite(resistance.div))
 
-resistance_2010<-merge(resistance_2010, dec_williams, by="sp", all.x=T)
+resistance_2010<-merge(resistance_2010, select(dec_williams, -"X"), by="sp", all.x=T)
 
@@
 
+#2015 drought---------
+
 resistance_2015 <- dendro_2015 %>%
@@
          resistance.div = inc_annual/inc_min1,
-         resistance.dif = inc_annual-inc_min1)%>%
+         resistance.dif = inc_annual-inc_min1,
+         resistance.prop = (inc_annual-inc_min1)/inc_min1)%>%
   filter(!is.na(inc_min1), !is.infinite(resistance.div))
 
-resistance_2015<-merge(resistance_2015, dec_williams, by="sp", all.x=T)
+resistance_2015<-merge(resistance_2015, select(dec_williams, -"X"), by="sp", all.x=T)
 
@@
 
-png("../doc/display/resistance_dif.png", width=8, height=4, units="in", res=300)
-grid.arrange(resistance_2010_dif, resistance_2015_dif, nrow=1)
+#2019 drought--------------
+
+resistance_2019 <- dendro_2019 %>%
+  group_by(Tag) %>% #grouping by treeID to calculate increment
+  mutate(inc_min1=dplyr::lag(inc_annual, n=1, default=NA, order_by=Tag),
+         resistance.div = inc_annual/inc_min1,
+         resistance.dif = inc_annual-inc_min1,
+         resistance.prop = (inc_annual-inc_min1)/inc_min1)%>%
+  filter(!is.na(inc_min1), !is.infinite(resistance.div))
+
+resistance_2019<-merge(resistance_2019, select(dec_williams, -"X"), by="sp", all.x=T)
+
+resistance_2019_div<-ggplot(data = resistance_2019, aes(x=resistance.div)) +
+  geom_density(alpha=0.5) +
+  theme_bw()
+
+resistance_2019_dif<-ggplot(data = resistance_2019, aes(x=resistance.dif)) +
+  geom_density(alpha=0.5) +
+  theme_bw()
+
+png("../doc/display/resistance_dif.png", width=4, height=8, units="in", res=300)
+grid.arrange(resistance_2010_dif, resistance_2015_dif, resistance_2019_dif, nrow=3)
 dev.off()
 
+```
+
+```{r resistance map}
+
+quantiles_2010_dif<-quantile(resistance_2010$resistance.dif, 
+                         probs=c(0.25, 0.5, 0.75, 1))
+
+quantiles_2015_dif<-quantile(resistance_2015$resistance.dif, 
+                         probs=c(0.25, 0.5, 0.75, 1))
+
+quantiles_2019_dif<-quantile(resistance_2019$resistance.dif, 
+                         probs=c(0.25, 0.5, 0.75, 1))
+
+resistance_2010$quantile<-ifelse(resistance_2010$resistance.dif<quantiles_2010_dif[1], 1, ifelse(resistance_2010$resistance.dif<quantiles_2010_dif[2], 2,
+            ifelse(resistance_2010$resistance.dif<quantiles_2010_dif[3], 3, 4))
+
+                                 )
+
+resistance_2015$quantile<-ifelse(resistance_2015$resistance.dif<quantiles_2015_dif[1], 1, ifelse(resistance_2015$resistance.dif<quantiles_2015_dif[2], 2,
+            ifelse(resistance_2015$resistance.dif<quantiles_2015_dif[3], 3, 4))
+
+                                 )
+
+resistance_2019$quantile<-ifelse(resistance_2019$resistance.dif<quantiles_2019_dif[1], 1, ifelse(resistance_2019$resistance.dif<quantiles_2019_dif[2], 2,
+            ifelse(resistance_2019$resistance.dif<quantiles_2019_dif[3], 3, 4))
+
+                                 )
+
+resistance_all<-rbind(resistance_2010, resistance_2015, resistance_2019)
+resistance_all$yr<-c(rep(2010, nrow(resistance_2010)), rep(2015, nrow(resistance_2015)), rep(2019, nrow(resistance_2019)))
+
+resistance_all<-merge(resistance_all, select(trees, -"SPCODE.UPDATE"), by="Tag")
+
+resistance_map<-ggplot(resistance_all, aes(x=X, y=Y, size=as.numeric(DBH.04), col=quantile))+
+  geom_point(alpha=0.5)+
+   scale_size(range = c(.1, 10), name="DBH (cm)")+
+  guides(color=guide_legend(title="resistance"))+
+  scale_colour_gradient(low="red", high="blue") +
+  facet_grid(yr~.)+
+  theme_minimal()+
+  coord_fixed()
+
+png("../doc/display/resistance_map.png", width=8, height=10, units="in", res=300)
+resistance_map
+dev.off()
 ```
@@
 
-resistance_all<-rbind(resistance_2010, resistance_2015)
-resistance_all$yr<-c(rep(2010, nrow(resistance_2010)), rep(2015, nrow(resistance_2015)))
+model_2019<-glm(resistance.dif ~ calcDBH_min1 + cii_min1 + williams_dec+
+                  calcDBH_min1*cii_min1 + 
+                  calcDBH_min1*williams_dec+
+                  cii_min1*williams_dec,
+                data=resistance_2019)
+
+summary(model_2019)
+
+resistance_all<-rbind(resistance_2010, resistance_2015, resistance_2019)
+resistance_all$yr<-c(rep(2010, nrow(resistance_2010)), rep(2015, nrow(resistance_2015)), rep(2019, nrow(resistance_2019)))
 
 resistance_model<-resistance_all %>%
-  select(resistance.dif, calcDBH_min1, cii_min1, williams_dec, yr)
+  select(resistance.dif, calcDBH_min1, cii_min1, williams_dec, yr, quantile)
 
@@
 
-library(nlme)
-library(ggplot2)
+model_all_cat<-lme(quantile ~ calcDBH_min1 + cii_min1 + williams_dec+
+                  calcDBH_min1*cii_min1 + 
+                  calcDBH_min1*williams_dec+
+                  cii_min1*williams_dec,
+                random=~1|yr,
+                data=resistance_model)
+
+model_all_cat
+anova(model_all_cat)
+```
+
+```{r predictions, echo=F}
+
+library(sjPlot)
+
+plot_model(model_all, type="pred", terms=c("calcDBH_min1","williams_dec"), show.data = T)
+plot_model(model_all, type="pred", terms=c("cii_min1"), show.data = T)
+plot_model(model_all, type="pred", terms=c("williams_dec"), show.data = T)
+plot_model(model_all, type="pred", terms=c("calcDBH_min1"), show.data = T)
+
+plot_model(model_all, type = "est")+
+  theme_bw()+ylim(c(-0.25, 0.25))+
+  geom_hline(yintercept = 0)
+
+plot_model(model_all_cat, type = "est")+
+  theme_bw()+ylim(c(-0.25, 0.25))+
+  geom_hline(yintercept = 0)
+
+plot_model(model_all_cat, type="pred", terms=c("calcDBH_min1","williams_dec"), show.data = T)
 
@@
   labs(x = "Deciduousness", y = "Drought resistance", title = "Predictions from LME Model")
-
-```
-
-```{r maps of species with tree ring records, echo=F, eval=F}
-library(ggplot2)
-
-pat<-as.character(read.table("git_pat.txt")[1,1])
-
-remotes:::download("stem1.RData",
-                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem1.rdata",
-                   auth_token = pat)
-remotes:::download("stem2.RData",
-                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem2.rdata",
-                   auth_token = pat)
-remotes:::download("stem3.RData",
-                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem3.rdata",
-                   auth_token = pat)
-remotes:::download("stem4.RData",
-                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem4.rdata",
-                   auth_token = pat)
-
-load("stem1.RData")
-load("stem2.RData")
-load("stem3.RData")
-load("stem4.RData")
-
-hkk_census<-bind_rows(hkk.stem1, hkk.stem2, hkk.stem3, hkk.stem4)
-
-hkk_census <- hkk_census %>%
-  filter(status=="A")
-
-hkk_census_ring_sp<-hkk_census %>%
-  filter(CensusID==3)%>%
-  filter(sp %in% ring.sp)
-
-hkk_census_sp <- hkk_census %>%
-  filter(CensusID==4)%>%
-  #group_by(sp, tag, CensusID) %>%
-  #dplyr::summarise(tot.agb=sum(agb, na.rm=T)) %>%
-  group_by(sp, tag)%>%
-  dplyr::summarise(tot.agb=sum(agb, na.rm=T)) %>%
-  group_by(sp) %>%
-  dplyr::summarise(n.tree=length(unique(tag)),
-                   sum.agb=sum(tot.agb))%>%
-  ungroup()%>%
-  dplyr::mutate(pct.agb=sum.agb*100/sum(sum.agb, na.rm=T))%>%
-  arrange(-pct.agb)%>%
-  dplyr::mutate(pct.agb.cum=cumsum(pct.agb))
-
-sp_top15<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA",
-            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA",
-            "GARCSP", "CROTRO", "GARUPI", "NEOLOB", "PHOEPA")
-
-#pct agb of the top 15 dendro species
-sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% sp_top15])
-
-```
-
-## Fastest and slowest in 2015 and 2016
-
-```{r echo=F}
-
-dendro.data.noouts <- dendro.data.noouts %>%
-  group_by(year) %>% 
-  dplyr::mutate(growth_quantile = ntile(diainc, 4),
-                resid_quantile = ntile (inc.resid, 4))
-
-growth_2015 <- dendro.data.noouts %>%
-  filter(year==2015)
-
-quantiles_2015<-quantile(growth_2015$diainc, probs=c(0.25, 0.5, 0.75, 1))
-
-growth_2015_plot<-ggplot(growth_2015, aes(x=diainc)) +
-  geom_density(fill="brown", col="brown", alpha=0.5) +
-  geom_vline(xintercept=quantiles_2015, col="brown")+ theme_bw()
-
-growth_2014 <- dendro.data.noouts %>%
-  filter(Cno==11)
-
-quantiles_2014<-quantile(growth_2014$diainc, probs=c(0.25, 0.5, 0.75, 1))
-
-growth_2016 <- dendro.data.noouts %>%
-  filter(year==2016)
-
-quantiles_2016<-quantile(growth_2016$diainc, probs=c(0.25, 0.5, 0.75, 1))
-
-growth_2016_plot<-ggplot(growth_2016, aes(x=diainc)) +
-  geom_density(fill="brown", col="brown", alpha=0.5) +
-  geom_vline(xintercept=quantiles_2016, col="brown")+ theme_bw()
-
-growth_2017 <- dendro.data.noouts %>%
-  filter(year==2017)
-
-quantiles_2017<-quantile(growth_2017$diainc, probs=c(0.25, 0.5, 0.75, 1))
-
-growth_2017_plot<-ggplot(growth_2017, aes(x=diainc)) +
-  geom_density(fill="brown", col="brown", alpha=0.5) +
-  geom_vline(xintercept=quantiles_2016, col="brown")+ theme_bw()
-
-growth_2014_2015_plot<-ggplot() +
-  geom_density(data = growth_2014, aes(x=diainc), fill="forestgreen", col="forestgreen", alpha=0.5) +
-  geom_vline(xintercept=quantiles_2014, col="forestgreen")+
-  geom_density(data= growth_2015, aes (x=diainc), fill="brown", col="brown", alpha=0.5) +
-  geom_text(aes(x=0.75, y=1, label="2014"), col="forestgreen", size=3)+
-  geom_text(aes(x=0.75, y=1.5, label="2015"), col="brown", size=3)+
-  geom_vline(xintercept=quantiles_2015, col="brown")+ 
-  xlab("diameter increment (cm)") + ylab("density")+
-  theme_bw()
-
-dendro.data.noouts$year_fac<-as.character(dendro.data.noouts$year)
-dendro.data.noouts$year_fac<-factor(dendro.data.noouts$year_fac, levels=as.character(rev(2009:2023)))
-
-growth_all_years<-ggplot(dendro.data.noouts, aes(x = diainc, y = year_fac, 
-                       fill = stat(x))) +
-  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
-  scale_fill_viridis(name = "growth", option = "H", direction=-1) +
-  labs(title = 'tree growth in HKK (cm)', y="year") +
-  #theme_ipsum() +
-  geom_vline(xintercept=0, lty=2)+
-  theme_bw()+
-    theme(
-      legend.position="none",
-      panel.spacing = unit(0.1, "lines"),
-      strip.text.x = element_text(size = 8)
-    )
-
-growth_all_years
-
-png("../doc/display/growth_2014_2015.png", width=4, height=4, units="in", res=300)
-growth_2014_2015_plot
-dev.off()
-
-png("../doc/display/growth_allyrs.png", width=4, height=4, units="in", res=300)
-growth_all_years
-dev.off()
 
@@
 ```
-
-## Size classes
-
-```{r echo=F}
-
-size_top10<-ggplot(dendro.top10.dat, aes(x=dbh))+
-  geom_density()+
-  geom_vline(aes(xintercept=35))+
-  geom_vline(aes(xintercept=75))
-
-size_top10
-
-#trying three categories
-# <50, 50-100, >100
-
-dendro.top10.dat<-dendro.top10.dat %>%
-  dplyr::mutate(size=ifelse(dbh<35, "< 35 cm dbh",
-                       ifelse(dbh<70, "35-75 cm dbh", ">75 cm dbh")))
-
-growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(sp, levels=sp_top10), y=diainc,
-                                              #color=factor(sp),
-                                          #group=factor(size, levels=c("small", "medium", "large")),
-                                              fill=factor(sp, levels=sp_top10)))+ 
-  #geom_violin(position = "dodge")+ 
-  geom_boxplot(position = "dodge")+ 
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
-  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-  facet_grid(factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh"))~.)+
-      theme(axis.text.y = element_blank())+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  ggtitle("HKK dendroband data from 2008 to 2023")+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))+coord_flip()
-
-growth_sp10
-
-growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh")), y=diainc,
-                                              #color=factor(sp),
-                                          #group=factor(size, levels=c("small", "medium", "large")),
-                                              fill=factor(sp, levels=sp_top10)))+
-
-  geom_violin(position = "dodge")+ 
-  geom_boxplot(position = "dodge", col="black", fill="white", width=0.1)+ 
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
-  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-  #facet_grid(factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh"))~.)+
-      theme(axis.text.x = element_text(angle=45))+
-  xlab("")+
-  facet_grid(.~factor(sp, levels=sp_top10))+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  ggtitle("HKK dendroband data from 2008 to 2023")+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))#+coord_flip()
-
-growth_sp10
-
-```
-
-## Canopy illumination
-
-```{r echo=F}
-
-cii_top10<-ggplot(dendro.top10.dat, aes(x=cii))+
-  geom_histogram()
-
-cii_top10
-
-#trying three categories
-# <50, 50-100, >100
-
-#dendro.top10.dat<-dendro.top10.dat %>%
-#  dplyr::mutate(size=ifelse(dbh<50, "small",
-#                       ifelse(dbh<100, "medium", "large")))
-
-growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
-                        aes(x=factor(sp, levels=sp_top10), y=diainc,
-                                              #color=factor(sp),
-                                          #group=factor(size, levels=c("small", "medium", "large")),
-                                              fill=factor(sp, levels=sp_top10)))+ 
-  #geom_violin(position = "dodge")+ 
-  geom_boxplot(position = "dodge")+ 
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
-  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-  facet_grid(cii~.)+
-      theme(axis.text.y = element_blank())+
-  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))+coord_flip()
-
-growth_sp10_cii
-
-growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
-                        aes(
-                          #x=factor(sp, levels=sp_top10), 
-                          x=factor(cii), 
-                            y=diainc,
-                                              #color=factor(sp),
-                                          #group=factor(size, levels=c("small", "medium", "large")),
-                                              fill=factor(sp, levels=sp_top10)))+ 
-  #geom_violin(position = "dodge")+ 
-  geom_boxplot(position = "dodge")+ 
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
-  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-  facet_grid(.~factor(sp, levels = sp_top10))+
-      theme(axis.text.y = element_blank())+
-  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  #coord_flip()+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))
-growth_sp10_cii
-
-```
-## spatial patterns
-
-```{r spatial patterns, echo=F}
-
-trees.top10 <- trees.top10 %>%
-  dplyr::mutate(treeID=paste0("HKK.", Tag),
-                plotID=paste0(Col, "_", Row))
-
-dendro.top10.dat<-merge(dendro.top10.dat, 
-                        dplyr::select(trees.top10, c("Col", "Row", "treeID", "X", "Y", "plotID")), 
-                        by="treeID", all.x=T)
-
-dendro.top10.dat<-dendro.top10.dat %>%
-  dplyr::mutate(plotID=paste0(Col, "_", Row))
-
-census_spatial<-hkk_census %>%
-  group_by(quadrat) %>%
-  dplyr::summarise(sp.no=length(unique(sp)),
-                   sp.top.10 = length(unique(sp %in% sp_top10)))%>%
-  dplyr::mutate(Col=as.numeric(str_sub(quadrat, 1, 2)),
-                Row=as.numeric(str_sub(quadrat, 3, 4)))
-
-tiles<-ggplot(census_spatial, aes(x=Col, y=Row, fill=sp.no))+
-  geom_tile()+
-  #facet_wrap(~year)+
-  coord_fixed()+
-  #xlim(c(0,49))+ylim(c(0, 24))+
-  theme_bw()
-
-tiles
-
-tiles<-ggplot(census_spatial, aes(x=Col, y=Row, fill=sp.top.10))+
-  geom_tile()+
-  #facet_wrap(~year)+
-  coord_fixed()+
-  #xlim(c(0,49))+ylim(c(0, 24))+
-  theme_bw()
-
-tiles
-
-spatial_summary<- dendro.top10.dat %>%
-  group_by(Col, Row, plotID, sp, year) %>%
-  dplyr::summarise(inc.sum=sum(inc.resid, na.rm=T)) %>%
-  group_by(Col, Row, plotID, year) %>%
-  dplyr::summarise(inc.diff=var(inc.sum, na.rm=T))
-
-tiles<-ggplot(spatial_summary, aes(x=Col, y=Row, fill=log(inc.diff, 10)))+
-  geom_tile()+
-  facet_wrap(~year)+coord_fixed()+
-  xlim(c(0,49))+ylim(c(0, 24))+
-  theme_bw()
-
-tiles
-
-```
-
-## evergreen/dec
-
-```{r evg/dec, echo=F}
-
-sp_top15<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA", 
-            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA",
-            "GARCSP", "CROTRO", "GARUPI", "NEOLOB", "PHOEPA")
-
-#these 15 species had the most dendrobands and together made up 51.22% of the agb in 2009
-
-#below values eyeballed from Williams et al. 2008 Fig 1.
-sp_top15_phenoscore<-c(2.5, 0.8, 1.3, 4, 0.7, 
-                       1, 1.25, 1.27, 0.2, 0.3,
-                       0.4, 0.65, 3.95, 0.15, 0.25)
-
-#no info on Alphonsea ventricosa in the paper. 
-#Apparently evergreen, so used 1 as value - middle of evergreen range
-
-pheno<-as.data.frame(cbind(sp_top15, sp_top15_phenoscore))
-colnames(pheno)<-c("sp", "phenoscore")
-
-dendro.top10.dat<-merge(dendro.top10.dat, pheno, by="sp")
-
-#categorising as evg and deciduous
-dendro.top10.dat$cat<-ifelse(dendro.top10.dat$phenoscore<=1, "evg", "dec")
-
-growth_yr_cat_detrend<-ggplot(dendro_top15, aes(x=factor(cat), y=inc.resid,
-                                              #group=c(year, sp),
-                                              #group=sp,
-                                              fill=factor(cat)))+ 
-  geom_violin(position = "dodge")+ 
-  ylab("Residuals of annualised diameter increment")+theme_bw()+
-  facet_grid(.~year)+
-  geom_hline(yintercept = 0)+
-  scale_fill_manual(values=c("sienna4", "darkgreen"))+
-  theme(axis.text.x = element_blank())
-
-growth_yr_cat_detrend
-
-growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
-                        aes(
-                          #x=factor(sp, levels=sp_top10), 
-                          x=factor(cii), 
-                            y=diainc,
-                                              #color=factor(sp),
-                                          #group=factor(size, levels=c("small", "medium", "large")),
-                                              fill=factor(sp, levels=sp_top10)))+ 
-  #geom_violin(position = "dodge")+ 
-  geom_boxplot(position = "dodge")+ 
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
-  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-  facet_grid(cat~factor(sp, levels = sp_top10))+
-      theme(axis.text.y = element_blank())+
-  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  #coord_flip()+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))
-growth_sp10_cii
-
-```
-
-## Repeat all steps for dry censuses
-
-```{r echo=F}
-
-#remove all data with decimal errors, misplaced decimal
-
-`%nin%` <- Negate(`%in%`)
-
-dendro.data<-dendrobyband %>%
-  dplyr::rename(sp=SPCODE.UPDATE)%>%
-  mutate(year=year(Date2)) %>%
-  #filter(year!=2209)%>% #quick fix to remove bad dates - there are 30 entries
-  #filter(start_date <= year & year <= end_date) %>%
-  filter(Cno %nin% wet_census)%>%
-  filter(is.na(DBH1_flag), DBH1_error=="none",
-         win.decimal.error=="none", is.na(win.flag))%>%
-  group_by(year, sp, Tag) %>%
-  dplyr::summarise(calcDBH=calcDBH[which(n.dendro==min(n.dendro))],
-                   Date2=Date2[which(n.dendro==min(n.dendro))],
-                   Cii=max(Cii, na.rm=T))
-
-dendro.data_all<-dendro.data%>%
-  group_by(Tag) %>% #grouping by treeID to calculate increment
-  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
-         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
-         inc=calcDBH-calcDBH_min1,
-         inc_annual=(inc/as.numeric(tdif))*365,
-         inc.time=year-lag(year),
-         #removing increment value if the previous measure was before the last year
-         inc_annual=ifelse(inc.time==1, inc_annual, NA),
-         #has.out=ifelse(1 %in% out, 1, 0),
-         treeID=paste0("HKK.", Tag))%>%
-  filter(!is.na(inc))%>%
-  ungroup()%>%
-  filter(!is.na(inc_annual))%>%
-  ungroup()%>%
-  #filter(start_date <= year & year <= end_date) %>%
-  group_by(sp, treeID, year)%>%
-  dplyr::summarise(
-    diainc = mean(inc_annual, na.rm=T),
-    Cii=max(Cii, na.rm=T))
-
-#first fitting a line through the time series
-obs.model.dbh<-nlme::lme(calcDBH~year, random=~1|Tag, data=dendro.data,
-                         na.action=na.omit,
-                         method="ML", control=(msMaxIter=100)) #had to increase the number of iterations
-
-#standardised residuals calculated https://stat.ethz.ch/R-manual/R-devel/library/nlme/html/residuals.lme.html
-
-std.res.dbh<-residuals(obs.model.dbh, type="pearson")
-hist(std.res.dbh)
-dendro.data$std.res<-rep(NA, nrow(dendro.data))
-dendro.data$std.res[!is.na(dendro.data$calcDBH)]<-std.res.dbh
-#defining all observation with abs(residual) >3 as an outlier
-dendro.data$out<-factor(ifelse(abs(dendro.data$std.res)<=3, 0, 1))
-
-#find the trees which have outlier observations
-outs.Tag<-unique(dendro.data$Tag[dendro.data$out==1])
-
-#plot trees that have outliers
-ggplot(dendro.data[dendro.data$Tag %in%outs.Tag,],
-       aes(x = year, y = calcDBH, col=factor(out), group = Tag))+
-  geom_line(alpha=0.5)+geom_point()+
-  scale_color_manual(values=c("grey20", "red", "grey"))
-
-dendro.data.noouts<-dendro.data %>%
-  mutate(dbh3=ifelse(abs(std.res)<=3, calcDBH, NA)) %>% #create an outlier removed column
-  group_by(Tag) %>% #grouping by treeID to calculate increment
-  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
-         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
-         inc=calcDBH-calcDBH_min1,
-         inc_annual=(inc/as.numeric(tdif))*365,
-         inc.time=year-lag(year),
-         #removing increment value if the previous measure was before the last year
-         #inc=ifelse(inc.time==1, inc, NA),
-         inc_annual=ifelse(inc.time==1, inc_annual, NA),
-         has.out=ifelse(1 %in% out, 1, 0),
-         treeID=paste0("HKK.", Tag))%>%
-  filter(!is.na(inc))%>%
-  ungroup()%>%
-  filter(!is.na(inc_annual))%>%
-  ungroup()%>%
-  #filter(start_date <= year & year <= end_date) %>%
-  group_by(sp, treeID, year)%>%
-  dplyr::summarise(
-    diainc = mean(inc_annual, na.rm=T),
-    calcDBH_min1 = mean(calcDBH_min1, na.rm=T),
-    dbh=mean(calcDBH, na.rm=T),
-    out = ifelse(1 %in% out, 1, 0),
-    has.out = ifelse(1 %in% has.out, 1, 0), 
-    cii=max(Cii, na.rm=T))%>%
-  filter(out==0)%>%
-  ungroup() %>%
-  dplyr::mutate(
-    diainc.scaled = scale(diainc),
-    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))%>%
-  filter(out.inc==0)
-
-```
-
-## Raw diameter increments of all species
-
-```{r echo=FALSE}
-
-library(ggplot2)
-
-growth_yr_withouts<-ggplot(dendro.data_all, aes(x=year, y=diainc, group=year))+
-  geom_violin()+
-  ylab("Annualised diameter increment (cm)")+theme_bw()
-
-growth_yr_withouts
-```
-
-## Removing outliers based (more than 3 SD)
-
-```{r echo=FALSE}
-
-growth_yr_noouts<-ggplot(dendro.data.noouts, aes(x=year, y=diainc, group=year))+ geom_violin()+ ylab("Annualised diameter increment (cm)")+theme_bw()
-
-growth_yr_noouts
-```
-
-## Tree ring species
-
-- Four species with tree ring chronologies in HKK
-
-```{r ring trees, echo=F}
-#These are species that have been cored near HKK and reported in Zuidema et al., Anderson-Teixeira et al., etc
-ring.sp<-c("AFZEXY", "MELIAZ", "TOONCI", "CHUKTA", "NEOLOB")
-
-trees.with.ring<-trees[trees$SPCODE.UPDATE %in% ring.sp,]
-
-dendro.ring<-dendro.data.noouts %>%
-  filter(sp %in% ring.sp)
-
-hkk_sp_ring<-trees.with.ring %>%
-  group_by(SPCODE.UPDATE)%>%
-  dplyr::summarise(n.tree=length(SPCODE.UPDATE))
-knitr::kable(hkk_sp_ring, format='html', table.attr="style='width:50%;'")
-
-```
-
-```{r maps of species with tree ring records, echo=F}
-library(ggplot2)
-
-pat<-as.character(read.table("git_pat.txt")[1,1])
-
-# remotes:::download("stem1.RData",
-#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem1.rdata",
-#                    auth_token = pat)
-# remotes:::download("stem2.RData",
-#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem2.rdata",
-#                    auth_token = pat)
-# remotes:::download("stem3.RData",
-#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem3.rdata",
-#                    auth_token = pat)
-# remotes:::download("stem4.RData",
-#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem4.rdata",
-#                    auth_token = pat)
-#
-
-load("stem1.RData")
-load("stem2.RData")
-load("stem3.RData")
-load("stem4.RData")
-
-hkk_census<-bind_rows(hkk.stem1, hkk.stem2, hkk.stem3, hkk.stem4)
-
-hkk_census <- hkk_census %>%
-  filter(status=="A")
-
-hkk_census_ring_sp<-hkk_census %>%
-  filter(CensusID==3)%>%
-  filter(sp %in% ring.sp)
-
-hkk_census_sp <- hkk_census %>%
-  filter(CensusID==4)%>%
-  #group_by(sp, tag, CensusID) %>%
-  #dplyr::summarise(tot.agb=sum(agb, na.rm=T)) %>%
-  group_by(sp, tag)%>%
-  dplyr::summarise(tot.agb=sum(agb, na.rm=T)) %>%
-  group_by(sp) %>%
-  dplyr::summarise(n.tree=length(unique(tag)),
-                   sum.agb=sum(tot.agb))%>%
-  ungroup()%>%
-  dplyr::mutate(pct.agb=sum.agb*100/sum(sum.agb, na.rm=T))%>%
-  arrange(-pct.agb)%>%
-  dplyr::mutate(pct.agb.cum=cumsum(pct.agb))
-
-sp_top15<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA",
-            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA",
-            "GARCSP", "CROTRO", "GARUPI", "NEOLOB", "PHOEPA")
-
-#pct agb of the top 15 dendro species
-sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% sp_top15])
-
-ring.sp.map.census<-ggplot(hkk_census_ring_sp, aes(x=gx, y=gy, size=as.numeric(dbh/10), color=factor(sp)))+
-  geom_point(alpha=0.5)+
-   scale_size(range = c(.1, 10), name="DBH (cm)")+
-  guides(color=guide_legend(title="Species"))+
-  #scale_colour_gradient(low="orange", high="purple") +
-  theme_minimal()+
-  coord_fixed()
-
-ring.sp.map.census
-
-ring.sp.map<-ggplot(trees.with.ring, aes(x=X, y=Y, size=as.numeric(DBH.04), color=factor(SPCODE.UPDATE)))+
-  geom_point(alpha=0.5)+
-   scale_size(range = c(.1, 10), name="DBH (cm)")+
-  guides(color=guide_legend(title="Species"))+
-  #scale_colour_gradient(low="orange", high="purple") +
-  theme_minimal()+
-  coord_fixed()
-
-ring.sp.map
-```
-Species with tree rings contribute `r round(sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% ring.sp]),2)` % aboveground biomass in 2004
-
-## Dendrobands on ring species
-
-```{r echo=F}
-
-growth_sp<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
-                                              fill=factor(sp),
-                                              color=factor(sp)))+ 
-  geom_violin(position = "dodge")+ 
-  geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #facet_grid(.~year)+
-  #geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
-  geom_hline(yintercept = median(dendro.ring$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-      theme(axis.text.x = element_blank())+
-  ggtitle("HKK dendroband data from 2008 to 2023")+
-  geom_text(aes(label=paste0("median=\n", round(median(dendro.ring$diainc, na.rm=T),2), " cm"), 
-                x=0.6, y=median(dendro.ring$diainc, na.rm=T)+0.15), col="black", size=3)+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))
-
-growth_sp
-
-growth_sp_box<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
-                                              #color=factor(sp),
-                                              fill=factor(sp)))+ 
-  geom_boxplot(position = "dodge")+ 
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #facet_grid(.~year)+
-  geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
-  geom_hline(yintercept = median(dendro.ring$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-      theme(axis.text.x = element_blank())+
-  ggtitle("HKK dendroband data from 2008 to 2023")+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))
-
-#growth_sp_box
-
-growth_yr_sp<-ggplot(dendro.ring, aes(x=factor(sp), y=diainc,
-                                              #color=factor(sp),
-                                              fill=factor(sp)))+ 
-  #geom_violin(position = "dodge")+ 
-  geom_boxplot(position="dodge")+
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  facet_grid(.~year)+
-  geom_hline(yintercept = mean(dendro.ring$diainc, na.rm=T))+
-  theme(axis.text.x = element_blank())+
-  guides(color="none")
-
-growth_yr_sp
-
-```
-
-## Detrending by tree with smooth splines
-
-```{r detrending, echo=F}
-
-#trying for one tree
-
-dendro.ring<-dendro.ring %>%
-  group_by(sp, treeID) %>%
-  dplyr::mutate(n.obs=length(treeID))%>%
-  filter(n.obs>=4)%>%
-  ungroup()%>%
-  group_by(sp, treeID)%>%
-  dplyr::mutate(inc.resid=residuals(smooth.spline(x=calcDBH_min1, y=diainc)),
-                inc.spline=fitted(smooth.spline(x=calcDBH_min1, y=diainc)))
-
-growth_sp_detrend<-ggplot(dendro.ring, aes(x=factor(sp), 
-                                           y=inc.resid, 
-                                           #color=factor(sp),
-                                           fill=factor(sp)))+ 
-  #geom_violin()+ 
-  geom_boxplot()+ 
-  ylab("Residuals of annualised diameter increment")+
-  theme_bw()+
-  theme(axis.text.x = element_text(angle=45))
-growth_sp_detrend
-
-growth_yr_detrend<-ggplot(dendro.ring, aes(x=year, y=inc.resid, group=year))+ geom_boxplot()+ 
-  ylab("Residuals of annualised diameter increment")+theme_bw()+
-  theme(axis.text.x = element_text(angle=45))
-growth_yr_detrend
-
-```
-
-## Detrended plots by species
-
-```{r echo=F}
-
-growth_yr_sp_detrend<-ggplot(dendro.ring, aes(x=factor(sp), y=inc.resid,
-                                              #group=c(year, sp),
-                                              #group=sp,
-                                              #color=factor(sp),
-                                              fill=factor(sp)))+ 
-  geom_boxplot(position = "dodge")+ 
-  ylab("Residuals of annualised diameter increment")+theme_bw()+
-  geom_hline(yintercept=0)+
-  facet_grid(.~year)+
-  theme(axis.text.x = element_blank())+
-  guides(color="none")
-
-growth_yr_sp_detrend
-
-```
-
-```{r echo=F}
-
-spag_all<-ggplot(dendro.ring, aes(x=year, y=inc.spline, group=treeID)) +
-  geom_line()+ggtitle("fitted values (smooth spline)")
-
-spag_all
-
-spag_resid<-ggplot(dendro.ring, aes(x=year, y=inc.resid, group=treeID)) +
-  geom_line()+ggtitle("residuals smooth spline")
-
-spag_resid
-```
-## Detrending all species
-
-```{r detrending all, echo=F}
-
-#trying for one tree
-
-dendro.data.noouts<-dendro.data.noouts %>%
-  group_by(sp, treeID) %>%
-  dplyr::mutate(n.obs=length(treeID))%>%
-  filter(n.obs>=4)%>%
-  ungroup()%>%
-  filter(treeID!="HKK.92562")%>% #removing an issue with spline for now
-  group_by(sp, treeID)%>%
-  dplyr::mutate(inc.resid=residuals(smooth.spline(x=calcDBH_min1, y=diainc)),
-                inc.spline=fitted(smooth.spline(x=calcDBH_min1, y=diainc)))
-```
-
-For `r length(unique(dendro.data.noouts$sp))` species with dendrobands, all individual trees detrended for DBH with smooth splines
-
-```{r echo=F}
-growth_yr_detrend_all<-ggplot(dendro.data.noouts, aes(x=year, y=inc.resid, group=year))+ 
-  #geom_violin()+ 
-  geom_boxplot()+ 
-  ylab("Residuals of annualised diameter increment")+theme_bw()+
-  theme(axis.text.x = element_text(angle=45))
-growth_yr_detrend_all
-
-```
-
-## top 10 species
-
-```{r echo=F}
-
-dendro.by.sp<-dendro.data.noouts %>%
-  group_by (sp, year) %>%
-  dplyr::summarise(n.inds=length(sp),
-    mean.inc = mean(inc.resid, na.rm=T),
-                   median.inc = median(inc.resid, na.rm=T),
-                   sd.inc = sd(inc.resid, na.rm=T),
-    lower=mean.inc-sd.inc,
-    upper=mean.inc+sd.inc)%>%
-  filter(n.inds>=10)%>%
-  arrange(year, -median.inc)
-
-sp_top5<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA")
-
-sp_top10<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA", 
-            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA")
-
-dendro.top5<-dendro.by.sp %>%
-  filter(sp %in% sp_top5)
-
-dendro.top5.dat<-dendro.data.noouts %>%
-  filter(sp %in% sp_top5)
-
-dendro.top10.dat<-dendro.data.noouts %>%
-  filter(sp %in% sp_top10)
-
-hkk_census_top10_sp<-hkk_census %>%
-  filter(CensusID==3)%>%
-  filter(sp %in% sp_top10)
-
-#pct agb of the top 15 dendro species
-sum(hkk_census_sp$pct.agb[hkk_census_sp$sp %in% sp_top15])
-
-top10sp.map.census<-ggplot(hkk_census_top10_sp, aes(x=gx, y=gy, 
-                                                    #size=as.numeric(dbh/10),
-                                                    color=factor(sp, levels=sp_top10)))+
-  geom_point(size=1)+
-   scale_size(range = c(.1, 10), name="dbh")+
-  guides(color=guide_legend(title="Species"))+
-  scale_color_viridis_d()+
-  ggtitle("top 10 species in 2004")+
-  #scale_colour_gradient(low="orange", high="purple") +
-  theme_minimal()+
-  coord_fixed()
-
-top10sp.map.census
-
-trees.top10<-trees[trees$SPCODE.UPDATE %in% sp_top10,]
-
-top10sp.map.dendro<-ggplot(trees.top10, aes(x=X, y=Y, 
-                                                    #size=as.numeric(dbh/10), 
-                                                    color=factor(SPCODE.UPDATE, levels=sp_top10)))+
-  geom_point(size=1)+
-   scale_size(range = c(.1, 10), name="dbh")+
-  guides(color=guide_legend(title="Species"))+
-  scale_color_viridis_d()+
-  ggtitle("top 10 species in 2004")+
-  #scale_colour_gradient(low="orange", high="purple") +
-  theme_minimal()+
-  coord_fixed()
-
-top10sp.map.dendro
-
-```
-
-## Growth of most abundant species
-
-```{r echo=F}
-
-growth_sp5<-ggplot(dendro.top5.dat, aes(x=factor(sp, levels=sp_top5), y=diainc,
-                                              fill=factor(sp, levels=sp_top5),
-                                              color=factor(sp, levels=sp_top5)))+ 
-  geom_violin(position = "dodge")+ 
-    geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
-ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #geom_hline(yintercept = mean(dendro.top5.dat$diainc, na.rm=T))+
-  geom_hline(yintercept = median(dendro.top5.dat$diainc, na.rm=T), lty=2)+
-  geom_text(aes(label=paste0("median=\n", round(median(dendro.top5.dat$diainc, na.rm=T),2), " cm"), 
-                x=0.6, y=median(dendro.top5.dat$diainc, na.rm=T)+0.15), col="black", size=3)+
-  geom_hline(yintercept = 0, lty=3)+
-      theme(axis.text.x = element_blank())+
-  ggtitle("HKK dendroband data from 2008 to 2023")+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))+
-  xlab("")
-
-growth_sp5
-
-growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(sp, levels=sp_top10), y=diainc,
-                                              fill=factor(sp, levels=sp_top10),
-                                              color=factor(sp, levels=sp_top10)))+ 
-  geom_violin(position = "dodge")+ 
-    geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
-ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
-  geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
-  geom_text(aes(label=paste0("median=\n", round(median(dendro.top10.dat$diainc, na.rm=T),2), " cm"), 
-                x=0.6, y=median(dendro.top10.dat$diainc, na.rm=T)+0.15), col="black", size=3)+
-  geom_hline(yintercept = 0, lty=3)+
-      theme(axis.text.x = element_blank())+
-  ggtitle("HKK dendroband data from 2008 to 2023")+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))+
-  xlab("")
-
-growth_sp10
-
-growth_yr_sp<-ggplot(dendro.top5, aes(x=year, y=mean.inc, 
-                                      color=sp))+ 
-  geom_point()+
-  geom_errorbar(aes(x=year, ymin=lower, ymax=upper))+
-  ylab("Residuals of annualised diameter increment")+theme_bw()+
-  theme(axis.text.x = element_text(angle=45))+
-  facet_grid(sp~.)+
-  guides(color="none")+
-  geom_hline(yintercept = 0)+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  geom_vline(xintercept=c(2010, 2016), lty=2)
-growth_yr_sp
-
-#png("plots/growth_by_sp.png", height=8, width=4, units="in", res=300)
-#growth_yr_sp
-#dev.off()
-
-```
-
-## Size classes
-
-```{r echo=F}
-
-size_top10<-ggplot(dendro.top10.dat, aes(x=dbh))+
-  geom_density()+
-  geom_vline(aes(xintercept=35))+
-  geom_vline(aes(xintercept=75))
-
-size_top10
-
-#trying three categories
-# <50, 50-100, >100
-
-dendro.top10.dat<-dendro.top10.dat %>%
-  dplyr::mutate(size=ifelse(dbh<35, "< 35 cm dbh",
-                       ifelse(dbh<70, "35-75 cm dbh", ">75 cm dbh")))
-
-growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(sp, levels=sp_top10), y=diainc,
-                                              #color=factor(sp),
-                                          #group=factor(size, levels=c("small", "medium", "large")),
-                                              fill=factor(sp, levels=sp_top10)))+ 
-  #geom_violin(position = "dodge")+ 
-  geom_boxplot(position = "dodge")+ 
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
-  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-  facet_grid(factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh"))~.)+
-      theme(axis.text.y = element_blank())+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  ggtitle("HKK dendroband data from 2008 to 2023")+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))+coord_flip()
-
-growth_sp10
-
-growth_sp10<-ggplot(dendro.top10.dat, aes(x=factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh")), y=diainc,
-                                              #color=factor(sp),
-                                          #group=factor(size, levels=c("small", "medium", "large")),
-                                              fill=factor(sp, levels=sp_top10)))+
-
-  geom_violin(position = "dodge")+ 
-  geom_boxplot(position = "dodge", col="black", fill="white", width=0.1)+ 
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
-  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-  #facet_grid(factor(size, levels=c("< 35 cm dbh", "35-75 cm dbh", ">75 cm dbh"))~.)+
-      theme(axis.text.x = element_text(angle=45))+
-  xlab("")+
-  facet_grid(.~factor(sp, levels=sp_top10))+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  ggtitle("HKK dendroband data from 2008 to 2023")+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))#+coord_flip()
-
-growth_sp10
-
-```
-
-## Canopy illumination
-
-```{r echo=F}
-
-cii_top10<-ggplot(dendro.top10.dat, aes(x=cii))+
-  geom_histogram()
-
-cii_top10
-
-#trying three categories
-# <50, 50-100, >100
-
-#dendro.top10.dat<-dendro.top10.dat %>%
-#  dplyr::mutate(size=ifelse(dbh<50, "small",
-#                       ifelse(dbh<100, "medium", "large")))
-
-growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
-                        aes(x=factor(sp, levels=sp_top10), y=diainc,
-                                              #color=factor(sp),
-                                          #group=factor(size, levels=c("small", "medium", "large")),
-                                              fill=factor(sp, levels=sp_top10)))+ 
-  #geom_violin(position = "dodge")+ 
-  geom_boxplot(position = "dodge")+ 
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
-  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-  facet_grid(cii~.)+
-      theme(axis.text.y = element_blank())+
-  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))+coord_flip()
-
-growth_sp10_cii
-
-growth_sp10_cii<-ggplot(dendro.top10.dat[dendro.top10.dat$cii %in% (1:5),], 
-                        aes(
-                          #x=factor(sp, levels=sp_top10), 
-                          x=factor(cii), 
-                            y=diainc,
-                                              #color=factor(sp),
-                                          #group=factor(size, levels=c("small", "medium", "large")),
-                                              fill=factor(sp, levels=sp_top10)))+ 
-  #geom_violin(position = "dodge")+ 
-  geom_boxplot(position = "dodge")+ 
-  ylab("Annualised diameter increment (cm)")+theme_bw()+
-  #geom_hline(yintercept = mean(dendro.top10.dat$diainc, na.rm=T))+
-  #geom_hline(yintercept = median(dendro.top10.dat$diainc, na.rm=T), lty=2)+
-  geom_hline(yintercept = 0, lty=3)+
-  facet_grid(.~factor(sp, levels = sp_top10))+
-      theme(axis.text.y = element_blank())+
-  ggtitle("HKK dendroband data from 2008 to 2023 - canopy illumination")+
-  scale_fill_viridis_d()+
-  scale_color_viridis_d()+
-  #coord_flip()+
-  guides(fill=guide_legend(title="species"),
-         color=guide_legend(title="species"))
-growth_sp10_cii
-
-```
-
-## Top 15 species split by leaf habit
-
-```{r evg/dec, echo=F}
-
-sp_top15<-c("SACCLI", "HOPEOD", "POLYVI", "TETRNU", "VATIHA", 
-            "ALPHVE", "DIPTAL", "DIOSWI", "ARYTLI", "BACCRA",
-            "GARCSP", "CROTRO", "GARUPI", "NEOLOB", "PHOEPA")
-
-#these 15 species had the most dendrobands and together made up 51.22% of the agb in 2009
-
-#below values eyeballed from Williams et al. 2008 Fig 1.
-sp_top15_phenoscore<-c(2.5, 0.8, 1.3, 4, 0.7, 
-                       1, 1.25, 1.27, 0.2, 0.3,
-                       0.4, 0.65, 3.95, 0.15, 0.25)
-
-#no info on Alphonsea ventricosa in the paper. 
-#Apparently evergreen, so used 1 as value - middle of evergreen range
-
-pheno<-as.data.frame(cbind(sp_top15, sp_top15_phenoscore))
-colnames(pheno)<-c("sp", "phenoscore")
-
-dendro_top15<-dendro.data.noouts[dendro.data.noouts$sp %in% sp_top15,]
-dendro_top15<-merge(dendro_top15, pheno, by="sp")
-
-#categorising as evg and deciduous
-dendro_top15$cat<-ifelse(dendro_top15$phenoscore<=2, "evg", "dec")
-
-growth_yr_cat_detrend<-ggplot(dendro_top15, aes(x=factor(cat), y=inc.resid,
-                                              #group=c(year, sp),
-                                              #group=sp,
-                                              fill=factor(cat)))+ 
-  geom_violin(position = "dodge")+ 
-  ylab("Residuals of annualised diameter increment")+theme_bw()+
-  facet_grid(.~year)+
-  geom_hline(yintercept = 0)+
-  scale_fill_manual(values=c("sienna4", "darkgreen"))+
-  theme(axis.text.x = element_blank())
-
-growth_yr_cat_detrend
-
-```
-
-- identified and removed growth release
-- hand pick high quality time series (10-20 per species maybe)
-- try VH's model - temporal autocorr
-- mean and median growth 
-- by month plots - time difference
-- different rates

### 2024-07-02 doc/manuscript.Rmd
manuscript edits

@@
 
-***Sites and data*** -
+***Sites and data***
+
+*Huai Kha Khaeng plot*
+- location
+- annual precipitation
+- ecosystem type/association
+- long-term trends from @vlam et al etc
+
+*Resistance*
+-Dendrometer band measurements - census protocol
+-Calculated DBH at each instance from first DBH and window size using Condit's functions.
+-Removed measures with large decimal errors, large outliers (> 3 SD). 
+- For all individuals with three points of measurements for each drought, calculated raw increment as difference between relevant censuses to get pre-drought increment and drought-year increment.
+- Resistance - difference between pre-drought and drought-year increment
+
+*Tree size*
+- calculated DBH for at the start of the pre-drought year from dendrometer band window
+
+*Deciduousness*
+-data from @williams_deciduousness_2008. 
+-Leaf phenology score - mean proportion of crown loss. 
+-Continuous scale from 1 to 4, where 4 is maximum crown loss or deciduousness. 
+-One value per species. 
+- 7 species out of the top 51 species in the dendroband census have no values. (currently excluded from analysis but can fill these with best estimates and add them back in)
+
+*Exposure*
+- crown illumination index for each individual, collected at each dendrometer band census
+- categorical variable with values 1 to 5 where 1 is least exposed and 5 is most exposed
+
+*Topographic Wetness Index*
+- calculated from topography
+- estimate of water availability at a location
 
 ***Models***
+
+For individual stems, we modelled drought resistance as:
+
+$$
+resistance \sim size + exposure + deciduousness + water + size*exposure + deciduousness*exposure + size*deciduousness + (1|drought.year)
+$$
 

### 2024-07-05 doc/manuscript.Rmd
add DAG.png

@@
 ## Figures
+fig_DAG=1
 
@@
 - Large trees tend to undergo greater growth declines during drought compared to smaller trees [e.g., @bennett_larger_2015; @mcgregor_tree_2021; @anderson-teixeira_joint_2022]. 
-- It remains poorly understood the extent to which this is shaped by tree size itself, crown exposure, water access, and traits that tend to covary with size (e.g., decidiousness) -- all of which interact to shape drought resistance
+- It remains poorly understood the extent to which this is shaped by tree size itself, crown exposure, water access, and traits that tend to covary with size (e.g., decidiousness) -- all of which interact to shape drought resistance (Fig.
+`r fig_DAG`).
 

### 2024-07-07 scripts/explore_analysis.Rmd
visualising top 4 species after workshop meeting

@@
 #reading the GitHub PAT (stored privately and locally)
-pat<-as.character(read.table("../data/HKK-dendro/git_pat.txt")[1,1])
+#pat<-as.character(read.table("../data/HKK-dendro/git_pat.txt")[1,1])
 
@@
 
-load("../data/HKK-dendro/DendroByBand.RData")
+load("growth-precip-thailand/data/HKK-dendro/DendroByBand.RData")
 
@@
 
-#this can be refined if needed
-#wet_census<-c(1, 3, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24, 26, 28)
-
 #getting climate data for all
@@
 
-chirps_hkk<-read.csv("../data/climate/CHIRPS_from_GEE.csv")
+#chirps_hkk<-read.csv("../data/climate/CHIRPS_from_GEE.csv")
+chirps_hkk<-read.csv("growth-precip-thailand/data/climate/CHIRPS_from_GEE.csv")
 colnames(chirps_hkk)[1]<-"time"
@@
 
-spei_hkk<-read.csv("../data/climate/SPEI_HKK_from_GEE.csv")
+spei_hkk<-read.csv("growth-precip-thailand/data/climate/SPEI_HKK_from_GEE.csv")
 colnames(spei_hkk)[1]<-"time"
@@
 pre_yr_census<-ggplot()+
-  #geom_area(data=chirps_hkk, aes(x=time, y=precipitation), fill="blue", alpha=0.5)+
   geom_line(data=chirps_hkk, aes(x=time, y=precipitation), col="blue", alpha=0.5)+
@@
 
-png("../doc/display/census_precip.png", width=6, height=3, res=300, units="in") 
+#png("../doc/display/census_precip.png", width=6, height=3, res=300, units="in") 
+#png("../doc/display/census_precip.png", width=6, height=3, res=300, units="in") 
 pre_yr_census
-dev.off()
+#dev.off()
 
@@
   mutate(year=year(Date2)) %>%
-  #filter(year!=2209)%>% #quick fix to remove bad dates - there are 30 entries
-  #filter(start_date <= year & year <= end_date) %>%
-  #filter(Cno %in% wet_census)%>%
   filter(is.na(DBH1_flag), DBH1_error=="none",
@@
 ```
-```{r calculating increments for specific years}
+```{r long-term growth rates}
+
+# remotes:::download("../data/HKK-dendro/stem1.RData",
+#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem1.rdata",
+#                    auth_token = pat)
+# remotes:::download("../data/HKK-dendro/stem2.RData",
+#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem2.rdata",
+#                    auth_token = pat)
+# remotes:::download("../data/HKK-dendro/stem3.RData",
+#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem3.rdata",
+#                    auth_token = pat)
+# remotes:::download("../data/HKK-dendro/stem4.RData",
+#                    "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem4.rdata",
+#                    auth_token = pat)
+
+load("growth-precip-thailand/data/HKK-dendro/stem1.RData")
+load("growth-precip-thailand/data/HKK-dendro/stem2.RData")
+load("growth-precip-thailand/data/HKK-dendro/stem3.RData")
+load("growth-precip-thailand/data/HKK-dendro/stem4.RData")
+
+hkk_census<-bind_rows(hkk.stem1, hkk.stem2, hkk.stem3, hkk.stem4)
+
+hkk_census <- hkk_census %>%
+  filter(status=="A")%>%
+  filter(tag %in% dendro.data$Tag)%>%
+  group_by(tag, StemTag)%>%
+  dplyr::mutate(dbhmin1 = dplyr::lag(dbh, n=1, default=NA, order_by=StemTag),
+                inc_annual = (dbh-dbhmin1)/5)%>%
+  filter(!is.na(dbhmin1))%>%
+  #average increment in cm
+  dplyr::summarise(avg_inc = mean(inc_annual/10, na.rm=T),
+                   sd_inc = sd(inc_annual/10, na.rm=T))
+
+#find the StemTags of trees that are in the dendro data
+#match using the DBH04 column
+
+#identify multistem trees from the census
+multistem<-hkk_census %>%
+  group_by(tag)%>%
+  dplyr::summarise(n.stem=n())%>%
+  rename(Tag=tag)
+
+multistem<-merge(multistem, trees, by="Tag", all.x=T)
+
+multistem<-multistem %>%
+rename(tag=Tag)%>%
+dplyr::mutate(dbh= (as.numeric(DBH.04)*10))
+
+hkk.stem3.dendro<-hkk.stem3 %>%
+  filter(tag %in% multistem$tag)
+
+#add the StemTag values from hkk.stem3.dendro to multistem 
+#match based on matching dbh and tag
+
+multistem<- multistem %>%
+dplyr::inner_join(select(hkk.stem3.dendro, c("tag", "StemTag", "dbh")), 
+by=c("tag", "dbh"))
+
+sum(multistem$tag != multistem$StemTag)
+#there are no stemtags with values different from Tag
+#all stems with dendrobands have them on the main stem?
+
+#hkk_census<-hkk_census %>%
+#  filter(StemTag %in% multistem$StemTag)
+
+#since dendrobands are installed on trees with tag==StemTag
+hkk_census<-hkk_census %>%
+filter(tag==StemTag)
+
+#there are 2271 trees in this dataset.
+
+inc_census<-ggplot(data = hkk_census, aes(x=avg_inc)) +
+  geom_density(alpha=0.5) +
+  xlab("mean annual increment (cm)")+
+  theme_bw()
+
+png("growth-precip-thailand/doc/display/census_increment.png", width=4, height=4, units="in", res=300)
+inc_census
+dev.off()
+
+#plot average increment for the top 4 species
+
+hkk_census2<-hkk_census %>%
+dplyr::rename(Tag=tag)
+
+hkk_census2<-merge(hkk_census2, trees, by="Tag")
+hkk_census2<-hkk_census2 %>%
+dplyr::rename(sp=SPCODE.UPDATE)
+
+png("growth-precip-thailand/doc/display/census_increment_species.png", width=4, height=4, units="in", res=300)
+ggplot(hkk_census2 %>% filter(sp %in% c("SACCLI", "HOPEOD", "TETRNU", "POLYVI")), 
+aes(x= factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), y = avg_inc, fill=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")))) +
+  geom_violin(position = "dodge")+ 
+    geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
+  scale_fill_viridis_d()+
+  labs(title = 'tree growth in HKK from census', 
+  y="Annualised diameter increment (cm)", x="species") +
+  theme_bw()+
+  #geom_vline(xintercept=0, lty=2)+
+    theme(
+      legend.position="none",
+      panel.spacing = unit(0.1, "lines"),
+      strip.text.x = element_text(size = 8)
+    )
+dev.off()
+```
+
+```{r calculating increments for all years}
+
+hkk_census<-hkk_census %>%
+rename(Tag=tag)
+
+#add long-term increment to dendro data
+dendro.data.noouts<-merge(dendro.data.noouts, hkk_census, by="Tag", all.x=T)
+dendro.data<-merge(dendro.data, hkk_census, by="Tag", all.x=T)
+
+#making increments for each year
+
+alt_census<-c(1, 3, 5, 7, 9, 11, 13, 15, 17, 19, 21, 23, 25, 27, 29)
+
+dendro_inc<- dendro.data %>%
+  filter(Cno %in% alt_census)%>%
+  group_by(Tag) %>% #grouping by treeID to calculate increment
+  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
+         cii_min1=dplyr::lag(Cii, n=1, default=NA, order_by=Tag),
+         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
+         inc=calcDBH-calcDBH_min1,
+         inc_annual=(inc/as.numeric(tdif))*365,
+         inc.time=Cno-lag(Cno),
+         #removing increment value if the previous measure was before the last year
+         inc_annual=ifelse(inc.time==2, inc_annual, NA),
+         treeID=paste0("HKK.", Tag))%>%
+  filter(!is.na(inc))%>%
+  filter(!is.na(inc_annual))%>%
+  ungroup()%>%
+  dplyr::mutate(
+    diainc.scaled = scale(inc_annual),
+    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))%>%
+  filter(out.inc==0)
+```
+```{r species wise increments, echo=F}
+
+#filter increments for two species
+
+saccli<-dendro_inc %>%
+  filter(sp == "SACCLI")
+
+hopeod <- dendro_inc %>%
+  filter(sp == "HOPEOD")
+
+tetrnu<-dendro_inc %>%
+  filter(sp == "TETRNU")
+
+polyvi <- dendro_inc %>%
+  filter(sp == "POLYVI")
+
+#spaghetti plot saccli and hopeod increments over time
+
+saccli_plot<-ggplot() +
+  geom_line(data = saccli, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag)) +
+  geom_line(data=saccli %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=mean_inc), col="red")+
+  xlab("year") + ylab("diameter increment (cm)")+
+  ggtitle("growth increments for SACCLI (deciduous)")+
+  theme_bw()
+
+hopeod_plot<-ggplot() +
+  geom_line(data = hopeod, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag)) +
+  geom_line(data=hopeod %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=mean_inc), col="red")+
+  xlab("year") + ylab("diameter increment (cm)")+
+  ggtitle("growth increments for HOPEOD (evergreen)")+
+  theme_bw()
+
+tetrnu_plot<-ggplot() +
+  geom_line(data = tetrnu, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag)) +
+  geom_line(data=tetrnu %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=mean_inc), col="red")+
+  xlab("year") + ylab("diameter increment (cm)")+
+  ggtitle("growth increments for TETRNU (deciduous)")+
+  theme_bw()
+
+polyvi_plot<-ggplot() +
+  geom_line(data = polyvi, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag)) +
+  geom_line(data=polyvi %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=mean_inc), col="red")+
+  xlab("year") + ylab("diameter increment (cm)")+
+  ggtitle("growth increments for POLYVI (evergreen)")+
+  theme_bw()
+
+png("growth-precip-thailand/doc/display/ind_increments.png", width=8, height=8, units="in", res=300)
+grid.arrange(hopeod_plot, tetrnu_plot, polyvi_plot, saccli_plot, nrow=2)
+dev.off()
+
+top4<-dendro_inc %>%
+filter(sp %in% c("SACCLI", "HOPEOD", "TETRNU", "POLYVI"))
+
+#sample size per species per year
+png("growth-precip-thailand/doc/display/top4_sample_size.png", width=8, height=4, units="in", res=300)
+top4 %>% 
+  group_by(sp, Cno) %>% 
+  dplyr::summarise(n=n())%>%
+  ggplot(aes(x=Cno, y=n, fill=factor(sp,levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
+  geom_bar(stat="identity")+
+  scale_fill_viridis_d()+
+  facet_grid(.~factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")))+
+  xlab("census number")+
+  ylab("sample size")+
+  guides(fill=guide_legend(title="species"))+
+  theme_bw()
+dev.off()
+
+png("growth-precip-thailand/doc/display/top4_increments.png", width=8, height=4, units="in", res=300)
+ggplot(top4, aes(x = inc_annual, y = factor(((Cno-3)/2)+2009), 
+                       fill = stat(x))) +
+  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
+  scale_fill_viridis(name = "growth", option = "H", direction=-1) +
+  labs(title = 'tree growth in HKK (cm)', y="year") +
+  #theme_ipsum() +
+  facet_grid(.~factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")))+
+  theme_bw()+
+  xlim(-1.5, 1.5)+
+  geom_vline(xintercept=0, lty=2)+
+    theme(
+      legend.position="none",
+      panel.spacing = unit(0.1, "lines"),
+      strip.text.x = element_text(size = 8)
+    )
+dev.off()
+
+growth_top4<-ggplot(top4, aes(x=factor(((Cno-3)/2)+2009), y=inc_annual,
+                                              fill=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")),
+                                              color=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))
+                                              ))+ 
+  geom_violin(position = "dodge")+ 
+    geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
+ylab("Annualised diameter increment (cm)")+theme_bw()+
+facet_grid(.~factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")))+
+  geom_hline(yintercept = 0, lty=3)+
+      theme(axis.text.x = element_text(angle=90))+
+  ggtitle("HKK dendroband data from 2008 to 2023")+
+  scale_fill_viridis_d()+
+  scale_color_viridis_d()+
+  guides(fill=guide_legend(title="species"),
+         color=guide_legend(title="species"))+
+         xlab("year")
+
+png("growth-precip-thailand/doc/display/top4_increments_violin.png", width=8, height=4, units="in", res=300)
+growth_top4
+dev.off()
+
+growth_top4
+
+growth_top4_sum<-ggplot(top4, aes(x=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), y=inc_annual,
+                                              fill=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")),
+                                              color=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))
+                                              ))+ 
+  geom_violin(position = "dodge")+ 
+    geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
+ylab("Annualised diameter increment (cm)")+
+xlab("species")+theme_bw()+
+  geom_hline(yintercept = 0, lty=3)+
+  ggtitle("tree growth in HKK from dendrobands")+
+  scale_fill_viridis_d()+
+  scale_color_viridis_d()+
+  guides(fill="none",
+         color="none")
+
+png("growth-precip-thailand/doc/display/top4_increments_sum.png", width=4, height=4, units="in", res=300)
+growth_top4_sum
+dev.off()
+
+growth_top4
+```
+
+```{r plot increments}
 
@@
          #removing increment value if the previous measure was before the last year
-         #inc=ifelse(inc.time==1, inc, NA),
          inc_annual=ifelse(inc.time==2, inc_annual, NA),
@@
 library(gridExtra)
-png("../doc/display/increments_raw.png", width=8, height=4, units="in", res=300)
-grid.arrange(dendro_2010_plot, dendro_2015_plot, dendro_2019_plot, nrow=1)
+png("growth-precip-thailand/doc/display/increments_raw.png", width=8, height=4, units="in", res=300)
+#grid.arrange(dendro_2010_plot, dendro_2015_plot, dendro_2019_plot, nrow=1)
+grid.arrange(dendro_2010_plot, dendro_2015_plot, nrow=1)
 dev.off()
@@
 
-```{r long-term growth rates}
+```{r top4 species drought resistance, echo=F}
 
-library(ggplot2)
+resistance_2010_top4 <- top4 %>%
+  filter(Cno %in% c(3, 5))%>%
+  group_by(Tag) %>% #grouping by treeID to calculate increment
+  dplyr::mutate(inc_min1=dplyr::lag(inc_annual, n=1, default=NA, order_by=Tag),
+         resistance.div = inc_annual/avg_inc,
+         resistance.dif = inc_annual-inc_min1,
+         resistance.prop = (inc_annual-inc_min1)/avg_inc)%>%
+  filter(!is.na(inc_min1), !is.infinite(resistance.div))
 
-pat<-as.character(read.table("../data/HKK-dendro/git_pat.txt")[1,1])
+resistance_2015_top4 <- top4 %>%
+  filter(Cno %in% c(11, 13, 15))%>%
+  group_by(Tag) %>% #grouping by treeID to calculate increment
+  mutate(inc_min1=dplyr::lag(inc_annual, n=1, default=NA, order_by=Tag),
+         resistance.div = inc_annual/avg_inc,
+         resistance.dif = inc_annual-inc_min1,
+         resistance.prop = (inc_annual-inc_min1)/avg_inc)%>%
+  filter(!is.na(inc_min1), !is.infinite(resistance.div))
 
-remotes:::download("../data/HKK-dendro/stem1.RData",
-                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem1.rdata",
-                   auth_token = pat)
-remotes:::download("../data/HKK-dendro/stem2.RData",
-                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem2.rdata",
-                   auth_token = pat)
-remotes:::download("../data/HKK-dendro/stem3.RData",
-                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem3.rdata",
-                   auth_token = pat)
-remotes:::download("../data/HKK-dendro/stem4.RData",
-                   "https://raw.githubusercontent.com/forestgeo/HKK-tree-growth-data/main/data/Census/raw_data/hkk.stem4.rdata",
-                   auth_token = pat)
+#calculate quantiles for resistance.div for both years
 
-load("../data/HKK-dendro/stem1.RData")
-load("../data/HKK-dendro/stem2.RData")
-load("../data/HKK-dendro/stem3.RData")
-load("../data/HKK-dendro/stem4.RData")
+quantiles_2010_top4<-quantile(resistance_2010_top4$resistance.div, 
+                         probs=c(0.25, 0.5, 0.75, 1), na.rm=T)
+quantiles_2015_top4<-quantile(resistance_2015_top4$resistance.div, 
+                         probs=c(0.25, 0.5, 0.75, 1), na.rm=T)
 
-hkk_census<-bind_rows(hkk.stem1, hkk.stem2, hkk.stem3, hkk.stem4)
+#split the data into categories based on quantiles
 
-hkk_census <- hkk_census %>%
-  filter(status=="A")%>%
-  filter(tag %in% dendro.data.noouts$Tag)%>%
-  group_by(StemTag)%>%
-  dplyr::mutate(dbhmin1 = dplyr::lag(dbh, n=1, default=NA, order_by=StemTag),
-                inc_annual = (dbh-dbhmin1)/5)%>%
-  filter(!is.na(dbhmin1))%>%
-  dplyr::summarise(avg_inc = mean(inc_annual, na.rm=T),
-                   sd_inc = sd(inc_annual, na.rm=T))
+resistance_2010_top4$quantile<-ifelse(resistance_2010_top4$resistance.div<quantiles_2010_top4[1], 1, ifelse(resistance_2010_top4$resistance.div<quantiles_2010_top4[2], 2, ifelse(resistance_2010_top4$resistance.div<quantiles_2010_top4[3], 3, 4)))
+resistance_2015_top4$quantile<-ifelse(resistance_2015_top4$resistance.div<quantiles_2015_top4[1], 1, ifelse(resistance_2015_top4$resistance.div<quantiles_2015_top4[2], 2, ifelse(resistance_2015_top4$resistance.div<quantiles_2015_top4[3], 3, 4)))
 
-inc_census<-ggplot(data = hkk_census, aes(x=avg_inc)) +
-  geom_density(alpha=0.5) +
-  xlab("mean annual increment (cm)")+
-  theme_bw()
+#map of resistance.div quantiles for top 4 species
+resistance_all<-rbind(resistance_2010_top4, resistance_2015_top4)
+resistance_all$yr<-c(rep(2010, nrow(resistance_2010_top4)), rep(2015, nrow(resistance_2015_top4)))
 
-png("../doc/display/census_increment.png", width=4, height=4, units="in", res=300)
-inc_census
+resistance_all<-merge(resistance_all, select(trees, -"SPCODE.UPDATE"), by="Tag")
+
+resistance_map<-ggplot(resistance_all, aes(x=X, y=Y, size=as.numeric(DBH.04), col=quantile))+
+  geom_point(alpha=0.5)+
+   scale_size(range = c(.1, 10), name="DBH (cm)")+
+  guides(color=guide_legend(title="resistance"))+
+  scale_colour_gradient(low="red", high="blue") +
+  facet_grid(sp~yr)+
+  theme_minimal()+
+  coord_fixed()
+
+png("growth-precip-thailand/doc/display/resistance_map_top4.png", width=8, height=8, units="in", res=300)
+resistance_map
 dev.off()
@@
 
-dec_williams<-read.csv("../data/HKK-dendro/species.csv")
+dec_williams<-read.csv("growth-precip-thailand/data/HKK-dendro/species.csv")
 
@@
   mutate(inc_min1=dplyr::lag(inc_annual, n=1, default=NA, order_by=Tag),
-         resistance.div = inc_annual/inc_min1,
+         resistance.div = inc_annual/avg_inc,
          resistance.dif = inc_annual-inc_min1,
-         resistance.prop = (inc_annual-inc_min1)/inc_min1)%>%
+         resistance.prop = (inc_annual-inc_min1)/avg_inc)%>%
   filter(!is.na(inc_min1), !is.infinite(resistance.div))
@@
   geom_density(alpha=0.5) +
+  geom_vline(xintercept = 0, lty=2)+
   theme_bw()
@@
 
+library(gridExtra)
+
+#png("growth-precip-thailand/doc/display/resistance_div.png", width=4, height=8, units="in", res=300)
+#grid.arrange(resistance_2010_div, resistance_2015_div, resistance_2019_div, nrow=3)
+#dev.off()
+
+png("growth-precip-thailand/doc/display/resistance_div.png", width=8, height=4, units="in", res=300)
+grid.arrange(resistance_2010_div, resistance_2015_div, nrow=1)
+dev.off()
+
 ```
@@
 
-resistance_all<-rbind(resistance_2010, resistance_2015, resistance_2019)
-resistance_all$yr<-c(rep(2010, nrow(resistance_2010)), rep(2015, nrow(resistance_2015)), rep(2019, nrow(resistance_2019)))
+quantiles_2010_div<-quantile(resistance_2010$resistance.div, 
+                         probs=c(0.25, 0.5, 0.75, 1), na.rm=T)
+
+quantiles_2015_div<-quantile(resistance_2015$resistance.div, 
+                         probs=c(0.25, 0.5, 0.75, 1), na.rm=T)
+
+quantiles_2019_div<-quantile(resistance_2019$resistance.div, 
+                         probs=c(0.25, 0.5, 0.75, 1), na.rm=T)
+
+resistance_2010$quantile.div<-ifelse(resistance_2010$resistance.div<quantiles_2010_div[1], 1, ifelse(resistance_2010$resistance.div<quantiles_2010_div[2], 2,
+            ifelse(resistance_2010$resistance.div<quantiles_2010_div[3], 3, 4))
+
+                                 )
+
+resistance_2015$quantile.div<-ifelse(resistance_2015$resistance.div<quantiles_2015_div[1], 1, ifelse(resistance_2015$resistance.div<quantiles_2015_div[2], 2,
+            ifelse(resistance_2015$resistance.div<quantiles_2015_div[3], 3, 4))
+
+                                 )
+
+resistance_2019$quantile.div<-ifelse(resistance_2019$resistance.div<quantiles_2019_div[1], 1, ifelse(resistance_2019$resistance.div<quantiles_2019_div[2], 2,
+            ifelse(resistance_2019$resistance.div<quantiles_2019_div[3], 3, 4))
+
+                                 )
+
+resistance_all<-rbind(resistance_2010, resistance_2015)
+resistance_all$yr<-c(rep(2010, nrow(resistance_2010)), rep(2015, nrow(resistance_2015)))
 
@@
 
-resistance_map<-ggplot(resistance_all, aes(x=X, y=Y, size=as.numeric(DBH.04), col=quantile))+
+resistance_map<-ggplot(resistance_all, aes(x=X, y=Y, size=as.numeric(DBH.04), col=quantile.div))+
   geom_point(alpha=0.5)+
@@
 
-png("../doc/display/resistance_map.png", width=8, height=10, units="in", res=300)
+png("growth-precip-thailand/doc/display/resistance_map.png", width=8, height=6, units="in", res=300)
 resistance_map
@@
 
-model_2010<-glm(resistance.dif ~ calcDBH_min1 + cii_min1 + williams_dec+
-                  calcDBH_min1*cii_min1 + 
+resistance_all$ciiF<-ordered(resistance_all$cii_min1)
+
+model_2010<-glm(resistance.div ~ calcDBH_min1 + ciiF + williams_dec+
+                  calcDBH_min1*ciiF + 
                   calcDBH_min1*williams_dec+
-                  cii_min1*williams_dec,
-                data=resistance_2010)
+                  ciiF*williams_dec,
+                data=resistance_all[resistance_all$yr==2010,])
 
@@
 
-model_2015<-glm(resistance.dif ~ calcDBH_min1 + cii_min1 + williams_dec+
-                  calcDBH_min1*cii_min1 + 
+model_2015<-glm(resistance.div ~ calcDBH_min1 + ciiF + williams_dec+
+                  calcDBH_min1*ciiF + 
                   calcDBH_min1*williams_dec+
-                  cii_min1*williams_dec,
-                data=resistance_2015)
+                  ciiF*williams_dec,
+                data=resistance_all[resistance_all$yr==2015,])
 
@@
 
-model_2019<-glm(resistance.dif ~ calcDBH_min1 + cii_min1 + williams_dec+
-                  calcDBH_min1*cii_min1 + 
-                  calcDBH_min1*williams_dec+
-                  cii_min1*williams_dec,
-                data=resistance_2019)
-
 summary(model_2019)
 
-resistance_all<-rbind(resistance_2010, resistance_2015, resistance_2019)
-resistance_all$yr<-c(rep(2010, nrow(resistance_2010)), rep(2015, nrow(resistance_2015)), rep(2019, nrow(resistance_2019)))
-
 resistance_model<-resistance_all %>%
-  select(resistance.dif, calcDBH_min1, cii_min1, williams_dec, yr, quantile)
+  select(resistance.div, calcDBH_min1, ciiF, williams_dec, yr, quantile)
 
@@
 library(nlme)
-model_all<-lme(resistance.dif ~ calcDBH_min1 + cii_min1 + williams_dec+
-                  calcDBH_min1*cii_min1 + 
+model_all<-lm(resistance.div ~ calcDBH_min1 + ciiF + williams_dec+
+                  calcDBH_min1*ciiF + 
                   calcDBH_min1*williams_dec+
-                  cii_min1*williams_dec,
-                random=~1|yr,
-                data=resistance_model)
+                  ciiF*williams_dec + yr,
+                data=resistance_model,
+                )
 
@@
 
-plot_model(model_all, type="pred", terms=c("calcDBH_min1","williams_dec"), show.data = T)
+plot_model(model_all, type="pred", terms=c("ciiF","williams_dec"), show.data = T)
 plot_model(model_all, type="pred", terms=c("cii_min1"), show.data = T)
@@
 
-```{r echo=F}
+```{r data distribution, echo=F}
+
+#growth data all years ridge plot
+png("growth-precip-thailand/doc/display/growth_allyrs2.png", width=4, height=4, units="in", res=300)
+ggplot(dendro_inc, aes(x = inc_annual, y = factor(((Cno-3)/2)+2009), 
+                       fill = stat(x))) +
+  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
+  scale_fill_viridis(name = "growth", option = "H", direction=-1) +
+  labs(title = 'tree growth in HKK (cm)', y="year") +
+  #theme_ipsum() +
+  theme_bw()+
+  xlim(-1.5, 1.5)+
+  geom_vline(xintercept=0, lty=2)+
+    theme(
+      legend.position="none",
+      panel.spacing = unit(0.1, "lines"),
+      strip.text.x = element_text(size = 8)
+    )
+dev.off()
 

### 2024-07-08 scripts/explore_analysis.Rmd
updated script for explore plots

@@
   filter(Cno %in% alt_census)%>%
+  arrange(Cno)%>%
   group_by(Tag) %>% #grouping by treeID to calculate increment
-  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
+  dplyr::mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
          cii_min1=dplyr::lag(Cii, n=1, default=NA, order_by=Tag),
-         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
+         tdif=difftime(Date2, dplyr::lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
          inc=calcDBH-calcDBH_min1,
          inc_annual=(inc/as.numeric(tdif))*365,
-         inc.time=Cno-lag(Cno),
+         inc.time=Cno-dplyr::lag(Cno, n=1, default=NA, order_by=Tag),
          #removing increment value if the previous measure was before the last year
@@
 saccli_plot<-ggplot() +
-  geom_line(data = saccli, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag)) +
+  geom_line(data = saccli, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag), alpha=0.5) +
   geom_line(data=saccli %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=mean_inc), col="red")+
+  geom_line(data=saccli %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=median_inc), col="blue")+
   xlab("year") + ylab("diameter increment (cm)")+
@@
 hopeod_plot<-ggplot() +
-  geom_line(data = hopeod, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag)) +
+  geom_line(data = hopeod, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag), alpha=0.5) +
   geom_line(data=hopeod %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=mean_inc), col="red")+
+  geom_line(data=hopeod %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=median_inc), col="blue")+
   xlab("year") + ylab("diameter increment (cm)")+
@@
 tetrnu_plot<-ggplot() +
-  geom_line(data = tetrnu, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag)) +
+  geom_line(data = tetrnu, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag), alpha=0.5) +
   geom_line(data=tetrnu %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=mean_inc), col="red")+
+  geom_line(data=tetrnu %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=median_inc), col="blue")+
   xlab("year") + ylab("diameter increment (cm)")+
@@
 polyvi_plot<-ggplot() +
-  geom_line(data = polyvi, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag)) +
+  geom_line(data = polyvi, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag), alpha=0.5) +
   geom_line(data=polyvi %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=mean_inc), col="red")+
+  geom_line(data=polyvi %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=median_inc), col="blue")+
   xlab("year") + ylab("diameter increment (cm)")+
@@
 
+library(gridExtra)
 png("growth-precip-thailand/doc/display/ind_increments.png", width=8, height=8, units="in", res=300)
@@
 filter(sp %in% c("SACCLI", "HOPEOD", "TETRNU", "POLYVI"))
+
+#spaghetti plot of all species together
+
+spagplot_top4<-ggplot() +
+  #species plots
+  geom_line(data=top4 %>% group_by(Cno, sp) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=mean_inc, 
+  group=factor(sp,levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), 
+  col=factor(sp,levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
+  geom_line(data=top4 %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=mean_inc), col="black", size=2)+
+  scale_color_viridis_d()+
+  guides(col=guide_legend(title="species"))+
+  xlab("year") + ylab("diameter increment (cm)")+
+  ggtitle("growth increments for top 4 species")+
+  theme_bw()
+
+png("growth-precip-thailand/doc/display/spaghetti_top4.png", width=4, height=4, units="in", res=300)
+spagplot_top4
+dev.off()
+
+spagplot_all<-ggplot() +
+  #species plots
+  geom_line(data=dendro_inc %>% group_by(Cno, sp) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=mean_inc, 
+  group=sp, col=sp))+
+  #mean of all trees
+  geom_line(data=dendro_inc %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=mean_inc), col="black", size=2)+
+  #mean of species
+  geom_line(data=dendro_inc %>% group_by(Cno, sp) %>% 
+  dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)) %>%
+  ungroup()%>%
+  group_by(Cno) %>% dplyr::summarise(mean_inc=mean(mean_inc, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=mean_inc), col="grey40", size=0.8)+
+  scale_color_viridis_d()+
+  guides(col="none", group="none")+
+  xlab("year") + ylab("diameter increment (cm)")+
+  ggtitle("growth increments for top 4 species")+
+  theme_bw()
+
+png("growth-precip-thailand/doc/display/spaghetti_all.png", width=4, height=4, units="in", res=300)
+spagplot_all
+dev.off()
 
@@
 growth_top4
+
 ```
@@
 
-png("growth-precip-thailand/doc/display/resistance_map_top4.png", width=8, height=8, units="in", res=300)
+png("growth-precip-thailand/doc/display/resistance_map_top4.png", width=10, height=8, units="in", res=300)
 resistance_map

### 2024-07-08 scripts/explore_analysis.Rmd
updated script

@@
 
+census_top4<-bind_rows(hkk.stem1, hkk.stem2, hkk.stem3, hkk.stem4) %>%
+  filter(status=="A")%>%
+  filter(sp %in% c("SACCLI", "HOPEOD", "TETRNU", "POLYVI"))%>%
+  filter(tag==StemTag) %>%
+  group_by(sp, tag)%>%
+  dplyr::mutate(dbhmin1 = dplyr::lag(dbh, n=1, default=NA, order_by=tag),
+                inc_annual = (dbh-dbhmin1)/5)%>%
+  filter(!is.na(dbhmin1))%>%
+  #average increment in cm
+  dplyr::summarise(avg_inc = mean(inc_annual/10, na.rm=T),
+                   sd_inc = sd(inc_annual/10, na.rm=T))
+
 #find the StemTags of trees that are in the dendro data
@@
   theme_bw()+
+  geom_text(data=hkk_census2 %>% filter(sp %in% c("SACCLI", "HOPEOD", "TETRNU", "POLYVI")) %>% group_by(sp) %>% dplyr::summarise(n=n()),
+   aes(x=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), y=2.2, 
+  label=n))+
   #geom_vline(xintercept=0, lty=2)+
@@
 dev.off()
+
+png("growth-precip-thailand/doc/display/census_increment_species_alltrees.png", width=4, height=4, units="in", res=300)
+ggplot(census_top4, 
+aes(x= factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), y = avg_inc, fill=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")))) +
+  geom_violin(position = "dodge")+ 
+    geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
+  scale_fill_viridis_d()+
+  labs(title = 'tree growth in HKK (all stems)', 
+  y="Annualised diameter increment (cm)", x="species") +
+  theme_bw()+
+  geom_text(data=census_top4 %>% group_by(sp) %>% dplyr::summarise(n=n()),
+   aes(x=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), y=2.2, 
+  label=n))+
+  #geom_vline(xintercept=0, lty=2)+
+    theme(
+      legend.position="none",
+      panel.spacing = unit(0.1, "lines"),
+      strip.text.x = element_text(size = 8)
+    )
+dev.off()
+
 ```
@@
 
+#spaghetti plot for top 10 species
+
+sp_top10<-c("SACCLI", "HOPEOD", "TETRNU", "POLYVI", "VATIHA", 
+"DIPTAL", "ALPHVE", "GARCSP", "GARUPI", "NEOLOB")
+
+spagplot_top10<-ggplot()+
+  #species plots
+  geom_line(data=dendro_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno, sp) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=mean_inc, 
+  group=sp, col=sp))+
+  #mean of all trees
+  geom_line(data=dendro_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=mean_inc), col="black", size=2)+
+  #mean of species
+  geom_line(data=dendro_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno, sp) %>% 
+  dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)) %>%
+  ungroup()%>%
+  group_by(Cno) %>% dplyr::summarise(mean_inc=mean(mean_inc, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=mean_inc), col="grey40", size=0.8)+
+  scale_color_viridis_d()+
+  guides(col=guide_legend("species"))+
+  xlab("year") + ylab("diameter increment (cm)")+
+  ggtitle("growth increments for top 10 species")+
+  theme_bw()
+
+png("growth-precip-thailand/doc/display/spaghetti_top10.png", width=4, height=4, units="in", res=300)
+spagplot_top10
+dev.off()
+
 spagplot_all<-ggplot() +
@@
   xlab("year") + ylab("diameter increment (cm)")+
-  ggtitle("growth increments for top 4 species")+
+  ggtitle("growth increments for all species")+
   theme_bw()
@@
 spagplot_all
-dev.off()
-
-#sample size per species per year
-png("growth-precip-thailand/doc/display/top4_sample_size.png", width=8, height=4, units="in", res=300)
-top4 %>% 
-  group_by(sp, Cno) %>% 
-  dplyr::summarise(n=n())%>%
-  ggplot(aes(x=Cno, y=n, fill=factor(sp,levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
-  geom_bar(stat="identity")+
-  scale_fill_viridis_d()+
-  facet_grid(.~factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")))+
-  xlab("census number")+
-  ylab("sample size")+
-  guides(fill=guide_legend(title="species"))+
-  theme_bw()
 dev.off()
@@
 ylab("Annualised diameter increment (cm)")+theme_bw()+
+geom_segment(data=top4 %>% 
+  group_by(sp, Cno) %>% 
+  dplyr::summarise(n=n()/500),
+aes(x=factor(((Cno-3)/2)+2009), xend=factor(((Cno-3)/2)+2009), y=-3, yend=n-3, 
+fill=factor(sp,levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))), size=3)+
 facet_grid(.~factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")))+
@@
 xlab("species")+theme_bw()+
+  geom_text(data=top4 %>% group_by(sp) %>% dplyr::summarise(n=n()),
+   aes(x=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), y=2.2, 
+  label=n))+
   geom_hline(yintercept = 0, lty=3)+
@@
 growth_top4
+
+```
+```{r top4 species growth contribution, echo=F}
+
+#calculate contribution of each species to average growth for each year
+
+top4_sum<-top4 %>%
+  group_by(Cno) %>%
+  dplyr::mutate(tot_inc_all=sum(inc_annual, na.rm=T))%>%
+  ungroup()%>%
+  group_by(Cno, sp) %>%
+  dplyr::summarise(contrib=sum(inc_annual, na.rm=T)/mean(tot_inc_all))
+
+#plot contributions across years
+
+contrib_top4<-ggplot(top4_sum, 
+aes(x=factor(((Cno-3)/2)+2009), 
+y=contrib, fill=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
+  geom_bar(stat="identity")+
+  scale_fill_viridis_d()+
+  xlab("year")+
+  ylab("contribution to total growth")+
+  guides(fill=guide_legend(title="species"))+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle=90))
+
+png("growth-precip-thailand/doc/display/top4_contributions.png", width=4, height=4, units="in", res=300)
+contrib_top4
+dev.off()
+
+#repeat contributions plot for all species
+
+species_sum <- dendro_inc %>%
+  group_by(Cno) %>%
+  dplyr::mutate(tot_inc_all=sum(inc_annual, na.rm=T))%>%
+  ungroup()%>%
+  group_by(Cno, sp) %>%
+  dplyr::summarise(contrib=sum(inc_annual, na.rm=T)/mean(tot_inc_all))
+
+#plot contributions across years
+
+contrib_all<-ggplot(species_sum, 
+aes(x=factor(((Cno-3)/2)+2009), 
+y=contrib, fill=factor(sp)))+
+  geom_bar(stat="identity")+
+  scale_fill_viridis_d()+
+  xlab("year")+
+  ylab("contribution to total growth")+
+  guides(fill=guide_legend(title="species"))+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle=90))
+
+png("growth-precip-thailand/doc/display/species_contributions.png", width=12, height=4, units="in", res=300)
+contrib_all
+dev.off()
 
@@
 
-png("growth-precip-thailand/doc/display/resistance_map_top4.png", width=10, height=8, units="in", res=300)
+png("growth-precip-thailand/doc/display/resistance_map_top4.png", width=9, height=8, units="in", res=300)
 resistance_map

### 2024-07-09 scripts/explore_analysis.Rmd
updated script

@@
 
-#first fitting a line through the time series
-obs.model.dbh<-nlme::lme(calcDBH~Cno, random=~1|Tag, data=dendro.data,
-                         na.action=na.omit,
-                         method="ML", control=(msMaxIter=100)) #had to increase the number of iterations
-
-#standardised residuals calculated https://stat.ethz.ch/R-manual/R-devel/library/nlme/html/residuals.lme.html
-
-std.res.dbh<-residuals(obs.model.dbh, type="pearson")
-hist(std.res.dbh)
-dendro.data$std.res<-rep(NA, nrow(dendro.data))
-dendro.data$std.res[!is.na(dendro.data$calcDBH)]<-std.res.dbh
-#defining all observation with abs(residual) >3 as an outlier
-dendro.data$out<-factor(ifelse(abs(dendro.data$std.res)<=3, 0, 1))
-
-#find the trees which have outlier observations
-outs.Tag<-unique(dendro.data$Tag[dendro.data$out==1])
-
-#plot trees that have outliers
-ggplot(dendro.data[dendro.data$Tag %in%outs.Tag,],
-       aes(x = Cno, y = calcDBH, col=factor(out), group = Tag))+
-  geom_line(alpha=0.5)+geom_point()+
-  scale_color_manual(values=c("grey20", "red", "grey"))
-
-dendro.data.noouts<-dendro.data %>%
-  filter(abs(std.res)<=3)%>%
-  select(-c(std.res, out))
-  #mutate(dbh3=ifelse(abs(std.res)<=3, calcDBH, NA)) #%>% #create an outlier removed column
+dbh.data<-dendrobyband %>%
+  dplyr::rename(sp=SPCODE.UPDATE)%>%
+  mutate(year=year(Date2)) %>%
+  filter(is.na(dbh.flag), dbh.decimal.error=="none")%>%
+  group_by(Cno, sp, Tag) %>%
+  dplyr::summarise(dbh=DbhDendro[which(n.dendro==min(n.dendro))],
+                   Date2=Date2[which(n.dendro==min(n.dendro))],
+                   Cii=max(Cii, na.rm=T))
 
@@
 #add long-term increment to dendro data
-dendro.data.noouts<-merge(dendro.data.noouts, hkk_census, by="Tag", all.x=T)
 dendro.data<-merge(dendro.data, hkk_census, by="Tag", all.x=T)
@@
   filter(out.inc==0)
+
+dbh_inc<- dbh.data %>%
+  filter(Cno %in% alt_census)%>%
+  arrange(Cno)%>%
+  group_by(Tag) %>% #grouping by treeID to calculate increment
+  dplyr::mutate(dbh_min1=dplyr::lag(dbh, n=1, default=NA, order_by=Tag),
+         cii_min1=dplyr::lag(Cii, n=1, default=NA, order_by=Tag),
+         tdif=difftime(Date2, dplyr::lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
+         inc=dbh-dbh_min1,
+         inc_annual=(inc/as.numeric(tdif))*365,
+         inc.time=Cno-dplyr::lag(Cno, n=1, default=NA, order_by=Tag),
+         #removing increment value if the previous measure was before the last year
+         inc_annual=ifelse(inc.time==2, inc_annual, NA),
+         treeID=paste0("HKK.", Tag))%>%
+  filter(!is.na(inc))%>%
+  filter(!is.na(inc_annual))%>%
+  ungroup()%>%
+  dplyr::mutate(
+    diainc.scaled = scale(inc_annual),
+    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))%>%
+  filter(out.inc==0)
+
 ```
@@
 
+top4_dbh<-dbh_inc %>%
+filter(sp %in% c("SACCLI", "HOPEOD", "TETRNU", "POLYVI"))
+
 #spaghetti plot of all species together
@@
   #species plots
-  geom_line(data=top4 %>% group_by(Cno, sp) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), 
-  aes(x=((Cno-3)/2)+2009, y=mean_inc, 
+  geom_line(data=top4 %>% group_by(Cno, sp) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=median_inc, 
   group=factor(sp,levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), 
   col=factor(sp,levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
-  geom_line(data=top4 %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), 
-  aes(x=((Cno-3)/2)+2009, y=mean_inc), col="black", size=2)+
+  geom_line(data=top4 %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=median_inc), col="black", size=2)+
   scale_color_viridis_d()+
@@
 
+spagplot_top4_dbh<-ggplot() +
+  #species plots
+  geom_line(data=top4_dbh %>% group_by(Cno, sp) %>% 
+  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=median_inc, 
+  group=factor(sp,levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), 
+  col=factor(sp,levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
+  geom_line(data=top4_dbh %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=median_inc), col="black", size=2)+
+  scale_color_viridis_d()+
+  guides(col=guide_legend(title="species"))+
+  xlab("year") + ylab("diameter increment (cm)")+
+  ggtitle("growth increments for top 4 species")+
+  theme_bw()
+
 png("growth-precip-thailand/doc/display/spaghetti_top4.png", width=4, height=4, units="in", res=300)
@@
 
+#individual panels using facet_wrap
+spagplot_dbh_ind<-ggplot() +
+  geom_line(data = top4_dbh, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag), alpha=0.5) +
+  geom_line(data=top4_dbh %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=mean_inc), col="red")+
+  geom_line(data=top4_dbh %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=median_inc), col="blue")+
+  facet_wrap(factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))~., 
+  nrow=2)+
+  xlab("year") + ylab("diameter increment (cm)")+
+  ggtitle("growth increments for top 4 species from DBH measurements")+
+  theme_bw()
+
+png("growth-precip-thailand/doc/display/spaghetti_top4_dbh_ind.png", width=8, height=8, units="in", res=300)
+spagplot_dbh_ind
+dev.off()
+
+png("growth-precip-thailand/doc/display/spaghetti_top4_dbh.png", width=4, height=4, units="in", res=300)
+spagplot_top4_dbh
+dev.off()
+
 #spaghetti plot for top 10 species
@@
   #species plots
-  geom_line(data=dendro_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno, sp) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), 
-  aes(x=((Cno-3)/2)+2009, y=mean_inc, 
+  geom_line(data=dendro_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno, sp) %>% 
+  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=median_inc, 
   group=sp, col=sp))+
   #mean of all trees
-  geom_line(data=dendro_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), 
-  aes(x=((Cno-3)/2)+2009, y=mean_inc), col="black", size=2)+
+  geom_line(data=dendro_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno) %>% 
+  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=median_inc), col="black", size=2)+
   #mean of species
   geom_line(data=dendro_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno, sp) %>% 
-  dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)) %>%
+  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)) %>%
   ungroup()%>%
-  group_by(Cno) %>% dplyr::summarise(mean_inc=mean(mean_inc, na.rm=T)), 
-  aes(x=((Cno-3)/2)+2009, y=mean_inc), col="grey40", size=0.8)+
+  group_by(Cno) %>% dplyr::summarise(median_inc=mean(median_inc, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=median_inc), col="grey40", size=0.8)+
   scale_color_viridis_d()+
@@
 
+spagplot_top10_dbh<-ggplot()+
+  #species plots
+  geom_line(data=dbh_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno, sp) %>% 
+  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=median_inc, 
+  group=sp, col=sp))+
+  #mean of all trees
+  geom_line(data=dbh_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno) %>% 
+  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=median_inc), col="black", size=2)+
+  #mean of species
+  geom_line(data=dbh_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno, sp) %>% 
+  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)) %>%
+  ungroup()%>%
+  group_by(Cno) %>% dplyr::summarise(median_inc=mean(median_inc, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=median_inc), col="grey40", size=0.8)+
+  scale_color_viridis_d()+
+  guides(col=guide_legend("species"))+
+  xlab("year") + ylab("diameter increment (cm)")+
+  ggtitle("growth increments for top 10 species")+
+  theme_bw()
+
+png("growth-precip-thailand/doc/display/spaghetti_top10_dbh.png", width=4, height=4, units="in", res=300)
+spagplot_top10_dbh
+dev.off()
+
 spagplot_all<-ggplot() +
   #species plots
-  geom_line(data=dendro_inc %>% group_by(Cno, sp) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), 
-  aes(x=((Cno-3)/2)+2009, y=mean_inc, 
+  geom_line(data=dendro_inc %>% group_by(Cno, sp) %>% 
+  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=median_inc, 
   group=sp, col=sp))+
   #mean of all trees
-  geom_line(data=dendro_inc %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), 
-  aes(x=((Cno-3)/2)+2009, y=mean_inc), col="black", size=2)+
+  geom_line(data=dendro_inc %>% group_by(Cno) %>% 
+  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=median_inc), col="black", size=2)+
   #mean of species
   geom_line(data=dendro_inc %>% group_by(Cno, sp) %>% 
-  dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)) %>%
+  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)) %>%
   ungroup()%>%
-  group_by(Cno) %>% dplyr::summarise(mean_inc=mean(mean_inc, na.rm=T)), 
-  aes(x=((Cno-3)/2)+2009, y=mean_inc), col="grey40", size=0.8)+
+  group_by(Cno) %>% dplyr::summarise(median_inc=mean(median_inc, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=median_inc), col="grey40", size=0.8)+
   scale_color_viridis_d()+
@@
 
-growth_top4
-
-```
+#growth_top4_sum using dbh_inc data
+
+growth_top4_sum_dbh<-ggplot(top4_dbh, aes(x=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), y=inc_annual,
+                                              fill=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")),
+                                              color=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))
+                                              ))+ 
+  geom_violin(position = "dodge")+ 
+    geom_boxplot(position="dodge", col="black", fill="white", width=0.1)+
+ylab("Annualised diameter increment (cm)")+
+xlab("species")+theme_bw()+
+  geom_text(data=top4_dbh %>% group_by(sp) %>% dplyr::summarise(n=n()),
+   aes(x=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), y=2.2, 
+  label=n))+
+  geom_hline(yintercept = 0, lty=3)+
+  ggtitle("tree growth in HKK from tape measure")+
+  scale_fill_viridis_d()+
+  scale_color_viridis_d()+
+  guides(fill="none",
+         color="none")
+
+png("growth-precip-thailand/doc/display/top4_increments_sum_dbh.png", width=4, height=4, units="in", res=300)
+growth_top4_sum_dbh
+dev.off()
+
+```
+```{r comparing dbh and dendro, echo=F}
+
+#merge dbh and dendro data by Tag and Cno
+
+inc_all<-merge(dendro_inc, dbh_inc, by=c("Tag", "Cno"), all=T)
+
+#plot increments from dbh and dendro data
+
+png("growth-precip-thailand/doc/display/dbh_dendro_compare.png", width=4, height=4, units="in", res=300)
+ggplot(inc_all %>% filter(sp.x %in% c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), aes(x=inc_annual.x, y=inc_annual.y, 
+col=factor(sp.x, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
+  geom_point()+
+  scale_color_viridis_d()+
+  geom_abline(intercept=0, slope=1)+
+  xlab("dendroband increment (cm)")+
+  ylab("dbh increment (cm)")+
+  guides(col=guide_legend(title="species"))+
+  ggtitle("comparison of dbh and dendro increments")+
+  theme_bw()
+dev.off()
+```
+
 ```{r top4 species growth contribution, echo=F}

### 2024-07-10 scripts/01_prepdata.R
create a data prep file for final analysis

@@ -1,0 +1,200 @@
+
+# load libraries---------------------------
+library(tidyverse)
+library(lubridate)
+library(httr)
+library(janitor)
+
+#reading the GitHub PAT (stored privately and locally)
+#pat<-as.character(read.table("../data/HKK-dendro/git_pat.txt")[1,1])
+
+# remotes:::download("DendroByBand.RData",
+#                    "https://raw.githubusercontent.com/forestgeo/HKK-dendrobands/main/data/modified_data/DendroByBand.RData",
+#                    auth_token = pat)
+
+load("growth-precip-thailand/data/HKK-dendro/DendroByBand.RData")
+
+
+dendrobyband<-merge(dendrobyband, trees, by="Tag")
+
+hkk_OBSID$year<-year(hkk_OBSID$Date2)
+hkk_OBSID$month<-month(hkk_OBSID$Date2)
+hkk_OBSID$day<-day(hkk_OBSID$Date2)
+
+#correcting two sets of wrong dates
+
+hkk_OBSID$year[hkk_OBSID$year==2209]<-2009
+hkk_OBSID$year[hkk_OBSID$month==1 & hkk_OBSID$Cno==5]<-2011
+
+dendrobyband<-merge(dendrobyband, hkk_OBSID, by=c("OBSID", "Cno", "Date2", "Tag"))
+
+dendrobyband$Date2<-as.Date(paste0(dendrobyband$year,"-", dendrobyband$month, "-", dendrobyband$day))
+
+
+# calculate dbh by stem-------------------------------
+
+#remove all data with decimal errors, misplaced decimal
+
+dendro.data<-dendrobyband %>%
+  dplyr::rename(sp=SPCODE.UPDATE)%>%
+  mutate(year=year(Date2)) %>%
+  filter(is.na(DBH1_flag), DBH1_error=="none",
+         win.decimal.error=="none", is.na(win.flag))%>%
+  group_by(Cno, sp, Tag) %>%
+  dplyr::summarise(calcDBH=calcDBH[which(n.dendro==min(n.dendro))],
+                   Date2=Date2[which(n.dendro==min(n.dendro))],
+                   n.dendro=min(n.dendro), #which dendroband is it from?
+                   Cii=max(Cii, na.rm=T))
+
+dbh.data<-dendrobyband %>%
+  dplyr::rename(sp=SPCODE.UPDATE)%>%
+  mutate(year=year(Date2)) %>%
[…]

### 2024-07-10 doc/manuscript.Rmd
update methods for calculating resistance

@@
 *Resistance*
--Dendrometer band measurements - census protocol
--Calculated DBH at each instance from first DBH and window size using Condit's functions.
--Removed measures with large decimal errors, large outliers (> 3 SD). 
-- For all individuals with three points of measurements for each drought, calculated raw increment as difference between relevant censuses to get pre-drought increment and drought-year increment.
-- Resistance - difference between pre-drought and drought-year increment
+- Dendrometer band measurements - census protocol
+- Calculated DBH at each instance from first DBH and window size using Condit's functions.
+- Removed measures with large decimal errors, potential band misID, large outliers (> 3 SD). 
+- calculated increments for each individual for each census year using dendroband data and tape measurement
+- tape measurement does not have accuracy required to measure this effect, but used tape increments to detect outliers.
+- removed all increment measurements with more than 0.5 cm deviation from the 1:1 line for dendro and tape increments 
+- For each drought year, used drought-year increment.
+- for each tree, long-term growth was calculated as the mean growth across 20 years of full-plot censuses
+- Resistance - ratio of drought-year increment to long-term growth for each individual tree
 
 *Tree size*
-- calculated DBH for at the start of the pre-drought year from dendrometer band window
+- calculated DBH for at the start of the drought year from dendrometer band window
 

### 2024-07-10 scripts/explore_analysis.Rmd
update exploratory analysis script

@@
                    Date2=Date2[which(n.dendro==min(n.dendro))],
+                   n.dendro=min(n.dendro), #which dendroband is it from?
                    Cii=max(Cii, na.rm=T))
@@
                    Date2=Date2[which(n.dendro==min(n.dendro))],
+                   n.dendro=min(n.dendro),
                    Cii=max(Cii, na.rm=T))
@@
 
-#find the StemTags of trees that are in the dendro data
-#match using the DBH04 column
-
-#identify multistem trees from the census
-multistem<-hkk_census %>%
-  group_by(tag)%>%
-  dplyr::summarise(n.stem=n())%>%
-  rename(Tag=tag)
-
-multistem<-merge(multistem, trees, by="Tag", all.x=T)
-
-multistem<-multistem %>%
-rename(tag=Tag)%>%
-dplyr::mutate(dbh= (as.numeric(DBH.04)*10))
-
-hkk.stem3.dendro<-hkk.stem3 %>%
-  filter(tag %in% multistem$tag)
-
-#add the StemTag values from hkk.stem3.dendro to multistem 
-#match based on matching dbh and tag
-
-multistem<- multistem %>%
-dplyr::inner_join(select(hkk.stem3.dendro, c("tag", "StemTag", "dbh")), 
-by=c("tag", "dbh"))
-
-sum(multistem$tag != multistem$StemTag)
-#there are no stemtags with values different from Tag
-#all stems with dendrobands have them on the main stem?
-
-#hkk_census<-hkk_census %>%
-#  filter(StemTag %in% multistem$StemTag)
-
 #since dendrobands are installed on trees with tag==StemTag
@@
 #plot average increment for the top 4 species
-
 hkk_census2<-hkk_census %>%
@@
 dendro.data<-merge(dendro.data, hkk_census, by="Tag", all.x=T)
+dbh.data<-merge(dbh.data, hkk_census, by="Tag", all.x=T)
 
@@
   dplyr::mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
+  n.dendro_diff=n.dendro-dplyr::lag(n.dendro, n=1, default=NA, order_by=Tag),
          cii_min1=dplyr::lag(Cii, n=1, default=NA, order_by=Tag),
@@
   dplyr::mutate(dbh_min1=dplyr::lag(dbh, n=1, default=NA, order_by=Tag),
+  n.dendro_diff=n.dendro-dplyr::lag(n.dendro, n=1, default=NA, order_by=Tag),
          cii_min1=dplyr::lag(Cii, n=1, default=NA, order_by=Tag),
@@
 ```
+
+```{r comparing dbh and dendro and identify outliers, echo=F}
+
+#merge dbh and dendro data by Tag and Cno
+inc_all<-merge(dendro_inc, dbh_inc, by=c("sp", "Tag", "Cno"), all=T)
+
+#Perpendicular distance from point 'a' to a line with 'slope' and 'intercept'
+dist_point_line <- function(a, slope, intercept) {
+    b = c(1, intercept+slope)
+    c = c(-intercept/slope,0)       
+    v1 <- b - c
+    v2 <- a - b
+    m <- cbind(v1,v2)
+    return(abs(det(m))/sqrt(sum(v1*v1)))
+}
+
+inc_all<-inc_all %>%
+  dplyr::rename(inc_annual.dendro=inc_annual.x, 
+  inc_annual.dbh=inc_annual.y,
+  band.diff.dendro=n.dendro_diff.x,
+  band.diff.dbh=n.dendro_diff.y) %>%
+  #split into three size classes
+  dplyr::mutate(size_class= ifelse(dbh<35, "small",
+  ifelse(dbh<70, "medium", "large")),
+  size_class=factor(size_class, levels=c("small", "medium", "large"))
+  #find the deviation of the increment from the 1:1 line
+  )%>%
+  select(sp, Tag, Cno, inc_annual.dendro, inc_annual.dbh, dbh, 
+  size_class, band.diff.dbh, band.diff.dendro)
+
+#using the distance to the line
+dev.1<-mapply(function(x)dist_point_line(c(inc_all$inc_annual.dendro[x], inc_all$inc_annual.dbh[x]), 1, 0), x=1:nrow(inc_all))
+inc_all$dev.1<-dev.1
+
+#compare dbh and dendro increments for all species
+
+cor.all<-cor(inc_all$inc_annual.dendro, inc_all$inc_annual.dbh, use="complete.obs")
+
+dbh_dendro_all<-ggplot(inc_all, aes(x=inc_annual.dendro, y=inc_annual.dbh, col=factor(band.diff.dendro+band.diff.dbh, levels=c(0, 1, 2))))+
+  geom_point(alpha=0.5)+
+  geom_abline(intercept=0, slope=1)+
+  scale_color_manual(values=c("black", "orange", "tomato", "grey40"))+
+  xlab("dendroband increment (cm)")+
+  ylab("dbh increment (cm)")+
+  geom_rug(alpha=0.3)+
+  geom_text(aes(x=2, y=4, label = paste0("r = ", round(cor.all,3))), vjust = -0.5) +
+  guides(col=guide_legend(title="band_diff"))+
+  ggtitle("comparison of dbh and dendro increments")+
+  theme_bw()
+
+#plot distribution of deviations
+
+pct.0.5<-round(sum(inc_all$dev.1>=0.5, na.rm=T)*100/nrow(inc_all),2)
+pct.1<-round(sum(inc_all$dev.1>=1, na.rm=T)*100/nrow(inc_all),2)
+pct.2<-round(sum(inc_all$dev.1>=2, na.rm=T)*100/nrow(inc_all),2)
+pct.0.1<-round(sum(inc_all$dev.1>=0.1, na.rm=T)*100/nrow(inc_all),2)
+
+png("growth-precip-thailand/doc/display/deviation_distribution.png", width=4, height=4, units="in", res=300)
+ggplot()+
+  geom_density(data=inc_all, aes(x=dev.1), size=1.2)+
+  ggtitle("deviation of dendroband and tape-based \nincrements from each other")+
+  geom_vline(xintercept=c(0.1, 0.5, 1, 2), lty=2, col="orange")+
+  geom_text(aes(x=c(0.1, 0.5, 1, 2), y=5, 
+  label = paste0(">",c(0.1, 0.5, 1, 2), " = ", c(pct.0.1, pct.0.5, pct.1, pct.2), "%")), 
+  angle=90, vjust = 1, size=2)+
+  xlab("distribution of deviations from 1:1 line")+
+  theme_bw()
+dev.off()
+```
+
+```{r separating outliers, echo=F}
+
+#outlier dataframe
+outliers<-inc_all %>% filter(dev.1>=0.5)
+write.csv(outliers, "growth-precip-thailand/data/HKK-dendro/inc_outliers.csv", row.names=F)
+
+#clean dataframe
+
+dendro_inc<-merge(dendro_inc, inc_all %>% select(Tag, Cno, sp, dev.1), 
+by=c("Tag", "Cno", "sp"), all.x=T)
+
+dendro_inc_clean<-dendro_inc %>% filter(dev.1<0.5)
+
+#plot increments from dbh and dendro data with outliers removed
+
+cor.noouts<-dendro_inc_clean %>%
+dplyr::summarise(cor=cor(inc_annual.dendro, inc_annual.dbh, use="complete.obs"))
+cor.noouts<-cor.noouts$cor[1]
+
+dbh_dendro_noouts<-ggplot(inc_all %>% filter(dev.1<0.5), aes(x=inc_annual.dendro, y=inc_annual.dbh, col=factor(band.diff.dendro+band.diff.dbh, levels=c(0, 1, 2))))+
+  geom_point(alpha=0.5)+
+  geom_abline(intercept=0, slope=1)+
+  scale_color_manual(values=c("black", "orange", "tomato", "grey40"))+
+  xlab("dendroband increment (cm)")+
+  ylab("dbh increment (cm)")+
+  geom_rug(alpha=0.3)+
+  geom_text(aes(x=2, y=4, label = paste0("r = ", round(cor.noouts,3))), vjust = -0.5) +
+  guides(col=guide_legend(title="band_diff"))+
+  ggtitle("dbh and dendro increments (outliers removed)")+
+  theme_bw()
+dev.off()
+
+library(gridExtra)
+png("growth-precip-thailand/doc/display/dbh_dendro_compare_all.png", width=8, height=4, units="in", res=300)
+grid.arrange(dbh_dendro_all, dbh_dendro_noouts, ncol=2)
+dev.off()
+
+```
+```{r compare deviations by species and size, echo=F}
+
+ggplot(inc_all %>% filter(sp.x %in% c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), aes(x=inc_annual.x, y=inc_annual.y, 
+col=factor(sp.x, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
+  geom_point()+
+  scale_color_viridis_d()+
+  geom_abline(intercept=0, slope=1)+
+  xlab("dendroband increment (cm)")+
+  ylab("dbh increment (cm)")+
+  guides(col=guide_legend(title="species"))+
+  ggtitle("comparison of dbh and dendro increments")+
+  theme_bw()
+
+#split this by size classes
+
+cors.size<-inc_all %>% 
+filter(!is.na(size_class))%>%
+  group_by(size_class) %>% 
+  dplyr::summarise(cor=cor(inc_annual.dendro, inc_annual.dbh, use="complete.obs"))
+
+png("growth-precip-thailand/doc/display/dbh_dendro_compare_size.png", width=8, height=4, units="in", res=300)
+ggplot(inc_all %>% filter(!is.na(size_class)), aes(x=inc_annual.dendro, y=inc_annual.dbh, 
+col=size_class))+
+  geom_point(alpha=0.5)+
+  geom_abline(intercept=0, slope=1)+
+  xlab("dendroband increment (cm)")+
+  ylab("dbh increment (cm)")+
+  facet_grid(.~size_class)+
+  geom_text(data=cors.size, aes(x=2, y=4, label = paste0("r = ", round(cor,3))), vjust = -0.5) +
+  guides(col=guide_legend(title="size class"))+
+  ggtitle("comparison of dbh and dendro increments")+
+  theme_bw()
+dev.off()
+
+```
+
 ```{r species wise increments, echo=F}
-
-#filter increments for two species
-
-saccli<-dendro_inc %>%
-  filter(sp == "SACCLI")
-
-hopeod <- dendro_inc %>%
-  filter(sp == "HOPEOD")
-
-tetrnu<-dendro_inc %>%
-  filter(sp == "TETRNU")
-
-polyvi <- dendro_inc %>%
-  filter(sp == "POLYVI")
-
-#spaghetti plot saccli and hopeod increments over time
-
-saccli_plot<-ggplot() +
-  geom_line(data = saccli, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag), alpha=0.5) +
-  geom_line(data=saccli %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=mean_inc), col="red")+
-  geom_line(data=saccli %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=median_inc), col="blue")+
-  xlab("year") + ylab("diameter increment (cm)")+
-  ggtitle("growth increments for SACCLI (deciduous)")+
-  theme_bw()
-
-hopeod_plot<-ggplot() +
-  geom_line(data = hopeod, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag), alpha=0.5) +
-  geom_line(data=hopeod %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=mean_inc), col="red")+
-  geom_line(data=hopeod %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=median_inc), col="blue")+
-  xlab("year") + ylab("diameter increment (cm)")+
-  ggtitle("growth increments for HOPEOD (evergreen)")+
-  theme_bw()
-
-tetrnu_plot<-ggplot() +
-  geom_line(data = tetrnu, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag), alpha=0.5) +
-  geom_line(data=tetrnu %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=mean_inc), col="red")+
-  geom_line(data=tetrnu %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=median_inc), col="blue")+
-  xlab("year") + ylab("diameter increment (cm)")+
-  ggtitle("growth increments for TETRNU (deciduous)")+
-  theme_bw()
-
-polyvi_plot<-ggplot() +
-  geom_line(data = polyvi, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag), alpha=0.5) +
-  geom_line(data=polyvi %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=mean_inc), col="red")+
-  geom_line(data=polyvi %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=median_inc), col="blue")+
-  xlab("year") + ylab("diameter increment (cm)")+
-  ggtitle("growth increments for POLYVI (evergreen)")+
-  theme_bw()
-
-library(gridExtra)
-png("growth-precip-thailand/doc/display/ind_increments.png", width=8, height=8, units="in", res=300)
-grid.arrange(hopeod_plot, tetrnu_plot, polyvi_plot, saccli_plot, nrow=2)
-dev.off()
 
@@
 
-top4_dbh<-dbh_inc %>%
+top4_noouts<-dendro_inc_clean %>%
 filter(sp %in% c("SACCLI", "HOPEOD", "TETRNU", "POLYVI"))
+
+#individual panels using facet_wrap
+spagplot_ind<-ggplot() +
+  geom_line(data = top4, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag), alpha=0.5) +
+  geom_line(data=top4 %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=mean_inc), col="red")+
+  geom_line(data=top4 %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=median_inc), col="blue")+
+  facet_wrap(factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))~., 
+  nrow=2)+
+  xlab("year") + ylab("diameter increment (cm)")+
+  ggtitle("growth increments for top 4 species")+
+  theme_bw()
+
+png("growth-precip-thailand/doc/display/spaghetti_top4_ind.png", width=8, height=8, units="in", res=300)
+spagplot_ind
+dev.off()
+
+#individual panels using facet_wrap for the no outs dataset
+spagplot_ind_noouts<-ggplot() +
+  geom_line(data = top4_noouts, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag), alpha=0.5) +
+  geom_line(data=top4_noouts %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=mean_inc), col="red")+
+  geom_line(data=top4_noouts %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=median_inc), col="blue")+
+  facet_wrap(factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))~., 
+  nrow=2)+
+  xlab("year") + ylab("diameter increment (cm)")+
+  ggtitle("growth increments for top 4 species (no outliers)")+
+  theme_bw()
+
+png("growth-precip-thailand/doc/display/spaghetti_top4_ind_noouts.png", width=8, height=8, units="in", res=300)
+spagplot_ind_noouts
+dev.off()
 
@@
 
-spagplot_top4_dbh<-ggplot() +
-  #species plots
-  geom_line(data=top4_dbh %>% group_by(Cno, sp) %>% 
-  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
-  aes(x=((Cno-3)/2)+2009, y=median_inc, 
-  group=factor(sp,levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), 
-  col=factor(sp,levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
-  geom_line(data=top4_dbh %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
-  aes(x=((Cno-3)/2)+2009, y=median_inc), col="black", size=2)+
-  scale_color_viridis_d()+
-  guides(col=guide_legend(title="species"))+
-  xlab("year") + ylab("diameter increment (cm)")+
-  ggtitle("growth increments for top 4 species")+
-  theme_bw()
-
 png("growth-precip-thailand/doc/display/spaghetti_top4.png", width=4, height=4, units="in", res=300)
 spagplot_top4
-dev.off()
-
-#individual panels using facet_wrap
-spagplot_dbh_ind<-ggplot() +
-  geom_line(data = top4_dbh, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag), alpha=0.5) +
-  geom_line(data=top4_dbh %>% group_by(Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=mean_inc), col="red")+
-  geom_line(data=top4_dbh %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=median_inc), col="blue")+
-  facet_wrap(factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))~., 
-  nrow=2)+
-  xlab("year") + ylab("diameter increment (cm)")+
-  ggtitle("growth increments for top 4 species from DBH measurements")+
-  theme_bw()
-
-png("growth-precip-thailand/doc/display/spaghetti_top4_dbh_ind.png", width=8, height=8, units="in", res=300)
-spagplot_dbh_ind
-dev.off()
-
-png("growth-precip-thailand/doc/display/spaghetti_top4_dbh.png", width=4, height=4, units="in", res=300)
-spagplot_top4_dbh
 dev.off()
@@
 
-spagplot_top10_dbh<-ggplot()+
+#spaghetti plot for top 10 species with outliers removed
+
+sp_top10<-c("SACCLI", "HOPEOD", "TETRNU", "POLYVI", "VATIHA", 
+"DIPTAL", "ALPHVE", "GARCSP", "GARUPI", "NEOLOB")
+
+spagplot_top10_noouts<-ggplot()+
   #species plots
-  geom_line(data=dbh_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno, sp) %>% 
+  geom_line(data=dendro_inc_clean %>% filter(sp %in% sp_top10)%>% group_by(Cno, sp) %>% 
   dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
@@
   #mean of all trees
-  geom_line(data=dbh_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno) %>% 
+  geom_line(data=dendro_inc_clean %>% filter(sp %in% sp_top10)%>% group_by(Cno) %>% 
   dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
@@
   #mean of species
-  geom_line(data=dbh_inc %>% filter(sp %in% sp_top10)%>% group_by(Cno, sp) %>% 
+  geom_line(data=dendro_inc_clean %>% filter(sp %in% sp_top10)%>% group_by(Cno, sp) %>% 
   dplyr::summarise(median_inc=median(inc_annual, na.rm=T)) %>%
@@
 
-png("growth-precip-thailand/doc/display/spaghetti_top10_dbh.png", width=4, height=4, units="in", res=300)
-spagplot_top10_dbh
-dev.off()
+png("growth-precip-thailand/doc/display/spaghetti_top10_noouts.png", width=4, height=4, units="in", res=300)
+spagplot_top10_noouts
+dev.off()
+
+#all species with and without outliers
 
@@
 spagplot_all
+dev.off()
+
+spagplot_all_noouts<-ggplot() +
+  #species plots
+  geom_line(data=dendro_inc_clean %>% group_by(Cno, sp) %>% 
+  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=median_inc, 
+  group=sp, col=sp))+
+  #mean of all trees
+  geom_line(data=dendro_inc_clean %>% group_by(Cno) %>% 
+  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=median_inc), col="black", size=2)+
+  #mean of species
+  geom_line(data=dendro_inc_clean %>% group_by(Cno, sp) %>% 
+  dplyr::summarise(median_inc=median(inc_annual, na.rm=T)) %>%
+  ungroup()%>%
+  group_by(Cno) %>% dplyr::summarise(median_inc=mean(median_inc, na.rm=T)), 
+  aes(x=((Cno-3)/2)+2009, y=median_inc), col="grey40", size=0.8)+
+  scale_color_viridis_d()+
+  guides(col="none", group="none")+
+  xlab("year") + ylab("diameter increment (cm)")+
+  ggtitle("growth increments for all species (no outs)")+
+  theme_bw()
+
+png("growth-precip-thailand/doc/display/spaghetti_all_noouts.png", width=4, height=4, units="in", res=300)
+spagplot_all_noouts
 dev.off()
@@
 ```
-```{r comparing dbh and dendro, echo=F}
-
-#merge dbh and dendro data by Tag and Cno
-
-inc_all<-merge(dendro_inc, dbh_inc, by=c("Tag", "Cno"), all=T)
-
-#plot increments from dbh and dendro data
-
-png("growth-precip-thailand/doc/display/dbh_dendro_compare.png", width=4, height=4, units="in", res=300)
-ggplot(inc_all %>% filter(sp.x %in% c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), aes(x=inc_annual.x, y=inc_annual.y, 
-col=factor(sp.x, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
-  geom_point()+
-  scale_color_viridis_d()+
-  geom_abline(intercept=0, slope=1)+
-  xlab("dendroband increment (cm)")+
-  ylab("dbh increment (cm)")+
-  guides(col=guide_legend(title="species"))+
-  ggtitle("comparison of dbh and dendro increments")+
-  theme_bw()
-dev.off()
-```
 
@@
 
-resistance_2010_top4 <- top4 %>%
-  filter(Cno %in% c(3, 5))%>%
+#drought years 2010 and 2015
+# 2010 - cno 5, 2015 - cno 15
+
+resistance_top4 <- top4 %>%
+  filter(Cno %in% c(5, 15))%>%
   group_by(Tag) %>% #grouping by treeID to calculate increment
-  dplyr::mutate(inc_min1=dplyr::lag(inc_annual, n=1, default=NA, order_by=Tag),
+  dplyr::mutate(
+    yr=ifelse(Cno==5, 2010, 2015),
         resistance.div = inc_annual/avg_inc,
-         resistance.dif = inc_annual-inc_min1,
-         resistance.prop = (inc_annual-inc_min1)/avg_inc)%>%
-  filter(!is.na(inc_min1), !is.infinite(resistance.div))
-
-resistance_2015_top4 <- top4 %>%
-  filter(Cno %in% c(11, 13, 15))%>%
-  group_by(Tag) %>% #grouping by treeID to calculate increment
-  mutate(inc_min1=dplyr::lag(inc_annual, n=1, default=NA, order_by=Tag),
-         resistance.div = inc_annual/avg_inc,
-         resistance.dif = inc_annual-inc_min1,
-         resistance.prop = (inc_annual-inc_min1)/avg_inc)%>%
-  filter(!is.na(inc_min1), !is.infinite(resistance.div))
+         resistance.dif = inc_annual-avg_inc,
+         resistance.prop = (inc_annual-avg_inc)/avg_inc)%>%
+  filter(!is.infinite(resistance.div))
 
@@
 #map of resistance.div quantiles for top 4 species
-resistance_all<-rbind(resistance_2010_top4, resistance_2015_top4)
-resistance_all$yr<-c(rep(2010, nrow(resistance_2010_top4)), rep(2015, nrow(resistance_2015_top4)))
-
-resistance_all<-merge(resistance_all, select(trees, -"SPCODE.UPDATE"), by="Tag")
+
+resistance_div<-ggplot(data = resistance_all, aes(x=resistance.div)) +
+  geom_density(alpha=0.5) +
+  facet_grid(.~yr)+
+  xlim(-5, 5)+
+  xlab("inc/long-term increment")+
+  geom_vline(xintercept = c(0,1), lty=2)+
+  theme_bw()
+
+png("growth-precip-thailand/doc/display/resistance_div_top4.png", width=6, height=4, units="in", res=300)
+resistance_div
+dev.off()
+
+resistance_div_sp<-ggplot(data = resistance_all, 
+aes(x=resistance.div, 
+col=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")),
+fill=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
+  geom_density(alpha=0.5) +
+  facet_grid(factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))~yr)+
+  scale_color_viridis_d()+ scale_fill_viridis_d()+
+  guides(col="none", fill="none")+
+  xlim(-5, 5)+
+  xlab("inc/long-term increment")+
+  geom_vline(xintercept = c(0,1), lty=2)+
+  theme_bw()
+
+png("growth-precip-thailand/doc/display/resistance_div_top4_sp.png", width=6, height=12, units="in", res=300)
+resistance_div_sp
+dev.off()
 
@@
 
-resistance_2010 <- dendro_2010 %>%
+resistance_2010 <- dendro_inc %>%
+filter(Cno %in% c(3, 5))%>%
   group_by(Tag) %>% #grouping by treeID to calculate increment

### 2024-07-10 scripts/explore_analysis.Rmd
updated resistance values for top 4 species
uses clean dataset

@@
 
-```{r plot increments}
+```{r plot increments old code, echo=F, eval=F}
 
@@
 
-resistance_top4 <- top4 %>%
+resistance_top4 <- top4_noouts %>%
   filter(Cno %in% c(5, 15))%>%
@@
          resistance.prop = (inc_annual-avg_inc)/avg_inc)%>%
-  filter(!is.infinite(resistance.div))
+         ungroup()%>%
+  filter(!is.infinite(resistance.div))%>%
+         group_by(Cno)%>%
+         #making quantiles for each year
+         dplyr::mutate(
+         quantiles_div=cut(resistance.div,
+         breaks=quantile(resistance.div, 
+         #probs=c(0.25, 0.5, 0.75, 1), 
+         probs=seq(0, 1, 0.25),
+         na.rm=T),
+         labels=c("1", "2", "3", "4")))
+
+#merge with tree attributes
+
+resistance_top4<-merge(resistance_top4, select(trees, -"SPCODE.UPDATE"), by="Tag", all.x=T)
 
@@
 
-quantiles_2010_top4<-quantile(resistance_2010_top4$resistance.div, 
+quantiles_2010_top4<-quantile(resistance_top4$resistance.div, 
                          probs=c(0.25, 0.5, 0.75, 1), na.rm=T)
@@
 
-#split the data into categories based on quantiles
-
-resistance_2010_top4$quantile<-ifelse(resistance_2010_top4$resistance.div<quantiles_2010_top4[1], 1, ifelse(resistance_2010_top4$resistance.div<quantiles_2010_top4[2], 2, ifelse(resistance_2010_top4$resistance.div<quantiles_2010_top4[3], 3, 4)))
-resistance_2015_top4$quantile<-ifelse(resistance_2015_top4$resistance.div<quantiles_2015_top4[1], 1, ifelse(resistance_2015_top4$resistance.div<quantiles_2015_top4[2], 2, ifelse(resistance_2015_top4$resistance.div<quantiles_2015_top4[3], 3, 4)))
-
 #map of resistance.div quantiles for top 4 species
 
-resistance_div<-ggplot(data = resistance_all, aes(x=resistance.div)) +
+resistance_div<-ggplot(data = resistance_top4, aes(x=resistance.div)) +
   geom_density(alpha=0.5) +
@@
 
-resistance_div_sp<-ggplot(data = resistance_all, 
+resistance_div_sp<-ggplot(data = resistance_top4, 
 aes(x=resistance.div, 
@@
 
-resistance_map<-ggplot(resistance_all, aes(x=X, y=Y, size=as.numeric(DBH.04), col=quantile))+
+resistance_map<-ggplot(resistance_top4, aes(x=X, y=Y, size=as.numeric(DBH.04), col=quantiles_div))+
   geom_point(alpha=0.5)+

### 2024-07-10 scripts/explore_analysis.Rmd
updated resistance values plot @teixeirak

@@
          dplyr::mutate(
-         quantiles_div=cut(resistance.div,
-         breaks=quantile(resistance.div, 
-         #probs=c(0.25, 0.5, 0.75, 1), 
-         probs=seq(0, 1, 0.25),
-         na.rm=T),
-         labels=c("1", "2", "3", "4")))
+         quantiles_div=ntile(resistance.div, 4))
 
@@
 
-resistance_map<-ggplot(resistance_top4, aes(x=X, y=Y, size=as.numeric(DBH.04), col=quantiles_div))+
+resistance_map<-ggplot(resistance_top4, aes(x=X, y=Y, size=as.numeric(DBH.04), 
+col=quantiles_div))+
   geom_point(alpha=0.5)+
@@
 
-#2010 drought--------
-
-resistance_2010 <- dendro_inc %>%
-filter(Cno %in% c(3, 5))%>%
+resistance <- dendro_inc_clean %>%
+  filter(Cno %in% c(5, 15))%>%
   group_by(Tag) %>% #grouping by treeID to calculate increment
-  mutate(inc_min1=dplyr::lag(inc_annual, n=1, default=NA, order_by=Tag),
+  dplyr::mutate(
+    yr=ifelse(Cno==5, 2010, 2015),
         resistance.div = inc_annual/avg_inc,
-         resistance.dif = inc_annual-inc_min1,
-         resistance.prop = (inc_annual-inc_min1)/avg_inc)%>%
-  filter(!is.na(inc_min1), !is.infinite(resistance.div))
-
-resistance_2010<-merge(resistance_2010, select(dec_williams, -"X"), by="sp", all.x=T)
-
-resistance_2010_div<-ggplot(data = resistance_2010, aes(x=resistance.div)) +
+         resistance.dif = inc_annual-avg_inc,
+         resistance.prop = (inc_annual-avg_inc)/avg_inc)%>%
+         ungroup()%>%
+  filter(!is.infinite(resistance.div))%>%
+         group_by(Cno)%>%
+         #making quantiles for each year
+         dplyr::mutate(
+         quantiles_div=ntile(resistance.div, 4),
+         quantiles_dif=ntile(resistance.dif, 4),
+         quantiles_prop=ntile(resistance.prop, 4))
+
+#merge with tree attributes
+
+resistance<-merge(resistance, select(trees, -"SPCODE.UPDATE"), by="Tag", all.x=T)
+resistance<-merge(resistance, dec_williams, by="sp", all.x=T)
+
+#distribution and map
+
+resistance_div<-ggplot(data = resistance, aes(x=resistance.div)) +
   geom_density(alpha=0.5) +
-  geom_vline(xintercept = 0, lty=2)+
-  theme_bw()
-
-resistance_2010_dif<-ggplot(data = resistance_2010, aes(x=resistance.dif)) +
-  geom_density(alpha=0.5) +
-  theme_bw()
-
-#2015 drought---------
-
-resistance_2015 <- dendro_2015 %>%
-  group_by(Tag) %>% #grouping by treeID to calculate increment
-  mutate(inc_min1=dplyr::lag(inc_annual, n=1, default=NA, order_by=Tag),
-         resistance.div = inc_annual/avg_inc,
-         resistance.dif = inc_annual-inc_min1,
-         resistance.prop = (inc_annual-inc_min1)/avg_inc)%>%
-  filter(!is.na(inc_min1), !is.infinite(resistance.div))
-
-resistance_2015<-merge(resistance_2015, select(dec_williams, -"X"), by="sp", all.x=T)
-
-resistance_2015_div<-ggplot(data = resistance_2015, aes(x=resistance.div)) +
-  geom_density(alpha=0.5) +
-  geom_vline(xintercept = 0, lty=2)+
-  theme_bw()
-
-resistance_2015_dif<-ggplot(data = resistance_2015, aes(x=resistance.dif)) +
-  geom_density(alpha=0.5) +
-  theme_bw()
-
-#2019 drought--------------
-
-resistance_2019 <- dendro_2019 %>%
-  group_by(Tag) %>% #grouping by treeID to calculate increment
-  mutate(inc_min1=dplyr::lag(inc_annual, n=1, default=NA, order_by=Tag),
-         resistance.div = inc_annual/inc_min1,
-         resistance.dif = inc_annual-inc_min1,
-         resistance.prop = (inc_annual-inc_min1)/inc_min1)%>%
-  filter(!is.na(inc_min1), !is.infinite(resistance.div))
-
-resistance_2019<-merge(resistance_2019, select(dec_williams, -"X"), by="sp", all.x=T)
-
-resistance_2019_div<-ggplot(data = resistance_2019, aes(x=resistance.div)) +
-  geom_density(alpha=0.5) +
-  theme_bw()
-
-resistance_2019_dif<-ggplot(data = resistance_2019, aes(x=resistance.dif)) +
-  geom_density(alpha=0.5) +
-  theme_bw()
-
-png("../doc/display/resistance_dif.png", width=4, height=8, units="in", res=300)
-grid.arrange(resistance_2010_dif, resistance_2015_dif, resistance_2019_dif, nrow=3)
-dev.off()
-
-library(gridExtra)
-
-#png("growth-precip-thailand/doc/display/resistance_div.png", width=4, height=8, units="in", res=300)
-#grid.arrange(resistance_2010_div, resistance_2015_div, resistance_2019_div, nrow=3)
-#dev.off()
-
-png("growth-precip-thailand/doc/display/resistance_div.png", width=8, height=4, units="in", res=300)
-grid.arrange(resistance_2010_div, resistance_2015_div, nrow=1)
+  facet_grid(.~yr)+
+  xlim(-5, 5)+
+  xlab("inc/long-term increment")+
+  ggtitle("resistance values for all trees")+
+  geom_vline(xintercept = c(0,1), lty=2)+
+  theme_bw()
+
+png("growth-precip-thailand/doc/display/resistance_div.png", width=6, height=4, units="in", res=300)
+resistance_div
 dev.off()
@@
 ```{r resistance map}
-
-quantiles_2010_dif<-quantile(resistance_2010$resistance.dif, 
-                         probs=c(0.25, 0.5, 0.75, 1))
-
-quantiles_2015_dif<-quantile(resistance_2015$resistance.dif, 
-                         probs=c(0.25, 0.5, 0.75, 1))
-
-quantiles_2019_dif<-quantile(resistance_2019$resistance.dif, 
-                         probs=c(0.25, 0.5, 0.75, 1))
-
-resistance_2010$quantile<-ifelse(resistance_2010$resistance.dif<quantiles_2010_dif[1], 1, ifelse(resistance_2010$resistance.dif<quantiles_2010_dif[2], 2,
-            ifelse(resistance_2010$resistance.dif<quantiles_2010_dif[3], 3, 4))
-
-                                 )
-
-resistance_2015$quantile<-ifelse(resistance_2015$resistance.dif<quantiles_2015_dif[1], 1, ifelse(resistance_2015$resistance.dif<quantiles_2015_dif[2], 2,
-            ifelse(resistance_2015$resistance.dif<quantiles_2015_dif[3], 3, 4))
-
-                                 )
-
-resistance_2019$quantile<-ifelse(resistance_2019$resistance.dif<quantiles_2019_dif[1], 1, ifelse(resistance_2019$resistance.dif<quantiles_2019_dif[2], 2,
-            ifelse(resistance_2019$resistance.dif<quantiles_2019_dif[3], 3, 4))
-
-                                 )
-
-quantiles_2010_div<-quantile(resistance_2010$resistance.div, 
-                         probs=c(0.25, 0.5, 0.75, 1), na.rm=T)
-
-quantiles_2015_div<-quantile(resistance_2015$resistance.div, 
-                         probs=c(0.25, 0.5, 0.75, 1), na.rm=T)
-
-quantiles_2019_div<-quantile(resistance_2019$resistance.div, 
-                         probs=c(0.25, 0.5, 0.75, 1), na.rm=T)
-
-resistance_2010$quantile.div<-ifelse(resistance_2010$resistance.div<quantiles_2010_div[1], 1, ifelse(resistance_2010$resistance.div<quantiles_2010_div[2], 2,
-            ifelse(resistance_2010$resistance.div<quantiles_2010_div[3], 3, 4))
-
-                                 )
-
-resistance_2015$quantile.div<-ifelse(resistance_2015$resistance.div<quantiles_2015_div[1], 1, ifelse(resistance_2015$resistance.div<quantiles_2015_div[2], 2,
-            ifelse(resistance_2015$resistance.div<quantiles_2015_div[3], 3, 4))
-
-                                 )
-
-resistance_2019$quantile.div<-ifelse(resistance_2019$resistance.div<quantiles_2019_div[1], 1, ifelse(resistance_2019$resistance.div<quantiles_2019_div[2], 2,
-            ifelse(resistance_2019$resistance.div<quantiles_2019_div[3], 3, 4))
-
-                                 )
-
-resistance_all<-rbind(resistance_2010, resistance_2015)
-resistance_all$yr<-c(rep(2010, nrow(resistance_2010)), rep(2015, nrow(resistance_2015)))
-
-resistance_all<-merge(resistance_all, select(trees, -"SPCODE.UPDATE"), by="Tag")
 

### 2024-07-10 scripts/01_prepdata.R

@@ -198,3 +198,39 @@ dendro_inc<-merge(dendro_inc, inc_all %>% select(Tag, Cno, sp, dev.1),
 by=c("Tag", "Cno", "sp"), all.x=T)
 
 dendro_inc_clean<-dendro_inc %>% filter(dev.1<0.5)
+
+
+#calculate resistance to drought-----------------------------
+
+#here we define drought years as 2010 and 2015
+
+resistance <- dendro_inc_clean %>%
+  filter(Cno %in% c(5, 15))%>%
+  group_by(Tag) %>% #grouping by treeID to calculate increment
+  dplyr::mutate(
+    yr=ifelse(Cno==5, 2010, 2015),
+        resistance.div = inc_annual/avg_inc,
+         resistance.dif = inc_annual-avg_inc,
+         resistance.prop = (inc_annual-avg_inc)/avg_inc)%>%
+         ungroup()%>%
+  filter(!is.infinite(resistance.div))%>%
+         group_by(Cno)%>%
+         #making quantiles for each year
+         dplyr::mutate(
+         quantiles_div=ntile(resistance.div, 4),
+         quantiles_dif=ntile(resistance.dif, 4),
+         quantiles_prop=ntile(resistance.prop, 4))
+
+#merge with tree attributes
+
+resistance<-merge(resistance, select(trees, -"SPCODE.UPDATE"), by="Tag", all.x=T)
+
+
+#read in the deciduousness values
+dec_williams<-read.csv("growth-precip-thailand/data/HKK-dendro/species.csv")
+
+dec_williams<-dec_williams %>%
+  filter(n.tree >= 5)
+
+
+resistance<-merge(resistance, dec_williams, by="sp", all.x=T)

### 2024-07-12 scripts/explore_analysis.Rmd
updated code, dag for individual

@@
 fill=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
-  geom_density(alpha=0.5) +
+  geom_density() +
   facet_grid(factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))~yr)+
@@
 
-```{r drought resistance}
-
+```{r associated datasets, echo=F}
+
+#deciduousness
 dec_williams<-read.csv("growth-precip-thailand/data/HKK-dendro/species.csv")
@@
   filter(n.tree >= 5)
+
+#habitat data
+habitat<-read.csv("growth-precip-thailand/data/HKK-other/HKK_habtype(in).csv")
+
+#plot a tile map of habitat types
+ggplot(habitat, aes(x=x, y=y, fill=factor(hab)))+
+  geom_tile()+
+  scale_fill_viridis_d(option="E", direction=-1)+
+  theme_minimal()+
+  coord_fixed()
+
+habitat<-habitat %>%
+  dplyr::mutate(Col = x/20,
+                Row = y/20)
+
+trees<-merge(trees, select(habitat, -c(X, x, y)), by=c("Col", "Row"), all.x=T)
+
+#elevation
+#although .rdata, this needs to be read using read.table - thanks Bier!
+
+hkk_elev<-read.table("growth-precip-thailand/data/HKK-other/CTFSElev_hkk.rdata")
+png("growth-precip-thailand/doc/display/elevation_resistance_map_2015.png", width=8, height=4, units="in", res=300)
+ggplot(hkk_elev, aes(x=col.x, y=col.y, fill=col.elev))+
+geom_tile()+
+scale_fill_viridis_c(option="E")+
+#add contours
+geom_contour(aes(z=col.elev), colour="black")+
+geom_point(data=resistance_clean %>% filter(yr==2015, sp %in% c("HOPEOD", "TETRNU", "POLYVI", "SACCLI")), 
+aes(x=X.x, y=Y, size=as.numeric(DBH.04)), inherit.aes=F, col="black", alpha=0.5)+
+facet_wrap(.~sp, nrow=2)+
+theme_bw()+
+coord_fixed()
+dev.off()
+
+#convert to topographic wetness index
+
+library(sf)
+library(terra)
+
+#first convert to raster
+elev_rast<-rast(hkk_elev$col.elev, nrows=100, ncols=100, xmin=0, xmax=100, ymin=0, ymax=100)
+```
+
+```{r drought resistance}
 
@@
          quantiles_dif=ntile(resistance.dif, 4),
-         quantiles_prop=ntile(resistance.prop, 4))
+         quantiles_prop=ntile(resistance.prop, 4))%>%
+         #remove trees that have -ve long-term increment
+          filter(avg_inc>0)
 
@@
 resistance_div
+dev.off()
+
+resistance_prop<-ggplot(data = resistance, aes(x=resistance.prop)) +
+  geom_density(alpha=0.5) +
+  facet_grid(.~yr)+
+  xlim(-5, 5)+
+  xlab("proportion of change from long-term mean")+
+  ggtitle("resistance values for all trees")+
+  geom_vline(xintercept = c(-1, 0,1), lty=2)+
+  theme_bw()
+
+png("growth-precip-thailand/doc/display/resistance_prop.png", width=6, height=4, units="in", res=300)
+resistance_prop
 dev.off()
@@
 ```
+
+```{r resistance with predictor variables, echo=F}
+
+resistance_clean <- resistance %>%
+filter(abs(resistance.div)<5)
+
+sum(resistance$resistance.div>5, na.rm=T)/nrow(resistance)
+
+#resistance across species
+
+sp_top10<-c("SACCLI", "HOPEOD", "TETRNU", "POLYVI", "VATIHA", 
+"DIPTAL", "ALPHVE", "GARCSP", "GARUPI", "NEOLOB")
+
+png("growth-precip-thailand/doc/display/resistance_bysp.png", width=8, height=6, units="in", res=300)
+ggplot(data=resistance_clean %>% filter(sp %in% sp_top10), aes(x=factor(sp), y=resistance.prop, fill=factor(sp)))+ 
+  geom_violin()+
+  geom_boxplot(fill="white", width=0.1)+
+  geom_hline(yintercept=c(-1, 0, 1), lty=2)+
+  facet_wrap(~yr)+
+  ylab("growth decrease from long term mean")+
+  #make ylab percentage
+  scale_y_continuous(labels=scales::percent)+ 
+  scale_fill_viridis_d()+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle=90))
+dev.off()
+
+#resistance with dbh
+
+ggplot()+
+  geom_point(data=resistance_clean, aes(x=calcDBH_min1, 
+  y=resistance.prop))+
+  geom_smooth(data=resistance_clean, aes(x=calcDBH_min1, 
+  y=resistance.prop), method="lm")+
+  facet_grid(yr~.)+
+  theme_bw()
+
+#resistance with cii
+
+ggplot(data=resistance_clean, aes(x=factor(cii_min1), fill=factor(cii_min1),
+  #y=resistance.div))+
+  y=resistance.prop))+
+  geom_violin()+
+  geom_boxplot(fill="white", width=0.1)+
+  facet_grid(yr~.)+
+  theme_bw()
+
+#resistance with deciduousness
+
+ggplot()+
+  geom_point(data=resistance_clean, aes(x=williams_dec, 
+  #y=resistance.div))+
+  y=resistance.prop))+
+  geom_smooth(data=resistance_clean, aes(x=williams_dec, 
+  #y=resistance.div), method="lm")+
+  y=resistance.prop), method="lm")+
+  facet_grid(yr~.)+
+  theme_bw()
+
+#resistance with habitat
+
+ggplot(data=resistance_clean, aes(x=factor(hab), 
+  y=resistance.div, fill=factor(hab)))+
+  geom_violin()+
+  geom_boxplot(fill="white", width=0.1)+
+  scale_fill_viridis_d(option="E", direction=-1)+
+  facet_grid(yr~.)+
+  theme_bw()
+
+```
 ```{r basic model}
 
-resistance_all$ciiF<-ordered(resistance_all$cii_min1)
-
-model_2010<-glm(resistance.div ~ calcDBH_min1 + ciiF + williams_dec+
+resistance$ciiF<-ordered(resistance$cii_min1)
+resistance$hab<-factor(resistance$hab)
+
+model_2010<-glm(resistance.div ~ calcDBH_min1 + ciiF + 
+williams_dec+hab+
                   calcDBH_min1*ciiF + 
@@
                   ciiF*williams_dec,
-                data=resistance_all[resistance_all$yr==2010,])
+                data=resistance[resistance$yr==2010,])
 
@@
 
-model_2015<-glm(resistance.div ~ calcDBH_min1 + ciiF + williams_dec+
+model_2015<-glm(resistance.div ~ calcDBH_min1 + ciiF + 
+williams_dec+hab+
                   calcDBH_min1*ciiF + 
@@
                   ciiF*williams_dec,
-                data=resistance_all[resistance_all$yr==2015,])
+                data=resistance[resistance$yr==2015,])
 
@@
 
-resistance_model<-resistance_all %>%
-  select(resistance.div, calcDBH_min1, ciiF, williams_dec, yr, quantile)
+resistance_model<-resistance %>%
+  select(resistance.div, calcDBH_min1, ciiF, williams_dec, hab, yr, quantiles_div)
 
@@
 library(nlme)
-model_all<-lm(resistance.div ~ calcDBH_min1 + ciiF + williams_dec+
+model_all<-lm(resistance.div ~ calcDBH_min1 + ciiF + hab+ williams_dec+
                   calcDBH_min1*ciiF + 
@@
 
-model_all
+summary(model_all)
 anova(model_all)
@@
 ```
+
+```{r dag, echo=F}
+
+library(dagitty)
+library(ggdag)
+growth.dag<-dagitty( 'dag {
+bb="0,0,1,1"
+crowding [pos="0.427,0.650"]
+exposure [exposure,pos="0.405,0.325"]
+growth [outcome,pos="0.583,0.335"]
+size [exposure,pos="0.409,0.151"]
+topography [exposure,pos="0.259,0.454"]
+water [latent,pos="0.415,0.471"]
+exposure -> growth
+size -> exposure
+topography -> water
+water -> crowding
+water -> growth
+}' )
+
+png("growth-precip-thailand/doc/display/dag_ind.png", width=4, height=4, units="in", res=300)
+ggdag_status(growth.dag, text = FALSE, use_labels = "name") +
+  guides(color = FALSE) +  # Turn off legend
+  theme_dag()
+dev.off()
+
+impliedConditionalIndependencies( growth.dag )
+```
+
+```{r species DAG, echo=F}
+
+sp.dag<-dagitty('dag {
+bb="0,0,1,1"
+"canopy position" [exposure,pos="0.276,0.255"]
+"median size" [exposure,pos="0.440,0.108"]
+deciduousness [exposure,pos="0.458,0.618"]
+exposure [latent,pos="0.458,0.238"]
+growth [outcome,pos="0.609,0.329"]
+water [latent,pos="0.449,0.427"]
+"canopy position" -> exposure
+"median size" -> exposure
+deciduousness -> water
+exposure -> growth
+water -> growth
+}')
+
+png("growth-precip-thailand/doc/display/dag_sp.png", width=4, height=4, units="in", res=300)
+ggdag_status(sp.dag, text = FALSE, use_labels = "name") +
+  guides(color = FALSE) +  # Turn off legend
+  theme_dag()
+dev.off()
+
+impliedConditionalIndependencies( sp.dag )
+
+```

### 2024-07-12 doc/manuscript.Rmd
fleshing out intro

@@
 
-Tree woody growth key is a key component of aboveground productivity and affects the global carbon carbon cycle, but the drivers of its variation are poorly understood.
-- Woody growth contributes x% of the aboveground woody productivity
-- Long-lasting effects on the terrestrial carbon pool
-- Intrinsic and extrinsic factors affecting variation.
+Tropical tree woody growth is a key component of aboveground productivity and affects the global carbon carbon cycle, but its sensitivity to ongoing global change is poorly understood.  
+- Woody growth has long-lasting effects on the terrestrial carbon pool but is sensitive to interannual variation in temperature and precipitation.  
+- Extreme events like drought, with temperature and precipitation outside normal ranges that can lead to growth reductions that varies across species and individuals [@bennett_larger_2015; @mcgregor_tree_2021].
+- the extent of this variation in drought responses is poorly understood.  
+- with changing climate patterns across the global tropics, including novel climatic regimes [@dahinden_future_2017], there is a need to understand the drivers of this variation, to be able to predict and respond to forest responses to drought.
 
+Tree species in seasonally dry forests have diverse allocation strategies and traits which can result in differential sensitivities to drought  
+- Drought events can be different from each other, but generally lead to low soil moisture, low water tables, low vapour pressure and high temperature [@chitra-tarak_hydraulicallyvulnerable_2021].  
+- Results in thermal stress on tissues, evaporative loss, leaf turgor loss and cavitation.  
+- Species can be conservative or acquisitive in water and nutrient uptake; differential allocation to growth/survival.  
+- Species with drought tolerant traits like deep roots, more negative turgor loss point and ... are more resistant to drought [@kunert_leaf_2021; @refs].  
+- Some evidence that both understory and emergent species can have reduced survival during drought [@machado_forest_2023].  
+- Deciduous species, with shorter duration with leaves can have acquisitive strategies during leaf on [@desouza_drought_2020].   
+-  Strategies also affects species distribution with evergreen species covarying with soil moisture [@kunert_leaf_2021] and potentially affecting survival and growth.   
+- Many hydraulic-related traits vary with tree height [ @vinod_thermal_2023], including the frequency of dry season deciduous leaf loss -- both within species and at the community level [@condit_ref; @meakem].
+
+Within species, size, exposure and location can affect sensitivity to drought.  
 - Large trees tend to undergo greater growth declines during drought compared to smaller trees [e.g., @bennett_larger_2015; @mcgregor_tree_2021; @anderson-teixeira_joint_2022].   
@@
 `r fig_DAG`).
-
 - There is theory and evidence that tree size itself matters.  
@@
 - Even when trees are accessing deeper water, this does not mean that they're in better shape during drought.
-Rather, trees that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@rutuja]
+Rather, trees that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021].
 - indeed, there is evidence that trees near streams undergo greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta] during drought
-- rather than necessarily helping during drought, water access will shape the size the size and traits of species living in habitat, with stream habitats tending to have larger trees [@ref] and more evergreen trees, and also their average growth rate
-
-- Many hydraulic-related traits vary with tree height [ @vinod_thermal_2023], including the frequency of dry season deciduous leaf loss -- both within species and at the community level [@condit_ref; @meakem].
-- (discuss drought strategies - avoidance / adaptation )
+- rather than necessarily helping during drought, water access will shape the size and traits of species living in habitat, with stream habitats tending to have larger trees [@ref] and more evergreen trees, and also their average growth rate
 
@@
 Here we use a 14-year record of dendrometer band measurements in seasonally dry forest in Thailand to test the vulnerability of tropical tree growth to drought.
-We expect variation in the absolute growth during drought years and drought resistance (growth in drought year/growth in a previous year with normal rainfall) across trees.
+We expect variation in drought sensitivity (growth in drought year/growth in a previous year with normal rainfall) across trees.
+We ask: i) How much do species vary in drought sensitivity?
+ii) How much of the variation among species is explained by deciduousness?
+iii) How much of the intraspecific variation is explained by tree height *per se*, crown exposure and the habitat of the tree (proxy for water and nutrient availability).
 We hypothesise that growth during drought is affected by:

### 2024-08-19 point_dendro_design.Rmd
point dendro individuals
@teixeirak notebook with interactive maps for point dendro selected individuals. Potential individuals for 4 species and 8 species design.

---
title: "design for point dendrometer study"
author: ""
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```
```{r libraries, echo=F}
library(tidyverse)
library(viridis)
library(ggplot2)
library(plotly)
```

```{r species list, echo=F}

#read dendro_inc_clean.csv

dendro_inc_clean <- read.csv("data/HKK-dendro/dendro_inc_clean.csv")
load("data/HKK-dendro/dendrobyband.RData")
shortlist<-c("DIPTAL", "HOPEOD", "TETRNU", "VATIHA", "CHUKTA", "POLYVI", "DIOSWI", "CROTRO", "SACCLI", "LAGETO", "GARUPI", "AFZEXY")

sp.list <- read.csv("data/HKK-dendro/point_dendro_sp.csv")

sp.list
```
```{r species chars, echo=F}

#make dendroband plots across years for each species

#subset data to these species
point_sp<-dendro_inc_clean %>%
filter(sp %in% shortlist)

point_sp_tree<-trees %>%
filter(SPCODE.UPDATE %in% shortlist)

spagplot_ind<-ggplot() +
  geom_line(data = point_sp, aes(x=((Cno-3)/2)+2009, y=inc_annual, group=Tag), alpha=0.5) +
  geom_line(data=point_sp %>% group_by(sp,Cno) %>% dplyr::summarise(mean_inc=mean(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=mean_inc), col="red")+
  geom_line(data=point_sp %>% group_by(sp,Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), aes(x=((Cno-3)/2)+2009, y=median_inc), col="blue")+
  facet_wrap(factor(sp)~., 
  nrow=3)+
  xlab("year") + ylab("diameter increment (cm)")+
  ggtitle("growth increments for candidate species")+
  theme_bw()

#png("spagplot_pointdendro_sp.png", width=12, height=8, units="in", res=300)
spagplot_ind
#dev.off()

spagplot_pointsp<-ggplot() +
  #species plots
  geom_line(data=point_sp %>% group_by(Cno, sp) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=median_inc, 
  group=factor(sp), 
  col=factor(sp)))+
  geom_line(data=point_sp %>% group_by(Cno) %>% dplyr::summarise(median_inc=median(inc_annual, na.rm=T)), 
  aes(x=((Cno-3)/2)+2009, y=median_inc), col="black", size=2)+
  scale_color_viridis_d()+
  guides(col=guide_legend(title="species"))+
  xlab("year") + ylab("diameter increment (cm)")+
  ggtitle("growth increments for candidate species")+
  theme_bw()

#png("spagplot_pointdendro_spsum.png", width=4, height=4, units="in", res=300)
spagplot_pointsp
#dev.off()

```
```{r elevation, echo=F}
#elevation data
hkk_elev<-read.table("data/HKK-other/CTFSElev_hkk.rdata")

```

```{r maps, echo=F}
# make maps of dendroband individuals of species
# base map is habitat map

#habitat data
habitat<-read.csv("data/HKK-other/HKK_habtype(in).csv")

#plot a tile map of habitat types
#png("dendro_spmap.png", width=16, height=8, units="in", res=300)
ggplot()+
  geom_tile(data=habitat, aes(x=x, y=y, fill=factor(hab)))+
  scale_fill_viridis_d(option="E", direction=-1)+
  geom_point(data=point_sp_tree,
  alpha=0.5, aes(x=X, y=Y, size=as.numeric(DBH.04)))+
  theme_minimal()+
  facet_wrap(factor(SPCODE.UPDATE)~.)+
  coord_fixed()
#dev.off()

habitat<-habitat %>%
  dplyr::mutate(Col = x/20,
                Row = y/20)

trees<-merge(trees, select(habitat, -c(X, x, y)), by=c("Col", "Row"), all.x=T)

```

```{r individual species, results='asis', echo=F}

p1<-ggplot()+
  geom_tile(data=habitat, aes(x=x, y=y, fill=factor(hab)))+
  scale_fill_viridis_d(option="E", direction=-1)+
  geom_point(data=point_sp_tree %>% filter(SPCODE.UPDATE=="CROTRO"),
  alpha=0.5, aes(x=X, y=Y, size=as.numeric(DBH.04), text=paste0("Tag: ",Tag)))+
  theme_minimal()+
  #facet_wrap(factor(SPCODE.UPDATE)~.)+
  coord_fixed()

ggplotly(p1, tooltip = c("text", "size"))

sps<-c("HOPEOD", "TETRNU", "SACCLI", "POLYVI")

for(i in 1:4){

p<-ggplot()+
  geom_tile(data=habitat, aes(x=x, y=y, fill=factor(hab)))+
  scale_fill_viridis_d(option="E", direction=-1)+
  geom_point(data=point_sp_tree %>% 
  #filter(SPCODE.UPDATE=="HOPEOD"),
  filter(SPCODE.UPDATE==sps[i]),
  alpha=0.5, aes(x=X, y=Y, size=as.numeric(DBH.04), text=paste0("Tag: ",Tag)))+
  theme_minimal()+
  ggtitle(paste0("species: ", sps[i]))+
  #facet_wrap(factor(SPCODE.UPDATE)~.)+
  coord_fixed()

print(htmltools::tagList(ggplotly(p, tooltip = c("text", "size"))))
}

```

```{r plot of selected trees for 4 species, results='asis', echo=F}

pdendro_trees<-read.csv("point_dendro_trees.csv")

sp4<-c("HOPEOD", "TETRNU", "SACCLI", "POLYVI")

pdendro_confirm<-pdendro_trees %>%
filter(sp %in% sp4) %>%
filter(confirm4=="y")

#filter the main dataset with these tags

point_trees<-trees %>%
filter(Tag %in% pdendro_confirm$Tag)

#plot these on the habitat map

p <- ggplot()+
  geom_tile(data=habitat, aes(x=x+10, y=y+10, fill=factor(hab)), alpha=0.5)+
  scale_fill_viridis_d(option="E", direction=-1)+
  geom_contour(data=hkk_elev, aes(x=col.x, y=col.y, z=col.elev), col="black")+
  geom_point(data=point_trees, aes(x=X, y=Y, 
  col=factor(SPCODE.UPDATE),
  size=as.numeric(DBH.04), 
  text=paste0("Tag: ",Tag)))+
  theme_minimal()+
  coord_fixed()+
  ggtitle("selected trees for 4 species design")+
  xlab("") + ylab("")+
  guides(fill="none", size="none",
  col=guide_legend(title="species"))

png("point_dendro_sp4.png", width=12, height=7, units="in", res=300)
p+geom_text(data=point_trees, aes(x=X, y=Y, label=Tag), size=3)
dev.off()

ggplotly(p, tooltip = c("text", "size"))

```

```{r plot of selected trees for 8 species, results='asis', echo=F}

pdendro_trees<-read.csv("point_dendro_trees.csv")

pdendro_confirm<-pdendro_trees %>%
#filter(sp %in% sp4) %>%
filter(confirm8=="y")

#filter the main dataset with these tags

point_trees<-trees %>%
filter(Tag %in% pdendro_confirm$Tag)

#plot these on the habitat map

p <- ggplot()+
  geom_tile(data=habitat, aes(x=x+10, y=y+10, fill=factor(hab)), alpha=0.5)+
  scale_fill_viridis_d(option="E", direction=-1)+
  geom_contour(data=hkk_elev, aes(x=col.x, y=col.y, z=col.elev), col="black")+
  geom_point(data=point_trees, aes(x=X, y=Y, 
  col=factor(SPCODE.UPDATE),
  size=as.numeric(DBH.04), 
  text=paste0("Tag: ",Tag)))+
  theme_minimal()+
  coord_fixed()+
  ggtitle("selected trees for 8 species design")+
  xlab("") + ylab("")+
  guides(fill="none", size="none",
  col=guide_legend(title="species"))

ggplotly(p, tooltip = c("text", "size"))

```

### 2024-08-19 doc/manuscript.Rmd
cleanup commits

@@
 
+Growth reductions from droughts are difficult to estimate reliably from whole-plot inventories.  
+- Droughts are different but generally leads to low soil moisture, low water tables, vapour pressure, high temperature.
+- Results in thermal stress, evaporative loss, leaf turgor loss, embolism/cavitation  High mortality and growth reductions across global forests, including the tropics  
+- Inventory data are often not at annual scales; smooths over drought/no-drought period.  
+- Reductions are often small (<1 mm), and difficult to detect with tape measurements.
+
 Tree species in seasonally dry forests have diverse allocation strategies and traits which can result in differential sensitivities to drought  
@@
 - @chen_hydraulic_2022  
-
 - There is theory and evidence that crown exposure matters [ @vinod_thermal_2023;  @refs_in_ vinod_thermal_2023; @scharnweber_confessions_2019].
@@
 - Trees with exposed crowns suffered significant crown dieback at greater rates in the 2012-16 CA drought [@ma_tree_2023].  
-
 - Water access....  
@@
 We ask: i) How much do species vary in drought sensitivity?
-ii) How much of the variation among species is explained by deciduousness?
+ii) How does deciduousness influence the drought sensitivity of a species?
 iii) How much of the intraspecific variation is explained by tree height *per se*, crown exposure and the habitat of the tree (proxy for water and nutrient availability).

### 2025-01-16 results/results.qmd
results presentation

---
title: "HKK drought sensitivity results"
author: "Krishna Anujan"
date: "2025-01-10"
format: 
  revealjs:
    theme: simple
    transition: slide
    slideNumber: true
    center: true
    embed-resources: true
---

## Sensitivity of tree growth to precipitation in seasonally dry tropical forests using long-term dendrometer band measurements

## Objective

::: {.columns}
::: {.column width="40%"}

Disentangling the effects of size, exposure and water availability on drought sensitivity of tree growth

:::

::: {.column width="60%"}

![](../doc/display/hypotheses.png)

:::
:::

## Defining drought sensitivity

$sensitivity_{tree} = \frac{drought.year.increment - mean.annual.increment_{tree}}{mean.annual.increment_{tree}}$

![](../doc/display/sens_definition.png)

## Data

::: {.columns}
::: {.column width="60%"}
| **Variable**   | **Description**                                      |
|----------------|------------------------------------------------------|
| <span style="background-color: #000000; color: #ffffff;"> sensitivity</span>    | Late wet/early dry dendrometer band measurements |
| <span style="background-color: #f7b731; color: #ffffff;">exposure   </span>    | Crown Illumination index at previous dendroband census |
| <span style="background-color: #e15f41; color: #ffffff;">size   </span>        | DBH at previous dendroband census |
| <span style="background-color: #546de5; color: #ffffff;">water</span>          | Topographic wetness index |
:::

::: {.column width="40%"}
![](../doc/display/hypotheses.png)
:::

:::

## ENSO drought years were associated with growth declines across species...

::: {.columns}
::: {.column width="50%}
![](../doc/display/spaghetti_top10_new.png)
:::

::: {.column width="50%"}
![](../results/plots/climate_data_timeseries_2007_2019.png)
:::
:::

## ...but we observed large variation in drought sensitivity within and among species

::: {.columns}
::: {.column width="50%"}
![](../doc/display/sens_all.png)
:::

::: {.column width="50%"}
![](../doc/display/sens_top10.png)
:::
:::

##

::: {.columns}
::: {.column width="50%"}
How well do species traits predict average sensitivity of species?

$sensitivity_{tree}~ 1+ (1|Species)$

![](../results/plots/non_negative/ranefs_intercept_bysp_new.png)
:::

::: {.column width="50%"}
![](../results/plots/non_negative/sp_intercept_plot.png)
:::
:::

## Models

We ran 3 different models

1. varying intercepts by species  
$sensitivity_{tree} = \beta_{DBH}DBH_{0} + \beta_{TWI}TWI + \beta_{CII}CII_{0} + (1|Species)$

2. varying intercepts + predictors scaled within species  
$sensitivity_{tree} = \beta_{DBH}DBH_sp_{0} + \beta_{TWI}TWI_sp + \beta_{CII}CII_sp_{0} + (1|Species)$

3. varying slopes + predictors scaled within species  
$sensitivity_{tree} = \beta_{DBH}DBH_sp_{0} + \beta_{TWI}TWI_sp + \beta_{CII}CII_sp_{0} + (1+DBH+CII+TWI|Species)$

##

::: {.columns}
::: {.column width="30%"}
Variation in sensitivity among individuals was explained more by crown exposure than size per se
:::

::: {.column width="70%"}
![](../results/plots/non_negative/coefs_all_plot.png)
:::
:::

## Higher exposure was associated with more negative growth sensitivities across species

##

::: {.columns}
::: {.column width="30%"}
To corroborate these results we modelled growth occurrence as a binary response.

And found similar, stronger influence of exposure on growth occurrence
:::

::: {.column width="70%"}
![](../results/plots/zero_growth/coefs_all_plot.png)
:::
:::

## When modelled without exposure, the effect of size was more negative as exposure is strongly predicted by size

::: {.columns}
::: {.column width="30%"}
![](../doc/display/hypotheses.png)
:::

::: {.column width="70%"}
![](../results/plots/non_negative/coefs_mediation.png)
:::
:::

This implies mediation effects of size on exposure

## Take-home messages

1. In 2015, the stronger ENSO drought, deciduous and slow-growing species had stronger negative sensitivities.
2. While two droughts decreased growth on most trees, more exposed trees were more negatively impacted than their less exposed counterparts
3. While exposure effects were stronger than size effects on drought sensitivity, we see mediation effects.

### 2025-01-16 scripts/01_prepdata.R
updated scripts with prep and models

@@ -256,7 +256,7 @@ ggplot(tree_incs, aes(x = avg_inc)) +
   theme_minimal()
 
 # how many trees with avg_inc < 0.001?
-#sum(tree_incs$avg_inc<0.001) #58 trees
+# sum(tree_incs$avg_inc<0.001) #58 trees
 # nrow(tree_incs)
 
 # remove these trees from the dendro_inc_clean dataset
@@ -406,3 +406,177 @@ sp_vars <- merge(max.DBH, dplyr::select(dec_williams, c("sp", "williams_dec")), 
 
 # make an RData object with these dataframes
 saveRDS(list(tree_vars = tree_vars, tree.time = tree.time, sp_vars = sp_vars), "data/HKK-dendro/sensitivity_data.RData")
+
+
+# environmental variables---------------------------
+
+# read in the environmental variables
+
+chirps <- read.csv("data/climate/CHIRPS_HKK.csv")
+spei <- read.csv("data/climate/SPEI_HKK_from_GEE.csv")
+ffstation <- read.csv("data/climate/WeatherData_ALL_forest_fire_research_station.csv")
+
+# make the ffstation data monthly
+
+head(ffstation)
+
+ffstation_monthly <- ffstation %>%
+  group_by(YearII, Month) %>%
+  dplyr::summarise(
+    precipitation = sum(Precipitation),
+    dry_days = sum(Precipitation == 0),
+    data = "ffstation"
+  ) %>%
+  rename(year = YearII, month = Month) %>%
+  select(year, month, precipitation, dry_days, data)
+
+
+# make the other data have year and month columns
+library(lubridate)
+chirps <- chirps %>%
+  dplyr::mutate(
+    Date = as.Date(system.time_start, format = "%B %d, %Y"),
+    year = year(Date),
+    month = month(Date),
+    data = "chirps"
+  ) %>%
+  rename(dry_days = dry_day) %>%
+  select(year, month, precipitation, dry_days, data)
+
+spei <- spei %>%
+  dplyr::mutate(
+    Date = as.Date(system.time_start, format = "%B %d, %Y"),
+    year = year(Date),
+    month = month(Date),
+    data = "spei_01"
+  ) %>%
+  rename(spei_val = SPEI_01_month) %>%
+  select(year, month, spei_val, data)
+
+
+# plot correlations between chirps and ffstation
+precip_all <- rbind(chirps, ffstation_monthly) %>%
[…]

### 2025-01-16 scripts/02_models.R

@@ -21,6 +21,8 @@ tree.time <- merge(tree.time, tree_vars, by = "Tag", all.x = TRUE)
 
 colnames(tree.time)
 
+colours <- c("#e15f41", "#546de5", "#f7b731")
+
 # standardise the variables within species and across all--------------------
 
 tree.time <- tree.time %>%
@@ -29,7 +31,8 @@ tree.time <- tree.time %>%
         cii_min1_scaled = scale(cii_min1, center = TRUE, scale = TRUE),
         cii_min1_scaled = ifelse(cii_min1_scaled == "NaN", 0, cii_min1_scaled),
         twi_scaled = scale(twi, center = TRUE, scale = TRUE),
-        median_inc_scaled = scale(median_inc, center = TRUE, scale = TRUE)
+        median_inc_scaled = scale(median_inc, center = TRUE, scale = TRUE),
+        avg_inc_tree_scaled = scale(avg_inc_tree, center = TRUE, scale = TRUE)
     ) %>%
     group_by(Species) %>%
     # scale while retaining original values
@@ -43,7 +46,8 @@ tree.time <- tree.time %>%
     # remove large outliers for each year
     group_by(yr) %>%
     # find sens.prop values that are 3 sds from the mean
-    dplyr::mutate(sens.prop = ifelse(sens.prop > mean(sens.prop, na.rm = TRUE) + 3 * sd(sens.prop, na.rm = TRUE), NA, sens.prop)) %>%
+    #    dplyr::mutate(sens.prop = ifelse(mean(sens.prop, na.rm = TRUE) + 3 * sd(sens.prop, na.rm = TRUE) > sens.prop & sens.prop > mean(sens.prop, na.rm = TRUE) - 3 * sd(sens.prop, na.rm = TRUE), sens.prop, NA)) %>%
+    dplyr::mutate(sens.prop = ifelse(mean(sens.prop, na.rm = TRUE) + 4 * sd(sens.prop, na.rm = TRUE) > sens.prop & sens.prop > mean(sens.prop, na.rm = TRUE) - 4 * sd(sens.prop, na.rm = TRUE), sens.prop, NA)) %>%
     filter(!is.na(sens.prop) & !is.na(cii_min1) & !is.na(calcDBH_min1) & !is.na(twi)) %>%
     ungroup()
 
@@ -90,6 +94,12 @@ spagplot_top10 <- ggplot() +
             dplyr::summarise(median_inc = median(inc_annual, na.rm = T)),
[…]

### 2025-01-16 scripts/04_models_noslow.R

@@ -44,6 +44,6 @@ tree.time <- tree.time %>%
     # remove large outliers for each year
     group_by(yr) %>%
     # find sens.prop values that are 3 sds from the mean
-    dplyr::mutate(sens.prop = ifelse(sens.prop > mean(sens.prop, na.rm = TRUE) + 3 * sd(sens.prop, na.rm = TRUE), NA, sens.prop)) %>%
+    # dplyr::mutate(sens.prop = ifelse(sens.prop > mean(sens.prop, na.rm = TRUE) + 3 * sd(sens.prop, na.rm = TRUE), NA, sens.prop)) %>%
     filter(!is.na(sens.prop) & !is.na(cii_min1) & !is.na(calcDBH_min1) & !is.na(twi)) %>%
     ungroup()

### 2025-01-16 scripts/05_models_zerogrowth.R

@@ -1,0 +1,556 @@
+# models using the zero growth framework
+
+# Load required libraries---------------------
+library(tidyverse)
+library(brms)
+
+# load data--------------------------
+rm(list = ls())
+datasets <- readRDS("data/HKK-dendro/sensitivity_data.RData")
+
+tree.time <- datasets$tree.time
+tree_vars <- datasets$tree_vars
+sp_vars <- datasets$sp_vars
+
+colnames(sp_vars)[1] <- "Species"
+
+# merge tree_vars and sp_vars
+tree_vars <- merge(tree_vars, sp_vars, by = "Species", all.x = TRUE)
+# merge tree_vars and tree.time
+tree.time <- merge(tree.time, tree_vars, by = "Tag", all.x = TRUE)
+
+colnames(tree.time)
+
+colours <- c("#e15f41", "#546de5", "#f7b731")
+
+# standardise the variables within species and across all--------------------
+
+tree.time <- tree.time %>%
+    dplyr::mutate(
+        calcDBH_min1_scaled = scale(calcDBH_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled = scale(cii_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled = ifelse(cii_min1_scaled == "NaN", 0, cii_min1_scaled),
+        twi_scaled = scale(twi, center = TRUE, scale = TRUE),
+        median_inc_scaled = scale(median_inc, center = TRUE, scale = TRUE),
+        avg_inc_tree_scaled = scale(avg_inc_tree, center = TRUE, scale = TRUE),
+        sens.bin = ifelse(sens.prop > 0, 1, 0)
+    ) %>%
+    group_by(Species) %>%
+    # scale while retaining original values
+    dplyr::mutate(
+        calcDBH_min1_scaled_sp = scale(calcDBH_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled_sp = scale(cii_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled_sp = ifelse(cii_min1_scaled_sp == "NaN", 0, cii_min1_scaled_sp),
+        twi_scaled_sp = scale(twi, center = TRUE, scale = TRUE)
+    ) %>%
+    ungroup() %>%
+    # remove large outliers for each year
+    group_by(yr) %>%
+    # find sens.prop values that are 3 sds from the mean
[…]

### 2025-01-16 doc/manuscript.Rmd
couple of conceptual figs, bib and RMD files

@@
 ## Figures
-fig_DAG=1
+fig_DAG <- 1
 
 # Read in files
-
 ```
@@
 set_flextable_defaults(font.family = "Arial", font.size = 9)
-team<- read.csv("display/authors.csv", stringsAsFactors = FALSE, check.names = FALSE)
+team <- read.csv("display/authors.csv", stringsAsFactors = FALSE, check.names = FALSE)
 
@@
 ft
-
 ```
@@
 
-# Abstract {-}
+# Summary {-}
+
+- rationale  
+- Methods  
+- key Results   
+- main conclusion
 
@@
 
-# Methods {-}
+# Materials and Methods {-}
 
@@
 
+# Acknowledgements {-}
+
+# Competing interests {-}
+
+# Author contributions {-}
+
+# Data availability
+
 \newpage

### 2025-01-21 doc/manuscript.Rmd
updated ms, fig 1

@@
 
-**Title:** Vulnerability of tree growth to precipitation in Thailand
+**Title:** Sensitivity of tree growth to drought in seasonally dry tropical forests using long-term dendrometer band measurements
 
@@
 
+![](display/Fig1.png)
+
 Here we use a 14-year record of dendrometer band measurements in seasonally dry forest in Thailand to test the vulnerability of tropical tree growth to drought.
@@
 
-- **habitat** : water availability is a key driver of tropical tree growth [@wagner_water_2012]. 
+<!-- - **habitat** : water availability is a key driver of tropical tree growth [@wagner_water_2012]. 
 - **leaf habit** : drought resistance is higher in deciduous species because of leaf strategies that minimise water loss during months of high vulnerability.
@@
 - **leaf habit x habitat** : tropical evergreen and deciduous species have different habitat preferences [@kunert_leaf_2021] that could affect their drought resistance.   
-- **habitat x exposure** : trees with high exposure in upland habitats are expected to have lowest drought resistance because of the compounding influence of abiotic stressors.
+- **habitat x exposure** : trees with high exposure in upland habitats are expected to have lowest drought resistance because of the compounding influence of abiotic stressors.   -->
 
@@
 
-ForestGEO data [@anderson-teixeira_ctfsforestgeo_2015]
-
 ***Sites and data***
 
-*Huai Kha Khaeng plot*
-- location
-- annual precipitation
-- ecosystem type/association
-- long-term trends from @vlam et al etc
+We analysed data from manual dendrometer band censuses from 2008-2023 in the Huai Kha Khaeng ForestGEO plot (lat/long).
+This is a 50-hectate hectare plot in mixed deciduous tropical forest, with a mean annual temperature of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
+The dominant canopy species are _Hopea odorata_, _Vatica harmandiana_, _Dipterocarpus alatus_ and _Tetrameles nudiflora_. 
+Long-term analysis of tree-ring data shows that tree growth at Huai Kha Khaeng has been influenced by temperature and rainfall [@vlam_temperature_2014].
 
-*Resistance*
-- Dendrometer band measurements - census protocol
-- Calculated DBH at each instance from first DBH and window size using Condit's functions.
-- Removed measures with large decimal errors, potential band misID, large outliers (> 3 SD). 
-- calculated increments for each individual for each census year using dendroband data and tape measurement
-- tape measurement does not have accuracy required to measure this effect, but used tape increments to detect outliers.
-- removed all increment measurements with more than 0.5 cm deviation from the 1:1 line for dendro and tape increments 
-- For each drought year, used drought-year increment.
-- for each tree, long-term growth was calculated as the mean growth across 20 years of full-plot censuses
-- Resistance - ratio of drought-year increment to long-term growth for each individual tree
+We used the late wet/early dry dendrometer band censuses to calculate annual growth for each tree. 
+We used the raw window size measurements combined with the diameter at breast height (DBH) measurement at installation to calculate DBH at each census using standard equations [condit].
+We removed measures with large decimal errors, potential misidentified bands (old band numbers that restarted after stopping) and large measurement outliers (> 3 standard deviations from the mean). 
+We then calculated annualised increments for each individual for each year by finding the difference in increment from the previous year and adjusting for the number of days between measurements.
+We repeated the same steps for annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement with these tape increments.  
+Finally, we excluded trees with negative increments over the whole timeseries, and included only species that had at least 10 individuals, resulting in a final dataset of 1820 individuals across 30 species.
+
+*Variables*
+
+We calculated the sensitivity of each tree to each drought as a growth anomaly from its mean annual increment.
+We first calculated the mean annual increment for each tree as the mean of the annualised increments across the whole timeseries.
+We then calculated sensitivity for each tree for each drought as 
+$ sensitivity_{i} \sim \frac{drought.year.increment_{i} - mean.annual.increment_{i}}{mean.annual.increment_{i}}$
+
+*Species-level traits*
+
+*Individual variables*
 

### 2025-01-21 scripts/explore_analysis.Rmd
cleanup commits
updated ms, fig 1

@@
 
-load("growth-precip-thailand/data/HKK-dendro/DendroByBand.RData")
+load("data/HKK-dendro/DendroByBand.RData")
 
@@
 #chirps_hkk<-read.csv("../data/climate/CHIRPS_from_GEE.csv")
-chirps_hkk<-read.csv("growth-precip-thailand/data/climate/CHIRPS_from_GEE.csv")
+chirps_hkk<-read.csv("data/climate/CHIRPS_from_GEE.csv")
 colnames(chirps_hkk)[1]<-"time"
@@
 
-spei_hkk<-read.csv("growth-precip-thailand/data/climate/SPEI_HKK_from_GEE.csv")
+spei_hkk<-read.csv("data/climate/SPEI_HKK_from_GEE.csv")
 colnames(spei_hkk)[1]<-"time"
@@
 ```
+```{r forest fire station, echo=F}
+
+rainfall_hkk<-read.csv("data/climate/WeatherData_ALL_forest_fire_research_station.csv")
+head(rainfall_hkk)
+max(rainfall_hkk$YearII)
+
+```
 ```{r echo=F}
@@
 
-load("growth-precip-thailand/data/HKK-dendro/stem1.RData")
-load("growth-precip-thailand/data/HKK-dendro/stem2.RData")
-load("growth-precip-thailand/data/HKK-dendro/stem3.RData")
-load("growth-precip-thailand/data/HKK-dendro/stem4.RData")
+load("data/HKK-dendro/stem1.RData")
+load("data/HKK-dendro/stem2.RData")
+load("data/HKK-dendro/stem3.RData")
+load("data/HKK-dendro/stem4.RData")
 
@@
 
-```{r plot increments old code, echo=F, eval=F}
-
-dendro_2010<- dendro.data.noouts %>%
-  filter(Cno %in% c(1, 3, 5))%>%
-  group_by(Tag) %>% #grouping by treeID to calculate increment
-  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
-         cii_min1=dplyr::lag(Cii, n=1, default=NA, order_by=Tag),
-         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
-         inc=calcDBH-calcDBH_min1,
-         inc_annual=(inc/as.numeric(tdif))*365,
-         inc.time=Cno-lag(Cno),
-         #removing increment value if the previous measure was before the last year
-         inc_annual=ifelse(inc.time==2, inc_annual, NA),
-         treeID=paste0("HKK.", Tag))%>%
-  filter(!is.na(inc))%>%
-  filter(!is.na(inc_annual))%>%
-  ungroup()%>%
-  dplyr::mutate(
-    diainc.scaled = scale(inc_annual),
-    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))%>%
-  filter(out.inc==0)
-
-quantiles_2009<-quantile(dendro_2010$inc_annual[dendro_2010$Cno==3], 
-                         probs=c(0.25, 0.5, 0.75, 1))
-
-quantiles_2010<-quantile(dendro_2010$inc_annual[dendro_2010$Cno==5], 
-                         probs=c(0.25, 0.5, 0.75, 1))
-
-dendro_2010_plot<-ggplot(data = dendro_2010, aes(x=inc_annual, group=factor(Cno), fill=factor(Cno), col=factor(Cno))) +
-  geom_density(alpha=0.5) +
-  scale_color_manual(values=c("forestgreen", "brown"))+
-  scale_fill_manual(values=c("forestgreen", "brown"))+
-  geom_vline(xintercept=quantiles_2009, col="forestgreen")+
-  geom_text(aes(x=0.75, y=1, label="2009"), col="forestgreen", size=3)+
-  geom_text(aes(x=0.75, y=1.5, label="2010"), col="brown", size=3)+
-  geom_vline(xintercept=quantiles_2010, col="brown")+ 
-  xlab("diameter increment (cm)") + ylab("density")+
-  guides(col="none", fill="none")+
-  theme_bw()
-
-dendro_2015<- dendro.data.noouts %>%
-  filter(Cno %in% c(11, 13, 15))%>%
-  group_by(Tag) %>% #grouping by treeID to calculate increment
-  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
-         cii_min1=dplyr::lag(Cii, n=1, default=NA, order_by=Tag),
-         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
-         inc=calcDBH-calcDBH_min1,
-         inc_annual=(inc/as.numeric(tdif))*365,
-         inc.time=Cno-lag(Cno),
-         #removing increment value if the previous measure was before the last year
-         #inc=ifelse(inc.time==1, inc, NA),
-         inc_annual=ifelse(inc.time==2, inc_annual, NA),
-         treeID=paste0("HKK.", Tag))%>%
-  filter(!is.na(inc))%>%
-  filter(!is.na(inc_annual))%>%
-  ungroup()%>%
-  dplyr::mutate(
-    diainc.scaled = scale(inc_annual),
-    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))%>%
-  filter(out.inc==0)
-
-quantiles_2014<-quantile(dendro_2015$inc_annual[dendro_2015$Cno==13], 
-                         probs=c(0.25, 0.5, 0.75, 1))
-
-quantiles_2015<-quantile(dendro_2015$inc_annual[dendro_2015$Cno==15], 
-                         probs=c(0.25, 0.5, 0.75, 1))
-
-dendro_2015_plot<-ggplot(data = dendro_2015, aes(x=inc_annual, group=factor(Cno), fill=factor(Cno), col=factor(Cno))) +
-  geom_density(alpha=0.5) +
-  scale_color_manual(values=c("forestgreen", "brown"))+
-  scale_fill_manual(values=c("forestgreen", "brown"))+
-  geom_vline(xintercept=quantiles_2014, col="forestgreen")+
-  geom_text(aes(x=0.75, y=1, label="2014"), col="forestgreen", size=3)+
-  geom_text(aes(x=0.75, y=1.5, label="2015"), col="brown", size=3)+
-  geom_vline(xintercept=quantiles_2015, col="brown")+ 
-  xlab("diameter increment (cm)") + ylab("density")+
-  guides(col="none", fill="none")+
-  theme_bw()
-
-dendro_2019<- dendro.data.noouts %>%
-  filter(Cno %in% c(19, 21, 23))%>%
-  group_by(Tag) %>% #grouping by treeID to calculate increment
-  mutate(calcDBH_min1=dplyr::lag(calcDBH, n=1, default=NA, order_by=Tag),
-         cii_min1=dplyr::lag(Cii, n=1, default=NA, order_by=Tag),
-         tdif=difftime(Date2, lag(Date2, n=1, default=NA, order_by=Tag), units = "days"),
-         inc=calcDBH-calcDBH_min1,
-         inc_annual=(inc/as.numeric(tdif))*365,
-         inc.time=Cno-lag(Cno),
-         #removing increment value if the previous measure was before the last year
-         #inc=ifelse(inc.time==1, inc, NA),
-         inc_annual=ifelse(inc.time==2, inc_annual, NA),
-         treeID=paste0("HKK.", Tag))%>%
-  filter(!is.na(inc))%>%
-  filter(!is.na(inc_annual))%>%
-  ungroup()%>%
-  dplyr::mutate(
-    diainc.scaled = scale(inc_annual),
-    out.inc = ifelse(abs(diainc.scaled)>3, 1, 0))%>%
-  filter(out.inc==0)
-
-quantiles_2018<-quantile(dendro_2019$inc_annual[dendro_2019$Cno==21], 
-                         probs=c(0.25, 0.5, 0.75, 1))
-
-quantiles_2019<-quantile(dendro_2019$inc_annual[dendro_2019$Cno==23], 
-                         probs=c(0.25, 0.5, 0.75, 1))
-
-dendro_2019_plot<-ggplot(data = dendro_2019, aes(x=inc_annual, group=factor(Cno), fill=factor(Cno), col=factor(Cno))) +
-  geom_density(alpha=0.5) +
-  scale_color_manual(values=c("forestgreen", "brown"))+
-  scale_fill_manual(values=c("forestgreen", "brown"))+
-  geom_vline(xintercept=quantiles_2018, col="forestgreen")+
-  geom_text(aes(x=0.75, y=1, label="2018"), col="forestgreen", size=3)+
-  geom_text(aes(x=0.75, y=1.5, label="2019"), col="brown", size=3)+
-  geom_vline(xintercept=quantiles_2019, col="brown")+ 
-  xlab("diameter increment (cm)") + ylab("density")+
-  guides(col="none", fill="none")+
-  theme_bw()
-
-library(gridExtra)
-png("growth-precip-thailand/doc/display/increments_raw.png", width=8, height=4, units="in", res=300)
-#grid.arrange(dendro_2010_plot, dendro_2015_plot, dendro_2019_plot, nrow=1)
-grid.arrange(dendro_2010_plot, dendro_2015_plot, nrow=1)
-dev.off()
-
-```
-
 ```{r top4 species drought resistance, echo=F}
@@
 
+```{r top 4 resistance and exposure, echo=F}
+
+#exposure vs resistance plot
+
+resistance_exposure<-ggplot(resistance_top4, aes(x=cii_min1, y=resistance.div, 
+col=factor(sp, levels=c("HOPEOD", "TETRNU", "POLYVI", "SACCLI"))))+
+  geom_boxplot()+
+  geom_smooth(method="lm", se=F)+
+  facet_grid(sp~yr)+
+  scale_color_viridis_d()+
+  xlab("exposure")+ylab("resistance")+theme_bw()
+
+resistance_exposure
+
+```
+
 ```{r associated datasets, echo=F}
@@
 #deciduousness
-dec_williams<-read.csv("growth-precip-thailand/data/HKK-dendro/species.csv")
+dec_williams<-read.csv("data/HKK-dendro/species.csv")
 
@@
 #habitat data
-habitat<-read.csv("growth-precip-thailand/data/HKK-other/HKK_habtype(in).csv")
+habitat<-read.csv("data/HKK-other/HKK_habtype(in).csv")
 
@@
 trees<-merge(trees, select(habitat, -c(X, x, y)), by=c("Col", "Row"), all.x=T)
-
+```
+```{r TWI from elev, echo=F}
 #elevation
@@
 
-hkk_elev<-read.table("growth-precip-thailand/data/HKK-other/CTFSElev_hkk.rdata")
+hkk_elev<-read.table("data/HKK-other/CTFSElev_hkk.rdata")
+
 png("growth-precip-thailand/doc/display/elevation_resistance_map_2015.png", width=8, height=4, units="in", res=300)
@@
 
+#following code from https://vt-hydroinformatics.github.io/rgeoraster.html
+
 library(sf)
 library(terra)
-
-#first convert to raster
-elev_rast<-rast(hkk_elev$col.elev, nrows=100, ncols=100, xmin=0, xmax=100, ymin=0, ymax=100)
+library(tidyverse)
+library(raster)
+library(tmap)
+
+library(whitebox)
+#whitebox::install_whitebox()
+whitebox::wbt_init()
+
+theme_set(theme_classic())
+
+#convert to raster
+
+#first convert hkk_elev$col.elev to matrix
+elev_mat<-matrix(hkk_elev$col.elev, nrow=201, ncol=101, byrow=T)
+
+#then convert to raster
+elev_rast<-raster(elev_mat, xmn=0, xmx=50, ymn=0, ymx=100,
+crs="+proj=utm +zone=47 +datum=WGS84")
+plot(elev_rast)
+
+writeRaster(elev_rast, "data/HKK-other/HKKDEMsm_5m.tif", overwrite = TRUE)
+
+tm_shape(elev_rast)+
+  tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE)+
+  tm_scale_bar()
+
+#hillshade
+
+wbt_hillshade(dem = "data/HKK-other/HKKDEMsm_5m.tif",
+              output = "data/HKK-other/HKK_hillshade.tif",
+              azimuth = 115)
+
+hillshade <- raster("data/HKK-other/HKK_hillshade.tif")
+
+tm_shape(hillshade)+
+  tm_raster(style = "cont", palette = "-Greys", legend.show = FALSE)+
+  tm_scale_bar()
+
+wbt_breach_depressions_least_cost(
+  dem = "data/HKK-other/HKKDEMsm_5m.tif",
+  output = "data/HKK-other/HKKDEMsm_breached.tif",
+  dist = 5,
+  fill = TRUE)
+
+wbt_fill_depressions_wang_and_liu(
+  dem = "data/HKK-other/HKKDEMsm_breached.tif",
+  output = "data/HKK-other/HKKDEMsm_filled_breached.tif"
+)
+
+filled_breached <- raster("data/HKK-other/HKKDEMsm_filled_breached.tif")
+
+## What did this do?
+difference <- elev_rast - filled_breached
+
+difference[difference == 0] <- NA
+
+tm_shape(hillshade)+
+  tm_raster(style = "cont",palette = "-Greys", legend.show = FALSE)+
+  tm_scale_bar()+
+tm_shape(difference)+
+  tm_raster(style = "cont",legend.show = TRUE)+
+  tm_scale_bar()
+
+#TWI
+
+wbt_d_inf_flow_accumulation(input = "data/HKK-other/HKKDEMsm_filled_breached.tif",
+                            output = "data/HKK-other/HKK_DinfFAsca.tif",
+                            out_type = "Specific Contributing Area")
+
+wbt_slope(dem = "data/HKK-other/HKKDEMsm_filled_breached.tif",
+          output = "data/HKK-other/demslope.tif",
+          units = "degrees")
+
+wbt_wetness_index(sca = "data/HKK-other/HKK_DinfFAsca.tif",
+                  slope = "data/HKK-other/demslope.tif",
+                  output = "data/HKK-other/TWI.tif")
+
+library(raster)
+twi <- raster("data/HKK-other/TWI.tif")
+
+#twi[twi > 0] <- NA
+
+tm_shape(hillshade)+
+  tm_raster(style = "cont",palette = "-Greys", legend.show = FALSE)+
+tm_shape(twi)+
+  tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE, alpha = 0.5)+
+  tm_scale_bar()
+```
+
+```{r crowding variables, echo=F}
+
+#find basal area around each tree
+
+#read tree locations
+load("data/HKK-dendro/stem1.RData")
+load("data/HKK-dendro/stem2.RData")
+load("data/HKK-dendro/stem3.RData")
+load("data/HKK-dendro/stem4.RData")
+
+hkk_census<-bind_rows(hkk.stem1, hkk.stem2, hkk.stem3, hkk.stem4)
+
+#for each tree in each census, find the basal area of trees within 5, 10, 15 and 20m
+library(sf)
+#first convert to sf object
+hkk.stem1_sf<-st_as_sf(hkk.stem1, coords=c("gx", "gy"))
+hkk.stem2_sf<-st_as_sf(hkk.stem2, coords=c("gx", "gy"))
+hkk.stem3_sf<-st_as_sf(hkk.stem3, coords=c("gx", "gy"))
+hkk.stem4_sf<-st_as_sf(hkk.stem4, coords=c("gx", "gy"))
+
+#then find the distance between each tree and all other trees
+hkk.stem1_dist<-st_distance(hkk.stem1_sf)
+
+hkk_census <- hkk_census %>%
+  dplyr::mutate(
+    basal_area_5m = NA,
+    basal_area_10m = NA,
+    basal_area_15m = NA,
+    basal_area_20m = NA
+  )
 ```

### 2025-01-21 scripts/06_figures.R

@@ -10,16 +10,27 @@ library(lubridate)
 spei <- read.csv("data/climate/SPEI_HKK_from_GEE.csv")
 ffstation <- read.csv("data/climate/WeatherData_ALL_forest_fire_research_station.csv")
 
+enso <- read.table("data/climate/ENSO_index_meiv2.data", header = F, skip = 1)
+
+# calculate vpd using ffstation data
+ffstation <- ffstation %>%
+    mutate(
+        vpdmax = 0.6108 * exp(17.27 * TempMax / (TempMax + 237.3)) * (1 - (Relative.Humidity / 100)),
+        vpdmin = 0.6108 * exp(17.27 * TempMin / (TempMin + 237.3)) * (1 - (Relative.Humidity / 100))
+    )
+
 ffstation_monthly <- ffstation %>%
     group_by(YearII, Month) %>%
     dplyr::summarise(
         precipitation = sum(Precipitation),
         dry_days = sum(Precipitation == 0),
         tmax = max(TempMax),
+        vpdmax = mean(vpdmax),
+        vpdmin = mean(vpdmin)
     ) %>%
     rename(year = YearII, month = Month) %>%
     dplyr::mutate(year = as.character(year)) %>%
-    select(year, month, precipitation, dry_days, tmax)
+    select(year, month, precipitation, dry_days, tmax, vpdmax, vpdmin)
 
 ffstation_lt <- ffstation_monthly %>%
     filter(year <= 2021, year >= 2009) %>%
@@ -29,9 +40,11 @@ ffstation_lt <- ffstation_monthly %>%
         precipitation = mean(precipitation, na.rm = T),
         dry_days = mean(dry_days, na.rm = T),
         tmax = mean(tmax, na.rm = T),
+        vpdmax = mean(vpdmax, na.rm = T),
+        vpdmin = mean(vpdmin, na.rm = T),
         year = "long-term"
     ) %>%
-    select(year, month, precipitation, dry_days, tmax)
+    select(year, month, precipitation, dry_days, tmax, vpdmax, vpdmin)
 
 # bind these two dataframes
 ffstation_full <- bind_rows(ffstation_monthly, ffstation_lt) %>%
@@ -70,11 +83,11 @@ clim <- merge(ffstation_full, spei_full, by = c("year", "month"))
 
 clim <- clim %>%
     pivot_longer(
-        cols = c(precipitation, dry_days, tmax, spei_val),
[…]

### 2025-01-22 doc/manuscript.Rmd
manuscript updates
@teixeirak - added draft abstract, site and data methods, fig 1. Intro, results are unchanged

@@
 **Authors**  
-```{r eval = TRUE, echo=FALSE, warning=FALSE, eval=F}
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
 library(flextable)
@@
 # Summary {-}
+<!-- 
+- Rationale  
+- Methods  
+- Key results   
+- Main conclusion   -->
 
-- rationale  
-- Methods  
-- key Results   
-- main conclusion
+- Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but how climate anomalies might impact the growth of tropical tree growth is poorly understood. 
+Standard methods like tree rings and plot inventories may either not represent most species or may not be accurate enough to capture variation in slow-growing species.  
+- We used a unique data record - manual dendrometer band measurements over 15 years on 1820 trees across 30 species in a seasonally dry tropical forest in Huai Kha Khaeng, Thailand, along with species traits (deciduousness and maximum size), individual characteristics (crown exposure, size) and environment-related water availability to predict the sensitivity of tree growth to drought in two ENSO drought events.
+We used Bayesian regressions in a causal modelling framework to predict variability among species and individuals in drought sensitivity.
+- In the more severe drought, deciduous species were more sensitive than evergreen species. 
+After accounting for species differences, more exposed trees were more likely to be negatively impacted than their less exposed counterparts, with exposure showing clear negative impacts compared to size *per se*.
+- Large differences among species and individuals in their sensitivity to drought, as well as climatic differences differences among droughts themselves can explain heterogeneity in ecological responses to drought.
+
+# Keywords {-}
 
@@
 
-Tropical tree woody growth is a key component of aboveground productivity and affects the global carbon carbon cycle, but its sensitivity to ongoing global change is poorly understood.  
+Tropical tree woody growth is a key component of aboveground productivity and affects the global carbon carbon cycle, but its sensitivity to climate anomalies is poorly understood.  
 - Woody growth has long-lasting effects on the terrestrial carbon pool but is sensitive to interannual variation in temperature and precipitation.  
@@
 
-*Variables*
-
 We calculated the sensitivity of each tree to each drought as a growth anomaly from its mean annual increment.
@@
 We then calculated sensitivity for each tree for each drought as 
-$ sensitivity_{i} \sim \frac{drought.year.increment_{i} - mean.annual.increment_{i}}{mean.annual.increment_{i}}$
+$$
+sensitivity_{i} = \frac{drought.year.increment_{i} - mean.annual.increment_{i}}{mean.annual.increment_{i}}
+$$
 
-*Species-level traits*
+We calculated two species-level traits : deciduousness and maximum size and one variable for each tree - topographic wetness index and two variables for each tree at each census - starting size and canopy exposure.
+We defined deciduousness as the mean proportion of canopy loss at maximum loss across individuals in the species, with a score ranging from 1 to 4 where 1 is most evergreen and 4 is most deciduous. 
+We used data published in @williams_deciduousness_2008 for the top 30 species. 
+Out of these, data was unavailable for Alphonsea ve...., an evergreen species, and so we manually assigned it a deciduousness score of 1, in the middle of the evergreen range. 
+An alternate measure of deciduousness is the duration spent at maximum crown loss.
+This data was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for a larger list of species. 
+At the species level, we also calculated the maximum size across all individuals in the plot as measured in the 2008 whole-plot census.
 
-*Individual variables*
+Across all individuals in the plot, we calculated topographic wetness index based on a 5 m x 5 m elevation layer collected at plot establishment.
+Topographic wetness index provides an expectation of the water availability across space based on the concavity/convexity of a location. 
+To model sensitivity among individuals, we also used tree size (DBH at previous dendroband census) and crown exposure measured as crown illumination index.
+Crown illumination index is an ordinal categorical variable from 1 to 5 (least to most exposed) measured at each dendroband census.
 
-*Tree size*
-- calculated DBH for at the start of the drought year from dendrometer band window
-
-*Deciduousness*
+<!-- *Deciduousness*
 -data from @williams_deciduousness_2008. 
@@
 -One value per species. 
-- 7 species out of the top 51 species in the dendroband census have no values. (currently excluded from analysis but can fill these with best estimates and add them back in)
+- 7 species out of the top 51 species in the dendroband census have no values. (currently excluded from analysis but can fill these with best estimates and add them back in) -->
 
-*Exposure*
+<!-- *Exposure*
 - crown illumination index for each individual, collected at each dendrometer band census
@@
 - calculated from topography
-- estimate of water availability at a location
+- estimate of water availability at a location -->
 
-***Models***
+***Statistical methods***
 
-For individual stems, we modelled drought resistance as:
+For individual stems, we modelled drought sensitivity as:
 
-$$
+<!-- $$
 resistance \sim size + exposure + deciduousness + water + size*exposure + deciduousness*exposure + size*deciduousness + (1|drought.year)
-$$
+$$ -->
 
@@
 
-# Data availability
+# Data availability {-}
 

### 2025-01-22 doc/manuscript.Rmd
draft fig 2 #11 + in the manuscript

@@
 
-![](display/Fig1.png)
+![***Figure 1.
+Climatic characteristics of drought years and average years*** in Huai Kha Khaeng ForestGEO plot.
+For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1.png)
 
@@
 Finally, we excluded trees with negative increments over the whole timeseries, and included only species that had at least 10 individuals, resulting in a final dataset of 1820 individuals across 30 species.
+
+![***Figure 2: Growth increment timeseries and drought sensitivity*** across trees and species in the Huai Kha Khaeng ForestGEO plot.
+a) Median of annualised growth increments across species (colours) and the whole plot (black) from 2009 - 2022 from late wet/early dry season dendroband measurements.
+2010 and 2015 were moderate and very strong ENSO events respectively.
+b) Dsitribution of drought sensitivity of tree growth across all individuals in the two drought years.
+For comments - https://github.com/forestgeo/growth-precip-thailand/issues/11](display/Fig2.png)
 

### 2025-01-22 scripts/06_figures.R

@@ -123,3 +123,137 @@ climplot
 png("doc/display/Fig1.png", width = 8, height = 4, units = "in", res = 300)
 climplot
 dev.off()
+
+# figure 2 - growth increments ENSO plot + sensitivity raw distributions
+
+# Load required libraries---------------------
+library(tidyverse)
+
+# load data--------------------------
+datasets <- readRDS("data/HKK-dendro/sensitivity_data.RData")
+
+tree.time <- datasets$tree.time
+tree_vars <- datasets$tree_vars
+sp_vars <- datasets$sp_vars
+
+colnames(sp_vars)[1] <- "Species"
+
+# merge tree_vars and sp_vars
+tree_vars <- merge(tree_vars, sp_vars, by = "Species", all.x = TRUE)
+# merge tree_vars and tree.time
+tree.time <- merge(tree.time, tree_vars, by = "Tag", all.x = TRUE)
+
+colnames(tree.time)
+
+# standardise the variables within species and across all--------------------
+
+tree.time <- tree.time %>%
+    dplyr::mutate(
+        calcDBH_min1_scaled = scale(calcDBH_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled = scale(cii_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled = ifelse(cii_min1_scaled == "NaN", 0, cii_min1_scaled),
+        twi_scaled = scale(twi, center = TRUE, scale = TRUE),
+        median_inc_scaled = scale(median_inc, center = TRUE, scale = TRUE),
+        avg_inc_tree_scaled = scale(avg_inc_tree, center = TRUE, scale = TRUE)
+    ) %>%
+    group_by(Species) %>%
+    # scale while retaining original values
+    dplyr::mutate(
+        calcDBH_min1_scaled_sp = scale(calcDBH_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled_sp = scale(cii_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled_sp = ifelse(cii_min1_scaled_sp == "NaN", 0, cii_min1_scaled_sp),
+        twi_scaled_sp = scale(twi, center = TRUE, scale = TRUE)
+    ) %>%
+    ungroup() %>%
+    # remove large outliers for each year
+    group_by(yr) %>%
+    # find sens.prop values that are 3 sds from the mean
[…]

### 2025-01-22 scripts/06_figures.R
draft with plot labels

@@ -254,6 +254,7 @@ spagplot_top10 <- ggplot() +
 # dev.off()
 
 library(gridExtra)
+library(patchwork)
 png("doc/display/Fig2.png", width = 4, height = 6, units = "in", res = 300)
-grid.arrange(spagplot_top10, sens.all, nrow = 2, heights = c(1.2, 1))
+spagplot_top10 / sens.all + plot_annotation(tag_levels = "a") + plot_layout(heights = c(1.2, 1))
 dev.off()

### 2025-01-27 scripts/01_prepdata.R
plots for TWI distributions

@@ -304,7 +304,7 @@ trees <- trees %>% filter(Tag %in% tree.time$Tag)
 
 # twi values
 library(raster)
-twi <- raster("data/HKK-other/TWI.tif")
+twi <- raster::raster("data/HKK-other/TWI.tif")
 
 # pick twi values for each tree
 trees$twi <- raster::extract(twi, 0.1 * trees[, c("Y", "X")]) # multiply by 0.1 to get the correct resolution
@@ -392,6 +392,28 @@ ggplot() +
   coord_fixed()
 dev.off()
 
+# plot distribution of twi across top 4 species
+png("doc/display/explore/twi_dist_top4.png", width = 5, height = 5, units = "in", res = 300)
+ggplot(hkk.stem4 %>% filter(sp %in% c("SACCLI", "HOPEOD", "POLYVI", "TETRNU")), aes(x = twi, fill = sp)) +
+  geom_density(alpha = 0.5) +
+  scale_fill_viridis_d() +
+  theme_minimal() +
+  labs(title = "Distribution of top 4 species in TWI space (HKK)")
+dev.off()
+
+# plot distribution as violin plot for top 10 species
+dec_william_sort <- dec_williams %>%
+  arrange(williams_dec)
+
+png("doc/display/explore/twi_violin_top10.png", width = 5, height = 5, units = "in", res = 300)
+ggplot(hkk.stem4 %>% filter(sp %in% dec_william_sort$sp), aes(x = factor(sp, levels = dec_william_sort$sp), y = twi, fill = sp)) +
+  geom_violin() +
+  geom_boxplot(width = 0.1) +
+  scale_fill_viridis_d() +
+  theme_minimal() +
+  labs(title = "Distribution of top 10 species in TWI space (HKK)")
+dev.off()
+
 sp.traits <- hkk.stem4 %>%
   filter(status == "A") %>%
   filter(sp %in% dec_williams$sp) %>%
@@ -412,6 +434,7 @@ sp.traits <- hkk.stem4 %>%
     hab6 = ifelse(is.na(sum(hab6, na.rm = T)) == T, 0, 1),
     # twi mean
     twi_mean = mean(twi, na.rm = T),
+    twi_median = median(twi, na.rm = T),
     # twi sd
     twi_sd = sd(twi, na.rm = T)
   )
@@ -423,6 +446,56 @@ sp_vars <- merge(sp.traits, dplyr::select(dec_williams, c("sp", "williams_dec"))
 
[…]

### 2025-01-27 doc/supplementary_information.Rmd
supplementary information update

@@
 suppress-bibliography: false
-csl: methods-in-ecology-and-evolution.csl
+csl: new-phytologist.csl
 header-includes:  \usepackage{float}
@@
 
-#Appendices
+# Appendices
 
-#SI table numbers
+# SI table numbers
 
-#SI table captions
+# SI table captions
 
-#SI figure numbers
+# SI figure numbers
 
-#SI figure captions
+# SI figure captions
 
-#https://github.com/EcoClimLab/Operations/blob/master/Data_Management-Scientific_Workflow/manuscript_prep_with_Rmd.md
+# https://github.com/EcoClimLab/Operations/blob/master/Data_Management-Scientific_Workflow/manuscript_prep_with_Rmd.md
 ```
@@
 # Read in files
+```
 
-```
 \newpage
+
+## Climate data and correlations
+
+![](../results/plots/climate_data_timeseries_2007_2019.png)
+
+![](../results/plots/climate_data_correlations.png)
+
+\newpage
+
+## Data cleaning additional methods
+
+\newpage
+
+## Species traits and correlations
+
+\newpage
+
+## Individual model diagnostics
+
+\newpage

### 2025-01-27 scripts/02_models.R
updated species traits plot with TWI

@@ -341,10 +341,11 @@ ranef_df <- merge(ranef_df, sp_vars, by = "Species", all.x = TRUE)
 
 # make long df for plotting
 ranef_df_long <- ranef_df %>%
-    pivot_longer(c("median_inc", "maxDBH", "williams_dec")) %>%
+    pivot_longer(c("twi_sd", "maxDBH", "williams_dec")) %>%
     # rename the variables
     dplyr::mutate(name = case_when(
-        name == "median_inc" ~ "median increment",
+        # name == "twi_median" ~ "median TWI",
+        name == "twi_sd" ~ "sd(TWI)",
         name == "maxDBH" ~ "maximum DBH",
         name == "williams_dec" ~ "deciduousness"
     ))

### 2025-01-27 scripts/06_figures.R
draft figure 3 #12
@teixeirak

@@ -189,7 +189,7 @@ sens.all <- ggplot(data = tree.time %>% filter(yr %in% yrs), aes(x = sens.prop))
 # sens.all
 # dev.off()
 
-# run the model for the top 10 species for 2015
+# get the top 10 species for 2015
 
 top_10_sp <- tree.time %>%
     filter(Cno == 15) %>%
@@ -257,4 +257,82 @@ library(gridExtra)
 library(patchwork)
 png("doc/display/Fig2.png", width = 4, height = 6, units = "in", res = 300)
 spagplot_top10 / sens.all + plot_annotation(tag_levels = "a") + plot_layout(heights = c(1.2, 1))
+dev.off()
+
+# figure 3 -
+
+# related issue - https://github.com/forestgeo/growth-precip-thailand/issues/12
+
+# order of species abundance
+tree.time$Species <- factor(tree.time$Species, levels = table(tree.time$Species) %>% sort(decreasing = T) %>% names())
+
+# plot of median sensitivities across species for 2010 and 2015
+sensplot <- ggplot(data = tree.time %>% filter(yr %in% c(2010, 2015)), aes(x = Species, y = sens.prop, color = factor(yr))) +
+    geom_point(position = position_jitterdodge(), alpha = 0.5) +
+    geom_boxplot(fill = NA) +
+    # scale_color_viridis_d() +
+    scale_color_manual(values = c("2010" = "indianred2", "2015" = "indianred4")) +
+    labs(
+        # title = "Distribution of drought sensitivities \nfor all species",
+        x = "Species", y = "Sensitivity"
+    ) +
+    theme_bw() +
+    guides(color = guide_legend(title = "Year")) +
+    theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+# coefs plot
+
+coefs_df <- readRDS("results/models/non_negative/sensitivity_model_intercept.RData")
+pred_df <- readRDS("results/models/non_negative/predictions_intercept.RData")
+
+# plot random effects
+ranef_df <- coefs_df %>% filter(grepl("r_Species", param))
+ranef_df$Species <- gsub("r_Species\\[|\\,Intercept\\]", "", ranef_df$param)
+
+# add intercept to ranef_df
+# make intercept vector
+intercepts <- coefs_df %>%
+    filter(param %in% "Intercept") %>%
+    dplyr::select(median) %>%
+    pull(median)
[…]

### 2025-01-30 scripts/000_twi.R
small updates to scripts

@@ -124,6 +124,8 @@ tm_shape(twi) +
   tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE, alpha = 0.5) +
   tm_scale_bar()
 
+png("doc/display/explore/twi.png", width = 4, height = 8, units = "in", res = 300)
 tm_shape(twi) +
-  tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE) +
+  tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE, alpha=1) +
   tm_scale_bar()
+dev.off()

### 2025-01-30 scripts/06_figures.R

@@ -181,11 +181,11 @@ climplot <- ggplot() +
     # theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
     # make every month show up on the x-axis
     # scale_x_continuous(breaks = 1:12) +
-    # add rectangles for dry season
-    # geom_rect(
-    #     data = data.frame(xmin = c(11, 1), xmax = c(12, 4), ymin = -Inf, ymax = Inf),
-    #     aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), fill = "grey", alpha = 0.3
-    # ) +
+    #add rectangles for dry season
+    geom_rect(
+        data = data.frame(xmin = c(305, 1), xmax = c(366, 120), ymin = -Inf, ymax = Inf),
+        aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), fill = "grey", alpha = 0.3
+    ) +
     guides(linetype = "none") +
     labs(x = "Day of year", y = "Value", color = "Year") +
     scale_color_manual(values = c("long-term" = "grey40", "2010" = "indianred2", "2015" = "indianred4"))

### 2025-01-30 doc/manuscript.Rmd
some more updates to methods

@@
 <!-- 
+New Phytologist Summary format
 - Rationale  
@@
 We expect variation in drought sensitivity (growth in drought year/growth in a previous year with normal rainfall) across trees.
-We ask: i) How much do species vary in drought sensitivity?
-ii) How does deciduousness influence the drought sensitivity of a species?
-iii) How much of the intraspecific variation is explained by tree height *per se*, crown exposure and the habitat of the tree (proxy for water and nutrient availability).
-We hypothesise that growth during drought is affected by:
+We ask: i) How much do species vary in drought sensitivity of tree growth?
+What drives this variation ii) What drives variation among individuals in the sensitivity of growth to drought?
+We hypothesise that growth sensitivity....
 
@@
 
-We analysed data from manual dendrometer band censuses from 2008-2023 in the Huai Kha Khaeng ForestGEO plot (lat/long).
-This is a 50-hectate hectare plot in mixed deciduous tropical forest, with a mean annual temperature of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
-The dominant canopy species are _Hopea odorata_, _Vatica harmandiana_, _Dipterocarpus alatus_ and _Tetrameles nudiflora_. 
+We analysed data from manual dendrometer band censuses from 2008-2023 in the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
+This is a 50-hectare plot in mixed deciduous tropical forest, with a mean annual temperature of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
+The dominant canopy species are _Hopea odorata_, _Vatica harmandiana_, _Dipterocarpus alatus_ and _Tetrameles nudiflora_ and lower layers of the canopy are dominated by sub-canopy evergreen and deciduous species like _Saccopetalum linneatum_ and _Polyalthia viridis_.
 Long-term analysis of tree-ring data shows that tree growth at Huai Kha Khaeng has been influenced by temperature and rainfall [@vlam_temperature_2014].
@@
 We used the raw window size measurements combined with the diameter at breast height (DBH) measurement at installation to calculate DBH at each census using standard equations [condit].
-We removed measures with large decimal errors, potential misidentified bands (old band numbers that restarted after stopping) and large measurement outliers (> 3 standard deviations from the mean). 
+We removed measures with likely data entry errors because of misplaced decimals, potential misidentified bands (old band numbers that restarted after stopping) and large measurement outliers (> 3 standard deviations from the mean across all observations).
 We then calculated annualised increments for each individual for each year by finding the difference in increment from the previous year and adjusting for the number of days between measurements.
@@
 
-We calculated the sensitivity of each tree to each drought as a growth anomaly from its mean annual increment.
+Combining this growth series with climatic information, we identified two drought years of interest - 2010 and 2015 - and calculated growth sensitivity for each tree to each drought as a growth anomaly from its mean annual increment.
 We first calculated the mean annual increment for each tree as the mean of the annualised increments across the whole timeseries.
@@
 
-We calculated two species-level traits : deciduousness and maximum size and one variable for each tree - topographic wetness index and two variables for each tree at each census - starting size and canopy exposure.
-We defined deciduousness as the mean proportion of canopy loss at maximum loss across individuals in the species, with a score ranging from 1 to 4 where 1 is most evergreen and 4 is most deciduous. 
-We used data published in @williams_deciduousness_2008 for the top 30 species. 
-Out of these, data was unavailable for Alphonsea ve...., an evergreen species, and so we manually assigned it a deciduousness score of 1, in the middle of the evergreen range. 
-An alternate measure of deciduousness is the duration spent at maximum crown loss.
-This data was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for a larger list of species. 
-At the species level, we also calculated the maximum size across all individuals in the plot as measured in the 2008 whole-plot census.
-
 Across all individuals in the plot, we calculated topographic wetness index based on a 5 m x 5 m elevation layer collected at plot establishment.
@@
 
-<!-- *Deciduousness*
--data from @williams_deciduousness_2008. 
--Leaf phenology score - mean proportion of crown loss. 
--Continuous scale from 1 to 4, where 4 is maximum crown loss or deciduousness. 
--One value per species. 
-- 7 species out of the top 51 species in the dendroband census have no values. (currently excluded from analysis but can fill these with best estimates and add them back in) -->
-
-<!-- *Exposure*
-- crown illumination index for each individual, collected at each dendrometer band census
-- categorical variable with values 1 to 5 where 1 is least exposed and 5 is most exposed
-
-*Topographic Wetness Index*
-- calculated from topography
-- estimate of water availability at a location -->
+We calculated three species-level traits : deciduousness, maximum size and niche breadth and one variable for each tree - topographic wetness index and two variables for each tree at each census - starting size and canopy exposure.
+We defined deciduousness as the mean proportion of canopy loss at maximum loss across individuals in the species, with a score ranging from 1 to 4 where 1 is most evergreen and 4 is most deciduous. 
+We used data published in @williams_deciduousness_2008 for the top 30 species. 
+Out of these, data was unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1, in the middle of the evergreen range. 
+An alternate measure of deciduousness is the duration spent at maximum crown loss.
+This data was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for a larger list of species. 
+At the species level, we also calculated the maximum size across all individuals in the plot as measured in the 2008 whole-plot census.
 
@@
 
+We first modelled sensitivity as a simple random intercept model
+
+$$
+sensitivity \sim 1 + (1 | Species)
+$$
+
+We then used the predicted sensitivities from this model and tested their correlations with three species-level traits - deciduousness, maximum size and niche breadth using simple linear models.
+
 For individual stems, we modelled drought sensitivity as:
 
-<!-- $$
-resistance \sim size + exposure + deciduousness + water + size*exposure + deciduousness*exposure + size*deciduousness + (1|drought.year)
-$$ -->
+$$
+sensitivity \sim DBH + CII + TWI + size + (DBH+CII+TWI | Species)
+$$
+
+All statistical analyses were performed using R (version). 
+Mixed models were fit through Bayesian regressions using MCMC methods using the package _brms_ [REF].
 
 # Results {-}
+
+## Interannual growth variation and drought sensitivity {-}
+
+![***Figure 3: Variation in sensitivity among species*** in the Huai Kha Khaeng ForestGEO plot during droughts in 2010 and 2015.
+For comments: https://github.com/forestgeo/growth-precip-thailand/issues/12](display/Fig3.png)
+
+## Variation in sensitivity among species {-}
+
+## Variation among individuals {-}
 

### 2025-01-30 scripts/06_figures.R
updated Fig 1, attempt at FIg 4 #13  #10

@@ -136,8 +136,8 @@ ffstation_monthly_rl <- ffstation %>%
     arrange(YearII, Month, Date, climvar) %>%
     group_by(climvar) %>%
     dplyr::mutate(
-        rlsum = zoo::rollsum(value, k = 15, fill = NA, align = "center"),
-        rlmean = zoo::rollmean(value, k = 15, fill = NA, align = "center"),
+        rlsum = zoo::rollsum(value, k = 30, fill = NA, align = "center"),
+        rlmean = zoo::rollmean(value, k = 30, fill = NA, align = "center"),
         rlval = ifelse(climvar == c("Precipitation"), rlsum, rlmean)
     ) %>%
     rename(year = YearII, month = Month) %>%
@@ -177,7 +177,7 @@ varnames <- as_labeller(c(
 
 climplot <- ggplot() +
     geom_line(data = ffstation_full_rl, aes(x = day_of_year, y = rlval, color = year, linetype = year), linewidth = 0.8) +
-    facet_wrap(~climvar, scales = "free_y", labeller = varnames) +
+    facet_wrap(~climvar, scales = "free_y", labeller = varnames, strip.position = "left") +
     theme_bw() +
     scale_linetype_manual(values = c("long-term" = "longdash", "2010" = "solid", "2015" = "solid")) +
     geom_ribbon(data = ffstation_lt_rl %>% filter(climvar %in% c("Precipitation", "dry_days", "TempMax", "vpdmax")), aes(x = day_of_year, ymin = rlval - rlse, ymax = rlval + rlse), alpha = 0.3) +
@@ -190,8 +190,12 @@ climplot <- ggplot() +
         aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), fill = "grey", alpha = 0.3
     ) +
     guides(linetype = "none") +
-    labs(x = "Day of year", y = "Value", color = "Year") +
-    scale_color_manual(values = c("long-term" = "grey40", "2010" = "indianred2", "2015" = "indianred4"))
+    labs(x = "Day of year", y = "", color = "Year") +
+    scale_color_manual(values = c("long-term" = "grey40", "2010" = "indianred2", "2015" = "indianred4")) +
+    theme(
+        strip.background = element_blank(),
+        strip.placement = "outside"
+    )
 
 climplot
 
@@ -206,8 +210,8 @@ speiplot <- ggplot() +
[…]

### 2025-01-30 scripts/06_figures.R
updated manuscript with fig 4

@@ -446,9 +446,7 @@ dag_img <- magick::image_read("doc/display/hypotheses.png")
 print(dag_img)
 dag_img
 
-dag_gg <- ggplot() +
-    ggpubr::background_image(dag_img) +
-    theme_minimal()
+dag_gg <- magick::image_ggplot(dag_img, interpolate = F)
 
 # read coefs
 coefs_df <- readRDS("results/models/non_negative/sensitivity_model_spre.RData")
@@ -479,7 +477,7 @@ coefs_all_sp <- ggplot(
     geom_hline(yintercept = 0, linetype = "dashed") +
     # facet_grid(~param, scales = "free", labeller = par_names) +
     facet_grid(~ factor(yr, levels = c(2010, 2015)), scales = "free") +
-    labs(title = "Effect of parameters on growth sensitivity", x = "", y = "coefficient") +
+    labs(title = "", x = "", y = "coefficient") +
     guides(color = "none") +
     theme_bw() +
     coord_flip()
@@ -489,6 +487,6 @@ library(patchwork)
 
 
 
-png("doc/display/Fig4.png", width = 4, height = 4, units = "in", res = 300)
-(dag_gg / coefs_all_sp) + plot_annotation(tag_levels = "a") + plot_layout(heights = c(1.5, 1))
+png("doc/display/Fig4.png", width = 6, height = 4, units = "in", res = 300)
+(dag_gg / coefs_all_sp) + plot_annotation(tag_levels = "a") + plot_layout(heights = c(2, 1))
 dev.off()

### 2025-02-05 scripts/06_figures.R
update fig 1 following comments #10
@teixeirak fyi cosmetic changes

@@ -21,6 +21,104 @@ ffstation <- ffstation %>%
         dry_days = ifelse(Precipitation == 0, 1, 0)
     )
 
+
+# figure 1 alt-------------------------------------------
+
+# make a figure with rolling monthly means
+ffstation_monthly_rl <- ffstation %>%
+    pivot_longer(cols = c(Precipitation, dry_days, TempMax, vpdmax, vpdmin, vpdmean), names_to = "climvar", values_to = "value") %>%
+    select(YearII, Month, Date, climvar, value) %>%
+    arrange(YearII, Month, Date, climvar) %>%
+    group_by(climvar) %>%
+    dplyr::mutate(
+        rlsum = zoo::rollsum(value, k = 30, fill = NA, align = "center"),
+        rlmean = zoo::rollmean(value, k = 30, fill = NA, align = "center"),
+        rlval = ifelse(climvar == c("Precipitation"), rlsum, rlmean)
+    ) %>%
+    rename(year = YearII, month = Month) %>%
+    dplyr::mutate(year = as.character(year)) %>%
+    dplyr::mutate(
+        Date2 = as.Date(paste0(year, "-", month, "-", Date)),
+        day_of_year = as.numeric(format(Date2, "%j"))
+    ) %>%
+    select(-c(rlsum, rlmean))
+
+ffstation_lt_rl <- ffstation_monthly_rl %>%
+    filter(year <= 2021, year >= 2009) %>%
+    ungroup() %>%
+    group_by(day_of_year, climvar) %>%
+    dplyr::summarise(
+        rlse = sd(rlval, na.rm = T) / sqrt(sum(!is.na(rlval))),
+        rlval = mean(rlval, na.rm = T)
+    ) %>%
+    ungroup() %>%
+    dplyr::mutate(year = "long-term")
+select(year, month, Date, climvar, rlval, rlse, day_of_year)
+
+# bind these two dataframes
+ffstation_full_rl <- bind_rows(ffstation_monthly_rl, select(ffstation_lt_rl, -("rlse"))) %>%
+    filter(year %in% c("long-term", "2010", "2015")) %>%
+    filter(climvar %in% c("Precipitation", "dry_days", "TempMax", "vpdmax"))
+
+varnames <- as_labeller(c(
+    Precipitation = "Precipitation (mm)",
+    dry_days = "Dry days",
+    TempMax = "Max temperature (°C)",
+    vpdmax = "VPD max (kPa)",
+    vpdmin = "VPD min (kPa)",
[…]

### 2025-02-05 scripts/06_figures.R
updated Fig 2 #11 cosmetic

@@ -82,8 +82,8 @@ climplot <- ggplot() +
     scale_linetype_manual(values = c("long-term" = "longdash", "2010" = "solid", "2015" = "solid")) +
     geom_ribbon(data = ffstation_lt_rl %>% filter(climvar %in% c("Precipitation", "dry_days", "TempMax", "vpdmax")), aes(x = day_of_year, ymin = rlval - rlse, ymax = rlval + rlse), alpha = 0.3) +
     # text for wet and dry season
-    annotate("text", x = -Inf, y = -Inf, label = "wet season", col = "black", hjust = -0.5, vjust = -1, fontface = "italic", size = 1.25) +
-    annotate("text", x = -Inf, y = -Inf, label = "dry season", col = "black", hjust = -2.5, vjust = -1, fontface = "italic", size = 1.25) +
+    annotate("text", x = -Inf, y = -Inf, label = "wet season", col = "black", hjust = -0.5, vjust = -1, fontface = "italic", size = 1.15) +
+    annotate("text", x = -Inf, y = -Inf, label = "dry season", col = "black", hjust = -2.5, vjust = -1, fontface = "italic", size = 1.15) +
     guides(linetype = "none") +
     labs(x = "Day of year", y = "", color = "Year") +
     scale_color_manual(values = c("long-term" = "grey60", "2010" = "indianred2", "2015" = "indianred4")) +
@@ -118,7 +118,7 @@ climplot + speiplot + plot_layout(widths = c(1.6, 1)) + plot_annotation(tag_leve
 dev.off()
 
 
-# monthly means as supplementary plot
+# monthly means as supplementary plot-------------
 ffstation_monthly <- ffstation %>%
     group_by(YearII, Month) %>%
     dplyr::summarise(
@@ -285,11 +285,14 @@ sens.all <- ggplot(data = tree.time %>% filter(yr %in% yrs), aes(x = sens.prop))
     facet_wrap(~yr) +
     # xlim(-5, 5)+
     labs(x = "Drought sensitivity", y = "Density") +
-    theme_bw()
+    theme_bw() +
+    # make the strip title bold
+    theme(
+        strip.background = element_blank(),
+        strip.placement = "outside",
+        strip.text = element_text(size = 12)
+    )
 
-# png("doc/display/sens_all.png", width = 6, height = 4, units = "in", res = 300)
-# sens.all
-# dev.off()
 
[…]

### 2025-02-06 scripts/06_figures.R
updated SI

@@ -85,8 +85,8 @@ climplot <- ggplot() +
     scale_linetype_manual(values = c("long-term" = "longdash", "2010" = "solid", "2015" = "solid")) +
     geom_ribbon(data = ffstation_lt_rl %>% filter(climvar %in% c("Precipitation", "dry_days", "TempMax", "vpdmax")), aes(x = day_of_year, ymin = rlval - rlse, ymax = rlval + rlse), alpha = 0.3) +
     # text for wet and dry season
-    annotate("text", x = -Inf, y = -Inf, label = "wet season", col = "black", hjust = -0.5, vjust = -1, fontface = "italic", size = 1.15) +
-    annotate("text", x = -Inf, y = -Inf, label = "dry season", col = "black", hjust = -2.5, vjust = -1, fontface = "italic", size = 1.15) +
+    annotate("text", x = -Inf, y = -Inf, label = "dry season", col = "black", hjust = -0.5, vjust = -1, fontface = "italic", size = 1.15) +
+    annotate("text", x = -Inf, y = -Inf, label = "wet season", col = "black", hjust = -2.5, vjust = -1, fontface = "italic", size = 1.15) +
     guides(linetype = "none") +
     labs(x = "Day of year", y = "", color = "Year") +
     scale_color_manual(values = c("long-term" = "grey60", "2010" = "indianred2", "2015" = "indianred4")) +
@@ -99,6 +99,34 @@ climplot <- ggplot() +
 # https://stackoverflow.com/questions/75007050/manually-adjusting-the-z-order-of-geom-layers
 
 climplot
+
+# spei
+
+spei_month <- spei %>%
+    dplyr::mutate(
+        Date = as.Date(system.time_start, format = "%B %d, %Y"),
+        year = year(Date),
+        month = month(Date)
+    ) %>%
+    rename(spei_val = SPEI_01_month) %>%
+    select(year, month, spei_val)
+
+spei_lt <- spei_month %>%
+    filter(year <= 2021, year >= 2009) %>%
+    ungroup() %>%
+    group_by(month) %>%
+    dplyr::mutate(
+        spei_val = mean(spei_val, na.rm = T),
+        year = "long-term"
+    ) %>%
+    select(year, month, spei_val)
+
+spei_month <- spei_month %>%
+    dplyr::mutate(year = as.character(year)) %>%
+    filter(year %in% c("long-term", "2010", "2015"))
+
[…]

### 2025-02-12 scripts/000_twi.R
TWI at the watershed  #16

@@ -126,6 +126,272 @@ tm_shape(twi) +
 
 png("doc/display/explore/twi.png", width = 4, height = 8, units = "in", res = 300)
 tm_shape(twi) +
-  tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE, alpha=1) +
+  tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE, alpha = 1) +
+  tm_scale_bar()
+dev.off()
+
+
+# make TWI with SRTM--------------------
+
+# load data-----------------------
+# elevation
+
+
+# convert to topographic wetness index
+# following code from https://vt-hydroinformatics.github.io/rgeoraster.html
+
+hkk_elev <- raster("data/HKK-other/N15E099.SRTMGL1.hgt/N15E099.hgt")
+
+theme_set(theme_classic())
+
+# read hkk shapefile
+hkkwls <- read_sf("data/HKK-other/HKK_boundary/shapefile/WDPA_WDOECM_Feb2025_Public_1407_shp-polygons.shp")
+
+# HKK plot as point
+# from the HKK site data repo (15.6324 N, 99.217 E)
+hkk_point <- st_point(c(99.217, 15.6324))
+
+# Convert hkk_point to an sf object
+hkk_point_sf <- st_sfc(hkk_point, crs = st_crs(hkkwls))
+hkk_point_sf <- st_sf(geometry = hkk_point_sf)
+
+writeRaster(hkk_elev, "data/HKK-other/HKKDEM_30m.tif", overwrite = TRUE)
+
+
+png("doc/display/explore/hkk_elev_30m.png", width = 8, height = 8, units = "in", res = 300)
+tm_shape(hkk_elev) +
+  tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE) +
+  tm_shape(hkkwls) +
+  tm_borders(lwd = 4) +
+  tm_shape(hkk_point_sf) +
+  tm_symbols(size = 2, shape = 15, border.lwd = 2, col = NA) +
+  tm_scale_bar()
+dev.off()
+
+# hillshade
+
+wbt_hillshade(
+  dem = "data/HKK-other/HKKDEM_30m.tif",
+  output = "data/HKK-other/HKK_hillshade_30m.tif",
+  azimuth = 115
+)
+
+hillshade <- raster("data/HKK-other/HKK_hillshade_30m.tif")
+
+tm_shape(hillshade) +
[…]

### 2025-02-12 scripts/08_species.R
isocline plots for combined model
combined model with cii, twi and dbh and their interactions with deciduousness + actual measurements

@@ -54,8 +54,6 @@ yrs <- c(2010, 2015)
 
 # species median and lms------------------------
 
-
-
 sens_sp <- tree.time %>%
     filter(yr %in% yrs) %>%
     group_by(Species, yr) %>%
@@ -225,6 +223,8 @@ saveRDS(pred, "results/models/non_negative/pred_isoclines_nore.RDS")
 saveRDS(coefs, "results/models/non_negative/coefs_isoclines_nore.RDS")
 saveRDS(new_preds, "results/models/non_negative/new_preds_isoclines_nore.RDS")
 
+coefs <- readRDS("results/models/non_negative/coefs_isoclines_nore.RDS")
+
 new_preds_df <- do.call(rbind, new_preds)
 coefs_df <- do.call(rbind, coefs)
 
@@ -233,6 +233,7 @@ coefs_df$signif <- ifelse(coefs_df$lwr < 0 & coefs_df$upr > 0, "no",
 )
 
 coefs_df
+
 
 library(ggplot2)
 # plot for four species
@@ -445,4 +446,232 @@ iso_plot <- ggplot(
 
 png("results/plots/non_negative/isocline_cii.png", width = 8, height = 4, units = "in", res = 300)
 iso_plot
+dev.off()
+
+
+# all coefs together------------
+
+coefs_twi <- readRDS("results/models/non_negative/coefs_isoclines_nore.RDS")
+coefs_dbh <- readRDS("results/models/non_negative/coefs_isoclines_dbh.RDS")
+coefs_cii <- readRDS("results/models/non_negative/coefs_isoclines_cii.RDS")
+
+coefs_df <- rbind(do.call(rbind, coefs_twi), do.call(rbind, coefs_dbh), do.call(rbind, coefs_cii))
+
+coefs_df$signif <- ifelse(coefs_df$lwr < 0 & coefs_df$upr > 0, "no",
+    ifelse(coefs_df$lwr > 0, "pos", "neg")
+)
+coefs_df$model <- rep(c("twi", "dbh", "cii"), each = nrow(coefs_df) / 3)
+
+# plot coefs
+# first make long df
+library(tidyverse)
+`%nin%` <- Negate(`%in%`)
+coefs_df <- coefs_df %>%
+    filter(param %nin% c("lp__", "lprior", "b_Intercept", "Intercept", "sigma"))
+
+coefs_interact <- ggplot(coefs_df) +
+    geom_point(aes(x = param, y = median, color = signif, shape = model)) +
[…]

### 2025-02-12 scripts/08_species.R
coef plots for deciduousness

@@ -479,6 +479,10 @@ coefs_interact <- ggplot(coefs_df) +
     geom_hline(yintercept = 0, linetype = "dashed") +
     coord_flip()
 
+png("results/plots/non_negative/coefs_interact.png", width = 8, height = 4, units = "in", res = 300)
+coefs_interact
+dev.off()
+
 # pred vs obs--------------------------------
 
 pred_twi <- readRDS("results/models/non_negative/pred_isoclines_nore.RDS")
@@ -606,6 +610,8 @@ saveRDS(pred, "results/models/non_negative/pred_isoclines_all_dec.RDS")
 saveRDS(coefs, "results/models/non_negative/coefs_isoclines_all_dec.RDS")
 saveRDS(new_preds, "results/models/non_negative/new_preds_isoclines_all_dec.RDS")
 
+coefs_df <- readRDS("results/models/non_negative/coefs_isoclines_all_dec.RDS")
+
 new_preds_df <- do.call(rbind, new_preds)
 coefs_df <- do.call(rbind, coefs)
 
@@ -631,8 +637,8 @@ iso_plot1 <- ggplot() +
         aes(x = williams_dec, y = calcDBH_min1, color = sens.prop)
     ) +
     theme_bw() +
-    scale_fill_gradientn(colors = c("#5a39fc", "white", "#ef4738"), values = c(0, 0.5, 1), limits = c(-2, 2)) +
-    scale_color_gradientn(colors = c("#5a39fc", "white", "#ef4738"), values = c(0, 0.5, 1), limits = c(-2, 2)) +
+    scale_fill_gradientn(colors = rev(c("#5a39fc", "white", "#ef4738")), values = c(0, 0.5, 1), limits = c(-2, 2)) +
+    scale_color_gradientn(colors = rev(c("#5a39fc", "white", "#ef4738")), values = c(0, 0.5, 1), limits = c(-2, 2)) +
     guides(color = "none")
 
 iso_plot2 <- ggplot() +
@@ -649,8 +655,8 @@ iso_plot2 <- ggplot() +
         aes(x = williams_dec, y = cii_min1, color = sens.prop)
     ) +
     theme_bw() +
-    scale_fill_gradientn(colors = c("#5a39fc", "white", "#ef4738"), values = c(0, 0.5, 1), limits = c(-2, 2)) +
[…]

### 2025-02-12 doc/manuscript.Rmd
updated manuscript with fig 4
update fig 1 following comments #10
@teixeirak fyi cosmetic changes
methods draft (869 words)

@@
     #pdf_document:
+    # engine: xelatex 
   #officedown::rdocx_document
@@
 Standard methods like tree rings and plot inventories may either not represent most species or may not be accurate enough to capture variation in slow-growing species.  
-- We used a unique data record - manual dendrometer band measurements over 15 years on 1820 trees across 30 species in a seasonally dry tropical forest in Huai Kha Khaeng, Thailand, along with species traits (deciduousness and maximum size), individual characteristics (crown exposure, size) and environment-related water availability to predict the sensitivity of tree growth to drought in two ENSO drought events.
+- We used a unique data record - manual dendrometer band measurements over 15 years on 1820 trees across 30 species in a seasonally dry tropical forest in Huai Kha Khaeng, Thailand, along with species traits (deciduousness and maximum size), individual characteristics (crown exposure, size) and environment-related water availability to predict the sensitivity of tree growth to drought in two ENSO drought events with different characteristics - a wet-season drought and a dry season drought.
 We used Bayesian regressions in a causal modelling framework to predict variability among species and individuals in drought sensitivity.
-- In the more severe drought, deciduous species were more sensitive than evergreen species. 
+- In the wet-season drought, deciduous species were more sensitive than evergreen species, especially in wetter microsites, while these patterns were not significant in the dry-season drought. 
 After accounting for species differences, more exposed trees were more likely to be negatively impacted than their less exposed counterparts, with exposure showing clear negative impacts compared to size *per se*.
@@
 $$
+By scaling the difference in increment to the mean annual growth of each tree, this metric allows us to compare across slow and fast growing individuals and species.
 
-Across all individuals in the plot, we calculated topographic wetness index based on a 5 m x 5 m elevation layer collected at plot establishment.
-Topographic wetness index provides an expectation of the water availability across space based on the concavity/convexity of a location. 
-To model sensitivity among individuals, we also used tree size (DBH at previous dendroband census) and crown exposure measured as crown illumination index.
-Crown illumination index is an ordinal categorical variable from 1 to 5 (least to most exposed) measured at each dendroband census.
+Across all individuals in the plot, we calculated topographic wetness index for an expectation of the water availability across space based on the concavity/convexity of a location. 
+We used a Digital Elevation Model from a 5 m x 5 m elevation layer collected at plot establishment.
+We used the *whitebox* package in R [@wu_whitebox_2022; @lindsay_whitebox_2016a] to fill and breach this layer, calculate flow accumulation at each cell and then calculate TWI as $log_{e}\frac{As}{tan(slope)}$.
+We defined tree size as the diameter at breast height (calculated from dendrometer band window measurement) at the previous census. 
+We estimated crown exposure for each tree at each census using crown illumination index, an ordinal categorical variable from 1 to 5 (least to most exposed) measured at each dendroband census.
 
@@
 At the species level, we also calculated the maximum size across all individuals in the plot as measured in the 2008 whole-plot census.
+We calculated niche breadth as the standard deviation of TWI across all observed individuals of each species in the whole plot census in 2008.
 
@@
 
-For individual stems, we modelled drought sensitivity as:
+To model the interactive effects of deciduousness on sensitivity, we modelled
 
 $$
-sensitivity \sim DBH + CII + TWI + size + (DBH+CII+TWI | Species)
+sensitivity \sim 1 + TWI + deciduousness + TWI:decidousness
 $$
+
+We then calculated predictions across all combinations of values from observed maximum to minimum of TWI and deciduousness within this dataset.
+
+To further disentange the drivers of stem-level drought sensitivity, we modelled:
+
+$$
+sensitivity \sim DBH + CII + TWI + (DBH+CII+TWI | Species)
+$$
+
+We chose to model a species random effect on all slopes here to account for variation across species instead of specific traits which may have different effects with each of the variables.
 
@@
 
+![***Figure 4: Drivers of variation in sensitivity among individuals***](display/Fig4.png)
+
 # Discussion {-}
@@
 # Acknowledgements {-}
+
+Huai Kha Khaeng forest monitoring team. 
+Ekaphan Kraichak, Kanisorn Chowtiwuttakorn. 
+Jose Medina-Vega, Valentine Herrmann, Eugenie Mas.
+Smithsonian Institution Postdoctoral Fellowship.
+ForestGEO Analytical Workshop, Asia 2024.
 

### 2025-02-13 scripts/02_models.R
looking at TWI slope effects with deciduosness #17 #13

@@ -1712,3 +1712,143 @@ library(gridExtra)
 png("results/plots/non_negative/coefs_mediation.png", width = 8, height = 4, units = "in", res = 300)
 grid.arrange(coefs_tree_no_cii, coefs_dbh_cii, ncol = 1, heights = c(2, 1))
 dev.off()
+
+
+# plot species random effect slope
+
+# read coefs
+coefs_df <- readRDS("results/models/non_negative/sensitivity_model_spre.RData")
+
+head(coefs_df)
+# species random effects on TWI slope
+
+# mean slope
+coefs_twi <- coefs_df %>%
+    filter(grepl("b_twi_scaled_sp", param))
+
+# random effect on slope
+coefs_sp_twi <- coefs_df %>%
+    filter(grepl("r_Species", param)) %>%
+    filter(grepl("twi", param)) %>%
+    filter(grepl("cii|DBH|Intercept", param) == FALSE) %>%
+    dplyr::mutate(
+        Species = gsub("r_Species\\[|\\,twi_scaled_sp\\]", "", param)
+    ) %>%
+    group_by(yr) %>%
+    dplyr::mutate(
+        total_effect = median + coefs_twi$median[match(yr, coefs_twi$yr)],
+        total_lwr = lwr + coefs_twi$median[match(yr, coefs_twi$yr)],
+        total_upr = upr + coefs_twi$median[match(yr, coefs_twi$yr)]
+    )
+
+# merge species vars
+coefs_sp_twi <- merge(coefs_sp_twi, sp_vars, by = "Species", all.x = TRUE)
+
+# cors
+twi_cor <- coefs_sp_twi %>%
+    group_by(yr) %>%
+    dplyr::summarise(
+        cor = cor.test(williams_dec, total_effect)[4]$estimate,
+        cor_p = cor.test(williams_dec, total_effect)[3]$p.value
+    )
+
+twi_cor
+# plot coefs
+
+library(ggpubr)
+twi_slopes_plot <- ggplot(
+    data = coefs_sp_twi,
+    # aes(x = reorder(Species, median), y = median),
+    aes(x = williams_dec, y = total_effect),
+    order = median
+) +
+    geom_point() +
+    # geom_smooth(method = "lm", col = "grey40") +
+    geom_errorbar(aes(ymin = total_lwr, ymax = total_upr), width = 0.1) +
+    geom_hline(yintercept = 0, linetype = "dashed") +
[…]

### 2025-02-13 doc/manuscript.Rmd
more fleshed out methods

@@
 
-We analysed data from manual dendrometer band censuses from 2008-2023 in the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
+We analysed data from manual dendrometer band censuses from 2008-2023 in the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
 This is a 50-hectare plot in mixed deciduous tropical forest, with a mean annual temperature of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
-The dominant canopy species are _Hopea odorata_, _Vatica harmandiana_, _Dipterocarpus alatus_ and _Tetrameles nudiflora_ and lower layers of the canopy are dominated by sub-canopy evergreen and deciduous species like _Saccopetalum linneatum_ and _Polyalthia viridis_.
+The dominant canopy species are _Hopea odorata_, _Vatica harmandiana_, _Dipterocarpus alatus_ and _Tetrameles nudiflora_ and lower layers of the canopy are dominated by sub-canopy evergreen and deciduous species like _Miliusa horsfieldii_ and _Polyalthia viridis_.
 Long-term analysis of tree-ring data shows that tree growth at Huai Kha Khaeng has been influenced by temperature and rainfall [@vlam_temperature_2014].
 
-We used the late wet/early dry dendrometer band censuses to calculate annual growth for each tree. 
-We used the raw window size measurements combined with the diameter at breast height (DBH) measurement at installation to calculate DBH at each census using standard equations [condit].
-We removed measures with likely data entry errors because of misplaced decimals, potential misidentified bands (old band numbers that restarted after stopping) and large measurement outliers (> 3 standard deviations from the mean across all observations).
-We then calculated annualised increments for each individual for each year by finding the difference in increment from the previous year and adjusting for the number of days between measurements.
-We repeated the same steps for annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement with these tape increments.  
+In the HKK plot, metal dendrometer bands were installed in 2008 on 2353 trees across 152 species following standard protocol [@muller-landau_metal_2008]. 
+Trees were selected were based on the relative abundance of species in the forest, with more individuals banded from the abundant species (ranging from 230 trees to 1 tree per species), and spanning the size and habitat distribution within the plot.
+On each tree, the dendrometer band was installed at 1.4 m from the ground, a height of 10 cm above the regular point of measure for the whole plot census (typically 1.3 m), except on irregular trunks or buttresses when they were shifted above.
+All dendrobands were censused two times a year - in the wet and dry season - typically in June and January. 
+During the census, the size of the window opening along the band was measured using Mitutoyo digital callipers (least count = 0.01 mm) and the diameter of the tree at the dendrometer band was measured using standard DBH tape.
+
+Using the late wet/early dry dendrometer band censuses, we calculated annualised growth for each year for each tree. 
+We first removed measurements with likely data entry errors because of misplaced decimals, potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses).
+We used the raw window size measurements combined with the diameter at breast height (DBH) measurement at installation to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[condit].
+Since this was dependent on the diameter measurement at installation, we used a direct measurement at the time of installation if it was available.
+If the tape measurement at the first window size measurement was not available for any band, we excluded window size measurements until the first available diameter measurement. 
+We then calculated annualised diameter increments for each individual for each year by finding the difference in diameter from the previous year ($inc = dia_{t} - dia_{t-1}$) and adjusting for the number of days between measurements ($inc_{annual} = inc * \frac{tdif}{365}$).
+To ensure that the data did not smooth over multiple years, we excluded increment measures with missing values for $dia_{t-1}$ and then excluded large measurement outliers (> 3 standard deviations from the mean across all observations).
+We repeated the same steps using annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement with these annualised increments calculated from tape measurements.
 Finally, we excluded trees with negative increments over the whole timeseries, and included only species that had at least 10 individuals, resulting in a final dataset of 1820 individuals across 30 species.
@@
 
-Combining this growth series with climatic information, we identified two drought years of interest - 2010 and 2015 - and calculated growth sensitivity for each tree to each drought as a growth anomaly from its mean annual increment.
+***Drought years and characteristics***
+
+Combining the growth timeseries from dendrometer band measurements with climatic information, we identified two drought years of interest - 2010 and 2015.
+To identify drought years, we identified years with low plot-level growth rate and combined this with information on ENSO events that affected the landscape. 
+We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected.
+Moreover, broad definitions based on climate variables (often not measured at the location), may not capture ecological realities within the plot. 
+Out of the two years with low plot-level growth, 2010 was a moderate ENSO event while 2015 was a very strong ENSO event. 
+These years also corresponded with expert knowledge of major droughts that affected the region.
+
+To characterise these two droughts, we used climate data from a weather station near the plot and publicly available climate data for the region. 
+We used daily weather data collected from 2001 to 2019 at the Forest Fire Research Station [CHECK WHERE], located ~20 km from the ForestGEO plot. 
+We first calculated daily values of Vapour Pressure Deficit using the equation $$
+0.6108 * e^{17.27 * \frac{t_{max}}{t_{max}+237.3}}* (1-\frac{relative.humidity}{100})
+$$
+We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 15 day periods across the whole year for each year.
+Finally, we calculated the long-term mean for each variable as the mean of the values of the rolling mean across all years of the dendroband measurements, and the standard error around this mean. 
+We also used the SPEIbase dataset [REF] to calculate the Standardised Precipitation-Evapotranspiration Index, a drought severity index, calculated at the monthly scale. 
+We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.
+
+***Variables for analysis***
+
+For the two drought years, 2010 and 2015, we calculated growth sensitivity for each tree to each drought as a growth anomaly from its mean annual increment.
 We first calculated the mean annual increment for each tree as the mean of the annualised increments across the whole timeseries.
@@
 
-We calculated three species-level traits : deciduousness, maximum size and niche breadth and one variable for each tree - topographic wetness index and two variables for each tree at each census - starting size and canopy exposure.
+We calculated three species-level traits : deciduousness, maximum size and niche breadth. 
 We defined deciduousness as the mean proportion of canopy loss at maximum loss across individuals in the species, with a score ranging from 1 to 4 where 1 is most evergreen and 4 is most deciduous. 
@@
 
-All statistical analyses were performed using R (version). 
+All statistical analyses were performed using R version 4.4.0 [@rcoreteam_language_2024]. 
 Mixed models were fit through Bayesian regressions using MCMC methods using the package _brms_ [REF].
+We ran models using gaussian priors for the data distribution.
+We ran 4 MCMC chains for 3000 iterations with 1000 of these set as warmup.
+We extracted draws, calculated distributions using functions from the _posterior_ package [REF].
 

### 2025-02-14 doc/manuscript.Rmd
reading data file into manuscript for results

@@
 These years also corresponded with expert knowledge of major droughts that affected the region. 
+We also note that we chose not to include 2020 in our analysis despite low growth, because a) declines in growth can occur due to various reasons which could have distinct ecological responses in species and individuals b) 2020 did not emerge as a drought year either in the examination of drought indicators or from expert knowledge.
 
@@
 
+## Drought characteristics {-}
+
+The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Fig 1).
+In 2010, drought characteristics of lower precipitation, higher number of dry days, higher maximum temperature, higher VPD than the long-term mean and SPEI values below -1 occurred between February and April, in the dry season.
+On the other hand, in 2015, similar deviations occurred in May and June, in the wet season. 
+In 2015, SPEI in May was -2, an anomalously low value that was not observed in any month in the entire dendroband census time period, suggesting a severe drought.
+
+```{r data summaries, echo=F, warning=F, message=F}
+library(tidyverse)
+tree.time <- readRDS("../data/HKK-dendro/sensitivity_data_formodels.RData")
+
+tree.time.sp <- tree.time %>%
+  group_by(Species) %>%
+  dplyr::summarise(inc_annual_med = round(median(inc_annual, na.rm = T), 2)) %>%
+  arrange(inc_annual_med)
+```
+
 ## Interannual growth variation and drought sensitivity {-}
+
+The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig 2a). 
+Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual, na.rm=T),2)` cm. 
+Median annual increment differed by species with `r tree.time.sp$Species[1]` having the lowest (`r tree.time.sp$inc_annual_med[1]` cm) and `r tree.time.sp$Species[nrow(tree.time.sp)]` the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` cm)among the `r nrow(tree.time.sp)` species analysed.
+The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015], na.rm=T),2)` cm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015], na.rm=T)-median(tree.time$inc_annual, na.rm=T))/median(tree.time$inc_annual, na.rm=T)),2)`% lower than the median across all years.
+In 2010, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010], na.rm=T),2)` cm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010], na.rm=T)-median(tree.time$inc_annual, na.rm=T))/median(tree.time$inc_annual, na.rm=T)),2)`% lower than the median across all years.
+While there were more species differences in median responses to the 2010 drought, all species had lower increment than a typical year in 2015.
+
+Drought sensitivity across all trees also showed negative sensitivities for both droughts with considerable variation around it (Fig 2b).
 

### 2025-02-14 scripts/02_models.R

@@ -51,6 +51,7 @@ tree.time <- tree.time %>%
     filter(!is.na(sens.prop) & !is.na(cii_min1) & !is.na(calcDBH_min1) & !is.na(twi)) %>%
     ungroup()
 
+saveRDS(tree.time, "data/HKK-dendro/sensitivity_data_formodels.RData")
 # plot the sensitivity values for all trees in 2010 and 2015
 yrs <- c(2010, 2015)
 

### 2025-02-14 doc/manuscript.Rmd
One section of results

@@
 tree.time <- readRDS("../data/HKK-dendro/sensitivity_data_formodels.RData")
+# tree.time <- readRDS("data/HKK-dendro/sensitivity_data_formodels.RData")
 
@@
   group_by(Species) %>%
-  dplyr::summarise(inc_annual_med = round(median(inc_annual, na.rm = T), 2)) %>%
+  dplyr::summarise(
+    inc_annual_med = round(median(inc_annual, na.rm = T), 2),
+    inc_annual_sd = round(sd(inc_annual, na.rm = T), 2)
+  ) %>%
   arrange(inc_annual_med)
@@
 The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig 2a). 
-Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual, na.rm=T),2)` cm. 
-Median annual increment differed by species with `r tree.time.sp$Species[1]` having the lowest (`r tree.time.sp$inc_annual_med[1]` cm) and `r tree.time.sp$Species[nrow(tree.time.sp)]` the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` cm)among the `r nrow(tree.time.sp)` species analysed.
+Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual, na.rm=T),2)` cm. 
+Median annual increment differed by species with `r tree.time.sp$Species[1]` having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` cm) and `r tree.time.sp$Species[nrow(tree.time.sp)]` the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` cm) among the `r nrow(tree.time.sp)` species analysed.
+
+Drought years were associated with lower growth across the plot. 
 The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015], na.rm=T),2)` cm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015], na.rm=T)-median(tree.time$inc_annual, na.rm=T))/median(tree.time$inc_annual, na.rm=T)),2)`% lower than the median across all years.
@@
 While there were more species differences in median responses to the 2010 drought, all species had lower increment than a typical year in 2015. 
+Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years. 
+Across the years, on average, `r round(100*mean(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% of the trees had negative growth, with a minimum of `r round(100*min(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% in `r names(sort(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))))[1]`.
+The drought year 2015 had the maximum proportion of trees with negative growth in the timeseries, `r round(100*sum(tree.time$inc_annual[tree.time$yr==2015]<0)/nrow(tree.time[tree.time$yr==2015,]),2)`%.
 

### 2025-02-14 scripts/02_models.R

@@ -52,6 +52,7 @@ tree.time <- tree.time %>%
     ungroup()
 
 saveRDS(tree.time, "data/HKK-dendro/sensitivity_data_formodels.RData")
+
 # plot the sensitivity values for all trees in 2010 and 2015
 yrs <- c(2010, 2015)
 

### 2025-02-18 scripts/06_figures.R
Fig 4 coloured by deciduousness #13

@@ -3,7 +3,7 @@
 # load libraries
 library(tidyverse)
 library(lubridate)
-
+library(ggplot2)
 
 
 # figure 1 ------------------------------------------------
@@ -608,7 +608,6 @@ dev.off()
 # figure 4 alternate----------------------------------
 # read coefs
 coefs_df <- readRDS("results/models/non_negative/sensitivity_model_spre.RData")
-
 par_names <- as_labeller(c("b_calcDBH_min1_scaled_sp" = "DBH effect", "b_cii_min1_scaled_sp" = "CII effect", "b_twi_scaled_sp" = "TWI effect"))
 
 coefs_all_sp <- ggplot(
@@ -702,9 +701,11 @@ preds_df <- readRDS("results/models/non_negative/predictions_spre.RData")
 
 # plot predictions
 pred_plot <- ggplot(data = preds_df, aes(x = twi_scaled_sp, y = median)) +
-    geom_smooth(aes(group = Species), method = "lm", col = "grey60") +
+    # geom_smooth(aes(group = Species, color=williams_dec), method = "lm", col = "grey60") +
+    geom_smooth(aes(group = Species, col = williams_dec), method = "lm") +
     geom_smooth(method = "lm", col = "black") +
     geom_point(aes(x = twi_scaled_sp, y = sens.prop), alpha = 0.1) +
+    scale_color_gradient(low = "darkgreen", high = "brown") +
     # geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.1) +
     geom_hline(yintercept = 0, linetype = "dashed") +
     facet_wrap(~ factor(yr, levels = c(2010, 2015)),
@@ -723,6 +724,8 @@ pred_plot
 str(preds)
 
 library(patchwork)
-png("doc/display/Fig4_alternate.png", width = 12, height = 8, units = "in", res = 300)
-(coefs_all_sp / twi_slopes_plot) | pred_plot + plot_annotation(tag_levels = "a") #+ plot_layout(heights = c(0.5, 1), widths = c(1, 1))
+png("doc/display/Fig4_alternate.png", width = 16, height = 8, units = "in", res = 300)
+(coefs_all_sp / twi_slopes_plot) | pred_plot + plot_annotation(tag_levels = "a") #+ plot_layout(widths = c(1, 1.5))
 dev.off()
+
+# fig 5-----------------------------------

### 2025-02-18 doc/manuscript.Rmd
Rough text of last intro para

@@
 
-Here we use a 14-year record of dendrometer band measurements in seasonally dry forest in Thailand to test the vulnerability of tropical tree growth to drought.
-We expect variation in drought sensitivity (growth in drought year/growth in a previous year with normal rainfall) across trees.
-We ask: i) How much do species vary in drought sensitivity of tree growth?
-What drives this variation ii) What drives variation among individuals in the sensitivity of growth to drought?
-We hypothesise that growth sensitivity....
+Here we use two known ENSO droughts, one in the dry season and another in the wet season, within a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand to test drivers of variation in the sensitivity of tropical tree growth to drought.
+<!-- We expect variation in drought sensitivity (growth in drought year/growth in a previous year with normal rainfall) across trees.  -->
+We ask: i) How much do species vary in drought sensitivity of tree growth? What drives this variation ii) What drives variation among individuals in the sensitivity of growth to drought? 
+We hypothesise that species and individuals have different sensitivities to wet and dry season droughts, based on their traits and microenvironments. 
+Specifically, we hypothesise that i) species growth sensitivities to wet and dry season droughts are influenced by deciduousness; deciduous species are negatively affected by wet season droughts because of shorter growing seasons.
+ii) the influence of drivers on drought sensitivity are distinct in wet and dry season droughts; canopy exposure could have stronger influence during dry season droughts while water availability has a stronger influence during wet season droughts.
 

### 2025-02-19 scripts/06_figures.R
cosmetic changes to Fig 4 alternate

@@ -617,14 +617,14 @@ coefs_all_sp <- ggplot(
         # col = factor(param, levels = c("b_calcDBH_min1_scaled_sp", "b_twi_scaled_sp", "b_cii_min1_scaled_sp"))
     )
 ) +
-    geom_point() +
+    geom_point(size = 5) +
     scale_x_discrete(labels = par_names) +
     # make error bars with narrow heads
     geom_errorbar(aes(
         ymin = lwr, ymax = upr # ,
         # col = factor(signif, levels = c("neg", "pos", "no"))
         # col = factor(param, levels = c("b_calcDBH_min1_scaled_sp", "b_twi_scaled_sp", "b_cii_min1_scaled_sp"))
-    ), width = 0.1) +
+    ), linewidth = 1.5, width = 0.05) +
     # scale_color_manual(values = c("red", "blue", "grey40"), drop = FALSE) +
     # scale_color_manual(values = rep(colours, 2), drop = FALSE) +
     geom_hline(yintercept = 0, linetype = "dashed") +
@@ -678,9 +678,11 @@ twi_slopes_plot <- ggplot(
     aes(x = williams_dec, y = total_effect),
     order = median
 ) +
-    geom_point() +
+    geom_point(size = 3, alpha = 0.7) +
     # geom_smooth(method = "lm", col = "grey40") +
-    geom_errorbar(aes(ymin = total_lwr, ymax = total_upr), width = 0.1) +
+    geom_errorbar(aes(ymin = total_lwr, ymax = total_upr),
+        width = 0.005, linewidth = 1, alpha = 0.7
+    ) +
     geom_hline(yintercept = 0, linetype = "dashed") +
     facet_wrap(~ factor(yr, levels = c(2010, 2015)),
         scales = "free", ncol = 2
@@ -725,7 +727,7 @@ str(preds)
 
 library(patchwork)
 png("doc/display/Fig4_alternate.png", width = 16, height = 8, units = "in", res = 300)
-(coefs_all_sp / twi_slopes_plot) | pred_plot + plot_annotation(tag_levels = "a") #+ plot_layout(widths = c(1, 1.5))
+(coefs_all_sp / twi_slopes_plot) | pred_plot # & plot_annotation(tag_levels = "a") & plot_layout(widths = c(1, 1.5))
 dev.off()
 
 # fig 5-----------------------------------

### 2025-02-19 scripts/03_negativevalues.R
script formatting

@@ -1,132 +1,144 @@
-#assessing and understanding the negative values in the dataset
+# assessing and understanding the negative values in the dataset
 
 # Load required libraries---------------------
 library(tidyverse)
 library(brms)
 
 # load data--------------------------
-rm(list=ls())
-datasets<-readRDS("data/HKK-dendro/sensitivity_data.RData")
+rm(list = ls())
+datasets <- readRDS("data/HKK-dendro/sensitivity_data.RData")
 
-tree.time<-datasets$tree.time
-tree_vars<-datasets$tree_vars
-sp_vars<-datasets$sp_vars
+tree.time <- datasets$tree.time
+tree_vars <- datasets$tree_vars
+sp_vars <- datasets$sp_vars
 
-colnames(sp_vars)[1]<-"Species"
+colnames(sp_vars)[1] <- "Species"
 
-#merge tree_vars and sp_vars
-tree_vars<-merge(tree_vars, sp_vars, by="Species", all.x=TRUE)
-#merge tree_vars and tree.time
-tree.time<-merge(tree.time, tree_vars, by="Tag", all.x=TRUE)
+# merge tree_vars and sp_vars
+tree_vars <- merge(tree_vars, sp_vars, by = "Species", all.x = TRUE)
+# merge tree_vars and tree.time
+tree.time <- merge(tree.time, tree_vars, by = "Tag", all.x = TRUE)
 
 colnames(tree.time)
 
-#standardise the variables within species and across all--------------------
+# standardise the variables within species and across all--------------------
 
-tree.time<-tree.time %>%
-dplyr::mutate(calcDBH_min1_scaled = scale(calcDBH_min1, center = TRUE, scale = TRUE),
-cii_min1_scaled = scale(cii_min1, center = TRUE, scale = TRUE),
-cii_min1_scaled = ifelse(cii_min1_scaled=="NaN", 0, cii_min1_scaled),
-twi_scaled = scale(twi, center = TRUE, scale = TRUE),
-median_inc_scaled = scale(median_inc, center = TRUE, scale=TRUE))%>%
-group_by(Species) %>%
-#scale while retaining original values
-dplyr::mutate(calcDBH_min1_scaled_sp = scale(calcDBH_min1, center = TRUE, scale = TRUE),
-cii_min1_scaled_sp = scale(cii_min1, center = TRUE, scale = TRUE),
-cii_min1_scaled_sp = ifelse(cii_min1_scaled_sp=="NaN", 0, cii_min1_scaled_sp),
[…]

### 2025-02-19 scripts/09_orderedcii.R
trying models with cii as ordered fac #18

@@ -1,0 +1,106 @@
+# models considering cii as an ordered factor
+
+# Load required libraries---------------------
+library(tidyverse)
+library(brms)
+
+# load data--------------------------
+rm(list = ls())
+datasets <- readRDS("data/HKK-dendro/sensitivity_data.RData")
+
+tree.time <- datasets$tree.time
+tree_vars <- datasets$tree_vars
+sp_vars <- datasets$sp_vars
+
+colnames(sp_vars)[1] <- "Species"
+
+# merge tree_vars and sp_vars
+tree_vars <- merge(tree_vars, sp_vars, by = "Species", all.x = TRUE)
+# merge tree_vars and tree.time
+tree.time <- merge(tree.time, tree_vars, by = "Tag", all.x = TRUE)
+
+colnames(tree.time)
+
+colours <- c("#e15f41", "#546de5", "#f7b731")
+
+# standardise the variables within species and across all--------------------
+yrs <- c(2010, 2015)
+
+tree.time <- tree.time %>%
+    dplyr::mutate(
+        calcDBH_min1_scaled = scale(calcDBH_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled = scale(cii_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled = ifelse(cii_min1_scaled == "NaN", 0, cii_min1_scaled),
+        twi_scaled = scale(twi, center = TRUE, scale = TRUE),
+        median_inc_scaled = scale(median_inc, center = TRUE, scale = TRUE),
+        avg_inc_tree_scaled = scale(avg_inc_tree, center = TRUE, scale = TRUE)
+    ) %>%
+    group_by(Species) %>%
+    # scale while retaining original values
+    dplyr::mutate(
+        calcDBH_min1_scaled_sp = scale(calcDBH_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled_sp = scale(cii_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled_sp = ifelse(cii_min1_scaled_sp == "NaN", 0, cii_min1_scaled_sp),
+        twi_scaled_sp = scale(twi, center = TRUE, scale = TRUE)
+    ) %>%
+    ungroup() %>%
+    # remove large outliers for each year
+    group_by(yr) %>%
[…]

### 2025-02-21 doc/manuscript.Rmd
manuscript edits to statistical models

@@
 
-To further disentange the drivers of stem-level drought sensitivity, we modelled:
+To further disentange the drivers of stem-level drought sensitivity, we simultaneously modelled:
 
@@
 $$ 
+and 
+$$
+CII \sim DBH
+$$
 
+Simultaneously modelling the effect of variables on sensitivity and the effect of DBH on CII in this way allows us to disentangle causual linkages between these processes. 
 We chose to model a species random effect on all slopes here to account for variation across species instead of specific traits which may have different effects with each of the variables. 
+However, we also ran simpler models with species random effects only on the intercept, the results of which are reported in the Supplementary Material. 
+Across all models, we used Gaussian priors for the distribution of the response variable sensitivity, considered CII as an monotonic predictor (ordered factor) and centred and scaled the other predictors.
+We modelled CII as an ordinal categorical variable with a cumulative logit link function. 
+In the combined form, the model therefore used a multivariate distribution of Gaussian and logit. 
+Finally, to test full mediation effects, or the hypothesis that DBH effects on sensitivity are fully explained by its effects on exposure, we tested the same models, removing the DBH as a predictor of sensitivity.
+We then compared the fits of the partial mediation and full mediation models using Leave-One-Out Cross Validation.
 
@@
 Mixed models were fit through Bayesian regressions using MCMC methods using the package _brms_ [REF].
-We ran models using gaussian priors for the data distribution.
 We ran 4 MCMC chains for 3000 iterations with 1000 of these set as warmup.
@@
 
+![***Figure 4: Drivers of variation in sensitivity among individuals***](display/Fig4_alternate.png)
+
 ## Variation among individuals {-}
 
-![***Figure 4: Drivers of variation in sensitivity among individuals***](display/Fig4.png)
+![***Figure 5: Effects of size, exposure and water availability***](display/Fig5.png)
 

### 2025-02-21 scripts/06_figures.R
some alternate figure options #15

@@ -793,3 +793,93 @@ coefs_plot_parmed <- ggplot(coefs_parmed %>% filter(param %in% pars.keep), aes(x
 png("doc/display/Fig5.png", width = 12, height = 8, units = "in", res = 300)
 (coefs_plot_parmed + parmed_gg) / (coefs_plot_fullmed + fullmed_gg)
 dev.off()
+
+# plot conditional effects of cii
+
+# https://discourse.mc-stan.org/t/conditional-effects-plot-for-monotonic-predictors-with-logit-link/35850/2
+
+# read fits
+fits <- readRDS("results/models/orderedcii/fits_partialmed_rel.rds")
+
+library(tidybayes)
+get_variables(fits[[1]])
+
+cii_fit1 <- fits[[1]] %>%
+    spread_draws(
+        b_sensprop_Intercept, bsp_sensprop_mocii_min1,
+        simo_sensprop_mocii_min11[i]
+    ) %>%
+    dplyr::mutate(yr = 2010)
+
+cii_fit2 <- fits[[2]] %>%
+    spread_draws(
+        b_sensprop_Intercept, bsp_sensprop_mocii_min1,
+        simo_sensprop_mocii_min11[i]
+    ) %>%
+    dplyr::mutate(yr = 2015)
+
+cii_fit <- bind_rows(cii_fit1, cii_fit2)
+
+cii_fit <- cii_fit %>%
+    group_by(yr) %>%
+    dplyr::mutate(
+        # D is equal to number of categories minus 1
+        D = length(unique(i) - 1)
+    ) %>%
+    group_by(.chain, .iteration) %>%
+    # add a row within each of these groups
+
+    dplyr::mutate(
+        cumsumi = cumsum(simo_sensprop_mocii_min11),
+        post_mu = b_sensprop_Intercept + (bsp_sensprop_mocii_min1 * D * cumsumi)
+    )
+
+# add a row per chain and iteration with i = 0
+cii_fit_add <- cii_fit %>%
+    group_by(yr, .chain, .iteration, .draw) %>%
+    dplyr::summarise(
+        b_sensprop_Intercept = b_sensprop_Intercept[1],
+        bsp_sensprop_mocii_min1 = bsp_sensprop_mocii_min1[1],
+        i = 0,
+        simo_sensprop_mocii_min11 = 0,
+        yr = yr[1],
+        cumsumi = 0,
[…]

### 2025-03-03 scripts/02_models.R
old changes

@@ -755,7 +755,16 @@ for (i in 1:length(yrs)) {
 
 saveRDS(fits, "results/models/non_negative/fits_treeisatree.RDS")
 
+fits <- readRDS("results/models/non_negative/fits_treeisatree.RDS")
 summary(fits[[1]])
+
+# conditional effects
+condeff <- conditional_effects(fits[[1]])
+str(condeff)
+
+plot(condeff, ask = F)
+
+
 
 png("results/plots/non_negative/fits_treeisatree.png")
 plot(fits[[1]], variable = "^b", regex = T)

### 2025-03-03 scripts/06_figures.R

@@ -424,7 +424,8 @@ coefs_dec_lms <- ranef_df %>%
     filter(yr %in% yrs) %>%
     filter(Species != "ALPHVE") %>%
     nest_by(yr) %>%
-    dplyr::mutate(mod = list(lm(median ~ williams_dec, data = data))) %>%
+    # dplyr::mutate(mod = list(lm(intercept ~ williams_dec, data = data))) %>%
+    dplyr::mutate(mod = list(cor.test(data$williams_dec, data$intercept))) %>%
     dplyr::reframe(broom::tidy(mod))
 
 coefs_dec_lms
@@ -441,7 +442,10 @@ dec_intercept_plot <- ggplot(ranef_df, aes(x = williams_dec, y = intercept, ymin
         ncol = 2
     ) +
     # add text with p value
-    geom_text(data = coefs_dec_lms %>% filter(term == "williams_dec"), inherit.aes = F, aes(x = c(3, 3), y = c(1.5, 1), label = paste("p = ", round(p.value, 2)), hjust = 0, vjust = 0)) +
+    geom_text(
+        data = coefs_dec_lms %>% filter(term == "williams_dec"), inherit.aes = F,
+        aes(x = c(3, 3), y = c(1, 1), label = paste("p = ", round(p.value, 2)), hjust = 0, vjust = 0)
+    ) +
     labs(x = "Deciduousness", y = "Predicted sensitivity") +
     guides(color = "none") +
     theme_bw() +

### 2025-03-03 doc/manuscript.Rmd
title options, basic changes to format
for Ecology Letters

@@
 bibliography: growth-precip-thailand.bib
-csl: new-phytologist.csl
+# csl: new-phytologist.csl
+csl: ecology-letters.csl 
 ---
@@
 
-**Title:** Sensitivity of tree growth to drought in seasonally dry tropical forests using long-term dendrometer band measurements
+**Title 1:** Sensitivity of tree growth to drought in seasonally dry tropical forests using long-term dendrometer band measurements
+
+**Title 2:** Contrasting effects of species and environmental predictors on tree growth sensitivity to different droughts in a seasonally dry tropical forest
+
+**Title 3:** Diversity of species- and environment-mediated drought responses in a seasonally dry tropical forest
+
+**Title 4:** Long-term dendrometer band measurements reveal diverse species- and environment-mediated drought responses in a seasonally dry tropical forest
+
+**Title 5:** Deciduousness may exacerbate/worsen/amplify tree growth sensitivity during wet season droughts in seasonally dry tropical forests
+
+**Title 6:**
 
@@
 
+**Running title:**
+
+**Article Type:** Letter
+
+**Number of Words in Abstract:**  
+**Number of Words in Main Text:**  
+**Number of References:**  
+**Number of Figures:**  
+**Number of Tables:**  
+**Number of Text boxes:**
+
+**Corresponding author address:**
+
+# Keywords {-}
+
+# Statement of Authorship {-}
+
+# Data Accessibility Statement {-}
+
 \newpage
 
-# Summary {-}
+<!-- Ecology Letters format 
+-->
+
+<!-- # Summary {-} -->
 <!-- 
@@
 - Main conclusion   -->
+# Abstract (150 words) {-}
 
@@
 
-# Keywords {-}
-
+<!-- # Keywords {-} -->
 \newpage
@@
 
-Tropical tree woody growth is a key component of aboveground productivity and affects the global carbon carbon cycle, but its sensitivity to climate anomalies is poorly understood.  
+<!-- Tropical tree woody growth is a key component of aboveground productivity and affects the global carbon carbon cycle, but its sensitivity to climate anomalies is poorly understood.  
 - Woody growth has long-lasting effects on the terrestrial carbon pool but is sensitive to interannual variation in temperature and precipitation.  
@@
 -High Crown exposure makes trees more vulnerable to drought, but drought deciduous habit or perennial water access allows them to escape this  
--Disentangling the effects of tree size, crown exposure, water access, and deciduousness on drought sensitivity
+-Disentangling the effects of tree size, crown exposure, water access, and deciduousness on drought sensitivity   -->
 
@@
 In the combined form, the model therefore used a multivariate distribution of Gaussian and logit. 
-Finally, to test full mediation effects, or the hypothesis that DBH effects on sensitivity are fully explained by its effects on exposure, we tested the same models, removing the DBH as a predictor of sensitivity.
-We then compared the fits of the partial mediation and full mediation models using Leave-One-Out Cross Validation.
+<!-- Finally, to test full mediation effects, or the hypothesis that DBH effects on sensitivity are fully explained by its effects on exposure, we tested the same models, removing the DBH as a predictor of sensitivity.
+We then compared the fits of the partial mediation and full mediation models using Leave-One-Out Cross Validation.   -->
 
@@
 tree.time.sp <- tree.time %>%
-  group_by(Species) %>%
+  group_by(spfull) %>%
   dplyr::summarise(
@@
   arrange(inc_annual_med)
+
+sens.sp <- tree.time %>%
+  filter(yr %in% c(2010, 2015)) %>%
+  group_by(spfull, yr) %>%
+  dplyr::summarise(
+    sens_med = round(median(sens.prop, na.rm = T), 2),
+    sens_mean = round(mean(sens.prop, na.rm = T), 2),
+    sens_sd = round(sd(sens.prop, na.rm = T), 2)
+  ) %>%
+  arrange(sens_med)
+
+# spread this for each year
+
+sens.sp.wide <- sens.sp %>%
+  ungroup() %>%
+  dplyr::select(spfull, yr, sens_med) %>%
+  pivot_wider(names_from = yr, values_from = sens_med, names_prefix = "sens.")
+
+# paired correlation of 2010 and 2015 sensitivities across species
+
+sens.sp.cor <- cor.test(sens.sp.wide$sens.2010, sens.sp.wide$sens.2015)
 ```
@@
 Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual, na.rm=T),2)` cm. 
-Median annual increment differed by species with `r tree.time.sp$Species[1]` having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` cm) and `r tree.time.sp$Species[nrow(tree.time.sp)]` the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` cm) among the `r nrow(tree.time.sp)` species analysed.
+Median annual increment differed by species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` cm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` cm) among the `r nrow(tree.time.sp)` species analysed.
 
@@
 
-Drought sensitivity across all trees also showed negative sensitivities for both droughts with considerable variation around it (Fig 2b).
+Trees on average had negative drought sensitivities for both droughts with considerable variation around it (Fig 2b). 
+Median drought sensitivity across all trees in 2010 was `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)` while in 2015 it was `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`.
+In 2010, *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* and *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/2]`* had the lowest and highest median sensitivity of `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` and `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/2]` respectively.
+In 2015, *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* and *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/2]`* had the lowest and highest median sensitivity of `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` and `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/2]` respectively.
 
@@
 
+```{r model summaries, echo=F, warning=F, message=F}
+# load data--------------------------
+datasets <- readRDS("../data/HKK-dendro/sensitivity_data.RData")
+
+tree.time <- datasets$tree.time
+tree_vars <- datasets$tree_vars
+sp_vars <- datasets$sp_vars
+
+colnames(sp_vars)[1] <- "Species"
+
+coefs_df <- readRDS("../results/models/non_negative/sensitivity_model_intercept.RData")
+ranef_df <- coefs_df %>% filter(grepl("r_Species", param))
+ranef_df$Species <- gsub("r_Species\\[|\\,Intercept\\]", "", ranef_df$param)
+
+# add intercept to ranef_df
+# make intercept vector
+intercepts <- coefs_df %>%
+  filter(param %in% "Intercept") %>%
+  dplyr::select(median) %>%
+  pull(median)
+intercepts <- rep(intercepts, each = nrow(ranef_df) / 2)
+ranef_df <- ranef_df %>%
+  dplyr::mutate(
+    intercept = intercepts + median,
+    lwr = intercepts + lwr,
+    upr = intercepts + upr
+  )
+
+# plot intercepts against species characteristics
+
+# join ranef_df with sp_vars
+ranef_df <- merge(ranef_df, sp_vars, by = "Species", all.x = TRUE)
+
+# lms
+coefs_dec_lms <- ranef_df %>%
+  # filter(yr %in% yrs) %>%
+  filter(Species != "ALPHVE") %>%
+  nest_by(yr) %>%
+  # dplyr::mutate(mod = list(lm(intercept ~ williams_dec, data = data))) %>%
+  dplyr::mutate(mod = list(cor.test(data$williams_dec, data$intercept))) %>%
+  dplyr::reframe(broom::tidy(mod))
+
+isocline_coefs <- readRDS("../results/models/non_negative/coefs_isoclines_nore.RDS")
+# isocline_coefs <- readRDS("results/models/non_negative/coefs_isoclines_nore.RDS")
+
+isocline_coefs <- do.call(rbind, isocline_coefs)
+
+isocline_coefs %>% filter(param == "b_williams_dec")
+
+# head(isocline_coefs)
+```
+
 ## Variation in sensitivity among species {-}
 
-![***Figure 4: Drivers of variation in sensitivity among individuals***](display/Fig4_alternate.png)
+Deciduousness was associated with lower negative sensitivities in 2015, but not in 2010 (Fig 3a). 
+Species median sensitivities in 2010 and 2015 were not correlated with each other (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`) but on average had lower mean and median sensitivities in 2015 than in 2010, in alignment with median sensitivities across all trees.
+In 2015, predicted sensitivities from intercept-only models of sensitivity with a species random effect on the intercept showed a significant negative association with deciduousness value (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`).
+On the other hand, in 2010, predicted sensitivities had no significant association with deciduousness (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`).
+
+Across all trees, predicted sensitivities in 2010 and 2015 was heterogenous with deciduousness and TWI (Fig 3b).
+
+Accounting for species differences in mean sensitivity, the effect of TWI on sensitivity was different in 2010 and 2015, with 2015 effects associated with deciduousness.
+
+![***Figure 4: Drivers of variation in sensitivity among individuals***](display/Fig4.png)
 
@@
 
-![***Figure 5: Effects of size, exposure and water availability***](display/Fig5.png)
+Crown exposure was associated with stronger negative sensitivities in both droughts.
+
+Accounting for effects of exposure, remaining effects of size on sensitivity was different between the two droughts.
+
+![***Figure 5: Effects of size, exposure and water availability***](display/Fig5_alternate2.png)
 

### 2025-03-03 scripts/06_figures.R
updated Fig 2 and 3 #12 #11 cosmetic

@@ -4,7 +4,7 @@
 library(tidyverse)
 library(lubridate)
 library(ggplot2)
-
+library(patchwork)
 
 # figure 1 ------------------------------------------------
 
@@ -376,7 +376,8 @@ spagplot_top10 <- ggplot() +
 spagplot_top10
 
 library(patchwork)
-png("doc/display/Fig2.png", width = 10, height = 6, units = "in", res = 300)
+# png("doc/display/Fig2.png", width = 10, height = 6, units = "in", res = 300)
+png("doc/display/Fig2.png", width = 10, height = 8, units = "in", res = 300)
 spagplot_top10 + sens.all + plot_annotation(tag_levels = "a") +
     # plot_layout(guides = "collect", widths = c(1, 2.2)) & theme(
     plot_layout(guides = "collect") & theme(
@@ -433,7 +434,10 @@ coefs_dec_lms
 # make plot with just deciduousness
 dec_intercept_plot <- ggplot(ranef_df, aes(x = williams_dec, y = intercept, ymin = lwr, ymax = upr)) +
     geom_pointrange() +
-    geom_smooth(method = "lm", col = "grey40") +
+    geom_smooth(
+        data = ranef_df %>% filter(yr == 2015),
+        method = "lm", col = "grey40"
+    ) +
     # geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.1) +
     geom_hline(yintercept = 0, linetype = "dashed") +
     # facet_grid(name ~ factor(yr, levels = c(2010, 2015)), scales = "free_x") +
@@ -443,8 +447,10 @@ dec_intercept_plot <- ggplot(ranef_df, aes(x = williams_dec, y = intercept, ymin
     ) +
     # add text with p value
     geom_text(
-        data = coefs_dec_lms %>% filter(term == "williams_dec"), inherit.aes = F,
-        aes(x = c(3, 3), y = c(1, 1), label = paste("p = ", round(p.value, 2)), hjust = 0, vjust = 0)
+        # data = coefs_dec_lms %>% filter(term == "williams_dec"),
+        data = coefs_dec_lms,
+        inherit.aes = F,
+        aes(x = c(2, 2), y = c(1, 1), label = paste("r = ", round(estimate, 2), " p = ", round(p.value, 2)), hjust = 0, vjust = 0)
     ) +
     labs(x = "Deciduousness", y = "Predicted sensitivity") +
     guides(color = "none") +

### 2025-03-03 doc/manuscript.Rmd
updated abstract and some more results

@@
 
-- Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but how climate anomalies might impact the growth of tropical tree growth is poorly understood. 
+<!-- - Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but how climate anomalies might impact the growth of tropical tree growth is poorly understood. 
 Standard methods like tree rings and plot inventories may either not represent most species or may not be accurate enough to capture variation in slow-growing species.  
@@
 - Large differences among species and individuals in their sensitivity to drought, as well as climatic differences differences among droughts themselves can explain heterogeneity in ecological responses to drought.
+-->
+
+Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but how climate anomalies like drought might impact the growth of tropical tree growth is poorly understood. 
+In seasonally dry tropical forests with strong seasonality, species evolutionary adaptations to hot dry seasons, like dry season deciduousness, or species distributions in relation to water availability might buffer growth sensitivity to drought.
+<!-- However, standard methods like tree rings and plot inventories may either not represent most species or may not be accurate enough to capture variation in slow-growing species.  -->
+We used a unique data record - manual dendrometer band measurements over 15 years on 1820 trees across 30 species in a seasonally dry tropical forest in <!---Huai Kha Khaeng, ---->Thailand, along with species and individual characteristics, using Bayesian regressions in a causal modelling framework to predict the sensitivity of tree growth to drought in two ENSO drought events with different characteristics - a wet-season drought and a dry season drought.
+We report a wide range of drought sensitivities across species and stems within a drought event, as well as contrasting responses across these two droughts. 
+Deciduous species were more sensitive than evergreen species in the wet-season drought, with this difference worsened in wetter microclimates. 
+Crown exposure was associated with stronger negative sensitivities in the dry season drought. 
+With observed regime shifts in precipitation across the global tropics and increasing incidence of climate anomalies, trees within seasonally dry tropical forests might experience divergent responses.
 
@@
 
--High Crown exposure makes trees more vulnerable to drought, but drought deciduous habit or perennial water access allows them to escape this  
+<!---High Crown exposure makes trees more vulnerable to drought, but drought deciduous habit or perennial water access allows them to escape this  
 -Disentangling the effects of tree size, crown exposure, water access, and deciduousness on drought sensitivity   -->
@@
 isocline_coefs <- do.call(rbind, isocline_coefs)
+isocline_coefs_dec <- isocline_coefs %>% filter(param == "b_williams_dec")
+isocline_coefs_twi <- isocline_coefs %>% filter(param == "b_twi")
+isocline_coefs_inter <- isocline_coefs %>% filter(param == "b_twi:williams_dec")
+# isocline_coefs_twi
+# head(isocline_coefs)
 
-isocline_coefs %>% filter(param == "b_williams_dec")
-
-# head(isocline_coefs)
+coefs_spre <- readRDS("../results/models/orderedcii/coefs_rel_spre.RDS")
+# coefs_spre <- readRDS("results/models/orderedcii/coefs_rel_spre.RDS")
+coefs_df_spre <- do.call(rbind, coefs_spre)
+coefs_spre_twi <- coefs_df_spre %>% filter(param == "b_sensprop_twi_scaled")
+# head(coefs_spre_twi)
 ```
@@
 On the other hand, in 2010, predicted sensitivities had no significant association with deciduousness (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`).
+Moreover, models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`) but not in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).
 
-Across all trees, predicted sensitivities in 2010 and 2015 was heterogenous with deciduousness and TWI (Fig 3b).
+The effect of TWI on sensitivity was different in 2010 and 2015, with 2015 effects interacting with deciduousness. 
+Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`).
+The interaction between TWI and deciduousness was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) but was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`).
+Combined, predicted sensitivities of deciduous trees was more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig 3b).
 
-Accounting for species differences in mean sensitivity, the effect of TWI on sensitivity was different in 2010 and 2015, with 2015 effects associated with deciduousness.
+The opposite effects of TWI on sensitivity in the wet and dry season drought, and the influence of deciduousness was also confirmed by models that controlled for crown exposure and size as well as species variation in intercept and slopes (Fig 4a).
+These models showed small negative sensitivities in wetter sites in the dry season drought (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but strong positive effect of wetness in the wet season drought (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`).
+The positive effect of TWI in 2015 was, however, mainly due to the response of evergreen species; deciduousness had a significant negative correlation with the slopes of the TWI effect in 2015 (Fig 4b).
+Predictions across the range of observed variables also showed that deciduous species generally had larger negative sensitivities in 2015, which did not show increase with wetness (Fig 4c). 
+On the other hand, neither the mean sensitivities across species nor their response to TWI was associated with deciduousness in 2010.
 
-![***Figure 4: Drivers of variation in sensitivity among individuals***](display/Fig4.png)
+![***Figure 4: Drivers of variation in sensitivity among individuals***. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/13](display/Fig4.png)
 
@@
 
-![***Figure 5: Effects of size, exposure and water availability***](display/Fig5_alternate2.png)
+![***Figure 5: Effects of size, exposure and water availability*** For comments - https://github.com/forestgeo/growth-precip-thailand/issues/14](display/Fig5_alternate2.png)
 

### 2025-03-04 scripts/06_figures.R
updated Fig 5 with some cosmetic changes #15

@@ -896,7 +896,8 @@ p_manual_ce <- ggplot(data = cii_fit %>% filter(i < 5), aes(y = post_mu, x = fac
     geom_hline(yintercept = 0, linetype = "dashed") +
     facet_wrap(~yr) +
     labs(x = "CII", y = "Sensitivity") +
-    theme_bw()
+    theme_bw() +
+    theme(strip.background = element_blank())
 
 png("doc/display/Fig5_alternate.png", width = 12, height = 8, units = "in", res = 300)
 (coefs_plot_parmed + parmed_gg) / p_manual_ce
@@ -911,7 +912,7 @@ dag_2015 <- magick::image_read("doc/display/dag_2015_vals.png")
 dag_2010_gg <- magick::image_ggplot(dag_2010, interpolate = F)
 dag_2015_gg <- magick::image_ggplot(dag_2015, interpolate = F)
 
-png("doc/display/Fig5_alternate2.png", width = 8, height = 8, units = "in", res = 300)
+png("doc/display/Fig5.png", width = 8, height = 8, units = "in", res = 300)
 (dag_2010_gg + dag_2015_gg) / p_manual_ce
 dev.off()
 

### 2025-03-04 doc/manuscript.Rmd
results draft

@@
 
+```{r exposure models, echo=F, warning=F, message=F}
+coefs_spre_cii <- coefs_df_spre %>% filter(param == "bsp_sensprop_mocii_min1")
+coefs_spre_ordcii <- coefs_df_spre %>% filter(param == "b_orderedciimin1_calcDBH_min1_scaled")
+```
+
 ## Variation among individuals {-}
@@
 Crown exposure was associated with stronger negative sensitivities in both droughts.
+In the same models with species random effects on intercept and slope, the simplex parameter for the ordered factor crown exposure in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 the negative effect was not as strong (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`).
+Predicted effects of crown exposure on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all crown exposure categories had negative sensitivities in 2015 (Fig 5b).
 
-Accounting for effects of exposure, remaining effects of size on sensitivity was different between the two droughts.
+Accounting for effects of exposure, remaining effects of size on sensitivity differed between the two droughts. 
+DBH was a significant predictor of exposure category, with larger DBH associated with higher exposure categories in 2010 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`) and 2015 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) (Fig 5a).
+Controlling for this direct effect of exposure, the effect of exposure, the remaining effect of size *per se* was significantly positive in 2010, while it was weakly negative in 2015 (Fig 4a, Fig 5a).
 
-![***Figure 5: Effects of size, exposure and water availability*** For comments - https://github.com/forestgeo/growth-precip-thailand/issues/14](display/Fig5_alternate2.png)
+![***Figure 5: Effects of size, exposure and water availability*** For comments - https://github.com/forestgeo/growth-precip-thailand/issues/14](display/Fig5.png)
 

### 2025-03-05 doc/manuscript_hkk_drought_sensitivity.Rmd
renamed manuscript file

--- 2025-03-04 doc/manuscript.Rmd
@@
 
-**Title 1:** Sensitivity of tree growth to drought in seasonally dry tropical forests using long-term dendrometer band measurements
+**Title 1:** Sensitivity of tree growth to drought in a seasonally dry tropical forest using long-term dendrometer band measurements
 
-**Title 2:** Contrasting effects of species and environmental predictors on tree growth sensitivity to different droughts in a seasonally dry tropical forest
+**Title 2:** Contrasting effects of species and environmental predictors on tree growth sensitivity to different/two droughts in a seasonally dry tropical forest
 
@@
 **Title 5:** Deciduousness may exacerbate/worsen/amplify tree growth sensitivity during wet season droughts in seasonally dry tropical forests
-
-**Title 6:**
 
@@
 Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but how climate anomalies like drought might impact the growth of tropical tree growth is poorly understood. 
-In seasonally dry tropical forests with strong seasonality, species evolutionary adaptations to hot dry seasons, like dry season deciduousness, or species distributions in relation to water availability might buffer growth sensitivity to drought.
+In seasonally dry tropical forests, species evolutionary adaptations to hot dry seasons, like dry season deciduousness, or species distributions in relation to water availability might buffer growth sensitivity to drought.
 <!-- However, standard methods like tree rings and plot inventories may either not represent most species or may not be accurate enough to capture variation in slow-growing species.  -->
-We used a unique data record - manual dendrometer band measurements over 15 years on 1820 trees across 30 species in a seasonally dry tropical forest in <!---Huai Kha Khaeng, ---->Thailand, along with species and individual characteristics, using Bayesian regressions in a causal modelling framework to predict the sensitivity of tree growth to drought in two ENSO drought events with different characteristics - a wet-season drought and a dry season drought.
-We report a wide range of drought sensitivities across species and stems within a drought event, as well as contrasting responses across these two droughts. 
+We used a unique data record - manual dendrometer band measurements over 15 years on 1820 trees across 30 species in a seasonally dry tropical forest in <!---Huai Kha Khaeng, ---->Thailand, along with species and individual characteristics, using Bayesian regressions in a causal modelling framework to predict the sensitivity of tree growth to drought in two ENSO drought events with different characteristics - a wet-/dry-season drought.
+Species and stems exhibited a wide range of drought sensitivities within a drought event, as well as contrasting responses across these two droughts. 
 Deciduous species were more sensitive than evergreen species in the wet-season drought, with this difference worsened in wetter microclimates. 
@@
 
-<!-- Tropical tree woody growth is a key component of aboveground productivity and affects the global carbon carbon cycle, but its sensitivity to climate anomalies is poorly understood.  
-- Woody growth has long-lasting effects on the terrestrial carbon pool but is sensitive to interannual variation in temperature and precipitation.  
-- Extreme events like drought, with temperature and precipitation outside normal ranges that can lead to growth reductions that varies across species and individuals [@bennett_larger_2015; @mcgregor_tree_2021].
-- the extent of this variation in drought responses is poorly understood.  
-- with changing climate patterns across the global tropics, including novel climatic regimes [@dahinden_future_2017], there is a need to understand the drivers of this variation, to be able to predict and respond to forest responses to drought.
+Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but it may be vulnerable to climate anomalies like drought.  
+- Woody growth has one of the longest residence times in terrestrial carbon pool, contributing to long-lasting impacts on carbon removal from the atmosphere [@yang_variations_2021; @muller-landau_patterns_2021].
+- Climate anomalies like drought can lead to stand-level growth reductions, along with increased mortality [@bennett_sensitivity_2023; @bauman_tropical_2022].  
+<!-- that varies across species and individuals [@bennett_larger_2015; @mcgregor_tree_2021].   -->
+- Seasonally dry tropical forests are especially vulnerable to ecological impacts of drought, being at the edge of plant thermal and water stress tolerance [@allen_will_2017].
+- Within a stand, drought sensitivity varies among species and individuals, but the extent and drivers of this variation in drought responses is poorly understood.  
+- With changing climate patterns across the global tropics, emergence of novel climatic regimes like "very extreme droughts" in recent decades [@dahinden_future_2017; @luo_tropical_2022], there is an urgent need to understand the drivers of this variation, to be able to predict and respond to forest drought responses.
 
-Growth reductions from droughts are difficult to estimate reliably from whole-plot inventories.  
+Seasonally dry tropical forests can be more vulnerable to droughts, despite trees being adapted to strong dry seasons.  
+- Trees in seasonally dry forests are routinely exposed to hot, dry conditions in the dry season and may have evolutionary strategies like dry-season deciduousness and drought deciduousness for survival.
+- However, across the global tropics, drier forests show stronger sensitivities to drought-associated water stress  [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019].
+- One potential reason - thermal/water stress tolerance thresholds crossed [@allen_will_2017].  
+- Another reason: drought events are climatically different in timing and intensity, with trees exposed to drought conditions in different parts of their seasonal phenologies.  
+- Morevoer, besides broad patterns at the stand level, species- and individual-level heterogeneity can lead to divergent responses with consequences for stand composition.   
+- Some of the discrepancies between expectations from trees and observations at the stand-level may be resolved by considering factors that determine drought sensitivity across species and individuals
+
+Trees in seasonally dry forests have strategies that span and interact across timescales to maximise growth and survival under hot/dry conditions.  
+- evolutionary timescales - e.g. dry season deciduousness   
+- Deciduous species, with shorter duration with leaves can have acquisitive strategies during leaf on [@desouza_drought_2020].    
+- ecological timescales - e.g. distribution of trees in relation to wetness  
+-  Species distribution of evergreen species covarying with soil moisture [@kunert_leaf_2021] and potentially affecting survival and growth.     
+- ecophysiological - e.g. negative TLPs, plasticity in deciduousness (root/shoot)
+- Species with drought tolerant traits like deep roots, more negative turgor loss point and ...
+can be more or less resistant to drought [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021].
+- Larger trees suffer more growth declines during drought [@bennett_larger_2015], exposed crowns -> more thermal stress [@vinod_thermal_2023] although crown exposure -> more light -> more growth in normal years.
+- Some of these adaptations may confer drought sensitivity and buffer stem growth declines during drought years, or may tradeoff with advantages in normal years. 
+- To understand drought sensitivity representative of stand diversity, trees across these strategies need to be analysed.
+<!-- - Drought events can be different from each other, but generally lead to low soil moisture, low water tables, low vapour pressure and high temperature [@chitra-tarak_hydraulicallyvulnerable_2021].
+-->
+<!-- - Results in thermal stress on tissues, evaporative loss, leaf turgor loss and cavitation.  
+- Species can be conservative or acquisitive in water and nutrient uptake; differential allocation to growth/survival.   -->
+
+Drought sensitivity representative of stand diversity in seasonally dry tropical forests is difficult to assess using standard methods.  
+- Annual tree rings - found in SDTFs but not many species,  
+- Species with rings represent specific strategy to water use and growth - and often affiliated to drier microenvironments.  
+- Tape measurements - more representative, but may not be annual because of effort  
+- Even when tape measures are annual, not enough precision; error for each tape measure is 1mm, error for growth is at least 2mm.
+(Median annual growth across trees in HKK is ~1.4 mm, deviation from annual growth needs more precision).
+- In sum, these methods may only be capturing the most dramatic effects - the most severe droughts that affect ring-forming species or produce large-enough declines in fast-growing individuals.  
+- Methods need to capture species strategies, distributions across microenvironments, and be precise enough.  
+<!---Growth reductions from droughts are difficult to estimate reliably from whole-plot inventories.  
 - Droughts are different but generally leads to low soil moisture, low water tables, vapour pressure, high temperature.
@@
 - Reductions are often small (<1 mm), and difficult to detect with tape measurements.
-
-Tree species in seasonally dry forests have diverse allocation strategies and traits which can result in differential sensitivities to drought  
-- Drought events can be different from each other, but generally lead to low soil moisture, low water tables, low vapour pressure and high temperature [@chitra-tarak_hydraulicallyvulnerable_2021].  
-- Results in thermal stress on tissues, evaporative loss, leaf turgor loss and cavitation.  
-- Species can be conservative or acquisitive in water and nutrient uptake; differential allocation to growth/survival.  
-- Species with drought tolerant traits like deep roots, more negative turgor loss point and ... are more resistant to drought [@kunert_leaf_2021; @refs].  
-- Some evidence that both understory and emergent species can have reduced survival during drought [@machado_forest_2023].  
-- Deciduous species, with shorter duration with leaves can have acquisitive strategies during leaf on [@desouza_drought_2020].   
--  Strategies also affects species distribution with evergreen species covarying with soil moisture [@kunert_leaf_2021] and potentially affecting survival and growth.   
-- Many hydraulic-related traits vary with tree height [ @vinod_thermal_2023], including the frequency of dry season deciduous leaf loss -- both within species and at the community level [@condit_ref; @meakem].
 
@@
 
-Here we use two known ENSO droughts, one in the dry season and another in the wet season, within a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand to test drivers of variation in the sensitivity of tropical tree growth to drought.
+Here we examine individual tree growth during two known ENSO drought years, one in the dry season and another in the wet season, within a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand to disentangle drivers of variation in drought sensitivity.
 <!-- We expect variation in drought sensitivity (growth in drought year/growth in a previous year with normal rainfall) across trees.  -->
@@
 By scaling the difference in increment to the mean annual growth of each tree, this metric allows us to compare across slow and fast growing individuals and species.
+
+To account for potential biases in annual increments from dendrobands from water-related growth or shrinkage, we also calculated growth occurrence using a zero-growth assumption [@zweifel]. 
+Under this assumption, positive DBH change is scored as 1 and any change <=0 is scored as 0. 
+Timing of measurement and seasonal dynamics in tree hydration and bark dynamics could contribute to negative values in dendrometer band increments, unlike woody growth from tree ring datasets [@chitra-tarak].
+To account for biases introduced by timing of measurements and seasonal dynamics, we used this simplifying assumption to corroborate the patterns observed with the sensitivity metric.
 
@@
 
+Unordered points from Introduction
+- Many hydraulic-related traits vary with tree height [ @vinod_thermal_2023], including the frequency of dry season deciduous leaf loss -- both within species and at the community level [@condit_ref; @meakem].
+*caveat for species-level traits*
+
+*for where else might they be allocating to?*
+- Some evidence that both understory and emergent species can have reduced survival during drought [@machado_forest_2023].  
+- ENSO droguht changed regeneration dynamics in Thailand [@nutiprapun_effects_2023]
+- deciduous species have higher survival under multi-year droughts [@aguirre-guiterrez refs]
+
+- (BCI light x size interaction in @rüger_growth_2011)
+
+- Afzelia xylocarpa, one of the tree ring species had the most distinct (positive) sensitivity in the dry season drought *on how representative sampling is important*
+
 # Acknowledgements {-}

### 2025-03-19 doc/manuscript_hkk_drought_sensitivity.Rmd
manuscript with refined hypotheses and abstract #20
@teixeirak - updated manuscript with info from this issue

@@
 
-**Title 1:** Sensitivity of tree growth to drought in a seasonally dry tropical forest using long-term dendrometer band measurements
+**Title:** Tropical tree drought sensitivity is jointly shaped by species adaptations, individual microenvironments, and drought characteristics
+
+<!-- **Title 1:** Sensitivity of tree growth to drought in a seasonally dry tropical forest using long-term dendrometer band measurements
 
@@
 
-**Title 5:** Deciduousness may exacerbate/worsen/amplify tree growth sensitivity during wet season droughts in seasonally dry tropical forests
+**Title 5:** Deciduousness may exacerbate/worsen/amplify tree growth sensitivity during wet season droughts in seasonally dry tropical forests -->
 
@@
 
-Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but how climate anomalies like drought might impact the growth of tropical tree growth is poorly understood. 
-In seasonally dry tropical forests, species evolutionary adaptations to hot dry seasons, like dry season deciduousness, or species distributions in relation to water availability might buffer growth sensitivity to drought.
-<!-- However, standard methods like tree rings and plot inventories may either not represent most species or may not be accurate enough to capture variation in slow-growing species.  -->
-We used a unique data record - manual dendrometer band measurements over 15 years on 1820 trees across 30 species in a seasonally dry tropical forest in <!---Huai Kha Khaeng, ---->Thailand, along with species and individual characteristics, using Bayesian regressions in a causal modelling framework to predict the sensitivity of tree growth to drought in two ENSO drought events with different characteristics - a wet-/dry-season drought.
-Species and stems exhibited a wide range of drought sensitivities within a drought event, as well as contrasting responses across these two droughts. 
+<!-- Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but how climate anomalies like drought might impact the growth of tropical tree growth is poorly understood.
+-->
+The frequency and intensity of droughts are increasing across tropical forests, altering tree demography and its contribution to the land carbon sink. 
+In seasonally dry forests, species adaptations to strong dry seasons might confer drought resistance, but this resistance may be variable with individual microenvironments and across drought events. 
+We analysed a unique data record - manual dendrometer band measurements over 15 years on 1820 trees across 30 species in a seasonally dry tropical forest in <!---Huai Kha Khaeng, ---->Thailand, <!--using Bayesian regressions in a causal modelling framework---> to predict the drought sensitivity of tree growth in two climatically distinct droughts.
+Species and stems exhibited a wide range of drought sensitivities within a drought event, and inconsistent responses across these two droughts. 
 Deciduous species were more sensitive than evergreen species in the wet-season drought, with this difference worsened in wetter microclimates. 
-Crown exposure was associated with stronger negative sensitivities in the dry season drought. 
-With observed regime shifts in precipitation across the global tropics and increasing incidence of climate anomalies, trees within seasonally dry tropical forests might experience divergent responses.
+Although exposed trees were more sensitive, access to water through larger size or topographic location buffered growth declines. 
+Heterogenous responses of species and individuals to droughts indicate potential spatiotemporal insurance effects in diverse forests in the face of increased climate anomalies.
 
@@
 
-Seasonally dry tropical forests can be more vulnerable to droughts, despite trees being adapted to strong dry seasons.  
+Seasonally dry tropical forests can be vulnerable to droughts, despite trees being adapted to strong dry seasons.  
 - Trees in seasonally dry forests are routinely exposed to hot, dry conditions in the dry season and may have evolutionary strategies like dry-season deciduousness and drought deciduousness for survival.
@@
 - indeed, there is evidence that trees near streams undergo greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta] during drought
-- rather than necessarily helping during drought, water access will shape the size and traits of species living in habitat, with stream habitats tending to have larger trees [@ref] and more evergreen trees, and also their average growth rate
+- rather than necessarily helping during drought, water access will shape the size and traits of species living in habitat, with stream habitats tending to have larger trees [@ref] and more evergreen trees, and also their average growth rate--->
 
+```{r table, echo=FALSE, warning=FALSE, message=FALSE}
+# figure 0-----------------------------
+
+# make a figure with summaries
+
+library(flextable)
+set_flextable_defaults(font.family = "Arial", font.size = 9)
+
+sumtab <- read.csv("display/summary_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+
+ft <- flextable(sumtab)
+ft <- set_header_labels(ft,
+  na = ""
+)
+# Merge the title rows across all columns
+# ft <- merge_at(ft, i = c(1, 3, 6), j = 1:3, part = "body")
+# # Add text to the merged row
+# ft <- compose(ft, i = c(1, 3, 6), j = 1, value = as_paragraph(ft[1, 1], ft[3, 1], ft[6, 1]))
+ft <- hline(ft, i = c(1, 4, 8), part = "body")
+ft <- line_spacing(ft, space = 1, part = "all")
+ft <- autofit(ft)
+ft <- bold(ft, i = c(1, 4, 8), bold = TRUE, part = "body")
+# ft <- width(ft, 1, width = 1.75)
+# ft <- width(ft, 2, width = 1.75)
+# ft <- width(ft, 3, width = 3)
+ft
+```
 <!---KAT framing---->
@@
 
-![***Figure 1.
-Climatic characteristics of drought years and average years*** in Huai Kha Khaeng ForestGEO plot.
-For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1.png)
-
-Here we examine individual tree growth during two known ENSO drought years, one in the dry season and another in the wet season, within a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand to disentangle drivers of variation in drought sensitivity.
-<!-- We expect variation in drought sensitivity (growth in drought year/growth in a previous year with normal rainfall) across trees.  -->
-We ask: i) How much do species vary in drought sensitivity of tree growth? What drives this variation ii) What drives variation among individuals in the sensitivity of growth to drought? 
-We hypothesise that species and individuals have different sensitivities to wet and dry season droughts, based on their traits and microenvironments. 
-Specifically, we hypothesise that i) species growth sensitivities to wet and dry season droughts are influenced by deciduousness; deciduous species are negatively affected by wet season droughts because of shorter growing seasons.
-ii) the influence of drivers on drought sensitivity are distinct in wet and dry season droughts; canopy exposure could have stronger influence during dry season droughts while water availability has a stronger influence during wet season droughts.
-
-<!----KAT - wetter habitats have higher growth reductions - because used to more water---->
+Here we examine individual tree growth during two known ENSO drought years within a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand to disentangle the drivers of tree growth sensitivity to drought.
+We hypothesise that tropical tree drought sensitivity is jointly shaped by species adaptations, individual microenvironments and drought characteristics. 
+<!-- We ask: i) How much do species vary in drought sensitivity of tree growth? What drives this variation ii) What drives variation among individuals in the sensitivity of growth to drought?  -->
+Specifically, we expect that i) both species and individuals respond differently to different droughts. 
+ii) Deciduous species, which rely on wet seasons for growth, are particularly sensitive to wet season drought. 
+iii) Exposed trees are more sensitive to drought, but water availability, either through larger size or topographic location, can buffer growth declines.
 
@@
 - **exposure** : trees with higher exposure because of their canopy position have lower drought resistance than trees with lower exposure due to the direct effects of temperature and vapour pressure deficit.
-- **competition** : trees in denser stands have lower drought resistance than trees in sparser stands because of more intense competition for groundwater.
-Differences in rooting depths could add complexities to this effect, however, we do not have direct measures that could test this effect.
-- **exposure x size** :  While larger trees are likely to have uniform high exposure, smaller trees can have high or low canopy position based on stand characteristics (stand density, presence of a canopy gap etc.) Drought resistance of smaller trees with high exposure is expected to be lower than that of trees with low exposure because of the direct influence of temperature and light that could lead to cavitation.
-(BCI light x size interaction in @rüger_growth_2011)
 - **leaf habit x exposure** : species canopy strategies along with their leaf habit could exacerbate or counteract drought vulnerability [@rahman_speciesspecific_2019].
 Under high exposure, deciduous species are potentially more drought resistant than evergreen species, while the pattern could be less clear under low exposure.
-- **leaf habit x habitat** : tropical evergreen and deciduous species have different habitat preferences [@kunert_leaf_2021] that could affect their drought resistance.   
-- **habitat x exposure** : trees with high exposure in upland habitats are expected to have lowest drought resistance because of the compounding influence of abiotic stressors.   -->
+- **leaf habit x habitat** : tropical evergreen and deciduous species have different habitat preferences [@kunert_leaf_2021] that could affect their drought resistance.   -->
 
@@
 Long-term analysis of tree-ring data shows that tree growth at Huai Kha Khaeng has been influenced by temperature and rainfall [@vlam_temperature_2014].
+
+![***Figure 1.
+Climatic characteristics of drought years and average years*** in Huai Kha Khaeng ForestGEO plot.
+For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1.png)
 

### 2025-03-20 doc/manuscript_hkk_drought_sensitivity.Rmd
results titles reworded, fig 1 replaced

@@
 **Number of References:**  
-**Number of Figures:**  
-**Number of Tables:**  
-**Number of Text boxes:**
+**Number of Figures:** 5  
+**Number of Tables:** 1  
+**Number of Text boxes:** 0
 
-**Corresponding author address:**
+**Corresponding author address:** anujank@si.edu
 
@@
 
-<!-- Ecology Letters format 
--->
-
-<!-- # Summary {-} -->
-<!-- 
-New Phytologist Summary format
-- Rationale  
-- Methods  
-- Key results   
-- Main conclusion   -->
 # Abstract (150 words) {-}
-
-<!-- - Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but how climate anomalies might impact the growth of tropical tree growth is poorly understood. 
-Standard methods like tree rings and plot inventories may either not represent most species or may not be accurate enough to capture variation in slow-growing species.  
-- We used a unique data record - manual dendrometer band measurements over 15 years on 1820 trees across 30 species in a seasonally dry tropical forest in Huai Kha Khaeng, Thailand, along with species traits (deciduousness and maximum size), individual characteristics (crown exposure, size) and environment-related water availability to predict the sensitivity of tree growth to drought in two ENSO drought events with different characteristics - a wet-season drought and a dry season drought.
-We used Bayesian regressions in a causal modelling framework to predict variability among species and individuals in drought sensitivity.
-- In the wet-season drought, deciduous species were more sensitive than evergreen species, especially in wetter microsites, while these patterns were not significant in the dry-season drought. 
-After accounting for species differences, more exposed trees were more likely to be negatively impacted than their less exposed counterparts, with exposure showing clear negative impacts compared to size *per se*.
-- Large differences among species and individuals in their sensitivity to drought, as well as climatic differences differences among droughts themselves can explain heterogeneity in ecological responses to drought.
--->
-
-<!-- Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but how climate anomalies like drought might impact the growth of tropical tree growth is poorly understood.
--->
 The frequency and intensity of droughts are increasing across tropical forests, altering tree demography and its contribution to the land carbon sink. 
@@
 
-<!-- # Keywords {-} -->
 \newpage
@@
 )
-# Merge the title rows across all columns
-# ft <- merge_at(ft, i = c(1, 3, 6), j = 1:3, part = "body")
-# # Add text to the merged row
-# ft <- compose(ft, i = c(1, 3, 6), j = 1, value = as_paragraph(ft[1, 1], ft[3, 1], ft[6, 1]))
-ft <- hline(ft, i = c(1, 4, 8), part = "body")
+ft <- hline(ft, i = c(1, 4, 7), part = "body")
 ft <- line_spacing(ft, space = 1, part = "all")
 ft <- autofit(ft)
-ft <- bold(ft, i = c(1, 4, 8), bold = TRUE, part = "body")
-# ft <- width(ft, 1, width = 1.75)
+# ft <- height(ft, height = 0.3)
+ft <- bold(ft, i = c(1, 4, 7), bold = TRUE, part = "body")
+ft <- width(ft, 1, width = 3.5)
 # ft <- width(ft, 2, width = 1.75)
@@
 Climatic characteristics of drought years and average years*** in Huai Kha Khaeng ForestGEO plot.
-For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1.png)
+For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1_anom_2.png)
 
@@
 
-## Interannual growth variation and drought sensitivity {-}
+## Drought sensitivity across distinct drought events {-}
 
@@
 
-## Variation in sensitivity among species {-}
+## Predictors of species-level variation in drought sensitivity {-}
 
@@
 
-## Variation among individuals {-}
+## Predictors of stem-level variation in drought sensitivity {-}
 

### 2025-03-20 doc/manuscript_hkk_drought_sensitivity.Rmd
manuscript with introduction points
@teixeirak ready for comment

@@
 
-Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but it may be vulnerable to climate anomalies like drought.  
+Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but it may be vulnerable to climate anomalies like drought [@tao_increasing_2022].  
 - Woody growth has one of the longest residence times in terrestrial carbon pool, contributing to long-lasting impacts on carbon removal from the atmosphere [@yang_variations_2021; @muller-landau_patterns_2021].
@@
 <!-- that varies across species and individuals [@bennett_larger_2015; @mcgregor_tree_2021].   -->
-- Seasonally dry tropical forests are especially vulnerable to ecological impacts of drought, being at the edge of plant thermal and water stress tolerance [@allen_will_2017].
-- Within a stand, drought sensitivity varies among species and individuals, but the extent and drivers of this variation in drought responses is poorly understood.  
+<!-- - Seasonally dry tropical forests are especially vulnerable to ecological impacts of drought, being at the edge of plant thermal and water stress tolerance [@allen_will_2017]. -->
+- Across the tropics, drought frequency and intensity are increasing, and there is some evidence that stand-level vulnerability is increasing [@luo_tropical_2022; @tao_increasing_2022].
+- Drought sensitivity is often understood as a species trait, but it can vary substantially among drought events [] and among individuals based on their characteristics[@bennett_larger_2015, @mcgregor_tree_2021], compounding uncertainties in predicting forest responses.
 - With changing climate patterns across the global tropics, emergence of novel climatic regimes like "very extreme droughts" in recent decades [@dahinden_future_2017; @luo_tropical_2022], there is an urgent need to understand the drivers of this variation, to be able to predict and respond to forest drought responses.
 
-Seasonally dry tropical forests can be vulnerable to droughts, despite trees being adapted to strong dry seasons.  
-- Trees in seasonally dry forests are routinely exposed to hot, dry conditions in the dry season and may have evolutionary strategies like dry-season deciduousness and drought deciduousness for survival.
-- However, across the global tropics, drier forests show stronger sensitivities to drought-associated water stress  [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019].
-- One potential reason - thermal/water stress tolerance thresholds crossed [@allen_will_2017].  
-- Another reason: drought events are climatically different in timing and intensity, with trees exposed to drought conditions in different parts of their seasonal phenologies.  
-- Morevoer, besides broad patterns at the stand level, species- and individual-level heterogeneity can lead to divergent responses with consequences for stand composition.   
-- Some of the discrepancies between expectations from trees and observations at the stand-level may be resolved by considering factors that determine drought sensitivity across species and individuals
+Drought characteristics can influence climatic drivers of tree growth and their timing, and affect drought sensitivity of tropical trees.
+- Tree growth in both everwet and seasonally dry tropical forests show predictable seasonal patterns [@garcía-cervigón_climate_2020, @giraldo_tree_2023], but with differences among these patterns.
+- Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.
+- The effect of drought timing can be more pronounced in seasonally dry tropical forests because of strong seasonality in climate and consequently on photosynthesis, allocation and growth (including dry seasons unfavourable to growth).
+- They are also at thermal/water stress tolerance thresholds, making them more vulnerable to drought [@allen_will_2017], potentially explaining stand-level patterns
+of drought-associated water stress  [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019].
+- However, trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress, which might confer differential advantages in drought conditions.
+<!-- - Trees in seasonally dry forests are routinely exposed to hot, dry conditions in the dry season and may have evolutionary strategies like dry-season deciduousness and <!-- - Morevoer, besides broad patterns at the stand level, species- and individual-level heterogeneity can lead to divergent responses with consequences for stand composition.
+- Some of the discrepancies between expectations from trees and observations at the stand-level may be resolved by considering factors that determine drought sensitivity across species and individuals  -->
 
@@
 - ecophysiological - e.g. negative TLPs, plasticity in deciduousness (root/shoot)
+- These spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  
 - Species with drought tolerant traits like deep roots, more negative turgor loss point and ...
@@
 - Larger trees suffer more growth declines during drought [@bennett_larger_2015], exposed crowns -> more thermal stress [@vinod_thermal_2023] although crown exposure -> more light -> more growth in normal years.
-- Some of these adaptations may confer drought sensitivity and buffer stem growth declines during drought years, or may tradeoff with advantages in normal years. 
-- To understand drought sensitivity representative of stand diversity, trees across these strategies need to be analysed.
-<!-- - Drought events can be different from each other, but generally lead to low soil moisture, low water tables, low vapour pressure and high temperature [@chitra-tarak_hydraulicallyvulnerable_2021].
--->
-<!-- - Results in thermal stress on tissues, evaporative loss, leaf turgor loss and cavitation.  
-- Species can be conservative or acquisitive in water and nutrient uptake; differential allocation to growth/survival.   -->
+- Some of these adaptations may confer drought sensitivity and buffer stem growth declines during drought years, or conversely, may tradeoff with advantages in normal years, leading to a diversity of responses within a drought event.
+- Moreover, drought characteristics can positively or negatively interfere with these factors, leading to consistent or heterogenous sensitivities across drought events.
 
-Drought sensitivity representative of stand diversity in seasonally dry tropical forests is difficult to assess using standard methods.  
-- Annual tree rings - found in SDTFs but not many species,  
-- Species with rings represent specific strategy to water use and growth - and often affiliated to drier microenvironments.  
-- Tape measurements - more representative, but may not be annual because of effort  
-- Even when tape measures are annual, not enough precision; error for each tape measure is 1mm, error for growth is at least 2mm.
+Given the diversity of factors that can influence drought sensitivity in seasonally dry tropical forests, sampling strategies need to effectively span this variation and have adequate precision.  
+- Species with rings are few, represent specific strategy to water use and growth - and often affiliated to drier microenvironments.  
+- Tape measurements can be representative (although often not annual), not enough precision; error for each tape measure is 1mm, error for growth is at least 2mm.
 (Median annual growth across trees in HKK is ~1.4 mm, deviation from annual growth needs more precision).
@@
 - Methods need to capture species strategies, distributions across microenvironments, and be precise enough.  
-<!---Growth reductions from droughts are difficult to estimate reliably from whole-plot inventories.  
-- Droughts are different but generally leads to low soil moisture, low water tables, vapour pressure, high temperature.
-- Results in thermal stress, evaporative loss, leaf turgor loss, embolism/cavitation  High mortality and growth reductions across global forests, including the tropics  
-- Inventory data are often not at annual scales; smooths over drought/no-drought period.  
-- Reductions are often small (<1 mm), and difficult to detect with tape measurements.
+- Manual dendrometer band timeseries can represent a wide range of species strategies with equipment precision necessary to detect signals.
 
- Within species, size, exposure and location can affect sensitivity to drought.  
+ <!-- Within species, size, exposure and location can affect sensitivity to drought.  
 - Large trees tend to undergo greater growth declines during drought compared to smaller trees [e.g., @bennett_larger_2015; @mcgregor_tree_2021; @anderson-teixeira_joint_2022].   

### 2025-03-21 doc/manuscript_hkk_drought_sensitivity.Rmd
Update manuscript_hkk_drought_sensitivity.Rmd
@krishnaanujan , FYI. I am going to make smaller pushes to make it easier for you to see what I've changed.

This push includes:
- tweaks to abstract
- fix to citation issue (need to separate refs with semicolon, not comma)
- commented out methods paragraph near end of intro

@@
 In seasonally dry forests, species adaptations to strong dry seasons might confer drought resistance, but this resistance may be variable with individual microenvironments and across drought events. 
-We analysed a unique data record - manual dendrometer band measurements over 15 years on 1820 trees across 30 species in a seasonally dry tropical forest in <!---Huai Kha Khaeng, ---->Thailand, <!--using Bayesian regressions in a causal modelling framework---> to predict the drought sensitivity of tree growth in two climatically distinct droughts.
-Species and stems exhibited a wide range of drought sensitivities within a drought event, and inconsistent responses across these two droughts. 
-Deciduous species were more sensitive than evergreen species in the wet-season drought, with this difference worsened in wetter microclimates. 
+We analysed a unique data record - manual dendrometer band measurements over 15 years on 1820 trees across 30 species in a seasonally dry tropical forest in <!---Huai Kha Khaeng, ---->Thailand, <!--using Bayesian regressions in a causal modelling framework--->to assess the drought sensitivity of tree growth in two climatically distinct droughts.
+Species and individuals exhibited a wide range of drought sensitivities within a drought event, and inconsistent responses across these two droughts. 
+Deciduous species were more sensitive than evergreen species in the wet-season drought, and this difference was amplified in wetter microclimates. 
 Although exposed trees were more sensitive, access to water through larger size or topographic location buffered growth declines. 
-Heterogenous responses of species and individuals to droughts indicate potential spatiotemporal insurance effects in diverse forests in the face of increased climate anomalies.
+Heterogeneous responses of species and individuals to droughts indicate potential spatiotemporal insurance effects in diverse forests in the face of increased climate anomalies.
 
@@
 - Across the tropics, drought frequency and intensity are increasing, and there is some evidence that stand-level vulnerability is increasing [@luo_tropical_2022; @tao_increasing_2022].
-- Drought sensitivity is often understood as a species trait, but it can vary substantially among drought events [] and among individuals based on their characteristics[@bennett_larger_2015, @mcgregor_tree_2021], compounding uncertainties in predicting forest responses.
+- Drought sensitivity is often understood as a species trait, but it can vary substantially among drought events [] and among individuals based on their characteristics [@bennett_larger_2015; @mcgregor_tree_2021], compounding uncertainties in predicting forest responses.
 - With changing climate patterns across the global tropics, emergence of novel climatic regimes like "very extreme droughts" in recent decades [@dahinden_future_2017; @luo_tropical_2022], there is an urgent need to understand the drivers of this variation, to be able to predict and respond to forest drought responses.
@@
 Drought characteristics can influence climatic drivers of tree growth and their timing, and affect drought sensitivity of tropical trees.
-- Tree growth in both everwet and seasonally dry tropical forests show predictable seasonal patterns [@garcía-cervigón_climate_2020, @giraldo_tree_2023], but with differences among these patterns.
+- Tree growth in both everwet and seasonally dry tropical forests show predictable seasonal patterns [@garcía-cervigón_climate_20200; @giraldo_tree_2023], but with differences among these patterns.
 - Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.
@@
 
+<!--[from KAT: While all good content, I find this paragraph unnecessary/ distracting. If you'd really like to include content here, I'd work it into the methods.]
 Given the diversity of factors that can influence drought sensitivity in seasonally dry tropical forests, sampling strategies need to effectively span this variation and have adequate precision.  
@@
 - Methods need to capture species strategies, distributions across microenvironments, and be precise enough.  
-- Manual dendrometer band timeseries can represent a wide range of species strategies with equipment precision necessary to detect signals.
+- Manual dendrometer band timeseries can represent a wide range of species strategies with equipment precision necessary to detect signals. -->
 

### 2025-03-21 doc/manuscript_hkk_drought_sensitivity.Rmd
Update manuscript_hkk_drought_sensitivity.Rmd
MACROSCOPIC ORGANIZATION OF FRAMING:
- annotated paragraphs with comments summarizing content
- rearranged wording of title and main hypothesis  to have consistent ordering of drought characteristics, species effects, and microenvironment (applied that order)
- cleaned up a few old comments

@@
 
-**Title:** Tropical tree drought sensitivity is jointly shaped by species adaptations, individual microenvironments, and drought characteristics
+**Title:** Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments
 
@@
 
-Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but it may be vulnerable to climate anomalies like drought [@tao_increasing_2022].  
+<!--PARAGRAPH ON IMPORTANCE OF DROUGHT, HOW OUR UNDERSTANDING REMAINS LIMITED, PARTICULARLY FOR SEASONAL TROPICAL FORESTS-->
+Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but is vulnerable to climate anomalies like drought [@tao_increasing_2022].  
 - Woody growth has one of the longest residence times in terrestrial carbon pool, contributing to long-lasting impacts on carbon removal from the atmosphere [@yang_variations_2021; @muller-landau_patterns_2021].
@@
 
+<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
 Drought characteristics can influence climatic drivers of tree growth and their timing, and affect drought sensitivity of tropical trees.
@@
 
+<!--PARAGRAPHS ON HOW SPECIES TRAITS AND INDIVIDUAL CHARACTERISTICS SHAPE DROUGHT RESPONSES (SEPARATE SPECIES AND MICROENVIRONMENT)-->
 Trees in seasonally dry forests have strategies that span and interact across timescales to maximise growth and survival under hot/dry conditions.  
@@
 ```
-<!---KAT framing---->
 
-<!---High Crown exposure makes trees more vulnerable to drought, but drought deciduous habit or perennial water access allows them to escape this  
--Disentangling the effects of tree size, crown exposure, water access, and deciduousness on drought sensitivity   -->
-
-Here we examine individual tree growth during two known ENSO drought years within a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand to disentangle the drivers of tree growth sensitivity to drought.
-We hypothesise that tropical tree drought sensitivity is jointly shaped by species adaptations, individual microenvironments and drought characteristics. 
-<!-- We ask: i) How much do species vary in drought sensitivity of tree growth? What drives this variation ii) What drives variation among individuals in the sensitivity of growth to drought?  -->
-Specifically, we expect that i) both species and individuals respond differently to different droughts. 
+<!--HYPOTHESIS PARAGRAPH-->
+We hypothesize that tropical tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, and individual microenvironments. 
+We test this by examining individual tree growth during two ENSO drought years within a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand. 
+Specifically, we test predictions that
+i) both species and individuals respond differently to different droughts. 
 ii) Deciduous species, which rely on wet seasons for growth, are particularly sensitive to wet season drought. 

### 2025-03-21 doc/manuscript_hkk_drought_sensitivity.Rmd
moved table 1 after hypothesis paragraph
Update manuscript_hkk_drought_sensitivity.Rmd
- added/ organized some intro content (including copying some content from grant proposal)

@@
 <!--PARAGRAPH ON IMPORTANCE OF DROUGHT, HOW OUR UNDERSTANDING REMAINS LIMITED, PARTICULARLY FOR SEASONAL TROPICAL FORESTS-->
-Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but is vulnerable to climate anomalies like drought [@tao_increasing_2022].  
+- Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but is vulnerable to climate anomalies like drought [@tao_increasing_2022].  
+- Every year, woody growth of trees in tropical forests removes roughly `r round (150*1/6,0)` Gt CO~2~ from the atmosphere [@beer_terrestrial_2010;@anderson-teixeira_carbon_2021;@anderson-teixeira_carbon_2016], contributing to a net biomass increase of ~7 Gt CO~2-eq~ [@harris_global_2021] that offsets ~`r round(7/36.3*100,-1)`% of anthropogenic fossil fuel emissions [@friedlingstein_global_2022].<!--copied from our grant proposal - something like this is more specific and perhaps more compelling-->
 - Woody growth has one of the longest residence times in terrestrial carbon pool, contributing to long-lasting impacts on carbon removal from the atmosphere [@yang_variations_2021; @muller-landau_patterns_2021].
-- Climate anomalies like drought can lead to stand-level growth reductions, along with increased mortality [@bennett_sensitivity_2023; @bauman_tropical_2022].  
-<!-- that varies across species and individuals [@bennett_larger_2015; @mcgregor_tree_2021].   -->
-<!-- - Seasonally dry tropical forests are especially vulnerable to ecological impacts of drought, being at the edge of plant thermal and water stress tolerance [@allen_will_2017]. -->
-- Across the tropics, drought frequency and intensity are increasing, and there is some evidence that stand-level vulnerability is increasing [@luo_tropical_2022; @tao_increasing_2022].
+- However, tree growth is increasingly vulnerable to drought, the frequency and intensity of which are increasing across the much of tropics [@tao_increasing_2022].
+<!--and there is some evidence that stand-level vulnerability is increasing [@luo_tropical_2022; @tao_increasing_2022].-->
+- (We don't know very much about woody growth responses, as few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022].)
 - Drought sensitivity is often understood as a species trait, but it can vary substantially among drought events [] and among individuals based on their characteristics [@bennett_larger_2015; @mcgregor_tree_2021], compounding uncertainties in predicting forest responses.
@@
 - Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.
+- Analyses to date have found that the woody growth of tropical trees <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly vulnerable to dry season precipitation[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates.
 - The effect of drought timing can be more pronounced in seasonally dry tropical forests because of strong seasonality in climate and consequently on photosynthesis, allocation and growth (including dry seasons unfavourable to growth).
@@
 
+<!--HYPOTHESIS PARAGRAPH-->
+We hypothesize that tropical tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, and individual microenvironments. 
+We test this by examining individual tree growth during two ENSO drought years within a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand. 
+Specifically, we test predictions that
+i) both species and individuals respond differently to different droughts. 
+ii) Deciduous species, which rely on wet seasons for growth, are particularly sensitive to wet season drought. 
+iii) Exposed trees are more sensitive to drought, but water availability, either through larger size or topographic location, can buffer growth declines.
+
+<!-- - **habitat** : water availability is a key driver of tropical tree growth [@wagner_water_2012]. 
+- **leaf habit** : drought resistance is higher in deciduous species because of leaf strategies that minimise water loss during months of high vulnerability.
+Deciduous and evergreen species have differential sensitivity to drought [@desouza_drought_2020].
+- **tree size** : larger trees face higher risk of mortality during drought, as well as greater growth reductions [e.g., @bennett_larger_2015; @mcgregor_tree_2021; @anderson-teixeira_joint_2022].  
+- **exposure** : trees with higher exposure because of their canopy position have lower drought resistance than trees with lower exposure due to the direct effects of temperature and vapour pressure deficit.
+- **leaf habit x exposure** : species canopy strategies along with their leaf habit could exacerbate or counteract drought vulnerability [@rahman_speciesspecific_2019].
+Under high exposure, deciduous species are potentially more drought resistant than evergreen species, while the pattern could be less clear under low exposure.
+- **leaf habit x habitat** : tropical evergreen and deciduous species have different habitat preferences [@kunert_leaf_2021] that could affect their drought resistance.   -->
+
 ```{r table, echo=FALSE, warning=FALSE, message=FALSE}
@@
 ```
-
-<!--HYPOTHESIS PARAGRAPH-->
-We hypothesize that tropical tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, and individual microenvironments. 
-We test this by examining individual tree growth during two ENSO drought years within a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand. 
-Specifically, we test predictions that
-i) both species and individuals respond differently to different droughts. 
-ii) Deciduous species, which rely on wet seasons for growth, are particularly sensitive to wet season drought. 
-iii) Exposed trees are more sensitive to drought, but water availability, either through larger size or topographic location, can buffer growth declines.
-
-<!-- - **habitat** : water availability is a key driver of tropical tree growth [@wagner_water_2012]. 
-- **leaf habit** : drought resistance is higher in deciduous species because of leaf strategies that minimise water loss during months of high vulnerability.
-Deciduous and evergreen species have differential sensitivity to drought [@desouza_drought_2020].
-- **tree size** : larger trees face higher risk of mortality during drought, as well as greater growth reductions [e.g., @bennett_larger_2015; @mcgregor_tree_2021; @anderson-teixeira_joint_2022].  
-- **exposure** : trees with higher exposure because of their canopy position have lower drought resistance than trees with lower exposure due to the direct effects of temperature and vapour pressure deficit.
-- **leaf habit x exposure** : species canopy strategies along with their leaf habit could exacerbate or counteract drought vulnerability [@rahman_speciesspecific_2019].
-Under high exposure, deciduous species are potentially more drought resistant than evergreen species, while the pattern could be less clear under low exposure.
-- **leaf habit x habitat** : tropical evergreen and deciduous species have different habitat preferences [@kunert_leaf_2021] that could affect their drought resistance.   -->
 

### 2025-03-25 doc/manuscript_hkk_drought_sensitivity.Rmd
Update manuscript_hkk_drought_sensitivity.Rmd
@krishnaanujan , FYI

@@
 
-<!--PARAGRAPH ON IMPORTANCE OF DROUGHT, HOW OUR UNDERSTANDING REMAINS LIMITED, PARTICULARLY FOR SEASONAL TROPICAL FORESTS-->
+<!--KAT suggestion for new intro paragraph (would replace next paragraph)-->
+- *In forests worldwide, intense drought is becoming more frequent, affecting forest demography, diversity, and  C cycling.*
+- *Understanding these impacts is an important research priority, particularly for tropical forests, which harbor high species diversity and are critical in the global C cycle.* 
+- *However, study of tropical tree woody growth responses to drought is inhibited by the fact that few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022].*
+- *We know that drought characteristics, species traits, and microenvironment all affect tree growth responses.* 
+- *However, we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought.*
+<!--The above statement is true at least to my knowledge offhand, but we need to check--> 
+- *Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, it is relatively straightforward to predict tree drought responses; for example, we'd expect a species that has proved relatively resistant to past droughts to be among the most drought-resistant species in any microhabitat and through any drought.
+*
+- *In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in the next.
+The resultant amplification of the complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify.*
+
+<!--PREVIOUS INTRO PARAGRAPH ON IMPORTANCE OF DROUGHT, HOW OUR UNDERSTANDING REMAINS LIMITED, PARTICULARLY FOR SEASONAL TROPICAL FORESTS-->
 - Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but is vulnerable to climate anomalies like drought [@tao_increasing_2022].  
-- Every year, woody growth of trees in tropical forests removes roughly `r round (150*1/6,0)` Gt CO~2~ from the atmosphere [@beer_terrestrial_2010;@anderson-teixeira_carbon_2021;@anderson-teixeira_carbon_2016], contributing to a net biomass increase of ~7 Gt CO~2-eq~ [@harris_global_2021] that offsets ~`r round(7/36.3*100,-1)`% of anthropogenic fossil fuel emissions [@friedlingstein_global_2022].<!--copied from our grant proposal - something like this is more specific and perhaps more compelling-->
 - Woody growth has one of the longest residence times in terrestrial carbon pool, contributing to long-lasting impacts on carbon removal from the atmosphere [@yang_variations_2021; @muller-landau_patterns_2021].
@@
 
+- Our finding that a wet season drought had bigger impact calls into question the generality of the findings of @zuidema_tropical_2022, which found that in species with tree-ring chronologies, dry season drought was most impactful.
+However, care must be taken in interpretation, as the 2015-16 ENSO was stronger.
 - Afzelia xylocarpa, one of the tree ring species had the most distinct (positive) sensitivity in the dry season drought *on how representative sampling is important*

### 2025-03-25 doc/manuscript_hkk_drought_sensitivity.Rmd
intro first para update, discussion first para

@@
 
-<!--PREVIOUS INTRO PARAGRAPH ON IMPORTANCE OF DROUGHT, HOW OUR UNDERSTANDING REMAINS LIMITED, PARTICULARLY FOR SEASONAL TROPICAL FORESTS-->
-- Tropical tree woody growth has long-lasting impacts on the terrestrial carbon pool, but is vulnerable to climate anomalies like drought [@tao_increasing_2022].  
-- Woody growth has one of the longest residence times in terrestrial carbon pool, contributing to long-lasting impacts on carbon removal from the atmosphere [@yang_variations_2021; @muller-landau_patterns_2021].
-- However, tree growth is increasingly vulnerable to drought, the frequency and intensity of which are increasing across the much of tropics [@tao_increasing_2022].
+<!--PARAGRAPH ON IMPORTANCE OF DROUGHT, HOW OUR UNDERSTANDING REMAINS LIMITED, PARTICULARLY FOR SEASONAL TROPICAL FORESTS-->
+
+Increasing drought risk is threatening demographic processes in global forests and consequently their consistent contributions to the land carbon sink [IPCC].  
+- Every year, woody growth of trees in tropical forests removes roughly `r round (150*1/6,0)` Gt CO~2~ from the atmosphere [@beer_terrestrial_2010;@anderson-teixeira_carbon_2021;@anderson-teixeira_carbon_2016], contributing to a net biomass increase of ~7 Gt CO~2-eq~ [@harris_global_2021] that offsets ~`r round(7/36.3*100,-1)`% of anthropogenic fossil fuel emissions [@friedlingstein_global_2022].<!--copied from our grant proposal - something like this is more specific and perhaps more compelling-->
+- Moreover, woody growth has one of the longest residence times in terrestrial carbon pool, contributing to long-lasting impacts on carbon removal from the atmosphere [@yang_variations_2021; @muller-landau_patterns_2021].
+- However, tree growth is vulnerable to drought, the frequency and intensity of which are increasing across the much of tropics [@tao_increasing_2022].
 <!--and there is some evidence that stand-level vulnerability is increasing [@luo_tropical_2022; @tao_increasing_2022].-->
@@
 - Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.
-- Analyses to date have found that the woody growth of tropical trees <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly vulnerable to dry season precipitation[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates.
 - The effect of drought timing can be more pronounced in seasonally dry tropical forests because of strong seasonality in climate and consequently on photosynthesis, allocation and growth (including dry seasons unfavourable to growth).
@@
 of drought-associated water stress  [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019].
-- However, trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress, which might confer differential advantages in drought conditions.
+- Analyses to date have found that the woody growth of tropical trees <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly vulnerable to dry season precipitation[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates.
+- Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress, which might confer differential advantages in drought conditions.
 <!-- - Trees in seasonally dry forests are routinely exposed to hot, dry conditions in the dry season and may have evolutionary strategies like dry-season deciduousness and <!-- - Morevoer, besides broad patterns at the stand level, species- and individual-level heterogeneity can lead to divergent responses with consequences for stand composition.
@@
 
-<!--PARAGRAPHS ON HOW SPECIES TRAITS AND INDIVIDUAL CHARACTERISTICS SHAPE DROUGHT RESPONSES (SEPARATE SPECIES AND MICROENVIRONMENT)-->
-Trees in seasonally dry forests have strategies that span and interact across timescales to maximise growth and survival under hot/dry conditions.  
+<!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
+Trees in seasonally dry forests have strategies that span and interact across timescales to maximise growth and survival under periodic hot/dry conditions.  
 - evolutionary timescales - e.g. dry season deciduousness   
 - Deciduous species, with shorter duration with leaves can have acquisitive strategies during leaf on [@desouza_drought_2020].    
+- Species with drought tolerant traits like deep roots, more negative turgor loss point and ...
+can be more or less resistant to drought [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021].
+- ecophysiological - e.g. negative TLPs, plasticity in deciduousness (root/shoot)
+- These spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.
+
+<!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
+Individual growth responses to drought can also depend on the specific microenvironments they experience that can lead to distict atmospheric and hydrological stresses with ecological consequences.  
+- Larger trees suffer more growth declines during drought [@bennett_larger_2015], exposed crowns -> more thermal stress [@vinod_thermal_2023] although crown exposure -> more light -> more growth in normal years.
 - ecological timescales - e.g. distribution of trees in relation to wetness  
 -  Species distribution of evergreen species covarying with soil moisture [@kunert_leaf_2021] and potentially affecting survival and growth.  
-- ecophysiological - e.g. negative TLPs, plasticity in deciduousness (root/shoot)
-- These spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  
-- Species with drought tolerant traits like deep roots, more negative turgor loss point and ...
-can be more or less resistant to drought [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021].
-- Larger trees suffer more growth declines during drought [@bennett_larger_2015], exposed crowns -> more thermal stress [@vinod_thermal_2023] although crown exposure -> more light -> more growth in normal years.
 - Some of these adaptations may confer drought sensitivity and buffer stem growth declines during drought years, or conversely, may tradeoff with advantages in normal years, leading to a diversity of responses within a drought event.
@@
 
+Our results show that tree drought sensitivity can be jointly shaped by drought characteristics, species adaptations and local environmental conditions.
+We found that some of the variation in sensitivity of tree growth to drought can be explained by a combination of species-level deciduousness, and individual level crown exposure, water availability and other size effects.
+Taken together, our analysis of tree growth in two different drought years, and spanning meaningful variation, shows that species and individuals can respond differently to different drought events.  
+Despite stand-level growth declines in drought years, the heterogeneity in individual and species responses suggest insurance effects in the community with consequences for ecosystem stability.
+
 Unordered points from Introduction

### 2025-03-25 results/results.qmd
intro first para update, discussion first para

@@
 
+##
+
 ## Models
@@
 1. varying intercepts by species  
-$sensitivity_{tree} = \beta_{DBH}DBH_{0} + \beta_{TWI}TWI + \beta_{CII}CII_{0} + (1|Species)$
+$sensitivity_{tree} \sim DBH_{0} + TWI + CII_{0} + (1|Species)$
 
 2. varying intercepts + predictors scaled within species  
-$sensitivity_{tree} = \beta_{DBH}DBH_sp_{0} + \beta_{TWI}TWI_sp + \beta_{CII}CII_sp_{0} + (1|Species)$
+$sensitivity_{tree} \sim DBH_sp_{0} + TWI_sp + CII_sp_{0} + (1|Species)$
 
 3. varying slopes + predictors scaled within species  
-$sensitivity_{tree} = \beta_{DBH}DBH_sp_{0} + \beta_{TWI}TWI_sp + \beta_{CII}CII_sp_{0} + (1+DBH+CII+TWI|Species)$
+$sensitivity_{tree} \sim DBH_sp_{0} + TWI_sp + CII_sp_{0} + (1+DBH+CII+TWI|Species)$
 

### 2025-03-25 scripts/trial.R

@@ -1,0 +1,73 @@
+library(brms)
+library(tidyverse)
+library(tidybayes)
+library(patchwork) # to look at plots side-by-side
+
+# example data from ?brms::mo
+
+# generate some data
+income_options <- c("below_20", "20_to_40", "40_to_100", "greater_100")
+income <- factor(sample(income_options, 100, TRUE),
+    levels = income_options, ordered = TRUE
+)
+mean_ls <- c(30, 60, 70, 75)
+ls <- mean_ls[income] + rnorm(100, sd = 7)
+dat <- data.frame(income, ls)
+
+# add some city-level variation
+dat$city <- rep(1:10, each = 10)
+var_city <- rnorm(10, sd = 10)
+dat$ls <- dat$ls + var_city[dat$city]
+
+
+# plot raw data:
+ggplot(dat, aes(income, ls)) +
+    geom_point()
+
+ggplot(dat, aes(income, ls, color = city)) +
+    geom_point()
+
+# fit a simple monotonic model
+get_prior(ls ~ mo(income), data = dat)
+fit1 <- brm(ls ~ mo(income), data = dat, chains = 4, cores = 4, backend = "cmdstanr")
+summary(fit1)
+p_ce <- plot(conditional_effects(fit1))
+
+# now try manually with spread_draws:
+get_variables(fit1) # what's available
+
+p_manual_ce <- fit1 %>%
+    spread_draws(
+        b_Intercept, bsp_moincome,
+        simo_moincome1[i]
+    ) %>%
+    # simplex parameters are grouped by variable `i`, pivot into same row
+    pivot_wider(names_from = `i`, names_prefix = "simo_moincome1_", values_from = simo_moincome1) %>%
+    mutate(
+        # D is equal to number of categories minus 1
+        D = 3,
+        # for each level of income, get posterior mu
+        mu_cat_1 = b_Intercept,
+        mu_cat_2 = b_Intercept + (bsp_moincome * D * (simo_moincome1_1)),
+        mu_cat_3 = b_Intercept + (bsp_moincome * D * (simo_moincome1_1 + simo_moincome1_2)),
+        mu_cat_4 = b_Intercept + (bsp_moincome * D * (simo_moincome1_1 + simo_moincome1_2 + simo_moincome1_3))
+    ) %>%
+    # at this point or after pivoting again, you could apply the inverse link function on mu
[…]

### 2025-03-25 doc/manuscript_hkk_drought_sensitivity.Rmd
removed conflict

@@
 <!--PARAGRAPH ON IMPORTANCE OF DROUGHT, HOW OUR UNDERSTANDING REMAINS LIMITED, PARTICULARLY FOR SEASONAL TROPICAL FORESTS-->
-
 Increasing drought risk is threatening demographic processes in global forests and consequently their consistent contributions to the land carbon sink [IPCC].  

### 2025-03-26 doc/manuscript_hkk_drought_sensitivity.Rmd
discussion skeleton

--- 2025-03-25 doc/manuscript_hkk_drought_sensitivity.Rmd
@@
 <!--KAT suggestion for new intro paragraph (would replace next paragraph)-->
-- *In forests worldwide, intense drought is becoming more frequent, affecting forest demography, diversity, and  C cycling.*
-- *Understanding these impacts is an important research priority, particularly for tropical forests, which harbor high species diversity and are critical in the global C cycle.* 
-- *However, study of tropical tree woody growth responses to drought is inhibited by the fact that few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022].*
-- *We know that drought characteristics, species traits, and microenvironment all affect tree growth responses.* 
-- *However, we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought.*
+In forests worldwide, intense drought is becoming more frequent, affecting forest demography, diversity, and  C cycling. [IPCC]
+- Understanding these impacts is an important research priority, particularly for tropical forests, which harbor high species diversity and are critical in the global C cycle.
+- However, study of tropical tree woody growth responses to drought is inhibited by the fact that few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022].
+- We know that drought characteristics, species traits, and microenvironment all affect tree growth responses.  
+- However, we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought.  
 <!--The above statement is true at least to my knowledge offhand, but we need to check--> 
-- *Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, it is relatively straightforward to predict tree drought responses; for example, we'd expect a species that has proved relatively resistant to past droughts to be among the most drought-resistant species in any microhabitat and through any drought.
-*
-- *In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in the next.
-The resultant amplification of the complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify.*
+- Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, it is relatively straightforward to predict tree drought responses; for example, we'd expect a species that has proved relatively resistant to past droughts to be among the most drought-resistant species in any microhabitat and through any drought.
+- In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in the next.
+The resultant amplification of the complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify.
 
 <!--PARAGRAPH ON IMPORTANCE OF DROUGHT, HOW OUR UNDERSTANDING REMAINS LIMITED, PARTICULARLY FOR SEASONAL TROPICAL FORESTS-->
-
-Increasing drought risk is threatening demographic processes in global forests and consequently their consistent contributions to the land carbon sink [IPCC].  
-- Every year, woody growth of trees in tropical forests removes roughly `r round (150*1/6,0)` Gt CO~2~ from the atmosphere [@beer_terrestrial_2010;@anderson-teixeira_carbon_2021;@anderson-teixeira_carbon_2016], contributing to a net biomass increase of ~7 Gt CO~2-eq~ [@harris_global_2021] that offsets ~`r round(7/36.3*100,-1)`% of anthropogenic fossil fuel emissions [@friedlingstein_global_2022].<!--copied from our grant proposal - something like this is more specific and perhaps more compelling-->
-- Moreover, woody growth has one of the longest residence times in terrestrial carbon pool, contributing to long-lasting impacts on carbon removal from the atmosphere [@yang_variations_2021; @muller-landau_patterns_2021].
+<!-- Increasing drought risk is threatening demographic processes in global forests and consequently their consistent contributions to the land carbon sink [IPCC].   -->
+<!-- - Every year, woody growth of trees in tropical forests removes roughly `r round (150*1/6,0)` Gt CO~2~ from the atmosphere [@beer_terrestrial_2010;@anderson-teixeira_carbon_2021;@anderson-teixeira_carbon_2016], contributing to a net biomass increase of ~7 Gt CO~2-eq~ [@harris_global_2021] that offsets ~`r round(7/36.3*100,-1)`% of anthropogenic fossil fuel emissions [@friedlingstein_global_2022].copied from our grant proposal - something like this is more specific and perhaps more compelling -->
+<!-- - Moreover, woody growth has one of the longest residence times in terrestrial carbon pool, contributing to long-lasting impacts on carbon removal from the atmosphere [@yang_variations_2021; @muller-landau_patterns_2021].
 - However, tree growth is vulnerable to drought, the frequency and intensity of which are increasing across the much of tropics [@tao_increasing_2022].
 <!--and there is some evidence that stand-level vulnerability is increasing [@luo_tropical_2022; @tao_increasing_2022].-->
-- (We don't know very much about woody growth responses, as few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022].)
+<!-- - (We don't know very much about woody growth responses, as few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022].)
 - Drought sensitivity is often understood as a species trait, but it can vary substantially among drought events [] and among individuals based on their characteristics [@bennett_larger_2015; @mcgregor_tree_2021], compounding uncertainties in predicting forest responses.
 - With changing climate patterns across the global tropics, emergence of novel climatic regimes like "very extreme droughts" in recent decades [@dahinden_future_2017; @luo_tropical_2022], there is an urgent need to understand the drivers of this variation, to be able to predict and respond to forest drought responses.
+ -->
 
@@
 - Tree growth in both everwet and seasonally dry tropical forests show predictable seasonal patterns [@garcía-cervigón_climate_20200; @giraldo_tree_2023], but with differences among these patterns. 
+- Drought definitions are many, but here we refer to hot dry conditions anomalous from long-term means
+- Drought is anomalously hot/dry conditios, but different types based on which variable is anomalous.  
 - Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.  
@@
 - Analyses to date have found that the woody growth of tropical trees <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly vulnerable to dry season precipitation[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates.
+- DRY SEASON IS NOT DROUGHT; dry season has hot, dry conditions periodic and predictable, but anomalously hot, dry conditions, unpredictable.  
 - Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress, which might confer differential advantages in drought conditions.
-<!-- - Trees in seasonally dry forests are routinely exposed to hot, dry conditions in the dry season and may have evolutionary strategies like dry-season deciduousness and <!-- - Morevoer, besides broad patterns at the stand level, species- and individual-level heterogeneity can lead to divergent responses with consequences for stand composition.
-- Some of the discrepancies between expectations from trees and observations at the stand-level may be resolved by considering factors that determine drought sensitivity across species and individuals  -->
 
@@
 -  Species distribution of evergreen species covarying with soil moisture [@kunert_leaf_2021] and potentially affecting survival and growth.  
+- Larger trees have larger root systems, but do not necessarily access deeper water [@ref_from_Panama]  
+- Even when trees are accessing deeper water, this does not mean that they're in better shape during drought.
+Rather, trees that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021].
+- indeed, there is evidence that trees near streams undergo greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta] during drought  
 - Some of these adaptations may confer drought sensitivity and buffer stem growth declines during drought years, or conversely, may tradeoff with advantages in normal years, leading to a diversity of responses within a drought event.
@@
  <!-- Within species, size, exposure and location can affect sensitivity to drought.  
-- Large trees tend to undergo greater growth declines during drought compared to smaller trees [e.g., @bennett_larger_2015; @mcgregor_tree_2021; @anderson-teixeira_joint_2022].   
-- It remains poorly understood the extent to which this is shaped by tree size itself, crown exposure, water access, and traits that tend to covary with size (e.g., decidiousness) -- all of which interact to shape drought resistance (Fig.
-`r fig_DAG`).
 - There is theory and evidence that tree size itself matters.  
@@
 - Trees with exposed crowns suffered significant crown dieback at greater rates in the 2012-16 CA drought [@ma_tree_2023].  
-- Water access....  
-- Larger trees have larger root systems, but do not necessarily access deeper water [@ref_from_Panama]  
-- Even when trees are accessing deeper water, this does not mean that they're in better shape during drought.
-Rather, trees that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021].
-- indeed, there is evidence that trees near streams undergo greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta] during drought
 - rather than necessarily helping during drought, water access will shape the size and traits of species living in habitat, with stream habitats tending to have larger trees [@ref] and more evergreen trees, and also their average growth rate--->
@@
 Deciduous and evergreen species have differential sensitivity to drought [@desouza_drought_2020].
-- **tree size** : larger trees face higher risk of mortality during drought, as well as greater growth reductions [e.g., @bennett_larger_2015; @mcgregor_tree_2021; @anderson-teixeira_joint_2022].  
-- **exposure** : trees with higher exposure because of their canopy position have lower drought resistance than trees with lower exposure due to the direct effects of temperature and vapour pressure deficit.
 - **leaf habit x exposure** : species canopy strategies along with their leaf habit could exacerbate or counteract drought vulnerability [@rahman_speciesspecific_2019].
@@
 
-Our results show that tree drought sensitivity can be jointly shaped by drought characteristics, species adaptations and local environmental conditions.
-We found that some of the variation in sensitivity of tree growth to drought can be explained by a combination of species-level deciduousness, and individual level crown exposure, water availability and other size effects.
+<!---CONCLUSIONS---->
+We show strong evidence that tropical tree drought sensitivity can be jointly and interactively shaped by drought characteristics, species adaptations and local environmental conditions (Table 1).
+Analysing two droughts in a 14-year dendrometer band timeseries, we found that tropical tree species and individuals grew less than average in drought years, reflecting in lower plot-level growth (Fig 2).
+Comparing across droughts, we found that despite the shorter duration of drought conditions, the wet-season drought elicited stronger responses across trees, but especially so for deciduous species that are expected to have stronger seasonality in growth phenology.
+Interestingly, in the wet-season drought, we also show that evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites.
+Further, besides consistent negative effects of crown exposure, we found that the drivers of variation among species and individuals in drought sensitivity were dissimilar between the drought events. 
+Water availability buffered exposure-associated growth declines through different mechanisms in the two different droughts - larger trees (with likely deeper roots) in the dry-season drought and trees in wetter sites in the wet-season drought suffered smaller declines in growth.
 Taken together, our analysis of tree growth in two different drought years, and spanning meaningful variation, shows that species and individuals can respond differently to different drought events.  
-Despite stand-level growth declines in drought years, the heterogeneity in individual and species responses suggest insurance effects in the community with consequences for ecosystem stability.
+Despite stand-level growth declines in drought years, we show that these broad patterns mask enormous heterogeneity in individual and species responses to drought in tropical forests with possible consequences for the maintenance of species diveristy and ecosystem stability.
 
-Unordered points from Introduction
+<!----Limitations in filling gap---->
+a: Although our analysis spans representative spatial and ecological variation, we are limited in quantitatively analysing differences among droughts and variation in drought characteristics, because of low sample size.
+-   Our analysis spans ecological and spatial/environmental variation realistic to tropical trees, and beyond previous such analyses.
+- But given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts.
+- Thus, quantitative analyses parsing out the specific effects of drought characteristics like relative influence of timing and degree of anomaly is currently impossible, limiting interpretations.  
+- However, as the record continues to grow, so does the potential to answer these questions.
+
+b: We analysed species traits as static, but we might be missing intraspecific and plastic variation in drought sensitivity traits, as well as other axes of trait variation that might explain variation in drought sensitivity.
 - Many hydraulic-related traits vary with tree height [ @vinod_thermal_2023], including the frequency of dry season deciduous leaf loss -- both within species and at the community level [@condit_ref; @meakem].
 *caveat for species-level traits*
+- While leaf deciduousness was the main axis of hydraulic conservative strategy that we considered, leaf-level stomatal regulation and belowground root loss etc.
+are other potential strategies for conservative water use, and our interpretation of species patterns may be limited by our understanding of these processes across species.
+- Surprisingly, although deciduousness may have had an advantage in the dry season drought, because this strategy -> conservative water use in the dry season, species differences in this drought were not predicted by deciduousness.
 
-*for where else might they be allocating to?*
-- Some evidence that both understory and emergent species can have reduced survival during drought [@machado_forest_2023].  
-- ENSO droguht changed regeneration dynamics in Thailand [@nutiprapun_effects_2023]
+<!----Limitations in generalisation---->
+a: Differential allocation to growth may not mean negative outcomes, it might mean demographic trade-offs
+- Differences in allocation to growth may not be the complete picture - may have altered allocation to mortality or reproduction.  
+- Some evidence that both understory and emergent species can have reduced survival during drought [@machado_forest_2023].  *for where else might they be allocating to?*
+- 2015 ENSO drought changed regeneration dynamics in Thailand [@nutiprapun_effects_2023]
+- also NSCs and then compensatory growth
 - deciduous species have higher survival under multi-year droughts [@aguirre-guiterrez refs]
 
-- (BCI light x size interaction in @rüger_growth_2011)
+b: Tropical dry forests also represent periodic hot/dry conditions, leading to adaptations or other genetic variation for resistance, but this might not hold in other bioclimatic zones where ecological variability is different.
+- Trait variation across forest and bioclimatic types
+- Careful interpretation across forest types with different adaptive pressures.  
+- However, some generalisation is possible because despite low seasonality in climatic conditions, everwet forest species also show seasonal growth patterns.
 
+<!----Contributions beyond---->
+Despite consistent growth declines at the plot level across drought events, we show that this scale obscures a variety of ecological responses at the species and tree level that are heterogeneous across drought events.
 - Our finding that a wet season drought had bigger impact calls into question the generality of the findings of @zuidema_tropical_2022, which found that in species with tree-ring chronologies, dry season drought was most impactful.
 However, care must be taken in interpretation, as the 2015-16 ENSO was stronger.
-- Afzelia xylocarpa, one of the tree ring species had the most distinct (positive) sensitivity in the dry season drought *on how representative sampling is important*
+- Influence of deciduousness was different in the two different droughts. 
+- *Afzelia xylocarpa*, one of the tree ring species had the most distinct (positive) sensitivity in the dry season drought, but a large negative sensitivity in the wet season drought - one of the most distinct responses.
+- More generally, tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may be biased towards fast growing (measurable using tape) or dry-associated species (that form annual tree rings) - which might have been capturing only the more dramatic effects.
+- Distinct responses of species and individuals to drought characteristics suggests potential spatiotemporal insurance effects among tropical tree response to drought.
+
+<!----Science is better now---->
+Drought sensitivity is often defined as a static species trait but we show dimensions beyond that with possible consequences for management and restoration 
+- Although net decrease at the plot level (agreement with landscape-scale in @aguirre-guiterrez), we show evidence that the same strategies are not uniformly successfully across spatiotemporal climatic variation.
+- Frameworks that include ecological complexity and covarying traits with environment like @trugman_why_2021 also assume individual trees to have consistent sensitivity across events, but our results demonstrate temporal dimensions of variation, at least partially attributable to drought characteristics.
+- Our results indicate that drought responses in tropical trees may not be a fixed response across events, but jointly shaped by...
+- Given these inconsistencies, diverse stands that may represent greater variation in ecological and evolutionary strategies, may have mechanisms for stable functioning under future droughts.
+
+<!---SPECIES EFFECTS--->
+
+- (BCI light x size interaction in @rüger_growth_2011)
 

### 2025-03-31 doc/manuscript_hkk_drought_sensitivity.Rmd
cleaned up front page, para 1 and 2

@@
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}
-library(flextable)
-set_flextable_defaults(font.family = "Arial", font.size = 9)
+# library(flextable)
+# set_flextable_defaults(font.family = "Arial", font.size = 9)
 team <- read.csv("display/authors.csv", stringsAsFactors = FALSE, check.names = FALSE)
 
-ft <- flextable(team)
-ft <- line_spacing(ft, space = 1, part = "all")
-ft <- autofit(ft)
-ft <- width(ft, 1, width = 1.75)
-ft <- width(ft, 2, width = 1.75)
-ft <- width(ft, 3, width = 3)
-ft
+# team <- read.csv("doc/display/authors.csv", stringsAsFactors = FALSE, check.names = FALSE)
+
+# ft <- flextable(team)
+# ft <- line_spacing(ft, space = 1, part = "all")
+# ft <- autofit(ft)
+# ft <- width(ft, 1, width = 1.75)
+# ft <- width(ft, 2, width = 1.75)
+# ft <- width(ft, 3, width = 3)
+# ft
 ```
+
+Krishna Anujan^1,2^*  ORCiD: `r team$ORCID[1]`  
+Sean McMahon^2,3^  ORCiD: `r team$ORCID[2]`  
+Sarayudh Bunyavejchewin^4^  ORCiD: `r team$ORCID[3]`  
+Stuart J. Davies^3^  ORCiD: `r team$ORCID[4]`  
+Helene C. Muller-Landau^3^  ORCiD : `r team$ORCID[5]`  
+Nantachai Pongpattananurak^5^  ORCiD: `r team$ORCID[6]`  
+Kristina Anderson-Teixeira^1,3^  ORCiD: `r team$ORCID[7]`
+
+**Affiliations:**
+
+^1^SCBI
+^2^SERC
+^3^STRI
+^4^National Park, Wildlife and Plant Conservation Department, Thailand
+^5^Kasetsart University, Thailand
+
+*=corresponding author
 
@@
 <!--KAT suggestion for new intro paragraph (would replace next paragraph)-->
-In forests worldwide, intense drought is becoming more frequent, affecting forest demography, diversity, and  C cycling. [IPCC]
-- Understanding these impacts is an important research priority, particularly for tropical forests, which harbor high species diversity and are critical in the global C cycle.
-- However, study of tropical tree woody growth responses to drought is inhibited by the fact that few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022].
-- We know that drought characteristics, species traits, and microenvironment all affect tree growth responses.  
-- However, we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought.  
+In forests worldwide, intense drought is becoming more frequent, affecting forest demography, diversity, and  C cycling [@intergovernmentalpanelonclimatechangeipcc_climate_2023]. 
+Understanding these impacts is an important research priority, particularly for tropical forests, which <!--harbor high species diversity and --> are critical C sinks; tropical tree woody growth removes roughly `r round (150*1/6,0)` Gt CO~2~ and offsets ~`r round(7/36.3*100,-1)`% of anthropogenic fossil fuel emissions [@beer_terrestrial_2010;@anderson-teixeira_carbon_2021;@anderson-teixeira_carbon_2016;@brando_droughts_2019].
+Drought events can threaten this sink at different temporal scales; the 2015-2016 ENSO drought in the Amazon temporarily shifted the region from a net sink to a net source of carbon [@brando_droughts_2019].
+However, the study of tropical tree woody growth responses to drought is inhibited by the fact that few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022]. 
+We know that drought characteristics [], species traits, and microenvironment [@bennett_larger_2015;@mcgregor_tree_2021] all affect tree growth responses. 
+However, we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought. 
 <!--The above statement is true at least to my knowledge offhand, but we need to check--> 
-- Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, it is relatively straightforward to predict tree drought responses; for example, we'd expect a species that has proved relatively resistant to past droughts to be among the most drought-resistant species in any microhabitat and through any drought.
-- In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in the next.
-The resultant amplification of the complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify.
+Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; for example, a species with low sensitivity to past droughts would be among the least sensitive species in any microhabitat and through any drought.
+In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in a different drought. 
+The resultant amplification of the complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022].
 
@@
 <!--and there is some evidence that stand-level vulnerability is increasing [@luo_tropical_2022; @tao_increasing_2022].-->
-<!-- - (We don't know very much about woody growth responses, as few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022].)
-- Drought sensitivity is often understood as a species trait, but it can vary substantially among drought events [] and among individuals based on their characteristics [@bennett_larger_2015; @mcgregor_tree_2021], compounding uncertainties in predicting forest responses.
+<!-- 
 - With changing climate patterns across the global tropics, emergence of novel climatic regimes like "very extreme droughts" in recent decades [@dahinden_future_2017; @luo_tropical_2022], there is an urgent need to understand the drivers of this variation, to be able to predict and respond to forest drought responses.
@@
 Drought characteristics can influence climatic drivers of tree growth and their timing, and affect drought sensitivity of tropical trees. 
-- Tree growth in both everwet and seasonally dry tropical forests show predictable seasonal patterns [@garcía-cervigón_climate_20200; @giraldo_tree_2023], but with differences among these patterns. 
-- Drought definitions are many, but here we refer to hot dry conditions anomalous from long-term means
-- Drought is anomalously hot/dry conditios, but different types based on which variable is anomalous.  
-- Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.  
-- The effect of drought timing can be more pronounced in seasonally dry tropical forests because of strong seasonality in climate and consequently on photosynthesis, allocation and growth (including dry seasons unfavourable to growth).
-- They are also at thermal/water stress tolerance thresholds, making them more vulnerable to drought [@allen_will_2017], potentially explaining stand-level patterns
-of drought-associated water stress  [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019].
-- Analyses to date have found that the woody growth of tropical trees <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly vulnerable to dry season precipitation[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates.
-- DRY SEASON IS NOT DROUGHT; dry season has hot, dry conditions periodic and predictable, but anomalously hot, dry conditions, unpredictable.  
-- Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress, which might confer differential advantages in drought conditions.
+Droughts are notoriously hard to define and encompass a variey of types based on the meteorogical/climatic variables affected [@slette_how_2019]. 
+Here we define drought as hot/dry conditions anomalously departing from long-term means during any given time of year. 
+<!-- Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.  -->
+<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
+Tree growth responses to drought timing can be more pronounced in seasonally dry tropical forests because of strong seasonality in climate and consequently on photosynthesis, allocation and growth (including dry seasons unfavourable to growth) [@garcía-cervigón_climate_2020].
+(Here we distinguish dry season from drought with differences in peridicity; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable and anomalous hot, dry conditions.)
+Seasonally dry tropical forests are also at thermal/water stress tolerance thresholds, making them more vulnerable to drought [@allen_will_2017], and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019].
+Analyses to date have found that the woody growth of tropical trees <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly vulnerable to dry season precipitation[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates.
+Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress, which might confer differential advantages in drought conditions.
 
@@
 <!----Limitations in filling gap---->
-a: Although our analysis spans representative spatial and ecological variation, we are limited in quantitatively analysing differences among droughts and variation in drought characteristics, because of low sample size.
+Although our analysis spans representative spatial and ecological variation, we are limited in quantitatively analysing differences among droughts and variation in drought characteristics, because of low sample size.
 -   Our analysis spans ecological and spatial/environmental variation realistic to tropical trees, and beyond previous such analyses.
@@
 
-b: We analysed species traits as static, but we might be missing intraspecific and plastic variation in drought sensitivity traits, as well as other axes of trait variation that might explain variation in drought sensitivity.
+<!-- ALT: We analysed species traits as static, but we might be missing intraspecific and plastic variation in drought sensitivity traits, as well as other axes of trait variation that might explain variation in drought sensitivity.
 - Many hydraulic-related traits vary with tree height [ @vinod_thermal_2023], including the frequency of dry season deciduous leaf loss -- both within species and at the community level [@condit_ref; @meakem].
@@
 - Surprisingly, although deciduousness may have had an advantage in the dry season drought, because this strategy -> conservative water use in the dry season, species differences in this drought were not predicted by deciduousness.
+-->
 
 <!----Limitations in generalisation---->
-a: Differential allocation to growth may not mean negative outcomes, it might mean demographic trade-offs
+<!-- ALT: Differential allocation to growth may not mean negative outcomes, it might mean demographic trade-offs
 - Differences in allocation to growth may not be the complete picture - may have altered allocation to mortality or reproduction.  
@@
 - also NSCs and then compensatory growth
-- deciduous species have higher survival under multi-year droughts [@aguirre-guiterrez refs]
+- deciduous species have higher survival under multi-year droughts [@aguirre-guiterrez refs] -->
 
-b: Tropical dry forests also represent periodic hot/dry conditions, leading to adaptations or other genetic variation for resistance, but this might not hold in other bioclimatic zones where ecological variability is different.
+Tropical dry forests also represent periodic hot/dry conditions, leading to adaptations or other genetic variation for resistance, but this might not hold in other bioclimatic zones where ecological variability is different.
 - Trait variation across forest and bioclimatic types
@@
 
-<!---SPECIES EFFECTS--->
+<!---Other points--->
 

### 2025-04-01 doc/manuscript_hkk_drought_sensitivity.Rmd
cleanish intro

@@
 In forests worldwide, intense drought is becoming more frequent, affecting forest demography, diversity, and  C cycling [@intergovernmentalpanelonclimatechangeipcc_climate_2023]. 
-Understanding these impacts is an important research priority, particularly for tropical forests, which <!--harbor high species diversity and --> are critical C sinks; tropical tree woody growth removes roughly `r round (150*1/6,0)` Gt CO~2~ and offsets ~`r round(7/36.3*100,-1)`% of anthropogenic fossil fuel emissions [@beer_terrestrial_2010;@anderson-teixeira_carbon_2021;@anderson-teixeira_carbon_2016;@brando_droughts_2019].
-Drought events can threaten this sink at different temporal scales; the 2015-2016 ENSO drought in the Amazon temporarily shifted the region from a net sink to a net source of carbon [@brando_droughts_2019].
+Understanding these impacts is an important research priority, particularly for tropical forests, which <!--harbor high species diversity and --> are critical C sinks; tropical tree woody growth removes roughly `r round (150*1/6,0)` Gt CO~2~ annually and offsets ~`r round(7/36.3*100,-1)`% of anthropogenic fossil fuel emissions [@beer_terrestrial_2010;@anderson-teixeira_carbon_2021;@anderson-teixeira_carbon_2016;@brando_droughts_2019].
+Drought events can threaten this sink; the 2015-2016 ENSO drought in the Amazon temporarily shifted the region from a net sink to a net source of carbon [@brando_droughts_2019].
 However, the study of tropical tree woody growth responses to drought is inhibited by the fact that few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022]. 
@@
 Drought characteristics can influence climatic drivers of tree growth and their timing, and affect drought sensitivity of tropical trees. 
-Droughts are notoriously hard to define and encompass a variey of types based on the meteorogical/climatic variables affected [@slette_how_2019]. 
-Here we define drought as hot/dry conditions anomalously departing from long-term means during any given time of year. 
+Droughts are notoriously hard to define and encompass a variety of types based on the meteorogical or climatic variables affected [@slette_how_2019]. 
+Here we define drought as hot/dry conditions anomalously departing from their long-term means during any given time of year. 
 <!-- Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.  -->
@@
 Tree growth responses to drought timing can be more pronounced in seasonally dry tropical forests because of strong seasonality in climate and consequently on photosynthesis, allocation and growth (including dry seasons unfavourable to growth) [@garcía-cervigón_climate_2020].
-(Here we distinguish dry season from drought with differences in peridicity; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable and anomalous hot, dry conditions.)
-Seasonally dry tropical forests are also at thermal/water stress tolerance thresholds, making them more vulnerable to drought [@allen_will_2017], and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019].
+(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable and anomalous hot, dry conditions beyond the seasonal expectations.)
+Seasonally dry tropical forests are also at thermal and water stress tolerance thresholds for many species, making them more vulnerable to drought [@allen_will_2017], and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019].
 Analyses to date have found that the woody growth of tropical trees <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly vulnerable to dry season precipitation[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates.
-Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress, which might confer differential advantages in drought conditions.
+Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
+These strategies might confer differential advantages under drought conditions.
 
 <!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
-Trees in seasonally dry forests have strategies that span and interact across timescales to maximise growth and survival under periodic hot/dry conditions.  
-- evolutionary timescales - e.g. dry season deciduousness   
-- Deciduous species, with shorter duration with leaves can have acquisitive strategies during leaf on [@desouza_drought_2020].    
-- Species with drought tolerant traits like deep roots, more negative turgor loss point and ...
-can be more or less resistant to drought [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021].
-- ecophysiological - e.g. negative TLPs, plasticity in deciduousness (root/shoot)
-- These spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.
+Tree species in seasonally dry forests have diverse evolutionary strategies <!--that span and interact across timescales --->to maximise growth and survival under periodic hot/dry conditions. 
+These strategies can range along a spectrum from hydraulic safety to efficiency leading to tolerance or avoidance of dry conditions[@gonzález-m._diverging_2021]. 
+Dry season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid water loss and have acquisitive strategies during leaf on periods [@desouza_drought_2020].
+Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf turgor loss point (hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), large vessels (hydraulic efficiency) [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021].
+Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021].
+This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics. 
+<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
+Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls; e.g.
+some species regulate negative TLPs, leaf deciduousness and root depth, based on environmental cues leading to intraspecific or interannual variation potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; REFS Lasky?
+Hulshof?].
 
 <!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
-Individual growth responses to drought can also depend on the specific microenvironments they experience that can lead to distict atmospheric and hydrological stresses with ecological consequences.  
-- Larger trees suffer more growth declines during drought [@bennett_larger_2015], exposed crowns -> more thermal stress [@vinod_thermal_2023] although crown exposure -> more light -> more growth in normal years.
-- ecological timescales - e.g. distribution of trees in relation to wetness  
--  Species distribution of evergreen species covarying with soil moisture [@kunert_leaf_2021] and potentially affecting survival and growth.  
-- Larger trees have larger root systems, but do not necessarily access deeper water [@ref_from_Panama]  
-- Even when trees are accessing deeper water, this does not mean that they're in better shape during drought.
-Rather, trees that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021].
-- indeed, there is evidence that trees near streams undergo greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta] during drought  
-- Some of these adaptations may confer drought sensitivity and buffer stem growth declines during drought years, or conversely, may tradeoff with advantages in normal years, leading to a diversity of responses within a drought event.
-- Moreover, drought characteristics can positively or negatively interfere with these factors, leading to consistent or heterogenous sensitivities across drought events.
-
+Individual growth responses to drought can depend on the specific microenvironments that trees experience leading to varied atmospheric and hydrological stresses with distinct ecological consequences.
+Physiologically, trees experience drought stress as a combination of higher evaporative demand and higher thermal stress on leaves, and greater stress on roots because of low soil moisture, leading to greater stress on trees that have exposed crowns [@vinod_thermal_2023], shallower roots or growing in drier microenvironments [].
+Consequently, larger trees suffer more growth declines during drought [@bennett_larger_2015]; larger trees have more exposed crowns although crown exposure can allow access to light and increase growth in normal years.
+To some extent trees with belowground investment in larger and deeper root systems, or are in wetter locations have lower growth declines during drought because of continued access to water []. 
+However, these responses can potentially be different across species or with drought severity.
+<!--Larger trees have larger root systems, but do not necessarily access deeper water [@ref_from_Panama] -->
+Trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
+Indeed, there is evidence that trees near streams undergo greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta] during drought. 
+Finally, distribution of evergreen species covarying with soil moisture [@kunert_leaf_2021] and potentially affecting survival and growth. 
+<!-- Some of these adaptations may confer drought sensitivity and buffer stem growth declines during drought years, or conversely, may tradeoff with advantages in normal years, leading to a diversity of responses within a drought event.
+-->
+Drought characteristics may positively or negatively interfere with these evolutionary and physiological strategies, leading to consistent or heterogenous sensitivities across drought events.  
 <!--[from KAT: While all good content, I find this paragraph unnecessary/ distracting. If you'd really like to include content here, I'd work it into the methods.]
@@
 - Manual dendrometer band timeseries can represent a wide range of species strategies with equipment precision necessary to detect signals. --> 
-
  <!-- Within species, size, exposure and location can affect sensitivity to drought.  
@@
 Finally, we calculated the long-term mean for each variable as the mean of the values of the rolling mean across all years of the dendroband measurements, and the standard error around this mean. 
-We also used the SPEIbase dataset [REF] to calculate the Standardised Precipitation-Evapotranspiration Index, a drought severity index, calculated at the monthly scale. 
+We also used the SPEIbase dataset [@vicente-serrano_multiscalar_2010] to calculate the Standardised Precipitation-Evapotranspiration Index, a drought severity index, calculated at the monthly scale. 
 We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.
@@
 
-To account for potential biases in annual increments from dendrobands from water-related growth or shrinkage, we also calculated growth occurrence using a zero-growth assumption [@zweifel]. 
+To account for potential biases in annual increments from dendrobands from water-related growth or shrinkage, we also calculated growth occurrence using a zero-growth assumption [@zweifel_are_2016]. 
 Under this assumption, positive DBH change is scored as 1 and any change <=0 is scored as 0. 
-Timing of measurement and seasonal dynamics in tree hydration and bark dynamics could contribute to negative values in dendrometer band increments, unlike woody growth from tree ring datasets [@chitra-tarak].
+Timing of measurement and seasonal dynamics in tree hydration and bark dynamics could contribute to negative values in dendrometer band increments, unlike woody growth from tree ring datasets [@chitra-tarak_it_2015].
 To account for biases introduced by timing of measurements and seasonal dynamics, we used this simplifying assumption to corroborate the patterns observed with the sensitivity metric.
@@
 All statistical analyses were performed using R version 4.4.0 [@rcoreteam_language_2024]. 
-Mixed models were fit through Bayesian regressions using MCMC methods using the package _brms_ [REF].
+Mixed models were fit through Bayesian regressions using MCMC methods using the package _brms_ [@bürkner_brms_2017].
 We ran 4 MCMC chains for 3000 iterations with 1000 of these set as warmup.
-We extracted draws, calculated distributions using functions from the _posterior_ package [REF].
+We extracted draws, calculated distributions using functions from the _posterior_ package [@vehtari_ranknormalization_2021; @bürkner_posterior_2023].
 
@@
 
-## Drought sensitivity across distinct drought events {-}
+## Growth timeseries and Drought sensitivity {-}
 
@@
 
-## Predictors of species-level variation in drought sensitivity {-}
+## Species effects on drought sensitivity {-}
 
@@
 Moreover, models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`) but not in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).
+
+## Microenvironment effects drought sensitivity {-}
 
@@
 ```
-
-## Predictors of stem-level variation in drought sensitivity {-}
 

### 2025-04-01 doc/manuscript_hkk_drought_sensitivity.Rmd
discussion first pass

@@
 In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in a different drought. 
-The resultant amplification of the complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022].
+The resultant amplification of the complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].
 
@@
 Although our analysis spans representative spatial and ecological variation, we are limited in quantitatively analysing differences among droughts and variation in drought characteristics, because of low sample size.
--   Our analysis spans ecological and spatial/environmental variation realistic to tropical trees, and beyond previous such analyses.
-- But given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts.
-- Thus, quantitative analyses parsing out the specific effects of drought characteristics like relative influence of timing and degree of anomaly is currently impossible, limiting interpretations.  
-- However, as the record continues to grow, so does the potential to answer these questions.
+Given the diversity of factors that can influence tree growth and drought sensitivity in tropical tree communities, our analysis of 30 species across size classes and habitats spans this ecological and spatial (environmental) variation realistically and beyond the scope of previous such analyses.
+<!-- Long-timescale analyses with tree rings are limited by the fact that species with rings are few, represent specific strategy to water use and growth - and often affiliated to drier microenvironments.
+Tape measurements of growth from long-term ecological dynamics plots can be representative of species strategies but lack precision to parse signal from noise; error for each tape measure is 1mm, error for growth is at least 2mm while median annual growth across trees in HKK is ~1.4 mm and measuring deviation from this annual growth needs more precision.
+-->
+<!-- Traditional methods may only be capturing the most dramatic effects - the most severe droughts that affect ring-forming species or produce large-enough declines in fast-growing individuals.  -->
+Further, although dendroband timeseries can have measurement biases because of processes besides woody growth like stem water shrinkage and bark peeling [@chitra-tarak_it_2015], our results are robust to simplifying assumptions that account for these biases (Fig S--).
+However, given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts.
+Thus, quantitative analyses parsing out the specific effects of drought characteristics like relative influence of timing and degree of anomaly are currently impossible, limiting interpretations. 
+In temperate forests, long-term records from tree ring analyses could disentangle species and individual sensitivities to differing drought characteristics, as annual rings are formed by a majority of the species.
+Indeed @mcgregor_tree_2021 found that species sensitivities were different across drought events. 
+However, as this accumulating data record continues to grow, so does the potential to answer these questions across a wider range of forested ecosystems.
 
@@
 
-Tropical dry forests also represent periodic hot/dry conditions, leading to adaptations or other genetic variation for resistance, but this might not hold in other bioclimatic zones where ecological variability is different.
-- Trait variation across forest and bioclimatic types
-- Careful interpretation across forest types with different adaptive pressures.  
-- However, some generalisation is possible because despite low seasonality in climatic conditions, everwet forest species also show seasonal growth patterns.
+Tropical dry forests represent periodic hot/dry conditions, creating selective pressures for adaptative variation towards drought resistance, but this might not hold in other bioclimatic zones where adaptive pressures and thus ecological variability is different.
+Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies and has resulted in diversification along hydraulic safety-efficiency axis with diverse strategies to maximise tree vital rates [@gonzález-m._diverging_2021].
+The degree of adaptive variation along the water use axis might be comparable to other dry climatic systems (e.g.
+temperate drylands reported in @vasey_intraspecific_2022) but it may not apply to the wet or everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different.
+Therefore, we caution that our results suggestive of community-level insurance effects under drought be interpreted carefully across forest types that may be structured primarily by nutrient limitation, fires, herbivory or pathogens.
+However, our inference on the effects of drought timing could be generalisable across wet and dry tropics in the light of emerging evidence that despite low seasonality in climatic conditions, everwet forest species also show seasonal growth patterns [@giraldo_tree_2023].
 
 <!----Contributions beyond---->
-Despite consistent growth declines at the plot level across drought events, we show that this scale obscures a variety of ecological responses at the species and tree level that are heterogeneous across drought events.
-- Our finding that a wet season drought had bigger impact calls into question the generality of the findings of @zuidema_tropical_2022, which found that in species with tree-ring chronologies, dry season drought was most impactful.
-However, care must be taken in interpretation, as the 2015-16 ENSO was stronger.
-- Influence of deciduousness was different in the two different droughts. 
-- *Afzelia xylocarpa*, one of the tree ring species had the most distinct (positive) sensitivity in the dry season drought, but a large negative sensitivity in the wet season drought - one of the most distinct responses.
-- More generally, tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may be biased towards fast growing (measurable using tape) or dry-associated species (that form annual tree rings) - which might have been capturing only the more dramatic effects.
-- Distinct responses of species and individuals to drought characteristics suggests potential spatiotemporal insurance effects among tropical tree response to drought.
+Despite consistent observations of growth declines at the plot level across drought events, we show that this scale obscures a variety of ecological responses at the species and tree level that are heterogeneous across drought events.
+Our finding that a wet season drought had bigger impact calls into question the generality of the findings of @zuidema_tropical_2022, which found that in species with tree-ring chronologies, dry season drought was most impactful.
+(Care must be taken in interpretation, however, as the wet season drought (2015-16 ENSO) was stronger (Fig 1)) 
+<!-- Given the limitations of tree ring methods and tape measurements in measuring accurately measuring growth differences across a range of species in the tropics, methods so far could have led to capturing only the most dramatic droughts that affected (dry-associated) tree ring species and fast-growing individuals.
+-->
+*Afzelia xylocarpa*, one of the few annual-ring-forming species in HKK is noteworthy with the most distinct response across the two events; it had the highest (positive) sensitivity in the dry season drought, but a large negative sensitivity in the wet season drought (Fig---).
+More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects.
+In contrast, our analyses reveal a range of responses, including the contrasting influence of deciduousness in the two different droughts, showing that ecological and evolutionary strategies may result in heterogeneous consequences <!--in line with not trait, but species effects seen in -->.
+Distinct responses of species and individuals to drought characteristics suggests potential spatiotemporal insurance effects among tropical tree response to drought.
 
 <!----Science is better now---->
-Drought sensitivity is often defined as a static species trait but we show dimensions beyond that with possible consequences for management and restoration 
-- Although net decrease at the plot level (agreement with landscape-scale in @aguirre-guiterrez), we show evidence that the same strategies are not uniformly successfully across spatiotemporal climatic variation.
-- Frameworks that include ecological complexity and covarying traits with environment like @trugman_why_2021 also assume individual trees to have consistent sensitivity across events, but our results demonstrate temporal dimensions of variation, at least partially attributable to drought characteristics.
-- Our results indicate that drought responses in tropical trees may not be a fixed response across events, but jointly shaped by...
-- Given these inconsistencies, diverse stands that may represent greater variation in ecological and evolutionary strategies, may have mechanisms for stable functioning under future droughts.
+Drought sensitivity is often defined as a static species trait but our study reveals expanded dimensions of drought responses with possible consequences for management and restoration. 
+Despite net decreased growth at the plot level (in agreement with landscape-scale in @aguirre-gutiérrez_functional_2022 and @bennett_sensitivity_2023), we show evidence that the same strategies are not uniformly successfully across spatiotemporal climatic variation.
+Static traits are often used to explain and predict forest responses to drought, especially in the tropics where long-term records are few [@guillemot_small_2022; @vico_how_2017a]. 
+More inclusive frameworks for drought responses like @trugman_why_2021 encompass ecological complexity and covarying traits with environment like, but they also consistent sensitivity of individual trees across events.
+Our results, however, demonstrate an additional temporal dimension of variation, at least partially attributable to drought characteristics. 
+Our results indicate that drought responses in tropical trees may not be a fixed response across events, but jointly shaped by species adaptations and local environmental conditions. 
+Given these inconsistencies, diverse stands that may represent greater variation in ecological and evolutionary strategies, may have mechanisms for stable functioning under future droughts.
 
@@
 
-- (BCI light x size interaction in @rüger_growth_2011)
+<!-- - (BCI light x size interaction in @rüger_growth_2011)   -->
 

### 2025-04-02 doc/manuscript_hkk_drought_sensitivity.Rmd
cleaned up intro and half methods

@@
 
-^1^SCBI  
-^2^SERC  
-^3^STRI  
-^4^National Park, Wildlife and Plant Conservation Department, Thailand  
-^5^Kasetsart University, Thailand
+^1^Smithsonian's National Zoo and Conservation Biology Institute, Front Royal, VA, USA  
+^2^Smithsonian Environmental Research Centre, Edgewater, MD, USA  
+^3^Smithsonian Tropical Research Institute, Panama  
+^4^National Park, Wildlife and Plant Conservation Department, Bangkok, Thailand  
+^5^Kasetsart University, Chatuchak, Bangkok, Thailand
 
@@
 # Abstract (150 words) {-}
-The frequency and intensity of droughts are increasing across tropical forests, altering tree demography and its contribution to the land carbon sink. 
-In seasonally dry forests, species adaptations to strong dry seasons might confer drought resistance, but this resistance may be variable with individual microenvironments and across drought events. 
+The frequency and intensity of droughts are increasing across forests worldwide, altering tree demography and its contribution to the land carbon sink. 
+In seasonally dry tropical forests, species adaptations to strong dry seasons might confer drought resistance, but this resistance may be variable with individual microenvironments and across drought events.
 We analysed a unique data record - manual dendrometer band measurements over 15 years on 1820 trees across 30 species in a seasonally dry tropical forest in <!---Huai Kha Khaeng, ---->Thailand, <!--using Bayesian regressions in a causal modelling framework--->to assess the drought sensitivity of tree growth in two climatically distinct droughts.
@@
 In forests worldwide, intense drought is becoming more frequent, affecting forest demography, diversity, and  C cycling [@intergovernmentalpanelonclimatechangeipcc_climate_2023]. 
-Understanding these impacts is an important research priority, particularly for tropical forests, which <!--harbor high species diversity and --> are critical C sinks; tropical tree woody growth removes roughly `r round (150*1/6,0)` Gt CO~2~ annually and offsets ~`r round(7/36.3*100,-1)`% of anthropogenic fossil fuel emissions [@beer_terrestrial_2010;@anderson-teixeira_carbon_2021;@anderson-teixeira_carbon_2016;@brando_droughts_2019].
+Understanding these impacts is an important research priority, particularly for tropical forests, which <!--harbor high species diversity and --> are critical C sinks; tropical tree woody growth removes roughly `r round (150*1/6,0)` Gt CO~2~ annually offsetting ~`r round(7/36.3*100,-1)`% of anthropogenic fossil fuel emissions [@beer_terrestrial_2010;@anderson-teixeira_carbon_2021;@anderson-teixeira_carbon_2016;@brando_droughts_2019].
 Drought events can threaten this sink; the 2015-2016 ENSO drought in the Amazon temporarily shifted the region from a net sink to a net source of carbon [@brando_droughts_2019].
-However, the study of tropical tree woody growth responses to drought is inhibited by the fact that few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022]. 
-We know that drought characteristics [], species traits, and microenvironment [@bennett_larger_2015;@mcgregor_tree_2021] all affect tree growth responses. 
-However, we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought. 
-<!--The above statement is true at least to my knowledge offhand, but we need to check--> 
-Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; for example, a species with low sensitivity to past droughts would be among the least sensitive species in any microhabitat and through any drought.
+However, nuanced study of tropical tree woody growth responses to drought is inhibited by the fact that few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022]. 
+Separate studies have shown that drought characteristics species traits, and microenvironment all affect tree growth responses [@bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought.
+Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., a species with low sensitivity to past droughts would have low sensitivity in any microhabitat and through any drought.
 In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in a different drought. 
@@
 <!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
-Tree growth responses to drought timing can be more pronounced in seasonally dry tropical forests because of strong seasonality in climate and consequently on photosynthesis, allocation and growth (including dry seasons unfavourable to growth) [@garcía-cervigón_climate_2020].
-(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable and anomalous hot, dry conditions beyond the seasonal expectations.)
+Tree growth responses to drought timing can be pronounced in seasonally dry tropical forests because of strong seasonality in climate and consequently on photosynthesis, allocation and growth (including dry seasons unfavourable to growth) [@garcía-cervigón_climate_2020].
+(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.)
 Seasonally dry tropical forests are also at thermal and water stress tolerance thresholds for many species, making them more vulnerable to drought [@allen_will_2017], and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019].
@@
 Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021].
-This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics. 
+<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.
+-->
 <!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
@@
 Individual growth responses to drought can depend on the specific microenvironments that trees experience leading to varied atmospheric and hydrological stresses with distinct ecological consequences.
-Physiologically, trees experience drought stress as a combination of higher evaporative demand and higher thermal stress on leaves, and greater stress on roots because of low soil moisture, leading to greater stress on trees that have exposed crowns [@vinod_thermal_2023], shallower roots or growing in drier microenvironments [].
-Consequently, larger trees suffer more growth declines during drought [@bennett_larger_2015]; larger trees have more exposed crowns although crown exposure can allow access to light and increase growth in normal years.
-To some extent trees with belowground investment in larger and deeper root systems, or are in wetter locations have lower growth declines during drought because of continued access to water []. 
+Physiologically, trees experience drought stress as a combination of higher evaporative demand and higher thermal stress on leaves, and greater stress on roots because of low soil moisture. 
+This results in greater stress on trees that have exposed crowns [@vinod_thermal_2023], shallower roots or growing in drier microenvironments []. 
+Consequently, larger trees, with more exposed crowns, suffer more growth declines during drought [@bennett_larger_2015]<!--; larger trees have more exposed crowns although crown exposure can allow access to light and increase growth in normal years-->.
+<!-- To some extent trees with belowground investment in larger and deeper root systems, or are in wetter locations have lower growth declines during drought because of continued access to water [].
+-->
 However, these responses can potentially be different across species or with drought severity.
@@
 Trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
-Indeed, there is evidence that trees near streams undergo greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta] during drought. 
-Finally, distribution of evergreen species covarying with soil moisture [@kunert_leaf_2021] and potentially affecting survival and growth. 
+Indeed, there is evidence that trees near streams undergo greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta_droughtinduced_2017] during drought. 
+Finally, species strategies may constrain tree light and water microenvironments; e.g.
+evergreen species can covary with soil moisture [@kunert_leaf_2021], shade tolerant and understory species may be rarely exposed to hot/dry air.
 <!-- Some of these adaptations may confer drought sensitivity and buffer stem growth declines during drought years, or conversely, may tradeoff with advantages in normal years, leading to a diversity of responses within a drought event.
 -->
-Drought characteristics may positively or negatively interfere with these evolutionary and physiological strategies, leading to consistent or heterogenous sensitivities across drought events.  
+Whether drought characteristics interfere positively or negatively with these evolutionary and physiological strategies and whether it leads to consistent or heterogenous sensitivities across drought events is as yet unknown.
 <!--[from KAT: While all good content, I find this paragraph unnecessary/ distracting. If you'd really like to include content here, I'd work it into the methods.]
@@
 
+***Calculating annualised growth increments***
+
 Using the late wet/early dry dendrometer band censuses, we calculated annualised growth for each year for each tree. 
 We first removed measurements with likely data entry errors because of misplaced decimals, potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses).
-We used the raw window size measurements combined with the diameter at breast height (DBH) measurement at installation to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[condit].
-Since this was dependent on the diameter measurement at installation, we used a direct measurement at the time of installation if it was available.
+We used the raw window size measurements combined with the diameter measurement at installation $(dia_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[condit].
+We used a direct measurement of $dia_{1}$ at the time of installation if it was available.
 If the tape measurement at the first window size measurement was not available for any band, we excluded window size measurements until the first available diameter measurement. 
 We then calculated annualised diameter increments for each individual for each year by finding the difference in diameter from the previous year ($inc = dia_{t} - dia_{t-1}$) and adjusting for the number of days between measurements ($inc_{annual} = inc * \frac{tdif}{365}$).
+
 To ensure that the data did not smooth over multiple years, we excluded increment measures with missing values for $dia_{t-1}$ and then excluded large measurement outliers (> 3 standard deviations from the mean across all observations).
@@
 
-Combining the growth timeseries from dendrometer band measurements with climatic information, we identified two drought years of interest - 2010 and 2015.
-To identify drought years, we identified years with low plot-level growth rate and combined this with information on ENSO events that affected the landscape. 
-We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected.
-Moreover, broad definitions based on climate variables (often not measured at the location), may not capture ecological realities within the plot. 
-Out of the two years with low plot-level growth, 2010 was a moderate ENSO event while 2015 was a very strong ENSO event. 
-These years also corresponded with expert knowledge of major droughts that affected the region. 
+We identified drought years of interest in the dendroband timeseries - 2010 and 2015 - by examining two drought indices and corroborating with plot-level ecological responses and expert information. 
+Within the dendroband timeseries, two ENSO events had occurred; 2010 was a moderate ENSO event while 2015 was a very strong ENSO event [REF]. 
+<!-- We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected.
+-->
+We also used the SPEIbase dataset [@vicente-serrano_multiscalar_2010] to calculate the Standardised Precipitation-Evapotranspiration Index, a drought severity index, calculated at the monthly scale.
+Defining drought as periods with two or more consecutive months with SPEI < -1 also resulted in 2010 and 2015 as drought events. 
+These years also corresponded with expert knowledge of major droughts that affected the region as well as with plot-level growth declines in the timeseries. 
+<!-- We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.  -->
+<!-- Moreover, broad definitions based on climate variables (often not measured at the location), may not capture ecological realities within the plot.  -->
 We also note that we chose not to include 2020 in our analysis despite low growth, because a) declines in growth can occur due to various reasons which could have distinct ecological responses in species and individuals b) 2020 did not emerge as a drought year either in the examination of drought indicators or from expert knowledge.
@@
 We used daily weather data collected from 2001 to 2019 at the Forest Fire Research Station [CHECK WHERE], located ~20 km from the ForestGEO plot. 
-We first calculated daily values of Vapour Pressure Deficit using the equation $$
-0.6108 * e^{17.27 * \frac{t_{max}}{t_{max}+237.3}}* (1-\frac{relative.humidity}{100})
+We first calculated daily values of Vapour Pressure Deficit (VPD) using the equation $$
+VPD = 0.6108 * e^{17.27 * \frac{t_{max}}{t_{max}+237.3}}* (1-\frac{relative.humidity}{100})
 $$
 We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 15 day periods across the whole year for each year.
-Finally, we calculated the long-term mean for each variable as the mean of the values of the rolling mean across all years of the dendroband measurements, and the standard error around this mean. 
-We also used the SPEIbase dataset [@vicente-serrano_multiscalar_2010] to calculate the Standardised Precipitation-Evapotranspiration Index, a drought severity index, calculated at the monthly scale. 
-We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.
+Finally, we calculated the anomaly for each variable on each day of 2010 and 2015 from the long-term expectation as $$anomaly = \frac{rolling.mean - rolling.mean_{long-term}}{rolling.mean{sd}}$$
 
@@
 Tropical dry forests represent periodic hot/dry conditions, creating selective pressures for adaptative variation towards drought resistance, but this might not hold in other bioclimatic zones where adaptive pressures and thus ecological variability is different.
-Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies and has resulted in diversification along hydraulic safety-efficiency axis with diverse strategies to maximise tree vital rates [@gonzález-m._diverging_2021].
+Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies and has resulted in diversification along hydraulic safety-efficiency axis with diverse strategies to maximise tree vital rates [@gonzález-m._diverging_2021;@oliviera].
 The degree of adaptive variation along the water use axis might be comparable to other dry climatic systems (e.g.
@@
 Static traits are often used to explain and predict forest responses to drought, especially in the tropics where long-term records are few [@guillemot_small_2022; @vico_how_2017a]. 
-More inclusive frameworks for drought responses like @trugman_why_2021 encompass ecological complexity and covarying traits with environment like, but they also consistent sensitivity of individual trees across events.
+<!--aligns with remote sensing based inferences across global forests @liu_drought_2022-->
+More inclusive frameworks for drought responses like @trugman_why_2021 (on mortality) encompass ecological complexity and covarying traits with environment like, but they assume consistent sensitivity of individual trees across events.
 Our results, however, demonstrate an additional temporal dimension of variation, at least partially attributable to drought characteristics. 

### 2025-04-02 doc/manuscript_hkk_drought_sensitivity.Rmd
methods cleaned

@@
 <!--[from KAT: While all good content, I find this paragraph unnecessary/ distracting. If you'd really like to include content here, I'd work it into the methods.]
-Given the diversity of factors that can influence drought sensitivity in seasonally dry tropical forests, sampling strategies need to effectively span this variation and have adequate precision.  
-- Species with rings are few, represent specific strategy to water use and growth - and often affiliated to drier microenvironments.  
-- Tape measurements can be representative (although often not annual), not enough precision; error for each tape measure is 1mm, error for growth is at least 2mm.
-(Median annual growth across trees in HKK is ~1.4 mm, deviation from annual growth needs more precision).
 - In sum, these methods may only be capturing the most dramatic effects - the most severe droughts that affect ring-forming species or produce large-enough declines in fast-growing individuals.  
@@
 To ensure that the data did not smooth over multiple years, we excluded increment measures with missing values for $dia_{t-1}$ and then excluded large measurement outliers (> 3 standard deviations from the mean across all observations).
-We repeated the same steps using annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement with these annualised increments calculated from tape measurements.
+Tape measurements themselves are not precise enough for calculating drought sensitivity; error for each tape measure is 1mm, error for growth is at least 2mm. 
+In comparison, median annual growth across trees in HKK is ~1.4 mm and calculating deviation from annual growth needs sub-mm precision. 
+However, tape measurements can be used to check for consistency in trends or identify outliers. 
+We calculated increments from annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement with these annualised increments calculated from tape measurements.
 Finally, we excluded trees with negative increments over the whole timeseries, and included only species that had at least 10 individuals, resulting in a final dataset of 1820 individuals across 30 species.
@@
 We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 15 day periods across the whole year for each year.
-Finally, we calculated the anomaly for each variable on each day of 2010 and 2015 from the long-term expectation as $$anomaly = \frac{rolling.mean - rolling.mean_{long-term}}{rolling.mean{sd}}$$
+Finally, we calculated daily rolling anomalies for each variable for 2010 and 2015 from the long-term daily expectation as $$anomaly_{day} = \frac{rolling.mean_{day} - \bar{rolling.mean_{day}}}{sd(rolling.mean_{day})}$$
 
@@
 
-For the two drought years, 2010 and 2015, we calculated growth sensitivity for each tree to each drought as a growth anomaly from its mean annual increment.
+For the two drought years, 2010 and 2015, we calculated growth sensitivity for each tree to each drought as a growth anomaly from its mean annual increment from calculated from dendrometer band window size.
+
 We first calculated the mean annual increment for each tree as the mean of the annualised increments across the whole timeseries.
-We then calculated sensitivity for each tree for each drought as 
+We then calculated sensitivity for each tree $i$ for each drought as 
 $$
@@
 
-To account for potential biases in annual increments from dendrobands from water-related growth or shrinkage, we also calculated growth occurrence using a zero-growth assumption [@zweifel_are_2016]. 
-Under this assumption, positive DBH change is scored as 1 and any change <=0 is scored as 0. 
+To account for potential biases in dendrometer band increments from water-related growth or shrinkage, we also calculated growth occurrence using a zero-growth assumption [@zweifel_are_2016]. 
 Timing of measurement and seasonal dynamics in tree hydration and bark dynamics could contribute to negative values in dendrometer band increments, unlike woody growth from tree ring datasets [@chitra-tarak_it_2015].
+Using a zero-growth model, we calculated growth occurrence as 1 under positive diameter change and any change <=0 as 0. 
 To account for biases introduced by timing of measurements and seasonal dynamics, we used this simplifying assumption to corroborate the patterns observed with the sensitivity metric.
@@
 
-We calculated three species-level traits : deciduousness, maximum size and niche breadth. 
+We calculated deciduousness as a species level trait using data published in @williams_deciduousness_2008 for the top 30 species. 
 We defined deciduousness as the mean proportion of canopy loss at maximum loss across individuals in the species, with a score ranging from 1 to 4 where 1 is most evergreen and 4 is most deciduous.  
-We used data published in @williams_deciduousness_2008 for the top 30 species. 
-Out of these, data was unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1, in the middle of the evergreen range. 
+Data was unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1, in the middle of the evergreen range. 
 An alternate measure of deciduousness is the duration spent at maximum crown loss.
 This data was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for a larger list of species.  
-At the species level, we also calculated the maximum size across all individuals in the plot as measured in the 2008 whole-plot census.
-We calculated niche breadth as the standard deviation of TWI across all observed individuals of each species in the whole plot census in 2008.
+<!-- At the species level, we also calculated the maximum size across all individuals in the plot as measured in the 2008 whole-plot census.
+We calculated niche breadth as the standard deviation of TWI across all observed individuals of each species in the whole plot census in 2008.  -->
 
@@
 
-We then used the predicted sensitivities from this model and tested their correlations with three species-level traits - deciduousness, maximum size and niche breadth using simple linear models.
+We then used the predicted sensitivities from this model and tested their correlations with deciduousness using simple linear models.
 
-To model the interactive effects of deciduousness on sensitivity, we modelled
+To model the interactive effects of deciduousness and wetness on sensitivity, we modelled
 
@@
 
-To further disentange the drivers of stem-level drought sensitivity, we simultaneously modelled:
+To further disentange the microenvironmental drivers of drought sensitivity, we simultaneously modelled:
 
@@
 We ran 4 MCMC chains for 3000 iterations with 1000 of these set as warmup.
-We extracted draws, calculated distributions using functions from the _posterior_ package [@vehtari_ranknormalization_2021; @bürkner_posterior_2023].
+We extracted draws, calculated distributions and checked convergence using functions from the _posterior_ package [@vehtari_ranknormalization_2021; @bürkner_posterior_2023].
 

### 2025-04-02 doc/manuscript_hkk_drought_sensitivity.Rmd
results done, acknowledgements draft
need to add supplementary figures

@@
 
-The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Fig 1).
+The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig 1).
 In 2010, drought characteristics of lower precipitation, higher number of dry days, higher maximum temperature, higher VPD than the long-term mean and SPEI values below -1 occurred between February and April, in the dry season.
@@
 While there were more species differences in median responses to the 2010 drought, all species had lower increment than a typical year in 2015. 
-Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years. 
+Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S--). 
 Across the years, on average, `r round(100*mean(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% of the trees had negative growth, with a minimum of `r round(100*min(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% in `r names(sort(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))))[1]`.
@@
 Deciduousness was associated with lower negative sensitivities in 2015, but not in 2010 (Fig 3a). 
-Species median sensitivities in 2010 and 2015 were not correlated with each other (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`) but on average had lower mean and median sensitivities in 2015 than in 2010, in alignment with median sensitivities across all trees.
+Species median sensitivities in 2010 and 2015 were not correlated with each other (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S--) but on average had lower mean and median sensitivities in 2015 than in 2010 (Fig S--), in alignment with median sensitivities across all trees.
 In 2015, predicted sensitivities from intercept-only models of sensitivity with a species random effect on the intercept showed a significant negative association with deciduousness value (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`).
@@
 
-The effect of TWI on sensitivity was different in 2010 and 2015, with 2015 effects interacting with deciduousness. 
+The effect of wetness on sensitivity was different in 2010 and 2015, with 2015 effects interacting with deciduousness. 
 Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`).
@@
 DBH was a significant predictor of exposure category, with larger DBH associated with higher exposure categories in 2010 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`) and 2015 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) (Fig 5a).
-Controlling for this direct effect of exposure, the effect of exposure, the remaining effect of size *per se* was significantly positive in 2010, while it was weakly negative in 2015 (Fig 4a, Fig 5a).
+Controlling for this direct effect of exposure, the remaining effect of size *per se* was significantly positive in 2010, while it was weakly negative in 2015 (Fig 4a, Fig 5a).
 
@@
 Analysing two droughts in a 14-year dendrometer band timeseries, we found that tropical tree species and individuals grew less than average in drought years, reflecting in lower plot-level growth (Fig 2).
-Comparing across droughts, we found that despite the shorter duration of drought conditions, the wet-season drought elicited stronger responses across trees, but especially so for deciduous species that are expected to have stronger seasonality in growth phenology.
-Interestingly, in the wet-season drought, we also show that evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites.
-Further, besides consistent negative effects of crown exposure, we found that the drivers of variation among species and individuals in drought sensitivity were dissimilar between the drought events. 
-Water availability buffered exposure-associated growth declines through different mechanisms in the two different droughts - larger trees (with likely deeper roots) in the dry-season drought and trees in wetter sites in the wet-season drought suffered smaller declines in growth.
+Comparing across droughts, we found that despite the shorter duration of drought conditions, the wet-season drought elicited stronger responses across trees, but especially so for deciduous species that are expected to have stronger seasonality in growth phenology (Fig 3).
+Interestingly, in the wet-season drought, we also show that evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites (Fig 3b, Fig 4b).
+Further, besides consistent negative effects of crown exposure, we found that the drivers of variation among species and individuals in drought sensitivity were dissimilar between the drought events (Fig 4a).
+Water availability buffered exposure-associated growth declines through different mechanisms in the two different droughts - larger trees (with likely deeper roots) in the dry-season drought and trees in wetter sites in the wet-season drought suffered smaller declines in growth (Fig 5).
 Taken together, our analysis of tree growth in two different drought years, and spanning meaningful variation, shows that species and individuals can respond differently to different drought events. 
@@
 
-Huai Kha Khaeng forest monitoring team.  
-Ekaphan Kraichak, Kanisorn Chowtiwuttakorn.  
-Jose Medina-Vega, Valentine Herrmann, Eugenie Mas. 
-Smithsonian Institution Postdoctoral Fellowship. 
-ForestGEO Analytical Workshop, Asia 2024.
+We acknowledge the continued contributions of the Huai Kha Khaeng forest monitoring team in collecting data and maintaining the dendrometer bands. 
+Ekaphan Kraichak and Kanisorn Chowtiwuttakorn helped with procuring elevation and climate datasets. 
+Jose Medina-Vega, Valentine Herrmann and Eugenie Mas provided input on improving analyses.  
+KA was supported by the Smithsonian Institution Postdoctoral Fellowship and the ForestGEO Analytical Workshop, Asia 2024 in completing this work.
 
@@
 
-# Author contributions {-}
-
-# Data availability {-}
-
 \newpage

### 2025-04-02 doc/manuscript_hkk_drought_sensitivity.Rmd
edited discussion

@@
 Taken together, our analysis of tree growth in two different drought years, and spanning meaningful variation, shows that species and individuals can respond differently to different drought events. 
-Despite stand-level growth declines in drought years, we show that these broad patterns mask enormous heterogeneity in individual and species responses to drought in tropical forests with possible consequences for the maintenance of species diveristy and ecosystem stability.
+Despite stand-level growth declines in drought years, we show that these broad patterns belie enormous heterogeneity in individual and species responses to drought in tropical forests with possible consequences for the maintenance of species diveristy and ecosystem stability.
 
 <!----Limitations in filling gap---->
-Although our analysis spans representative spatial and ecological variation, we are limited in quantitatively analysing differences among droughts and variation in drought characteristics, because of low sample size.
+Our analysis spans representative spatial and ecological variation, but we are limited in quantitatively analysing differences among droughts and variation in drought characteristics, because of low sample size.
 Given the diversity of factors that can influence tree growth and drought sensitivity in tropical tree communities, our analysis of 30 species across size classes and habitats spans this ecological and spatial (environmental) variation realistically and beyond the scope of previous such analyses.
@@
 <!-- Traditional methods may only be capturing the most dramatic effects - the most severe droughts that affect ring-forming species or produce large-enough declines in fast-growing individuals.  -->
-Further, although dendroband timeseries can have measurement biases because of processes besides woody growth like stem water shrinkage and bark peeling [@chitra-tarak_it_2015], our results are robust to simplifying assumptions that account for these biases (Fig S--).
+Further, although estimates of woody growth from dendroband timeseries can have measurement biases because of stem water shrinkage and bark peeling [@chitra-tarak_it_2015], our results are robust to simplifying assumptions that account for these biases (Fig S--).
 However, given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts.
@@
 Indeed @mcgregor_tree_2021 found that species sensitivities were different across drought events. 
-However, as this accumulating data record continues to grow, so does the potential to answer these questions across a wider range of forested ecosystems.
+However, as this accumulating dendrometer data record continues to grow, so does the potential to answer these questions across a wider range of forested ecosystems.
 
@@
 Tropical dry forests represent periodic hot/dry conditions, creating selective pressures for adaptative variation towards drought resistance, but this might not hold in other bioclimatic zones where adaptive pressures and thus ecological variability is different.
-Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies and has resulted in diversification along hydraulic safety-efficiency axis with diverse strategies to maximise tree vital rates [@gonzález-m._diverging_2021;@oliviera].
+Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies and has resulted in diversification along the hydraulic safety-efficiency axis with diverse strategies to maximise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021].
 The degree of adaptive variation along the water use axis might be comparable to other dry climatic systems (e.g.
@@
 Our finding that a wet season drought had bigger impact calls into question the generality of the findings of @zuidema_tropical_2022, which found that in species with tree-ring chronologies, dry season drought was most impactful.
-(Care must be taken in interpretation, however, as the wet season drought (2015-16 ENSO) was stronger (Fig 1)) 
+(Care must be taken in interpretation, however, as the 2015-2016 wet season ENSO drought was stronger (Fig 1)). 
 <!-- Given the limitations of tree ring methods and tape measurements in measuring accurately measuring growth differences across a range of species in the tropics, methods so far could have led to capturing only the most dramatic droughts that affected (dry-associated) tree ring species and fast-growing individuals.
 -->
-*Afzelia xylocarpa*, one of the few annual-ring-forming species in HKK is noteworthy with the most distinct response across the two events; it had the highest (positive) sensitivity in the dry season drought, but a large negative sensitivity in the wet season drought (Fig---).
+*Afzelia xylocarpa*, one of the few annual-ring-forming species in HKK is noteworthy with the most distinct response across the two events; it had the highest (positive) sensitivity in the dry season drought, but a large negative sensitivity in the wet season drought (Fig S--).
 More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects.
-In contrast, our analyses reveal a range of responses, including the contrasting influence of deciduousness in the two different droughts, showing that ecological and evolutionary strategies may result in heterogeneous consequences <!--in line with not trait, but species effects seen in -->.
-Distinct responses of species and individuals to drought characteristics suggests potential spatiotemporal insurance effects among tropical tree response to drought.
+In contrast, our analyses reveal a range of responses, including the contrasting influence of deciduousness in the two different droughts, showing that ecological and evolutionary strategies may result in heterogeneous consequences.
+Distinct responses of species and individuals to drought characteristics suggests potential spatiotemporal insurance effects among tropical tree response to drought in alignment with recent global analyses (@liu_drought_2022).
 

### 2025-04-02 doc/supplementary_information.Rmd
some images into SI file
updated SI
add corr plot to supplement
updated SI file

@@
 ---
-title: "Supplementary Information: Vulnerability of tree growth"
+title: "Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments"
 author: 
@@
 
+**Authors**  
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+# library(flextable)
+# set_flextable_defaults(font.family = "Arial", font.size = 9)
+team <- read.csv("display/authors.csv", stringsAsFactors = FALSE, check.names = FALSE)
+
+# team <- read.csv("doc/display/authors.csv", stringsAsFactors = FALSE, check.names = FALSE)
+
+# ft <- flextable(team)
+# ft <- line_spacing(ft, space = 1, part = "all")
+# ft <- autofit(ft)
+# ft <- width(ft, 1, width = 1.75)
+# ft <- width(ft, 2, width = 1.75)
+# ft <- width(ft, 3, width = 3)
+# ft
+```
+
+Krishna Anujan^1,2^*  ORCiD: `r team$ORCID[1]`  
+Sean McMahon^2,3^  ORCiD: `r team$ORCID[2]`  
+Sarayudh Bunyavejchewin^4^  ORCiD: `r team$ORCID[3]`  
+Stuart J. Davies^3^  ORCiD: `r team$ORCID[4]`  
+Helene C. Muller-Landau^3^  ORCiD : `r team$ORCID[5]`  
+Nantachai Pongpattananurak^5^  ORCiD: `r team$ORCID[6]`  
+Kristina Anderson-Teixeira^1,3^  ORCiD: `r team$ORCID[7]`
+
 ```{r read files, echo=F}
@@
 
+<!-- ![](display/env_vars_SI.png) -->
+
 \newpage
@@
 
+![](display/Fig_SI_species_sensitivities.png)
+
+![](display/Fig_SI_sensitivity_cor_dec.png)
+
+![](display/Fig_trait_intercept_plot.png)
+
+\newpage
+
+## Individual correlations
+
+![](display/sens_corr.png)
+
+## Variables and their distributions
+
+![](display/explore/twi.png)
+
+![](display/explore/dbh_cii.png)
+
 \newpage
@@
 \newpage
+
+## Additional models

### 2025-04-03 doc/manuscript_hkk_drought_sensitivity.Rmd
updated figures with panel numbers, Fig 1 background

@@
 # Data Accessibility Statement {-}
+
+# Conflict of Interest {-}
 
@@
 ![***Figure 1.
-Climatic characteristics of drought years and average years*** in Huai Kha Khaeng ForestGEO plot.
-For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1_anom_2.png)
+Climatic characteristics of drought years and average years*** in the Huai Kha Khaeng ForestGEO plot.
+For four daily climate variables dry days, precipitation, maximum temperature and VPD, panel a) represents rolling means for 30 days around the focal day for that year.
+Grey line represents the mean of this rolling mean from 2008 to 2019 and ribbons represent standard deviation around this mean.
+Panel b) represents anomalies of this rolling mean from the values across the timeseries.
+Values, therefore represents the number of standard deviations from the mean on a given day.
+c) shows Standardised Precipitation Evapotranspiration Index (SPEI) values for each month in 2010 and 2015.
+Across all panels, light blue background represents the average wet season, from May to October.
+For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1.png)
 
@@
 
-# Competing interests {-}
-
 \newpage

### 2025-04-03 scripts/06_figures.R

@@ -1,6 +1,6 @@
 # script to code for figures
 
-# load libraries
+# load libraries ---------------------------
 library(tidyverse)
 library(lubridate)
 library(ggplot2)
@@ -96,18 +96,22 @@ varnames <- as_labeller(c(
 
 
 climplot <- ggplot() +
+    # geom_rect(
+    #     data = data.frame(xmin = c(305, 1), xmax = c(366, 120), ymin = -Inf, ymax = Inf),
+    #     aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), fill = "grey", alpha = 0.3
+    # ) +
     geom_rect(
-        data = data.frame(xmin = c(305, 1), xmax = c(366, 120), ymin = -Inf, ymax = Inf),
-        aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), fill = "grey", alpha = 0.3
+        data = data.frame(xmin = 121, xmax = 304, ymin = -Inf, ymax = Inf),
+        aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), fill = "lightblue", alpha = 0.3
     ) +
     geom_line(data = ffstation_full_rl, aes(x = day_of_year, y = rlval, color = year, linetype = year), linewidth = 0.8) +
     facet_wrap(~climvar, scales = "free_y", labeller = varnames, strip.position = "left", nrow = 4) +
     theme_bw() +
     scale_linetype_manual(values = c("long-term" = "longdash", "2010" = "solid", "2015" = "solid")) +
     geom_ribbon(data = ffstation_lt_rl %>% filter(climvar %in% c("Precipitation", "dry_days", "TempMax", "vpdmax")), aes(x = day_of_year, ymin = rlval - rlsd, ymax = rlval + rlsd), alpha = 0.3) +
-    # text for wet and dry season
-    annotate("text", x = -Inf, y = -Inf, label = "dry season", col = "black", hjust = -0.5, vjust = -1, fontface = "italic", size = 1.15) +
-    annotate("text", x = -Inf, y = -Inf, label = "wet season", col = "black", hjust = -2.5, vjust = -1, fontface = "italic", size = 1.15) +
+    # # text for wet and dry season
+    # annotate("text", x = -Inf, y = -Inf, label = "dry season", col = "black", hjust = -0.5, vjust = -1, fontface = "italic", size = 1.15) +
[…]

### 2025-04-03 doc/manuscript_hkk_drought_sensitivity.Rmd
figure captions #22

@@
 2010 and 2015 were moderate and very strong ENSO events respectively.
-b) Dsitribution of drought sensitivity of tree growth across all individuals in the two drought years.
+b) Distribution of drought sensitivity of tree growth across all individuals in the two drought years.
 For comments - https://github.com/forestgeo/growth-precip-thailand/issues/11](display/Fig2.png)
@@
 ![***Figure 3: Variation in sensitivity among species*** in the Huai Kha Khaeng ForestGEO plot during droughts in 2010 and 2015.
+a) Relationship of the mean species sensitivities from model fits of an intercept-only model, with mean species deciduousness values.
+Line shown for significant correlation.
+b) Predicted sensitivity values across the observed range of deciduousness and topographic wetness index from a model with TWI, deciduosness and their interaction across all trees modelled for each drought year separately.
 For comments: https://github.com/forestgeo/growth-precip-thailand/issues/12](display/Fig3.png)
@@
 
-![***Figure 4: Drivers of variation in sensitivity among individuals***. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/13](display/Fig4.png)
+![***Figure 4: Drivers of variation in sensitivity among individuals***.
+All panels represent model results from models (one for each year) predicting sensitivity with microenvironment conditions with species random effect on all slopes.
+a) Coefficient plots showing median effects and 90% CI for wetness, exposure and DBH of tree.
+b) wetness effect across all species and its relationship with deciduousness value.
+Line represents a significant correlation.
+c) Predicted relationship of drought sensitivity with topographic wetness index across all species derived from model predictions.
+Black line represents overall relationship; coloured lines for each species reflects deciduousness values.
+For comments - https://github.com/forestgeo/growth-precip-thailand/issues/13](display/Fig4.png)
 
@@
 
-![***Figure 5: Effects of size, exposure and water availability*** For comments - https://github.com/forestgeo/growth-precip-thailand/issues/14](display/Fig5.png)
+![***Figure 5: Effects of size, exposure and water availability*** from combined models.
+Hypothesised relationships between microenvironment variables and drought sensitivity with their modelled effects in the HKK ForestGEO plot in the droughts of a) 2010 and b) 2015.
+Solid lines represent relationships where 90% credible intervals do not overlap 0, blue and red lines represent positive and negative effects respectively and line thickness is scaled to the effect size.
+c) Modelled sensitivity across crown exposure classes in 2010 and 2015.
+For comments - https://github.com/forestgeo/growth-precip-thailand/issues/14](display/Fig5.png)
 

### 2025-04-03 scripts/06_figures.R
new SI plots with full climate data series

@@ -46,17 +46,17 @@ ffstation_monthly_rl <- ffstation %>%
     ) %>%
     select(-c(rlsum, rlmean))
 
-ffstation_lt_rl <- ffstation_monthly_rl %>%
-    filter(year <= 2021, year >= 2009) %>%
-    ungroup() %>%
-    group_by(day_of_year, climvar) %>%
-    dplyr::summarise(
-        # rlse = sd(rlval, na.rm = T) / sqrt(sum(!is.na(rlval))),
-        rlval = mean(rlval, na.rm = T)
-    ) %>%
-    ungroup() %>%
-    dplyr::mutate(year = "long-term")
-# select(year, month, Date, climvar, rlval, rlse, day_of_year)
+# ffstation_lt_rl <- ffstation_monthly_rl %>%
+#     filter(year <= 2021, year >= 2009) %>%
+#     ungroup() %>%
+#     group_by(day_of_year, climvar) %>%
+#     dplyr::summarise(
+#         # rlse = sd(rlval, na.rm = T) / sqrt(sum(!is.na(rlval))),
+#         rlval = mean(rlval, na.rm = T)
+#     ) %>%
+#     ungroup() %>%
+#     dplyr::mutate(year = "long-term")
+# # select(year, month, Date, climvar, rlval, rlse, day_of_year)
 
 ffstation_lt_rl <- ffstation_monthly_rl %>%
     filter(year <= 2021, year >= 2009) %>%
@@ -74,14 +74,13 @@ ffstation_lt_rl_long <- ffstation_lt_rl %>%
     dplyr::mutate(year = ifelse(stat == "rlsd", "long-term.sd", year)) %>%
     select(-stat)
 
-
 # bind these two dataframes
 ffstation_full_rl <- bind_rows(ffstation_monthly_rl, ffstation_lt_rl) %>%
     filter(year %in% c("long-term", "2010", "2015")) %>%
     filter(climvar %in% c("Precipitation", "dry_days", "TempMax", "vpdmax"))
 
 ffstation_full_rl_long <- bind_rows(ffstation_monthly_rl, ffstation_lt_rl_long) %>%
-    filter(year %in% c("long-term", "long-term.sd", "2010", "2015")) %>%
+    # filter(year %in% c("long-term", "long-term.sd", "2010", "2015")) %>%
     filter(climvar %in% c("Precipitation", "dry_days", "TempMax", "vpdmax"))
 
[…]

### 2025-04-03 doc/manuscript_hkk_drought_sensitivity.Rmd
updated Fig 1 with code fixed
updated doc with accents to names

@@
 # ft <- width(ft, 3, width = 3)
+ft <- set_caption(ft, caption = "Summary table of hypotheses tested, effects of variables on drought sensitivity and corresponding evidence")
 ft
@@
 Ekaphan Kraichak and Kanisorn Chowtiwuttakorn helped with procuring elevation and climate datasets. 
-Jose Medina-Vega, Valentine Herrmann and Eugenie Mas provided input on improving analyses.  
+José Medina-Vega, Valentine Herrmann and Eugenie Más provided input on improving analyses. 
 KA was supported by the Smithsonian Institution Postdoctoral Fellowship and the ForestGEO Analytical Workshop, Asia 2024 in completing this work.

### 2025-04-03 doc/supplementary_information.Rmd
supplementary with updated clim figs

@@
 # Read in files
+
+# figure numbers
+
+# figure titles
 ```
@@
 
-![](../results/plots/climate_data_timeseries_2007_2019.png)
+![Figure S1: Raw climate data timeseries](display/climvars_fullseries.png)
 
-![](../results/plots/climate_data_correlations.png)
-
-<!-- ![](display/env_vars_SI.png) -->
-
+![Figure S2: Climate anomalies across the whole timeseries](display/climvars_anomalies_fullseries.png)
 \newpage
@@
 ## Data cleaning additional methods
+
+- Data cleaning steps
+- Calculating DBH equation
+- Corroboration with tape measurements
+- error distribution
 
@@
 
-## Individual correlations
+## Individual growth and correlations
+
+- Negative growth proportions across the years
 
@@
 
-## Individual model diagnostics
+<!-- ## Individual model diagnostics
 
-\newpage
+To add:
 
-## Additional models
+\newpage -->
+
+## Growth occurrence models
+
+To add: zero growth models

### 2025-04-03 scripts/06_figures.R
updated SI plots

@@ -612,13 +612,13 @@ spagplot_top10_anom <- ggplot() +
         position = "dodge"
     ) +
     # make all years show on x-axis
-    scale_x_discrete() +
+    scale_x_continuous(breaks = unique(tree.time$yr)) +
     scale_fill_viridis_d() +
     geom_hline(yintercept = c(-2, -1, 0, 1, 2), linetype = "dashed") +
     # add text on these lines
     geom_text(aes(x = c(2010, 2015), y = 1, label = "ENSO drought"), hjust = 0.8, vjust = -0.2, angle = 90) +
     guides(fill = guide_legend("species"), nrow = 3) +
-    xlab("species") +
+    xlab("year") +
     ylab("anomaly") +
     # ggtitle("growth increments for top 10 species") +
     theme_bw() +
@@ -632,14 +632,99 @@ spagplot_top10_anom <- ggplot() +
 
 spagplot_top10_anom
 
-png("doc/display/Fig2_anom.png", width = 10, height = 8, units = "in", res = 300)
-spagplot_top10_anom + sens.all + plot_annotation(tag_levels = "a") +
-    # plot_layout(guides = "collect", widths = c(1, 2.2)) & theme(
-    plot_layout(guides = "collect") & theme(
-    legend.position = "bottom",
-    legend.margin = margin(),
-    legend.text = element_text(face = "italic")
-)
+png("doc/display/growth_anom_fullseries.png", width = 8, height = 8, units = "in", res = 300)
+spagplot_top10_anom
+dev.off()
+
+## SI plots - negative growth increments
+neg.incs.sp <- tree.time %>%
+    # filter(inc_annual < 0) %>%
+    group_by(yr, Species) %>%
+    dplyr::summarise(
+        negs = sum(inc_annual < 0, na.rm = T),
+        n = n()
+    ) %>%
+    ungroup() %>%
+    dplyr::mutate(
+        Species = factor(Species, levels = unique(Species))
+    )
+
+neg.incs <- neg.incs.sp %>%
+    group_by(yr) %>%
+    dplyr::summarise(
+        negs = sum(negs, na.rm = T),
+        n = sum(n)
+    ) %>%
+    ungroup() %>%
+    dplyr::mutate(
+        Species = "all",
+        forcol = ifelse(yr == 2010, "2010",
+            ifelse(yr == 2015, "2015", "other")
+        ),
[…]

### 2025-04-03 doc/supplementary_information.Rmd
updated SI plots
updated SI with some fig titles

@@
 
-## Species traits and correlations
+## Growth timeseries and anomalies
 
-![](display/Fig_SI_species_sensitivities.png)
+![Fig S3: Growth anomalies across the timeseries](display/growth_anom_fullseries.png)
 
-![](display/Fig_SI_sensitivity_cor_dec.png)
+![Fig S4: Proportion of negative growth increments](display/neg_incs.png)
 
-![](display/Fig_trait_intercept_plot.png)
+## Drought year growth of species and individuals
 
-\newpage
+![Fig S4: Species sensitivities in drought years](display/Fig_SI_species_sensitivities.png)
 
-## Individual growth and correlations
+![Fig S5: Correlations of species sensitivities across the two years](display/Fig_SI_sensitivity_cor_dec.png)
 
-- Negative growth proportions across the years
+<!-- ![](display/Fig_trait_intercept_plot.png) -->
 
-![](display/sens_corr.png)
+![Fig S6: Correlation of individual sensitivities across the two years](display/sens_corr.png)
 
@@
 ![](display/explore/dbh_cii.png)
+
+- Conditional dependence of variables
 

### 2025-04-03 doc/manuscript_hkk_drought_sensitivity.Rmd
current figure numbers from SI into table and text #22

@@
 While there were more species differences in median responses to the 2010 drought, all species had lower increment than a typical year in 2015. 
-Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S--). 
+Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S4). 
 Across the years, on average, `r round(100*mean(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% of the trees had negative growth, with a minimum of `r round(100*min(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% in `r names(sort(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))))[1]`.
@@
 Deciduousness was associated with lower negative sensitivities in 2015, but not in 2010 (Fig 3a). 
-Species median sensitivities in 2010 and 2015 were not correlated with each other (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S--) but on average had lower mean and median sensitivities in 2015 than in 2010 (Fig S--), in alignment with median sensitivities across all trees.
+Species median sensitivities in 2010 and 2015 were not correlated with each other (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S5) but on average had lower mean and median sensitivities in 2015 than in 2010 (Fig S5), in alignment with median sensitivities across all trees.
 In 2015, predicted sensitivities from intercept-only models of sensitivity with a species random effect on the intercept showed a significant negative association with deciduousness value (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`).
@@
 The effect of wetness on sensitivity was different in 2010 and 2015, with 2015 effects interacting with deciduousness. 
+Individual tree responses to drought were not correlated with each other (Fig S6). 
 Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`).
@@
 -->
-*Afzelia xylocarpa*, one of the few annual-ring-forming species in HKK is noteworthy with the most distinct response across the two events; it had the highest (positive) sensitivity in the dry season drought, but a large negative sensitivity in the wet season drought (Fig S--).
+*Afzelia xylocarpa*, one of the few annual-ring-forming species in HKK is noteworthy with the most distinct response across the two events; it had the highest (positive) sensitivity in the dry season drought, but a large negative sensitivity in the wet season drought (Fig S5).
 More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects.

### 2025-04-03 doc/manuscript_hkk_drought_sensitivity.Rmd
updated with word counts etc

@@
 
-**Number of Words in Abstract:**  
-**Number of Words in Main Text:**  
-**Number of References:**  
+**Number of Words in Abstract:** 173  
+**Number of Words in Main Text:** 4989  
+**Number of References:** 46 
 **Number of Figures:** 5  

### 2025-04-05 doc/manuscript_hkk_drought_sensitivity.Rmd
condensing abstract

@@
 # Abstract (150 words) {-}
-The frequency and intensity of droughts are increasing across forests worldwide, altering tree demography and its contribution to the land carbon sink. 
-In seasonally dry tropical forests, species adaptations to strong dry seasons might confer drought resistance, but this resistance may be variable with individual microenvironments and across drought events.
-We analysed a unique data record - manual dendrometer band measurements over 15 years on 1820 trees across 30 species in a seasonally dry tropical forest in <!---Huai Kha Khaeng, ---->Thailand, <!--using Bayesian regressions in a causal modelling framework--->to assess the drought sensitivity of tree growth in two climatically distinct droughts.
+Droughts are increasingly impacting forests worldwide, altering tree demography and its contribution to the land carbon sink. 
+In tropical dry forests, species adaptations to strong dry seasons (e.g., deciduousness) might confer drought resistance, but this resistance may be variable across microenvironments and drought events.
+We analysed 15 years of dendrometer band measurements on 1820 trees across 30 species in a seasonally dry tropical forest in <!---Huai Kha Khaeng, ---->Thailand <!--using Bayesian regressions in a causal modelling framework--->to assess the drought sensitivity of tree growth in two climatically distinct droughts.
 Species and individuals exhibited a wide range of drought sensitivities within a drought event, and inconsistent responses across these two droughts. 

### 2025-04-05 doc/manuscript_hkk_drought_sensitivity.Rmd
edits to 1st paragraph
@krishnaanujan , I deleted some text with an estimate of C uptake by tropical tree woody growth. This was a very rough, back-of-envelope estimate that I pieced together for the internal grant proposal, but re-reading in this context I realized I'm not comfortable giving that estimate here. (PLus some minor edits.)

@@
 In forests worldwide, intense drought is becoming more frequent, affecting forest demography, diversity, and  C cycling [@intergovernmentalpanelonclimatechangeipcc_climate_2023]. 
-Understanding these impacts is an important research priority, particularly for tropical forests, which <!--harbor high species diversity and --> are critical C sinks; tropical tree woody growth removes roughly `r round (150*1/6,0)` Gt CO~2~ annually offsetting ~`r round(7/36.3*100,-1)`% of anthropogenic fossil fuel emissions [@beer_terrestrial_2010;@anderson-teixeira_carbon_2021;@anderson-teixeira_carbon_2016;@brando_droughts_2019].
-Drought events can threaten this sink; the 2015-2016 ENSO drought in the Amazon temporarily shifted the region from a net sink to a net source of carbon [@brando_droughts_2019].
+Understanding these impacts is an important research priority, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019].
 However, nuanced study of tropical tree woody growth responses to drought is inhibited by the fact that few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022]. 
-Separate studies have shown that drought characteristics species traits, and microenvironment all affect tree growth responses [@bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought.
+Separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [@bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought.
 Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., a species with low sensitivity to past droughts would have low sensitivity in any microhabitat and through any drought.

### 2025-04-05 doc/manuscript_hkk_drought_sensitivity.Rmd
2nd paragraph of intro
@krishnaanujan , I did a bit of rearranging and rewording. I ended up commenting out the last few sentences because I'm not sure what they add - they seem to overlap a lot with the intro to the following paragraph.

@@
 Understanding these impacts is an important research priority, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019].
-However, nuanced study of tropical tree woody growth responses to drought is inhibited by the fact that few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022]. 
+However, study of tropical tree woody growth responses to drought is inhibited by the fact that few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022]. 
 Separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [@bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought.
@@
 Drought characteristics can influence climatic drivers of tree growth and their timing, and affect drought sensitivity of tropical trees. 
-Droughts are notoriously hard to define and encompass a variety of types based on the meteorogical or climatic variables affected [@slette_how_2019]. 
-Here we define drought as hot/dry conditions anomalously departing from their long-term means during any given time of year. 
+Droughts, or hot/dry conditions anomalously departing from their long-term means during any given time of year, are notoriously hard to define and encompass a variety of types based on the meteorogical or climatic variables affected [@slette_how_2019].
+<!--Here we define drought as hot/dry conditions anomalously departing from their long-term means during any given time of year. -->
 <!-- Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.  -->
 <!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
-Tree growth responses to drought timing can be pronounced in seasonally dry tropical forests because of strong seasonality in climate and consequently on photosynthesis, allocation and growth (including dry seasons unfavourable to growth) [@garcía-cervigón_climate_2020].
-(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.)
-Seasonally dry tropical forests are also at thermal and water stress tolerance thresholds for many species, making them more vulnerable to drought [@allen_will_2017], and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019].
+In tropical seasonal forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically in relation to the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020].
+<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
 Analyses to date have found that the woody growth of tropical trees <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly vulnerable to dry season precipitation[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates.
+<!--
+Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019].
+<!--the last phrase of this sentence and how it relates to Bennett et al.
+is unclear to me.
 Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
-These strategies might confer differential advantages under drought conditions.
+These strategies might confer differential advantages under drought conditions. -->
 

### 2025-04-05 doc/manuscript_hkk_drought_sensitivity.Rmd
more edits to paragraphs 1 & 2, starting on 3

@@
 
-<!--KAT suggestion for new intro paragraph (would replace next paragraph)-->
 In forests worldwide, intense drought is becoming more frequent, affecting forest demography, diversity, and  C cycling [@intergovernmentalpanelonclimatechangeipcc_climate_2023]. 
@@
 However, study of tropical tree woody growth responses to drought is inhibited by the fact that few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022]. 
-Separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [@bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought.
+For forests worldwide, separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought.
 Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., a species with low sensitivity to past droughts would have low sensitivity in any microhabitat and through any drought.
@@
 <!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
-Drought characteristics can influence climatic drivers of tree growth and their timing, and affect drought sensitivity of tropical trees. 
-Droughts, or hot/dry conditions anomalously departing from their long-term means during any given time of year, are notoriously hard to define and encompass a variety of types based on the meteorogical or climatic variables affected [@slette_how_2019].
+Drought, or hot/dry conditions anomalously departing from climatic means during any given time of year, is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the severity and duration of the climatic anomalies [@slette_how_2019].
+While forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @refs ], droughts with different characteristics may in fact have very different effects on tree growth [@refs].
 <!--Here we define drought as hot/dry conditions anomalously departing from their long-term means during any given time of year. -->
@@
 Analyses to date have found that the woody growth of tropical trees <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly vulnerable to dry season precipitation[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates.
+It remains unknown how diverse assemblages of tree species in tropical dry forests respond to different types of droughts, including those occurring at different times relative to the regular seasonal cycle of wet and dry seasons.
 <!--
@@
 <!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
-Tree species in seasonally dry forests have diverse evolutionary strategies <!--that span and interact across timescales --->to maximise growth and survival under periodic hot/dry conditions. 
-These strategies can range along a spectrum from hydraulic safety to efficiency leading to tolerance or avoidance of dry conditions[@gonzález-m._diverging_2021]. 
+As in forests globally, tree species in seasonally dry forests have diverse evolutionary strategies <!--that span and interact across timescales --->to cope with drought, often leading to classification of species as "drought tolerant" or "drought sensitive" [@refs].
+In tropical seasonal forests in particular, trees have evolved to maximise growth and survival under periodic hot/dry conditions, with strategies ranging along a spectrum from hydraulic safety to efficiency leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
 Dry season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid water loss and have acquisitive strategies during leaf on periods [@desouza_drought_2020].

### 2025-04-06 doc/manuscript_hkk_drought_sensitivity.Rmd
Update manuscript_hkk_drought_sensitivity.Rmd
@krishnaanujan , one of the changes here was changing author from "Intergovernmental Panel On Climate Change (Ipcc)" to just "IPCC". I did this in Zotero and changed the citation key-- hopefully that comes through for you.

Other changes were refining/ condensing text that I'd already worked on.

@@
 # Abstract (150 words) {-}
-Droughts are increasingly impacting forests worldwide, altering tree demography and its contribution to the land carbon sink. 
-In tropical dry forests, species adaptations to strong dry seasons (e.g., deciduousness) might confer drought resistance, but this resistance may be variable across microenvironments and drought events.
-We analysed 15 years of dendrometer band measurements on 1820 trees across 30 species in a seasonally dry tropical forest in <!---Huai Kha Khaeng, ---->Thailand <!--using Bayesian regressions in a causal modelling framework--->to assess the drought sensitivity of tree growth in two climatically distinct droughts.
-Species and individuals exhibited a wide range of drought sensitivities within a drought event, and inconsistent responses across these two droughts. 
-Deciduous species were more sensitive than evergreen species in the wet-season drought, and this difference was amplified in wetter microclimates. 
-Although exposed trees were more sensitive, access to water through larger size or topographic location buffered growth declines. 
-Heterogeneous responses of species and individuals to droughts indicate potential spatiotemporal insurance effects in diverse forests in the face of increased climate anomalies.
+Droughts are increasingly impacting forests worldwide, threatening their critical carbon sink. 
+Tree species adaptations to hot and dry conditions -- for example, dry season deciduousness in tropical seasonal forests -- might confer drought resistance, but such resistance may be variable across microenvironments and drought events.
+We analysed dendrometer band measurements on 1820 trees across 30 species in a seasonally dry forest in <!---Huai Kha Khaeng, ---->Thailand <!--using Bayesian regressions in a causal modelling framework--->to assess tree growth responses to two climatically distinct droughts.
+Species and individuals exhibited a wide range of drought responses within each drought and inconsistent responses across droughts. 
+Deciduous species were more sensitive to wet-season drought than evergreen species, and this difference was amplified in wetter microclimates. 
+Exposed trees were more sensitive, but greater water access through larger size or topographic location buffered growth declines. 
+Heterogeneous drought responses of species and individuals indicate potential spatiotemporal insurance effects in diverse forests in the face of increased drought.
 
@@
 
-In forests worldwide, intense drought is becoming more frequent, affecting forest demography, diversity, and  C cycling [@intergovernmentalpanelonclimatechangeipcc_climate_2023]. 
-Understanding these impacts is an important research priority, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019].
-However, study of tropical tree woody growth responses to drought is inhibited by the fact that few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022]. 
+In forests worldwide, intense drought is becoming more frequent, affecting forest demography, diversity, and  C cycling [@ipcc_climate_2023]. 
+Understanding these impacts is an important research priority, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022].
 For forests worldwide, separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought.
@@
 <!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
-In tropical seasonal forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically in relation to the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020].
+In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020].
 <!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
-Analyses to date have found that the woody growth of tropical trees <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly vulnerable to dry season precipitation[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates.
+Analyses to date have found that tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates.
 It remains unknown how diverse assemblages of tree species in tropical dry forests respond to different types of droughts, including those occurring at different times relative to the regular seasonal cycle of wet and dry seasons.
@@
 <!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
-As in forests globally, tree species in seasonally dry forests have diverse evolutionary strategies <!--that span and interact across timescales --->to cope with drought, often leading to classification of species as "drought tolerant" or "drought sensitive" [@refs].
+Tree species have diverse evolutionary strategies <!--that span and interact across timescales --->to cope with drought, often leading to their classification as "drought tolerant" or "drought sensitive" [@refs].
 In tropical seasonal forests in particular, trees have evolved to maximise growth and survival under periodic hot/dry conditions, with strategies ranging along a spectrum from hydraulic safety to efficiency leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].

### 2025-04-07 doc/manuscript_hkk_drought_sensitivity.Rmd
Update manuscript_hkk_drought_sensitivity.Rmd
refining intro & a bit in discussion. Just pushing to keep pushes small/easier to review

@@
 <!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
-Tree species have diverse evolutionary strategies <!--that span and interact across timescales --->to cope with drought, often leading to their classification as "drought tolerant" or "drought sensitive" [@refs].
+Tree species have diverse evolutionary strategies <!--that span and interact across timescales --->to cope with dry conditions, often leading to their classification as "drought tolerant" or "drought sensitive" [@refs].
 In tropical seasonal forests in particular, trees have evolved to maximise growth and survival under periodic hot/dry conditions, with strategies ranging along a spectrum from hydraulic safety to efficiency leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
-Dry season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid water loss and have acquisitive strategies during leaf on periods [@desouza_drought_2020].
-Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf turgor loss point (hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), large vessels (hydraulic efficiency) [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021].
-Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021].
+Dry season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss and have acquisitive strategies during leaf-on periods [@desouza_drought_2020].
+Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and *large vessels (hydraulic efficiency)*<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021].
+Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021].
 <!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.
@@
 <!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
-Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls; e.g.
-some species regulate negative TLPs, leaf deciduousness and root depth, based on environmental cues leading to intraspecific or interannual variation potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; REFS Lasky?
+Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls.
+For example, some species regulate negative $\pi_{tlp}$s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation potentially and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; REFS Lasky?
 Hulshof?].
@@
 <!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
-Individual growth responses to drought can depend on the specific microenvironments that trees experience leading to varied atmospheric and hydrological stresses with distinct ecological consequences.
+Individual growth responses to drought can depend on the specific microenvironments that trees experience, leading to varied atmospheric and hydrological stresses with distinct ecological consequences.
 Physiologically, trees experience drought stress as a combination of higher evaporative demand and higher thermal stress on leaves, and greater stress on roots because of low soil moisture. 
-This results in greater stress on trees that have exposed crowns [@vinod_thermal_2023], shallower roots or growing in drier microenvironments []. 
-Consequently, larger trees, with more exposed crowns, suffer more growth declines during drought [@bennett_larger_2015]<!--; larger trees have more exposed crowns although crown exposure can allow access to light and increase growth in normal years-->.
+This results in greater biophysical stress on trees that have exposed crowns [@vinod_thermal_2023] or rely on more ephemeral (typically shallower) water sources that are severely diminished under drought [@refs].<!--I (KAT) made a substantive change here.
+It's not necessarily the case that shallow-rooted trees or trees in dry microenvironments suffer more, as they'd also tend to have adaptations to handle those conditions.-->
+Consequently, larger trees, with more exposed crowns, suffer more growth declines during drought [@bennett_larger_2015] (**see issue [#24](https://github.com/forestgeo/growth-precip-thailand/issues/24)**)<!--; larger trees have more exposed crowns although crown exposure can allow access to light and increase growth in normal years-->.
 <!-- To some extent trees with belowground investment in larger and deeper root systems, or are in wetter locations have lower growth declines during drought because of continued access to water [].
@@
 Analysing two droughts in a 14-year dendrometer band timeseries, we found that tropical tree species and individuals grew less than average in drought years, reflecting in lower plot-level growth (Fig 2).
-Comparing across droughts, we found that despite the shorter duration of drought conditions, the wet-season drought elicited stronger responses across trees, but especially so for deciduous species that are expected to have stronger seasonality in growth phenology (Fig 3).
+Comparing across droughts, we found that despite the *shorter duration of drought conditions*<!--careful here - didn't this continue into 2016?--> and greater absolute extremes of temperature, vapor pressure deficit, and lack of precipitation (Fig.
+2a), the wet-season drought elicited stronger responses across trees, especially deciduous species that are expected to have stronger seasonality in growth phenology (Fig 3).
 Interestingly, in the wet-season drought, we also show that evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites (Fig 3b, Fig 4b).

### 2025-04-08 doc/manuscript_hkk_drought_sensitivity.Rmd
refining intro (and abstract)
@krishnaanujan , biggest change here is near complete re-write of paragraph on microenvironments. Please see what you think.

@@
 Species and individuals exhibited a wide range of drought responses within each drought and inconsistent responses across droughts. 
-Deciduous species were more sensitive to wet-season drought than evergreen species, and this difference was amplified in wetter microclimates. 
-Exposed trees were more sensitive, but greater water access through larger size or topographic location buffered growth declines. 
+Deciduous species were more sensitive to early wet-season drought than evergreen species, and this difference was amplified in wetter microclimates. 
+Exposed trees were more sensitive, but presumed greater water access through larger size or wetter topographic positions buffered growth declines. 
 Heterogeneous drought responses of species and individuals indicate potential spatiotemporal insurance effects in diverse forests in the face of increased drought.
@@
 
-In forests worldwide, intense drought is becoming more frequent, affecting forest demography, diversity, and  C cycling [@ipcc_climate_2023]. 
-Understanding these impacts is an important research priority, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022].
+Globally, intense drought is becoming more frequent, affecting forest demography, diversity, and  carbon (C) cycling [@ipcc_climate_2023]. 
+However, our ability to predict drought impacts on tree woody growth and resultant ecosystem C sequestration remains limited, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022].
 For forests worldwide, separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought.
@@
 <!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
-Drought, or hot/dry conditions anomalously departing from climatic means during any given time of year, is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the severity and duration of the climatic anomalies [@slette_how_2019].
+Drought, or hot/dry conditions anomalously departing from climatic means during any time of year, is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the severity and duration of the climatic anomalies [@slette_how_2019].
 While forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @refs ], droughts with different characteristics may in fact have very different effects on tree growth [@refs].
-<!--Here we define drought as hot/dry conditions anomalously departing from their long-term means during any given time of year. -->
-<!-- Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.  -->
 <!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
@@
 <!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
-Analyses to date have found that tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates.
-It remains unknown how diverse assemblages of tree species in tropical dry forests respond to different types of droughts, including those occurring at different times relative to the regular seasonal cycle of wet and dry seasons.
+Analyses to date have found that tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates.
+It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those occurring at different times relative to regular seasonal cycles<!-- of wet and dry seasons-->.
 <!--
@@
 <!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
-Tree species have diverse evolutionary strategies <!--that span and interact across timescales --->to cope with dry conditions, often leading to their classification as "drought tolerant" or "drought sensitive" [@refs].
-In tropical seasonal forests in particular, trees have evolved to maximise growth and survival under periodic hot/dry conditions, with strategies ranging along a spectrum from hydraulic safety to efficiency leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
+Tree species adaptations to cope with drought vary widely in type and strength, often leading to classification of species as "drought tolerant" or "drought sensitive" [@refs].
+In tropical seasonal forests in particular, trees have evolved to maximise growth, reproduction, and survival under periodic hot/dry conditions, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
 Dry season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss and have acquisitive strategies during leaf-on periods [@desouza_drought_2020].
@@
 Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls.
-For example, some species regulate negative $\pi_{tlp}$s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation potentially and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; REFS Lasky?
+For example, some species regulate negative $\pi_{tlp}$'s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; REFS Lasky?
 Hulshof?].
@@
 <!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
-Individual growth responses to drought can depend on the specific microenvironments that trees experience, leading to varied atmospheric and hydrological stresses with distinct ecological consequences.
-Physiologically, trees experience drought stress as a combination of higher evaporative demand and higher thermal stress on leaves, and greater stress on roots because of low soil moisture. 
-This results in greater biophysical stress on trees that have exposed crowns [@vinod_thermal_2023] or rely on more ephemeral (typically shallower) water sources that are severely diminished under drought [@refs].<!--I (KAT) made a substantive change here.
-It's not necessarily the case that shallow-rooted trees or trees in dry microenvironments suffer more, as they'd also tend to have adaptations to handle those conditions.-->
-Consequently, larger trees, with more exposed crowns, suffer more growth declines during drought [@bennett_larger_2015] (**see issue [#24](https://github.com/forestgeo/growth-precip-thailand/issues/24)**)<!--; larger trees have more exposed crowns although crown exposure can allow access to light and increase growth in normal years-->.
-<!-- To some extent trees with belowground investment in larger and deeper root systems, or are in wetter locations have lower growth declines during drought because of continued access to water [].
+Finally, the microenvironments that individual trees experience influence growth responses to drought.
+<!--leading to varied atmospheric and hydrological stresses with distinct ecological consequences.-->
+Across horizontal axes, microenvironments defined by topographic and edaphic variation shape tree drought responses. 
+While habitats with relatively moist soils (e.g., low-lying topographic positions) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted [e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021] or acclimatized [@refs].
+Indeed, there are observations of trees near streams undergoing greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta_droughtinduced_2017] during drought. 
+Across vertical axes, trees of different sizes have their crowns and root systems in sometimes dramatically divergent microenvironments.
+Aboveground, tree crowns in canopy or emergent positions are exposed to higher evaporative demand and thermal stress driven by higher solar radiation, wind speed, and vapor pressure deficit [@vinod_thermal_2023], and such exposure makes trees more vulnerable to drought [@scharnweber_confessions_2019; @ma_tree_2023].
+<!--In contrast, understory trees may be rarely exposed to direct solar radiation and extreme evaporative demand.-->
+These aboveground gradients in crown exposure are at least partially offset by the tendency for large trees to have larger and often deeper root systems [@refs] and thereby potential to access more reliable water sources [@refs].
+However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
+While the overall tendency is that larger trees suffer greater growth declines during drought [@bennett_larger_2015], it remains unclear whether this is driven primarily by crown exposure, by greater height itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium<!--, as described by Darcy's law -->[@mcdowell_darcy_2015; @fernández-de-uña_role_inpress], or by trait differences across vertical axis [@vinod_thermal_2023].
+Thus, horizontal and vertical microenvironmental gradients modify tree drought responses, likely in interaction with drought characteristics and species traits.
+
+<!--Whether drought characteristics interfere positively or negatively with these evolutionary and physiological strategies and whether it leads to consistent or heterogenous sensitivities across drought events is as yet unknown.
 -->
-However, these responses can potentially be different across species or with drought severity.
-<!--Larger trees have larger root systems, but do not necessarily access deeper water [@ref_from_Panama] -->
-Trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
-Indeed, there is evidence that trees near streams undergo greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta_droughtinduced_2017] during drought. 
-Finally, species strategies may constrain tree light and water microenvironments; e.g.
-evergreen species can covary with soil moisture [@kunert_leaf_2021], shade tolerant and understory species may be rarely exposed to hot/dry air.
-<!-- Some of these adaptations may confer drought sensitivity and buffer stem growth declines during drought years, or conversely, may tradeoff with advantages in normal years, leading to a diversity of responses within a drought event.
--->
-Whether drought characteristics interfere positively or negatively with these evolutionary and physiological strategies and whether it leads to consistent or heterogenous sensitivities across drought events is as yet unknown.
-<!--[from KAT: While all good content, I find this paragraph unnecessary/ distracting. If you'd really like to include content here, I'd work it into the methods.]
-- In sum, these methods may only be capturing the most dramatic effects - the most severe droughts that affect ring-forming species or produce large-enough declines in fast-growing individuals.  
-- Methods need to capture species strategies, distributions across microenvironments, and be precise enough.  
-- Manual dendrometer band timeseries can represent a wide range of species strategies with equipment precision necessary to detect signals. --> 
- <!-- Within species, size, exposure and location can affect sensitivity to drought.  
-- There is theory and evidence that tree size itself matters.  
-- Theoretically, we expect that greater height makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium, as described by Darcy's law [@mcdowell_darcy_2015; @fernández-de-uña_role_inpress].
-- Height... [see refs in @vinod_thermal_2023; @olson_plant_2018; @couvreurWaterTransportTall2018].
-- As tree size increases, leaves exert lower control over hydraulic resistance [@wolfe_leaves_2023].
-- @chen_hydraulic_2022  
-- There is theory and evidence that crown exposure matters [ @vinod_thermal_2023;  @refs_in_ vinod_thermal_2023; @scharnweber_confessions_2019].
-- Microclimate buffering leads to cooler, moister understory air [@vinod_thermal_2023].
-- Soils under closed canopies would also be cooler during hot times of the year [@lembrechts_global_2022].
-- Reduced evaporative demand would also make them moister, and this might be added to by hydraulic redistribution.
-- Trees with exposed crowns suffered significant crown dieback at greater rates in the 2012-16 CA drought [@ma_tree_2023].  
-- rather than necessarily helping during drought, water access will shape the size and traits of species living in habitat, with stream habitats tending to have larger trees [@ref] and more evergreen trees, and also their average growth rate--->
 
 <!--HYPOTHESIS PARAGRAPH-->
-We hypothesize that tropical tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, and individual microenvironments. 
+We hypothesize that tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, and microenvironments, resulting in variable drought responses of species and individuals across droughts, and of species across microhabitats.
 We test this by examining individual tree growth during two ENSO drought years within a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand. 

### 2025-04-08 doc/manuscript_hkk_drought_sensitivity.Rmd
Update manuscript_hkk_drought_sensitivity.Rmd

@@
 
-We analysed data from manual dendrometer band censuses from 2008-2023 in the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
+Our study site was the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
 This is a 50-hectare plot in mixed deciduous tropical forest, with a mean annual temperature of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].

### 2025-04-08 doc/manuscript_hkk_drought_sensitivity.Rmd
updates with @teixeirak edits and #21 edits

@@
 
-**Number of Words in Abstract:** 173  
+**Number of Words in Abstract:** 150  
 **Number of Words in Main Text:** 4989  
@@
 )
-ft <- hline(ft, i = c(1, 4, 7), part = "body")
+ft <- hline(ft, i = c(1, 7, 10), part = "body")
 ft <- line_spacing(ft, space = 1, part = "all")
@@
 # ft <- height(ft, height = 0.3)
-ft <- bold(ft, i = c(1, 4, 7), bold = TRUE, part = "body")
+ft <- bold(ft, i = c(1, 7, 10), bold = TRUE, part = "body")
 ft <- width(ft, 1, width = 3.5)
@@
 # ft <- width(ft, 3, width = 3)
-ft <- set_caption(ft, caption = "Summary table of hypotheses tested, effects of variables on drought sensitivity and corresponding evidence")
+ft <- set_caption(ft, caption = "Summary table of hypotheses tested, effects of variables on drought sensitivity and corresponding evidence.
+For each variable x year combination, \"+\" indicates a positive effect, \"-\" indicates a negative effect and \"n.s.\" indicates that the effect was not statistically significant.")
 ft

### 2025-04-09 doc/manuscript_hkk_drought_sensitivity.Rmd
methods, fig numbering
@krishnaanujan , I put in an example of how I like to number figures. Not necessary to change to this, but makes it a lot easier if you change your figure set/order.

@@
 ## Figures
-fig_DAG <- 1
+droughts = 1
+drought_responses = 2
 
@@
 This is a 50-hectare plot in mixed deciduous tropical forest, with a mean annual temperature of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
+The region is periodically affected by droughts associated with the *El Niño* Southern Oscillation [ENSO, @refs_ENSO].
 The dominant canopy species are _Hopea odorata_, _Vatica harmandiana_, _Dipterocarpus alatus_ and _Tetrameles nudiflora_ and lower layers of the canopy are dominated by sub-canopy evergreen and deciduous species like _Miliusa horsfieldii_ and _Polyalthia viridis_.
@@
 
-![***Figure 1.
+![***Figure `r droughts`.
 Climatic characteristics of drought years and average years*** in the Huai Kha Khaeng ForestGEO plot.
@@
 
-![***Figure 2: Growth increment timeseries and drought sensitivity*** across trees and species in the Huai Kha Khaeng ForestGEO plot.
+![***Figure `r drought_responses`: Growth increment timeseries and drought sensitivity*** across trees and species in the Huai Kha Khaeng ForestGEO plot.
 a) Median of annualised growth increments across species (colours) and the whole plot (black) from 2009 - 2022 from late wet/early dry season dendroband measurements.
@@
 
-We identified drought years of interest in the dendroband timeseries - 2010 and 2015 - by examining two drought indices and corroborating with plot-level ecological responses and expert information. 
-Within the dendroband timeseries, two ENSO events had occurred; 2010 was a moderate ENSO event while 2015 was a very strong ENSO event [REF]. 
+We identified drought years of interest in the dendroband timeseries - 2010 and 2015 - based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses.
+Within the dendroband timeseries, two ENSO events had occurred; 2010 was a moderate ENSO event while 2015 was a very strong ENSO event [@REF]. 
 <!-- We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected.
 -->
-We also used the SPEIbase dataset [@vicente-serrano_multiscalar_2010] to calculate the Standardised Precipitation-Evapotranspiration Index, a drought severity index, calculated at the monthly scale.
-Defining drought as periods with two or more consecutive months with SPEI < -1 also resulted in 2010 and 2015 as drought events. 
-These years also corresponded with expert knowledge of major droughts that affected the region as well as with plot-level growth declines in the timeseries. 
+We also calculate Standardised Precipitation-Evapotranspiration Index (SPEI) at the monthly scale using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
+Defining drought as periods with two or more consecutive months with SPEI < -1 also resulted in identification of 2010 and 2015 as drought events. 
+These years also corresponded with expert knowledge of major droughts that affected the region as well as meteorological records (see next paragraph; Fig.
+`r droughts`) and plot-level growth declines in the timeseries (Fig.
+`r drought_responses`).
 <!-- We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.  -->
 <!-- Moreover, broad definitions based on climate variables (often not measured at the location), may not capture ecological realities within the plot.  -->
-We also note that we chose not to include 2020 in our analysis despite low growth, because a) declines in growth can occur due to various reasons which could have distinct ecological responses in species and individuals b) 2020 did not emerge as a drought year either in the examination of drought indicators or from expert knowledge.
+We note that although tree growth was unusally low in 2020, this year did not emerge as a drought year either in the examination of drought indicators or from expert knowledge.
 
-To characterise these two droughts, we used climate data from a weather station near the plot and publicly available climate data for the region. 
+To characterise the two focal droughts, we used data from a weather station near the plot and publicly available climate data for the region. 
 We used daily weather data collected from 2001 to 2019 at the Forest Fire Research Station [CHECK WHERE], located ~20 km from the ForestGEO plot. 
+*(SPECIFY WHICH VARIABLES WERE OBTAINED FROM WHICH SOURCE, AND OTHER SOURCES.)*
 We first calculated daily values of Vapour Pressure Deficit (VPD) using the equation $$
@@
 
-The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig 1).
+The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r droughts`).
 In 2010, drought characteristics of lower precipitation, higher number of dry days, higher maximum temperature, higher VPD than the long-term mean and SPEI values below -1 occurred between February and April, in the dry season.

### 2025-04-09 doc/manuscript_hkk_drought_sensitivity.Rmd
update with figure numbers as vars

@@
 ## Figures
-droughts = 1
-drought_responses = 2
+fig_droughts <- 1
+fig_drought_responses <- 2
+fig_species <- 3
+fig_microenvs <- 4
+fig_dag <- 5
 
@@
 
-![***Figure `r droughts`.
+![***Figure `r fig_droughts`.
 Climatic characteristics of drought years and average years*** in the Huai Kha Khaeng ForestGEO plot.
@@
 
-![***Figure `r drought_responses`: Growth increment timeseries and drought sensitivity*** across trees and species in the Huai Kha Khaeng ForestGEO plot.
+![***Figure `r fig_drought_responses`: Growth increment timeseries and drought sensitivity*** across trees and species in the Huai Kha Khaeng ForestGEO plot.
 a) Median of annualised growth increments across species (colours) and the whole plot (black) from 2009 - 2022 from late wet/early dry season dendroband measurements.
@@
 These years also corresponded with expert knowledge of major droughts that affected the region as well as meteorological records (see next paragraph; Fig.
-`r droughts`) and plot-level growth declines in the timeseries (Fig.
-`r drought_responses`).
+`r fig_droughts`) and plot-level growth declines in the timeseries (Fig.
+`r fig_drought_responses`).
 <!-- We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.  -->
@@
 
-The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r droughts`).
+The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r fig_droughts`).
 In 2010, drought characteristics of lower precipitation, higher number of dry days, higher maximum temperature, higher VPD than the long-term mean and SPEI values below -1 occurred between February and April, in the dry season.
@@
 
-The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig 2a). 
+The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a). 
 Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual, na.rm=T),2)` cm. 
@@
 
-Trees on average had negative drought sensitivities for both droughts with considerable variation around it (Fig 2b). 
+Trees on average had negative drought sensitivities for both droughts with considerable variation around it (Fig `r fig_drought_responses`b). 
 Median drought sensitivity across all trees in 2010 was `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)` while in 2015 it was `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`.
@@
 
-![***Figure 3: Variation in sensitivity among species*** in the Huai Kha Khaeng ForestGEO plot during droughts in 2010 and 2015.
+![***Figure `r fig_species`: Variation in sensitivity among species*** in the Huai Kha Khaeng ForestGEO plot during droughts in 2010 and 2015.
 a) Relationship of the mean species sensitivities from model fits of an intercept-only model, with mean species deciduousness values.
@@
 
-Deciduousness was associated with lower negative sensitivities in 2015, but not in 2010 (Fig 3a). 
+Deciduousness was associated with lower negative sensitivities in 2015, but not in 2010 (Fig `r fig_species`a). 
 Species median sensitivities in 2010 and 2015 were not correlated with each other (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S5) but on average had lower mean and median sensitivities in 2015 than in 2010 (Fig S5), in alignment with median sensitivities across all trees.
@@
 The interaction between TWI and deciduousness was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) but was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`).
-Combined, predicted sensitivities of deciduous trees was more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig 3b).
+Combined, predicted sensitivities of deciduous trees was more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig `r fig_species`b).
 
-The opposite effects of TWI on sensitivity in the wet and dry season drought, and the influence of deciduousness was also confirmed by models that controlled for crown exposure and size as well as species variation in intercept and slopes (Fig 4a).
+The opposite effects of TWI on sensitivity in the wet and dry season drought, and the influence of deciduousness was also confirmed by models that controlled for crown exposure and size as well as species variation in intercept and slopes (Fig `r fig_microenvs`a).
 These models showed small negative sensitivities in wetter sites in the dry season drought (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but strong positive effect of wetness in the wet season drought (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`).
-The positive effect of TWI in 2015 was, however, mainly due to the response of evergreen species; deciduousness had a significant negative correlation with the slopes of the TWI effect in 2015 (Fig 4b).
-Predictions across the range of observed variables also showed that deciduous species generally had larger negative sensitivities in 2015, which did not show increase with wetness (Fig 4c). 
+The positive effect of TWI in 2015 was, however, mainly due to the response of evergreen species; deciduousness had a significant negative correlation with the slopes of the TWI effect in 2015 (Fig `r fig_microenvs`b).
+Predictions across the range of observed variables also showed that deciduous species generally had larger negative sensitivities in 2015, which did not show increase with wetness (Fig `r fig_microenvs`c).
 On the other hand, neither the mean sensitivities across species nor their response to TWI was associated with deciduousness in 2010.
 
-![***Figure 4: Drivers of variation in sensitivity among individuals***.
+![***Figure `r fig_microenvs`: Drivers of variation in sensitivity among individuals***.
 All panels represent model results from models (one for each year) predicting sensitivity with microenvironment conditions with species random effect on all slopes.
@@
 In the same models with species random effects on intercept and slope, the simplex parameter for the ordered factor crown exposure in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 the negative effect was not as strong (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`).
-Predicted effects of crown exposure on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all crown exposure categories had negative sensitivities in 2015 (Fig 5b).
+Predicted effects of crown exposure on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all crown exposure categories had negative sensitivities in 2015 (Fig `r fig_dag`b).
 
@@
 DBH was a significant predictor of exposure category, with larger DBH associated with higher exposure categories in 2010 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`) and 2015 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) (Fig 5a).
-Controlling for this direct effect of exposure, the remaining effect of size *per se* was significantly positive in 2010, while it was weakly negative in 2015 (Fig 4a, Fig 5a).
+Controlling for this direct effect of exposure, the remaining effect of size *per se* was significantly positive in 2010, while it was weakly negative in 2015 (Fig `r fig_microenvs`a, Fig `r fig_dag`a).
 
-![***Figure 5: Effects of size, exposure and water availability*** from combined models.
+![***Figure `r fig_dag`: Effects of size, exposure and water availability*** from combined models.
 Hypothesised relationships between microenvironment variables and drought sensitivity with their modelled effects in the HKK ForestGEO plot in the droughts of a) 2010 and b) 2015.
@@
 We show strong evidence that tropical tree drought sensitivity can be jointly and interactively shaped by drought characteristics, species adaptations and local environmental conditions (Table 1).
-Analysing two droughts in a 14-year dendrometer band timeseries, we found that tropical tree species and individuals grew less than average in drought years, reflecting in lower plot-level growth (Fig 2).
+Analysing two droughts in a 14-year dendrometer band timeseries, we found that tropical tree species and individuals grew less than average in drought years, reflecting in lower plot-level growth (Fig `r fig_drought_responses`).
 Comparing across droughts, we found that despite the *shorter duration of drought conditions*<!--careful here - didn't this continue into 2016?--> and greater absolute extremes of temperature, vapor pressure deficit, and lack of precipitation (Fig.
-2a), the wet-season drought elicited stronger responses across trees, especially deciduous species that are expected to have stronger seasonality in growth phenology (Fig 3).
-Interestingly, in the wet-season drought, we also show that evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites (Fig 3b, Fig 4b).
-Further, besides consistent negative effects of crown exposure, we found that the drivers of variation among species and individuals in drought sensitivity were dissimilar between the drought events (Fig 4a).
-Water availability buffered exposure-associated growth declines through different mechanisms in the two different droughts - larger trees (with likely deeper roots) in the dry-season drought and trees in wetter sites in the wet-season drought suffered smaller declines in growth (Fig 5).
+2a), the wet-season drought elicited stronger responses across trees, especially deciduous species that are expected to have stronger seasonality in growth phenology (Fig `r fig_species`).
+Interestingly, in the wet-season drought, we also show that evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites (Fig `r fig_species`b, Fig `r fig_microenvs`b).
+Further, besides consistent negative effects of crown exposure, we found that the drivers of variation among species and individuals in drought sensitivity were dissimilar between the drought events (Fig `r fig_microenvs`a).
+Water availability buffered exposure-associated growth declines through different mechanisms in the two different droughts - larger trees (with likely deeper roots) in the dry-season drought and trees in wetter sites in the wet-season drought suffered smaller declines in growth (Fig `r fig_dag`).
 Taken together, our analysis of tree growth in two different drought years, and spanning meaningful variation, shows that species and individuals can respond differently to different drought events. 

### 2025-04-10 doc/manuscript_hkk_drought_sensitivity.Rmd
Update manuscript_hkk_drought_sensitivity.Rmd
@krishnaanujan - I edited the latter part of methods, results, and 1st par of discussion.

@@
 However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
-While the overall tendency is that larger trees suffer greater growth declines during drought [@bennett_larger_2015], it remains unclear whether this is driven primarily by crown exposure, by greater height itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium<!--, as described by Darcy's law -->[@mcdowell_darcy_2015; @fernández-de-uña_role_inpress], or by trait differences across vertical axis [@vinod_thermal_2023].
+While the overall tendency is that larger trees suffer greater growth declines during drought [@bennett_larger_2015], it remains unclear whether this is driven primarily by crown exposure, by greater height itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium <!--, as described by Darcy's law -->[@mcdowell_darcy_2015; @fernández-de-uña_role_inpress], or by trait differences across vertical axis [@vinod_thermal_2023].
 Thus, horizontal and vertical microenvironmental gradients modify tree drought responses, likely in interaction with drought characteristics and species traits.
@@
 Values, therefore represents the number of standard deviations from the mean on a given day.
-c) shows Standardised Precipitation Evapotranspiration Index (SPEI) values for each month in 2010 and 2015.
+c) shows 1-month Standardised Precipitation Evapotranspiration Index (SPEI) values for each month in 2010 and 2015.
 Across all panels, light blue background represents the average wet season, from May to October.
@@
 
-![***Figure `r fig_drought_responses`: Growth increment timeseries and drought sensitivity*** across trees and species in the Huai Kha Khaeng ForestGEO plot.
-a) Median of annualised growth increments across species (colours) and the whole plot (black) from 2009 - 2022 from late wet/early dry season dendroband measurements.
-2010 and 2015 were moderate and very strong ENSO events respectively.
+![***Figure `r fig_drought_responses`: Growth increment timeseries and drought responses*** across trees and species in the Huai Kha Khaeng ForestGEO plot.
+a) Median of annualised growth increments across ten dominant species (colours) and all trees (black) from 2009 - 2022.
 b) Distribution of drought sensitivity of tree growth across all individuals in the two drought years.
@@
 -->
-We also calculate Standardised Precipitation-Evapotranspiration Index (SPEI) at the monthly scale using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
-Defining drought as periods with two or more consecutive months with SPEI < -1 also resulted in identification of 2010 and 2015 as drought events. 
+We also calculate Standardised Precipitation-Evapotranspiration Index (SPEI) at the 1-month scale using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
+Defining drought as periods with two or more consecutive months with 1-month SPEI < -1 also resulted in identification of 2010 and 2015 as drought events. 
 These years also corresponded with expert knowledge of major droughts that affected the region as well as meteorological records (see next paragraph; Fig.
@@
 
+*(See [issue #28](https://github.com/forestgeo/growth-precip-thailand/issues/28))*
 To account for potential biases in dendrometer band increments from water-related growth or shrinkage, we also calculated growth occurrence using a zero-growth assumption [@zweifel_are_2016]. 
@@
 We calculated deciduousness as a species level trait using data published in @williams_deciduousness_2008 for the top 30 species. 
-We defined deciduousness as the mean proportion of canopy loss at maximum loss across individuals in the species, with a score ranging from 1 to 4 where 1 is most evergreen and 4 is most deciduous.  
+We defined deciduousness as the mean proportion of crown leaf loss at maximum loss across individuals in the species, with a score ranging from 1 to 4 where 1 is most evergreen and 4 is most deciduous.
 Data was unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1, in the middle of the evergreen range. 
 An alternate measure of deciduousness is the duration spent at maximum crown loss.
-This data was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for a larger list of species.  
+This was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for a larger list of species.  
 <!-- At the species level, we also calculated the maximum size across all individuals in the plot as measured in the 2008 whole-plot census.
@@
 
-The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r fig_droughts`).
-In 2010, drought characteristics of lower precipitation, higher number of dry days, higher maximum temperature, higher VPD than the long-term mean and SPEI values below -1 occurred between February and April, in the dry season.
-On the other hand, in 2015, similar deviations occurred in May and June, in the wet season. 
-In 2015, SPEI in May was -2, an anomalously low value that was not observed in any month in the entire dendroband census time period, suggesting a severe drought.
+The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 an early wet-season drought (Table 1, Fig `r fig_droughts`).
+The 2010 drought peaked between February and April (dry season), with drought characteristics including lower precipitation, higher number of dry days, higher maximum temperature, and higher VPD relative to the long-term mean and 1-month SPEI values below -1.
+In 2015, similar deviations occurred in May and June (early wet season), with 1-month SPEI in May reaching -2, the lowest in the the entire dendroband census time period.
 
@@
 
-## Growth timeseries and Drought sensitivity {-}
+## Growth timeseries and drought responses {-}
 
@@
 Median annual increment differed by species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` cm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` cm) among the `r nrow(tree.time.sp)` species analysed.
+*(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))*
 
 Drought years were associated with lower growth across the plot. 
+In 2010, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010], na.rm=T),2)` cm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010], na.rm=T)-median(tree.time$inc_annual, na.rm=T))/median(tree.time$inc_annual, na.rm=T)),2)`% lower than the median across all years.
 The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015], na.rm=T),2)` cm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015], na.rm=T)-median(tree.time$inc_annual, na.rm=T))/median(tree.time$inc_annual, na.rm=T)),2)`% lower than the median across all years.
-In 2010, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010], na.rm=T),2)` cm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010], na.rm=T)-median(tree.time$inc_annual, na.rm=T))/median(tree.time$inc_annual, na.rm=T)),2)`% lower than the median across all years.
-While there were more species differences in median responses to the 2010 drought, all species had lower increment than a typical year in 2015. 
+While median species responses to the 2010 drought were variable, all species had lower increment than a typical year in 2015. 
 Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S4). 
@@
 The drought year 2015 had the maximum proportion of trees with negative growth in the timeseries, `r round(100*sum(tree.time$inc_annual[tree.time$yr==2015]<0)/nrow(tree.time[tree.time$yr==2015,]),2)`%.
+*(See [issue #28](https://github.com/forestgeo/growth-precip-thailand/issues/28))*
 
-Trees on average had negative drought sensitivities for both droughts with considerable variation around it (Fig `r fig_drought_responses`b). 
+On average, trees had negative drought sensitivities for both droughts, but with considerable variation (Fig `r fig_drought_responses`b). 
 Median drought sensitivity across all trees in 2010 was `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)` while in 2015 it was `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`.
-In 2010, *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* and *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/2]`* had the lowest and highest median sensitivity of `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` and `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/2]` respectively.
-In 2015, *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* and *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/2]`* had the lowest and highest median sensitivity of `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` and `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/2]` respectively.
-
-![***Figure `r fig_species`: Variation in sensitivity among species*** in the Huai Kha Khaeng ForestGEO plot during droughts in 2010 and 2015.
-a) Relationship of the mean species sensitivities from model fits of an intercept-only model, with mean species deciduousness values.
-Line shown for significant correlation.
-b) Predicted sensitivity values across the observed range of deciduousness and topographic wetness index from a model with TWI, deciduosness and their interaction across all trees modelled for each drought year separately.
-For comments: https://github.com/forestgeo/growth-precip-thailand/issues/12](display/Fig3.png)
+In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/2]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/2]`*.
+In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/2]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/2]`*.
 
@@
 
-## Species effects on drought sensitivity {-}
+## Species effects {-}
 
-Deciduousness was associated with lower negative sensitivities in 2015, but not in 2010 (Fig `r fig_species`a). 
 Species median sensitivities in 2010 and 2015 were not correlated with each other (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S5) but on average had lower mean and median sensitivities in 2015 than in 2010 (Fig S5), in alignment with median sensitivities across all trees.
+
+Deciduousness species had lower negative sensitivities in 2015, but not in 2010 (Fig `r fig_species`a). 
 In 2015, predicted sensitivities from intercept-only models of sensitivity with a species random effect on the intercept showed a significant negative association with deciduousness value (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`).
@@
 
-## Microenvironment effects drought sensitivity {-}
+![***Figure `r fig_species`: Variation in sensitivity among species*** in the Huai Kha Khaeng ForestGEO plot during droughts in 2010 and 2015.
+a) Relationship of the mean species sensitivities from model fits of an intercept-only model, with mean species deciduousness values.
+Line shown for significant correlation.
+b) Predicted sensitivity values across the observed range of deciduousness and topographic wetness index from a model with TWI, deciduosness and their interaction across all trees modelled for each drought year separately.
+For comments: https://github.com/forestgeo/growth-precip-thailand/issues/12](display/Fig3.png)
 
-The effect of wetness on sensitivity was different in 2010 and 2015, with 2015 effects interacting with deciduousness. 
+## Microenvironment and individual effects {-}
+
 Individual tree responses to drought were not correlated with each other (Fig S6).
-Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`).
+
+TWI had no significant on sensitivity in 2010 but a positive effect in 2015. 
+Specifically, across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`).
 The interaction between TWI and deciduousness was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) but was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`).
@@
 
-The opposite effects of TWI on sensitivity in the wet and dry season drought, and the influence of deciduousness was also confirmed by models that controlled for crown exposure and size as well as species variation in intercept and slopes (Fig `r fig_microenvs`a).
-These models showed small negative sensitivities in wetter sites in the dry season drought (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but strong positive effect of wetness in the wet season drought (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`).
+The differing effects of TWI on sensitivity in the two droughts, along with the influence of deciduousness, were confirmed by models that controlled for crown exposure and DBH as well as species variation in intercept and slopes (Fig `r fig_microenvs`a).
+These models showed small negative sensitivities in wetter sites in 2010 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but a strong positive effect of TWI in 2015 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`).
 The positive effect of TWI in 2015 was, however, mainly due to the response of evergreen species; deciduousness had a significant negative correlation with the slopes of the TWI effect in 2015 (Fig `r fig_microenvs`b).
-Predictions across the range of observed variables also showed that deciduous species generally had larger negative sensitivities in 2015, which did not show increase with wetness (Fig `r fig_microenvs`c).
-On the other hand, neither the mean sensitivities across species nor their response to TWI was associated with deciduousness in 2010.
+Predictions across the range of observed variables also showed that deciduous species generally had larger negative sensitivities in 2015, which did not  increase with TWI (Fig `r fig_microenvs`c). 
+On the other hand, neither the mean sensitivities across species nor their response to TWI were associated with deciduousness in 2010.
 
@@
 c) Predicted relationship of drought sensitivity with topographic wetness index across all species derived from model predictions.
-Black line represents overall relationship; coloured lines for each species reflects deciduousness values.
+Black line represents overall relationship (panel a); coloured lines for each species reflects deciduousness values.
 For comments - https://github.com/forestgeo/growth-precip-thailand/issues/13](display/Fig4.png)
@@
 
-Accounting for effects of exposure, remaining effects of size on sensitivity differed between the two droughts. 
-DBH was a significant predictor of exposure category, with larger DBH associated with higher exposure categories in 2010 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`) and 2015 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) (Fig 5a).
-Controlling for this direct effect of exposure, the remaining effect of size *per se* was significantly positive in 2010, while it was weakly negative in 2015 (Fig `r fig_microenvs`a, Fig `r fig_dag`a).
+DBH was a significant predictor of crown exposure, with larger DBH associated with higher exposure categories in 2010 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`) and 2015 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) (Fig 5a).
+Controlling for this direct effect of exposure, the remaining effect of DBH *per se* was significantly positive in 2010, while it was weakly negative in 2015 (Fig `r fig_microenvs`a, Fig `r fig_dag`a).
 
@@
 <!---CONCLUSIONS---->
-We show strong evidence that tropical tree drought sensitivity can be jointly and interactively shaped by drought characteristics, species adaptations and local environmental conditions (Table 1).
-Analysing two droughts in a 14-year dendrometer band timeseries, we found that tropical tree species and individuals grew less than average in drought years, reflecting in lower plot-level growth (Fig `r fig_drought_responses`).
-Comparing across droughts, we found that despite the *shorter duration of drought conditions*<!--careful here - didn't this continue into 2016?--> and greater absolute extremes of temperature, vapor pressure deficit, and lack of precipitation (Fig.
-2a), the wet-season drought elicited stronger responses across trees, especially deciduous species that are expected to have stronger seasonality in growth phenology (Fig `r fig_species`).
-Interestingly, in the wet-season drought, we also show that evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites (Fig `r fig_species`b, Fig `r fig_microenvs`b).
-Further, besides consistent negative effects of crown exposure, we found that the drivers of variation among species and individuals in drought sensitivity were dissimilar between the drought events (Fig `r fig_microenvs`a).
-Water availability buffered exposure-associated growth declines through different mechanisms in the two different droughts - larger trees (with likely deeper roots) in the dry-season drought and trees in wetter sites in the wet-season drought suffered smaller declines in growth (Fig `r fig_dag`).
-Taken together, our analysis of tree growth in two different drought years, and spanning meaningful variation, shows that species and individuals can respond differently to different drought events. 
-Despite stand-level growth declines in drought years, we show that these broad patterns belie enormous heterogeneity in individual and species responses to drought in tropical forests with possible consequences for the maintenance of species diveristy and ecosystem stability.
+We show that, at least in one tropical seasonal forest, the drought sensitivity of tree growth is jointly and interactively shaped by drought characteristics, species adaptations, and local environmental conditions (Table 1).
+Analysing two droughts in a 14-year dendrometer band timeseries, we found lower-than-average growth in drought years, reflecting in lower plot-level growth (Fig `r fig_drought_responses`). 
+Comparing across droughts, there was little consistency in species- or individual-level drought responses, reflecting dissimilar effects of a key species trait (deciduousness) and variables linked to water access (TWI, DBH) under differing drought conditions.
+Despite lower absolute extremes of temperature, vapor pressure deficit, and number of dry days (Fig.
+2a), drought in the early wet season (2015) elicited stronger responses across trees, especially deciduous species that are expected to have stronger seasonality in growth phenology (Fig `r fig_species`).
+Interestingly, in 2015 (but not 2010), evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites (Fig `r fig_species`b, Fig `r fig_microenvs`b).
+While trees with more exposed crowns were consistently more sensitive to drought, additional effects of DBH -- presumably including rooting volume and depth -- were dissimilar between the drought events (Fig `r fig_microenvs`a).
+In other words, water availability buffered exposure-associated growth declines through different mechanisms in the two different droughts; larger trees (with likely deeper roots) fared better in the dry-season drought while trees in wetter sites suffered smaller growth declines in the wet-season drought (Fig `r fig_dag`).
+Taken together, these results indicate that joint and interactive effects of drought characteristics, species traits, and individual microenvironments produce enormous heterogeneity in individual and species drought responses of tropical trees.
+This finding that individual droughts have divergent effects on species and individuals may contribute to the maintenance of species diveristy and ecosystem stability.
 

### 2025-04-10 doc/manuscript_hkk_drought_sensitivity.Rmd
Update manuscript_hkk_drought_sensitivity.Rmd
discussion

@@
 ii) Deciduous species, which rely on wet seasons for growth, are particularly sensitive to wet season drought. 
-iii) Exposed trees are more sensitive to drought, but water availability, either through larger size or topographic location, can buffer growth declines.
+iii) Exposed trees are more sensitive to drought, but water availability, either through larger topographic location or tree size (and corresponding rooting volume), can buffer growth declines.
 
@@
 
-The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 an early wet-season drought (Table 1, Fig `r fig_droughts`).
+The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r fig_droughts`).
 The 2010 drought peaked between February and April (dry season), with drought characteristics including lower precipitation, higher number of dry days, higher maximum temperature, and higher VPD relative to the long-term mean and 1-month SPEI values below -1.
@@
 <!----Limitations in filling gap---->
-Our analysis spans representative spatial and ecological variation, but we are limited in quantitatively analysing differences among droughts and variation in drought characteristics, because of low sample size.
-Given the diversity of factors that can influence tree growth and drought sensitivity in tropical tree communities, our analysis of 30 species across size classes and habitats spans this ecological and spatial (environmental) variation realistically and beyond the scope of previous such analyses.
+While our analysis of 30 species across size classes and habitats spans representative spatial and ecological variation realistically and beyond the scope of previous such analyses, we are limited in quantitatively analysing differences associated with drought characteristics because of low sample size.
 <!-- Long-timescale analyses with tree rings are limited by the fact that species with rings are few, represent specific strategy to water use and growth - and often affiliated to drier microenvironments.
@@
 -->
-<!-- Traditional methods may only be capturing the most dramatic effects - the most severe droughts that affect ring-forming species or produce large-enough declines in fast-growing individuals.  -->
+<!-- Traditional methods may only be capturing the most dramatic effects - the most severe droughts that affect ring-forming species or produce large-enough declines in fast-growing individuals.  
 Further, although estimates of woody growth from dendroband timeseries can have measurement biases because of stem water shrinkage and bark peeling [@chitra-tarak_it_2015], our results are robust to simplifying assumptions that account for these biases (Fig S--).
-However, given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts.
-Thus, quantitative analyses parsing out the specific effects of drought characteristics like relative influence of timing and degree of anomaly are currently impossible, limiting interpretations. 
-In temperate forests, long-term records from tree ring analyses could disentangle species and individual sensitivities to differing drought characteristics, as annual rings are formed by a majority of the species.
+However, given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts.-->
+The two droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, and magnitude and duration of meteorological extremes and anomalies (Fig. `r droughts`).  
+Thus, quantitative analyses parsing out the specific effects of drought characteristics (e.g., timing, degree of anomaly) are currently impossible.
+While the meteorological data justify our characterization of these droughts as "dry-season" (2010) and "wet-season" (2015), growth records from numerous additional droughts would be necessary to rigorously generalize differences based on the seasonality of drought effects.
+Given the relative rarity of drought and the lack of multi-decadal records of annual tree growth for diverse tropical forests, there is currently no data set on which such an analysis could be performed.
+In temperate and boreal forests, where the majority of species form annual growth rings, tree-ring analysis can disentangle species and individual sensitivities to larger numbers of droughts (although sample sizes still remain limited).
+*(KAT paused editing here.)*
 Indeed @mcgregor_tree_2021 found that species sensitivities were different across drought events. 

### 2025-04-10 doc/manuscript_hkk_drought_sensitivity.Rmd
@teixeirak - knit version of today's changes
cm into mm in results #27

@@
   group_by(spfull, yr) %>%
-  dplyr::summarise(
-    sens_med = round(median(sens.prop, na.rm = T), 2),
-    sens_mean = round(mean(sens.prop, na.rm = T), 2),
-    sens_sd = round(sd(sens.prop, na.rm = T), 2)
+  dplyr::summarise( # all values in mm
+    sens_med = round(median(sens.prop / 10, na.rm = T), 2),
+    sens_mean = round(mean(sens.prop / 10, na.rm = T), 2),
+    sens_sd = round(sd(sens.prop / 10, na.rm = T), 2)
   ) %>%
@@
 The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a). 
-Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual, na.rm=T),2)` cm. 
-Median annual increment differed by species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` cm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` cm) among the `r nrow(tree.time.sp)` species analysed.
+Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual/10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual/10, na.rm=T),2)` mm. 
+Median annual increment differed by species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm) among the `r nrow(tree.time.sp)` species analysed.
 *(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))*
@@
 Drought years were associated with lower growth across the plot. 
-In 2010, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010], na.rm=T),2)` cm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010], na.rm=T)-median(tree.time$inc_annual, na.rm=T))/median(tree.time$inc_annual, na.rm=T)),2)`% lower than the median across all years.
-The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015], na.rm=T),2)` cm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015], na.rm=T)-median(tree.time$inc_annual, na.rm=T))/median(tree.time$inc_annual, na.rm=T)),2)`% lower than the median across all years.
+In 2010, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010]/10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]/10, na.rm=T)-median(tree.time$inc_annual/10, na.rm=T))/median(tree.time$inc_annual/10, na.rm=T)),2)`% lower than the median across all years.
+The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]/10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]/10, na.rm=T)-median(tree.time$inc_annual/10, na.rm=T))/median(tree.time$inc_annual/10, na.rm=T)),2)`% lower than the median across all years.
 While median species responses to the 2010 drought were variable, all species had lower increment than a typical year in 2015. 
@@
 However, given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts.-->
-The two droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, and magnitude and duration of meteorological extremes and anomalies (Fig. `r droughts`).  
+The two droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, and magnitude and duration of meteorological extremes and anomalies (Fig. `r fig_droughts`).  
 Thus, quantitative analyses parsing out the specific effects of drought characteristics (e.g., timing, degree of anomaly) are currently impossible.

### 2025-04-10 doc/manuscript_hkk_drought_sensitivity.Rmd
fixed issue with mm calculation

@@
   group_by(spfull) %>%
-  dplyr::summarise(
-    inc_annual_med = round(median(inc_annual, na.rm = T), 2),
-    inc_annual_sd = round(sd(inc_annual, na.rm = T), 2)
+  dplyr::summarise( # change to mm
+    inc_annual_med = round(median(inc_annual * 10, na.rm = T), 2),
+    inc_annual_sd = round(sd(inc_annual * 10, na.rm = T), 2)
   ) %>%
@@
   group_by(spfull, yr) %>%
-  dplyr::summarise( # all values in mm
-    sens_med = round(median(sens.prop / 10, na.rm = T), 2),
-    sens_mean = round(mean(sens.prop / 10, na.rm = T), 2),
-    sens_sd = round(sd(sens.prop / 10, na.rm = T), 2)
+  dplyr::summarise(
+    sens_med = round(median(sens.prop, na.rm = T), 2),
+    sens_mean = round(mean(sens.prop, na.rm = T), 2),
+    sens_sd = round(sd(sens.prop, na.rm = T), 2)
   ) %>%
@@
 The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a). 
-Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual/10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual/10, na.rm=T),2)` mm. 
+Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual*10, na.rm=T),2)` mm. 
 Median annual increment differed by species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm) among the `r nrow(tree.time.sp)` species analysed.
-*(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))*
+<!-- *(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))* -->
 
 Drought years were associated with lower growth across the plot. 
-In 2010, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010]/10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]/10, na.rm=T)-median(tree.time$inc_annual/10, na.rm=T))/median(tree.time$inc_annual/10, na.rm=T)),2)`% lower than the median across all years.
-The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]/10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]/10, na.rm=T)-median(tree.time$inc_annual/10, na.rm=T))/median(tree.time$inc_annual/10, na.rm=T)),2)`% lower than the median across all years.
+In 2010, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
+The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
 While median species responses to the 2010 drought were variable, all species had lower increment than a typical year in 2015. 
@@
 The drought year 2015 had the maximum proportion of trees with negative growth in the timeseries, `r round(100*sum(tree.time$inc_annual[tree.time$yr==2015]<0)/nrow(tree.time[tree.time$yr==2015,]),2)`%.
-*(See [issue #28](https://github.com/forestgeo/growth-precip-thailand/issues/28))*
+<!-- *(See [issue #28](https://github.com/forestgeo/growth-precip-thailand/issues/28))* -->
 

### 2025-04-10 doc/manuscript_hkk_drought_sensitivity.Rmd
remove text on neg growth #28

@@
 
-*(See [issue #28](https://github.com/forestgeo/growth-precip-thailand/issues/28))*
-To account for potential biases in dendrometer band increments from water-related growth or shrinkage, we also calculated growth occurrence using a zero-growth assumption [@zweifel_are_2016]. 
+<!-- *(See [issue #28](https://github.com/forestgeo/growth-precip-thailand/issues/28))* -->
+<!-- To account for potential biases in dendrometer band increments from water-related growth or shrinkage, we also calculated growth occurrence using a zero-growth assumption [@zweifel_are_2016]. 
 Timing of measurement and seasonal dynamics in tree hydration and bark dynamics could contribute to negative values in dendrometer band increments, unlike woody growth from tree ring datasets [@chitra-tarak_it_2015].
 Using a zero-growth model, we calculated growth occurrence as 1 under positive diameter change and any change <=0 as 0. 
-To account for biases introduced by timing of measurements and seasonal dynamics, we used this simplifying assumption to corroborate the patterns observed with the sensitivity metric.
+To account for biases introduced by timing of measurements and seasonal dynamics, we used this simplifying assumption to corroborate the patterns observed with the sensitivity metric.   -->
 
@@
 While median species responses to the 2010 drought were variable, all species had lower increment than a typical year in 2015. 
-Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S4). 
-Across the years, on average, `r round(100*mean(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% of the trees had negative growth, with a minimum of `r round(100*min(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% in `r names(sort(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))))[1]`.
-The drought year 2015 had the maximum proportion of trees with negative growth in the timeseries, `r round(100*sum(tree.time$inc_annual[tree.time$yr==2015]<0)/nrow(tree.time[tree.time$yr==2015,]),2)`%.
+<!-- Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S4).  -->
+<!-- Across the years, on average, `r round(100*mean(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% of the trees had negative growth, with a minimum of `r round(100*min(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% in `r names(sort(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))))[1]`.
+-->
+<!-- The drought year 2015 had the maximum proportion of trees with negative growth in the timeseries, `r round(100*sum(tree.time$inc_annual[tree.time$yr==2015]<0)/nrow(tree.time[tree.time$yr==2015,]),2)`%.
+-->
 <!-- *(See [issue #28](https://github.com/forestgeo/growth-precip-thailand/issues/28))* -->

### 2025-04-11 doc/manuscript_hkk_drought_sensitivity.Rmd
Update manuscript_hkk_drought_sensitivity.Rmd
working on discussion

@@
 
-<!----Limitations in filling gap---->
+<!----Limitations in filling gap/ comparing only two droughts---->
 While our analysis of 30 species across size classes and habitats spans representative spatial and ecological variation realistically and beyond the scope of previous such analyses, we are limited in quantitatively analysing differences associated with drought characteristics because of low sample size.
@@
 Thus, quantitative analyses parsing out the specific effects of drought characteristics (e.g., timing, degree of anomaly) are currently impossible.
-While the meteorological data justify our characterization of these droughts as "dry-season" (2010) and "wet-season" (2015), growth records from numerous additional droughts would be necessary to rigorously generalize differences based on the seasonality of drought effects.
+While the meteorological data justify our characterization of these droughts as "dry-season" (2010) and "wet-season" (2015), growth records from numerous additional droughts would be necessary to rigorously generalize differences based on the seasonality of drought.
 Given the relative rarity of drought and the lack of multi-decadal records of annual tree growth for diverse tropical forests, there is currently no data set on which such an analysis could be performed.
 In temperate and boreal forests, where the majority of species form annual growth rings, tree-ring analysis can disentangle species and individual sensitivities to larger numbers of droughts (although sample sizes still remain limited).
-*(KAT paused editing here.)*
-Indeed @mcgregor_tree_2021 found that species sensitivities were different across drought events. 
-However, as this accumulating dendrometer data record continues to grow, so does the potential to answer these questions across a wider range of forested ecosystems.
+Studies in temperate forests have confirmed that drought characteristics including seasonal timing, duration, and severity affect tree growth responses [@gao_dynamic_2018; @dorangeville_drought_2018] and that effects of species, species traits, tree size, and microenvironment can differ across droughts [@mcgregor_tree_2021].
+As tropical dendrometer band records continues to grow, so will the potential to better understand how drought characteristics interact with species traits and microenironment to shape growth respones.
 
@@
 
-<!----Limitations in generalisation---->
+<!----Limitations in generalisation/ species examined here may have greater diversity of drought adaptaitons than elsewhere in the tropics---->
+We have demonstrated that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry tropical forest.
+Tropical dry forests represent periodic hot/dry conditions, creating selective pressures for adaptative variation towards drought resistance, and this might mean that species in tropical dry forests exhibit a greater diversity of drought responess than species in other bioclimatic zones where adaptive pressures and thus ecological variability are different.
+Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies and has resulted in diversification along the hydraulic safety-efficiency axis with diverse strategies to maximise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021].
+The degree of adaptive variation along the water use axis might be comparable to other dry climatic systems (e.g.
+temperate drylands reported in @vasey_intraspecific_2022), but it may not apply to the wet or everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different.
+Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different traits respond to different types of drought, and how these responses may be modified by microenvironment.
+Again, continued expansion of dendrometer band records and ecophysiological study of tropical tree drought responses will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.
+<!--Therefore, we caution that our results suggestive of community-level insurance effects under drought be interpreted carefully across forest types that may be structured primarily by nutrient limitation, fires, herbivory or pathogens.
+However, our inference on the effects of drought timing could be generalisable across wet and dry tropics in the light of emerging evidence that despite low seasonality in climatic conditions, everwet forest species also show seasonal growth patterns [@giraldo_tree_2023].
+-->
+
 <!-- ALT: Differential allocation to growth may not mean negative outcomes, it might mean demographic trade-offs
@@
 - deciduous species have higher survival under multi-year droughts [@aguirre-guiterrez refs] -->
-
-Tropical dry forests represent periodic hot/dry conditions, creating selective pressures for adaptative variation towards drought resistance, but this might not hold in other bioclimatic zones where adaptive pressures and thus ecological variability is different.
-Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies and has resulted in diversification along the hydraulic safety-efficiency axis with diverse strategies to maximise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021].
-The degree of adaptive variation along the water use axis might be comparable to other dry climatic systems (e.g.
-temperate drylands reported in @vasey_intraspecific_2022) but it may not apply to the wet or everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different.
-Therefore, we caution that our results suggestive of community-level insurance effects under drought be interpreted carefully across forest types that may be structured primarily by nutrient limitation, fires, herbivory or pathogens.
-However, our inference on the effects of drought timing could be generalisable across wet and dry tropics in the light of emerging evidence that despite low seasonality in climatic conditions, everwet forest species also show seasonal growth patterns [@giraldo_tree_2023].
 

### 2025-04-11 doc/manuscript_hkk_drought_sensitivity.Rmd
added paragraph on crown exposure
addresses #24

@@
 
+<!-- contributions / disentangling exposure from DBH-->
+One variable examined here -- crown exposure -- had a consistent effect across the two droughts (Table 1, Figs. `r fig_microenvs`, `r fig_dag`). 
+Trees with more exposed crowns exhibited greater growth declines in both droughts, while sub-canopy trees grew more than average during the 2010 drought.
+This finding is consistent with the concept that larger trees tend to suffer more during drought [@bennett_larger_2015], but adds the important insight that it is crown position -- as opposed to size *per se* -- that is primarily responsible for this pattern.
+The strong and consistent negative effect of crown position makes sense in light of pronounced gradients in drought stress across forest vertical profiles [@vinod_thermal_2023].
+We show that residual effects of size, presumably including root water access and biophysical challenges linked to height itself, may act in different directions depending on the drought characteristics (Figs.
+`r fig_microenvs`, `r fig_dag`).
+Together, these findings imply that while canopy and emergent trees face the greatest stress and exhibit the most negative growth responses in almost any drought, a suite of species traits and microenvironmental characteristics modify growth responses in interaction with drought characteristics.
+
 <!----Contributions beyond---->

### 2025-04-11 scripts/06_figures.R
SPEI plots for 1, 3, 6 and 12 months #26
@teixeirak - trying out different SPEI timescales. Moving them all to the same panel was confusing - this might be the cleanest.

@@ -16,7 +16,7 @@ library(patchwork)
 spei <- read.csv("data/climate/SPEI_HKK_from_GEE.csv")
 ffstation <- read.csv("data/climate/WeatherData_ALL_forest_fire_research_station.csv")
 
-enso <- read.table("data/climate/ENSO_index_meiv2.data", header = F, skip = 1)
+# enso <- read.table("data/climate/ENSO_index_meiv2.data", header = F, skip = 1)
 head(ffstation)
 # calculate vpd using ffstation data
 ffstation <- ffstation %>%
@@ -179,26 +179,30 @@ spei_month <- spei %>%
     dplyr::mutate(
         Date = as.Date(system.time_start, format = "%B %d, %Y"),
         year = year(Date),
+        year = as.character(year),
         month = month(Date)
     ) %>%
-    rename(spei_val = SPEI_01_month) %>%
-    select(year, month, spei_val)
+    # rename(spei_val = SPEI_01_month) %>%
+    # rename(spei_val = SPEI_03_month) %>% # results with only 3 month SPEI
+    filter(year %in% c("2010", "2015")) %>%
+    pivot_longer(cols = c(SPEI_01_month, SPEI_03_month, SPEI_06_month, SPEI_12_month), names_to = "spei_var", values_to = "spei_val") %>%
+    select(year, month, spei_var, spei_val)
 
-spei_lt <- spei_month %>%
-    filter(year <= 2021, year >= 2009) %>%
-    ungroup() %>%
-    group_by(month) %>%
-    dplyr::mutate(
-        spei_val = mean(spei_val, na.rm = T),
-        year = "long-term"
-    ) %>%
-    select(year, month, spei_val)
+# spei_lt <- spei_month %>%
+#     filter(year <= 2021, year >= 2009) %>%
+#     ungroup() %>%
+#     group_by(month) %>%
+#     dplyr::mutate(
+#         spei_val = mean(spei_val, na.rm = T),
+#         year = "long-term"
+#     ) %>%
+#     select(year, month, spei_val)
 
-spei_month <- spei_month %>%
-    dplyr::mutate(year = as.character(year)) %>%
-    filter(year %in% c("long-term", "2010", "2015"))
+# spei_month <- spei_month %>%
+#     dplyr::mutate(year = as.character(year)) %>%
+#     filter(year %in% c("long-term", "2010", "2015"))
 
-spei_full <- bind_rows(spei_month, spei_lt)
[…]

### 2025-04-11 scripts/06_figures.R
small code save

@@ -273,7 +273,7 @@ speiplot_line <- ggplot() +
     scale_x_continuous(breaks = 1:12) +
     # add ribbon
     # geom_ribbon(data = bounds, aes(x = month, ymin = ymin, ymax = ymax, fill = fill), fill = "grey20", alpha = 0.4) +
-    geom_area(data = bounds2, aes(x = x, y = y), fill = "grey20", alpha = 0.3, stat = "identity") +
+    # geom_area(data = bounds2, aes(x = x, y = y), fill = "grey20", alpha = 0.3, stat = "identity") +
     guides(linetype = "none") +
     guides(color = "none") +
     theme(

### 2025-04-12 doc/manuscript_hkk_drought_sensitivity.Rmd
updated with species n #25
updates based on additional SPEI metrics
#26

@krishnaanujan , I think the new SPEI calculations add a lot of insight into these droughts. I've made a few changes:
- in methods added detail to the description of each drought (bold text indicates where it would be improved with real values)
- in table 1, removed detail on the droughts, added SPEI classification of strength
- added sentence explaining SPEI time scales to Fig. 1 caption
- changed back to referring to the 2015 drought as just "wet season", as opposed to "early wet season"

Note: the multi-month SPEI metr

@@
 Species and individuals exhibited a wide range of drought responses within each drought and inconsistent responses across droughts. 
-Deciduous species were more sensitive to early wet-season drought than evergreen species, and this difference was amplified in wetter microclimates. 
+Deciduous species were more sensitive to wet-season drought than evergreen species, and this difference was amplified in wetter microclimates. 
 Exposed trees were more sensitive, but presumed greater water access through larger size or wetter topographic positions buffered growth declines. 
@@
 The dominant canopy species are _Hopea odorata_, _Vatica harmandiana_, _Dipterocarpus alatus_ and _Tetrameles nudiflora_ and lower layers of the canopy are dominated by sub-canopy evergreen and deciduous species like _Miliusa horsfieldii_ and _Polyalthia viridis_.
+For analysis, we focussed on 30 of the most abundant species in the plot.   
 Long-term analysis of tree-ring data shows that tree growth at Huai Kha Khaeng has been influenced by temperature and rainfall [@vlam_temperature_2014].
@@
 Values, therefore represents the number of standard deviations from the mean on a given day.
-c) shows 1-month Standardised Precipitation Evapotranspiration Index (SPEI) values for each month in 2010 and 2015.
+c) shows Standardised Precipitation Evapotranspiration Index (SPEI) values at four different time scales for each month in 2010 and 2015.
+A _k_-month SPEI integrates over _k_ months ending in the month current month.
 Across all panels, light blue background represents the average wet season, from May to October.
@@
 -->
-We also calculate Standardised Precipitation-Evapotranspiration Index (SPEI) at the 1-month scale using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
+We also calculate Standardised Precipitation-Evapotranspiration Index (SPEI) at 1-, 3-, 6- and 12- month scales using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
 Defining drought as periods with two or more consecutive months with 1-month SPEI < -1 also resulted in identification of 2010 and 2015 as drought events. 
@@
 The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r fig_droughts`).
-The 2010 drought peaked between February and April (dry season), with drought characteristics including lower precipitation, higher number of dry days, higher maximum temperature, and higher VPD relative to the long-term mean and 1-month SPEI values below -1.
-In 2015, similar deviations occurred in May and June (early wet season), with 1-month SPEI in May reaching -2, the lowest in the the entire dendroband census time period.
+The 2010 drought peaked between February and May (dry season and beginning of wet season), with 1-month SPEI values below -1 and meteorological anomalies on the 15-day scale including lower precipitation (min anomaly **<-1.5** SD), higher number of dry days (max anomaly **>1.5** SD), higher maximum temperature (max anomaly +0.8 SD), and higher VPD (max anomaly **>2** SD).
+Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions.
+1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought.
+
+In 2015, pronounced meteorological anomalies on the 15-day scale occurred during the wet season, including in precipitation (min anomaly **<-1.5** SD), number of dry days (max anomaly **>1.5** SD), maximum temperature (max anomaly +1.5 SD), and VPD (max anomaly +2 SD).
+1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census time period. 
+Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season.
+1- to 6-month SPEI values reached minima <2, indicative of extreme drought, while the 12-month values remained just above -2, indicative of severe drought.
 
@@
 Despite lower absolute extremes of temperature, vapor pressure deficit, and number of dry days (Fig.
-2a), drought in the early wet season (2015) elicited stronger responses across trees, especially deciduous species that are expected to have stronger seasonality in growth phenology (Fig `r fig_species`).
+2a), drought in the wet season (2015) elicited stronger responses across trees, especially deciduous species that are expected to have stronger seasonality in growth phenology (Fig `r fig_species`).
 Interestingly, in 2015 (but not 2010), evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites (Fig `r fig_species`b, Fig `r fig_microenvs`b).
@@
 However, given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts.-->
-The two droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, and magnitude and duration of meteorological extremes and anomalies (Fig. `r fig_droughts`).
+The two droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, magnitude and duration of meteorological extremes and anomalies, and severity as classified the SPEI metric (Fig.
+`r fig_droughts`).
 Thus, quantitative analyses parsing out the specific effects of drought characteristics (e.g., timing, degree of anomaly) are currently impossible.

### 2025-04-12 doc/manuscript_hkk_drought_sensitivity.Rmd
Update manuscript_hkk_drought_sensitivity.Rmd
worked on discussion

@@
 <!----Contributions beyond---->
-Despite consistent observations of growth declines at the plot level across drought events, we show that this scale obscures a variety of ecological responses at the species and tree level that are heterogeneous across drought events.
-Our finding that a wet season drought had bigger impact calls into question the generality of the findings of @zuidema_tropical_2022, which found that in species with tree-ring chronologies, dry season drought was most impactful.
-(Care must be taken in interpretation, however, as the 2015-2016 wet season ENSO drought was stronger (Fig 1)). 
+Despite consistent observations of growth declines at the plot level across drought events, we show that this scale obscures a variety of ecological responses at the species and tree level that are heterogeneous across droughts.
+An extremely important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings [n=4 at HKK; @vlam_temperature_2014] are unlikely to be representative of the drought responses of the entire community -- particularly because ring formation is caused by dry-season dormancy.
+Our finding that a wet-season drought had greater impact than a dry-season drought calls into question the generality of previous findings based on tree-ring chronologies from HKK  [@vlam_temperature_2014;@anderson-teixeira_joint_2022] and across the tropics [@zuidema_tropical_2022] that tree growth is most sensitive to dry-season precipitation.
+Care must be taken in interpretation, however, as the 2015 drought was stronger (Fig 1).
 <!-- Given the limitations of tree ring methods and tape measurements in measuring accurately measuring growth differences across a range of species in the tropics, methods so far could have led to capturing only the most dramatic droughts that affected (dry-associated) tree ring species and fast-growing individuals.
 -->
+*(KAT paused here.)*
 *Afzelia xylocarpa*, one of the few annual-ring-forming species in HKK is noteworthy with the most distinct response across the two events; it had the highest (positive) sensitivity in the dry season drought, but a large negative sensitivity in the wet season drought (Fig S5).

### 2025-04-12 doc/manuscript_hkk_drought_sensitivity.Rmd
edits to discussion
@krishnaanujan , I've now been through the entire manuscript. I may provide some more edits later, but please proceed with passing along to other coauthors

@@
 Thus, quantitative analyses parsing out the specific effects of drought characteristics (e.g., timing, degree of anomaly) are currently impossible.
-While the meteorological data justify our characterization of these droughts as "dry-season" (2010) and "wet-season" (2015), growth records from numerous additional droughts would be necessary to rigorously generalize differences based on the seasonality of drought.
-Given the relative rarity of drought and the lack of multi-decadal records of annual tree growth for diverse tropical forests, there is currently no data set on which such an analysis could be performed.
-In temperate and boreal forests, where the majority of species form annual growth rings, tree-ring analysis can disentangle species and individual sensitivities to larger numbers of droughts (although sample sizes still remain limited).
-Studies in temperate forests have confirmed that drought characteristics including seasonal timing, duration, and severity affect tree growth responses [@gao_dynamic_2018; @dorangeville_drought_2018] and that effects of species, species traits, tree size, and microenvironment can differ across droughts [@mcgregor_tree_2021].
-As tropical dendrometer band records continues to grow, so will the potential to better understand how drought characteristics interact with species traits and microenironment to shape growth respones.
+While our finding that a wet-season drought (2015) had greater impact on woody growth than a dry-season drought (2010) calls into question the generality of previous findings based on tree-ring chronologies from HKK [@vlam_temperature_2014;@anderson-teixeira_joint_2022] and across the tropics [@zuidema_tropical_2022] that tree growth is most sensitive to dry-season precipitation, care must be taken in interpretation given that the 2015 drought was more anomalous on multi-month time scales (Fig.
+`r fig_droughts`).
+Rigorous generalization on tropical tree growth responses to any specific drought characteristic would require annual growth records from numerous species over numerous droughts, but no such data set currently exists.
+Studies in temperate forests, where extensive tree-ring records are available, have confirmed that drought characteristics including seasonal timing, duration, and severity affect tree growth responses [@gao_dynamic_2018; @dorangeville_drought_2018] and that effects of species, species traits, tree size, and microenvironment can differ across droughts [@mcgregor_tree_2021].
+As tropical dendrometer band records continues to grow, so will the potential to better understand how specific drought characteristics interact with species traits and microenironment to shape growth respones.
 
@@
 Despite consistent observations of growth declines at the plot level across drought events, we show that this scale obscures a variety of ecological responses at the species and tree level that are heterogeneous across droughts.
-An extremely important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings [n=4 at HKK; @vlam_temperature_2014] are unlikely to be representative of the drought responses of the entire community -- particularly because ring formation is caused by dry-season dormancy.
-Our finding that a wet-season drought had greater impact than a dry-season drought calls into question the generality of previous findings based on tree-ring chronologies from HKK  [@vlam_temperature_2014;@anderson-teixeira_joint_2022] and across the tropics [@zuidema_tropical_2022] that tree growth is most sensitive to dry-season precipitation.
-Care must be taken in interpretation, however, as the 2015 drought was stronger (Fig 1).
-<!-- Given the limitations of tree ring methods and tape measurements in measuring accurately measuring growth differences across a range of species in the tropics, methods so far could have led to capturing only the most dramatic droughts that affected (dry-associated) tree ring species and fast-growing individuals.
+An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings [n=4 at HKK; @vlam_temperature_2014] are unlikely to be representative of the drought responses of the entire community.
+For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the two events; it's growth nearly doubled in the dry-season drought but was severely reduced in the wet-season drought (Fig.
+S5).
+More generally, tree-ring records are likely to give a biased picture of drought responses, as ring formation is caused by dry-season dormancy (often associated with deciduous leaf habit) and sampling often targets large trees with exposed crowns and dry microenvironments [@speer_fundamentals_2010]-- all factors shown here to directly and interactively influence drought sensitivity.<!-- More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects.
 -->
-*(KAT paused here.)*
-*Afzelia xylocarpa*, one of the few annual-ring-forming species in HKK is noteworthy with the most distinct response across the two events; it had the highest (positive) sensitivity in the dry season drought, but a large negative sensitivity in the wet season drought (Fig S5).
-More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects.
-In contrast, our analyses reveal a range of responses, including the contrasting influence of deciduousness in the two different droughts, showing that ecological and evolutionary strategies may result in heterogeneous consequences.
-Distinct responses of species and individuals to drought characteristics suggests potential spatiotemporal insurance effects among tropical tree response to drought in alignment with recent global analyses (@liu_drought_2022).
+Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in the two different droughts, showing that ecological and evolutionary strategies make particular species and trees resistant to some droughts but vulnerable to others.
+These distinct responses of species and individuals to drought characteristics suggests potential spatiotemporal insurance effects among tropical tree response to drought, in alignment with recent global analyses (@liu_drought_2022).
 
 <!----Science is better now---->
-Drought sensitivity is often defined as a static species trait but our study reveals expanded dimensions of drought responses with possible consequences for management and restoration. 
+Our study reveals that drought sensitivity is not a static trait associated to particular species, traits, or microenvironments, but rather a dynamic response shaped interactively by these factors and drought characteristics.
 Despite net decreased growth at the plot level (in agreement with landscape-scale in @aguirre-gutiérrez_functional_2022 and @bennett_sensitivity_2023), we show evidence that the same strategies are not uniformly successfully across spatiotemporal climatic variation.
@@
 <!--aligns with remote sensing based inferences across global forests @liu_drought_2022-->
-More inclusive frameworks for drought responses like @trugman_why_2021 (on mortality) encompass ecological complexity and covarying traits with environment like, but they assume consistent sensitivity of individual trees across events.
-Our results, however, demonstrate an additional temporal dimension of variation, at least partially attributable to drought characteristics. 
-Our results indicate that drought responses in tropical trees may not be a fixed response across events, but jointly shaped by species adaptations and local environmental conditions. 
-Given these inconsistencies, diverse stands that may represent greater variation in ecological and evolutionary strategies, may have mechanisms for stable functioning under future droughts.
+More inclusive frameworks for drought responses encompass ecological complexity and covarying traits with environment [e.g., @trugman_why_2021] but still assume consistent sensitivity of individual trees across events.
+However, we demonstrate that drought responses of tropical trees are not linearly predictable across drought events, but rather are shaped by the unique characteristics of each drought based on interactions between species adaptations and local environmental conditions.
+Without overriding the net negative effects of drought on tropical tree growth, these complexities reduce the likelihood of any given species or tree suffering severe growth declines under multiple consecutive droughts, thereby promoting diversity and stabilizing functioning of diverse tropical forests under drought.
 

### 2025-04-14 doc/manuscript_hkk_drought_sensitivity.Rmd
small responses to KAT comments

@@
 We first removed measurements with likely data entry errors because of misplaced decimals, potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses).
-We used the raw window size measurements combined with the diameter measurement at installation $(dia_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[condit].
+We used the raw window size measurements combined with the diameter measurement at installation $(dia_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[@condit_dendrometer_].
 We used a direct measurement of $dia_{1}$ at the time of installation if it was available.
@@
 To characterise the two focal droughts, we used data from a weather station near the plot and publicly available climate data for the region. 
-We used daily weather data collected from 2001 to 2019 at the Forest Fire Research Station [CHECK WHERE], located ~20 km from the ForestGEO plot. 
-*(SPECIFY WHICH VARIABLES WERE OBTAINED FROM WHICH SOURCE, AND OTHER SOURCES.)*
+We used daily weather data on maximum temperate, relative humidity and total precipitation collected from 2001 to 2019 at the Forest Fire Research Station <!--[CHECK WHERE]-->, located ~20 km from the ForestGEO plot.
+<!-- *(SPECIFY WHICH VARIABLES WERE OBTAINED FROM WHICH SOURCE, AND OTHER SOURCES.)* -->
 We first calculated daily values of Vapour Pressure Deficit (VPD) using the equation $$
@@
 $$
-We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 15 day periods across the whole year for each year.
+We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 30 day periods across the whole year for each year.
 Finally, we calculated daily rolling anomalies for each variable for 2010 and 2015 from the long-term daily expectation as $$anomaly_{day} = \frac{rolling.mean_{day} - \bar{rolling.mean_{day}}}{sd(rolling.mean_{day})}$$
@@
 The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r fig_droughts`).
-The 2010 drought peaked between February and May (dry season and beginning of wet season), with 1-month SPEI values below -1 and meteorological anomalies on the 15-day scale including lower precipitation (min anomaly **<-1.5** SD), higher number of dry days (max anomaly **>1.5** SD), higher maximum temperature (max anomaly +0.8 SD), and higher VPD (max anomaly **>2** SD).
+The 2010 drought peaked between February and May (dry season and beginning of wet season), with 1-month SPEI values below -1 and meteorological anomalies on the 30-day scale including lower precipitation (min anomaly **<-1.5** SD), higher number of dry days (max anomaly **>1.5** SD), higher maximum temperature (max anomaly +0.8 SD), and higher VPD (max anomaly **>2** SD).
 Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions.
@@
 
-In 2015, pronounced meteorological anomalies on the 15-day scale occurred during the wet season, including in precipitation (min anomaly **<-1.5** SD), number of dry days (max anomaly **>1.5** SD), maximum temperature (max anomaly +1.5 SD), and VPD (max anomaly +2 SD).
+In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including in precipitation (min anomaly **<-1.5** SD), number of dry days (max anomaly **>1.5** SD), maximum temperature (max anomaly +1.5 SD), and VPD (max anomaly +2 SD).
 1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census time period. 

### 2025-04-14 doc/manuscript_hkk_drought_sensitivity.Rmd
updated table

@@
 )
-ft <- hline(ft, i = c(1, 7, 10), part = "body")
+ft <- hline(ft, i = c(1, 4, 7), part = "body")
 ft <- line_spacing(ft, space = 1, part = "all")
@@
 # ft <- height(ft, height = 0.3)
-ft <- bold(ft, i = c(1, 7, 10), bold = TRUE, part = "body")
+ft <- bold(ft, i = c(1, 4, 7), bold = TRUE, part = "body")
 ft <- width(ft, 1, width = 3.5)

### 2025-04-15 doc/manuscript_hkk_drought_sensitivity.Rmd
replace fig 1 #10  to reflect #26
intro refs added

@@
 Drought, or hot/dry conditions anomalously departing from climatic means during any time of year, is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the severity and duration of the climatic anomalies [@slette_how_2019].
-While forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @refs ], droughts with different characteristics may in fact have very different effects on tree growth [@refs].
+While forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different characteristics may in fact have very different effects on tree growth [@gao_dynamic_2018].
 <!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
@@
 <!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
-Tree species adaptations to cope with drought vary widely in type and strength, often leading to classification of species as "drought tolerant" or "drought sensitive" [@refs].
+Tree species adaptations to cope with drought vary widely in type and strength, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
 In tropical seasonal forests in particular, trees have evolved to maximise growth, reproduction, and survival under periodic hot/dry conditions, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
@@
 Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls.
-For example, some species regulate negative $\pi_{tlp}$'s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; REFS Lasky?
-Hulshof?].
+For example, some species regulate negative $\pi_{tlp}$'s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; @hulshof_variation_2010].
 
@@
 Across horizontal axes, microenvironments defined by topographic and edaphic variation shape tree drought responses. 
-While habitats with relatively moist soils (e.g., low-lying topographic positions) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted [e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021] or acclimatized [@refs].
+While habitats with relatively moist soils (e.g., low-lying topographic positions) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted (e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021) or acclimatized [@refs].
 Indeed, there are observations of trees near streams undergoing greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta_droughtinduced_2017] during drought. 
@@
 <!--In contrast, understory trees may be rarely exposed to direct solar radiation and extreme evaporative demand.-->
-These aboveground gradients in crown exposure are at least partially offset by the tendency for large trees to have larger and often deeper root systems [@refs] and thereby potential to access more reliable water sources [@refs].
+These aboveground gradients in crown exposure are at least partially offset by the tendency for large trees to have larger and often deeper root systems and greater stem water storage [@fernández-de-uña_role_2023] and thereby potential to access more reliable water sources [@stahl_depth_2013].
 However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
@@
 Across all panels, light blue background represents the average wet season, from May to October.
-For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1.png)
+For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1_new.png)
 
@@
 We identified drought years of interest in the dendroband timeseries - 2010 and 2015 - based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses.
-Within the dendroband timeseries, two ENSO events had occurred; 2010 was a moderate ENSO event while 2015 was a very strong ENSO event [@REF]. 
+Within the dendroband timeseries, two ENSO events had occurred; 2010 was a moderate ENSO event while 2015 was a very strong ENSO event (Oceanic Niño Index @bamston_documentation_1997). 
 <!-- We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected.
@@
 Despite consistent observations of growth declines at the plot level across drought events, we show that this scale obscures a variety of ecological responses at the species and tree level that are heterogeneous across droughts.
-An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings [n=4 at HKK; @vlam_temperature_2014] are unlikely to be representative of the drought responses of the entire community.
+An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings (n=4 at HKK; @vlam_temperature_2014) are unlikely to be representative of the drought responses of the entire community.
 For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the two events; it's growth nearly doubled in the dry-season drought but was severely reduced in the wet-season drought (Fig.

### 2025-04-15 doc/manuscript_hkk_drought_sensitivity.Rmd
updated doc with some refs fixed + word count

@@
 **Number of Words in Abstract:** 150  
-**Number of Words in Main Text:** 4989  
+**Number of Words in Main Text:** 5200  
 **Number of References:** 46  
@@
 However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
-While the overall tendency is that larger trees suffer greater growth declines during drought [@bennett_larger_2015], it remains unclear whether this is driven primarily by crown exposure, by greater height itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium <!--, as described by Darcy's law -->[@mcdowell_darcy_2015; @fernández-de-uña_role_inpress], or by trait differences across vertical axis [@vinod_thermal_2023].
+While the overall tendency is that larger trees suffer greater growth declines during drought [@bennett_larger_2015], it remains unclear whether this is driven primarily by crown exposure, by greater height itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium <!--, as described by Darcy's law -->[@mcdowell_darcys_2015; @fernández-de-uña_role_2023], or by trait differences across vertical axis [@vinod_thermal_2023].
 Thus, horizontal and vertical microenvironmental gradients modify tree drought responses, likely in interaction with drought characteristics and species traits.
@@
 This is a 50-hectare plot in mixed deciduous tropical forest, with a mean annual temperature of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
-The region is periodically affected by droughts associated with the *El Niño* Southern Oscillation [ENSO, @refs_ENSO].
+The region is periodically affected by droughts associated with the *El Niño* Southern Oscillation [ENSO, @räsänen_spatial_2016].
 The dominant canopy species are _Hopea odorata_, _Vatica harmandiana_, _Dipterocarpus alatus_ and _Tetrameles nudiflora_ and lower layers of the canopy are dominated by sub-canopy evergreen and deciduous species like _Miliusa horsfieldii_ and _Polyalthia viridis_.
 For analysis, we focussed on 30 of the most abundant species in the plot.   
-Long-term analysis of tree-ring data shows that tree growth at Huai Kha Khaeng has been influenced by temperature and rainfall [@vlam_temperature_2014].
+Long-term analysis of tree-ring data shows that tree growth at HKK has been influenced by temperature and rainfall [@vlam_temperature_2014].
 
@@
 
-In the HKK plot, metal dendrometer bands were installed in 2008 on 2353 trees across 152 species following standard protocol [@muller-landau_metal_2008]. 
+In HKK, metal dendrometer bands were installed in 2008 on 2353 trees across 152 species following standard protocol [@muller-landau_metal_2008]. 
 Trees were selected were based on the relative abundance of species in the forest, with more individuals banded from the abundant species (ranging from 230 trees to 1 tree per species), and spanning the size and habitat distribution within the plot.
@@
 We first removed measurements with likely data entry errors because of misplaced decimals, potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses).
-We used the raw window size measurements combined with the diameter measurement at installation $(dia_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[@condit_dendrometer_].
-We used a direct measurement of $dia_{1}$ at the time of installation if it was available.
+We used the raw window size measurements combined with the diameter measurement at installation $(d_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[@condit_dendrometer_].
+We used a direct measurement of $d_{1}$ at the time of installation if it was available.
 If the tape measurement at the first window size measurement was not available for any band, we excluded window size measurements until the first available diameter measurement. 
-We then calculated annualised diameter increments for each individual for each year by finding the difference in diameter from the previous year ($inc = dia_{t} - dia_{t-1}$) and adjusting for the number of days between measurements ($inc_{annual} = inc * \frac{tdif}{365}$).
+We then calculated annualised diameter increments for each individual $i$ for each year $t$ by finding the difference in diameter from the previous year ($inc_{i,t} = d_{i,t} - d_{i,t-1}$) and adjusting for the number of days ($tdif$) between measurements ($ainc_{i, t} = inc_{i,t} * \frac{tdif}{365}$).
 
-To ensure that the data did not smooth over multiple years, we excluded increment measures with missing values for $dia_{t-1}$ and then excluded large measurement outliers (> 3 standard deviations from the mean across all observations).
-Tape measurements themselves are not precise enough for calculating drought sensitivity; error for each tape measure is 1mm, error for growth is at least 2mm. 
+To ensure that the data did not smooth over multiple years, we excluded increment measures with missing values for $d_{t-1}$ and then excluded large measurement outliers (> 3 standard deviations from the mean across all observations).
+Tape measurements themselves are not precise enough for calculating drought sensitivity; error for each tape measure is 1 mm, error for growth is at least 2 mm. 
 In comparison, median annual growth across trees in HKK is ~1.4 mm and calculating deviation from annual growth needs sub-mm precision. 
@@
 We also calculate Standardised Precipitation-Evapotranspiration Index (SPEI) at 1-, 3-, 6- and 12- month scales using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
-Defining drought as periods with two or more consecutive months with 1-month SPEI < -1 also resulted in identification of 2010 and 2015 as drought events. 
+Defining drought as periods with two or more consecutive months with SPEI < -1 also resulted in identification of 2010 and 2015 as drought events. 
 These years also corresponded with expert knowledge of major droughts that affected the region as well as meteorological records (see next paragraph; Fig.
@@
 To characterise the two focal droughts, we used data from a weather station near the plot and publicly available climate data for the region. 
-We used daily weather data on maximum temperate, relative humidity and total precipitation collected from 2001 to 2019 at the Forest Fire Research Station <!--[CHECK WHERE]-->, located ~20 km from the ForestGEO plot.
+We used daily weather data on maximum temperate, relative humidity and total precipitation collected from 2001 to 2019 at the Forest Fire Research Station <!--[CHECK WHERE]-->, located ~20 km from HKK, and calculated an additional variables vapour pressure deficit (VPD, @raesch_pvldcurve_2020).
 <!-- *(SPECIFY WHICH VARIABLES WERE OBTAINED FROM WHICH SOURCE, AND OTHER SOURCES.)* -->
-We first calculated daily values of Vapour Pressure Deficit (VPD) using the equation $$
-VPD = 0.6108 * e^{17.27 * \frac{t_{max}}{t_{max}+237.3}}* (1-\frac{relative.humidity}{100})
-$$
-We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 30 day periods across the whole year for each year.
+<!-- We first calculated daily values of Vapour Pressure Deficit (VPD) using the equation $$ -->
+<!-- VPD = 0.6108 * e^{17.27 * \frac{t_{max}}{t_{max}+237.3}}* (1-\frac{relative.humidity}{100}) -->
+<!-- $$ -->
+We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 30 day periods across the whole timeseries.
 Finally, we calculated daily rolling anomalies for each variable for 2010 and 2015 from the long-term daily expectation as $$anomaly_{day} = \frac{rolling.mean_{day} - \bar{rolling.mean_{day}}}{sd(rolling.mean_{day})}$$
@@
 
-We first calculated the mean annual increment for each tree as the mean of the annualised increments across the whole timeseries.
-We then calculated sensitivity for each tree $i$ for each drought as 
+We first  calculated the mean annual increment $\bar{a.inc_{i}}$ for each tree from annualised increment $ainc_{i,t}$<!-- as $\frac{\[ 
+  \sum_{t}^{}{ainc_{i,t}} 
+  \]}{\sum{t}}$-->.
+We then calculated sensitivity for each tree $i$ for each drought year $y$ as 
 $$
-sensitivity_{i} = \frac{drought.year.increment_{i} - mean.annual.increment_{i}}{mean.annual.increment_{i}}
+sensitivity_{i, t=y} = \frac{ainc_{i, t=y} - \bar{a.inc_{i}}}{\bar{a.inc_{i}}}
 $$

### 2025-04-15 doc/manuscript_hkk_drought_sensitivity.Rmd
updating SI

@@
 
-Across all individuals in the plot, we calculated topographic wetness index for an expectation of the water availability across space based on the concavity/convexity of a location. 
+Across all individuals in the plot, we calculated Topographic Wetness Index (TWI) for an expectation of the water availability across space based on the concavity/convexity of a location. 
 We used a Digital Elevation Model from a 5 m x 5 m elevation layer collected at plot establishment.
@@
 We defined tree size as the diameter at breast height (calculated from dendrometer band window measurement) at the previous census. 
-We estimated crown exposure for each tree at each census using crown illumination index, an ordinal categorical variable from 1 to 5 (least to most exposed) measured at each dendroband census.
+We estimated crown exposure for each tree at each census using crown illumination index (CII), an ordinal categorical variable from 1 to 5 (least to most exposed) measured at each dendroband census.
 

### 2025-04-15 scripts/06_figures.R

@@ -962,6 +962,53 @@ png("doc/display/Fig_trait_intercept_plot.png", width = 6, height = 6, units = "
 dec_intercept_plot
 dev.off()
 
+# SI figure - map of dendroband trees ---------------
+
+dendro_tree_map_sp <- ggplot() +
+    geom_point(data = tree.time %>% filter(yr %in% c(2010, 2015)), aes(x = Y, y = X, size = calcDBH_min1), alpha = 0.3) +
+    scale_color_viridis_d() +
+    theme_bw() +
+    facet_wrap(spfull ~ yr, nrow = 4) +
+    # theme(
+    #     legend.position = "none",
+    #     axis.text.x = element_blank(),
+    #     axis.text.y = element_blank(),
+    #     axis.ticks.x = element_blank(),
+    #     axis.ticks.y = element_blank()
+    # ) +
+    coord_fixed()
+
+png("doc/display/dendro_tree_map_sp.png", width = 24, height = 16, units = "in", res = 300)
+dendro_tree_map_sp
+dev.off()
+
+
+dendro_tree_map <- ggplot() +
+    geom_point(data = tree.time %>% filter(yr %in% c(2010, 2015)), aes(x = Y, y = X, size = calcDBH_min1, col = spfull), alpha = 0.5) +
+    scale_color_viridis_d() +
+    theme_bw() +
+    facet_wrap(~yr) +
+    xlab("metres east") +
+    ylab("metres north") +
+    guides(
+        size = guide_legend(title = "DBH"),
+        color = guide_legend(title = "Species")
+    ) +
+    # theme(
+    #     legend.position = "none",
+    #     axis.text.x = element_blank(),
+    #     axis.text.y = element_blank(),
+    #     axis.ticks.x = element_blank(),
+    #     axis.ticks.y = element_blank()
+    # ) +
+    coord_fixed()
+
+png("doc/display/dendro_tree_map.png", width = 16, height = 14, units = "in", res = 300)
+dendro_tree_map
+dev.off()
+
+
+
 # SI figure - species all species sensitivities---------------
 
 # order of species abundance

### 2025-04-15 scripts/06_figures.R
update SI with better cond dep figs

@@ -1533,3 +1533,85 @@ library(patchwork)
 png("doc/display/Fig_SI_sensitivity_cor_dec.png", width = 16, height = 8, units = "in", res = 300)
 sens.sp_plot + sens_rangeplot
 dev.off()
+
+
+# SI conditional dependencies -----------------------
+
+## test conditional independences for each species-----------------------------
+
+# DBH | TWI
+# CII | TWI
+
+# 2015 data only
+
+cond_dep <- tree.time %>%
+    filter(Cno == 15) %>%
+    group_by(Species) %>%
+    dplyr::summarise(
+        DBH_TWI = cor.test(calcDBH_min1, twi)[4]$estimate,
+        DBH_TWI_p = cor.test(calcDBH_min1, twi)[3]$p.value,
+        CII_TWI = cor.test(cii_min1, twi)[4]$estimate,
+        CII_TWI_p = cor.test(cii_min1, twi)[3]$p.value
+    )
+
+# these variables are conditionally independent for the most part
+
+# plot the conditional independencies
+
+library(ggpubr)
+cond_dep_dbh_twi <- ggscatter(
+    data = tree.time %>% filter(Cno == 15),
+    x = "calcDBH_min1", y = "twi",
+    add = "reg.line", conf.int = TRUE,
+    cor.coef = TRUE, cor.method = "spearman",
+    xlab = "DBH", ylab = "TWI"
+) +
+    facet_wrap(~ factor(Species, levels = names(sort(table(Species), decreasing = T))))
+
+png("doc/display/explore/cond_dep_dbh_twi.png", width = 12, height = 12, units = "in", res = 300)
+cond_dep_dbh_twi
+dev.off()
+
+# CII | TWI
+cond_dep_cii_twi <- ggscatter(
+    data = tree.time %>% filter(Cno == 15),
+    x = "cii_min1", y = "twi",
+    add = "reg.line", conf.int = TRUE,
+    cor.coef = TRUE, cor.method = "pearson",
+    xlab = "CII", ylab = "TWI"
+) +
+    facet_wrap(~ factor(Species, levels = names(sort(table(Species), decreasing = T))))
+
+png("doc/display/explore/cond_dep_cii_twi.png", width = 12, height = 12, units = "in", res = 300)
+cond_dep_cii_twi
+dev.off()
+
+# plot conditional dependencies across all trees
[…]

### 2025-04-16 doc/supplementary_information.Rmd
@teixeirak - knit version of today's changes
remove neg growth from SI #28
updating SI
update SI with better cond dep figs
updated SI figures, captions from 1-6

--- 2025-04-03 doc/supplementary_information.Rmd
@@
 
-![Figure S1: Raw climate data timeseries](display/climvars_fullseries.png)
+![*Figure S1:* Raw climate data timeseries from 2009 to 2020 from the Forest Fire Research Station near HKK.
+Measurements represent daily rolling means over a 30-day time window for precipitation, max temperature and VPD and rolling sum for number of dry days.](display/climvars_fullseries.png)
 
-![Figure S2: Climate anomalies across the whole timeseries](display/climvars_anomalies_fullseries.png)
+![*Figure S2:* Climate anomalies for all variables from 2009 to 2020from the Forest Fire Research Station near HKK.
+Values represent number of standard deviations from the long-term mean for daily rolling mean values.](display/climvars_anomalies_fullseries.png)
+
 \newpage
@@
 
-## Species traits and correlations
+## Growth timeseries and anomalies
 
-![](display/Fig_SI_species_sensitivities.png)
+![*Figure S3:* Growth anomalies across the timeseries.
+Growth anomalies for each year calculated as the number of standard deviations in growth each year from mean growth across all years and summarised for plot and species.](display/growth_anom_fullseries.png)
 
-![](display/Fig_SI_sensitivity_cor_dec.png)
-
-![](display/Fig_trait_intercept_plot.png)
+<!-- ![Fig S4: Proportion of negative growth increments](display/neg_incs.png) -->
 
@@
 
-## Individual growth and correlations
+## Drought year growth of species and individuals
 
-- Negative growth proportions across the years
+![*Figure S4:* Species sensitivities in drought years 2010 and 2015.
+Boxplots represent 25th, 50th (median) and 75th percentile of sensitivity for each species in each year and whiskers represent 95% CIs.
+All individual sensitivities represented by jittered points.](display/Fig_SI_species_sensitivities.png)
 
-![](display/sens_corr.png)
+![*Figure S5:* Correlations of species sensitivities across the two years.
+Panel a shows correlation of species median sensitivities in the two years and Pearson's correlation coefficient.
+Colour gradient represents deciduousness scale and dotted line is the 1:1 line.
+Panel b shows difference in magnitude of species median sensitivites among the two years](display/Fig_SI_sensitivity_cor_dec.png)
+
+<!-- ![](display/Fig_trait_intercept_plot.png) -->
+
+![*Figure S6:* Correlation of individual sensitivities across the two years along with Pearson's correlation coefficient and the 1:1 line.](display/sens_corr.png)
+
+\newpage
 
@@
 
-![](display/explore/twi.png)
+Map of trees
 
-![](display/explore/dbh_cii.png)
+![Fig S7:](display/dendro_tree_map.png)
+
+![Fig S8:](display/explore/twi.png)
+
+![Fig S9: ](display/explore/dbh_cii.png)
+
+\newpage
+
+## Conditional dependencies
+
+![Fig S10:](display/explore/cond_dep_alltrees.png)
+
+![Fig S11: ](display/explore/cond_dep_dbh_twi.png)
+
+![Fig S12:](display/explore/cond_dep_cii_twi.png)
 
@@
 
-## Growth occurrence models
+<!-- ## Growth occurrence models
 
-To add: zero growth models
+To add: zero growth models -->

### 2025-04-16 scripts/06_figures.R

@@ -250,6 +250,11 @@ bounds2 <- spei_month %>%
 
 bounds2
 
+df <- data.frame(
+    x = rep(9.5, 4), y = c(0.75, -0.7, -1.5, -2.5),
+    label = c("wetter", "drier", "drought", "severe drought")
+)
+
 # spei plot with lines
 speiplot_line <- ggplot() +
     # geom_rect(data = data.frame(xmin = c(10.5, 0.5), xmax = c(12.5, 4.5), ymin = -Inf, ymax = Inf), aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), fill = "grey", alpha = 0.3) +
@@ -263,17 +268,31 @@ speiplot_line <- ggplot() +
         #  linetype = spei_var,
         color = year
     ), linewidth = 1.5) +
+    # add text for wet, dry drought regions
+    # annotate("text",
+    #     x = c(12, 12, 12, 12), y = c(0.75, -0.75, -1.5, -2.2),
+    #     label = c("wetter", "drier", "drought", "severe drought"),
+    #     col = "black", hjust = -0.5, vjust = -1, fontface = "italic", size = 4
+    # ) +
+    # geom_text(inherit.aes = F, aes(
+    #     x = c(10, 10, 10, 10), y = c(0.75, -0.75, -1.5, -2.2),
+    #     label = c("wetter", "drier", "drought", "severe drought"), fontface = "italic"
+    # )) +
+    geom_text(
+        data = df, aes(x = x, y = y, label = label),
+        col = "black", hjust = 0, vjust = 0, fontface = "italic", size = 3
+    ) +
     ggtitle("SPEI") +
     scale_color_manual(values = c("long-term" = "grey40", "2010" = "indianred2", "2015" = "indianred4")) +
     # ggh4x::stat_difference(data = spei_month, aes(x = month, ymin = -1, ymax = spei_val), levels = c("above", "below")) +
     # scale_fill_manual(limits = c("above", "below"), values = c("grey40", "grey60")) +
-    facet_wrap(~spei_var, scales = "free_y", strip.position = "left", nrow = 4, labeller = varnames) +
+    facet_wrap(~spei_var,
+        # scales = "free_y",
+        strip.position = "left", nrow = 4, labeller = varnames
+    ) +
     theme_bw() +
     labs(x = "Month", y = "SPEI", fill = "Year") +
[…]

### 2025-04-16 doc/supplementary_information.Rmd
draft captions for all figures

@@
                   \captionsetup[table]{font=footnotesize}
-                  \captionsetup[figure]{font=footnotesize}
                   \captionsetup[figure]{labelformat=empty}
@@
 
-![*Figure S1:* Raw climate data timeseries from 2009 to 2020 from the Forest Fire Research Station near HKK.
+![**Figure S1:** Raw climate data timeseries from 2009 to 2020 from the Forest Fire Research Station near HKK.
 Measurements represent daily rolling means over a 30-day time window for precipitation, max temperature and VPD and rolling sum for number of dry days.](display/climvars_fullseries.png)
 
-![*Figure S2:* Climate anomalies for all variables from 2009 to 2020from the Forest Fire Research Station near HKK.
+![**Figure S2:** Climate anomalies for all variables from 2009 to 2020from the Forest Fire Research Station near HKK.
 Values represent number of standard deviations from the long-term mean for daily rolling mean values.](display/climvars_anomalies_fullseries.png)
@@
 
-![*Figure S3:* Growth anomalies across the timeseries.
+![**Figure S3:** Growth anomalies across the timeseries.
 Growth anomalies for each year calculated as the number of standard deviations in growth each year from mean growth across all years and summarised for plot and species.](display/growth_anom_fullseries.png)
@@
 
-![*Figure S4:* Species sensitivities in drought years 2010 and 2015.
+![**Figure S4:** Species sensitivities in drought years 2010 and 2015.
 Boxplots represent 25th, 50th (median) and 75th percentile of sensitivity for each species in each year and whiskers represent 95% CIs.
@@
 
-![*Figure S5:* Correlations of species sensitivities across the two years.
+![**Figure S5:** Correlations of species sensitivities across the two years.
 Panel a shows correlation of species median sensitivities in the two years and Pearson's correlation coefficient.
@@
 
-![*Figure S6:* Correlation of individual sensitivities across the two years along with Pearson's correlation coefficient and the 1:1 line.](display/sens_corr.png)
+![**Figure S6:** Correlation of individual sensitivities across the two years along with Pearson's correlation coefficient and the 1:1 line.](display/sens_corr.png)
 
@@
 
-![Fig S7:](display/dendro_tree_map.png)
+![**Figure S7:** Map of trees with dendrobands in HKK, included in analyses in the 2010 and 2015 datasets](display/dendro_tree_map.png)
 
-![Fig S8:](display/explore/twi.png)
+![**Figure S8:** Calculated Topographic Wetness Index (TWI) values for the HKK plot](display/explore/twi.png)
 
-![Fig S9: ](display/explore/dbh_cii.png)
+![**Figure S9:** Distribution of DBH with Crown Illumination Index (CII) in 2014.](display/explore/dbh_cii.png)
 
@@
 
-![Fig S10:](display/explore/cond_dep_alltrees.png)
+To analyse the influence of microenvironmental variables on growth sensitivity, we first created a Directed Acyclic Graph describing the relationships.
+To be able to test the DAG with the dataset, we used *dagitty* to derive the conditional dependencies that need to be tested, namely CII $\perp\!\!\!\perp$ TWI and DBH $\perp\!\!\!\perp$ TWI.
 
-![Fig S11: ](display/explore/cond_dep_dbh_twi.png)
+![**Figure S10:** Testing correlations of variables across all individuals](display/explore/cond_dep_alltrees.png)
 
-![Fig S12:](display/explore/cond_dep_cii_twi.png)
+There is low correlation between these two variable pairs across all individuals. We then tested conditional dependencies at the species level.
+
+![**Figure S11:** Testing correlations for DBH $\perp\!\!\!\perp$ TWI by species](display/explore/cond_dep_dbh_twi.png)
+
+![**Figure S12:** Testing correlations for CII $\perp\!\!\!\perp$ TWI by species](display/explore/cond_dep_cii_twi.png)
+
+Most species (barring a few) had low correlation between these variables in our dataset, allowing us to proceed with analysis and interpretation.
 

### 2025-04-18 doc/manuscript_hkk_drought_sensitivity.Rmd
Abstract and Introduction commented and edited by Sean

@@
 # Abstract (150 words) {-}
-Droughts are increasingly impacting forests worldwide, threatening their critical carbon sink. 
-Tree species adaptations to hot and dry conditions -- for example, dry season deciduousness in tropical seasonal forests -- might confer drought resistance, but such resistance may be variable across microenvironments and drought events.
-We analysed dendrometer band measurements on 1820 trees across 30 species in a seasonally dry forest in <!---Huai Kha Khaeng, ---->Thailand <!--using Bayesian regressions in a causal modelling framework--->to assess tree growth responses to two climatically distinct droughts.
-Species and individuals exhibited a wide range of drought responses within each drought and inconsistent responses across droughts. 
-Deciduous species were more sensitive to wet-season drought than evergreen species, and this difference was amplified in wetter microclimates. 
-Exposed trees were more sensitive, but presumed greater water access through larger size or wetter topographic positions buffered growth declines. 
+An observed increase in the frequency and severity of droughts threatens forest health worldwide, with important implications for this critical carbon sink. 
+Tree species adaptations to hot and dry conditions--for example, dry season deciduousness in tropical seasonal forests--might confer drought resistance, but such resistance may be variable across species, microenvironments, and drought events.
+We analysed dendrometer band measurements on 1820 trees across 30 species in a seasonally dry forest in Northern <!---Huai Kha Khaeng, ---->Thailand <!--using Bayesian regressions in a causal modelling framework---> to assess tree growth responses to two climatically distinct droughts.
+Species and individuals exhibited a wide range of growth responses within each drought, but these responses differed in intensity and affect  across droughts. 
+Deciduous species were more sensitive to wet-season drought than evergreen species, and wetter microclimates amplified this difference. 
+Exposed trees showed more sensitivity to drought, but  larger sized trees or trees in wetter topographic positions seemed to buffer growth declines. 
 Heterogeneous drought responses of species and individuals indicate potential spatiotemporal insurance effects in diverse forests in the face of increased drought.
@@
 Globally, intense drought is becoming more frequent, affecting forest demography, diversity, and  carbon (C) cycling [@ipcc_climate_2023]. 
-However, our ability to predict drought impacts on tree woody growth and resultant ecosystem C sequestration remains limited, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022].
+However, our ability to predict drought impacts on tree woody growth and the resultant ecosystem C sequestration remains limited, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022].
 For forests worldwide, separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought.
@@
 
+<!-- From Sean: I like the detail of this proposed paragraph.
+It almost operates as a first paragraph?
+I guess the difference is going Ecology picture of droughts -> drought importance in to the carbon cycle to 'how to detect a drought vs.
+big picture droughts, ways it can affect ecology, to how you would measure ...
+:::: either way, there is some redundancy between these two paragraphs.
+-->
+
 <!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
+<!-- Sean::: This paragraph is a bit redundant and all over the place.
+1) in the tropics, seasonal dry periods (often defined as months with <= 100 mm rainfall) can have complex influences on tree phenology, from leaf turnover to growth allocation.
+Droughts, which can exacerbate or extend a dry season, or reduce cloud cover ina rainy season, would be expected interact with seasonal phenology also in complex ways.
+[Then go into some details with citations on ?] -->
 Drought, or hot/dry conditions anomalously departing from climatic means during any time of year, is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the severity and duration of the climatic anomalies [@slette_how_2019].
-While forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different characteristics may in fact have very different effects on tree growth [@gao_dynamic_2018].
+Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different timing, duration, or intensity may in fact have very different effects on tree growth, especially in seasonal forests that have evolved phenological strategies to handle regular periods of climatic stress [@albert_ecology_letters_cryptic_phenology] [@gao_dynamic_2018].
 <!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
+<!-- Sean::: don't think the above is true universally. Laura Boeschoten and Erick Esteban have separate papers well in prep showing some species demonstarating seasonal growth and others not -->
 In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020].
+<!---- agreed but see above ref I put in to [@albert_ecology_letters_cryptic_phenology].
+It seems like these arguments are a bit redundant.
+Or meandering.
+---->
 <!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
@@
 <!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
+
+<!-- Sean::: mention rooting depth? [@chitra-tarak: root of the drought and chitra-tarak: new_phyt 2022] -->
 Tree species adaptations to cope with drought vary widely in type and strength, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
@@
 <!--In contrast, understory trees may be rarely exposed to direct solar radiation and extreme evaporative demand.-->
-These aboveground gradients in crown exposure are at least partially offset by the tendency for large trees to have larger and often deeper root systems and greater stem water storage [@fernández-de-uña_role_2023] and thereby potential to access more reliable water sources [@stahl_depth_2013].
+These aboveground gradients in crown exposure are at least partially offset by the tendency for large trees to have larger and often deeper root systems and greater stem water storage [@fernández-de-uña_role_2023] and therefore the potential to access more reliable water sources [@stahl_depth_2013].
+<!-- Sean::: I need to look at this source.
+Is this true?
+it doesn't really make sense unless coupled with other traits (e.g., deciduousness shifts accross ontogeny) as 'putting down a tap root' is one of the first things a seedling of a deep rooted species does (i.e., it's not allometry but strategy) -->
+
 However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
@@
 We hypothesize that tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, and microenvironments, resulting in variable drought responses of species and individuals across droughts, and of species across microhabitats.
-We test this by examining individual tree growth during two ENSO drought years within a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand. 
-Specifically, we test predictions that
-i) both species and individuals respond differently to different droughts. 
-ii) Deciduous species, which rely on wet seasons for growth, are particularly sensitive to wet season drought. 
-iii) Exposed trees are more sensitive to drought, but water availability, either through larger topographic location or tree size (and corresponding rooting volume), can buffer growth declines.
+Specifically, we test predictions that:
+i) both species and individuals respond differently to different droughts;
+ii) deciduous species, which rely on wet seasons for growth, are particularly sensitive to wet season drought; and,
+iii) exposed trees are more sensitive to drought, but water availability, either through larger topographic location or tree size (and corresponding rooting volume), can buffer growth declines.
+We tested these hypothses using  a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand.
+With data on 1820 trees across 30 species we built Bayesian regression models in a causal modelling framework to account for potential confounding relationships among the variables.
+This analysis adds important insight into the complexity of seasonal dry forest response to drought in  species-rich tropical forests.
 

### 2025-04-24 doc/cover_letter.Rmd
@teixeirak cover letter for review
https://onlinelibrary.wiley.com/page/journal/14610248/homepage/forauthors.html#tips8

not sure if novelty statement needs to be under a separate section
(can remove these files from the repo later)

---
title: ""
output: pdf_document
---

April 22, 2025
\newline
\newline

Dear Editor,

We are pleased to submit our manuscript "Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments" for consideration for publication in *Ecology Letters*.

Across global forests, droughts are increasing in frequency and intensity, threatening forest demography and its contribution to the global carbon cycle. 
While plot-level observations show that severe droughts lead to lower biomass gains, species and tree level growth declines could be influenced by multiple drivers that could either amplify or buffer growth declines during drought years. 
Some tree species adaptations to predictable hot, dry conditions such as dry season deciduousness in seasonal tropical forests could confer drought resistance but this could differ with droughts, species and individual tree microenvironments. 
Current knowledge on tropical tree responses to drought however is limited to understanding from tree ring records, but very few species in the tropics show annual rings.

Here, we use dendrometer band measurements from 1820 trees across 30 representative species in a tropical seasonal forest in northern Thailand to show that tree growth responses to drought can be influenced by drought characteristics, microenvironments and species traits. 
In two ENSO drought years within a 14-year timeseries, species and trees showed heterogeneous drought sensitivity across drought events analysed, as opposed to consistent or correlated responses scaled with drought intensity.
Exposed trees consistently suffered greater growth declines, but the effects of individual and species drivers related to tree-water relations - topographic wetness, tree size and leaf deciduousness - were all heterogenous between a dry- and wet-season drought. 
Our findings imply that although drought resistance is often regarded as a static trait, the species and tree level responses show a rich and dynamic diversity of responses across drought events. 
In sum, our findings of heterogenous responses indicate potential spatiotemporal insurance effects in diverse forests that may ameliorate effects of intensifying drought. 
This is a significant finding that will be of interest to the broad readership of *Ecology Letters*.

We believe our manuscript to be an appropriate length for *Ecology Letters*, with 4989 words and 6 display items in the main article. The article is accompanied by a Supplementary Information file (12 figures).

Thank you for considering our submission for *Ecology Letters*.

Sincerely,
\newline
\newline

Krishna Anujan, on behalf of all co-authors.

### 2025-05-02 doc/manuscript_hkk_drought_sensitivity.Rmd
results and part of discussion from sean

@@
 
-In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including in precipitation (min anomaly **<-1.5** SD), number of dry days (max anomaly **>1.5** SD), maximum temperature (max anomaly +1.5 SD), and VPD (max anomaly +2 SD).
+In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including precipitation (min anomaly **<-1.5** SD), number of dry days (max anomaly **>1.5** SD), maximum monthly mean temperatures (max anomaly +1.5 SD), and monthly maximum VPD (max anomaly +2 SD).
+<!-- make sure my adjustments above in details are correct ... I think we should present a precise description  -->
 1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census time period. 
@@
 
-Individual tree responses to drought were not correlated with each other (Fig S6).
+The test of whether individual trees responded similarly to the two drought events showed no correlation (Fig S6).
 
-TWI had no significant on sensitivity in 2010 but a positive effect in 2015. 
-Specifically, across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`).
+TWI showed no significant effect on sensitivity in 2010, but did show a positive effect in 2015. 
+Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`).
 The interaction between TWI and deciduousness was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) but was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`).
@@
 <!---CONCLUSIONS---->
-We show that, at least in one tropical seasonal forest, the drought sensitivity of tree growth is jointly and interactively shaped by drought characteristics, species adaptations, and local environmental conditions (Table 1).
-Analysing two droughts in a 14-year dendrometer band timeseries, we found lower-than-average growth in drought years, reflecting in lower plot-level growth (Fig `r fig_drought_responses`). 
-Comparing across droughts, there was little consistency in species- or individual-level drought responses, reflecting dissimilar effects of a key species trait (deciduousness) and variables linked to water access (TWI, DBH) under differing drought conditions.
+We show that, in one tropical seasonal forest, the drought sensitivity of tree growth is jointly and interactively shaped by drought characteristics, species adaptations, and local environmental conditions (Table 1).
+Analysing two droughts in a 14-year dendrometer band time series, we found lower-than-average growth in drought years, resulting in lower plot-level growth (Fig `r fig_drought_responses`). 
+Across droughts, there was little consistency in species- or individual-level drought responses, reflecting dissimilar effects of a key species trait (deciduousness) and variables linked to water access (TWI, DBH) under differing drought conditions.
 Despite lower absolute extremes of temperature, vapor pressure deficit, and number of dry days (Fig.
-2a), drought in the wet season (2015) elicited stronger responses across trees, especially deciduous species that are expected to have stronger seasonality in growth phenology (Fig `r fig_species`).
+2a), drought in the wet season (2015) elicited stronger responses across trees, especially deciduous species, which are expected to have stronger seasonality in growth phenology (Fig `r fig_species`).
 Interestingly, in 2015 (but not 2010), evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites (Fig `r fig_species`b, Fig `r fig_microenvs`b).
-While trees with more exposed crowns were consistently more sensitive to drought, additional effects of DBH -- presumably including rooting volume and depth -- were dissimilar between the drought events (Fig `r fig_microenvs`a).
-In other words, water availability buffered exposure-associated growth declines through different mechanisms in the two different droughts; larger trees (with likely deeper roots) fared better in the dry-season drought while trees in wetter sites suffered smaller growth declines in the wet-season drought (Fig `r fig_dag`).
-Taken together, these results indicate that joint and interactive effects of drought characteristics, species traits, and individual microenvironments produce enormous heterogeneity in individual and species drought responses of tropical trees.
+Although trees with more exposed crowns were consistently more sensitive to drought, additional effects of DBH -- presumably including rooting volume and depth -- were dissimilar between the drought events (Fig `r fig_microenvs`a).
+It is possible that water availability buffered exposure-associated growth declines through different mechanisms in the two different droughts; larger trees fared better in the dry-season drought while trees in wetter sites suffered smaller growth declines in the wet-season drought (Fig `r fig_dag`).
+<!-- we can talk about it ... but I don't think that we can know that larger trees have deeper roots.  -->
+Taken together, these results indicate that joint and interactive effects of drought characteristics, species traits, and individual microenvironments produce important heterogeneity in individual and species drought responses of tropical trees.
 This finding that individual droughts have divergent effects on species and individuals may contribute to the maintenance of species diveristy and ecosystem stability.
@@
 `r fig_droughts`).
-Thus, quantitative analyses parsing out the specific effects of drought characteristics (e.g., timing, degree of anomaly) are currently impossible.
-While our finding that a wet-season drought (2015) had greater impact on woody growth than a dry-season drought (2010) calls into question the generality of previous findings based on tree-ring chronologies from HKK [@vlam_temperature_2014;@anderson-teixeira_joint_2022] and across the tropics [@zuidema_tropical_2022] that tree growth is most sensitive to dry-season precipitation, care must be taken in interpretation given that the 2015 drought was more anomalous on multi-month time scales (Fig.
-`r fig_droughts`).
-Rigorous generalization on tropical tree growth responses to any specific drought characteristic would require annual growth records from numerous species over numerous droughts, but no such data set currently exists.
-Studies in temperate forests, where extensive tree-ring records are available, have confirmed that drought characteristics including seasonal timing, duration, and severity affect tree growth responses [@gao_dynamic_2018; @dorangeville_drought_2018] and that effects of species, species traits, tree size, and microenvironment can differ across droughts [@mcgregor_tree_2021].
+This is a central finding of our study.
+<!-- Quantitative analyses parsing out the specific effects of drought characteristics (e.g., timing, degree of anomaly) are currently impossible.
+I don't understand this, as we have done this, no?
+-->
+However, our finding that a wet-season drought in 2015 had greater impact on woody growth than a dry-season drought in 2010 questions the generality of previous findings.
+Tree ring studies from this site [@vlam_temperature_2014;@anderson-teixeira_joint_2022] and across the tropics [@zuidema_tropical_2022] have asserted that tree growth is most sensitive to dry-season precipitation.
+But  given that the 2015 drought was more anomalous on multi-month time scales (Fig.
+`r fig_droughts`) does not align with this assumption.
+Generalization of tropical tree growth responses to any specific drought seems to require fine-scale monitoring of growth (i.e., dendrometer band measurements or dendrochronology when possible) from species that might span importance and drought strategies over different timings and intensities of droughts.
+Advancing efforts to collect this level of data is critical to our assessment and prediction of drought responses of tropical forets to changing precipitation patterns.
+<!-- [is this necessary?]  Studies in temperate forests, where extensive tree-ring records are available, have confirmed that drought characteristics including seasonal timing, duration, and severity affect tree growth responses [@gao_dynamic_2018; @dorangeville_drought_2018] and that effects of species, species traits, tree size, and microenvironment can differ across droughts [@mcgregor_tree_2021].
 As tropical dendrometer band records continues to grow, so will the potential to better understand how specific drought characteristics interact with species traits and microenironment to shape growth respones.
+-->
 

### 2025-05-05 doc/manuscript_hkk_drought_sensitivity.Rmd
knit version with SM comments and HM edit to fig legend

@@
 ![***Figure `r fig_droughts`.
-Climatic characteristics of drought years and average years*** in the Huai Kha Khaeng ForestGEO plot.
-For four daily climate variables dry days, precipitation, maximum temperature and VPD, panel a) represents rolling means for 30 days around the focal day for that year.
-Grey line represents the mean of this rolling mean from 2008 to 2019 and ribbons represent standard deviation around this mean.
-Panel b) represents anomalies of this rolling mean from the values across the timeseries.
-Values, therefore represents the number of standard deviations from the mean on a given day.
-c) shows Standardised Precipitation Evapotranspiration Index (SPEI) values at four different time scales for each month in 2010 and 2015.
-A _k_-month SPEI integrates over _k_ months ending in the month current month.
+Climatic characteristics of the focal drought years (red in 2010, brown in 2015) compared with the average across years*** 2008-2019 (grey dashed) and its standard deviation (grey shading) in the Huai Kha Khaeng ForestGEO plot, for four daily climate variables: dry days, precipitation, maximum temperature and VPD.
+(a) 30-day running means.
+(b) Anomalies of the drought year running means from the averages, in units of standard deviations.
+(c) Standardised Precipitation Evapotranspiration Index (SPEI) values for the drought years at four different time scales for each month in 2010 and 2015.
+A k-month SPEI integrates over k months ending in the month current month.
 Across all panels, light blue background represents the average wet season, from May to October.

### 2025-05-09 doc/manuscript_hkk_drought_sensitivity.Rmd
simple text edits to methods and results from HM

@@
 We analysed dendrometer band measurements on 1820 trees across 30 species in a seasonally dry forest in Northern <!---Huai Kha Khaeng, ---->Thailand <!--using Bayesian regressions in a causal modelling framework---> to assess tree growth responses to two climatically distinct droughts.
-Species and individuals exhibited a wide range of growth responses within each drought, but these responses differed in intensity and affect  across droughts. 
+Species and individuals exhibited a wide range of growth responses within each drought, and these responses differed in intensity and affect between the two drought events. 
 Deciduous species were more sensitive to wet-season drought than evergreen species, and wetter microclimates amplified this difference. 
@@
 
-Globally, intense drought is becoming more frequent, affecting forest demography, diversity, and  carbon (C) cycling [@ipcc_climate_2023]. 
+Globally, droughts are becoming more frequent and intense, affecting forest demography, diversity, and  carbon (C) cycling [@ipcc_climate_2023]. 
 However, our ability to predict drought impacts on tree woody growth and the resultant ecosystem C sequestration remains limited, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022].
@@
 Our study site was the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
-This is a 50-hectare plot in mixed deciduous tropical forest, with a mean annual temperature of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
+This is a 50-hectare plot in mixed deciduous tropical forest, with a mean annual precipitation of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
 The region is periodically affected by droughts associated with the *El Niño* Southern Oscillation [ENSO, @räsänen_spatial_2016].
@@
 For analysis, we focussed on 30 of the most abundant species in the plot.   
-Long-term analysis of tree-ring data shows that tree growth at HKK has been influenced by temperature and rainfall [@vlam_temperature_2014].
+Analysis of long-term tree-ring data shows that interannual variation tree growth at HKK is influenced by temperature and rainfall [@vlam_temperature_2014].
 
@@
 
-In HKK, metal dendrometer bands were installed in 2008 on 2353 trees across 152 species following standard protocol [@muller-landau_metal_2008]. 
-Trees were selected were based on the relative abundance of species in the forest, with more individuals banded from the abundant species (ranging from 230 trees to 1 tree per species), and spanning the size and habitat distribution within the plot.
-On each tree, the dendrometer band was installed at 1.4 m from the ground, a height of 10 cm above the regular point of measure for the whole plot census (typically 1.3 m), except on irregular trunks or buttresses when they were shifted above.
-All dendrobands were censused two times a year - in the wet and dry season - typically in June and January. 
-During the census, the size of the window opening along the band was measured using Mitutoyo digital callipers (least count = 0.01 mm) and the diameter of the tree at the dendrometer band was measured using standard DBH tape.
+In HKK, metal dendrometer bands were installed in 2008 on 2353 trees across 152 species following the standard ForestGEO protocol [@muller-landau_metal_2008]. 
+Selected were a size-stratified and spatially stratified random sample, with additional trees chosen to fill out sample sizes for selected species of interest.
+Thus relative abudances among the selected trees largely reflected abundance variation in the plot as a whole, with more individuals banded from the abundant species (n=1 to 230 trees per species). 
+On each tree, the dendrometer band was installed at a height of 10 cm above the regular point of measurement for the whole plot census meaning installation at 1.4 m on most trees, and at higher heights on buttressed trees with higher measurement points.
+Dendrobands were censused twice a year with the measurement periods typically centred in June and January. 
+During each census, the size of the window opening along the band was measured using Mitutoyo digital callipers (precision = 0.01 mm) and the diameter of the tree at the dendrometer band was measured using standard DBH tape.
 
@@
 Using the late wet/early dry dendrometer band censuses, we calculated annualised growth for each year for each tree. 
-We first removed measurements with likely data entry errors because of misplaced decimals, potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses).
+We first conducted quality control on the dataset, removing measurements that appeared likely to be data entry errors (misplaced decimals), potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses).
 We used the raw window size measurements combined with the diameter measurement at installation $(d_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[@condit_dendrometer_].
@@
 
-To ensure that the data did not smooth over multiple years, we excluded increment measures with missing values for $d_{t-1}$ and then excluded large measurement outliers (> 3 standard deviations from the mean across all observations).
-Tape measurements themselves are not precise enough for calculating drought sensitivity; error for each tape measure is 1 mm, error for growth is at least 2 mm. 
+To ensure that the data did not smooth over multiple years, we excluded increment measures spanning missing values for $d_{t-1}$.
+We also excluded large measurement outliers (> 3 standard deviations from the mean increment across all observations). 
+Diameter tape measurements themselves are not precise enough for calculating drought sensitivity; error for each tape measure is 1 mm, error for growth is at least 2 mm. 
 In comparison, median annual growth across trees in HKK is ~1.4 mm and calculating deviation from annual growth needs sub-mm precision. 
-However, tape measurements can be used to check for consistency in trends or identify outliers. 
+However, tape measurements to flag likely errors in the dendrometer dataset. 
 We calculated increments from annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement with these annualised increments calculated from tape measurements.
-Finally, we excluded trees with negative increments over the whole timeseries, and included only species that had at least 10 individuals, resulting in a final dataset of 1820 individuals across 30 species.
+Finally, we excluded trees with negative increments over the whole timeseries, and included only species that had at least 10 individuals, resulting in a final dataset of 1820 individuals of 30 species.
 
 ![***Figure `r fig_drought_responses`: Growth increment timeseries and drought responses*** across trees and species in the Huai Kha Khaeng ForestGEO plot.
-a) Median of annualised growth increments across ten dominant species (colours) and all trees (black) from 2009 - 2022.
-b) Distribution of drought sensitivity of tree growth across all individuals in the two drought years.
+a) Median of annualised growth increments across ten dominant species (colours) and all trees (black) from 2009 to 2022.
+b) Distribution of drought sensitivity of tree growth across all individuals in the two drought years (see main text for equation).
 For comments - https://github.com/forestgeo/growth-precip-thailand/issues/11](display/Fig2.png)
@@
 
-We identified drought years of interest in the dendroband timeseries - 2010 and 2015 - based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses.
-Within the dendroband timeseries, two ENSO events had occurred; 2010 was a moderate ENSO event while 2015 was a very strong ENSO event (Oceanic Niño Index @bamston_documentation_1997). 
+We identified 2010 and 2015 as drought years of interest in the dendroband timeseries based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses.
+Both were El Niño years; 2010 was a moderate El Niño event while 2015 was a very strong El Niño event (Oceanic Niño Index @bamston_documentation_1997). 
 <!-- We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected.
@@
 
-For the two drought years, 2010 and 2015, we calculated growth sensitivity for each tree to each drought as a growth anomaly from its mean annual increment from calculated from dendrometer band window size.
+For the two drought years, 2010 and 2015, we calculated growth sensitivity for each tree to each drought as a growth anomaly from its mean annual increment over the entire time series.
 
@@
 
-We calculated deciduousness as a species level trait using data published in @williams_deciduousness_2008 for the top 30 species. 
-We defined deciduousness as the mean proportion of crown leaf loss at maximum loss across individuals in the species, with a score ranging from 1 to 4 where 1 is most evergreen and 4 is most deciduous.
+We assigned deciduousness as a species level trait using data published in @williams_deciduousness_2008 for the top 30 species. 
+We defined species deciduousness as the mean over trees of the proportion of crown leaf loss at maximum, with a score ranging from 1 to 4 where 1 is most evergreen and 4 is most deciduous.
 Data was unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1, in the middle of the evergreen range. 
-An alternate measure of deciduousness is the duration spent at maximum crown loss.
-This was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for a larger list of species.  
+We also considered using the duration of time spent at maximum crown loss as an alternative measure of deciduousness.
+This was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for more species.  
 <!-- At the species level, we also calculated the maximum size across all individuals in the plot as measured in the 2008 whole-plot census.
@@
 
+We fit alternative models for sensitivty including a null model with just species random effects, a model including only fixed effects of species deciduousness and habitat wetness and a model including size (DBH), light environment (CII) and habitat wetness as well as species random effects on all these.
 We first modelled sensitivity as a simple random intercept model
@@
 Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual*10, na.rm=T),2)` mm. 
-Median annual increment differed by species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm) among the `r nrow(tree.time.sp)` species analysed.
+Median annual increment varied among species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm) among the `r nrow(tree.time.sp)` species analysed.
 <!-- *(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))* -->
@@
 The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
-While median species responses to the 2010 drought were variable, all species had lower increment than a typical year in 2015. 
+While median species responses to the 2010 drought were variable, all species had lower median increment than a typical year in 2015. 
 <!-- Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S4).  -->
@@
 
-Species median sensitivities in 2010 and 2015 were not correlated with each other (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S5) but on average had lower mean and median sensitivities in 2015 than in 2010 (Fig S5), in alignment with median sensitivities across all trees.
+On average, species has lower mean and median sensitivies in 2015 than in 2010 (Figure S5), in alignment with median sensitivities across all trees. 
+Species median sensitivities in 2010were not correlated with median sensitivities in 2015 (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S5).
 
-Deciduousness species had lower negative sensitivities in 2015, but not in 2010 (Fig `r fig_species`a). 
+Deciduousness species responded more negatively than evergreen species in 2015, but not in 2010 (Fig `r fig_species`a). 
 In 2015, predicted sensitivities from intercept-only models of sensitivity with a species random effect on the intercept showed a significant negative association with deciduousness value (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`).
-On the other hand, in 2010, predicted sensitivities had no significant association with deciduousness (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`).
+In 2010, predicted sensitivities had no significant association with deciduousness (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`).
 Moreover, models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`) but not in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).
@@
 The interaction between TWI and deciduousness was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) but was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`).
-Combined, predicted sensitivities of deciduous trees was more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig `r fig_species`b).
+In 2015, evergreen trees responded more negatively in drier than in wetter areas, whereas deciduous trees - which were more negatively affected overall - responded more negatively in wetter than in drier areas.
+Combined, predicted sensitivities of deciduous trees were more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig `r fig_species`b).
 

### 2025-06-02 doc/manuscript_hkk_drought_sensitivity.Rmd
update with NP and SB edits

@@
 Helene C. Muller-Landau^3^  ORCiD : `r team$ORCID[5]`  
-Nantachai Pongpattananurak^5^  ORCiD: `r team$ORCID[6]`  
+Nantachai Pongpattananurak^4,5^  ORCiD: `r team$ORCID[6]`  
 Kristina Anderson-Teixeira^1,3^  ORCiD: `r team$ORCID[7]`
@@
 ^3^Smithsonian Tropical Research Institute, Panama  
-^4^National Park, Wildlife and Plant Conservation Department, Bangkok, Thailand  
-^5^Kasetsart University, Chatuchak, Bangkok, Thailand
+^4^4Thai Long-Term Forest Ecological Research, Kasetsart University, Bangkok, Thailand  
+^5^5Department of Forest Biology, Faculty of Forestry, Kasetsart University, Bangkok 10900, Thailand
 
@@
 Tree species adaptations to hot and dry conditions--for example, dry season deciduousness in tropical seasonal forests--might confer drought resistance, but such resistance may be variable across species, microenvironments, and drought events.
-We analysed dendrometer band measurements on 1820 trees across 30 species in a seasonally dry forest in Northern <!---Huai Kha Khaeng, ---->Thailand <!--using Bayesian regressions in a causal modelling framework---> to assess tree growth responses to two climatically distinct droughts.
+We analysed dendrometer band measurements on 1820 trees across 30 species in a seasonally dry tropical forest in Western <!---Huai Kha Khaeng, ---->Thailand <!--using Bayesian regressions in a causal modelling framework---> to assess tree growth responses to two climatically distinct droughts.
 Species and individuals exhibited a wide range of growth responses within each drought, and these responses differed in intensity and affect between the two drought events. 
@@
 iii) exposed trees are more sensitive to drought, but water availability, either through larger topographic location or tree size (and corresponding rooting volume), can buffer growth declines.
-We tested these hypothses using  a 14-year record of dendrometer band measurements in a seasonally dry forest in Thailand.
+We tested these hypotheses using  a 14-year record of dendrometer band measurements in a dry seasonal evergreen forest [@bunyavejchewin_seasonally_2011] in Thailand.
 With data on 1820 trees across 30 species we built Bayesian regression models in a causal modelling framework to account for potential confounding relationships among the variables.
@@
 Our study site was the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
-This is a 50-hectare plot in mixed deciduous tropical forest, with a mean annual precipitation of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
+This is a 50-hectare plot in dry seasonal evergreen forest, with a mean annual precipitation of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
 The region is periodically affected by droughts associated with the *El Niño* Southern Oscillation [ENSO, @räsänen_spatial_2016].
 The dominant canopy species are _Hopea odorata_, _Vatica harmandiana_, _Dipterocarpus alatus_ and _Tetrameles nudiflora_ and lower layers of the canopy are dominated by sub-canopy evergreen and deciduous species like _Miliusa horsfieldii_ and _Polyalthia viridis_.
-For analysis, we focussed on 30 of the most abundant species in the plot.   
+For analysis, we focused on 30 of the most abundant species in the plot.   
 Analysis of long-term tree-ring data shows that interannual variation tree growth at HKK is influenced by temperature and rainfall [@vlam_temperature_2014].
@@
 <!----Limitations in generalisation/ species examined here may have greater diversity of drought adaptaitons than elsewhere in the tropics---->
-We have demonstrated that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry tropical forest.
+We have demonstrated that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.
 Tropical dry forests represent periodic hot/dry conditions, creating selective pressures for adaptative variation towards drought resistance, and this might mean that species in tropical dry forests exhibit a greater diversity of drought responess than species in other bioclimatic zones where adaptive pressures and thus ecological variability are different.
@@
 
-We acknowledge the continued contributions of the Huai Kha Khaeng forest monitoring team in collecting data and maintaining the dendrometer bands. 
+We acknowledge the continued contributions of the monitoring team of the 50-ha forest dynamics plot at HKK in collecting data and maintaining the dendrometer bands. 
 Ekaphan Kraichak and Kanisorn Chowtiwuttakorn helped with procuring elevation and climate datasets. 

### 2025-06-02 doc/manuscript_hkk_drought_sensitivity.Rmd
SJD basic text edits

@@
 
-Globally, droughts are becoming more frequent and intense, affecting forest demography, diversity, and  carbon (C) cycling [@ipcc_climate_2023]. 
-However, our ability to predict drought impacts on tree woody growth and the resultant ecosystem C sequestration remains limited, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022].
+Intense droughts are becoming more frequent, affecting the demography, diversity, and  carbon (C) cycling of forests globally [@ipcc_climate_2023]. 
+However, our ability to predict drought impacts on tree woody growth and ecosystem C sequestration remains limited, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022].
 For forests worldwide, separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought.
@@
 In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in a different drought. 
-The resultant amplification of the complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].
+The complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].
 
@@
 Drought, or hot/dry conditions anomalously departing from climatic means during any time of year, is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the severity and duration of the climatic anomalies [@slette_how_2019].
-Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different timing, duration, or intensity may in fact have very different effects on tree growth, especially in seasonal forests that have evolved phenological strategies to handle regular periods of climatic stress [@albert_ecology_letters_cryptic_phenology] [@gao_dynamic_2018].
+Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different timing, duration, or intensity may have very different effects on tree growth, especially in seasonal forests that have evolved phenological strategies to handle regular periods of climatic stress [@albert_ecology_letters_cryptic_phenology] [@gao_dynamic_2018].
 <!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
@@
 We assigned deciduousness as a species level trait using data published in @williams_deciduousness_2008 for the top 30 species. 
-We defined species deciduousness as the mean over trees of the proportion of crown leaf loss at maximum, with a score ranging from 1 to 4 where 1 is most evergreen and 4 is most deciduous.
-Data was unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1, in the middle of the evergreen range. 
+We defined species deciduousness as the mean over trees of the proportion of crown leaf loss at maximum, with a score ranging from 0 to 4 where 0 is most evergreen and 4 is most deciduous.
+Data were unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1. 
 We also considered using the duration of time spent at maximum crown loss as an alternative measure of deciduousness.
@@
 
-The two droughts analysed had distinct temporal climatic characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r fig_droughts`).
+The two droughts analysed had distinct temporal characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r fig_droughts`).
 The 2010 drought peaked between February and May (dry season and beginning of wet season), with 1-month SPEI values below -1 and meteorological anomalies on the 30-day scale including lower precipitation (min anomaly **<-1.5** SD), higher number of dry days (max anomaly **>1.5** SD), higher maximum temperature (max anomaly +0.8 SD), and higher VPD (max anomaly **>2** SD).
@@
 <!-- make sure my adjustments above in details are correct ... I think we should present a precise description  -->
-1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census time period. 
+1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period. 
 Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season.
@@
 
-## Growth timeseries and drought responses {-}
+## Growth responses to drought {-}
 
-The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a). 
-Median annualised growth rate across the years of measurement for the whole plot was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual*10, na.rm=T),2)` mm. 
-Median annual increment varied among species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm) among the `r nrow(tree.time.sp)` species analysed.
+<!-- The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a).  -->
+Median annual growth rate across all years of measurement for the whole plot was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual*10, na.rm=T),2)` mm. 
+In the 2010 drought, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
+The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
+On average, trees had negative drought sensitivities for both droughts, but with considerable variation (Fig `r fig_drought_responses`b). 
+Median drought sensitivity across all trees in 2010 was `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)` while in 2015 it was `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`.
 <!-- *(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))* -->
 
-Drought years were associated with lower growth across the plot. 
-In 2010, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
-The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
-While median species responses to the 2010 drought were variable, all species had lower median increment than a typical year in 2015. 
+<!-- Drought years were associated with lower growth across the plot.  -->
 <!-- Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S4).  -->
@@
 <!-- *(See [issue #28](https://github.com/forestgeo/growth-precip-thailand/issues/28))* -->
-
-On average, trees had negative drought sensitivities for both droughts, but with considerable variation (Fig `r fig_drought_responses`b). 
-Median drought sensitivity across all trees in 2010 was `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)` while in 2015 it was `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`.
-In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/2]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/2]`*.
-In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/2]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/2]`*.
 
@@
 
-## Species effects {-}
+## Species and functional group drought responses {-}
 
-On average, species has lower mean and median sensitivies in 2015 than in 2010 (Figure S5), in alignment with median sensitivities across all trees. 
-Species median sensitivities in 2010were not correlated with median sensitivities in 2015 (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S5).
+Median annual increment varied among species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm) among the `r nrow(tree.time.sp)` species analysed.
+While median species responses to the 2010 drought were variable, all species had lower median increment than a typical year in 2015. 
+In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/2]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/2]`*.
+In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/2]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/2]`*.
 
-Deciduousness species responded more negatively than evergreen species in 2015, but not in 2010 (Fig `r fig_species`a). 
-In 2015, predicted sensitivities from intercept-only models of sensitivity with a species random effect on the intercept showed a significant negative association with deciduousness value (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`).
-In 2010, predicted sensitivities had no significant association with deciduousness (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`).
-Moreover, models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`) but not in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).
+Species median sensitivities to the 2010 drought were not correlated with median sensitivities in 2015 (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S5) but on average, species had more negative sensitivities to the 2015 drought than the 2010 drought, in alignment with median sensitivities across all trees.
+
+<!-- Deciduousness species responded more negatively than evergreen species in 2015, but not in 2010 (Fig `r fig_species`a).  -->
+Predicted sensitivities from intercept-only models with a species random effect showed that  deciduous species were more drought sensitive than evergreen species in 2015 (Fig `r fig_species`a, r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`), but not in 2010 (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`).
+Models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`) but not in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).
 
@@
 In 2015, evergreen trees responded more negatively in drier than in wetter areas, whereas deciduous trees - which were more negatively affected overall - responded more negatively in wetter than in drier areas.
-Combined, predicted sensitivities of deciduous trees were more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig `r fig_species`b).
+Predicted sensitivities of deciduous trees were more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig `r fig_species`b).
 
@@
 Predictions across the range of observed variables also showed that deciduous species generally had larger negative sensitivities in 2015, which did not  increase with TWI (Fig `r fig_microenvs`c). 
-On the other hand, neither the mean sensitivities across species nor their response to TWI were associated with deciduousness in 2010.
+Mean sensitivities across species and their response to TWI were not associated with deciduousness in 2010.
 
@@
 
-DBH was a significant predictor of crown exposure, with larger DBH associated with higher exposure categories in 2010 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`) and 2015 (median effect =`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) (Fig 5a).
-Controlling for this direct effect of exposure, the remaining effect of DBH *per se* was significantly positive in 2010, while it was weakly negative in 2015 (Fig `r fig_microenvs`a, Fig `r fig_dag`a).
+<!--DBH was a significant predictor of crown exposure, -->Trees with larger DBH had higher crown exposure in 2010 (median effect = `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`) and 2015 (median effect = `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) (Fig 5a).
+Controlling for the direct effect of exposure, the remaining effect of DBH *per se* was significantly positive in 2010, while it was weakly negative in 2015 (Fig `r fig_microenvs`a, Fig `r fig_dag`a).
 
@@
 `r fig_droughts`).
-This is a central finding of our study.
+<!-- This is a central finding of our study.  -->
 <!-- Quantitative analyses parsing out the specific effects of drought characteristics (e.g., timing, degree of anomaly) are currently impossible.
@@
 <!-- contributions / disentangling exposure from DBH-->
-One variable examined here -- crown exposure -- had a consistent effect across the two droughts (Table 1, Figs. `r fig_microenvs`, `r fig_dag`). 
+<!-- One variable examined here -- crown exposure -- had a consistent effect across the two droughts (Table 1, Figs. `r fig_microenvs`, `r fig_dag`).  -->
 Trees with more exposed crowns exhibited greater growth declines in both droughts, while sub-canopy trees grew more than average during the 2010 drought.
@@
 S5).
-More generally, tree-ring records are likely to give a biased picture of drought responses, as ring formation is caused by dry-season dormancy (often associated with deciduous leaf habit) and sampling often targets large trees with exposed crowns and dry microenvironments [@speer_fundamentals_2010]-- all factors shown here to directly and interactively influence drought sensitivity.<!-- More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects.
+More generally, tree-ring records are likely to give a biased picture of drought responses, as ring formation is caused by dry-season dormancy (often associated with deciduous leaf habit) and sampling often targets large trees with exposed crowns and dry microenvironments [@speer_fundamentals_2010]-- all factors shown here to <!--directly and interactively -->influence drought sensitivity.<!-- More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects.
 -->
@@
 José Medina-Vega, Valentine Herrmann and Eugenie Más provided input on improving analyses. 
-KA was supported by the Smithsonian Institution Postdoctoral Fellowship and the ForestGEO Analytical Workshop, Asia 2024 in completing this work.
+KA was supported by the Smithsonian Institution Postdoctoral Fellowship and the ForestGEO Analytical Workshop, Asia 2024 (supported by NSF grant DEB-[CHECK] to SJD) in completing this work.
 

### 2025-06-06 doc/manuscript_hkk_drought_sensitivity.Rmd
manuscript edits

@@
 # Statement of Authorship {-}
+KA, SM, SJD and KAT conceptualised the analysis. 
+SB, SJD and NP have been leading long-term data collection at HKK, while HMM created and helped execute the dendrometer band protocol. 
+KA cleaned the data and conducted formal analysis with support from SM, SJD, HMM and KAT. 
+KA wrote the original draft of the manuscript with support from KAT and SM. 
+All authors contributed to revision, review and editing of the manuscript.
 
@@
 # Conflict of Interest {-}
+The authors declare no conflict of interest.
 
@@
 Drought, or hot/dry conditions anomalously departing from climatic means during any time of year, is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the severity and duration of the climatic anomalies [@slette_how_2019].
-Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different timing, duration, or intensity may have very different effects on tree growth, especially in seasonal forests that have evolved phenological strategies to handle regular periods of climatic stress [@albert_ecology_letters_cryptic_phenology] [@gao_dynamic_2018].
+Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different timing, duration, or intensity may have very different effects on tree growth, especially in seasonal forests that have evolved phenological strategies to handle regular periods of climatic stress [@albert_cryptic_2019] [@gao_dynamic_2018].
 <!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
@@
 <!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
-Finally, the microenvironments that individual trees experience influence growth responses to drought.
-<!--leading to varied atmospheric and hydrological stresses with distinct ecological consequences.-->
-Across horizontal axes, microenvironments defined by topographic and edaphic variation shape tree drought responses. 
+<!-- Finally, the microenvironments that individual trees experience influence growth responses to drought leading to varied atmospheric and hydrological stresses with distinct ecological consequences.
+-->
+Microenvironments defined by horizontal topographic and edaphic variation, and vertical light and thermal variation further shape individual tree experience of and response to drought. 
 While habitats with relatively moist soils (e.g., low-lying topographic positions) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted (e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021) or acclimatized [@refs].
-Indeed, there are observations of trees near streams undergoing greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta_droughtinduced_2017] during drought. 
-Across vertical axes, trees of different sizes have their crowns and root systems in sometimes dramatically divergent microenvironments.
-Aboveground, tree crowns in canopy or emergent positions are exposed to higher evaporative demand and thermal stress driven by higher solar radiation, wind speed, and vapor pressure deficit [@vinod_thermal_2023], and such exposure makes trees more vulnerable to drought [@scharnweber_confessions_2019; @ma_tree_2023].
+Indeed, during drought trees near streams experienced greater growth declines in a temperate forest [@mcgregor_tree_2021] and increased mortality in a tropical forest [@zuleta_droughtinduced_2017]. 
+<!-- Across vertical axes, trees of different sizes have their crowns and root systems in sometimes dramatically divergent microenvironments. -->
+Tree crowns in canopy or emergent positions are exposed to higher evaporative demand and thermal stress driven by higher solar radiation, wind speed, and vapor pressure deficit [@vinod_thermal_2023], and such exposure makes trees more vulnerable to drought [@scharnweber_confessions_2019; @ma_tree_2023].
 <!--In contrast, understory trees may be rarely exposed to direct solar radiation and extreme evaporative demand.-->
@@
 it doesn't really make sense unless coupled with other traits (e.g., deciduousness shifts accross ontogeny) as 'putting down a tap root' is one of the first things a seedling of a deep rooted species does (i.e., it's not allometry but strategy) -->
-
 However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 

### 2025-06-06 scripts/000_twi.R
tpi stuff

@@ -395,3 +395,11 @@ tm_shape(clipped_raster) +
   # tm_symbols(size = 2, shape = 15, border.lwd = 2, col = NA) +
   tm_scale_bar()
 dev.off()
+
+
+# TPI ---------------------------
+install.packages("spatialEco")
+
+# convert elev_rast to a terra spatraster
+elev_rast_terra <-
+  tpi <- spatialEco::tpi(elev_rast)

### 2025-06-06 scripts/01_prepdata.R

@@ -311,6 +311,18 @@ twi <- raster::raster("data/HKK-other/TWI.tif")
 # pick twi values for each tree
 trees$twi <- raster::extract(twi, 0.1 * trees[, c("Y", "X")]) # multiply by 0.1 to get the correct resolution
 
+# tpi
+tpi <- raster::raster("data/HKK-other/TPI_7.tif")
+trees$tpi <- raster::extract(tpi, 0.1 * trees[, c("Y", "X")])
+
+# plot twi and tpi
+
+ggplot(trees, aes(x = twi, y = -tpi)) +
+  geom_point() +
+  geom_abline() +
+  theme_bw()
+
+
 # habitat---------------------------
 
 # read habitat data
@@ -334,7 +346,7 @@ trees <- merge(trees, dplyr::select(nci, -"X"), by = "Tag", all.x = T)
 
 tree_vars <- trees %>%
   rename(Species = SPCODE.UPDATE) %>%
-  dplyr::select(Tag, Species, X, Y, twi, habitat, nci_15)
+  dplyr::select(Tag, Species, X, Y, twi, tpi, habitat, nci_15)
 
 
 # species table---------------------------
@@ -380,6 +392,7 @@ hkk_census_growth <- hkk_census %>%
 
 hkk.stem4$habitat <- raster::extract(habitat_raster, hkk.stem4[c("gx", "gy")])
 hkk.stem4$twi <- raster::extract(twi, 0.1 * hkk.stem4[, c("gy", "gx")]) # multiply by 0.1 to get the correct resolution
+hkk.stem4$tpi <- raster::extract(tpi, 0.1 * hkk.stem4[, c("gy", "gx")]) # multiply by 0.1 to get the correct resolution
 head(hkk.stem4)
 
 # plot SACCLI trees on twi to make sure that the scale is right
@@ -446,7 +459,20 @@ sp.traits <- hkk.stem4 %>%
     # twi weighted mean with agb
     twi_agb_mean = mean(twi * agb, na.rm = T),
     twi_agb_median = median(twi * agb, na.rm = T),
-    twi_agb_sd = sd(twi * agb, na.rm = T)
+    twi_agb_sd = sd(twi * agb, na.rm = T),
+    # tpi mean
+    tpi_mean = mean(tpi, na.rm = T),
+    tpi_median = median(tpi, na.rm = T),
+    # twi sd
[…]

### 2025-06-06 scripts/07_treerings.R

@@ -3,16 +3,12 @@
 # load libraries
 # load the packages used by this script
 # library(plyr)
-library(tidyverse)
+# library(tidyverse)
 library(dplR)
-library(patchwork)
-library(ggplot2)
 # library(reshape2)
 # library(janitor)
 # library(lubridate)
 # library(httr)
-
-# Update rlang package
 
 # global variables
 start_date <- 1960
@@ -129,6 +125,8 @@ sens.all.tr <- ggplot(data = rw.hkk.2010, aes(x = sens.prop)) +
 png("doc/display/sens_all_tr.png", width = 4, height = 4, units = "in", res = 300)
 sens.all.tr
 dev.off()
+
+library(tidyverse)
 
 # load data--------------------------
 datasets <- readRDS("data/HKK-dendro/sensitivity_data.RData")
@@ -277,19 +275,9 @@ dbh_cii_sp <- ggplot(rw.hkk.2010, aes(
         strip.text.x = element_text(size = 8)
     )
 
-# make a table from the meta data
-sp.cii.table <- rw.hkk.2010 %>%
-    dplyr::mutate(Dawk = factor(Dawk, levels = c("2", "3", "4", "5", "NA"))) %>%
-    group_by(sp, Dawk) %>%
-    dplyr::summarise(n = n()) %>%
-    pivot_wider(names_from = Dawk, values_from = n, values_fill = 0, names_sort = F) %>%
-    ungroup() %>%
-    select("sp", "2", "3", "4", "5", "NA")
-
-sp.cii.table
-
+library(patchwork)
 png("doc/display/explore/tree_rings_dbh_cii.png", width = 8, height = 4, units = "in", res = 300)
-dbh_cii_all + gt::gt(sp.cii.table) + dbh_cii_sp
+(gt::gt(rw.hkk.2010[, c("sp", "Dawk")]) / dbh_cii_all) + dbh_cii_sp
 dev.off()
 
 # plot sensitivity distibution by canopy illumination for all and by species
@@ -322,7 +310,7 @@ sens_tr_ciisp <- ggplot(data = rw.hkk.2010, aes(x = Dawk, y = sens.prop, fill = 
 # save plots
 
 png("doc/display/explore/tree_ring_cii_sens.png", width = 8, height = 4, units = "in", res = 300)
-sens_tr_cii + gt::gt(sp.cii.table) + sens_tr_ciisp
[…]

### 2025-06-06 scripts/08_species.R

@@ -29,6 +29,7 @@ tree.time <- tree.time %>%
         cii_min1_scaled = scale(cii_min1, center = TRUE, scale = TRUE),
         cii_min1_scaled = ifelse(cii_min1_scaled == "NaN", 0, cii_min1_scaled),
         twi_scaled = scale(twi, center = TRUE, scale = TRUE),
+        tpi_scaled = scale(tpi, center = TRUE, scale = TRUE),
         median_inc_scaled = scale(median_inc, center = TRUE, scale = TRUE),
         avg_inc_tree_scaled = scale(avg_inc_tree, center = TRUE, scale = TRUE)
     ) %>%
@@ -38,7 +39,8 @@ tree.time <- tree.time %>%
         calcDBH_min1_scaled_sp = scale(calcDBH_min1, center = TRUE, scale = TRUE),
         cii_min1_scaled_sp = scale(cii_min1, center = TRUE, scale = TRUE),
         cii_min1_scaled_sp = ifelse(cii_min1_scaled_sp == "NaN", 0, cii_min1_scaled_sp),
-        twi_scaled_sp = scale(twi, center = TRUE, scale = TRUE)
+        twi_scaled_sp = scale(twi, center = TRUE, scale = TRUE),
+        tpi_scaled_sp = scale(tpi, center = TRUE, scale = TRUE)
     ) %>%
     ungroup() %>%
     # remove large outliers for each year
@@ -160,6 +162,7 @@ library(brms)
 isocline_model <- bf(sens.prop ~ 1 + twi + williams_dec + (1 + twi | Species))
 
 isocline_model_twi <- bf(sens.prop ~ 1 + twi + williams_dec + twi:williams_dec)
+isocline_model_tpi <- bf(sens.prop ~ 1 + tpi + williams_dec + tpi:williams_dec)
 
 
 coefs <- list()
@@ -185,10 +188,16 @@ newdata <- expand.grid(
     williams_dec = seq(min(tree.time$williams_dec, na.rm = T), max(tree.time$williams_dec, na.rm = T), length.out = 15)
 )
 
+newdata <- expand.grid(
+    tpi = seq(min(tree.time$tpi, na.rm = T), max(tree.time$tpi, na.rm = T), length.out = 15),
+    williams_dec = seq(min(tree.time$williams_dec, na.rm = T), max(tree.time$williams_dec, na.rm = T), length.out = 15)
+)
+
 # i<-2
 for (i in 1:length(yrs)) {
     # yrs<-c(2010, 2015, 2020)
[…]

### 2025-06-06 scripts/000_twi.R
update TPI calculations

@@ -7,6 +7,7 @@ library(terra)
 library(tidyverse)
 library(raster)
 library(tmap)
+library(spatialEco)
 
 library(whitebox)
 # whitebox::install_whitebox()
@@ -398,8 +399,42 @@ dev.off()
 
 
 # TPI ---------------------------
-install.packages("spatialEco")
+# library(spatialEco)
 
-# convert elev_rast to a terra spatraster
-elev_rast_terra <-
-  tpi <- spatialEco::tpi(elev_rast)
+# read elev_rast as terra SpatRaster
+
+elev_rast_terra <- terra::rast("data/HKK-other/HKKDEMsm_5m.tif")
+tpi_5 <- spatialEco::tpi(elev_rast_terra, scale = 5, win = "circle")
+
+png("doc/display/explore/tpi_5.png", width = 8, height = 8, units = "in", res = 300)
+tm_shape(tpi_5) +
+  tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE, alpha = 0.5) +
+  tm_scale_bar()
+dev.off()
+
+tpi_3 <- spatialEco::tpi(elev_rast_terra, scale = 3, win = "circle")
+png("doc/display/explore/tpi_3.png", width = 8, height = 8, units = "in", res = 300)
+tm_shape(tpi_3) +
+  tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE, alpha = 0.5) +
+  tm_scale_bar()
+dev.off()
+
+tpi_7 <- spatialEco::tpi(elev_rast_terra, scale = 7, win = "circle")
+png("doc/display/explore/tpi_7.png", width = 8, height = 8, units = "in", res = 300)
+tm_shape(tpi_7) +
+  tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE, alpha = 0.5) +
+  tm_scale_bar()
+dev.off()
+
+tpi_1 <- spatialEco::tpi(elev_rast_terra, scale = 1, win = "circle")
+png("doc/display/explore/tpi_1.png", width = 8, height = 8, units = "in", res = 300)
+tm_shape(tpi_1) +
+  tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE, alpha = 0.5) +
+  tm_scale_bar()
+dev.off()
+
+# write these as rasters
[…]

### 2025-06-17 scripts/06_figures.R
updated figures and old figs moved

@@ -211,7 +211,7 @@ speiplot <- ggplot() +
         data = data.frame(xmin = 4.5, xmax = 10.5, ymin = -Inf, ymax = Inf),
         aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), fill = "lightblue", alpha = 0.3
     ) +
-    geom_bar(data = spei_month %>% filter(year != "long-term"), aes(x = month, y = spei_val, fill = year), position = "dodge", stat = "identity") +
+    geom_bar(data = spei_month %>% filter(year %in% c("2010", "2015")), aes(x = month, y = spei_val, fill = year), position = "dodge", stat = "identity") +
     scale_fill_manual(values = c("long-term" = "grey40", "2010" = "indianred2", "2015" = "indianred4")) +
     theme_bw() +
     labs(x = "Month", y = "SPEI", fill = "Year") +
@@ -221,24 +221,6 @@ speiplot <- ggplot() +
     guides(fill = "none")
 
 
-speiplot_3yr <- ggplot() +
-    # geom_rect(data = data.frame(xmin = c(10.5, 0.5), xmax = c(12.5, 4.5), ymin = -Inf, ymax = Inf), aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), fill = "grey", alpha = 0.3) +
-    geom_rect(
-        data = data.frame(xmin = 4.5, xmax = 10.5, ymin = -Inf, ymax = Inf),
-        aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), fill = "lightblue", alpha = 0.3
-    ) +
-    geom_bar(data = spei_month %>% filter(year != "long-term", spei_var == "SPEI_01_month"), aes(x = month, y = spei_val, fill = year), position = "dodge", stat = "identity") +
-    # scale_fill_manual(values = c("long-term" = "grey40", "2010" = "indianred2", "2015" = "indianred4")) +
-    theme_bw() +
-    labs(x = "Month", y = "SPEI", fill = "Year") +
-    scale_x_continuous(breaks = 1:12) +
-    guides(linetype = "none") +
-    geom_hline(yintercept = c(0, -1, -2), linetype = "dashed") #+
-guides(fill = "none")
-
-png("doc/display/spei_plot_3yrs.png", width = 4, height = 4, units = "in", res = 300)
-speiplot_3yr
-dev.off()
 
[…]

### 2025-06-17 doc/manuscript_hkk_drought_sensitivity.Rmd
manuscript updates to figures, methods and results

@@
 (b) Anomalies of the drought year running means from the averages, in units of standard deviations.
-(c) Standardised Precipitation Evapotranspiration Index (SPEI) values for the drought years at four different time scales for each month in 2010 and 2015.
+(c) Standardised Precipitation Evapotranspiration Index (SPEI) values for the drought years at four different time scales for each month for each drought year.
 A k-month SPEI integrates over k months ending in the month current month.
 Across all panels, light blue background represents the average wet season, from May to October.
-For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1_new.png)
+For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1.png)
 
@@
 
-We identified 2010 and 2015 as drought years of interest in the dendroband timeseries based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses.
-Both were El Niño years; 2010 was a moderate El Niño event while 2015 was a very strong El Niño event (Oceanic Niño Index @bamston_documentation_1997). 
+<!-- We identified 2010 and 2015 as drought years of interest in the dendroband timeseries based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses.
+-->
+We identified 2010, 2015 and 2020 as drought years of interest in the dendroband timeseries based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses.
+Both 2010 and 2015 were El Niño years; 2010 was a moderate El Niño event while 2015 was a very strong El Niño event (Oceanic Niño Index @bamston_documentation_1997). 
+The drought conditions in 2020 were not associated with an El Niño event.
 <!-- We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected.
@@
 We also calculate Standardised Precipitation-Evapotranspiration Index (SPEI) at 1-, 3-, 6- and 12- month scales using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
-Defining drought as periods with two or more consecutive months with SPEI < -1 also resulted in identification of 2010 and 2015 as drought events. 
+Defining drought as periods with two or more consecutive months with SPEI < -1 also resulted in identification of 2010, 2015 and 2020 as drought events. 
 These years also corresponded with expert knowledge of major droughts that affected the region as well as meteorological records (see next paragraph; Fig.
@@
 <!-- Moreover, broad definitions based on climate variables (often not measured at the location), may not capture ecological realities within the plot.  -->
-We note that although tree growth was unusally low in 2020, this year did not emerge as a drought year either in the examination of drought indicators or from expert knowledge.
+<!-- We note that although tree growth was unusally low in 2020, this year did not emerge as a drought year either in the examination of drought indicators or from expert knowledge.     -->
 
 To characterise the two focal droughts, we used data from a weather station near the plot and publicly available climate data for the region. 
-We used daily weather data on maximum temperate, relative humidity and total precipitation collected from 2001 to 2019 at the Forest Fire Research Station <!--[CHECK WHERE]-->, located ~20 km from HKK, and calculated an additional variables vapour pressure deficit (VPD, @raesch_pvldcurve_2020).
-<!-- *(SPECIFY WHICH VARIABLES WERE OBTAINED FROM WHICH SOURCE, AND OTHER SOURCES.)* -->
-<!-- We first calculated daily values of Vapour Pressure Deficit (VPD) using the equation $$ -->
-<!-- VPD = 0.6108 * e^{17.27 * \frac{t_{max}}{t_{max}+237.3}}* (1-\frac{relative.humidity}{100}) -->
-<!-- $$ -->
+<!-- We used daily weather data on maximum temperate, relative humidity and total precipitation collected from 2001 to 2019 at the Forest Fire Research Station [CHECK WHERE], located ~20 km from HKK, and calculated an additional variables vapour pressure deficit (VPD, @raesch_pvldcurve_2020).
+-->
+We used daily weather data obtained from remotely sensed data sources CHIRPS and ERA5Land to calculate daily maximum temperature, total precipitation and vapour pressure deficit (VPD, @raesch_pvldcurve_2020).
 We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 30 day periods across the whole timeseries.
-Finally, we calculated daily rolling anomalies for each variable for 2010 and 2015 from the long-term daily expectation as $$anomaly_{day} = \frac{rolling.mean_{day} - \bar{rolling.mean_{day}}}{sd(rolling.mean_{day})}$$
+Finally, we calculated daily rolling anomalies for each variable for the drought years from the long-term daily expectation as $$anomaly_{day} = \frac{rolling.mean_{day} - \bar{rolling.mean_{day}}}{sd(rolling.mean_{day})}$$
 
@@
 
-For the two drought years, 2010 and 2015, we calculated growth sensitivity for each tree to each drought as a growth anomaly from its mean annual increment over the entire time series.
+For the three drought years, we calculated growth sensitivity for each tree to each drought as a growth anomaly from its mean annual increment over the entire time series.
 
@@
 
-Across all individuals in the plot, we calculated Topographic Wetness Index (TWI) for an expectation of the water availability across space based on the concavity/convexity of a location. 
+Across all individuals in the plot, we calculated Topographic Wetness Index (TWI) for an expectation of the water availability across space based on the total upslope area and the slope at the location.
 We used a Digital Elevation Model from a 5 m x 5 m elevation layer collected at plot establishment.
 We used the *whitebox* package in R [@wu_whitebox_2022; @lindsay_whitebox_2016a] to fill and breach this layer, calculate flow accumulation at each cell and then calculate TWI as $log_{e}\frac{As}{tan(slope)}$.
+As an alternate measure of water availability, we also calculated Topographic Position Index (TPI) based on the concavity/convexity of a location. 
 We defined tree size as the diameter at breast height (calculated from dendrometer band window measurement) at the previous census. 
@@
 
+We analysed data for each drought year by fitting separate models. 
 We fit alternative models for sensitivty including a null model with just species random effects, a model including only fixed effects of species deciduousness and habitat wetness and a model including size (DBH), light environment (CII) and habitat wetness as well as species random effects on all these.
@@
 
-The two droughts analysed had distinct temporal characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r fig_droughts`).
+<!-- The two droughts analysed had distinct temporal characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r fig_droughts`).
 The 2010 drought peaked between February and May (dry season and beginning of wet season), with 1-month SPEI values below -1 and meteorological anomalies on the 30-day scale including lower precipitation (min anomaly **<-1.5** SD), higher number of dry days (max anomaly **>1.5** SD), higher maximum temperature (max anomaly +0.8 SD), and higher VPD (max anomaly **>2** SD).
@@
 1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought.
+ -->
 
+<!-- In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including precipitation (min anomaly **<-1.5** SD), number of dry days (max anomaly **>1.5** SD), maximum monthly mean temperatures (max anomaly +1.5 SD), and monthly maximum VPD (max anomaly +2 SD).
+1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period. 
+Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season.
+1- to 6-month SPEI values reached minima <2, indicative of extreme drought, while the 12-month values remained just above -2, indicative of severe drought. -->
+
+The two droughts analysed had distinct climatic and temporal characteristics, with 2010 primarily a dry season drought, 2015 a wet-season drought and 2020 showing mixed characteristics (Table 1, Fig `r fig_droughts`).
+The 2010 drought peaked between February and May (dry season and beginning of wet season), with 1-month SPEI values below -1 and meteorological anomalies on the 30-day scale including lower precipitation (min anomaly **<-1.5** SD), higher number of dry days (max anomaly **>1.5** SD), higher maximum temperature (max anomaly +0.8 SD), and higher VPD (max anomaly **>2** SD).
+Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions.
+1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought. 
 In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including precipitation (min anomaly **<-1.5** SD), number of dry days (max anomaly **>1.5** SD), maximum monthly mean temperatures (max anomaly +1.5 SD), and monthly maximum VPD (max anomaly +2 SD).
-<!-- make sure my adjustments above in details are correct ... I think we should present a precise description  -->
 1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period. 
@@
 1- to 6-month SPEI values reached minima <2, indicative of extreme drought, while the 12-month values remained just above -2, indicative of severe drought. 
+In 2020, meteorological anomalied were pronounced both in the dry and wet seasons, including maximum temperature (max anomaly **> 2** SD) and VPD (max anomaly **> 2** SD). 
+1- to 6-month SPEI values reached minima ~2 in February, indicative of borderline severe drought.
+12-month SPEI showed severe to moderate drought conditions from January to August, spanning both dry and wet seasons.
 
@@
 sens.sp <- tree.time %>%
-  filter(yr %in% c(2010, 2015)) %>%
+  filter(yr %in% c(2010, 2015, 2020)) %>%
   group_by(spfull, yr) %>%
@@
 
-# paired correlation of 2010 and 2015 sensitivities across species
+# paired correlation of 2010, 2015 and 2020 sensitivities across species
 
-sens.sp.cor <- cor.test(sens.sp.wide$sens.2010, sens.sp.wide$sens.2015)
+sens.sp.cor_10_15 <- cor.test(sens.sp.wide$sens.2010, sens.sp.wide$sens.2015)
+sens.sp.cor_15_20 <- cor.test(sens.sp.wide$sens.2015, sens.sp.wide$sens.2020)
+sens.sp.cor_10_20 <- cor.test(sens.sp.wide$sens.2010, sens.sp.wide$sens.2020)
 ```
@@
 In the 2010 drought, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
+In the 2020 drought, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
 The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
-On average, trees had negative drought sensitivities for both droughts, but with considerable variation (Fig `r fig_drought_responses`b). 
-Median drought sensitivity across all trees in 2010 was `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)` while in 2015 it was `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`.
+On average, trees had negative drought sensitivities during drought years, but with considerable variation (Fig `r fig_drought_responses`b). 
+Median drought sensitivity across all trees in 2010 was `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)` in 2015  was `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)` and in 2020 was `r round(median(tree.time$sens.prop[tree.time$yr==2020]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2020]), 2)`.
 <!-- *(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))* -->
@@
 <!-- *(See [issue #28](https://github.com/forestgeo/growth-precip-thailand/issues/28))* -->
+
+## Species and functional group drought responses {-}
+
+Median annual increment varied among species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm) among the `r nrow(tree.time.sp)` species analysed.
+While median species responses to the 2010 and 2020 droughts were variable, all species had lower median increment than a typical year in 2015. 
+In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/3]`*.
+In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/3]`*.
+In 2020, sensitivities ranged from `r sens.sp[sens.sp$yr==2020,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2020,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[nrow(sens.sp)/3]`*.
+
+Species median sensitivities to the three droughts were not correlated with each other although on average, species had more negative sensitivities to the 2015 drought than the 2010 or 2020 droughts, in alignment with median sensitivities across all trees (for 2010 and 2015 Pearson's correlation coefficient r = `r round(sens.sp.cor_10_15[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_15[3]$p.value,2)`, for 2015 and 2020 was r = `r round(sens.sp.cor_15_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_15_20[3]$p.value,2)`, and for  2010 and 2020 r = `r round(sens.sp.cor_10_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_20[3]$p.value,2)`, Fig S5).
 
@@
   pull(median)
-intercepts <- rep(intercepts, each = nrow(ranef_df) / 2)
+intercepts <- rep(intercepts, each = nrow(ranef_df) / 3)
 ranef_df <- ranef_df %>%
@@
 
-isocline_coefs <- readRDS("../results/models/non_negative/coefs_isoclines_nore.RDS")
+isocline_coefs <- readRDS("../results/models/non_negative/coefs_isoclines_twi.RDS")
 # isocline_coefs <- readRDS("results/models/non_negative/coefs_isoclines_nore.RDS")
@@
 
-## Species and functional group drought responses {-}
+<!-- Deciduousness species responded more negatively than evergreen species in 2015, but not in 2010 (Fig `r fig_species`a).  -->
+Predicted sensitivities from intercept-only models with a species random effect showed heterogenous sensitivities of deciduous and evergreen species to different droughts.
+Deciduous species were more drought sensitive than evergreen species in 2015 (Fig `r fig_species`a, r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`), but the opposite was true in 2020 (`r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2020], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2020], 2)`).
+There was no difference in 2010 (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`). 
+Models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`,`r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`), positive in 2020 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2020],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2020],2)`,`r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2020],2)`), but no relationship in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).
 
-Median annual increment varied among species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm) among the `r nrow(tree.time.sp)` species analysed.
-While median species responses to the 2010 drought were variable, all species had lower median increment than a typical year in 2015. 
-In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/2]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/2]`*.
-In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/2]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/2]`*.
-
-Species median sensitivities to the 2010 drought were not correlated with median sensitivities in 2015 (Pearson's correlation coefficient r = `r round(sens.sp.cor[4]$estimate, 2)`, p = `r round(sens.sp.cor[3]$p.value,2)`, Fig S5) but on average, species had more negative sensitivities to the 2015 drought than the 2010 drought, in alignment with median sensitivities across all trees.
-
-<!-- Deciduousness species responded more negatively than evergreen species in 2015, but not in 2010 (Fig `r fig_species`a).  -->
-Predicted sensitivities from intercept-only models with a species random effect showed that  deciduous species were more drought sensitive than evergreen species in 2015 (Fig `r fig_species`a, r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`), but not in 2010 (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`).
-Models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`) but not in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).
-
-![***Figure `r fig_species`: Variation in sensitivity among species*** in the Huai Kha Khaeng ForestGEO plot during droughts in 2010 and 2015.
+![***Figure `r fig_species`: Variation in sensitivity among species*** in the Huai Kha Khaeng ForestGEO plot during droughts in 2010, 2015 and 2020.
 a) Relationship of the mean species sensitivities from model fits of an intercept-only model, with mean species deciduousness values.
@@
 
-The test of whether individual trees responded similarly to the two drought events showed no correlation (Fig S6).
+The test of whether individual trees responded similarly to the three drought events showed no correlation (Fig S6).
 
-TWI showed no significant effect on sensitivity in 2010, but did show a positive effect in 2015. 
-Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`).
-The interaction between TWI and deciduousness was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) but was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`).
-In 2015, evergreen trees responded more negatively in drier than in wetter areas, whereas deciduous trees - which were more negatively affected overall - responded more negatively in wetter than in drier areas.
+TWI showed significant positive effect in 2015 and 2020 but no significant effect on sensitivity in 2010. 
+Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`) and 2020 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2020],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2020],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2020],2)`).
+The interaction between TWI and deciduousness was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`) but was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) or 2020, (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2020],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2020],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2020],2)`)
+<!-- In 2015, evergreen trees responded more negatively in drier than in wetter areas, whereas deciduous trees - which were more negatively affected overall - responded more negatively in wetter than in drier areas.
+-->
 Predicted sensitivities of deciduous trees were more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig `r fig_species`b). 
+In 2020, predicted sensitivities of deciduous trees were more positive than evergreen trees and this effect was observable in drier sites (Fig `r fig_species`b).
 
-The differing effects of TWI on sensitivity in the two droughts, along with the influence of deciduousness, were confirmed by models that controlled for crown exposure and DBH as well as species variation in intercept and slopes (Fig `r fig_microenvs`a).
-These models showed small negative sensitivities in wetter sites in 2010 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but a strong positive effect of TWI in 2015 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`).
-The positive effect of TWI in 2015 was, however, mainly due to the response of evergreen species; deciduousness had a significant negative correlation with the slopes of the TWI effect in 2015 (Fig `r fig_microenvs`b).
+The differing effects of TWI on sensitivity in the three droughts, along with the influence of deciduousness, were confirmed by models that controlled for crown exposure and DBH as well as species variation in intercept and slopes (Fig `r fig_microenvs`a).
+These models showed small negative effect of wetness in 2010 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but a positive effect of wetness in 2015 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`) and 2020 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2020],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2020],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2020],2)`).
+<!-- The positive effect of TWI in 2015 was, however, mainly due to the response of evergreen species; deciduousness had a significant negative correlation with the slopes of the TWI effect in 2015 (Fig `r fig_microenvs`b).
+-->
 Predictions across the range of observed variables also showed that deciduous species generally had larger negative sensitivities in 2015, which did not  increase with TWI (Fig `r fig_microenvs`c).  
-Mean sensitivities across species and their response to TWI were not associated with deciduousness in 2010.
+Mean sensitivities across species and their response to TWI were not associated with deciduousness in 2010 or 2020, mirroring the response in the interaction model.
 
@@
 a) Coefficient plots showing median effects and 90% CI for wetness, exposure and DBH of tree.
-b) wetness effect across all species and its relationship with deciduousness value.
-Line represents a significant correlation.
-c) Predicted relationship of drought sensitivity with topographic wetness index across all species derived from model predictions.
+b) Predicted relationship of drought sensitivity with topographic wetness index across all species derived from model predictions.
 Black line represents overall relationship (panel a); coloured lines for each species reflects deciduousness values.
@@
 
-Crown exposure was associated with stronger negative sensitivities in both droughts.
-In the same models with species random effects on intercept and slope, the simplex parameter for the ordered factor crown exposure in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 the negative effect was not as strong (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`).
+Crown exposure was associated with more negative sensitivities in all three droughts.
+In the same models with species random effects on intercept and slope, the simplex parameter for the ordered factor crown exposure in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 and 2020 the negative effect was not as strong (2015 median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`; 2020 median effect = `r round(coefs_spre_cii$median[coefs_spre_cii$yr==2020],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2020],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2020],2)`).
 Predicted effects of crown exposure on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all crown exposure categories had negative sensitivities in 2015 (Fig `r fig_dag`b).
 
-<!--DBH was a significant predictor of crown exposure, -->Trees with larger DBH had higher crown exposure in 2010 (median effect = `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`) and 2015 (median effect = `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) (Fig 5a).
-Controlling for the direct effect of exposure, the remaining effect of DBH *per se* was significantly positive in 2010, while it was weakly negative in 2015 (Fig `r fig_microenvs`a, Fig `r fig_dag`a).
+<!--DBH was a significant predictor of crown exposure, -->Trees with larger DBH had higher crown exposure in 2010 (median effect = `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`), 2015 (`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) and 2020 (`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2020],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2020],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2020],2)`) (Fig 5a).
+Controlling for the direct effect of exposure, the remaining effect of DBH *per se* was significantly positive in 2010, and not significant in 2015 or 2020 (Fig `r fig_microenvs`a, Fig `r fig_dag`a).
 
 ![***Figure `r fig_dag`: Effects of size, exposure and water availability*** from combined models.
-Hypothesised relationships between microenvironment variables and drought sensitivity with their modelled effects in the HKK ForestGEO plot in the droughts of a) 2010 and b) 2015.
+Hypothesised relationships between microenvironment variables and drought sensitivity with their modelled effects in the HKK ForestGEO plot in the droughts of a) 2010, b) 2015 and c) 2020.
 Solid lines represent relationships where 90% credible intervals do not overlap 0, blue and red lines represent positive and negative effects respectively and line thickness is scaled to the effect size.
-c) Modelled sensitivity across crown exposure classes in 2010 and 2015.
+d) Modelled sensitivity across crown exposure classes in 2010, 2015 and 2020.
 For comments - https://github.com/forestgeo/growth-precip-thailand/issues/14](display/Fig5.png)

### 2025-06-18 doc/manuscript_hkk_drought_sensitivity.Rmd
updated to summary table, fig1 and manus

@@
 Tree species adaptations to hot and dry conditions--for example, dry season deciduousness in tropical seasonal forests--might confer drought resistance, but such resistance may be variable across species, microenvironments, and drought events.
-We analysed dendrometer band measurements on 1820 trees across 30 species in a seasonally dry tropical forest in Western <!---Huai Kha Khaeng, ---->Thailand <!--using Bayesian regressions in a causal modelling framework---> to assess tree growth responses to two climatically distinct droughts.
-Species and individuals exhibited a wide range of growth responses within each drought, and these responses differed in intensity and affect between the two drought events. 
+We analysed dendrometer band measurements on 1820 trees across 30 species in a seasonally dry tropical forest in Western <!---Huai Kha Khaeng, ---->Thailand <!--using Bayesian regressions in a causal modelling framework---> to assess tree growth responses to three climatically distinct droughts.
+Species and individuals exhibited a wide range of growth responses within each drought, and these responses differed in intensity and affect among the drought events. 
 Deciduous species were more sensitive to wet-season drought than evergreen species, and wetter microclimates amplified this difference. 
@@
 For analysis, we focused on 30 of the most abundant species in the plot.   
-Analysis of long-term tree-ring data shows that interannual variation tree growth at HKK is influenced by temperature and rainfall [@vlam_temperature_2014].
+<!-- Analysis of long-term tree-ring data shows that interannual variation tree growth at HKK is influenced by temperature and rainfall [@vlam_temperature_2014].   -->
 
@@
 We identified 2010, 2015 and 2020 as drought years of interest in the dendroband timeseries based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses.
-Both 2010 and 2015 were El Niño years; 2010 was a moderate El Niño event while 2015 was a very strong El Niño event (Oceanic Niño Index @bamston_documentation_1997). 
-The drought conditions in 2020 were not associated with an El Niño event.
-<!-- We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected.
--->
-We also calculate Standardised Precipitation-Evapotranspiration Index (SPEI) at 1-, 3-, 6- and 12- month scales using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
-Defining drought as periods with two or more consecutive months with SPEI < -1 also resulted in identification of 2010, 2015 and 2020 as drought events. 
+We calculated Standardised Precipitation-Evapotranspiration Index (SPEI) at 1-, 3-, 6- and 12- month scales using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
+Defining drought as periods with long-term SPEI < -1 resulted in identification of 2010, 2015 and 2020 as drought events. 
 These years also corresponded with expert knowledge of major droughts that affected the region as well as meteorological records (see next paragraph; Fig.
@@
 `r fig_drought_responses`).
+Among these, 2010 and 2015 were El Niño years; 2010 was a moderate El Niño event while 2015 was a very strong El Niño event (Oceanic Niño Index @bamston_documentation_1997). 
+The drought in 2020 was not associated with an El Niño event.
+<!-- We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected.
+-->
 <!-- We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.  -->
@@
 
-To characterise the two focal droughts, we used data from a weather station near the plot and publicly available climate data for the region. 
+<!-- To characterise the two focal droughts, we used data from a weather station near the plot and publicly available climate data for the region.  -->
 <!-- We used daily weather data on maximum temperate, relative humidity and total precipitation collected from 2001 to 2019 at the Forest Fire Research Station [CHECK WHERE], located ~20 km from HKK, and calculated an additional variables vapour pressure deficit (VPD, @raesch_pvldcurve_2020).
 -->
-We used daily weather data obtained from remotely sensed data sources CHIRPS and ERA5Land to calculate daily maximum temperature, total precipitation and vapour pressure deficit (VPD, @raesch_pvldcurve_2020).
+To characterise the three droughts, we used daily weather data obtained from publicly available data sources. 
+We used daily weather data obtained from remotely sensed data source CHIRPS [@funk_climate_2015] for daily total precipitation and ERA5Land [@muñoz-sabater_era5land_2021] to calculate daily maximum temperature and vapour pressure deficit (VPD, @raesch_pvldcurve_2020).
 We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 30 day periods across the whole timeseries.
@@
 We analysed data for each drought year by fitting separate models. 
-We fit alternative models for sensitivty including a null model with just species random effects, a model including only fixed effects of species deciduousness and habitat wetness and a model including size (DBH), light environment (CII) and habitat wetness as well as species random effects on all these.
+We fit alternative models for sensitivity including a null model with just species random effects, a model including only fixed effects of species deciduousness and habitat wetness and a model including size (DBH), light environment (CII) and habitat wetness as well as species random effects on all these.
 We first modelled sensitivity as a simple random intercept model
@@
 
-The two droughts analysed had distinct climatic and temporal characteristics, with 2010 primarily a dry season drought, 2015 a wet-season drought and 2020 showing mixed characteristics (Table 1, Fig `r fig_droughts`).
-The 2010 drought peaked between February and May (dry season and beginning of wet season), with 1-month SPEI values below -1 and meteorological anomalies on the 30-day scale including lower precipitation (min anomaly **<-1.5** SD), higher number of dry days (max anomaly **>1.5** SD), higher maximum temperature (max anomaly +0.8 SD), and higher VPD (max anomaly **>2** SD).
-Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions.
+The three droughts analysed had distinct climatic and temporal characteristics, with 2010 primarily a dry season drought, 2015 a wet-season drought and 2020 showing mixed characteristics (Table 1, Fig `r fig_droughts`).
+The 2010 drought peaked between February and May with meteorological anomalies on the 30-day scale including lower precipitation (min anomaly **<-1.5** SD), higher number of dry days (max anomaly **>1.5** SD), higher maximum temperature (max anomaly +0.8 SD), and higher VPD (max anomaly **>2** SD).
+<!-- Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions. -->
 1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought. 
@@
 1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period. 
-Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season.
+<!-- Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season. -->
 1- to 6-month SPEI values reached minima <2, indicative of extreme drought, while the 12-month values remained just above -2, indicative of severe drought. 
@@
 <!-- The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a).  -->
+Drought years were associated with lower plot-level median growth and greater tree-level sensitivities, with the strongest response in 2015 (Fig `r fig_drought_responses`). 
 Median annual growth rate across all years of measurement for the whole plot was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual*10, na.rm=T),2)` mm. 
-In the 2010 drought, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
-In the 2020 drought, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
 The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
-On average, trees had negative drought sensitivities during drought years, but with considerable variation (Fig `r fig_drought_responses`b). 
-Median drought sensitivity across all trees in 2010 was `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)` in 2015  was `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)` and in 2020 was `r round(median(tree.time$sens.prop[tree.time$yr==2020]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2020]), 2)`.
+In the 2010 and 2020 droughts, the median annual increments were `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm and `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm respectively, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% and `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
+<!-- In the 2020 drought, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
+-->
+<!-- On average, trees had negative drought sensitivities during drought years, but with considerable variation (Fig `r fig_drought_responses`b).  -->
+Median drought sensitivity across all trees in 2010, 2015 and 2020 were `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)`, `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)` and `r round(median(tree.time$sens.prop[tree.time$yr==2020]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2020]), 2)` respectively.
 <!-- *(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))* -->
@@
 However, given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts.-->
-The two droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, magnitude and duration of meteorological extremes and anomalies, and severity as classified the SPEI metric (Fig.
+The three droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, magnitude and duration of meteorological extremes and anomalies, and severity as classified the SPEI metric (Fig.
 `r fig_droughts`).
@@
 -->
-Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in the two different droughts, showing that ecological and evolutionary strategies make particular species and trees resistant to some droughts but vulnerable to others.
+Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in three different droughts, showing that ecological and evolutionary strategies make particular species and trees resistant to some droughts but vulnerable to others.
 These distinct responses of species and individuals to drought characteristics suggests potential spatiotemporal insurance effects among tropical tree response to drought, in alignment with recent global analyses (@liu_drought_2022).

### 2025-06-18 scripts/06_figures.R

@@ -574,9 +574,8 @@ speiplot_line <- ggplot() +
         col = "black", hjust = 0, vjust = 0, fontface = "italic", size = 3
     ) +
     ggtitle("SPEI") +
-    # scale_color_manual(values = c("long-term" = "grey40", "2010" = "indianred2", "2015" = "indianred4")) +
-    # ggh4x::stat_difference(data = spei_month, aes(x = month, ymin = -1, ymax = spei_val), levels = c("above", "below")) +
-    # scale_fill_manual(limits = c("above", "below"), values = c("grey40", "grey60")) +
+    # scale_color_manual(values = c("2010" = "indianred2", "2015" = "indianred4")) +
+    viridis::scale_color_viridis(discrete = T) +
     facet_wrap(~spei_var,
         # scales = "free_y",
         strip.position = "left", nrow = 4, labeller = varnames
@@ -667,6 +666,7 @@ climsat_plot <- ggplot(climsat_rlmean %>% filter(year %in% c("2010", "2015", "20
     guides(linetype = "none", color = "none") +
     ggtitle("Climate variables (remote)") +
     # scale_color_manual(values = c("2010" = "indianred2", "2015" = "indianred4")) +
+    viridis::scale_color_viridis(discrete = T) +
     theme(
         strip.background = element_blank(),
         strip.placement = "outside"
@@ -692,6 +692,7 @@ climsat_anomaly_plot <- ggplot(climsat_rlmean %>% filter(year %in% c("2010", "20
     # scale_x_continuous(breaks = 1:12) +
     guides(linetype = "none") +
     ggtitle("Anomalies") +
+    viridis::scale_color_viridis(discrete = T) +
     # scale_color_manual(values = c("2010" = "indianred2", "2015" = "indianred4")) +
     theme(
         strip.background = element_blank(),
@@ -713,6 +714,61 @@ ABC
 png("doc/display/Fig1.png", width = 8, height = 8, units = "in", res = 300)
 climsat_plot + climsat_anomaly_plot + speiplot_line + plot_layout(design = layout, guides = "collect") + plot_annotation(tag_levels = "a") & theme(legend.position = "bottom", legend.text = element_text(size = 18), legend.title = element_text(size = 18))
 dev.off()
+
+# correlation between remote and ground data and anomalies----------------------
[…]

### 2025-06-18 scripts/06_figures.R
Fig 1 with better colours

@@ -546,6 +546,9 @@ df <- data.frame(
     label = c("wetter", "drier", "drought", "severe drought")
 )
 
+cols <- viridis::viridis(4, option = "F")
+
+
 # spei plot with lines
 speiplot_line <- ggplot() +
     # geom_rect(data = data.frame(xmin = c(10.5, 0.5), xmax = c(12.5, 4.5), ymin = -Inf, ymax = Inf), aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), fill = "grey", alpha = 0.3) +
@@ -575,7 +578,8 @@ speiplot_line <- ggplot() +
     ) +
     ggtitle("SPEI") +
     # scale_color_manual(values = c("2010" = "indianred2", "2015" = "indianred4")) +
-    viridis::scale_color_viridis(discrete = T) +
+    # viridis::scale_color_viridis(discrete = T, option="H") +
+    scale_color_manual(values = cols[1:3]) +
     facet_wrap(~spei_var,
         # scales = "free_y",
         strip.position = "left", nrow = 4, labeller = varnames
@@ -666,7 +670,8 @@ climsat_plot <- ggplot(climsat_rlmean %>% filter(year %in% c("2010", "2015", "20
     guides(linetype = "none", color = "none") +
     ggtitle("Climate variables (remote)") +
     # scale_color_manual(values = c("2010" = "indianred2", "2015" = "indianred4")) +
-    viridis::scale_color_viridis(discrete = T) +
+    # viridis::scale_color_viridis(discrete = T) +
+    scale_color_manual(values = cols[1:3]) +
     theme(
         strip.background = element_blank(),
         strip.placement = "outside"
@@ -692,8 +697,9 @@ climsat_anomaly_plot <- ggplot(climsat_rlmean %>% filter(year %in% c("2010", "20
     # scale_x_continuous(breaks = 1:12) +
     guides(linetype = "none") +
     ggtitle("Anomalies") +
-    viridis::scale_color_viridis(discrete = T) +
+    # viridis::scale_color_viridis(discrete = T) +
     # scale_color_manual(values = c("2010" = "indianred2", "2015" = "indianred4")) +
+    scale_color_manual(values = cols[1:3]) +
     theme(
         strip.background = element_blank(),
         strip.placement = "outside"

### 2025-06-18 scripts/06_figures.R
docx knit with new Fig1

@@ -721,6 +721,8 @@ png("doc/display/Fig1.png", width = 8, height = 8, units = "in", res = 300)
 climsat_plot + climsat_anomaly_plot + speiplot_line + plot_layout(design = layout, guides = "collect") + plot_annotation(tag_levels = "a") & theme(legend.position = "bottom", legend.text = element_text(size = 18), legend.title = element_text(size = 18))
 dev.off()
 
+# supplementary figures with remote data----------------------
+
 # correlation between remote and ground data and anomalies----------------------
 head(ffstation_anomalies_full)
 head(climsat_rlmean)

### 2025-06-18 scripts/06_figures.R
climate timeseries for SI updated

@@ -721,7 +721,110 @@ png("doc/display/Fig1.png", width = 8, height = 8, units = "in", res = 300)
 climsat_plot + climsat_anomaly_plot + speiplot_line + plot_layout(design = layout, guides = "collect") + plot_annotation(tag_levels = "a") & theme(legend.position = "bottom", legend.text = element_text(size = 18), legend.title = element_text(size = 18))
 dev.off()
 
-# supplementary figures with remote data----------------------
+# SI figures with remote data----------------------
+
+varnames <- as_labeller(c(
+    precipitation = "Precipitation (mm)",
+    dry_day = "Dry days",
+    tmax = "Max temperature (°C)",
+    vpdmax = "VPD max (kPa)",
+    vpdmin = "VPD min (kPa)",
+    vpdmean = "VPD mean (kPa)",
+    spei_val = "SPEI"
+))
+
+
+climplot_SI <- ggplot(
+    climsat_rlmean %>%
+        filter(climvar %in% c("precipitation", "tmax", "dry_day", "vpdmax")) %>%
+        filter(year >= 2009)
+) +
+    annotate(
+        geom = "rect",
+        xmin = as.Date("2015-01-01"), xmax = as.Date("2016-01-01"), ymin = -Inf, ymax = Inf,
+        fill = cols[2], alpha = 0.5, col = cols[2]
+    ) +
+    annotate(
+        geom = "rect",
+        xmin = as.Date("2010-01-01"), xmax = as.Date("2011-01-01"), ymin = -Inf, ymax = Inf,
+        fill = cols[1], alpha = 0.5, col = cols[1]
+    ) +
+    annotate(
+        geom = "rect",
+        xmin = as.Date("2020-01-01"), xmax = as.Date("2021-01-01"), ymin = -Inf, ymax = Inf,
+        fill = cols[3], alpha = 0.5, col = cols[3]
+    ) +
+    geom_line(aes(x = date, y = rlval), size = 0.15) +
+    facet_wrap(~climvar, scales = "free_y", labeller = varnames, strip.position = "left", nrow = 4) +
+    # annotate("rect",
+    #     fill = "indianred2", alpha = 0.5,
+    #     xmin = 2010, xmax = 2011, ymin = -Inf, ymax = Inf
+    # ) +
+    theme_bw() +
+    ylab("") +
+    xlab("Date") +
+    # show all years on x-axis
+    scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
+    theme(
[…]

### 2025-06-18 scripts/06_figures.R
SI figures until Fig 3

@@ -1302,7 +1302,7 @@ dec_intercept_plot
 new_preds <- readRDS("results/models/non_negative/new_preds_isoclines_twi.RDS")
 
 new_preds_df <- do.call(rbind, new_preds)
-coefs_df <- do.call(rbind, coefs)
+# coefs_df <- do.call(rbind, coefs)
 
 iso_plot_twi <- ggplot(
     new_preds_df,
@@ -1415,11 +1415,11 @@ dec_intercept_plot <- ggplot(ranef_df_long, aes(x = value, y = intercept)) +
     geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.1) +
     geom_hline(yintercept = 0, linetype = "dashed") +
     # facet_grid(name ~ factor(yr, levels = c(2010, 2015)), scales = "free_x") +
-    facet_wrap(name ~ factor(yr, levels = c(2010, 2015)),
-        scales = "free", ncol = 2, strip.position = "left"
+    facet_wrap(name ~ factor(yr, levels = c(2010, 2015, 2020)),
+        scales = "free", ncol = 3, strip.position = "left"
     ) +
     # add text with p value
-    geom_text(data = coefs_dec_lms %>% filter(term == "value"), aes(x = 3, y = 1, label = paste("p = ", round(p.value, 2)), hjust = 0, vjust = 0)) +
+    geom_text(data = coefs_dec_lms %>% filter(term == "value"), aes(x = 1, y = 0.5, label = paste("p = ", round(p.value, 2)), hjust = 0, vjust = 0)) +
     labs(x = "species trait value", y = "intercept") +
     guides(color = "none") +
     theme_bw() +
@@ -1431,7 +1431,7 @@ dec_intercept_plot <- ggplot(ranef_df_long, aes(x = value, y = intercept)) +
 
 
 library(patchwork)
-png("doc/display/Fig_trait_intercept_plot.png", width = 6, height = 6, units = "in", res = 300)
+png("doc/display/Fig_trait_intercept_plot.png", width = 8, height = 6, units = "in", res = 300)
 # sensplot + sp_intercept_plot + plot_annotation(tag_levels = "a") + plot_layout(widths = c(1.8, 1))
 dec_intercept_plot
 dev.off()
@@ -1439,7 +1439,7 @@ dev.off()
 # SI figure - map of dendroband trees ---------------
 
 dendro_tree_map_sp <- ggplot() +
[…]

### 2025-06-18 doc/supplementary_information.Rmd
updated SI figures and SI file

@@
 
-![**Figure S1:** Raw climate data timeseries from 2009 to 2020 from the Forest Fire Research Station near HKK.
-Measurements represent daily rolling means over a 30-day time window for precipitation, max temperature and VPD and rolling sum for number of dry days.](display/climvars_fullseries.png)
+![**Figure S1:** Raw climate data timeseries from 2009 to 2023 from ERA5Land and CHIRPS.
+Measurements represent daily rolling means over a 30-day time window for precipitation, max temperature and VPD and rolling sum for number of dry days.](display/climvars_fullseries_remote.png)
 
-![**Figure S2:** Climate anomalies for all variables from 2009 to 2020from the Forest Fire Research Station near HKK.
-Values represent number of standard deviations from the long-term mean for daily rolling mean values.](display/climvars_anomalies_fullseries.png)
+![**Figure S2:** Climate anomalies for all variables from 2009 to 2020.
+Values represent number of standard deviations from the long-term mean for daily rolling mean values.](display/climvars_anomalies_fullseries_remote.png)
 

### 2025-06-18 scripts/06_figures.R

@@ -1197,22 +1197,44 @@ rval_2_3 <- cor.test(forcor$sens_2015, forcor$sens_2020, use = "pairwise.complet
 # vals[4]
 
 # plot the correlation
-corplot <- ggplot(forcor, aes(x = sens_2010, y = sens_2015)) +
+corplot_1 <- ggplot(forcor, aes(x = sens_2010, y = sens_2015)) +
     geom_point(alpha = 0.5) +
     geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
     xlab("Sensitivity in 2010") +
     ylab("Sensitivity in 2015") +
     # add text with r value
     geom_text(
-        data = data.frame(x = -3, y = 3, label = paste0("R = ", round(rval, 2), "\np = ", round(pval, 3))),
+        data = data.frame(x = -3, y = 3, label = paste0("R = ", round(rval_1_2, 2), "\np = ", round(pval_1_2, 3))),
         aes(x = x, y = y, label = label), hjust = 0.5, vjust = 0.5
     ) +
     theme_bw()
 
-corplot
+corplot_2 <- ggplot(forcor, aes(x = sens_2010, y = sens_2020)) +
+    geom_point(alpha = 0.5) +
+    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+    xlab("Sensitivity in 2010") +
+    ylab("Sensitivity in 2020") +
+    # add text with r value
+    geom_text(
+        data = data.frame(x = -3, y = 3, label = paste0("R = ", round(rval_1_3, 2), "\np = ", round(pval_1_3, 3))),
+        aes(x = x, y = y, label = label), hjust = 0.5, vjust = 0.5
+    ) +
+    theme_bw()
 
-png("doc/display/sens_corr.png", width = 4, height = 4, units = "in", res = 300)
-corplot
+corplot_3 <- ggplot(forcor, aes(x = sens_2015, y = sens_2020)) +
+    geom_point(alpha = 0.5) +
+    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+    xlab("Sensitivity in 2015") +
+    ylab("Sensitivity in 2020") +
+    # add text with r value
+    geom_text(
+        data = data.frame(x = -3, y = 3, label = paste0("R = ", round(rval_2_3, 2), "\np = ", round(pval_2_3, 3))),
[…]

### 2025-06-18 doc/manuscript_hkk_drought_sensitivity.Rmd
text snipping
current count - 5106

@@
 **Number of Words in Abstract:** 150  
-**Number of Words in Main Text:** 5200  
-**Number of References:** 46  
+**Number of Words in Main Text:** 5106  
+**Number of References:** 54  
 **Number of Figures:** 5  
@@
 For forests worldwide, separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought.
-Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., a species with low sensitivity to past droughts would have low sensitivity in any microhabitat and through any drought.
-In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in a different drought. 
+Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., a species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought.
+In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove more vulnerable in a different drought. 
 The complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].
@@
 [Then go into some details with citations on ?] -->
-Drought, or hot/dry conditions anomalously departing from climatic means during any time of year, is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the severity and duration of the climatic anomalies [@slette_how_2019].
-Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different timing, duration, or intensity may have very different effects on tree growth, especially in seasonal forests that have evolved phenological strategies to handle regular periods of climatic stress [@albert_cryptic_2019] [@gao_dynamic_2018].
+Drought is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the timing, severity and duration of the climatic anomalies [@slette_how_2019].
+<!-- ; we define it here as hot/dry conditions anomalously departing from climatic means during any time of year  -->
+Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different timing, duration, or intensity may have distinct effects on tree growth, especially in seasonal forests that have evolved phenological strategies to handle regular periods of climatic stress [@albert_cryptic_2019] [@gao_dynamic_2018].
 <!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
@@
 Analyses to date have found that tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates.
-It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those occurring at different times relative to regular seasonal cycles<!-- of wet and dry seasons-->.
+It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those that intensify or conflict with regular seasonal cycles<!-- of wet and dry seasons-->.
 <!--
@@
 Tree species adaptations to cope with drought vary widely in type and strength, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
-In tropical seasonal forests in particular, trees have evolved to maximise growth, reproduction, and survival under periodic hot/dry conditions, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
-Dry season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss and have acquisitive strategies during leaf-on periods [@desouza_drought_2020].
+In tropical seasonal forests in particular, trees have evolved to maximise growth, reproduction, and survival under dry seasons, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
+Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss and have acquisitive strategies during leaf-on periods [@desouza_drought_2020].
 Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and *large vessels (hydraulic efficiency)*<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021].
@@
 Our study site was the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
-This is a 50-hectare plot in dry seasonal evergreen forest, with a mean annual precipitation of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
+This is a 50-hectare plot in dry seasonal evergreen forest, with mean annual precipitation of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
 The region is periodically affected by droughts associated with the *El Niño* Southern Oscillation [ENSO, @räsänen_spatial_2016].
 The dominant canopy species are _Hopea odorata_, _Vatica harmandiana_, _Dipterocarpus alatus_ and _Tetrameles nudiflora_ and lower layers of the canopy are dominated by sub-canopy evergreen and deciduous species like _Miliusa horsfieldii_ and _Polyalthia viridis_.
-For analysis, we focused on 30 of the most abundant species in the plot.   
+<!-- For analysis, we focused on 30 of the most abundant species in the plot.    -->
 <!-- Analysis of long-term tree-ring data shows that interannual variation tree growth at HKK is influenced by temperature and rainfall [@vlam_temperature_2014].   -->
@@
 
-Using the late wet/early dry dendrometer band censuses, we calculated annualised growth for each year for each tree. 
+Using the late wet/early dry dendrometer band censuses, we calculated annualised growth for each year and tree. 
 We first conducted quality control on the dataset, removing measurements that appeared likely to be data entry errors (misplaced decimals), potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses).
-We used the raw window size measurements combined with the diameter measurement at installation $(d_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[@condit_dendrometer_].
-We used a direct measurement of $d_{1}$ at the time of installation if it was available.
-If the tape measurement at the first window size measurement was not available for any band, we excluded window size measurements until the first available diameter measurement. 
-We then calculated annualised diameter increments for each individual $i$ for each year $t$ by finding the difference in diameter from the previous year ($inc_{i,t} = d_{i,t} - d_{i,t-1}$) and adjusting for the number of days ($tdif$) between measurements ($ainc_{i, t} = inc_{i,t} * \frac{tdif}{365}$).
+We used window size measurements combined with the diameter measurement at installation $(d_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[@condit_dendrometer_].
+We used a direct measurement of $d_{1}$ at the time of installation if it was available; if not, we excluded window size measurements until the first available diameter measurement. 
+We calculated annualised diameter increments for each individual $i$ for each year $t$ by finding the difference in diameter from the previous year ($inc_{i,t} = d_{i,t} - d_{i,t-1}$) and adjusting for the number of days ($tdif$) between measurements ($ainc_{i, t} = inc_{i,t} * \frac{tdif}{365}$).
 
-To ensure that the data did not smooth over multiple years, we excluded increment measures spanning missing values for $d_{t-1}$.
-We also excluded large measurement outliers (> 3 standard deviations from the mean increment across all observations). 
-Diameter tape measurements themselves are not precise enough for calculating drought sensitivity; error for each tape measure is 1 mm, error for growth is at least 2 mm. 
+<!-- To ensure that the data did not smooth over multiple years, we excluded increment measures spanning missing values for $d_{t-1}$. -->
+We excluded large measurement outliers (> 3 standard deviations from the mean increment across all observations). 
+Diameter tape measurements themselves are not precise enough for calculating drought sensitivity; error for each measurement is 1 mm, error for growth is 2 mm. 
 In comparison, median annual growth across trees in HKK is ~1.4 mm and calculating deviation from annual growth needs sub-mm precision. 
-However, tape measurements to flag likely errors in the dendrometer dataset. 
-We calculated increments from annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement with these annualised increments calculated from tape measurements.
+However, we used tape measurements to flag likely errors in the dendrometer dataset. 
+We calculated increments from annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement. 
 Finally, we excluded trees with negative increments over the whole timeseries, and included only species that had at least 10 individuals, resulting in a final dataset of 1820 individuals of 30 species.
@@
 Among these, 2010 and 2015 were El Niño years; 2010 was a moderate El Niño event while 2015 was a very strong El Niño event (Oceanic Niño Index @bamston_documentation_1997). 
-The drought in 2020 was not associated with an El Niño event.
+<!-- The drought in 2020 was not associated with an El Niño event. -->
 <!-- We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected.
@@
 To characterise the three droughts, we used daily weather data obtained from publicly available data sources. 
-We used daily weather data obtained from remotely sensed data source CHIRPS [@funk_climate_2015] for daily total precipitation and ERA5Land [@muñoz-sabater_era5land_2021] to calculate daily maximum temperature and vapour pressure deficit (VPD, @raesch_pvldcurve_2020).
-We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 30 day periods across the whole timeseries.
+We used data from CHIRPS [@funk_climate_2015] for daily total precipitation and ERA5Land [@muñoz-sabater_era5land_2021] to calculate daily maximum temperature and vapour pressure deficit (VPD, @raesch_pvldcurve_2020).
+We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 30-day periods across the whole timeseries.
 Finally, we calculated daily rolling anomalies for each variable for the drought years from the long-term daily expectation as $$anomaly_{day} = \frac{rolling.mean_{day} - \bar{rolling.mean_{day}}}{sd(rolling.mean_{day})}$$
@@
 
-We first  calculated the mean annual increment $\bar{a.inc_{i}}$ for each tree from annualised increment $ainc_{i,t}$<!-- as $\frac{\[ 
-  \sum_{t}^{}{ainc_{i,t}} 
-  \]}{\sum{t}}$-->.
+We first  calculated the mean annual increment $\bar{a.inc_{i}}$ for each tree from annualised increment $ainc_{i,t}$.
 We then calculated sensitivity for each tree $i$ for each drought year $y$ as 
@@
 $$
-By scaling the difference in increment to the mean annual growth of each tree, this metric allows us to compare across slow and fast growing individuals and species.
+By scaling the increment deviation to the mean annual growth of each tree, this metric allows us to compare across slow and fast growing individuals and species.
 
@@
 
-Across all individuals in the plot, we calculated Topographic Wetness Index (TWI) for an expectation of the water availability across space based on the total upslope area and the slope at the location.
-We used a Digital Elevation Model from a 5 m x 5 m elevation layer collected at plot establishment.
-We used the *whitebox* package in R [@wu_whitebox_2022; @lindsay_whitebox_2016a] to fill and breach this layer, calculate flow accumulation at each cell and then calculate TWI as $log_{e}\frac{As}{tan(slope)}$.
-As an alternate measure of water availability, we also calculated Topographic Position Index (TPI) based on the concavity/convexity of a location. 
+Across all individuals, we calculated Topographic Wetness Index (TWI) for an expectation of the water availability across space based on the total upslope area and the slope at the location. 
+We used a Digital Elevation Model from a 5 m x 5 m elevation layer collected at plot establishment and the *whitebox* package in R [@wu_whitebox_2022; @lindsay_whitebox_2016a] to calculate flow accumulation at each cell and then calculate TWI as $log_{e}\frac{As}{tan(slope)}$.
+<!-- As an alternate measure of water availability, we also calculated Topographic Position Index (TPI) based on the concavity/convexity of a location.  -->
+For an alternate estimate of wetness (Topographic Position Index) and associated results, see supplementary material. 
 We defined tree size as the diameter at breast height (calculated from dendrometer band window measurement) at the previous census. 
 We estimated crown exposure for each tree at each census using crown illumination index (CII), an ordinal categorical variable from 1 to 5 (least to most exposed) measured at each dendroband census. 
+We centred and scaled TWI and DBH across the range of observed values.
 
@@
 
-We analysed data for each drought year by fitting separate models. 
+We analysed tree responses to individual droughts by fitting separate models. 
 We fit alternative models for sensitivity including a null model with just species random effects, a model including only fixed effects of species deciduousness and habitat wetness and a model including size (DBH), light environment (CII) and habitat wetness as well as species random effects on all these.
@@
 
-The three droughts analysed had distinct climatic and temporal characteristics, with 2010 primarily a dry season drought, 2015 a wet-season drought and 2020 showing mixed characteristics (Table 1, Fig `r fig_droughts`).
-The 2010 drought peaked between February and May with meteorological anomalies on the 30-day scale including lower precipitation (min anomaly **<-1.5** SD), higher number of dry days (max anomaly **>1.5** SD), higher maximum temperature (max anomaly +0.8 SD), and higher VPD (max anomaly **>2** SD).
+The droughts analysed had distinct climatic and temporal characteristics, with 2010 primarily a dry-season drought, 2015 a wet-season drought and 2020 showing mixed characteristics (Table 1, Fig `r fig_droughts`).
+The 2010 drought showed 30-day-mean meteorological anomalies between February and May including lower precipitation (min anomaly **<-1** SD), higher number of dry days (anomaly **~1.5** SD), higher maximum temperature (anomaly **>2** SD), and higher VPD (anomaly **>2** SD), including in the wet season.
 <!-- Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions. -->
 1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought. 
-In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including precipitation (min anomaly **<-1.5** SD), number of dry days (max anomaly **>1.5** SD), maximum monthly mean temperatures (max anomaly +1.5 SD), and monthly maximum VPD (max anomaly +2 SD).
-1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period. 
+In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including precipitation (anomaly **<-1.5** SD), number of dry days (max anomaly **>2** SD), maximum monthly temperatures (anomaly **+1.5** SD), and monthly maximum VPD (max anomaly **+2** SD).
+<!-- 1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period.  -->
 <!-- Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season. -->
-1- to 6-month SPEI values reached minima <2, indicative of extreme drought, while the 12-month values remained just above -2, indicative of severe drought. 
-In 2020, meteorological anomalied were pronounced both in the dry and wet seasons, including maximum temperature (max anomaly **> 2** SD) and VPD (max anomaly **> 2** SD). 
-1- to 6-month SPEI values reached minima ~2 in February, indicative of borderline severe drought.
-12-month SPEI showed severe to moderate drought conditions from January to August, spanning both dry and wet seasons.
+1- to 12-month SPEI values crossed minima -2, lowest in the dendroband census period and indicative of severe drought. 
+In 2020, meteorological anomalies were pronounced both in the dry and wet seasons, including maximum temperature (**>2** SD) and VPD (**>2** SD). 
+3- and 6-month SPEI values reached minima ~2 in February, indicative of borderline severe drought.
+12-month SPEI showed severe drought conditions from January to April, and moderate drought from May to September, spanning both dry and wet seasons.
 
@@
 <!-- The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a).  -->
-Drought years were associated with lower plot-level median growth and greater tree-level sensitivities, with the strongest response in 2015 (Fig `r fig_drought_responses`). 
-Median annual growth rate across all years of measurement for the whole plot was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual*10, na.rm=T),2)` mm. 
-The drought year 2015 had the lowest median annual increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
-In the 2010 and 2020 droughts, the median annual increments were `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm and `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm respectively, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% and `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
+Drought years were associated with lower plot-level growth and stronger tree-level sensitivities, with the strongest negative response in 2015 (Fig `r fig_drought_responses`). 
+Median plot-level annual growth rate across years of measurement was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual*10, na.rm=T),2)` mm. 
+The drought year 2015 had the lowest median increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
+In the 2010 and 2020 droughts, the median increments were `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm and `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm respectively, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% and `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
 <!-- In the 2020 drought, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
@@
 
-Species median sensitivities to the three droughts were not correlated with each other although on average, species had more negative sensitivities to the 2015 drought than the 2010 or 2020 droughts, in alignment with median sensitivities across all trees (for 2010 and 2015 Pearson's correlation coefficient r = `r round(sens.sp.cor_10_15[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_15[3]$p.value,2)`, for 2015 and 2020 was r = `r round(sens.sp.cor_15_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_15_20[3]$p.value,2)`, and for  2010 and 2020 r = `r round(sens.sp.cor_10_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_20[3]$p.value,2)`, Fig S5).
+Species median sensitivities to the three droughts were not correlated with each other, although on average, species had more negative sensitivities to the 2015 drought than the 2010 or 2020 droughts, in alignment with median sensitivities across all trees (for 2010 and 2015 Pearson's correlation coefficient r = `r round(sens.sp.cor_10_15[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_15[3]$p.value,2)`, for 2015 and 2020 was r = `r round(sens.sp.cor_15_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_15_20[3]$p.value,2)`, and for  2010 and 2020 r = `r round(sens.sp.cor_10_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_20[3]$p.value,2)`, Fig S5).
 
@@
 <!-- Deciduousness species responded more negatively than evergreen species in 2015, but not in 2010 (Fig `r fig_species`a).  -->
-Predicted sensitivities from intercept-only models with a species random effect showed heterogenous sensitivities of deciduous and evergreen species to different droughts.
+Deciduous and evergreen species showed heterogenous responses to different droughts.
 Deciduous species were more drought sensitive than evergreen species in 2015 (Fig `r fig_species`a, r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`), but the opposite was true in 2020 (`r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2020], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2020], 2)`).
@@
 
-Crown exposure was associated with more negative sensitivities in all three droughts.
+Crown exposure was associated with greater growth declines in all three droughts.
 In the same models with species random effects on intercept and slope, the simplex parameter for the ordered factor crown exposure in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 and 2020 the negative effect was not as strong (2015 median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`; 2020 median effect = `r round(coefs_spre_cii$median[coefs_spre_cii$yr==2020],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2020],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2020],2)`).
@@
 We show that, in one tropical seasonal forest, the drought sensitivity of tree growth is jointly and interactively shaped by drought characteristics, species adaptations, and local environmental conditions (Table 1).
-Analysing two droughts in a 14-year dendrometer band time series, we found lower-than-average growth in drought years, resulting in lower plot-level growth (Fig `r fig_drought_responses`). 
+Analysing three droughts in a 14-year dendrometer band time series, we found lower-than-average growth in drought years, resulting in lower plot-level growth (Fig `r fig_drought_responses`). 
 Across droughts, there was little consistency in species- or individual-level drought responses, reflecting dissimilar effects of a key species trait (deciduousness) and variables linked to water access (TWI, DBH) under differing drought conditions.
 Despite lower absolute extremes of temperature, vapor pressure deficit, and number of dry days (Fig.
-2a), drought in the wet season (2015) elicited stronger responses across trees, especially deciduous species, which are expected to have stronger seasonality in growth phenology (Fig `r fig_species`).
-Interestingly, in 2015 (but not 2010), evergreen and deciduous species had divergent responses along a wetness gradient; evergreen trees fared better and deciduous species suffered worse declines in wetter sites (Fig `r fig_species`b, Fig `r fig_microenvs`b).
+2a), severe drought in the wet season (2015) elicited stronger responses across trees, especially deciduous species, which are expected to have stronger seasonality in growth phenology (Fig `r fig_species`).
+Interestingly, in 2015 and 2020 (but not 2010), evergreen and deciduous species had divergent responses, especially along a wetness gradient; deciduous species suffered worse declines in wetter sites in 2015 and evergreen species suffered in drier sites in 2020 (Fig `r fig_species`b, Fig `r fig_microenvs`b).
 Although trees with more exposed crowns were consistently more sensitive to drought, additional effects of DBH -- presumably including rooting volume and depth -- were dissimilar between the drought events (Fig `r fig_microenvs`a).
@@
 <!----Limitations in filling gap/ comparing only two droughts---->
-While our analysis of 30 species across size classes and habitats spans representative spatial and ecological variation realistically and beyond the scope of previous such analyses, we are limited in quantitatively analysing differences associated with drought characteristics because of low sample size.
+While our analysis of 30 species across size classes and habitats spans spatial and ecological variation realistically and beyond the scope of previous analyses, we are limited in quantitatively analysing differences associated with drought characteristics because of limited sample size.
 <!-- Long-timescale analyses with tree rings are limited by the fact that species with rings are few, represent specific strategy to water use and growth - and often affiliated to drier microenvironments.
@@
 However, given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts.-->
-The three droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, magnitude and duration of meteorological extremes and anomalies, and severity as classified the SPEI metric (Fig.
+The three droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, magnitude and duration of meteorological extremes and anomalies, and severity as classified with SPEI (Fig.
 `r fig_droughts`).
@@
 Tree ring studies from this site [@vlam_temperature_2014;@anderson-teixeira_joint_2022] and across the tropics [@zuidema_tropical_2022] have asserted that tree growth is most sensitive to dry-season precipitation.
-But  given that the 2015 drought was more anomalous on multi-month time scales (Fig.
-`r fig_droughts`) does not align with this assumption.
-Generalization of tropical tree growth responses to any specific drought seems to require fine-scale monitoring of growth (i.e., dendrometer band measurements or dendrochronology when possible) from species that might span importance and drought strategies over different timings and intensities of droughts.
-Advancing efforts to collect this level of data is critical to our assessment and prediction of drought responses of tropical forets to changing precipitation patterns.
+But given that the 2015 drought was more anomalous on multi-month time scales (Fig. `r fig_droughts`) does not align with this assumption.
+Generalization of tropical tree growth responses to any specific drought seems to require fine-scale monitoring of growth (i.e., dendrometer band measurements or dendrochronology when possible) from species that might span drought strategies over different timings and intensities of droughts.
+Advancing efforts to collect these data is critical to our assessment and prediction of drought responses of tropical forets to changing precipitation patterns.
 <!-- [is this necessary?]  Studies in temperate forests, where extensive tree-ring records are available, have confirmed that drought characteristics including seasonal timing, duration, and severity affect tree growth responses [@gao_dynamic_2018; @dorangeville_drought_2018] and that effects of species, species traits, tree size, and microenvironment can differ across droughts [@mcgregor_tree_2021].
@@
 <!----Limitations in generalisation/ species examined here may have greater diversity of drought adaptaitons than elsewhere in the tropics---->
+Periodic hot/dry conditions in tropical dry forests could create selective pressures for adaptative variation towards drought resistance potnetially resulting in a greater diversity of drought responses here compared to other bioclimatic zones where adaptive pressures and thus ecological variability are different.
 We have demonstrated that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.
-Tropical dry forests represent periodic hot/dry conditions, creating selective pressures for adaptative variation towards drought resistance, and this might mean that species in tropical dry forests exhibit a greater diversity of drought responess than species in other bioclimatic zones where adaptive pressures and thus ecological variability are different.
 Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies and has resulted in diversification along the hydraulic safety-efficiency axis with diverse strategies to maximise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021].
@@
 Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different traits respond to different types of drought, and how these responses may be modified by microenvironment.
-Again, continued expansion of dendrometer band records and ecophysiological study of tropical tree drought responses will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.
+Again, continued expansion of dendrometer band records and ecophysiological studies will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.
 <!--Therefore, we caution that our results suggestive of community-level insurance effects under drought be interpreted carefully across forest types that may be structured primarily by nutrient limitation, fires, herbivory or pathogens.
@@
 <!-- One variable examined here -- crown exposure -- had a consistent effect across the two droughts (Table 1, Figs. `r fig_microenvs`, `r fig_dag`).  -->
-Trees with more exposed crowns exhibited greater growth declines in both droughts, while sub-canopy trees grew more than average during the 2010 drought.
+Trees with more exposed crowns exhibited greater growth declines, while sub-canopy trees grew more than average during the 2010 drought.
 This finding is consistent with the concept that larger trees tend to suffer more during drought [@bennett_larger_2015], but adds the important insight that it is crown position -- as opposed to size *per se* -- that is primarily responsible for this pattern.
@@
 `r fig_microenvs`, `r fig_dag`).
-Together, these findings imply that while canopy and emergent trees face the greatest stress and exhibit the most negative growth responses in almost any drought, a suite of species traits and microenvironmental characteristics modify growth responses in interaction with drought characteristics.
+Together, these findings imply that while canopy and emergent trees face the greatest stress and exhibit the most negative growth responses, a suite of species traits and microenvironmental characteristics modify growth responses in interaction with drought characteristics.
 
 <!----Contributions beyond---->
-Despite consistent observations of growth declines at the plot level across drought events, we show that this scale obscures a variety of ecological responses at the species and tree level that are heterogeneous across droughts.
+Despite consistent observations of plot-level growth declines across drought events, we show that this scale obscures a variety of species- and tree-level ecological responses that are heterogeneous across droughts.
 An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings (n=4 at HKK; @vlam_temperature_2014) are unlikely to be representative of the drought responses of the entire community.
-For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the two events; it's growth nearly doubled in the dry-season drought but was severely reduced in the wet-season drought (Fig.
+For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the three events; it's growth nearly doubled in the 2010 and 2020 droughts but was severely reduced in the wet-season drought (Fig.
 S5).
@@
 Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in three different droughts, showing that ecological and evolutionary strategies make particular species and trees resistant to some droughts but vulnerable to others.
-These distinct responses of species and individuals to drought characteristics suggests potential spatiotemporal insurance effects among tropical tree response to drought, in alignment with recent global analyses (@liu_drought_2022).
+These distinct responses of species and individuals to droughts suggests potential spatiotemporal insurance effects among tropical tree response, in alignment with recent global analyses (@liu_drought_2022).
 
@@
 Our study reveals that drought sensitivity is not a static trait associated to particular species, traits, or microenvironments, but rather a dynamic response shaped interactively by these factors and drought characteristics.
-Despite net decreased growth at the plot level (in agreement with landscape-scale in @aguirre-gutiérrez_functional_2022 and @bennett_sensitivity_2023), we show evidence that the same strategies are not uniformly successfully across spatiotemporal climatic variation.
+Despite net decreased growth at the plot level (in agreement with landscape-scale in @aguirre-gutiérrez_functional_2022 and @bennett_sensitivity_2023), we show evidence that the same strategies are not uniformly successful across spatiotemporal climatic variation.
 Static traits are often used to explain and predict forest responses to drought, especially in the tropics where long-term records are few [@guillemot_small_2022; @vico_how_2017a]. 

### 2025-06-18 doc/manuscript_hkk_drought_sensitivity.Rmd
words shaved to 5011

@@
 ^3^Smithsonian Tropical Research Institute, Panama  
-^4^4Thai Long-Term Forest Ecological Research, Kasetsart University, Bangkok, Thailand  
-^5^5Department of Forest Biology, Faculty of Forestry, Kasetsart University, Bangkok 10900, Thailand
+^4^Thai Long-Term Forest Ecological Research, Kasetsart University, Bangkok, Thailand  
+^5^Department of Forest Biology, Faculty of Forestry, Kasetsart University, Bangkok 10900, Thailand
 
@@
 **Number of Words in Abstract:** 150  
-**Number of Words in Main Text:** 5106  
+**Number of Words in Main Text:** 5011  
 **Number of References:** 54  
@@
 Intense droughts are becoming more frequent, affecting the demography, diversity, and  carbon (C) cycling of forests globally [@ipcc_climate_2023]. 
-However, our ability to predict drought impacts on tree woody growth and ecosystem C sequestration remains limited, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022].
-For forests worldwide, separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought.
-Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., a species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought.
-In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove more vulnerable in a different drought. 
+However, drought impacts on tree woody growth and ecosystem C sequestration remain difficult to predict, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022].
+For forests worldwide, studies have separately shown that drought characteristics, species traits, and microenvironment affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter understanding on forest responses to drought.
+Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought.
+In contrast, if these factors interact, species and individuals that exhibit resistance to one drought may prove more vulnerable in a different drought. 
 The complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].
@@
 [Then go into some details with citations on ?] -->
-Drought is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the timing, severity and duration of the climatic anomalies [@slette_how_2019].
+Drought is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the timing, severity and duration of climatic anomalies [@slette_how_2019] with potentially distinct ecological consequences.
 <!-- ; we define it here as hot/dry conditions anomalously departing from climatic means during any time of year  -->
-Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different timing, duration, or intensity may have distinct effects on tree growth, especially in seasonal forests that have evolved phenological strategies to handle regular periods of climatic stress [@albert_cryptic_2019] [@gao_dynamic_2018].
+Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different timing, duration, or intensity may have distinct effects on tree growth, especially in seasonal forests with adaptive phenological strategies to handle regular periods of climatic stress [@albert_cryptic_2019] [@gao_dynamic_2018].
 <!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
@@
 <!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
-Analyses to date have found that tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates.
+Analyses to date have found that tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent a very limited set of tropical tree species and climates.
 It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those that intensify or conflict with regular seasonal cycles<!-- of wet and dry seasons-->.
@@
 <!-- Sean::: mention rooting depth? [@chitra-tarak: root of the drought and chitra-tarak: new_phyt 2022] -->
-Tree species adaptations to cope with drought vary widely in type and strength, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
+Tree species adaptations to drought vary widely in type and strength, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
 In tropical seasonal forests in particular, trees have evolved to maximise growth, reproduction, and survival under dry seasons, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
 Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss and have acquisitive strategies during leaf-on periods [@desouza_drought_2020].
-Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and *large vessels (hydraulic efficiency)*<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021].
+Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and large vessels (hydraulic efficiency)<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021].
 Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021].
@@
 <!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
-Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls.
+Moreover, besides being adaptively conserved, these strategies can also have developmental or ecophysiological controls.
 For example, some species regulate negative $\pi_{tlp}$'s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; @hulshof_variation_2010].
@@
 -->
-Microenvironments defined by horizontal topographic and edaphic variation, and vertical light and thermal variation further shape individual tree experience of and response to drought. 
-While habitats with relatively moist soils (e.g., low-lying topographic positions) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted (e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021) or acclimatized [@refs].
+Microenvironments defined by horizontal topographic and edaphic variation, and vertical light and thermal variation further shape individual tree experience of, and response to, drought. 
+While habitats with relatively moist soils (e.g., low-lying topography) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted (e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021).
 Indeed, during drought trees near streams experienced greater growth declines in a temperate forest [@mcgregor_tree_2021] and increased mortality in a tropical forest [@zuleta_droughtinduced_2017]. 
@@
 <!--In contrast, understory trees may be rarely exposed to direct solar radiation and extreme evaporative demand.-->
-These aboveground gradients in crown exposure are at least partially offset by the tendency for large trees to have larger and often deeper root systems and greater stem water storage [@fernández-de-uña_role_2023] and therefore the potential to access more reliable water sources [@stahl_depth_2013].
+These aboveground gradients in crown exposure could be partially offset by the tendency for large trees to have larger/deeper root systems and greater stem water storage [@fernández-de-uña_role_2023] and therefore the potential to access more reliable water sources [@stahl_depth_2013].
 <!-- Sean::: I need to look at this source.
@@
 We defined tree size as the diameter at breast height (calculated from dendrometer band window measurement) at the previous census. 
-We estimated crown exposure for each tree at each census using crown illumination index (CII), an ordinal categorical variable from 1 to 5 (least to most exposed) measured at each dendroband census. 
 We centred and scaled TWI and DBH across the range of observed values. 
+We estimated crown exposure for each tree and year using ground measurements of crown illumination index (CII), an ordinal variable from 1 to 5 (least to most exposed).
 
-We assigned deciduousness as a species level trait using data published in @williams_deciduousness_2008 for the top 30 species. 
-We defined species deciduousness as the mean over trees of the proportion of crown leaf loss at maximum, with a score ranging from 0 to 4 where 0 is most evergreen and 4 is most deciduous.
+We assigned deciduousness as a species-level trait using data from @williams_deciduousness_2008. 
+We defined deciduousness as the mean over trees of the proportion of crown leaf loss at maximum, with a score ranging from 0 to 4 where 0 is most evergreen and 4 is most deciduous. 
 Data were unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1. 
-We also considered using the duration of time spent at maximum crown loss as an alternative measure of deciduousness.
+We also considered an alternative measure of deciduousness - the duration of time spent at maximum crown loss.
 This was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for more species.  
@@
 
-Median annual increment varied among species with *`r tree.time.sp$spfull[1]`* having the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm) among the `r nrow(tree.time.sp)` species analysed.
-While median species responses to the 2010 and 2020 droughts were variable, all species had lower median increment than a typical year in 2015. 
+Median annual increment varied among species with *`r tree.time.sp$spfull[1]`* the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm) among the `r nrow(tree.time.sp)` species analysed.
+Median species responses to the 2010 and 2020 droughts were variable, but all species had lower median increments than a typical year in 2015. 
 In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/3]`*.
-In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/3]`*.
-In 2020, sensitivities ranged from `r sens.sp[sens.sp$yr==2020,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2020,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[nrow(sens.sp)/3]`*.
+In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/3]`* and in 2020 from `r sens.sp[sens.sp$yr==2020,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2020,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[nrow(sens.sp)/3]`*.
 
@@
 
-The test of whether individual trees responded similarly to the three drought events showed no correlation (Fig S6).
+Individual trees did not respond similarly to the three drought events (Fig S6).
 
@@
 
-The differing effects of TWI on sensitivity in the three droughts, along with the influence of deciduousness, were confirmed by models that controlled for crown exposure and DBH as well as species variation in intercept and slopes (Fig `r fig_microenvs`a).
-These models showed small negative effect of wetness in 2010 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but a positive effect of wetness in 2015 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`) and 2020 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2020],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2020],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2020],2)`).
+The differing effects of TWI on sensitivity in the three droughts, along with the influence of deciduousness, were confirmed by models that controlled for crown exposure, DBH and species variation (Fig `r fig_microenvs`a).
+Wetness had a small negative effect in 2010 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but a positive effect in 2015 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`) and 2020 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2020],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2020],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2020],2)`).
 <!-- The positive effect of TWI in 2015 was, however, mainly due to the response of evergreen species; deciduousness had a significant negative correlation with the slopes of the TWI effect in 2015 (Fig `r fig_microenvs`b).
 -->
-Predictions across the range of observed variables also showed that deciduous species generally had larger negative sensitivities in 2015, which did not  increase with TWI (Fig `r fig_microenvs`c).  
-Mean sensitivities across species and their response to TWI were not associated with deciduousness in 2010 or 2020, mirroring the response in the interaction model.
+Predictions across the range of observed variables also showed that deciduous species experienced greater growth declines in 2015 across the wetness gradient (Fig `r fig_microenvs`c). 
+<!-- Mean sensitivities across species and their response to TWI were not associated with deciduousness in 2010 or 2020, mirroring the response in the interaction model.   -->
 
@@
 
-Crown exposure was associated with greater growth declines in all three droughts.
+Crown exposure was associated with greater growth declines in 2010.
 In the same models with species random effects on intercept and slope, the simplex parameter for the ordered factor crown exposure in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 and 2020 the negative effect was not as strong (2015 median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`; 2020 median effect = `r round(coefs_spre_cii$median[coefs_spre_cii$yr==2020],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2020],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2020],2)`).
-Predicted effects of crown exposure on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all crown exposure categories had negative sensitivities in 2015 (Fig `r fig_dag`b).
+Predicted effects of crown exposure on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all crown exposure categories had negative sensitivities in 2015 and 2020 (Fig `r fig_dag`b).
 
@@
 Across droughts, there was little consistency in species- or individual-level drought responses, reflecting dissimilar effects of a key species trait (deciduousness) and variables linked to water access (TWI, DBH) under differing drought conditions.
-Despite lower absolute extremes of temperature, vapor pressure deficit, and number of dry days (Fig.
+Despite lower extremes of temperature, vapor pressure deficit, and number of dry days (Fig.
 2a), severe drought in the wet season (2015) elicited stronger responses across trees, especially deciduous species, which are expected to have stronger seasonality in growth phenology (Fig `r fig_species`).
@@
 Although trees with more exposed crowns were consistently more sensitive to drought, additional effects of DBH -- presumably including rooting volume and depth -- were dissimilar between the drought events (Fig `r fig_microenvs`a).
-It is possible that water availability buffered exposure-associated growth declines through different mechanisms in the two different droughts; larger trees fared better in the dry-season drought while trees in wetter sites suffered smaller growth declines in the wet-season drought (Fig `r fig_dag`).
+It is possible that water availability buffered exposure-associated growth declines through different mechanisms in the different droughts; larger trees fared better in the dry-season drought while trees in wetter sites suffered smaller growth declines during wet season droughts (Fig `r fig_dag`).
 <!-- we can talk about it ... but I don't think that we can know that larger trees have deeper roots.  -->
 Taken together, these results indicate that joint and interactive effects of drought characteristics, species traits, and individual microenvironments produce important heterogeneity in individual and species drought responses of tropical trees.
-This finding that individual droughts have divergent effects on species and individuals may contribute to the maintenance of species diveristy and ecosystem stability.
+This finding that individual droughts have divergent effects on species and individuals may contribute to the maintenance of species diversity and ecosystem stability.
 
 <!----Limitations in filling gap/ comparing only two droughts---->
-While our analysis of 30 species across size classes and habitats spans spatial and ecological variation realistically and beyond the scope of previous analyses, we are limited in quantitatively analysing differences associated with drought characteristics because of limited sample size.
+Our analysis of 30 species across size classes and habitats spans realistic spatial and ecological variation beyond the scope of previous analyses, but we are limited in disentangling the effects of drought characteristics because of limited sample size.
 <!-- Long-timescale analyses with tree rings are limited by the fact that species with rings are few, represent specific strategy to water use and growth - and often affiliated to drier microenvironments.
@@
 However, given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts.-->
-The three droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, magnitude and duration of meteorological extremes and anomalies, and severity as classified with SPEI (Fig.
+The three droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, magnitude and duration of meteorological anomalies, and severity as classified with SPEI (Fig.
 `r fig_droughts`).
@@
 Tree ring studies from this site [@vlam_temperature_2014;@anderson-teixeira_joint_2022] and across the tropics [@zuidema_tropical_2022] have asserted that tree growth is most sensitive to dry-season precipitation.
-But given that the 2015 drought was more anomalous on multi-month time scales (Fig. `r fig_droughts`) does not align with this assumption.
-Generalization of tropical tree growth responses to any specific drought seems to require fine-scale monitoring of growth (i.e., dendrometer band measurements or dendrochronology when possible) from species that might span drought strategies over different timings and intensities of droughts.
-Advancing efforts to collect these data is critical to our assessment and prediction of drought responses of tropical forets to changing precipitation patterns.
+But the strength of the ecological response to the 2015 wet-season drought (Fig. `r fig_droughts`) does not align with this assumption.
+Generalization of tropical tree growth responses to any specific drought might require fine-scale monitoring of growth (i.e., dendrometer band measurements or dendrochronology when possible) on species that span strategies and over different timings and intensities of droughts.
+Advancing efforts to collect these data is critical to our assessment and prediction of tropical forest responses to changing precipitation patterns.
 <!-- [is this necessary?]  Studies in temperate forests, where extensive tree-ring records are available, have confirmed that drought characteristics including seasonal timing, duration, and severity affect tree growth responses [@gao_dynamic_2018; @dorangeville_drought_2018] and that effects of species, species traits, tree size, and microenvironment can differ across droughts [@mcgregor_tree_2021].
@@
 Periodic hot/dry conditions in tropical dry forests could create selective pressures for adaptative variation towards drought resistance potnetially resulting in a greater diversity of drought responses here compared to other bioclimatic zones where adaptive pressures and thus ecological variability are different.
-We have demonstrated that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.
-Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies and has resulted in diversification along the hydraulic safety-efficiency axis with diverse strategies to maximise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021].
-The degree of adaptive variation along the water use axis might be comparable to other dry climatic systems (e.g.
-temperate drylands reported in @vasey_intraspecific_2022), but it may not apply to the wet or everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different.
-Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different traits respond to different types of drought, and how these responses may be modified by microenvironment.
+We demonstrate that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.
+Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies resulting in diversification along the hydraulic safety-efficiency axis with diverse strategies to maximise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021].
+The degree of adaptive variation along the water-use axis might be comparable to other dry climatic systems (e.g.
+temperate drylands reported in @vasey_intraspecific_2022), but may be distinct from the wet/everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different.
+Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different traits respond to different types of drought, and how these responses may be modified by microenvironments.
 Again, continued expansion of dendrometer band records and ecophysiological studies will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.
@@
 S5).
-More generally, tree-ring records are likely to give a biased picture of drought responses, as ring formation is caused by dry-season dormancy (often associated with deciduous leaf habit) and sampling often targets large trees with exposed crowns and dry microenvironments [@speer_fundamentals_2010]-- all factors shown here to <!--directly and interactively -->influence drought sensitivity.<!-- More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects.
+More generally, tree-ring records likely present a biased picture of drought responses, as ring formation is caused by dry-season dormancy (often associated with deciduous leaf habit) and sampling often targets large trees with exposed crowns and dry microenvironments [@speer_fundamentals_2010]-- all factors shown here to <!--directly and interactively -->influence drought sensitivity.<!-- More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects.
 -->

### 2025-06-18 doc/supplementary_information.Rmd
updates to supplementary figures and appendix

@@
 
-![**Figure S10:** Testing correlations of variables across all individuals](display/explore/cond_dep_alltrees.png)
+![**Figure S10:** Testing correlations of variables across all individuals](display/cond_dep_alltrees.png)
 
@@
 
+## Visualising sensitivity with predictors
+
+![](display/sens_var_rels.png)
+
+## Alternate models using Topographic Position Index
+
+As an alternate measure of wetness to Topographic Wetness Index (TWI) that requires the total upslope area, we used Topographic Position Index (TPI), a localised convexity/concavity-based metric of water availability as a predictor.
+We calculated TPI using the package *spatialEco* [@evans_spatialeco_2023] using a circular buffer window of 1, 3, 5 and 7 pixels. 
+Larger window sizes while providing a smoother surface on the plot, clip out larger portions of the plot margin where the window might bleed outside. 
+We chose to use the 5-pixel version because of the balance between smoothness across the landscape and data loss.
+
+TPI and TWI had better resolution in different parts of the wetness gradient, with TWI capturing larger resolution among wetter locations and TPI capturing larger resolution among drier locations. 
+Interaction models of deciduousness with TPI showed qualitatively similar results with the TWI models across the three drought years.
+
+![](display/FigSI_TPI_preds.png)
+
+Similarly, models accounting for DBH and CII showed similar directional effects of wetness whether TWI or TPI was used. 
+Please note that left to right represents wet to dry, given the way that TPI is calculated.
+
+![](display/Fig4_tpi_3yrs.png)
+
 <!-- ## Individual model diagnostics
@@
 To add: zero growth models -->
+
+# References {-}

### 2025-06-18 scripts/06_figures.R

@@ -1237,6 +1237,98 @@ png("doc/display/sens_corr.png", width = 12, height = 4, units = "in", res = 300
 corplot_1 + corplot_2 + corplot_3 + plot_layout(ncol = 3)
 dev.off()
 
+# conditional dependencies-------------------------
+
+cond_dep_all <- tree.time %>%
+    # filter(Cno == 15) %>%
+    filter(Cno %in% c(5, 15, 25)) %>%
+    group_by(Cno) %>%
+    pivot_longer(cols = c("calcDBH_min1", "cii_min1"), names_to = "var", values_to = "value") %>%
+    dplyr::mutate(varnames = ifelse(var == "calcDBH_min1", "DBH", "CII"))
+
+# these variables are conditionally independent for the most part
+
+# plot the conditional independencies
+
+library(ggpubr)
+cond_dep_all_plot <- ggscatter(
+    data = cond_dep_all,
+    x = "value", y = "twi",
+    add = "reg.line", conf.int = TRUE, alpha = 0.3,
+    cor.coef = TRUE, cor.method = "spearman",
+    xlab = "variable", ylab = "TWI"
+) +
+    facet_wrap(varnames ~ yr, scales = "free")
+
+cond_dep_all_plot
+
+png("doc/display/cond_dep_alltrees.png", width = 8, height = 8, units = "in", res = 300)
+cond_dep_all_plot
+dev.off()
+
+# raw distribution plots--------------------------
+
+# plot of sensitivity against CII
+sens_cii <- ggplot(
+    tree.time %>% filter(yr %in% c(2010, 2015, 2020)),
+    aes(x = factor(cii_min1), y = sens.prop)
+) +
+    geom_boxplot() +
+    geom_jitter(alpha = 0.3) +
+    # geom_smooth(method = "lm", col = "grey40") +
+    facet_wrap(~yr) +
+    xlab("CII") +
+    ylab("Sensitivity") +
+    theme_bw() +
+    theme(
+        strip.background = element_blank(),
+        strip.placement = "outside",
+        strip.text = element_text(size = 12)
+    )
+
+sens_cii
+
+sens_dbh <- ggplot(
+    tree.time %>% filter(yr %in% c(2010, 2015, 2020)),
+    aes(x = calcDBH_min1, y = sens.prop)
+) +
+    geom_point(alpha = 0.3) +
+    geom_smooth(method = "lm", col = "grey40") +
+    facet_wrap(~yr) +
[…]

### 2025-06-18 doc/manuscript_hkk_drought_sensitivity.Rmd
keywords and data accessibility

@@
 # Keywords {-}
+drought response, tree growth, tropical forest, microenvironment buffering, deciduousness
 
@@
 # Data Accessibility Statement {-}
+This manuscript does not use novel code. 
+The dendrometer band data used for this manuscript is archived in the ForestGEO data repository and will be made available upon request. 
+The climate data used in this work are publicly available. 
+The code used to analyse the data will be made available in Zenodo upon acceptance of the manuscript.
 

### 2025-06-19 doc/manuscript_hkk_drought_sensitivity.Rmd
some more comments from HML

@@
 <!-- ; we define it here as hot/dry conditions anomalously departing from climatic means during any time of year  -->
-Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different timing, duration, or intensity may have distinct effects on tree growth, especially in seasonal forests with adaptive phenological strategies to handle regular periods of climatic stress [@albert_cryptic_2019] [@gao_dynamic_2018].
+Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different seasonal timing, duration, or intensity may have distinct effects on tree growth, especially in seasonal forests with diverse adaptive strategies to handle regular dry periods [@albert_cryptic_2019;@gao_dynamic_2018].
 <!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
@@
 <!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
-Analyses to date have found that tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent a very limited set of tropical tree species and climates.
+Analyses to date have found that interannual variation in tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent a very limited set of tropical tree species and climates.
 It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those that intensify or conflict with regular seasonal cycles<!-- of wet and dry seasons-->.
@@
 In tropical seasonal forests in particular, trees have evolved to maximise growth, reproduction, and survival under dry seasons, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
-Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss and have acquisitive strategies during leaf-on periods [@desouza_drought_2020].
+Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss [@desouza_drought_2020]. 
 Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and large vessels (hydraulic efficiency)<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021].
@@
 Our study site was the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
-This is a 50-hectare plot in dry seasonal evergreen forest, with mean annual precipitation of 1400 mm and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
+This is a 50-hectare plot in dry seasonal evergreen forest, with mean annual precipitation of 1400 mm, mean annual temperature of 23.5 C and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
 The region is periodically affected by droughts associated with the *El Niño* Southern Oscillation [ENSO, @räsänen_spatial_2016].
@@
 Using the late wet/early dry dendrometer band censuses, we calculated annualised growth for each year and tree. 
-We first conducted quality control on the dataset, removing measurements that appeared likely to be data entry errors (misplaced decimals), potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses).
-We used window size measurements combined with the diameter measurement at installation $(d_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[@condit_dendrometer_].
+<!-- We first conducted quality control on the dataset, removing measurements that appeared likely to be data entry errors (misplaced decimals), potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses).
+-->
+We first conducted quality control on window size measurements (see supplementary methods) and combined them with the diameter measurement at installation $(d_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[@detto_correcting_2023].
 We used a direct measurement of $d_{1}$ at the time of installation if it was available; if not, we excluded window size measurements until the first available diameter measurement. 
@@
 <!-- To ensure that the data did not smooth over multiple years, we excluded increment measures spanning missing values for $d_{t-1}$. -->
-We excluded large measurement outliers (> 3 standard deviations from the mean increment across all observations). 
-Diameter tape measurements themselves are not precise enough for calculating drought sensitivity; error for each measurement is 1 mm, error for growth is 2 mm. 
+<!-- We excluded large measurement outliers (> 3 standard deviations from the mean increment across all observations).  -->
+We conducted further quality control on increments (see supplementary methods for details) to assemble the final dataset. 
+Diameter tape measurements themselves are not precise enough for calculating drought sensitivity; error for growth is twice measurement error = 2 mm. 
 In comparison, median annual growth across trees in HKK is ~1.4 mm and calculating deviation from annual growth needs sub-mm precision. 
 However, we used tape measurements to flag likely errors in the dendrometer dataset. 
-We calculated increments from annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement. 
+<!-- We calculated increments from annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement.  -->
 Finally, we excluded trees with negative increments over the whole timeseries, and included only species that had at least 10 individuals, resulting in a final dataset of 1820 individuals of 30 species.
@@
 Across all individuals, we calculated Topographic Wetness Index (TWI) for an expectation of the water availability across space based on the total upslope area and the slope at the location. 
-We used a Digital Elevation Model from a 5 m x 5 m elevation layer collected at plot establishment and the *whitebox* package in R [@wu_whitebox_2022; @lindsay_whitebox_2016a] to calculate flow accumulation at each cell and then calculate TWI as $log_{e}\frac{As}{tan(slope)}$.
+We used a Digital Elevation Model at 5 m spatial grain for the area of the plot for fine resolution and ability to locate the trees on the layer, but this omits upslope area that is off the plot. 
+Using the *whitebox* package in R [@wu_whitebox_2022; @lindsay_whitebox_2016a] we calculated flow accumulation at each cell and then calculate TWI as $log_{e}\frac{As}{tan(slope)}$ where $As$ is the total upslope area within the plot and $slope$ is the absolute value of slope calculated at the 5m grain.
 <!-- As an alternate measure of water availability, we also calculated Topographic Position Index (TPI) based on the concavity/convexity of a location.  -->
@@
 We defined deciduousness as the mean over trees of the proportion of crown leaf loss at maximum, with a score ranging from 0 to 4 where 0 is most evergreen and 4 is most deciduous. 
+For each species, these were calculated from observations of individual trees that were assigned categorical scores, and averaged, allowing species to take continuous, decimal values in the 0-4 range.
 Data were unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1. 
@@
 
-We then used the predicted sensitivities from this model and tested their correlations with deciduousness using simple linear models.
+We then used the species effects (intercept + random effect) from this model and tested their correlations with deciduousness using simple linear models.
 
@@
 
-We then calculated predictions across all combinations of values from observed maximum to minimum of TWI and deciduousness within this dataset.
+<!-- We then calculated predictions across all combinations of values from observed maximum to minimum of TWI and deciduousness within this dataset. -->
 
-To further disentange the microenvironmental drivers of drought sensitivity, we simultaneously modelled:
+To further disentange the microenvironmental drivers of drought sensitivity, we used a Bayesian causal model to simultaneously estimate fits for:
 
 $$
-sensitivity \sim DBH + CII + TWI + (DBH+CII+TWI | Species)
+sensitivity \sim 1 + DBH + CII + TWI + (1+DBH+CII+TWI | Species)
 $$ 

### 2025-06-19 doc/supplementary_information.Rmd
some more comments from HML

@@
 
-- Data cleaning steps
-- Calculating DBH equation
-- Corroboration with tape measurements
-- error distribution
+<!-- - Data cleaning steps -->
+For raw measurements of dendrometer band window size, we conducted QAQC steps as follows:
+1.
+Removed potentially misidentified individuals.
+We removed individuals with conflicting metadata on tag or location across censuses.
+We did not remove individuals with conflicting species identification because these are often updated during the ForestGEO censuses.
+Therefore, we used the latest version of species identification for each tag shared by the PIs of the HKK plot.
+2.
+Removed potential misidentified bands.
+Each new dendroband installed on a tree is numbered sequentially starting from 1.
+We removed any measurements made on bands old bands after a new band series had begun.
+3.
+Removing measurements that appeared likely to be data entry errors.
+We identified misplaced decimals by checking if the ratio between adjacent window size measurements were closer to 1 or 10 or 100.
+We removed measurements that had ratios closer to 10 or 100 than 1, assuming that these were likely misplaced decimals.
+
+After calculating DBH from window size measurements and calculating annualised growth increments, we conducted further QAQC on these increments to create the final dataset:
+1. We excluded large measurement outliers, defined as > 3 standard deviations from the mean increment across all observations. 
+2.
+We excluded trees with negative increments or increments close to zero over the whole timeseries of observations.
+These low growing trees may be indicative of stress or mortality, and likely to bias analysis of interannual growth variation.
+3.
+We used concurrent tape measurements on dendrobanded trees to flag likely errors.
+We calculated annualised increments from annual tape measurements made on each dendrobanded tree at each census, using similar methods as described for dendrometer bands.
+For each tree and year, we calculated the degree of deviation of these measurements from each other as the Euclidean distance from the 1:1 line.
+We found that 90% of the increments were within 5 mm of of deviation, which we retained as a high confidence dataset.
+We excluded the remaining 10% from the analysis, although we were unable to ascertain whether the discrepancy was because of errors in tape or dendroband measurements.
 

### 2025-06-19 doc/manuscript_hkk_drought_sensitivity.Rmd
updated Fig 3 and 4 with axis direction

@@
 **Number of Words in Abstract:** 150  
-**Number of Words in Main Text:** 5011  
+**Number of Words in Main Text:** 5016  
+<!-- Words to subtract: 2628-->
 **Number of References:** 54  
@@
 S5).
-More generally, tree-ring records likely present a biased picture of drought responses, as ring formation is caused by dry-season dormancy (often associated with deciduous leaf habit) and sampling often targets large trees with exposed crowns and dry microenvironments [@speer_fundamentals_2010]-- all factors shown here to <!--directly and interactively -->influence drought sensitivity.<!-- More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects.
+More generally, tree-ring records likely present a biased picture of drought responses, as ring formation is caused by dry-season dormancy (often associated with deciduous leaf habit) and sampling often targets large trees with exposed crowns and dry microenvironments [@speer_fundamentals_2010]-- all factors shown here to <!--directly and interactively -->influence drought sensitivity.
+<!-- More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects.
 -->

### 2025-06-19 scripts/06_figures.R

@@ -1400,7 +1400,7 @@ dec_intercept_plot <- ggplot(ranef_df, aes(x = williams_dec, y = intercept, ymin
             label = paste("r = ", round(estimate, 2), "\np = ", round(p.value, 2)), hjust = 0, vjust = 0
         )
     ) +
-    labs(x = "Deciduousness", y = "Predicted sensitivity") +
+    labs(x = "Deciduousness", y = "Species sensitivity") +
     guides(color = "none") +
     theme_bw() +
     theme(
@@ -1423,7 +1423,11 @@ iso_plot_twi <- ggplot(
     aes(x = williams_dec, y = twi, fill = Estimate)
 ) +
     geom_tile() +
-    labs(x = "Deciduousness", y = "Topographic Wetness Index", fill = "Sensitivity") +
+    labs(
+        x = "Deciduousness",
+        fill = "Sensitivity"
+    ) +
+    ylab("Topographic Wetness Index\n drier \u2194 wetter") +
     # geom_contour(aes(z = Estimate), colour = "black") +
     # facet_grid(yr ~ Species) +
     facet_wrap(~yr) +
@@ -1434,14 +1438,18 @@ iso_plot_twi <- ggplot(
         strip.placement = "outside",
         strip.text = element_text(size = 12)
     )
-# iso_plot
+
+iso_plot_twi
 
 
 library(patchwork)
 # png("doc/display/Fig3.png", width = 8, height = 8, units = "in", res = 300)
-png("doc/display/Fig3_3yrs.png", width = 8, height = 8, units = "in", res = 300)
+png("doc/display/Fig3.png", width = 8, height = 8, units = "in", res = 300)
 dec_intercept_plot / iso_plot_twi + plot_annotation(tag_levels = "a") + plot_layout(heights = c(1.8, 1))
 dev.off()
+
+# SI figure with points and deciduousness--------------------
+
 
 
 # Figure 3 panel with TPI -----------------------------------------
@@ -1802,7 +1810,7 @@ pred_plot <- ggplot(data = preds_df, aes(x = twi_scaled, y = median)) +
         # scales = "free",
         ncol = 3
     ) +
-    labs(x = "Topographic Wetness Index", y = "Predicted sensitivity") +
+    labs(x = "Scaled Topographic Wetness Index\n drier \u2194 wetter", y = "Predicted sensitivity") +
     guides(color = guide_legend(title = "Deciduousness")) +
     theme_bw() +
     theme(
[…]

### 2025-06-19 scripts/06_figures.R
plot of residuals added to SI

@@ -1450,7 +1450,57 @@ dev.off()
 
 # SI figure with points and deciduousness--------------------
 
+# preds <- readRDS("results/models/non_negative/pred_isoclines_twi.RDS")
+# pred_df <- do.call(rbind, preds)
 
+# fits <- readRDS("results/models/non_negative/fits_isoclines_twi.RDS")
+# library(brms)
+
+# res1 <- residuals(fits[[1]])
+# res2 <- residuals(fits[[2]])
+# res3 <- residuals(fits[[3]])
+
+# head(res1)
+# head(pred_df)
+
+pred_df <- readRDS("results/models/non_negative/predictions_intercept.RData")
+str(pred_df)
+
+fits <- readRDS("results/models/non_negative/fits_interceptonly.RDS")
+res1 <- residuals(fits[[1]])
+res2 <- residuals(fits[[2]])
+res3 <- residuals(fits[[3]])
+
+residuals <- rbind(res1, res2, res3)
+
+# bind residuals to preds
+pred_res <- cbind(pred_df, residuals)
+
+# plot residuals with variables
+
+# first make long df
+pred_res_long <- pred_res %>%
+    pivot_longer(cols = c(cii_min1, calcDBH_min1_scaled, twi_scaled, williams_dec)) %>%
+    select(Tag, treeID, yr, sens.prop, median, lwr, upr, Estimate, Est.Error, Q2.5, Q97.5, name, value) %>%
+    dplyr::mutate(name = ifelse(name == "calcDBH_min1_scaled", "scaled DBH",
+        ifelse(name == "cii_min1", "CII",
+            ifelse(name == "twi_scaled", "scaled TWI", "deciduousness")
+        )
+    ))
+
+# plot these
+
+
+plot_res <- ggplot(pred_res_long, aes(x = value, y = Estimate)) +
+    geom_point(alpha = 0.3) +
+    geom_smooth() +
+    ylab("Residual") +
+    facet_grid(name ~ yr) +
+    theme_bw()
+
+png("doc/display/residuals.png", width = 6, height = 8, units = "in", res = 300)
+plot_res
+dev.off()
 
 # Figure 3 panel with TPI -----------------------------------------
 

### 2025-06-19 doc/supplementary_information.Rmd
plot of residuals added to SI
updated SI file with fig captions

@@
 
+## Raw sensitivity and residuals with predictors
+
+![**Figure S10:** Distribution of raw sensitivity values across years and their relationship with DBH, CII and TWI. Lines are linear model fits.](display/sens_var_rels.png)
+
+![**Figure S11:** Distribution of residuals of species intercept models of sensitivity and their relationship with scaled DBH, TWI and CII and deciduousness.
+Species intercept model results are reported in main text.](display/residuals.png)
+
+\newpage
+
 ## Conditional dependencies
@@
 
-![**Figure S10:** Testing correlations of variables across all individuals](display/cond_dep_alltrees.png)
+![**Figure S12:** Testing correlations of variables across all individuals](display/cond_dep_alltrees.png)
 
@@
 
-![**Figure S11:** Testing correlations for DBH $\perp\!\!\!\perp$ TWI by species](display/explore/cond_dep_dbh_twi.png)
+![**Figure S13:** Testing correlations for DBH $\perp\!\!\!\perp$ TWI by species](display/explore/cond_dep_dbh_twi.png)
 
-![**Figure S12:** Testing correlations for CII $\perp\!\!\!\perp$ TWI by species](display/explore/cond_dep_cii_twi.png)
+![**Figure S14:** Testing correlations for CII $\perp\!\!\!\perp$ TWI by species](display/explore/cond_dep_cii_twi.png)
 
@@
 \newpage
-
-## Visualising sensitivity with predictors
-
-![](display/sens_var_rels.png)
 
@@
 
-![](display/FigSI_TPI_preds.png)
+![**Figure S15:** Correlation of modelled species sensitivities with Topographic Position Index](display/FigSI_TPI_preds.png)
 
@@
 
-![](display/Fig4_tpi_3yrs.png)
+![**Figure S16:** Model results from Bayesian causal model of sensitivity as a function of DBH, CII and TPI.](display/Fig4_tpi_3yrs.png)
 

### 2025-06-19 doc/manuscript_hkk_drought_sensitivity.Rmd
add fig legends to bottom

@@
 Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in three different droughts, showing that ecological and evolutionary strategies make particular species and trees resistant to some droughts but vulnerable to others.
-These distinct responses of species and individuals to droughts suggests potential spatiotemporal insurance effects among tropical tree response, in alignment with recent global analyses (@liu_drought_2022).
+These distinct responses of species and individuals to droughts suggests potential spatiotemporal insurance effects among tropical tree response, in alignment with recent global analyses [@liu_drought_2022].
 
@@
 \newpage
+
+# Figure Legends {-}
+
+***Figure `r fig_droughts`.
+Climatic characteristics of the focal drought years (red in 2010, brown in 2015) compared with the average across years*** 2008-2019 (grey dashed) and its standard deviation (grey shading) in the Huai Kha Khaeng ForestGEO plot, for four daily climate variables: dry days, precipitation, maximum temperature and VPD.
+(a) 30-day running means.
+(b) Anomalies of the drought year running means from the averages, in units of standard deviations.
+(c) Standardised Precipitation Evapotranspiration Index (SPEI) values for the drought years at four different time scales for each month for each drought year.
+A k-month SPEI integrates over k months ending in the month current month.
+Across all panels, light blue background represents the average wet season, from May to October.
+
+***Figure `r fig_drought_responses`: Growth increment timeseries and drought responses*** across trees and species in the Huai Kha Khaeng ForestGEO plot.
+a) Median of annualised growth increments across ten dominant species (colours) and all trees (black) from 2009 to 2022.
+b) Distribution of drought sensitivity of tree growth across all individuals in the two drought years (see main text for equation).
+
+***Figure `r fig_species`: Variation in sensitivity among species*** in the Huai Kha Khaeng ForestGEO plot during droughts in 2010, 2015 and 2020.
+a) Relationship of the mean species sensitivities from model fits of an intercept-only model, with mean species deciduousness values.
+Line shown for significant correlation.
+b) Predicted sensitivity values across the observed range of deciduousness and topographic wetness index from a model with TWI, deciduosness and their interaction across all trees modelled for each drought year separately.
+
+***Figure `r fig_microenvs`: Drivers of variation in sensitivity among individuals***.
+All panels represent model results from models (one for each year) predicting sensitivity with microenvironment conditions with species random effect on all slopes.
+a) Coefficient plots showing median effects and 90% CI for wetness, exposure and DBH of tree.
+b) Predicted relationship of drought sensitivity with topographic wetness index across all species derived from model predictions.
+Black line represents overall relationship (panel a); coloured lines for each species reflects deciduousness values.
+
+***Figure `r fig_dag`: Effects of size, exposure and water availability*** from combined models.
+Hypothesised relationships between microenvironment variables and drought sensitivity with their modelled effects in the HKK ForestGEO plot in the droughts of a) 2010, b) 2015 and c) 2020.
+Solid lines represent relationships where 90% credible intervals do not overlap 0, blue and red lines represent positive and negative effects respectively and line thickness is scaled to the effect size.
+d) Modelled sensitivity across crown exposure classes in 2010, 2015 and 2020.
+
+\newpage
+
 # References  {-}

### 2025-06-19 scripts/06_figures.R
better fig 5 and alternate Fig 2

@@ -949,8 +949,53 @@ sens.all <- ggplot(data = tree.time %>% filter(yr %in% yrs), aes(x = sens.prop))
         strip.text = element_text(size = 12)
     )
 
-sens.all
+cols <- viridis::viridis(4, option = "F")
 
+sens.all.2 <- ggplot(data = tree.time %>% filter(yr %in% yrs), aes(x = sens.prop)) +
+    geom_density(alpha = 0.5, aes(col = factor(yr), fill = factor(yr))) +
+    scale_color_manual(values = cols) +
+    scale_fill_manual(values = cols) +
+    geom_vline(xintercept = c(-1, 0, 1), linetype = "dashed") +
+    # facet_wrap(~yr) +
+    facet_wrap(~ factor(yr), ncol = 3) +
+    # xlim(-5, 5)+
+    labs(x = "Drought sensitivity", y = "Density") +
+    theme_bw() +
+    guides(color = "none", fill = "none") +
+    # make the strip title bold
+    theme(
+        strip.background = element_blank(),
+        strip.placement = "outside",
+        strip.text = element_text(size = 12)
+    )
+
+
+sens.all.2
+
+sens.all.3 <- ggplot(data = tree.time %>% filter(yr %in% yrs), aes(x = sens.prop)) +
+    geom_density(alpha = 0.5, aes(col = factor(yr), fill = factor(yr))) +
+    scale_color_manual(values = cols) +
+    scale_fill_manual(values = cols) +
+    geom_vline(xintercept = c(-1, 0, 1), linetype = "dashed") +
+    # facet_wrap(~factor(yr), ncol = 1) +
+    # xlim(-5, 5)+
+    labs(x = "Drought sensitivity", y = "Density") +
+    theme_bw() +
+    # guides(color="none", fill="none")+
+    # make the strip title bold
+    theme(
+        strip.background = element_blank(),
+        strip.placement = "outside",
+        strip.text = element_text(size = 12),
+        legend.position = c(0.75, 0.75),
+        legend.background = element_rect(fill = alpha("white", 0.7), color = NA)
+    )
+
+sens.all.3
+
+png("doc/display/sens_all_1panel.png", width = 4, height = 4, units = "in", res = 300)
+sens.all.3
+dev.off()
 
 # get the top 10 species for 2015
 
@@ -1028,6 +1073,82 @@ spagplot_top10 <- ggplot() +
 
 spagplot_top10
 
[…]

### 2025-06-24 doc/manuscript_hkk_drought_sensitivity.Rmd
version sent to coauthors for final look

@@
 
-**Title:** Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments
+**Title:** Drought response is not a species trait: tropical tree drought sensitivity is shaped by drought characteristics, species adaptations, and individual microenvironments
 
@@
 Krishna Anujan^1,2^*  ORCiD: `r team$ORCID[1]`  
-Sean McMahon^2,3^  ORCiD: `r team$ORCID[2]`  
+Sean M. McMahon^2,3^  ORCiD: `r team$ORCID[2]`  
 Sarayudh Bunyavejchewin^4^  ORCiD: `r team$ORCID[3]`  
@@
 Nantachai Pongpattananurak^4,5^  ORCiD: `r team$ORCID[6]`  
-Kristina Anderson-Teixeira^1,3^  ORCiD: `r team$ORCID[7]`
+Kristina J. Anderson-Teixeira^1,3^  ORCiD: `r team$ORCID[7]`
 
@@
 # Statement of Authorship {-}
-KA, SM, SJD and KAT conceptualised the analysis. 
+KA, SMM, SJD and KAT conceptualised the analysis. 
 SB, SJD and NP have been leading long-term data collection at HKK, while HMM created and helped execute the dendrometer band protocol. 
-KA cleaned the data and conducted formal analysis with support from SM, SJD, HMM and KAT. 
-KA wrote the original draft of the manuscript with support from KAT and SM. 
+KA cleaned the data and conducted formal analysis with support from SMM, SJD, HMM and KAT. 
+KA wrote the original draft of the manuscript with support from KAT and SMM. 
 All authors contributed to revision, review and editing of the manuscript.
@@
 However, drought impacts on tree woody growth and ecosystem C sequestration remain difficult to predict, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022].
-For forests worldwide, studies have separately shown that drought characteristics, species traits, and microenvironment affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter understanding on forest responses to drought.
+For forests worldwide, studies have separately shown that drought characteristics, species traits, and microenvironment affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter understanding of forest responses to drought.
 Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought.
-In contrast, if these factors interact, species and individuals that exhibit resistance to one drought may prove more vulnerable in a different drought. 
-The complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].
+In contrast, if these factors interact, species and individuals that exhibit resistance to one drought may prove more vulnerable to a drought with different characteristics. 
+The complexity and diversity of drought responses may act as a mechanism for maintaining species diversity and buffering ecosystem function even as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].
 
@@
 <!-- Sean::: mention rooting depth? [@chitra-tarak: root of the drought and chitra-tarak: new_phyt 2022] -->
-Tree species adaptations to drought vary widely in type and strength, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
-In tropical seasonal forests in particular, trees have evolved to maximise growth, reproduction, and survival under dry seasons, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
+Tree species adaptations to drought vary widely, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
+In tropical seasonal forests in particular, trees have evolved to optimize growth, reproduction, and survival (i.e., fitness) under dry seasons, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
 Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss [@desouza_drought_2020]. 
@@
 We tested these hypotheses using  a 14-year record of dendrometer band measurements in a dry seasonal evergreen forest [@bunyavejchewin_seasonally_2011] in Thailand. 
-With data on 1820 trees across 30 species we built Bayesian regression models in a causal modelling framework to account for potential confounding relationships among the variables. 
+During this period, the forest experienced three different dry periods, two associated with ENSO events, and all three differing with respect to timing and duration (Table 1). 
+With data on 1820 trees across 30 species we built Bayesian regression models exploring how species characteristics and individual spatial context influenced growth responses to these droughts. 
+We used a causal modelling framework to account for potential confounding relationships among the variables. 
 This analysis adds important insight into the complexity of seasonal dry forest response to drought in species-rich tropical forests.
@@
 )
-ft <- hline(ft, i = c(1, 4, 7), part = "body")
+ft <- hline(ft, i = c(1, 4, 8), part = "body")
 ft <- line_spacing(ft, space = 1, part = "all")
@@
 # ft <- height(ft, height = 0.3)
-ft <- bold(ft, i = c(1, 4, 7), bold = TRUE, part = "body")
+ft <- bold(ft, i = c(1, 4, 8), bold = TRUE, part = "body")
 ft <- width(ft, 1, width = 3.5)
@@
 b) Distribution of drought sensitivity of tree growth across all individuals in the two drought years (see main text for equation).
-For comments - https://github.com/forestgeo/growth-precip-thailand/issues/11](display/Fig2.png)
+For comments - https://github.com/forestgeo/growth-precip-thailand/issues/11](display/Fig2_alt.png)
 
@@
 <!---CONCLUSIONS---->
-We show that, in one tropical seasonal forest, the drought sensitivity of tree growth is jointly and interactively shaped by drought characteristics, species adaptations, and local environmental conditions (Table 1).
+We show that, in a tropical seasonal forest, the drought sensitivity of tree growth is jointly and interactively shaped by drought characteristics, species adaptations, and local environmental conditions (Table 1).
 Analysing three droughts in a 14-year dendrometer band time series, we found lower-than-average growth in drought years, resulting in lower plot-level growth (Fig `r fig_drought_responses`). 
@@
 2a), severe drought in the wet season (2015) elicited stronger responses across trees, especially deciduous species, which are expected to have stronger seasonality in growth phenology (Fig `r fig_species`).
-Interestingly, in 2015 and 2020 (but not 2010), evergreen and deciduous species had divergent responses, especially along a wetness gradient; deciduous species suffered worse declines in wetter sites in 2015 and evergreen species suffered in drier sites in 2020 (Fig `r fig_species`b, Fig `r fig_microenvs`b).
+Interestingly, in 2015 and 2020 (but not 2010), evergreen and deciduous species had divergent responses, especially along a wetness gradient; deciduous species showed greater growth declines in wetter sites in 2015 while evergreen species grew less in drier sites in 2020 (Fig `r fig_species`b, Fig `r fig_microenvs`b).
 Although trees with more exposed crowns were consistently more sensitive to drought, additional effects of DBH -- presumably including rooting volume and depth -- were dissimilar between the drought events (Fig `r fig_microenvs`a).
@@
 However, given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts.-->
-The three droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, magnitude and duration of meteorological anomalies, and severity as classified with SPEI (Fig.
+The three droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, magnitude and duration of meteorological anomalies, and severity as classified with SPEI (Table 1, Fig.
 `r fig_droughts`).
@@
 But the strength of the ecological response to the 2015 wet-season drought (Fig. `r fig_droughts`) does not align with this assumption.
-Generalization of tropical tree growth responses to any specific drought might require fine-scale monitoring of growth (i.e., dendrometer band measurements or dendrochronology when possible) on species that span strategies and over different timings and intensities of droughts.
+Generalization of tropical tree growth responses to any specific drought likely requires fine-scale monitoring of growth (i.e., dendrometer band measurements or dendrochronology when possible) on species that span strategies and over different timings and intensities of droughts.
 Advancing efforts to collect these data is critical to our assessment and prediction of tropical forest responses to changing precipitation patterns.
@@
 José Medina-Vega, Valentine Herrmann and Eugenie Más provided input on improving analyses. 
-KA was supported by the Smithsonian Institution Postdoctoral Fellowship and the ForestGEO Analytical Workshop, Asia 2024 (supported by NSF grant DEB-[CHECK] to SJD) in completing this work.
+KA was supported by the Smithsonian Institution Postdoctoral Fellowship and the ForestGEO Analytical Workshop, Asia 2024 (supported by NSF grant DEB-2020424 to SJD) in completing this work.
 

### 2025-06-24 scripts/06_figures.R

@@ -1021,7 +1021,7 @@ spagplot_top10 <- ggplot() +
         aes(
             x = yr, y = median_inc,
             group = spname, col = spname
-        )
+        ), alpha = 0.7
     ) +
     # mean of all trees
     geom_line(
@@ -1081,7 +1081,7 @@ spagplot_top10_names <- ggplot() +
         aes(
             x = yr, y = median_inc,
             group = spname, col = spname
-        )
+        ), alpha = 0.7
     ) +
     # mean of all trees
     geom_line(
@@ -2401,6 +2401,11 @@ sens.sp_plot_1 <- ggplot(sens.sp.wide, aes(x = sens.2010, y = sens.2015, group =
     geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
     # add species names to points with the name starting at the point
     geom_text(aes(label = spfull, col = williams_dec), hjust = 0, vjust = 0) +
+    # add correlation values
+    geom_text(aes(
+        label = paste0("r = ", round(sens.sp.cor_10_15$estimate, 2), "\np = ", round(sens.sp.cor_10_15$p.value, 2)),
+        x = -0.7, y = 0
+    )) +
     xlim(c(-1, 1)) +
     guides(color = guide_legend(title = "Deciduousness")) +
     labs(x = "2010 sensitivity", y = "2015 sensitivity") +
@@ -2413,7 +2418,12 @@ sens_sp_plot_2 <- ggplot(sens.sp.wide, aes(x = sens.2015, y = sens.2020, group =
     geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
     # add species names to points with the name starting at the point
     geom_text(aes(label = spfull, col = williams_dec), hjust = 0, vjust = 0) +
-    xlim(c(-1, 1)) +
+    # add correlation values
+    geom_text(aes(
+        label = paste0("r = ", round(sens.sp.cor_15_20$estimate, 2), "\np = ", round(sens.sp.cor_15_20$p.value, 2)),
+        x = -0.7, y = 0.5
+    )) +
+    xlim(c(-1, 0.7)) +
     guides(color = guide_legend(title = "Deciduousness")) +
     labs(x = "2015 sensitivity", y = "2020 sensitivity") +
     theme_bw()
@@ -2425,7 +2435,12 @@ sens_sp_plot_3 <- ggplot(sens.sp.wide, aes(x = sens.2010, y = sens.2020, group =
[…]

### 2025-06-25 doc/manuscript_hkk_drought_sensitivity.Rmd
update with word count and submission format

@@
 
-**Number of Words in Abstract:** 150  
-**Number of Words in Main Text:** 5016  
+**Number of Words in Abstract:** 144  
+**Number of Words in Main Text:** 5019  
 <!-- Words to subtract: 2628-->
@@
 
-**Corresponding author address:** anujank@si.edu
+**Corresponding author address:**  
+Krishna Anujan,
+ForestGEO Ecosystems and Climate Lab,  
+Smithsonian's National Zoo and Conservation Biology Institute,  
+Front Royal, VA 22630  
+USA  
+Phone: 
+Email: anujank@si.edu
 
@@
 # Abstract (150 words) {-}
-An observed increase in the frequency and severity of droughts threatens forest health worldwide, with important implications for this critical carbon sink. 
-Tree species adaptations to hot and dry conditions--for example, dry season deciduousness in tropical seasonal forests--might confer drought resistance, but such resistance may be variable across species, microenvironments, and drought events.
-We analysed dendrometer band measurements on 1820 trees across 30 species in a seasonally dry tropical forest in Western <!---Huai Kha Khaeng, ---->Thailand <!--using Bayesian regressions in a causal modelling framework---> to assess tree growth responses to three climatically distinct droughts.
-Species and individuals exhibited a wide range of growth responses within each drought, and these responses differed in intensity and affect among the drought events. 
-Deciduous species were more sensitive to wet-season drought than evergreen species, and wetter microclimates amplified this difference. 
-Exposed trees showed more sensitivity to drought, but  larger sized trees or trees in wetter topographic positions seemed to buffer growth declines. 
+Increased frequency and severity of droughts threatens forest health worldwide, with important implications for this critical carbon sink. 
+Tree species adaptations to hot and dry conditions--e.g., dry-season deciduousness in tropical seasonal forests--might confer drought resistance, but such resistance may be variable across species, microenvironments, and drought events.
+We analysed dendrometer band measurements on 1820 trees across 30 species in a seasonally dry tropical forest in Western Thailand to assess tree growth responses to three climatically distinct droughts.
+Species and individuals exhibited a wide range of growth responses within each drought, and differences in response intensity and affect among the drought events. 
+Deciduous and evergreen species were more sensitive to wet- and dry-season drought respectively, and wetter microclimates amplified this difference. 
+<!-- Exposed trees showed more sensitivity to drought, but  larger sized trees or trees in wetter topographic positions seemed to buffer growth declines.  -->
 Heterogeneous drought responses of species and individuals indicate potential spatiotemporal insurance effects in diverse forests in the face of increased drought.
@@
 For forests worldwide, studies have separately shown that drought characteristics, species traits, and microenvironment affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter understanding of forest responses to drought.
-Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought.
-In contrast, if these factors interact, species and individuals that exhibit resistance to one drought may prove more vulnerable to a drought with different characteristics. 
+Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and linearly predictable; i.e., species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought.
+In contrast, if these factors interact, species and individuals that exhibit resistance to one drought may prove vulnerable to a drought with different characteristics. 
 The complexity and diversity of drought responses may act as a mechanism for maintaining species diversity and buffering ecosystem function even as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].
@@
 <!-- Sean::: don't think the above is true universally. Laura Boeschoten and Erick Esteban have separate papers well in prep showing some species demonstarating seasonal growth and others not -->
-In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020].
+In tropical forests with dry-seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020].
 <!---- agreed but see above ref I put in to [@albert_ecology_letters_cryptic_phenology].
@@
 <!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
-Analyses to date have found that interannual variation in tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent a very limited set of tropical tree species and climates.
+Analyses to date have found that interannual variation in tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry-season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent a very limited set of tropical tree species and climates.
 It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those that intensify or conflict with regular seasonal cycles<!-- of wet and dry seasons-->.
@@
 In tropical seasonal forests in particular, trees have evolved to optimize growth, reproduction, and survival (i.e., fitness) under dry seasons, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
-Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss [@desouza_drought_2020]. 
-Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and large vessels (hydraulic efficiency)<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021].
+Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry-season to avoid foliar water loss [@desouza_drought_2020]. 
+Other strategies include deep roots (hydraulic safety, to access deeper water in the dry-season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and large vessels (hydraulic efficiency)<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021].
 Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021].
@@
 i) both species and individuals respond differently to different droughts;
-ii) deciduous species, which rely on wet seasons for growth, are particularly sensitive to wet season drought; and,
+ii) deciduous species, which rely on wet-seasons for growth, are particularly sensitive to wet-season drought; and,
 iii) exposed trees are more sensitive to drought, but water availability, either through larger topographic location or tree size (and corresponding rooting volume), can buffer growth declines.
@@
 
-```{r table, echo=FALSE, warning=FALSE, message=FALSE}
+```{r table, eval=F, echo=FALSE, warning=FALSE, message=FALSE}
 # figure 0-----------------------------
@@
 Our study site was the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
-This is a 50-hectare plot in dry seasonal evergreen forest, with mean annual precipitation of 1400 mm, mean annual temperature of 23.5 C and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
+This is a 50-hectare plot in dry seasonal evergreen forest, with mean annual precipitation of 1400 mm, mean annual temperature of 23.5 C and a strong dry-season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
 The region is periodically affected by droughts associated with the *El Niño* Southern Oscillation [ENSO, @räsänen_spatial_2016].
@@
 
-![***Figure `r fig_droughts`.
+<!-- ![***Figure `r fig_droughts`.
 Climatic characteristics of the focal drought years (red in 2010, brown in 2015) compared with the average across years*** 2008-2019 (grey dashed) and its standard deviation (grey shading) in the Huai Kha Khaeng ForestGEO plot, for four daily climate variables: dry days, precipitation, maximum temperature and VPD.
@@
 Across all panels, light blue background represents the average wet season, from May to October.
-For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1.png)
+For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1.png) -->
 
-In HKK, metal dendrometer bands were installed in 2008 on 2353 trees across 152 species following the standard ForestGEO protocol [@muller-landau_metal_2008]. 
-Selected were a size-stratified and spatially stratified random sample, with additional trees chosen to fill out sample sizes for selected species of interest.
+In HKK, metal dendrometer bands were installed in 2008 on 2353 trees across 152 species following standard ForestGEO protocol [@muller-landau_metal_2008]. 
+Selected were a size-stratified and spatially-stratified random sample, with additional trees chosen to fill out sample sizes for selected species of interest.
 Thus relative abudances among the selected trees largely reflected abundance variation in the plot as a whole, with more individuals banded from the abundant species (n=1 to 230 trees per species). 
-On each tree, the dendrometer band was installed at a height of 10 cm above the regular point of measurement for the whole plot census meaning installation at 1.4 m on most trees, and at higher heights on buttressed trees with higher measurement points.
+On each tree, the dendrometer band was installed at a height of 10 cm above the regular point of measurement for the whole-plot census meaning installation at 1.4 m on most trees, and at higher heights on buttressed trees with higher measurement points.
 Dendrobands were censused twice a year with the measurement periods typically centred in June and January. 
@@
 -->
-We first conducted quality control on window size measurements (see supplementary methods) and combined them with the diameter measurement at installation $(d_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc, assuming a circular cross section[@detto_correcting_2023].
+We first conducted quality control on window size measurements (see supplementary methods) and combined them with the diameter measurement at installation $(d_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc[@detto_correcting_2023].
 We used a direct measurement of $d_{1}$ at the time of installation if it was available; if not, we excluded window size measurements until the first available diameter measurement. 
@@
 We conducted further quality control on increments (see supplementary methods for details) to assemble the final dataset. 
-Diameter tape measurements themselves are not precise enough for calculating drought sensitivity; error for growth is twice measurement error = 2 mm. 
-In comparison, median annual growth across trees in HKK is ~1.4 mm and calculating deviation from annual growth needs sub-mm precision. 
+Diameter tape measurements are not precise enough for calculating drought sensitivity. 
+Error for growth is twice measurement error = 2 mm; in comparison, median annual growth in HKK is ~1.4 mm and calculating deviation from annual growth needs sub-mm precision. 
 However, we used tape measurements to flag likely errors in the dendrometer dataset. 
@@
 
-![***Figure `r fig_drought_responses`: Growth increment timeseries and drought responses*** across trees and species in the Huai Kha Khaeng ForestGEO plot.
+<!-- ![***Figure `r fig_drought_responses`: Growth increment timeseries and drought responses*** across trees and species in the Huai Kha Khaeng ForestGEO plot.
 a) Median of annualised growth increments across ten dominant species (colours) and all trees (black) from 2009 to 2022.
 b) Distribution of drought sensitivity of tree growth across all individuals in the two drought years (see main text for equation).
-For comments - https://github.com/forestgeo/growth-precip-thailand/issues/11](display/Fig2_alt.png)
+For comments - https://github.com/forestgeo/growth-precip-thailand/issues/11](display/Fig2_alt.png) -->
 
@@
 The droughts analysed had distinct climatic and temporal characteristics, with 2010 primarily a dry-season drought, 2015 a wet-season drought and 2020 showing mixed characteristics (Table 1, Fig `r fig_droughts`).
-The 2010 drought showed 30-day-mean meteorological anomalies between February and May including lower precipitation (min anomaly **<-1** SD), higher number of dry days (anomaly **~1.5** SD), higher maximum temperature (anomaly **>2** SD), and higher VPD (anomaly **>2** SD), including in the wet season.
+The 2010 drought showed 30-day-mean meteorological anomalies between February and May including lower precipitation (min anomaly **<-1** SD), higher number of dry days (anomaly **~1.5** SD), higher maximum temperature (anomaly **>2** SD), and higher VPD (anomaly **>2** SD), including in the wet-season.
 <!-- Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions. -->
 1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought. 
-In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including precipitation (anomaly **<-1.5** SD), number of dry days (max anomaly **>2** SD), maximum monthly temperatures (anomaly **+1.5** SD), and monthly maximum VPD (max anomaly **+2** SD).
+In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet-season, including precipitation (anomaly **<-1.5** SD), number of dry days (max anomaly **>2** SD), maximum monthly temperatures (anomaly **+1.5** SD), and monthly maximum VPD (max anomaly **+2** SD).
 <!-- 1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period.  -->
@@
 1- to 12-month SPEI values crossed minima -2, lowest in the dendroband census period and indicative of severe drought. 
-In 2020, meteorological anomalies were pronounced both in the dry and wet seasons, including maximum temperature (**>2** SD) and VPD (**>2** SD). 
+In 2020, meteorological anomalies were pronounced both in the dry- and wet-seasons, including maximum temperature (**>2** SD) and VPD (**>2** SD). 
 3- and 6-month SPEI values reached minima ~2 in February, indicative of borderline severe drought.
-12-month SPEI showed severe drought conditions from January to April, and moderate drought from May to September, spanning both dry and wet seasons.
+12-month SPEI showed severe drought conditions from January to April, and moderate drought from May to September, spanning both dry- and wet-seasons.
 
@@
 
-![***Figure `r fig_species`: Variation in sensitivity among species*** in the Huai Kha Khaeng ForestGEO plot during droughts in 2010, 2015 and 2020.
+<!-- ![***Figure `r fig_species`: Variation in sensitivity among species*** in the Huai Kha Khaeng ForestGEO plot during droughts in 2010, 2015 and 2020.
 a) Relationship of the mean species sensitivities from model fits of an intercept-only model, with mean species deciduousness values.
@@
 b) Predicted sensitivity values across the observed range of deciduousness and topographic wetness index from a model with TWI, deciduosness and their interaction across all trees modelled for each drought year separately.
-For comments: https://github.com/forestgeo/growth-precip-thailand/issues/12](display/Fig3.png)
+For comments: https://github.com/forestgeo/growth-precip-thailand/issues/12](display/Fig3.png) -->
 
@@
 
-![***Figure `r fig_microenvs`: Drivers of variation in sensitivity among individuals***.
+<!-- ![***Figure `r fig_microenvs`: Drivers of variation in sensitivity among individuals***.
 All panels represent model results from models (one for each year) predicting sensitivity with microenvironment conditions with species random effect on all slopes.
@@
 Black line represents overall relationship (panel a); coloured lines for each species reflects deciduousness values.
-For comments - https://github.com/forestgeo/growth-precip-thailand/issues/13](display/Fig4.png)
+For comments - https://github.com/forestgeo/growth-precip-thailand/issues/13](display/Fig4.png) -->
 
@@
 
-![***Figure `r fig_dag`: Effects of size, exposure and water availability*** from combined models.
+<!-- ![***Figure `r fig_dag`: Effects of size, exposure and water availability*** from combined models.
 Hypothesised relationships between microenvironment variables and drought sensitivity with their modelled effects in the HKK ForestGEO plot in the droughts of a) 2010, b) 2015 and c) 2020.
@@
 d) Modelled sensitivity across crown exposure classes in 2010, 2015 and 2020.
-For comments - https://github.com/forestgeo/growth-precip-thailand/issues/14](display/Fig5.png)
+For comments - https://github.com/forestgeo/growth-precip-thailand/issues/14](display/Fig5.png) -->
 
@@
 Despite lower extremes of temperature, vapor pressure deficit, and number of dry days (Fig.
-2a), severe drought in the wet season (2015) elicited stronger responses across trees, especially deciduous species, which are expected to have stronger seasonality in growth phenology (Fig `r fig_species`).
+2a), severe drought in the wet-season (2015) elicited stronger responses across trees, especially deciduous species, which are expected to have stronger seasonality in growth phenology (Fig `r fig_species`).
 Interestingly, in 2015 and 2020 (but not 2010), evergreen and deciduous species had divergent responses, especially along a wetness gradient; deciduous species showed greater growth declines in wetter sites in 2015 while evergreen species grew less in drier sites in 2020 (Fig `r fig_species`b, Fig `r fig_microenvs`b).
 Although trees with more exposed crowns were consistently more sensitive to drought, additional effects of DBH -- presumably including rooting volume and depth -- were dissimilar between the drought events (Fig `r fig_microenvs`a).
-It is possible that water availability buffered exposure-associated growth declines through different mechanisms in the different droughts; larger trees fared better in the dry-season drought while trees in wetter sites suffered smaller growth declines during wet season droughts (Fig `r fig_dag`).
+It is possible that water availability buffered exposure-associated growth declines through different mechanisms in the different droughts; larger trees fared better in the dry-season drought while trees in wetter sites suffered smaller growth declines during wet-season droughts (Fig `r fig_dag`).
 <!-- we can talk about it ... but I don't think that we can know that larger trees have deeper roots.  -->
-Taken together, these results indicate that joint and interactive effects of drought characteristics, species traits, and individual microenvironments produce important heterogeneity in individual and species drought responses of tropical trees.
-This finding that individual droughts have divergent effects on species and individuals may contribute to the maintenance of species diversity and ecosystem stability.
+Taken together, these results indicate that combined effects of drought characteristics, species traits, and individual microenvironments produce important heterogeneity in individual and species drought responses of tropical trees.
+Divergent effects of species and individuals to droughts have  may contribute to the maintenance of species diversity and ecosystem stability.
 
@@
 But the strength of the ecological response to the 2015 wet-season drought (Fig. `r fig_droughts`) does not align with this assumption.
-Generalization of tropical tree growth responses to any specific drought likely requires fine-scale monitoring of growth (i.e., dendrometer band measurements or dendrochronology when possible) on species that span strategies and over different timings and intensities of droughts.
-Advancing efforts to collect these data is critical to our assessment and prediction of tropical forest responses to changing precipitation patterns.
+Generalization of tropical tree growth responses likely requires fine-scale monitoring of growth (i.e., dendrometer band measurements or dendrochronology when possible) on species that span strategies and over different timings and intensities of droughts.
+Advancing efforts to collect these data is critical to assessment and prediction of tropical forest responses to changing precipitation patterns.
 <!-- [is this necessary?]  Studies in temperate forests, where extensive tree-ring records are available, have confirmed that drought characteristics including seasonal timing, duration, and severity affect tree growth responses [@gao_dynamic_2018; @dorangeville_drought_2018] and that effects of species, species traits, tree size, and microenvironment can differ across droughts [@mcgregor_tree_2021].
@@
 This finding is consistent with the concept that larger trees tend to suffer more during drought [@bennett_larger_2015], but adds the important insight that it is crown position -- as opposed to size *per se* -- that is primarily responsible for this pattern.
-The strong and consistent negative effect of crown position makes sense in light of pronounced gradients in drought stress across forest vertical profiles [@vinod_thermal_2023].
+The consistent negative effect of crown position aligns with pronounced gradients in drought stress across forest vertical profiles [@vinod_thermal_2023].
 We show that residual effects of size, presumably including root water access and biophysical challenges linked to height itself, may act in different directions depending on the drought characteristics (Figs.
 `r fig_microenvs`, `r fig_dag`).
-Together, these findings imply that while canopy and emergent trees face the greatest stress and exhibit the most negative growth responses, a suite of species traits and microenvironmental characteristics modify growth responses in interaction with drought characteristics.
+These findings imply that while canopy and emergent trees experience the greatest stress and the most negative growth responses, a suite of species traits and microenvironmental characteristics modify growth responses in interaction with drought characteristics.
 
@@
 Our study reveals that drought sensitivity is not a static trait associated to particular species, traits, or microenvironments, but rather a dynamic response shaped interactively by these factors and drought characteristics.
-Despite net decreased growth at the plot level (in agreement with landscape-scale in @aguirre-gutiérrez_functional_2022 and @bennett_sensitivity_2023), we show evidence that the same strategies are not uniformly successful across spatiotemporal climatic variation.
+Despite net decreased growth at the plot-level (in agreement with landscape-scale in @aguirre-gutiérrez_functional_2022 and @bennett_sensitivity_2023), we show evidence that the same strategies are not uniformly successful across spatiotemporal climatic variation.
 Static traits are often used to explain and predict forest responses to drought, especially in the tropics where long-term records are few [@guillemot_small_2022; @vico_how_2017a]. 
@@
 
+# References  {-}
+
+<div id="refs"></div>
+
+\newpage
+
+# Table Caption {-}
+
+**Table 1.
+Summary table of hypotheses tested, effects of variables on drought sensitivity and corresponding evidence.** For each variable x year combination, "+" indicates a positive effect, "-" indicates a negative effect and "n.s." indicates that the effect was not statistically significant.
+
+\newpage
+
 # Figure Legends {-}
@@
 ***Figure `r fig_droughts`.
-Climatic characteristics of the focal drought years (red in 2010, brown in 2015) compared with the average across years*** 2008-2019 (grey dashed) and its standard deviation (grey shading) in the Huai Kha Khaeng ForestGEO plot, for four daily climate variables: dry days, precipitation, maximum temperature and VPD.
+Climatic characteristics of the focal drought years (black in 2010, purple in 2015, orange in 2020) compared with the average across years*** 2008-2019 (grey dashed) and its standard deviation (grey shading) in the Huai Kha Khaeng ForestGEO plot, for four daily climate variables: dry days, precipitation, maximum temperature and VPD.
 (a) 30-day running means.
@@
 A k-month SPEI integrates over k months ending in the month current month.
-Across all panels, light blue background represents the average wet season, from May to October.
+Across all panels, light blue background represents the average wet-season, from May to October.
 
@@
 a) Median of annualised growth increments across ten dominant species (colours) and all trees (black) from 2009 to 2022.
-b) Distribution of drought sensitivity of tree growth across all individuals in the two drought years (see main text for equation).
+b) Distribution of drought sensitivity of tree growth across all individuals in the three drought years (see main text for equation).
 
@@
 d) Modelled sensitivity across crown exposure classes in 2010, 2015 and 2020.
-
-\newpage
-
-# References  {-}

### 2025-06-25 doc/manuscript_hkk_drought_sensitivity.Rmd
updated manuscript with table

@@
 - **leaf habit x habitat** : tropical evergreen and deciduous species have different habitat preferences [@kunert_leaf_2021] that could affect their drought resistance.   -->
-
-```{r table, eval=F, echo=FALSE, warning=FALSE, message=FALSE}
-# figure 0-----------------------------
-
-# make a figure with summaries
-
-library(flextable)
-set_flextable_defaults(font.family = "Arial", font.size = 9)
-
-sumtab <- read.csv("display/summary_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-
-ft <- flextable(sumtab)
-ft <- set_header_labels(ft,
-  na = ""
-)
-ft <- hline(ft, i = c(1, 4, 8), part = "body")
-ft <- line_spacing(ft, space = 1, part = "all")
-ft <- autofit(ft)
-# ft <- height(ft, height = 0.3)
-ft <- bold(ft, i = c(1, 4, 8), bold = TRUE, part = "body")
-ft <- width(ft, 1, width = 3.5)
-# ft <- width(ft, 2, width = 1.75)
-# ft <- width(ft, 3, width = 3)
-ft <- set_caption(ft, caption = "Summary table of hypotheses tested, effects of variables on drought sensitivity and corresponding evidence.
-For each variable x year combination, \"+\" indicates a positive effect, \"-\" indicates a negative effect and \"n.s.\" indicates that the effect was not statistically significant.")
-ft
-```
 
@@
 
-# Table Caption {-}
+# Tables {-}
 
-**Table 1.
-Summary table of hypotheses tested, effects of variables on drought sensitivity and corresponding evidence.** For each variable x year combination, "+" indicates a positive effect, "-" indicates a negative effect and "n.s." indicates that the effect was not statistically significant.
+```{r table, echo=FALSE, warning=FALSE, message=FALSE}
+# figure 0-----------------------------
+
+# make a figure with summaries
+
+library(flextable)
+set_flextable_defaults(font.family = "Arial", font.size = 9)
+
+sumtab <- read.csv("display/summary_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+
+ft <- flextable(sumtab)
+ft <- set_header_labels(ft,
+  na = ""
+)
+ft <- hline(ft, i = c(1, 4, 8), part = "body")
+ft <- line_spacing(ft, space = 1, part = "all")
+ft <- autofit(ft)
+# ft <- height(ft, height = 0.3)
+ft <- bold(ft, i = c(1, 4, 8), bold = TRUE, part = "body")
+ft <- width(ft, 1, width = 3.5)
+# ft <- width(ft, 2, width = 1.75)
+# ft <- width(ft, 3, width = 3)
+ft <- set_caption(ft, caption = "Summary table of hypotheses tested, effects of variables on drought sensitivity and corresponding evidence.
+For each variable x year combination, \"+\" indicates a positive effect, \"-\" indicates a negative effect and \"n.s.\" indicates that the effect was not statistically significant.")
+ft
+```
 

### 2025-06-26 doc/manuscript_hkk_drought_sensitivity.Rmd
update Fig. 1 caption
#10

@@
 ***Figure `r fig_droughts`.
-Climatic characteristics of the focal drought years (black in 2010, purple in 2015, orange in 2020) compared with the average across years*** 2008-2019 (grey dashed) and its standard deviation (grey shading) in the Huai Kha Khaeng ForestGEO plot, for four daily climate variables: dry days, precipitation, maximum temperature and VPD.
+Climatic characteristics of the focal drought years compared with the average across years*** in the Huai Kha Khaeng ForestGEO plot, for four daily climate variables: dry days, precipitation, maximum temperature and VPD.
 (a) 30-day running means.
-(b) Anomalies of the drought year running means from the averages, in units of standard deviations.
+(b) Anomalies of the drought year running means from the 2008-2019 averages, in units of standard deviations.
 (c) Standardised Precipitation Evapotranspiration Index (SPEI) values for the drought years at four different time scales for each month for each drought year.

### 2025-06-26 scripts/06_figures.R
@teixeirak fixed Fig 1 based on #10

@@ -542,8 +542,8 @@ bounds2 <- spei_month %>%
 bounds2
 
 df <- data.frame(
-    x = rep(9.5, 4), y = c(0.75, -0.7, -1.5, -2.5),
-    label = c("wetter", "drier", "drought", "severe drought")
+    x = rep(9.5, 4), y = c(0.75, -0.7, -1.5, -3),
+    label = c("wetter", "drier", "drought", "severe\ndrought")
 )
 
 cols <- viridis::viridis(4, option = "F")
@@ -668,7 +668,7 @@ climsat_plot <- ggplot(climsat_rlmean %>% filter(year %in% c("2010", "2015", "20
     labs(x = "Dy of year", y = "", color = "Year") +
     # scale_x_continuous(breaks = 1:12) +
     guides(linetype = "none", color = "none") +
-    ggtitle("Climate variables (remote)") +
+    ggtitle("Climate variables") +
     # scale_color_manual(values = c("2010" = "indianred2", "2015" = "indianred4")) +
     # viridis::scale_color_viridis(discrete = T) +
     scale_color_manual(values = cols[1:3]) +

### 2025-06-26 doc/manuscript_hkk_drought_sensitivity.Rmd
abstract suggestions

@@
 Increased frequency and severity of droughts threatens forest health worldwide, with important implications for this critical carbon sink. 
-Tree species adaptations to hot and dry conditions--e.g., dry-season deciduousness in tropical seasonal forests--might confer drought resistance, but such resistance may be variable across species, microenvironments, and drought events.
-We analysed dendrometer band measurements on 1820 trees across 30 species in a seasonally dry tropical forest in Western Thailand to assess tree growth responses to three climatically distinct droughts.
+Tree species adaptations to hot and dry conditions -- e.g., dry-season deciduousness in tropical seasonal forests -- can confer drought resistance, but such resistance may be variable across species, microenvironments, and drought events.
+Here, we assess growth responses of 1820 trees across 30 species to three climatically distinct droughts in a seasonally dry tropical forest in Western Thailand. 
 Species and individuals exhibited a wide range of growth responses within each drought, and differences in response intensity and affect among the drought events. 
-Deciduous and evergreen species were more sensitive to wet- and dry-season drought respectively, and wetter microclimates amplified this difference. 
+Deciduous and evergreen species were more sensitive to wet- and dry-season drought respectively<!--, and wetter microclimates amplified this difference-->.
+While crown exposure consistently increased drought sensitivity, topographic wetness had variable effects.
 <!-- Exposed trees showed more sensitivity to drought, but  larger sized trees or trees in wetter topographic positions seemed to buffer growth declines.  -->

### 2025-06-26 doc/manuscript_hkk_drought_sensitivity.Rmd
changes to intro

@@
 
-Intense droughts are becoming more frequent, affecting the demography, diversity, and  carbon (C) cycling of forests globally [@ipcc_climate_2023]. 
-However, drought impacts on tree woody growth and ecosystem C sequestration remain difficult to predict, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022].
+Intense droughts are becoming more frequent [@ipcc_climate_2023], affecting the demography, diversity, and  carbon (C) cycling of forests globally. 
+However, drought impacts on tree woody growth and ecosystem C sequestration remain difficult to predict [@evans_scaling_2025], particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022].
 For forests worldwide, studies have separately shown that drought characteristics, species traits, and microenvironment affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter understanding of forest responses to drought.
@@
 [Then go into some details with citations on ?] -->
-Drought is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the timing, severity and duration of climatic anomalies [@slette_how_2019] with potentially distinct ecological consequences.
+Drought is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the timing, severity and duration of climatic anomalies [@slette_how_2019].
 <!-- ; we define it here as hot/dry conditions anomalously departing from climatic means during any time of year  -->
-Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different seasonal timing, duration, or intensity may have distinct effects on tree growth, especially in seasonal forests with diverse adaptive strategies to handle regular dry periods [@albert_cryptic_2019;@gao_dynamic_2018].
+Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], differences in seasonal timing, duration, or intensity of droughts may result in variable tree growth responses, especially in seasonal forests with diverse adaptive strategies to handle regular dry periods [@albert_cryptic_2019;@gao_dynamic_2018].
 <!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
 <!-- Sean::: don't think the above is true universally. Laura Boeschoten and Erick Esteban have separate papers well in prep showing some species demonstarating seasonal growth and others not -->
-In tropical forests with dry-seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020].
+In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020].
 <!---- agreed but see above ref I put in to [@albert_ecology_letters_cryptic_phenology].
@@
 Moreover, besides being adaptively conserved, these strategies can also have developmental or ecophysiological controls.
-For example, some species regulate negative $\pi_{tlp}$'s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; @hulshof_variation_2010].
+For example, some species regulate $\pi_{tlp}$, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; @hulshof_variation_2010].
 
@@
 -->
-Microenvironments defined by horizontal topographic and edaphic variation, and vertical light and thermal variation further shape individual tree experience of, and response to, drought. 
-While habitats with relatively moist soils (e.g., low-lying topography) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted (e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021).
+Microenvironments defined by horizontal topographic and edaphic variation and vertical light and thermal variation further shape individual tree experience of, and response to, drought. 
+While habitats with relatively moist soils (e.g., low-lying topography) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted [e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021].
 Indeed, during drought trees near streams experienced greater growth declines in a temperate forest [@mcgregor_tree_2021] and increased mortality in a tropical forest [@zuleta_droughtinduced_2017]. 
@@
 However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
-While the overall tendency is that larger trees suffer greater growth declines during drought [@bennett_larger_2015], it remains unclear whether this is driven primarily by crown exposure, by greater height itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium <!--, as described by Darcy's law -->[@mcdowell_darcys_2015; @fernández-de-uña_role_2023], or by trait differences across vertical axis [@vinod_thermal_2023].
+While the overall tendency is that larger trees suffer greater growth declines during drought [@bennett_larger_2015], this may be driven primarily by crown exposure, by greater height itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium <!--, as described by Darcy's law -->[@mcdowell_darcys_2015; @fernández-de-uña_role_2023], or by trait differences across vertical axis [@vinod_thermal_2023].
 Thus, horizontal and vertical microenvironmental gradients modify tree drought responses, likely in interaction with drought characteristics and species traits.
@@
 i) both species and individuals respond differently to different droughts;
-ii) deciduous species, which rely on wet-seasons for growth, are particularly sensitive to wet-season drought; and,
-iii) exposed trees are more sensitive to drought, but water availability, either through larger topographic location or tree size (and corresponding rooting volume), can buffer growth declines.
+ii) deciduous species, which rely on wet seasons for growth, are particularly sensitive to wet-season drought; and,
+iii) exposed trees are more sensitive to drought, whereas water access, either through topographic location or tree size (and corresponding rooting volume), has variable effects.
 We tested these hypotheses using  a 14-year record of dendrometer band measurements in a dry seasonal evergreen forest [@bunyavejchewin_seasonally_2011] in Thailand. 

### 2025-06-26 doc/manuscript_hkk_drought_sensitivity.Rmd
Update manuscript_hkk_drought_sensitivity.Rmd
@krishnaanujan , pushing one small edit where I commented out a part that wasn't clear to me. You may want to clarify/ reinstate.

@@
 The droughts analysed had distinct climatic and temporal characteristics, with 2010 primarily a dry-season drought, 2015 a wet-season drought and 2020 showing mixed characteristics (Table 1, Fig `r fig_droughts`).
-The 2010 drought showed 30-day-mean meteorological anomalies between February and May including lower precipitation (min anomaly **<-1** SD), higher number of dry days (anomaly **~1.5** SD), higher maximum temperature (anomaly **>2** SD), and higher VPD (anomaly **>2** SD), including in the wet-season.
+In 2010, 30-day-mean meteorological anomalies between February and May included lower precipitation (min anomaly **<-1** SD), higher number of dry days (anomaly **~1.5** SD), higher maximum temperature (anomaly **>2** SD), and higher VPD (anomaly **>2** SD)<!--, including in the wet-season [KAT: commenting this out because I find it unclear/ don't understand what it means]-->.
 <!-- Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions. -->

### 2025-06-26 doc/manuscript_hkk_drought_sensitivity.Rmd
report fewer digits on % change estimates
(This seemed like false precision)

@@
 <!-- Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season. -->
-1- to 12-month SPEI values crossed minima -2, lowest in the dendroband census period and indicative of severe drought. 
-In 2020, meteorological anomalies were pronounced both in the dry- and wet-seasons, including maximum temperature (**>2** SD) and VPD (**>2** SD). 
-3- and 6-month SPEI values reached minima ~2 in February, indicative of borderline severe drought.
+1- to 12-month SPEI fell below -2, lowest in the dendroband census period and indicative of severe drought. 
+In 2020, meteorological anomalies were pronounced both in the dry and wet seasons, including maximum temperature (**>2** SD) and VPD (**>2** SD). 
+3- and 6-month SPEI values approached -2 in February, indicative of borderline severe drought.
 12-month SPEI showed severe drought conditions from January to April, and moderate drought from May to September, spanning both dry- and wet-seasons.
@@
 Median plot-level annual growth rate across years of measurement was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual*10, na.rm=T),2)` mm. 
-The drought year 2015 had the lowest median increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
-In the 2010 and 2020 droughts, the median increments were `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm and `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm respectively, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% and `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
+The drought year 2015 had the lowest median increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% lower than the median across all years.
+In the 2010 and 2020 droughts, the median increments were `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm and `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm respectively, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% and `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% lower than the median across all years.
 <!-- In the 2020 drought, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.

### 2025-06-26 doc/manuscript_hkk_drought_sensitivity.Rmd
results

@@
 
-Median annual increment varied among species with *`r tree.time.sp$spfull[1]`* the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm) among the `r nrow(tree.time.sp)` species analysed.
-Median species responses to the 2010 and 2020 droughts were variable, but all species had lower median increments than a typical year in 2015. 
+Median annual increment varied among species, with *`r tree.time.sp$spfull[1]`* the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm) among the `r nrow(tree.time.sp)` species analysed.
+Median species responses to the 2010 and 2020 droughts were variable, but no species had higher median increments than a typical year in 2015.
+<!--KAT reworded here because Mitrephora thorelii was very slightly positive-->
 In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/3]`*.
-In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/3]`* and in 2020 from `r sens.sp[sens.sp$yr==2020,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2020,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[nrow(sens.sp)/3]`*.
+In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/3]`*, and in 2020 from `r sens.sp[sens.sp$yr==2020,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2020,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[nrow(sens.sp)/3]`*.
 

### 2025-06-26 doc/manuscript_hkk_drought_sensitivity.Rmd
working on predictions and organization around predictions
@krishnaanujan , here are suggested changes to predictions and organization around these predictions (linking back to predictions in results section). Still in the middle of this, but the little guy is starting to wake...

Along with this, you might make some minor changes to table 1 (referring to predictions?, labeling "species and functional group effects", perhaps moving deciduousness x TWI interaction up?)

@@
 Deciduous and evergreen species were more sensitive to wet- and dry-season drought respectively<!--, and wetter microclimates amplified this difference-->.
-While crown exposure consistently increased drought sensitivity, topographic wetness had variable effects.
+While individuals with more exposed crowns tended to grow less in all droughts, tree size and topographic wetness had variable effects.
 <!-- Exposed trees showed more sensitivity to drought, but  larger sized trees or trees in wetter topographic positions seemed to buffer growth declines.  -->
@@
 <!--HYPOTHESIS PARAGRAPH-->
-We hypothesize that tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, and microenvironments, resulting in variable drought responses of species and individuals across droughts, and of species across microhabitats.
+We hypothesize that tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, <!--tree size--> and microenvironment, resulting in variable drought responses of species and individuals across droughts, and of species across microhabitats.
 Specifically, we test predictions that:
 i) both species and individuals respond differently to different droughts;
-ii) deciduous species, which rely on wet seasons for growth, are particularly sensitive to wet-season drought; and,
-iii) exposed trees are more sensitive to drought, whereas water access, either through topographic location or tree size (and corresponding rooting volume), has variable effects.
+ii) deciduous and evergreen species exhibit differential sensitivity depending on the timing of drought and topographic location,
+iii) the influences of topographic location, crown exposure, and other size effects on individual drought responses are variable across droughts.
 We tested these hypotheses using  a 14-year record of dendrometer band measurements in a dry seasonal evergreen forest [@bunyavejchewin_seasonally_2011] in Thailand. 
@@
 
-Species median sensitivities to the three droughts were not correlated with each other, although on average, species had more negative sensitivities to the 2015 drought than the 2010 or 2020 droughts, in alignment with median sensitivities across all trees (for 2010 and 2015 Pearson's correlation coefficient r = `r round(sens.sp.cor_10_15[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_15[3]$p.value,2)`, for 2015 and 2020 was r = `r round(sens.sp.cor_15_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_15_20[3]$p.value,2)`, and for  2010 and 2020 r = `r round(sens.sp.cor_10_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_20[3]$p.value,2)`, Fig S5).
+Supporting prediction *i*, species median sensitivities to the three droughts were not correlated with each other, although on average, species had more negative sensitivities to the 2015 drought than the 2010 or 2020 droughts, in alignment with median sensitivities across all trees (for 2010 and 2015 Pearson's correlation coefficient r = `r round(sens.sp.cor_10_15[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_15[3]$p.value,2)`, for 2015 and 2020 was r = `r round(sens.sp.cor_15_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_15_20[3]$p.value,2)`, and for  2010 and 2020 r = `r round(sens.sp.cor_10_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_20[3]$p.value,2)`, Fig S5).
 
@@
 <!-- Deciduousness species responded more negatively than evergreen species in 2015, but not in 2010 (Fig `r fig_species`a).  -->
-Deciduous and evergreen species showed heterogenous responses to different droughts.
+Supporting prediction *ii*, deciduous and evergreen species showed heterogenous responses to different droughts.
 Deciduous species were more drought sensitive than evergreen species in 2015 (Fig `r fig_species`a, r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`), but the opposite was true in 2020 (`r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2020], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2020], 2)`).
@@
 
-Individual trees did not respond similarly to the three drought events (Fig S6).
+Supporting prediction *i*, individual trees did not respond similarly to the three drought events (Fig S6).
 
-TWI showed significant positive effect in 2015 and 2020 but no significant effect on sensitivity in 2010. 
+Supporting prediction *iii*, TWI had different effects across droughts. 
 Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`) and 2020 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2020],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2020],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2020],2)`).
@@
 
+Supporting prediction *iii*, ...
 Crown exposure was associated with greater growth declines in 2010.

### 2025-06-26 scripts/06_figures.R
knit doc with @teixeirak edits

@@ -584,6 +584,7 @@ speiplot_line <- ggplot() +
         # scales = "free_y",
         strip.position = "left", nrow = 4, labeller = varnames
     ) +
+    ylim(-3, 1.5)+
     theme_bw() +
     labs(x = "Month", y = "SPEI", fill = "Year") +
     scale_x_continuous(breaks = 1:12) +

### 2025-06-27 doc/manuscript_hkk_drought_sensitivity.Rmd
Update manuscript_hkk_drought_sensitivity.Rmd
a couple tweaks to discussion
@krishnaanujan , FYI

@@
 
-Supporting prediction *iii*, ...
+Supporting prediction *iii*, drought sensitivities of individuals with different crown exposures varied across droughts (Fig `r fig_dag`).
 Crown exposure was associated with greater growth declines in 2010.
@@
 -->
-However, our finding that a wet-season drought in 2015 had greater impact on woody growth than a dry-season drought in 2010 questions the generality of previous findings. 
+However, our finding that a wet-season drought in 2015 had greater impact on woody growth than dry-season droughts in 2010 and 2020 questions the generality of previous findings. 
 Tree ring studies from this site [@vlam_temperature_2014;@anderson-teixeira_joint_2022] and across the tropics [@zuidema_tropical_2022] have asserted that tree growth is most sensitive to dry-season precipitation.
-But the strength of the ecological response to the 2015 wet-season drought (Fig. `r fig_droughts`) does not align with this assumption.
+However, the strength of the ecological response to the 2015 wet-season drought (Fig. `r fig_droughts`) does not align with this assumption.
 Generalization of tropical tree growth responses likely requires fine-scale monitoring of growth (i.e., dendrometer band measurements or dendrochronology when possible) on species that span strategies and over different timings and intensities of droughts.

### 2025-06-27 doc/cover_letter.Rmd
updated cover letter

@@
 
-April 22, 2025
+June 27, 2025
 \newline
@@
 
-We are pleased to submit our manuscript "Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments" for consideration for publication in *Ecology Letters*.
+We are pleased to submit our manuscript “Drought response is not a species trait: tropical tree drought sensitivity is shaped by drought characteristics, species adaptations, and individual microenvironments” for consideration for publication in
+*Ecology Letters*.
 
@@
 While plot-level observations show that severe droughts lead to lower biomass gains, species and tree level growth declines could be influenced by multiple drivers that could either amplify or buffer growth declines during drought years.
-Some tree species adaptations to predictable hot, dry conditions such as dry season deciduousness in seasonal tropical forests could confer drought resistance but this could differ with droughts, species and individual tree microenvironments.
+Some tree species adaptations to predictable
+hot, dry conditions such as dry season deciduousness in seasonal tropical forests could confer drought resistance but this could differ with droughts, species and individual tree microenvironments. 
 Current knowledge on tropical tree responses to drought however is limited to understanding from tree ring records, but very few species in the tropics show annual rings.
@@
 Here, we use dendrometer band measurements from 1820 trees across 30 representative species in a tropical seasonal forest in northern Thailand to show that tree growth responses to drought can be influenced by drought characteristics, microenvironments and species traits.
-In two ENSO drought years within a 14-year timeseries, species and trees showed heterogeneous drought sensitivity across drought events analysed, as opposed to consistent or correlated responses scaled with drought intensity.
-Exposed trees consistently suffered greater growth declines, but the effects of individual and species drivers related to tree-water relations - topographic wetness, tree size and leaf deciduousness - were all heterogenous between a dry- and wet-season drought.
+In three drought years within a 14-year timeseries, species and trees showed heterogeneous drought sensitivity across drought events analysed, as opposed to consistent or correlated responses scaled with drought intensity.
+While exposed trees suffered greater growth declines, the effects of individual and species drivers related to tree-water relations - topographic wetness, tree size and leaf deciduousness - were all heterogenous between droughts that differed in climatic characteristics.
 Our findings imply that although drought resistance is often regarded as a static trait, the species and tree level responses show a rich and dynamic diversity of responses across drought events. 
@@
 
-We believe our manuscript to be an appropriate length for *Ecology Letters*, with 4989 words and 6 display items in the main article.
-The article is accompanied by a Supplementary Information file (12 figures).
+We believe our manuscript to be an appropriate length for *Ecology Letters*, with 5019 words and 6 display items in the main article.
+The article is accompanied by a Supplementary Information file (16 figures).
 

### 2025-06-27 doc/manuscript_hkk_drought_sensitivity.Rmd
Update manuscript_hkk_drought_sensitivity.Rmd
working on #33

@@
 # Abstract (150 words) {-}
-Increased frequency and severity of droughts threatens forest health worldwide, with important implications for this critical carbon sink. 
-Tree species adaptations to hot and dry conditions -- e.g., dry-season deciduousness in tropical seasonal forests -- can confer drought resistance, but such resistance may be variable across species, microenvironments, and drought events.
+Increased frequency and severity of droughts threatens forest health worldwide<!--, with important implications for this critical carbon sink-->. 
+Tree species adaptations <!--to hot and dry conditions-->-- e.g., dry-season deciduousness in tropical seasonal forests -- and individual traits -- e.g., size, crown position -- shape drought resistance, but such resistance may be variable across species, microenvironments, and drought events.
 Here, we assess growth responses of 1820 trees across 30 species to three climatically distinct droughts in a seasonally dry tropical forest in Western Thailand. 
@@
 Deciduous and evergreen species were more sensitive to wet- and dry-season drought respectively<!--, and wetter microclimates amplified this difference-->.
-While individuals with more exposed crowns tended to grow less in all droughts, tree size and topographic wetness had variable effects.
+While individuals with more exposed crowns tended to grow less in all droughts, stem diameter and topographic wetness had variable effects.
 <!-- Exposed trees showed more sensitivity to drought, but  larger sized trees or trees in wetter topographic positions seemed to buffer growth declines.  -->
@@
 Intense droughts are becoming more frequent [@ipcc_climate_2023], affecting the demography, diversity, and  carbon (C) cycling of forests globally. 
-However, drought impacts on tree woody growth and ecosystem C sequestration remain difficult to predict [@evans_scaling_2025], particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019] but for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022].
-For forests worldwide, studies have separately shown that drought characteristics, species traits, and microenvironment affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter understanding of forest responses to drought.
-Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and linearly predictable; i.e., species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought.
-In contrast, if these factors interact, species and individuals that exhibit resistance to one drought may prove vulnerable to a drought with different characteristics. 
+However, drought impacts on tree woody growth and ecosystem C sequestration remain difficult to predict [@evans_scaling_2025], particularly for tropical forests, <!--whose critical C sink is threatened by drought [@brando_droughts_2019] but--> for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022].
+For forests worldwide, studies have separately shown that drought characteristics, species traits, tree size and microenvironment affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter understanding of forest responses to drought.
+Specifically, if these factors act independently, tree drought responses to may be conserved, and linearly predictable; i.e., species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought.
+In contrast, if these factors interact, species and individuals that exhibit resistance to one drought in one particular environment may prove vulnerable to a drought with different characteristics or in a different environment.
 The complexity and diversity of drought responses may act as a mechanism for maintaining species diversity and buffering ecosystem function even as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].
@@
 <!--HYPOTHESIS PARAGRAPH-->
-We hypothesize that tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, <!--tree size--> and microenvironment, resulting in variable drought responses of species and individuals across droughts, and of species across microhabitats.
+We hypothesize that tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, tree size and microenvironment, resulting in variable drought responses of species and individuals across droughts, and of species across microhabitats.
 Specifically, we test predictions that:
 i) both species and individuals respond differently to different droughts;
-ii) deciduous and evergreen species exhibit differential sensitivity depending on the timing of drought and topographic location,
-iii) the influences of topographic location, crown exposure, and other size effects on individual drought responses are variable across droughts.
+ii) on the species level, deciduous and evergreen species exhibit differential sensitivity depending on the timing of drought and topographic location;
+iii) on the individual level, the influences of topographic location, crown exposure, and other size effects on drought responses are variable across droughts.
 We tested these hypotheses using  a 14-year record of dendrometer band measurements in a dry seasonal evergreen forest [@bunyavejchewin_seasonally_2011] in Thailand. 
@@
 <!---CONCLUSIONS---->
-We show that, in a tropical seasonal forest, the drought sensitivity of tree growth is jointly and interactively shaped by drought characteristics, species adaptations, and local environmental conditions (Table 1).
+We show that, in a tropical seasonal forest, the drought sensitivity of tree growth is jointly and interactively shaped by drought characteristics, species adaptations, tree size and microenvironment (Table 1).
 Analysing three droughts in a 14-year dendrometer band time series, we found lower-than-average growth in drought years, resulting in lower plot-level growth (Fig `r fig_drought_responses`). 
@@
 <!----Limitations in generalisation/ species examined here may have greater diversity of drought adaptaitons than elsewhere in the tropics---->
-Periodic hot/dry conditions in tropical dry forests could create selective pressures for adaptative variation towards drought resistance potnetially resulting in a greater diversity of drought responses here compared to other bioclimatic zones where adaptive pressures and thus ecological variability are different.
+Periodic hot/dry conditions in tropical dry forests could create selective pressures for adaptative variation towards drought resistance potentially resulting in a greater diversity of drought responses here compared to other bioclimatic zones where adaptive pressures and thus ecological variability are different.
 We demonstrate that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.

### 2025-06-27 doc/manuscript_hkk_drought_sensitivity.Rmd
fixed "wet-season" hyphenation

@@
 
-**Number of Words in Abstract:** 144  
-**Number of Words in Main Text:** 5019  
+**Number of Words in Abstract:** 146  
+**Number of Words in Main Text:** 5026  
 <!-- Words to subtract: 2628-->
@@
 In tropical seasonal forests in particular, trees have evolved to optimize growth, reproduction, and survival (i.e., fitness) under dry seasons, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
-Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry-season to avoid foliar water loss [@desouza_drought_2020]. 
-Other strategies include deep roots (hydraulic safety, to access deeper water in the dry-season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and large vessels (hydraulic efficiency)<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021].
+Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss [@desouza_drought_2020]. 
+Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and large vessels (hydraulic efficiency)<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021].
 Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021].
@@
 Our study site was the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
-This is a 50-hectare plot in dry seasonal evergreen forest, with mean annual precipitation of 1400 mm, mean annual temperature of 23.5 C and a strong dry-season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
+This is a 50-hectare plot in dry seasonal evergreen forest, with mean annual precipitation of 1400 mm, mean annual temperature of 23.5 C and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
 The region is periodically affected by droughts associated with the *El Niño* Southern Oscillation [ENSO, @räsänen_spatial_2016].
@@
 1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought. 
-In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet-season, including precipitation (anomaly **<-1.5** SD), number of dry days (max anomaly **>2** SD), maximum monthly temperatures (anomaly **+1.5** SD), and monthly maximum VPD (max anomaly **+2** SD).
+In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including precipitation (anomaly **<-1.5** SD), number of dry days (max anomaly **>2** SD), maximum monthly temperatures (anomaly **+1.5** SD), and monthly maximum VPD (max anomaly **+2** SD).
 <!-- 1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period.  -->
@@
 3- and 6-month SPEI values approached -2 in February, indicative of borderline severe drought.
-12-month SPEI showed severe drought conditions from January to April, and moderate drought from May to September, spanning both dry- and wet-seasons.
+12-month SPEI showed severe drought conditions from January to April, and moderate drought from May to September, spanning both dry and wet seasons.
 
@@
 Despite lower extremes of temperature, vapor pressure deficit, and number of dry days (Fig.
-2a), severe drought in the wet-season (2015) elicited stronger responses across trees, especially deciduous species, which are expected to have stronger seasonality in growth phenology (Fig `r fig_species`).
+2a), severe drought in the wet season (2015) elicited stronger responses across trees, especially deciduous species, which are expected to have stronger seasonality in growth phenology (Fig `r fig_species`).
 Interestingly, in 2015 and 2020 (but not 2010), evergreen and deciduous species had divergent responses, especially along a wetness gradient; deciduous species showed greater growth declines in wetter sites in 2015 while evergreen species grew less in drier sites in 2020 (Fig `r fig_species`b, Fig `r fig_microenvs`b).
@@
 A k-month SPEI integrates over k months ending in the month current month.
-Across all panels, light blue background represents the average wet-season, from May to October.
+Across all panels, light blue background represents the average wet season, from May to October.
 

### 2025-06-27 doc/manuscript_hkk_drought_sensitivity.Rmd
intro minor edits

@@
     bookdown::word_document2: 
-      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
+      #reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
+      reference_docx: word-styles-reference-singlespace.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
     #pdf_document:
@@
 **Number of Words in Abstract:** 146  
-**Number of Words in Main Text:** 5026  
-<!-- Words to subtract: 2628-->
-**Number of References:** 54  
+**Number of Words in Main Text:** 5048  
+**Number of References:** 55  
 **Number of Figures:** 5  
@@
 
-# Abstract (150 words) {-}
+# Abstract {-}
 Increased frequency and severity of droughts threatens forest health worldwide<!--, with important implications for this critical carbon sink-->. 
@@
 For forests worldwide, studies have separately shown that drought characteristics, species traits, tree size and microenvironment affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter understanding of forest responses to drought.
-Specifically, if these factors act independently, tree drought responses to may be conserved, and linearly predictable; i.e., species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought.
-In contrast, if these factors interact, species and individuals that exhibit resistance to one drought in one particular environment may prove vulnerable to a drought with different characteristics or in a different environment.
+Specifically, if these factors act independently, tree drought responses may be conserved, and linearly predictable; i.e., species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought.
+In contrast, if these factors interact, species and individuals that exhibit resistance to one drought may prove vulnerable to a drought with different characteristics. 
 The complexity and diversity of drought responses may act as a mechanism for maintaining species diversity and buffering ecosystem function even as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].
@@
 Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and large vessels (hydraulic efficiency)<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021].
-Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021].
+Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought vulnerability, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021].
 <!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.
@@
 While habitats with relatively moist soils (e.g., low-lying topography) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted [e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021].
-Indeed, during drought trees near streams experienced greater growth declines in a temperate forest [@mcgregor_tree_2021] and increased mortality in a tropical forest [@zuleta_droughtinduced_2017]. 
+Indeed, during drought, trees near streams experienced greater growth declines in a temperate forest [@mcgregor_tree_2021] and increased mortality in a tropical forest [@zuleta_droughtinduced_2017]. 
 <!-- Across vertical axes, trees of different sizes have their crowns and root systems in sometimes dramatically divergent microenvironments. -->
@@
 <!--In contrast, understory trees may be rarely exposed to direct solar radiation and extreme evaporative demand.-->
-These aboveground gradients in crown exposure could be partially offset by the tendency for large trees to have larger/deeper root systems and greater stem water storage [@fernández-de-uña_role_2023] and therefore the potential to access more reliable water sources [@stahl_depth_2013].
+Tree size can mediate microenvironments by influencing both crown exposure through tree height and belowground water access through rooting depth/size. 
+For large trees, aboveground gradients in crown exposure could be partially offset by the tendency to have larger/deeper root systems and greater stem water storage [@fernández-de-uña_role_2023] and therefore the potential to access more reliable water sources [@stahl_depth_2013].
 <!-- Sean::: I need to look at this source.
@@
 However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
-While the overall tendency is that larger trees suffer greater growth declines during drought [@bennett_larger_2015], this may be driven primarily by crown exposure, by greater height itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium <!--, as described by Darcy's law -->[@mcdowell_darcys_2015; @fernández-de-uña_role_2023], or by trait differences across vertical axis [@vinod_thermal_2023].
-Thus, horizontal and vertical microenvironmental gradients modify tree drought responses, likely in interaction with drought characteristics and species traits.
+While larger trees tend to suffer greater growth declines during drought [@bennett_larger_2015], this may be driven by crown exposure, by height <!--itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium --><!--, as described by Darcy's law -->and its pressures on stem water flow[@mcdowell_darcys_2015; @fernández-de-uña_role_2023], or by trait differences across vertical axis [@vinod_thermal_2023].
+Thus, horizontal and vertical microenvironmental gradients and tree size modify tree drought responses, likely in interaction with drought characteristics and species traits.
 
@@
 During this period, the forest experienced three different dry periods, two associated with ENSO events, and all three differing with respect to timing and duration (Table 1). 
-With data on 1820 trees across 30 species we built Bayesian regression models exploring how species characteristics and individual spatial context influenced growth responses to these droughts. 
+With data on 1820 trees across 30 species, we built Bayesian regression models exploring how species characteristics and individual spatial context influenced growth responses to these droughts. 
 We used a causal modelling framework to account for potential confounding relationships among the variables. 
-This analysis adds important insight into the complexity of seasonal dry forest response to drought in species-rich tropical forests.
+This analysis adds important insight into the complexity of drought responses in species-rich seasonal dry tropical forests.
 

### 2025-06-27 doc/manuscript_hkk_drought_sensitivity.Rmd
results minor edits

@@
 
-Our study site was the Huai Kha Khaeng ForestGEO plot (15.6324 N, 99.217 E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
-This is a 50-hectare plot in dry seasonal evergreen forest, with mean annual precipitation of 1400 mm, mean annual temperature of 23.5 C and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
+Our study site was the Huai Kha Khaeng ForestGEO plot (15.63° N, 99.22° E), hereafter "HKK", in the Huai Kha Khaeng Wildlife Sanctuary, Uthai Thani, Thailand.
+This is a 50-hectare plot in dry seasonal evergreen forest, with mean annual precipitation of 1400 mm, mean annual temperature of 23.5° C and a strong dry season from November to April [@anderson-teixeira_ctfsforestgeo_2015].
 The region is periodically affected by droughts associated with the *El Niño* Southern Oscillation [ENSO, @räsänen_spatial_2016].
@@
 -->
-We first conducted quality control on window size measurements (see supplementary methods) and combined them with the diameter measurement at installation $(d_{1})$ to calculate diameter at each dendroband census using standard equations that uses the chord across the measurement window to estimate the arc[@detto_correcting_2023].
-We used a direct measurement of $d_{1}$ at the time of installation if it was available; if not, we excluded window size measurements until the first available diameter measurement. 
-We calculated annualised diameter increments for each individual $i$ for each year $t$ by finding the difference in diameter from the previous year ($inc_{i,t} = d_{i,t} - d_{i,t-1}$) and adjusting for the number of days ($tdif$) between measurements ($ainc_{i, t} = inc_{i,t} * \frac{tdif}{365}$).
+We first conducted quality control on window size measurements (see supplementary methods) and combined them with the diameter measurement at installation $(dia_{1})$ to calculate diameter at each dendroband census for each tree $(d_{i, t})$ using standard equations that uses the chord across the measurement window to estimate the arc[@detto_correcting_2023].
+We used a direct measurement of $dia_{1}$ at the time of installation if it was available; if not, we excluded window size measurements until the first available diameter measurement. 
+We calculated annualised diameter increments for each individual $i$ for each year $t$ by finding the difference in diameter from the previous year ($inc_{i,t} = d_{i,t} - d_{i,t-1}$) and adjusting for the number of days ($\Delta day$) between measurements ($ainc_{i, t} = inc_{i,t} * \frac{\Delta day}{365}$).
 
@@
 We conducted further quality control on increments (see supplementary methods for details) to assemble the final dataset. 
-Diameter tape measurements are not precise enough for calculating drought sensitivity. 
+Diameter tape measurements $dia_{i, t}$ are not precise enough for calculating drought sensitivity. 
 Error for growth is twice measurement error = 2 mm; in comparison, median annual growth in HKK is ~1.4 mm and calculating deviation from annual growth needs sub-mm precision. 
@@
 We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 30-day periods across the whole timeseries.
-Finally, we calculated daily rolling anomalies for each variable for the drought years from the long-term daily expectation as $$anomaly_{day} = \frac{rolling.mean_{day} - \bar{rolling.mean_{day}}}{sd(rolling.mean_{day})}$$
+Finally, we calculated daily rolling anomalies from the long-term daily expectation for each variable for the drought years as $$anomaly_{day} = \frac{rolling.mean_{day} - \bar{rolling.mean_{day}}}{sd(rolling.mean_{day})}$$
 
@@
 We first  calculated the mean annual increment $\bar{a.inc_{i}}$ for each tree from annualised increment $ainc_{i,t}$.
-We then calculated sensitivity for each tree $i$ for each drought year $y$ as 
+We then calculated growth sensitivity for each tree $i$ for each drought year $y$ as 
 $$
@@
 Across all individuals, we calculated Topographic Wetness Index (TWI) for an expectation of the water availability across space based on the total upslope area and the slope at the location. 
-We used a Digital Elevation Model at 5 m spatial grain for the area of the plot for fine resolution and ability to locate the trees on the layer, but this omits upslope area that is off the plot. 
-Using the *whitebox* package in R [@wu_whitebox_2022; @lindsay_whitebox_2016a] we calculated flow accumulation at each cell and then calculate TWI as $log_{e}\frac{As}{tan(slope)}$ where $As$ is the total upslope area within the plot and $slope$ is the absolute value of slope calculated at the 5m grain.
+We used a Digital Elevation Model at 5 m spatial grain for the area of the plot and the *whitebox* package in R [@wu_whitebox_2022; @lindsay_whitebox_2016a] to calculate flow accumulation at each cell and then calculate TWI as $log_{e}\frac{As}{tan(slope)}$ where $As$ is the total upslope area within the plot and $slope$ is the absolute value of slope calculated at the 5m grain.
 <!-- As an alternate measure of water availability, we also calculated Topographic Position Index (TPI) based on the concavity/convexity of a location.  -->
-For an alternate estimate of wetness (Topographic Position Index) and associated results, see supplementary material. 
-We defined tree size as the diameter at breast height (calculated from dendrometer band window measurement) at the previous census. 
+This geolocated layer allows fine resolution and precise tree mapping with topography, but this omits upslope area beyond the plot.
+For an alternate estimate of wetness independent of total upslope area (Topographic Position Index) and associated results, see supplementary material. 
+We defined tree size as the diameter at breast height (calculated from dendrometer bands) at the previous census. 
 We centred and scaled TWI and DBH across the range of observed values. 
-We estimated crown exposure for each tree and year using ground measurements of crown illumination index (CII), an ordinal variable from 1 to 5 (least to most exposed).
+We estimated crown exposure for each tree and year using ground measurements of crown illumination index (CII) at the previous census, an ordinal variable from 1 to 5 (least to most exposed).
 
@@
 
-We analysed tree responses to individual droughts by fitting separate models. 
-We fit alternative models for sensitivity including a null model with just species random effects, a model including only fixed effects of species deciduousness and habitat wetness and a model including size (DBH), light environment (CII) and habitat wetness as well as species random effects on all these.
+We analysed tree drought responses by fitting separate models for each drought. 
+We fit alternative models for sensitivity including a null model with just species random effects, a model including only fixed effects of species deciduousness and habitat wetness (TWI) and a model including size (DBH), light environment (CII) and TWI as well as species random effects on all these.
 We first modelled sensitivity as a simple random intercept model
@@
 
-Simultaneously modelling the effect of variables on sensitivity and the effect of DBH on CII in this way allows us to disentangle causual linkages between these processes. 
+Simultaneously modelling the effect of variables on sensitivity and the effect of DBH on CII in this way allows us to disentangle causual linkages while accounting for confounding effects. 
 We chose to model a species random effect on all slopes here to account for variation across species instead of specific traits which may have different effects with each of the variables. 
 However, we also ran simpler models with species random effects only on the intercept, the results of which are reported in the Supplementary Material. 
-Across all models, we used Gaussian priors for the distribution of the response variable sensitivity, considered CII as an monotonic predictor (ordered factor) and centred and scaled the other predictors.
-We modelled CII as an ordinal categorical variable with a cumulative logit link function. 
+Across all models, we used Gaussian priors for the distribution of the response variable sensitivity, considered CII as a monotonic predictor (ordered factor) and centred and scaled the other predictors.
+We modelled CII response as an ordinal categorical variable with a cumulative logit link function. 
 In the combined form, the model therefore used a multivariate distribution of Gaussian and logit. 
@@
 All statistical analyses were performed using R version 4.4.0 [@rcoreteam_language_2024]. 
-Mixed models were fit through Bayesian regressions using MCMC methods using the package _brms_ [@bürkner_brms_2017].
+Mixed models were fit through Bayesian regressions using MCMC methods from the package _brms_ [@bürkner_brms_2017].
 We ran 4 MCMC chains for 3000 iterations with 1000 of these set as warmup.
@@
 <!-- Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions. -->
-1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought. 
+1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of drought, while the 12-month values remained above -1, indicative of mild drought. 
 In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including precipitation (anomaly **<-1.5** SD), number of dry days (max anomaly **>2** SD), maximum monthly temperatures (anomaly **+1.5** SD), and monthly maximum VPD (max anomaly **+2** SD).
@@
 <!-- The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a).  -->
-Drought years were associated with lower plot-level growth and stronger tree-level sensitivities, with the strongest negative response in 2015 (Fig `r fig_drought_responses`). 
+Drought years were associated with lower plot-level growth and negative tree-level sensitivities, with the strongest negative response in 2015 (Fig `r fig_drought_responses`). 
 Median plot-level annual growth rate across years of measurement was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual*10, na.rm=T),2)` mm. 
@@
 
-Median annual increment varied among species, with *`r tree.time.sp$spfull[1]`* the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm) among the `r nrow(tree.time.sp)` species analysed.
+Median annual increment across yeaes varied among the `r nrow(tree.time.sp)` species analysed, with *`r tree.time.sp$spfull[1]`* the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm).
 Median species responses to the 2010 and 2020 droughts were variable, but no species had higher median increments than a typical year in 2015.
@@
 
-Supporting prediction *i*, species median sensitivities to the three droughts were not correlated with each other, although on average, species had more negative sensitivities to the 2015 drought than the 2010 or 2020 droughts, in alignment with median sensitivities across all trees (for 2010 and 2015 Pearson's correlation coefficient r = `r round(sens.sp.cor_10_15[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_15[3]$p.value,2)`, for 2015 and 2020 was r = `r round(sens.sp.cor_15_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_15_20[3]$p.value,2)`, and for  2010 and 2020 r = `r round(sens.sp.cor_10_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_20[3]$p.value,2)`, Fig S5).
+Species median sensitivities to the three droughts were not correlated with each other, supporting prediction *i*. 
+However on average, species had more negative sensitivities to the 2015 drought than the 2010 or 2020 droughts, in alignment with median sensitivities across all trees (for 2010 and 2015 Pearson's correlation coefficient r = `r round(sens.sp.cor_10_15[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_15[3]$p.value,2)`, for 2015 and 2020, r = `r round(sens.sp.cor_15_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_15_20[3]$p.value,2)`, and for 2010 and 2020, r = `r round(sens.sp.cor_10_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_20[3]$p.value,2)`, Fig S5).
 
@@
 <!-- Deciduousness species responded more negatively than evergreen species in 2015, but not in 2010 (Fig `r fig_species`a).  -->
-Supporting prediction *ii*, deciduous and evergreen species showed heterogenous responses to different droughts.
+Deciduous and evergreen species showed heterogenous responses to different droughts, supporting prediction *ii*.
 Deciduous species were more drought sensitive than evergreen species in 2015 (Fig `r fig_species`a, r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`), but the opposite was true in 2020 (`r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2020], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2020], 2)`).
@@
 
-Supporting prediction *i*, individual trees did not respond similarly to the three drought events (Fig S6).
+Individual trees did not respond similarly to the three drought events, supporting prediction *i* (Fig S6).
 
-Supporting prediction *iii*, TWI had different effects across droughts. 
+TWI had different effects across droughts, supporting prediction *iii*. 
 Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`) and 2020 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2020],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2020],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2020],2)`).
@@
 
-The differing effects of TWI on sensitivity in the three droughts, along with the influence of deciduousness, were confirmed by models that controlled for crown exposure, DBH and species variation (Fig `r fig_microenvs`a).
+<!-- The differing effects of TWI on sensitivity in the three droughts, along with the influence of deciduousness, were confirmed by models that controlled for crown exposure, DBH and species variation (Fig `r fig_microenvs`a).
+-->
+Broadly, the effects of wetness and size were different across droughts while exposure tended to decrease growth with varying strengths across droughts (Fig `r fig_microenvs`a). 
 Wetness had a small negative effect in 2010 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but a positive effect in 2015 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`) and 2020 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2020],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2020],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2020],2)`).
@@
 -->
-Predictions across the range of observed variables also showed that deciduous species experienced greater growth declines in 2015 across the wetness gradient (Fig `r fig_microenvs`c). 
-<!-- Mean sensitivities across species and their response to TWI were not associated with deciduousness in 2010 or 2020, mirroring the response in the interaction model.   -->
+<!-- Predictions across the range of observed variables also showed that deciduous species experienced greater growth declines in 2015 across the wetness gradient (Fig `r fig_microenvs`c).  -->
+Deciduous species experiences greater growth declines in  wetter sites in 2015, while these interactions were less clear in 2010 and 2020 (Fig `r fig_microenvs`c).
 
@@
 
-Supporting prediction *iii*, drought sensitivities of individuals with different crown exposures varied across droughts (Fig `r fig_dag`).
+Drought sensitivities of individuals with different crown exposures varied across droughts (Fig `r fig_dag`), supporting prediction *iii*.
 Crown exposure was associated with greater growth declines in 2010.
-In the same models with species random effects on intercept and slope, the simplex parameter for the ordered factor crown exposure in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 and 2020 the negative effect was not as strong (2015 median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`; 2020 median effect = `r round(coefs_spre_cii$median[coefs_spre_cii$yr==2020],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2020],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2020],2)`).
+In the same models describd above, the simplex parameter for the ordered factor crown exposure in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 and 2020 the negative effect was not as strong (2015 median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`; 2020 median effect = `r round(coefs_spre_cii$median[coefs_spre_cii$yr==2020],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2020],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2020],2)`).
 Predicted effects of crown exposure on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all crown exposure categories had negative sensitivities in 2015 and 2020 (Fig `r fig_dag`b).
 
-<!--DBH was a significant predictor of crown exposure, -->Trees with larger DBH had higher crown exposure in 2010 (median effect = `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`), 2015 (`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) and 2020 (`r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2020],2)`, 90% CI `r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2020],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2020],2)`) (Fig 5a).
+<!--DBH was a significant predictor of crown exposure, -->Trees with larger DBH had higher crown exposure across all years; median effect and 90% CI for 2010, 2015 and 2020 respectively were `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)` (`r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`), `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)` (`r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) and `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2020],2)`(`r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2020],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2020],2)`) (Fig 5a).
 Controlling for the direct effect of exposure, the remaining effect of DBH *per se* was significantly positive in 2010, and not significant in 2015 or 2020 (Fig `r fig_microenvs`a, Fig `r fig_dag`a).

### 2025-06-27 doc/manuscript_hkk_drought_sensitivity.Rmd
updated text with minor edits and word count

@@
     bookdown::word_document2: 
-      #reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
-      reference_docx: word-styles-reference-singlespace.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
+      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
+      #reference_docx: word-styles-reference-singlespace.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
     #pdf_document:
@@
 **Number of Words in Abstract:** 146  
-**Number of Words in Main Text:** 5048  
+**Number of Words in Main Text:** 5028  
 **Number of References:** 55  
@@
 USA  
-Phone: 
+Phone: +1 (202) 633-1836  
 Email: anujank@si.edu
@@
 We show that, in a tropical seasonal forest, the drought sensitivity of tree growth is jointly and interactively shaped by drought characteristics, species adaptations, tree size and microenvironment (Table 1).
-Analysing three droughts in a 14-year dendrometer band time series, we found lower-than-average growth in drought years, resulting in lower plot-level growth (Fig `r fig_drought_responses`). 
-Across droughts, there was little consistency in species- or individual-level drought responses, reflecting dissimilar effects of a key species trait (deciduousness) and variables linked to water access (TWI, DBH) under differing drought conditions.
+Analysing three droughts in a 14-year dendrometer band time series, we found lower-than-average growth during drought years, resulting in lower plot-level growth (Fig `r fig_drought_responses`). 
+Across droughts, there was little consistency in species- or individual-level drought responses, reflecting dissimilar effects of a key species trait (deciduousness) and variables linked to tree drought experience (TWI, DBH, CII) under differing drought conditions.
 Despite lower extremes of temperature, vapor pressure deficit, and number of dry days (Fig.
@@
 Interestingly, in 2015 and 2020 (but not 2010), evergreen and deciduous species had divergent responses, especially along a wetness gradient; deciduous species showed greater growth declines in wetter sites in 2015 while evergreen species grew less in drier sites in 2020 (Fig `r fig_species`b, Fig `r fig_microenvs`b).
-Although trees with more exposed crowns were consistently more sensitive to drought, additional effects of DBH -- presumably including rooting volume and depth -- were dissimilar between the drought events (Fig `r fig_microenvs`a).
+Although trees with more exposed crowns tended to grow less under drought, additional effects of DBH -- presumably including rooting volume and depth -- were dissimilar between the drought events (Fig `r fig_microenvs`a).
 It is possible that water availability buffered exposure-associated growth declines through different mechanisms in the different droughts; larger trees fared better in the dry-season drought while trees in wetter sites suffered smaller growth declines during wet-season droughts (Fig `r fig_dag`).
@@
 Taken together, these results indicate that combined effects of drought characteristics, species traits, and individual microenvironments produce important heterogeneity in individual and species drought responses of tropical trees.
-Divergent effects of species and individuals to droughts have  may contribute to the maintenance of species diversity and ecosystem stability.
+Divergent effects of species and individuals to droughts may contribute to the maintenance of species diversity and ecosystem stability.
 
@@
 We demonstrate that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.
-Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies resulting in diversification along the hydraulic safety-efficiency axis with diverse strategies to maximise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021].
+Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies resulting in diversification along the hydraulic safety-efficiency axis with diverse strategies to optimise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021].
 The degree of adaptive variation along the water-use axis might be comparable to other dry climatic systems (e.g.
@@
 Trees with more exposed crowns exhibited greater growth declines, while sub-canopy trees grew more than average during the 2010 drought.
-This finding is consistent with the concept that larger trees tend to suffer more during drought [@bennett_larger_2015], but adds the important insight that it is crown position -- as opposed to size *per se* -- that is primarily responsible for this pattern.
-The consistent negative effect of crown position aligns with pronounced gradients in drought stress across forest vertical profiles [@vinod_thermal_2023].
+This finding is consistent with observations that larger trees tend to suffer more during drought [@bennett_larger_2015], but adds the important insight that it is crown position -- as opposed to size *per se* -- that is primarily responsible for this pattern.
+The consistent negative direction of the effect of crown position aligns with pronounced gradients in drought stress across forest vertical profiles [@vinod_thermal_2023].
 We show that residual effects of size, presumably including root water access and biophysical challenges linked to height itself, may act in different directions depending on the drought characteristics (Figs.
@@
 -->
-Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in three different droughts, showing that ecological and evolutionary strategies make particular species and trees resistant to some droughts but vulnerable to others.
-These distinct responses of species and individuals to droughts suggests potential spatiotemporal insurance effects among tropical tree response, in alignment with recent global analyses [@liu_drought_2022].
+Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in three different droughts, showing that ecological and evolutionary strategies can make particular species and trees resistant to some droughts but vulnerable to others.
+These distinct responses of species and individuals to droughts suggests potential spatiotemporal insurance effects among tropical tree responses, in alignment with recent global analyses [@liu_drought_2022].
 

### 2025-06-27 scripts/06_figures.R
updated SI file with one more figure

@@ -584,7 +584,7 @@ speiplot_line <- ggplot() +
         # scales = "free_y",
         strip.position = "left", nrow = 4, labeller = varnames
     ) +
-    ylim(-3, 1.5)+
+    ylim(-3, 1.5) +
     theme_bw() +
     labs(x = "Month", y = "SPEI", fill = "Year") +
     scale_x_continuous(breaks = 1:12) +
@@ -2104,6 +2104,74 @@ png("doc/display/Fig4_tpi_3yrs.png", width = 8, height = 8, units = "in", res = 
 coefs_all_sp + pred_plot + plot_layout(design = layout, guides = "collect") + plot_annotation(tag_levels = "a") & theme(legend.position = "bottom", legend.text = element_text(size = 18), legend.title = element_text(size = 18))
 dev.off()
 
+# SI figure with random intercept only----------------
+
+coefs <- readRDS("results/models/orderedcii/coefs_tree.rds")
+coefs_df <- do.call(rbind, coefs)
+
+# par_names <- as_labeller(c("b_calcDBH_min1_scaled_sp" = "DBH effect", "b_cii_min1_scaled_sp" = "CII effect", "b_twi_scaled_sp" = "TWI effect"))
+
+par_names <- as_labeller(c("b_sensprop_calcDBH_min1_scaled" = "DBH effect", "bsp_sensprop_mocii_min1" = "exposure effect", "b_sensprop_twi_scaled" = "wetness effect"))
+
+coefs_all_sp <- ggplot(
+    data = coefs_df %>% filter(param %in% c("b_sensprop_calcDBH_min1_scaled", "bsp_sensprop_mocii_min1", "b_sensprop_twi_scaled")),
+    aes(
+        x = factor(param, levels = c("b_sensprop_calcDBH_min1_scaled", "bsp_sensprop_mocii_min1", "b_sensprop_twi_scaled")),
+        y = median,
+        ymin = lwr, ymax = upr
+        # col = factor(param, levels = c("b_calcDBH_min1_scaled_sp", "b_twi_scaled_sp", "b_cii_min1_scaled_sp"))
+    )
+) +
+    geom_pointrange() +
+    scale_x_discrete(labels = par_names) +
+    geom_hline(yintercept = 0, linetype = "dashed") +
+    facet_grid(~ factor(yr, levels = c(2010, 2015, 2020)), scales = "free") +
+    labs(title = "", x = "", y = "coefficient") +
+    guides(color = "none") +
[…]

### 2025-06-27 scripts/09_orderedcii.R

@@ -457,3 +457,44 @@ saveRDS(fits, file = paste0(models_dir, "/fits_rel_spre.rds"))
 saveRDS(coefs, file = paste0(models_dir, "/coefs_rel_spre.rds"))
 saveRDS(pred_sens, file = paste0(models_dir, "/pred_sens_rel_spre.rds"))
 saveRDS(pred_cii, file = paste0(models_dir, "/pred_cii_rel_spre.rds"))
+
+# Model 2: species random effect on intercept only ---------------------------------
+
+tree_model <- bf(sens.prop ~ 1 + calcDBH_min1_scaled + mo(cii_min1) + twi_scaled + (1 + calcDBH_min1_scaled + twi_scaled + mo(cii_min1) | Species))
+cii_model <- bf(ordered(cii_min1) ~ calcDBH_min1_scaled, family = cumulative(link = "logit"))
+
+# run the model
+coefs <- list()
+pred_sens <- list()
+pred_cii <- list()
+fits <- list()
+
+for (i in 1:length(yrs)) {
+    fit <- brm(tree_model + cii_model + set_rescor(FALSE),
+        data = tree.time %>% filter(yr == yrs[i]), family = gaussian(),
+        chains = 4, iter = 4000, warmup = 2000, cores = 4,
+        control = list(adapt_delta = 0.9)
+    )
+    fits[[i]] <- fit
+    post <- posterior_samples(fit)
+    post_sum <- as.data.frame(t(apply(post, 2, quantile, probs = c(.5, .05, .95))))
+    colnames(post_sum) <- c("median", "lwr", "upr")
+    post_sum$param <- rownames(post_sum)
+    post_sum$yr <- rep(yrs[i], nrow(post_sum))
+    coefs[[i]] <- post_sum
+
+    # make predictions
+    preds <- posterior_predict(fit)
+    # this makes a dataframe with 4000 rows (chains* sampling iterations) and 1449 columns (number of trees)
+    pred_sens_df <- as.data.frame(t(apply(preds[, , 1], 2, quantile, probs = c(.5, .05, .95))))
+    pred_cii_df <- as.data.frame(t(apply(preds[, , 2], 2, quantile, probs = c(.5, .05, .95))))
+    colnames(pred_sens_df) <- colnames(pred_cii_df) <- c("median", "lwr", "upr")
+    # add this to the observation data
[…]

### 2025-06-27 doc/manuscript_hkk_drought_sensitivity.Rmd
addressing #34

@@
 **Number of Words in Abstract:** 146  
-**Number of Words in Main Text:** 5028  
+**Number of Words in Main Text:** 5036  
 **Number of References:** 55  
@@
 Despite lower extremes of temperature, vapor pressure deficit, and number of dry days (Fig.
-2a), severe drought in the wet season (2015) elicited stronger responses across trees, especially deciduous species, which are expected to have stronger seasonality in growth phenology (Fig `r fig_species`).
+2a), the severe drought in 2015 that peaked in the wet season elicited stronger responses across trees, especially deciduous species, which are expected to have stronger seasonality in growth phenology (Fig `r fig_species`).
 Interestingly, in 2015 and 2020 (but not 2010), evergreen and deciduous species had divergent responses, especially along a wetness gradient; deciduous species showed greater growth declines in wetter sites in 2015 while evergreen species grew less in drier sites in 2020 (Fig `r fig_species`b, Fig `r fig_microenvs`b).
 Although trees with more exposed crowns tended to grow less under drought, additional effects of DBH -- presumably including rooting volume and depth -- were dissimilar between the drought events (Fig `r fig_microenvs`a).
-It is possible that water availability buffered exposure-associated growth declines through different mechanisms in the different droughts; larger trees fared better in the dry-season drought while trees in wetter sites suffered smaller growth declines during wet-season droughts (Fig `r fig_dag`).
+It is possible that water availability buffered exposure-associated growth declines through different mechanisms in the different droughts; larger trees fared better in the dry-season drought while trees in wetter sites suffered smaller growth declines during droughts that extended into the wet season (Fig `r fig_dag`).
 <!-- we can talk about it ... but I don't think that we can know that larger trees have deeper roots.  -->
@@
 -->
-However, our finding that a wet-season drought in 2015 had greater impact on woody growth than dry-season droughts in 2010 and 2020 questions the generality of previous findings. 
+However, our finding that a severe wet-season drought in 2015 had greater impact on woody growth than dry-season drought in 2010 and dry-to-wet-season drought in 2020 questions the generality of previous findings.
 Tree ring studies from this site [@vlam_temperature_2014;@anderson-teixeira_joint_2022] and across the tropics [@zuidema_tropical_2022] have asserted that tree growth is most sensitive to dry-season precipitation.

### 2025-06-28 scripts/06_figures.R
updated SI figure

@@ -2168,9 +2168,14 @@ pred_plot
 
 library(patchwork)
 
+layout <- "
+AA
+BB
+"
 
-png("doc/display/FigSI_coefs_nospre.png", width = 8, height = 4, units = "in", res = 300)
-coefs_all_sp
+
+png("doc/display/FigSI_coefs_nospre.png", width = 8, height = 8, units = "in", res = 300)
+coefs_all_sp + pred_plot + plot_layout(design = layout, guides = "collect") + plot_annotation(tag_levels = "a") & theme(legend.position = "bottom", legend.text = element_text(size = 18), legend.title = element_text(size = 18))
 dev.off()
 
 

### 2025-06-28 scripts/09_orderedcii.R

@@ -460,7 +460,7 @@ saveRDS(pred_cii, file = paste0(models_dir, "/pred_cii_rel_spre.rds"))
 
 # Model 2: species random effect on intercept only ---------------------------------
 
-tree_model <- bf(sens.prop ~ 1 + calcDBH_min1_scaled + mo(cii_min1) + twi_scaled + (1 + calcDBH_min1_scaled + twi_scaled + mo(cii_min1) | Species))
+tree_model <- bf(sens.prop ~ 1 + calcDBH_min1_scaled + mo(cii_min1) + twi_scaled + (1 | Species))
 cii_model <- bf(ordered(cii_min1) ~ calcDBH_min1_scaled, family = cumulative(link = "logit"))
 
 # run the model

### 2025-08-22 doc/response_to_reviewer_R1.Rmd
set up for revisions
added reviewer responses doc, renamed submitted docx as v1 to track changes later, reknit word doc.

---
title: ''
output: 
    bookdown::word_document2: 
    #   reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
      reference_docx: word-styles-reference-singlespace.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    #pdf_document:
    # engine: xelatex 
  #officedown::rdocx_document
bibliography: growth-precip-thailand.bib
# csl: new-phytologist.csl
csl: ecology-letters.csl 
---

Referees' comments to the author(s):

Referee: 1

Comments for the Authors
General comment

The response of tropical forests to drought is a central topic in global ecology and climate change, directly aligned with the interests of Ecology Letters. The novel of this work is the central argument that drought sensitivity is not a fixed species trait, but rather the result of the interaction between drought characteristics, adaptations, and the microenvironment, provides conceptual originality. The authors used a 14-year series of dendrometer bands in a big plot with 1,820 trees of 30 species, which is a high-value dataset. In addition, the statistical analysis based on Bayesian models and a causal framework to disentangle trait, size, and microenvironment effects appears to be a sufficiently robust approach to address predictions. However, I have some concerns regarding the with the current version that should be addressed by the authors, which are appointed below.

RESPONSE:

Main comments:

•	It is always helpful to know the study site. I recommend including some pictures of the forest and study area.

RESPONSE:

•	The methodology should clearly state whether the 30 species selected for the study represent the functional groups. It should include the number of species per Deciduousness group.

RESPONSE:

•	I think it is important to add the average of non-dry years in Figure 1a (without 2010, 2015 and 2020). This helps clarify the difference between a normal/rainy year and a dry year.

RESPONSE:

•	It confuses many acronyms. It's necessary to rewrite their description at least once in the results section, e.g. CII doesn't appear in this section.

RESPONSE:

•	Predictions (i–iii) are clearly formulated, but the Results section is at times overly dense with statistical detail. It would be helpful to present more explicitly which findings support each hypothesis.

RESPONSE:

•	276: This section is important, but it's not a relevant result; rather, it's a methodological aspect that allowed the study to be conducted. I think it should be included in the materials and methods section.

RESPONSE:

•	Since the study is based on only one site, it is relevant to improve the discussion section with other already published study sites (add more references), mainly link the global implications, such mortality rate and vulnerability in tropical forests. See https://urldefense.proofpoint.com/v2/url?u=https-3A__doi.org_10.1126_science.adq6607&d=DwIFaQ&c=009klHSCxuh5AI1vNQzSO0KGjl4nbi2Q0M1QLJX9BeE&r=6fH468T58at0TtM59sMf109A_dRgo6XTfGMOVX5Ewrw&m=fPE_4y1UPsTl9St_Y44IUO7VIaQJOJIUiHtK15-Nr7g4UiVKbtYQ-TE0043SLAj6&s=SxqVZgFkbqAZ0-VEvdQntXWpGEUu9lr25rgxQtwaVdw&e=

RESPONSE:

•	There was something missing to discuss about the species in particular, some that have been more affected than others and why, especially those in conservation categories.

RESPONSE:

•	It is necessary to discuss further why the results indicate a "spatiotemporal insurance", which ecophysiological mechanisms or processes are relevant in these tropical forests.

RESPONSE:

Minor comments:

55: Could you briefly explain the three climatically distinct droughts?

RESPONSE:

162-163: Until when was the measurement? It is important to add the last year of measurement.

RESPONSE:

239-242: I think that this extra information is unnecessary, since although this measurement is interesting, it was not the one applied in this study.

RESPONSE:

301-307: The figure that supports this paragraph needs to be cited. I would think it would be figure 2a, but the species Gluta obovata is not there. I did not see the analysis by dominant species anymore, if the study does not focus on these species, I recommend removing it from Figure 2 since it confuses the reader.

RESPONSE:

339-346: I had a hard time understanding this result, on the one hand it says that Crown exposure was associated with greater growth declines in 2010, but not for 2015 and 2020. However, according Fig 5b all all crown exposure categories had negative sensitivities in 2015 and 2020.

RESPONSE:

Table 1: “other size effect” is DBH?

RESPONSE:

Figure 1a: “Dy” is not correct

RESPONSE:

Fig. 2. It is unnecessary to include the years 2023 and 2024 on the y-axis, as they were not evaluated. I think this figure could be analyzed further. For example, it took two years after 2014 for growth to recover to previous year (for ALL species) unlike in 2010 and 2020, when it rebounded within a year to levels equal to or higher than the previous year.

RESPONSE:

Referee: 2

Comments for the Authors
This study investigated the impact of three different drought events on tropical tree radial growth based on the band dendrometer monitoring of 50 ha tropical seasonally dry plot in western Thailand. The study found that wet season drought in 2015 had the most negative impact on the tree growth compared with the other two drought events. In addition, the results also showed that the impact of deciduousness, soil availability and canopy exposure had significant impact on the growth response to drought. The present study is useful to understand the impact of drought events. The data is very strong and the manuscript is well written, the analysis were also well done. The present study is helpful to understand the impact of drought events on the growth of tropical trees. I had some minor comments hope can help to improve the manuscript.

RESPONSE:

1 Line 175 “Using the late wet/early dry dendrometer…” Since there were two time of censuses (January and June) for the measurement, can the author clarify which time do they use to calculate the annual growth? Based on the measurement of January or June?

RESPONSE:

2 The topographic wetness index. Are there any measured soil moisture data in this plot? Since this index is only based on topography, it will be good if the author could provide some measured soil moisture data to calibrate the prediction based on model prediction.

RESPONSE:

3 In the case of deciduousness, the authors indicated that they used the deciduous index from Williams et al. (2008). Since different species might shed leaves in different season. Did the authors consider the impact of the timing of deciduousness on the growth?

RESPONSE:

Fig. S1 What does the shade area indicate? The three different drought events?

RESPONSE:

Fig S6 The R Square is too low, so it is difficult to detect the linear relationship. The same issue for Fig. S11 and Fig. S12. For the non-significant correlation, it’s better use dashed lines instead of solid line.

RESPONSE:

Fig S7 “…included in the analyses in the 2010 and 2015 datasets”, do you mean “2010, 2015 and 2020”?

RESPONSE:

Fig. S8 …need to add X and Y axes for the figures.

RESPONSE:

******************************************
Editor's comments to the author(s):

Editor
Editors Comments for the Author(s):
Two reviewers have reviewed this study and see a lot of value and importance in the data and conclusions. They provide valuable comments that should be able to be addressed in a major revision.

RESPONSE:

### 2025-08-22 doc/manuscript_hkk_drought_sensitivity.Rmd
update manuscript with Zenodo doi
set up for revisions
added reviewer responses doc, renamed submitted docx as v1 to track changes later, reknit word doc.

@@
 KA, SMM, SJD and KAT conceptualised the analysis. 
-SB, SJD and NP have been leading long-term data collection at HKK, while HMM created and helped execute the dendrometer band protocol. 
-KA cleaned the data and conducted formal analysis with support from SMM, SJD, HMM and KAT. 
+SB, SJD and NP have been leading long-term data collection at HKK, while HCM created and helped execute the dendrometer band protocol. 
+KA cleaned the data and conducted formal analysis with support from SMM, SJD, HCM and KAT. 
 KA wrote the original draft of the manuscript with support from KAT and SMM. 
@@
 The climate data used in this work are publicly available. 
-The code used to analyse the data will be made available in Zenodo upon acceptance of the manuscript.
+The code used to analyse the data is archived in Zenodo with doi:10.5281/zenodo.15777974.
 

### 2025-08-22 doc/response_to_reviewer_R1.Rmd
Few comments addressed

@@
 
-RESPONSE:
+**RESPONSE: Thanks for this comment.
+We have now included the long-term mean trend along with the long-term standard deviation to for Fig 1a to show the deviation more clearly.
+Fig 1b now follows more closely, showing the deviation in SD units from the long-term mean.**
 
@@
 
-RESPONSE:
+**RESPONSE: We have now removed this statement regarding the alternate measure of deciduousness that was considered.**
 
@@
 
-RESPONSE:
+**RESPONSE: Thanks for noticing this error. We have fixed this to say "Day of year".**
 
@@
 
-RESPONSE:
+**RESPONSE: Yes, the shaded area indicates the three different drought years. We have made edited the text to reflect this.**
 
@@
 
-RESPONSE:
+**RESPONSE: Thanks for notiicing this error. We have changed this to "2010, 2015 and 2020".**
 

### 2025-08-22 doc/manuscript_hkk_drought_sensitivity.Rmd
Few comments addressed

@@
 Data were unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1. 
-We also considered an alternative measure of deciduousness - the duration of time spent at maximum crown loss.
-This was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for more species.  
+<!-- We also considered an alternative measure of deciduousness - the duration of time spent at maximum crown loss. -->
+<!-- This was strongly correlated with the proportion of crown loss; we chose to use the proportion of crown loss because it was available for more species.   -->
 <!-- At the species level, we also calculated the maximum size across all individuals in the plot as measured in the 2008 whole-plot census.

### 2025-08-22 doc/supplementary_information.Rmd
updated SI file with one more figure
Few comments addressed

@@
 ![**Figure S1:** Raw climate data timeseries from 2009 to 2023 from ERA5Land and CHIRPS.
-Measurements represent daily rolling means over a 30-day time window for precipitation, max temperature and VPD and rolling sum for number of dry days.](display/climvars_fullseries_remote.png)
+Measurements represent daily rolling means over a 30-day time window for precipitation, max temperature and VPD and rolling sum for number of dry days.
+The shaded areas indicate the three drought years analysed - 2010, 2015 and 2020.](display/climvars_fullseries_remote.png)
 
 ![**Figure S2:** Climate anomalies for all variables from 2009 to 2020.
-Values represent number of standard deviations from the long-term mean for daily rolling mean values.](display/climvars_anomalies_fullseries_remote.png)
+Values represent number of standard deviations from the long-term mean for daily rolling mean values.
+The shaded areas indicate the three drought years analysed - 2010, 2015 and 2020.](display/climvars_anomalies_fullseries_remote.png)
 
@@
 
-![**Figure S4:** Species sensitivities in drought years 2010 and 2015.
+![**Figure S4:** Species sensitivities in drought years 2010, 2015 and 2020.
 Boxplots represent 25th, 50th (median) and 75th percentile of sensitivity for each species in each year and whiskers represent 95% CIs.
@@
 
-![**Figure S7:** Map of trees with dendrobands in HKK, included in analyses in the 2010 and 2015 datasets](display/dendro_tree_map.png)
+![**Figure S7:** Map of trees with dendrobands in HKK, included in analyses in the 2010, 2015 and 2020 datasets](display/dendro_tree_map.png)
 
@@
 
+# Alternate model using species random effect only on intercept
+
+![**Figure S17:** Model results from Bayesian causal model of sensitivity as a function of DBH, CII and TWI with species random effect only on intercept.](display/FigSI_coefs_nospre.png)
+
 <!-- ## Individual model diagnostics

### 2025-08-22 scripts/05_figures.R
figure changes

@@ -653,20 +653,22 @@ varnames <- as_labeller(c(
 
 # plot values
 
-climsat_plot <- ggplot(climsat_rlmean %>% filter(year %in% c("2010", "2015", "2020", "long-term")), col = year) +
+climsat_plot <- ggplot(climsat_rlmean %>% filter(year %in% c("2010", "2015", "2020")), col = year) +
     geom_rect(
         data = data.frame(xmin = 121, xmax = 304, ymin = -Inf, ymax = Inf),
         aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), fill = "lightblue", alpha = 0.3
     ) +
     # geom_bar(aes(x = month, y = value, fill = factor(year)), position = "dodge", stat = "identity") +
     geom_line(aes(x = doy, y = rlval, color = factor(year)), linewidth = 0.8) +
+    geom_line(aes(x = doy, y = long.term), col = "grey40") +
+    geom_ribbon(aes(x = doy, ymin = long.term - long.term.sd, ymax = long.term + long.term.sd), col = "grey60", alpha = 0.5) +
     facet_wrap(~climvar,
         scales = "free_y", ncol = 1,
         strip.position = "left",
         labeller = varnames
     ) +
     theme_bw() +
-    labs(x = "Dy of year", y = "", color = "Year") +
+    labs(x = "Day of year", y = "", color = "Year") +
     # scale_x_continuous(breaks = 1:12) +
     guides(linetype = "none", color = "none") +
     ggtitle("Climate variables") +

### 2025-08-26 scripts/05_figures.R
updated Fig 1 #37 #10

@@ -660,8 +660,8 @@ climsat_plot <- ggplot(climsat_rlmean %>% filter(year %in% c("2010", "2015", "20
     ) +
     # geom_bar(aes(x = month, y = value, fill = factor(year)), position = "dodge", stat = "identity") +
     geom_line(aes(x = doy, y = rlval, color = factor(year)), linewidth = 0.8) +
-    geom_line(aes(x = doy, y = long.term), col = "grey40") +
-    geom_ribbon(aes(x = doy, ymin = long.term - long.term.sd, ymax = long.term + long.term.sd), col = "grey60", alpha = 0.5) +
+    geom_line(aes(x = doy, y = long.term), col = "grey40", lty = "dashed") +
+    geom_ribbon(aes(x = doy, ymin = long.term - long.term.sd, ymax = long.term + long.term.sd), col = "grey60", alpha = 0.2) +
     facet_wrap(~climvar,
         scales = "free_y", ncol = 1,
         strip.position = "left",

### 2025-09-02 doc/response_to_reviewer_R1.Rmd
updated Fig 1 #37 #10
updates #40

@@
 
-RESPONSE:
+**RESPONSE: Thanks for this comment.
+We agree that it is useful to include the functional representativeness of the species.
+As we mention in the methods, our metric of deciduousness is a continuous variable and not categorical.
+And so, reporting the number per deciduousness group would require arbitrarily discretising this continuous variable.
+Instead, we now report a supplementary figure showing the distribution of deciduousness among the selected species.**
 
@@
 
-RESPONSE:
+**RESPONSE: Thanks for this comment.
+We have revised Figure 2 to remove species lines.
+Originally, we had chosen to include the top 10 species in the plot to illustrate both synchrony and divergence in species responses, but the results text reported on growth and sensitivity values across all species considered (n=30), shown in the Supplementary Appendix.
+We agree that this misalignment in species choice was confusing and have removed it from Figure 2 and referenced the relevant figure in the results text.**
 
@@
 
-RESPONSE:
+***RESPONSE: Since all crown exposure categories had negative growth sensitivities in 2015 and 2020, crown exposure was not associated with large differences in the magnitude or direction of sensitivity.
+In other words, more exposed trees had more or less similar growth declines compared to less exposed/understory trees.
+TO CLARIFY IN TEXT***
 
@@
 
-**RESPONSE: Yes, the shaded area indicates the three different drought years. We have made edited the text to reflect this.**
+**RESPONSE: Yes, the shaded area indicates the three different drought years. We have edited the text to reflect this.**
 
@@
 
-**RESPONSE: Thanks for notiicing this error. We have changed this to "2010, 2015 and 2020".**
+**RESPONSE: Thanks for noticing this error. We have changed this to "2010, 2015 and 2020".**
 
@@
 
-RESPONSE:
+**RESPONSE: Thanks for your favourable review of the manuscript.
+We have now addressed all the reviewer comments, which have improved the manuscript.
+We hope that the manuscript can be considered for publication in Ecology Letters.**

### 2025-09-02 scripts/05_figures.R

@@ -1641,6 +1641,33 @@ png("doc/display/Fig3.png", width = 8, height = 8, units = "in", res = 300)
 dec_intercept_plot / iso_plot_twi + plot_annotation(tag_levels = "a") + plot_layout(heights = c(1.8, 1))
 dev.off()
 
+# SI figure with distribution of deciduousness------------------
+
+str(sp_vars)
+
+# make a histogram of williams_dec
+
+dec_hist <- ggplot(sp_vars, aes(x = williams_dec)) +
+    geom_histogram(bins = 10, alpha = 0.7) +
+    ylab("Number of species") +
+    xlab("Deciduousness") +
+    scale_y_continuous(breaks = 1:10) +
+    theme_bw()
+
+png("doc/display/Fig_dec_dist.png", width = 4, height = 4, units = "in", res = 300)
+dec_hist
+dev.off()
+
+str(tree_vars)
+tree_n <- tree_vars %>%
+    group_by(Species) %>%
+    dplyr::summarise(n = n())
+
+str(tree_n)
+sp_n <- merge(sp_vars, tree_n, by = "Species")
+
+
+
 # SI figure with points and deciduousness--------------------
 
 # preds <- readRDS("results/models/non_negative/pred_isoclines_twi.RDS")

### 2025-09-02 doc/response_to_reviewer_R1.Rmd
updated summary table and some draft responses

@@
 
-RESPONSE:
+**RESPONSE: Thank you for your favourable review of the manuscript and for the constructive comments.
+We have now revised the manuscript based on these suggestions and that of the other reviewer which have helped improve the mansucript.
+Our detailed comments are below.**
 
@@
 
-RESPONSE:
+**RESPONSE: Our analysis includes measurements until January 2023, making 2022 the last growing year considered. We have now added this detail. It now reads:  
+*Using the late wet/early dry dendrometer band censuses (centred around January), we calculated annualised growth for each tree and year from 2009 to 2022.***
 
@@
 
-RESPONSE:
+**RESPONSE: Yes, other size effect is DBH, accounting for crown exposure. We have now made that clear.**
 
@@
 
-RESPONSE:
+*RESPONSE: Thanks for this comment.
+We have removed 2023 and 2024 from the x-axis.
+While we agree that there is a lot more to discuss and analyse from this figure, we wanted to be careful not to overinterpret the growth patterns without adequate analysis of other covariates.
+The pattern that you describe for the 2015 drought is interesting, but this is primarily because drought conditions were prolonged into 2016 - this was a multi-year drought - and the ENSO conditions switched only late in 2016, leading to a late monsoon.
+Our analysis does not have the power to analyse multi-year droughts separately from single year droughts, and therefore we chose not to analyse and interpret these cycles.*
 
@@
 
-RESPONSE:
+**RESPONSE: Thank you for your favourable review of our manuscript and for the constructive comments.
+This revision following comments has improved the clarity and interpretability of our manuscript.
+Our detailed responses to your comments are given below.**
 
@@
 
-RESPONSE:
+**RESPONSE: The January measurement was used to calculate annual growth. We have now included this detail at in now reads:  
+*Using the late wet/early dry dendrometer band censuses (centred around January), we calculated annualised growth for each tree and year from 2009 to 2022.***
 
@@
 
-RESPONSE:
+*RESPONSE: Thanks for this interesting comment.
+The timing of deciduousness may be an important factor to consider, however, from the data reported in Williams et al.
+(2008), leaf flush for most species is reported during the month after the spring equinox in late March, independent of the duration of leaflessness.
+Moreover, the timing of deciduousness can differ from year to year, and would require direct measurement at the tree to answer this interesting question.*
 

### 2025-09-03 scripts/000_twi.R
update with figures in the reviewer comments

@@ -123,12 +123,18 @@ max(twi[], na.rm = TRUE)
 #   tm_raster(style = "cont",palette = "-Greys", legend.show = FALSE)+
 tm_shape(twi) +
   tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE, alpha = 0.5) +
-  tm_scale_bar()
+  tm_scale_bar() +
+  tm_xlab("Grids East") +
+  tm_ylab("Grids North") +
+  tm_grid(lines = F)
 
 png("doc/display/explore/twi.png", width = 4, height = 8, units = "in", res = 300)
 tm_shape(twi) +
-  tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE, alpha = 1) +
-  tm_scale_bar()
+  tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE, alpha = 0.5) +
+  tm_scale_bar() +
+  tm_xlab("Grids East") +
+  tm_ylab("Grids North") +
+  tm_grid(lines = F)
 dev.off()
 
 

### 2025-09-03 doc/response_to_reviewer_R1.Rmd
update with figures in the reviewer comments
some more comments

@@
 
+![](display/Fig_dec_dist.png)
+
 •	I think it is important to add the average of non-dry years in Figure 1a (without 2010, 2015 and 2020). This helps clarify the difference between a normal/rainy year and a dry year.
@@
 Fig 1b now follows more closely, showing the deviation in SD units from the long-term mean.**
+
+![](display/Fig1.png)
 
@@
 
-RESPONSE:
+**RESPONSE: Thanks for this suggestion.
+We have now moved the description of the droughts to the methods section and have only retained a brief description of the differences among droughts in the results.**
 
@@
 
-RESPONSE:
+**RESPONSE: The three climatically distinct droughts**
 
@@
 **RESPONSE: Our analysis includes measurements until January 2023, making 2022 the last growing year considered. We have now added this detail. It now reads:  
-*Using the late wet/early dry dendrometer band censuses (centred around January), we calculated annualised growth for each tree and year from 2009 to 2022.***
+"Using the late wet/early dry dendrometer band censuses (centred around January), we calculated annualised growth for each tree and year from 2009 to 2022."**
 
@@
 
+![](display/Fig2_alt2.png)
+
 339-346: I had a hard time understanding this result, on the one hand it says that Crown exposure was associated with greater growth declines in 2010, but not for 2015 and 2020.
@@
 
-**RESPONSE: Yes, other size effect is DBH, accounting for crown exposure. We have now made that clear.**
+**RESPONSE: Yes, other size effect is DBH, accounting for crown exposure. We have now made that clear in the table.**
 
@@
 **RESPONSE: The January measurement was used to calculate annual growth. We have now included this detail at in now reads:  
-*Using the late wet/early dry dendrometer band censuses (centred around January), we calculated annualised growth for each tree and year from 2009 to 2022.***
+"Using the late wet/early dry dendrometer band censuses (centred around January), we calculated annualised growth for each tree and year from 2009 to 2022."**
 
@@
 
-RESPONSE:
+*RESPONSE: Unfortunately soil moisture data from the plot is not available for the years of study across which dendroband data are available.
+HKK, being fairly remote, has limited access to power sources and other infrastructure that are needed to maintain such data collection.
+We are currently piloting a few basic soil moisture sensors, however, this will have limited value for calibration as they only measure the top layer of the soil and are limited in spatial scale compared to the spatial resolution of TWI data.*
 
@@
 
-RESPONSE:
+**RESPONSE: Thanks for noticing this. We have made the change.**
 

### 2025-09-03 doc/manuscript_hkk_drought_sensitivity.Rmd
updated summary table and some draft responses
some more comments

@@
 
-Using the late wet/early dry dendrometer band censuses, we calculated annualised growth for each year and tree. 
+Using the late wet/early dry dendrometer band censuses (centred around January), we calculated annualised growth for each tree and year from 2009 to 2022. 
 <!-- We first conducted quality control on the dataset, removing measurements that appeared likely to be data entry errors (misplaced decimals), potential misidentified bands (old band numbers that restarted after stopping) or misidentified individuals (individuals with conflicting metadata across censuses).
@@
 Finally, we calculated daily rolling anomalies from the long-term daily expectation for each variable for the drought years as $$anomaly_{day} = \frac{rolling.mean_{day} - \bar{rolling.mean_{day}}}{sd(rolling.mean_{day})}$$
+
+<!-- ## Drought characteristics {-} -->
+
+<!-- The two droughts analysed had distinct temporal characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r fig_droughts`).
+The 2010 drought peaked between February and May (dry season and beginning of wet season), with 1-month SPEI values below -1 and meteorological anomalies on the 30-day scale including lower precipitation (min anomaly **<-1.5** SD), higher number of dry days (max anomaly **>1.5** SD), higher maximum temperature (max anomaly +0.8 SD), and higher VPD (max anomaly **>2** SD).
+Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions.
+1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought.
+ -->
+
+<!-- In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including precipitation (min anomaly **<-1.5** SD), number of dry days (max anomaly **>1.5** SD), maximum monthly mean temperatures (max anomaly +1.5 SD), and monthly maximum VPD (max anomaly +2 SD).
+1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period. 
+Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season.
+1- to 6-month SPEI values reached minima <2, indicative of extreme drought, while the 12-month values remained just above -2, indicative of severe drought. -->
+
+The droughts analysed had distinct climatic and temporal characteristics, with 2010 primarily a dry-season drought, 2015 a wet-season drought and 2020 showing mixed characteristics (Table 1, Fig `r fig_droughts`).
+In 2010, 30-day-mean meteorological anomalies between February and May included lower precipitation (min anomaly **<-1** SD), higher number of dry days (anomaly **~1.5** SD), higher maximum temperature (anomaly **>2** SD), and higher VPD (anomaly **>2** SD)<!--, including in the wet-season [KAT: commenting this out because I find it unclear/ don't understand what it means]-->.
+<!-- Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions. -->
+1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of drought, while the 12-month values remained above -1, indicative of mild drought. 
+In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including precipitation (anomaly **<-1.5** SD), number of dry days (max anomaly **>2** SD), maximum monthly temperatures (anomaly **+1.5** SD), and monthly maximum VPD (max anomaly **+2** SD).
+<!-- 1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period.  -->
+<!-- Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season. -->
+1- to 12-month SPEI fell below -2, lowest in the dendroband census period and indicative of severe drought. 
+In 2020, meteorological anomalies were pronounced both in the dry and wet seasons, including maximum temperature (**>2** SD) and VPD (**>2** SD). 
+3- and 6-month SPEI values approached -2 in February, indicative of borderline severe drought.
+12-month SPEI showed severe drought conditions from January to April, and moderate drought from May to September, spanning both dry and wet seasons.
 
@@
 
-## Drought characteristics {-}
-
-<!-- The two droughts analysed had distinct temporal characteristics, with 2010 primarily a dry season drought and 2015 a wet-season drought (Table 1, Fig `r fig_droughts`).
-The 2010 drought peaked between February and May (dry season and beginning of wet season), with 1-month SPEI values below -1 and meteorological anomalies on the 30-day scale including lower precipitation (min anomaly **<-1.5** SD), higher number of dry days (max anomaly **>1.5** SD), higher maximum temperature (max anomaly +0.8 SD), and higher VPD (max anomaly **>2** SD).
-Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions.
-1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of severe drought, while the 12-month values remained above -1, indicative of mild drought.
- -->
-
-<!-- In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including precipitation (min anomaly **<-1.5** SD), number of dry days (max anomaly **>1.5** SD), maximum monthly mean temperatures (max anomaly +1.5 SD), and monthly maximum VPD (max anomaly +2 SD).
-1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period. 
-Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season.
-1- to 6-month SPEI values reached minima <2, indicative of extreme drought, while the 12-month values remained just above -2, indicative of severe drought. -->
-
-The droughts analysed had distinct climatic and temporal characteristics, with 2010 primarily a dry-season drought, 2015 a wet-season drought and 2020 showing mixed characteristics (Table 1, Fig `r fig_droughts`).
-In 2010, 30-day-mean meteorological anomalies between February and May included lower precipitation (min anomaly **<-1** SD), higher number of dry days (anomaly **~1.5** SD), higher maximum temperature (anomaly **>2** SD), and higher VPD (anomaly **>2** SD)<!--, including in the wet-season [KAT: commenting this out because I find it unclear/ don't understand what it means]-->.
-<!-- Multi-month SPEI values reached their minima in May, after which unusually wet conditions led to a fairly rapid recovery to normal conditions. -->
-1- to 6-month SPEI values reached minima between -1.5 and -2, indicative of drought, while the 12-month values remained above -1, indicative of mild drought. 
-In 2015, pronounced meteorological anomalies on the 30-day scale occurred during the wet season, including precipitation (anomaly **<-1.5** SD), number of dry days (max anomaly **>2** SD), maximum monthly temperatures (anomaly **+1.5** SD), and monthly maximum VPD (max anomaly **+2** SD).
-<!-- 1-month SPEI was <-1 in May and June, reaching -2 in May, the lowest in the the entire dendroband census period.  -->
-<!-- Multi-month SPEI values reached their minima between June and August and remained negative into the subsequent dry season. -->
-1- to 12-month SPEI fell below -2, lowest in the dendroband census period and indicative of severe drought. 
-In 2020, meteorological anomalies were pronounced both in the dry and wet seasons, including maximum temperature (**>2** SD) and VPD (**>2** SD). 
-3- and 6-month SPEI values approached -2 in February, indicative of borderline severe drought.
-12-month SPEI showed severe drought conditions from January to April, and moderate drought from May to September, spanning both dry and wet seasons.
-
 ```{r data summaries, echo=F, warning=F, message=F}
@@
 
-## Growth responses to drought {-}
+## Tree growth responses to drought {-}
 
@@
 
-Median annual increment across yeaes varied among the `r nrow(tree.time.sp)` species analysed, with *`r tree.time.sp$spfull[1]`* the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm).
+Median annual increment across years varied among the `r nrow(tree.time.sp)` species analysed, with *`r tree.time.sp$spfull[1]`* the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm).
 Median species responses to the 2010 and 2020 droughts were variable, but no species had higher median increments than a typical year in 2015.

### 2025-09-04 doc/response_to_reviewer_R1.Rmd
changing SI plots to only have lines for significant cors

@@
 
-RESPONSE:
+**RESPONSE: Thanks for this suggestion. **
 
@@
 
-*RESPONSE: Unfortunately soil moisture data from the plot is not available for the years of study across which dendroband data are available.
+**RESPONSE: Unfortunately soil moisture data from the plot is not available for the years of study across which dendroband data are available.
 HKK, being fairly remote, has limited access to power sources and other infrastructure that are needed to maintain such data collection.
-We are currently piloting a few basic soil moisture sensors, however, this will have limited value for calibration as they only measure the top layer of the soil and are limited in spatial scale compared to the spatial resolution of TWI data.*
+We are currently piloting a few basic soil moisture sensors, however, this will have limited value for calibration as they only measure the top layer of the soil and are limited in spatial scale compared to the spatial resolution of TWI data.**
 
@@
 
-RESPONSE:
+**RESPONSE: Thanks for this suggestion.
+The ditted line in Fig S6 was originally the 1:1 line - a strict null expectation.
+We have now added dashed lines for the significant correlations.
+Following the same logic, for S11 and S12, we have now changed the non-significant correlations to dashed lines.**
 

### 2025-09-04 scripts/05_figures.R

@@ -1367,7 +1367,7 @@ dev.off()
 
 forcor <- tree.time %>%
     filter(yr %in% c(2010, 2015, 2020)) %>%
-    select(Tag, Species, yr, sens.prop) %>%
+    dplyr::select(Tag, Species, yr, sens.prop) %>%
     group_by(Tag) %>%
     pivot_wider(
         names_from = yr,
@@ -1392,7 +1392,8 @@ rval_2_3 <- cor.test(forcor$sens_2015, forcor$sens_2020, use = "pairwise.complet
 # plot the correlation
 corplot_1 <- ggplot(forcor, aes(x = sens_2010, y = sens_2015)) +
     geom_point(alpha = 0.5) +
-    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+    geom_abline(slope = 1, intercept = 0, linetype = "dashed", col = "grey40") +
+    # geom_smooth(method="lm", col="grey20", linetype)+
     xlab("Sensitivity in 2010") +
     ylab("Sensitivity in 2015") +
     # add text with r value
@@ -1404,7 +1405,8 @@ corplot_1 <- ggplot(forcor, aes(x = sens_2010, y = sens_2015)) +
 
 corplot_2 <- ggplot(forcor, aes(x = sens_2010, y = sens_2020)) +
     geom_point(alpha = 0.5) +
-    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+    geom_abline(slope = 1, intercept = 0, linetype = "dashed", col = "grey20") +
+    geom_smooth(method = "lm", col = "grey20") +
     xlab("Sensitivity in 2010") +
     ylab("Sensitivity in 2020") +
     # add text with r value
@@ -1416,7 +1418,8 @@ corplot_2 <- ggplot(forcor, aes(x = sens_2010, y = sens_2020)) +
 
 corplot_3 <- ggplot(forcor, aes(x = sens_2015, y = sens_2020)) +
     geom_point(alpha = 0.5) +
-    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+    geom_abline(slope = 1, intercept = 0, linetype = "dashed", col = "grey20") +
+    geom_smooth(method = "lm", col = "grey20") +
     xlab("Sensitivity in 2015") +
     ylab("Sensitivity in 2020") +
     # add text with r value
[…]

### 2025-09-04 doc/supplementary_information.Rmd
updated SI doc with new figs

@@
 ---
-title: "Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments"
+title: "Supplementary Information: Drought sensitivity is not a species trait: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments"
 author: 
@@
 
-![**Figure S8:** Calculated Topographic Wetness Index (TWI) values for the HKK plot](display/explore/twi.png)
+![**Figure S8:** Calculated Topographic Wetness Index (TWI) values for the HKK plot](display/twi.png)
 
-![**Figure S9:** Distribution of DBH with Crown Illumination Index (CII) in 2014.](display/explore/dbh_cii.png)
+![**Figure S9:** Distribution of DBH with Crown Illumination Index (CII) in 2014.](display/dbh_cii.png)
 
@@
 
-![**Figure S13:** Testing correlations for DBH $\perp\!\!\!\perp$ TWI by species](display/explore/cond_dep_dbh_twi.png)
+![**Figure S13:** Testing correlations for DBH $\perp\!\!\!\perp$ TWI by species](display/cond_dep_dbh_twi.png)
 
-![**Figure S14:** Testing correlations for CII $\perp\!\!\!\perp$ TWI by species](display/explore/cond_dep_cii_twi.png)
+![**Figure S14:** Testing correlations for CII $\perp\!\!\!\perp$ TWI by species](display/cond_dep_cii_twi.png)
 

### 2025-09-04 scripts/000_twi.R

@@ -128,7 +128,7 @@ tm_shape(twi) +
   tm_ylab("Grids North") +
   tm_grid(lines = F)
 
-png("doc/display/explore/twi.png", width = 4, height = 8, units = "in", res = 300)
+png("doc/display/twi.png", width = 4, height = 8, units = "in", res = 300)
 tm_shape(twi) +
   tm_raster(style = "cont", palette = "PuOr", legend.show = TRUE, alpha = 0.5) +
   tm_scale_bar() +

### 2025-09-04 doc/response_to_reviewer_R1.Rmd
updated responses

@@
 ---
-title: ''
+title: 'Responses to the Reviewer'
 output: 
@@
 ---
+
+**Title:** Drought response is not a species trait: tropical tree drought sensitivity is shaped by drought characteristics, species adaptations, and individual microenvironments
 
@@
 
-*RESPONSE: Thanks for this comment.
+*RESPONSE: Thanks for this suggestion.
 We have removed 2023 and 2024 from the x-axis.
 While we agree that there is a lot more to discuss and analyse from this figure, we wanted to be careful not to overinterpret the growth patterns without adequate analysis of other covariates.
-The pattern that you describe for the 2015 drought is interesting, but this is primarily because drought conditions were prolonged into 2016 - this was a multi-year drought - and the ENSO conditions switched only late in 2016, leading to a late monsoon.
-Our analysis does not have the power to analyse multi-year droughts separately from single year droughts, and therefore we chose not to analyse and interpret these cycles.*
+For example, the pattern of suppressed growth after the 2015 drought is not necessarily because of lag times in recovery but more likely because drought conditions were prolonged into 2016 - this was a multi-year drought - and El Nino conditions switched to La Nina only late in 2016, leading to a late monsoon.
+Our analysis was focussed on episodic droughts, and therefore we chose not to analyse and interpret the whole timeseries in this study.
+We recognise however, that lagged responses and compensatory responses would be important in such timeseries and hope to answer these as the dataset accumulates.*
 
@@
 **RESPONSE: Thanks for this suggestion.
-The ditted line in Fig S6 was originally the 1:1 line - a strict null expectation.
+The dotted line in Fig S6 was originally the 1:1 line - a strict null expectation.
 We have now added dashed lines for the significant correlations.
-Following the same logic, for S11 and S12, we have now changed the non-significant correlations to dashed lines.**
+Following the same logic, for S11 and S12, we have now removed lines for the non-significant correlations.**
 

### 2025-09-08 doc/response_to_reviewer_R1.Rmd
update with table for species sensitivity medians and IUCN status

@@
 
-**RESPONSE: Thanks for this suggestion. **
+*RESPONSE:*
 
@@
 
-RESPONSE:
+**RESPONSE: Thanks for this suggestion.
+We have now included a supplementary table that details the response of specific species to specific drought events.
+This also highlights our main message that the species responses are dissimilar acorss drought events.**
 
@@
 *RESPONSE: Thanks for this interesting comment.
-The timing of deciduousness may be an important factor to consider, however, from the data reported in Williams et al.
-(2008), leaf flush for most species is reported during the month after the spring equinox in late March, independent of the duration of leaflessness.
-Moreover, the timing of deciduousness can differ from year to year, and would require direct measurement at the tree to answer this interesting question.*
+We agree that there are several hydraulic strategies both above and belowground that might impact tree growth sensitivity to drought.
+We chose deciduousness as a simple, illustrative example of a strategy that might be associated with some of this variation.
+We agree that the timing of deciduousness may be an important factor to consider, as might leaf turgor loss point, wood density etc.
+and considering these would be a key direction for future analyses, but beyond the scope of the current analysis.*
 

### 2025-09-08 doc/supplementary_information.Rmd
update with table for species sensitivity medians and IUCN status

@@
 
+**Table S1: Species characteristics and their median sensitivities** for the three drought years.
+Species characteristics reported include accepted scientific names of species analysed, their ForestGEO codes, deciduousness values (amount of leaf loss at maximum) and IUCN status as accessed in September 2025.
+For each species, median growth sensitivity +/- standard deviation is reported across the individuals analysed in 2010, 2015, and 2020 - the three drought years considered.
+
+```{r summary table, echo=F, warning=F, message=F}
+library(tidyverse)
+datasets <- readRDS("../data/HKK-dendro/sensitivity_data.RData")
+
+tree.time <- datasets$tree.time
+tree_vars <- datasets$tree_vars
+sp_vars <- datasets$sp_vars
+
+sp_new <- read.csv("../data/HKK-dendro/species.csv")
+
+colnames(sp_vars)[1] <- "Species"
+sp_vars$iucn <- sp_new$iucn[match(sp_vars$Species, sp_new$sp)]
+
+# merge tree_vars and sp_vars
+tree_vars <- merge(tree_vars, sp_vars, by = "Species", all.x = TRUE)
+# merge tree_vars and tree.time
+tree.time <- merge(tree.time, tree_vars, by = "Tag", all.x = TRUE)
+
+# colnames(tree.time)
+
+# standardise the variables within species and across all--------------------
+
+tree.time <- tree.time %>%
+  # remove large outliers for each year
+  group_by(yr) %>%
+  dplyr::mutate(sens.prop = ifelse(mean(sens.prop, na.rm = TRUE) + 4 * sd(sens.prop, na.rm = TRUE) > sens.prop & sens.prop > mean(sens.prop, na.rm = TRUE) - 4 * sd(sens.prop, na.rm = TRUE), sens.prop, NA)) %>%
+  filter(!is.na(sens.prop) & !is.na(cii_min1) & !is.na(calcDBH_min1) & !is.na(twi)) %>%
+  ungroup()
+
+# make a species table for each year
+
+sens_sp <- tree.time %>%
+  filter(yr %in% c(2010, 2015, 2020)) %>%
+  group_by(spfull, yr) %>%
+  dplyr::summarise(
+    n = n(),
+    med_sens = round(median(sens.prop, na.rm = T), 2),
+    sd_sens = round(sd(sens.prop, na.rm = T), 2)
+  ) %>%
+  ungroup()
+
+sens_sp <- merge(sens_sp, dplyr::select(sp_vars, c(spfull, Species, williams_dec, iucn)), by = "spfull", all.x = T)
+
+# head(sens_sp)
+# make the column
+sens_sp$sens <- paste0(sens_sp$med_sens, " +/- ", sens_sp$sd_sens)
+
+# pivot wider by year
+
+sens_sp <- sens_sp %>%
+  dplyr::select(-med_sens, -sd_sens, -n) %>%
+  pivot_wider(values_from = sens, names_from = yr) %>%
+  dplyr::rename(
+    deciduousness = williams_dec,
+    "IUCN status" = iucn,
+    "Species name" = spfull,
+    "Species code" = Species
+  )
+# head(sens_sp)
+
+knitr::kable(sens_sp)
+```
+
 \newpage

### 2025-09-10 doc/response_to_reviewer_R1.Rmd
updated reviewer comments and manuscript

@@
 
-RESPONSE:
+**RESPONSE: We have now carefully reviewed the text and rewritten the description of acronyms (TWI and CII) in the results section to make it clearer for readers.**
 
@@
 
-RESPONSE:
+**RESPONSE: Thanks for this suggestion.
+We have now revised the Results section to remove some redundant statistical details that are reported in figures.
+We have also reorganised some text so that the statements that correspond to each prediction are highlighted at the start of each section.
+We hope that the results and their alignment with the predictions are clearer now.**
 
@@
 **RESPONSE: Thanks for this suggestion.
-We have now included a supplementary table that details the response of specific species to specific drought events.
-This also highlights our main message that the species responses are dissimilar acorss drought events.**
+We have now included a supplementary table (Table S1) that details the response of specific species to specific drought events.
+This summary highlights our main message that the species responses are dissimilar across drought events.
+This table also includes available information on the IUCN categories of the species analysed (accessed in September 2025) where information was avaialable.
+Two of the species analysed are "Endangered", including *Afzelia xylocarpa*, the species that was discussed.
+There are also "Vulnerable" and "Data Deficient" species in the list, some of which have consistent negative drought sensitivities.
+This table would therefore be useful for forest managers and conservation professionals and to guide species-specific research directions.**
 
@@
 
-**RESPONSE: The three climatically distinct droughts**
+**RESPONSE: The three climatically distinct droughts are summarised in Table 1.
+In summary, 2010 was a moderate drought that peaked in the dry season, 2015 was a severe drought that peaked in the wet season and 2020 was a moderate drought that peaked in the transition from dry to wet season.
+After moving the drought descriptions to the methods section, we have added this summary to set context for the results section.**
 
@@
 
-***RESPONSE: Since all crown exposure categories had negative growth sensitivities in 2015 and 2020, crown exposure was not associated with large differences in the magnitude or direction of sensitivity.
+**RESPONSE: Since all crown exposure categories had negative growth sensitivities in 2015 and 2020, crown exposure was not associated with large differences in the magnitude or direction of sensitivity.
 In other words, more exposed trees had more or less similar growth declines compared to less exposed/understory trees.
-TO CLARIFY IN TEXT***
+We have now clarified this in the text which now reads:**
+
+"Predicted effects of CII on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all CII categories in 2015 and 2020 has similar predictions, showing that exposure was not associated with altered sensitivity."
 

### 2025-09-10 doc/manuscript_hkk_drought_sensitivity.Rmd
updated reviewer comments and manuscript

@@
 **Number of Words in Abstract:** 146  
-**Number of Words in Main Text:** 5036  
+**Number of Words in Main Text:** 4999   
 **Number of References:** 55  
@@
 <!-- The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a).  -->
+In summary, the three climatically distinct droughts can be characterised thus: 2010 was a moderate, dry-season drought, 2015 a severe wet-season drought and 2020, a moderate drought during the dry-to-wet transition (Table 1).
+
 Drought years were associated with lower plot-level growth and negative tree-level sensitivities, with the strongest negative response in 2015 (Fig `r fig_drought_responses`). 
@@
 
+Species median sensitivities and individual tree sensitivities to the three droughts were not correlated with each other, supporting prediction *i* (Fig S5, Fig S6). 
 Median annual increment across years varied among the `r nrow(tree.time.sp)` species analysed, with *`r tree.time.sp$spfull[1]`* the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm).
-Median species responses to the 2010 and 2020 droughts were variable, but no species had higher median increments than a typical year in 2015.
-<!--KAT reworded here because Mitrephora thorelii was very slightly positive-->
 In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/3]`*.
 In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/3]`*, and in 2020 from `r sens.sp[sens.sp$yr==2020,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2020,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[nrow(sens.sp)/3]`*.
-
-Species median sensitivities to the three droughts were not correlated with each other, supporting prediction *i*. 
-However on average, species had more negative sensitivities to the 2015 drought than the 2010 or 2020 droughts, in alignment with median sensitivities across all trees (for 2010 and 2015 Pearson's correlation coefficient r = `r round(sens.sp.cor_10_15[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_15[3]$p.value,2)`, for 2015 and 2020, r = `r round(sens.sp.cor_15_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_15_20[3]$p.value,2)`, and for 2010 and 2020, r = `r round(sens.sp.cor_10_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_20[3]$p.value,2)`, Fig S5).
+Species had more negative sensitivities to the 2015 drought than the 2010 or 2020 droughts, in alignment with median sensitivities across all trees <!--(for 2010 and 2015 Pearson's correlation coefficient r = `r round(sens.sp.cor_10_15[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_15[3]$p.value,2)`, for 2015 and 2020, r = `r round(sens.sp.cor_15_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_15_20[3]$p.value,2)`, and for 2010 and 2020, r = `r round(sens.sp.cor_10_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_20[3]$p.value,2)`, -->(Fig S5).
+Median species sensitivities to the 2010 and 2020 droughts were variable, but no species had higher median increments than a typical year in 2015.
+<!--KAT reworded here because Mitrephora thorelii was very slightly positive-->
 
@@
 Deciduous and evergreen species showed heterogenous responses to different droughts, supporting prediction *ii*.
-Deciduous species were more drought sensitive than evergreen species in 2015 (Fig `r fig_species`a, r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`), but the opposite was true in 2020 (`r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2020], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2020], 2)`).
+Deciduous species were more drought sensitive than evergreen species in 2015 <!--(Fig `r fig_species`a, Pearson's correlation coefficient r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`), but the opposite was true in 2020 (`r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2020], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2020], 2)`)-->(Fig `r fig_species`a).
 There was no difference in 2010 (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`). 
-Models of all trees with TWI and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`,`r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`), positive in 2020 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2020],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2020],2)`,`r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2020],2)`), but no relationship in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).
+Models of all trees with Topographic Wetness Index (TWI) and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`,`r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`), positive in 2020 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2020],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2020],2)`,`r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2020],2)`), but no relationship in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).
 
@@
 
-## Microenvironment and individual effects {-}
+## Microenvironment <!--and individual -->effects {-}
 
-Individual trees did not respond similarly to the three drought events, supporting prediction *i* (Fig S6).
-
+<!-- Individual trees did not respond similarly to the three drought events, supporting prediction *i* (Fig S6).  -->
 TWI had different effects across droughts, supporting prediction *iii*. 
@@
 Drought sensitivities of individuals with different crown exposures varied across droughts (Fig `r fig_dag`), supporting prediction *iii*.
-Crown exposure was associated with greater growth declines in 2010.
-In the same models describd above, the simplex parameter for the ordered factor crown exposure in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 and 2020 the negative effect was not as strong (2015 median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`; 2020 median effect = `r round(coefs_spre_cii$median[coefs_spre_cii$yr==2020],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2020],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2020],2)`).
-Predicted effects of crown exposure on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all crown exposure categories had negative sensitivities in 2015 and 2020 (Fig `r fig_dag`b).
+Crown Illumination index (CII) was associated with greater growth declines in 2010.
+In the same models described above, the simplex parameter for the ordered factor CII in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 and 2020 the negative effect was not as strong (2015 median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`; 2020 median effect = `r round(coefs_spre_cii$median[coefs_spre_cii$yr==2020],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2020],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2020],2)`).
+Predicted effects of CII on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all CII categories in 2015 and 2020 has similar predictions, showing that exposure was not associated with altered sensitivity (Fig `r fig_dag`b).
 
-<!--DBH was a significant predictor of crown exposure, -->Trees with larger DBH had higher crown exposure across all years; median effect and 90% CI for 2010, 2015 and 2020 respectively were `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)` (`r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`), `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)` (`r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) and `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2020],2)`(`r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2020],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2020],2)`) (Fig 5a).
+<!--DBH was a significant predictor of crown exposure, -->Trees with larger DBH had higher CII across all years; median effect and 90% CI for 2010, 2015 and 2020 respectively were `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2010],2)` (`r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2010],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2010],2)`), `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2015],2)` (`r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2015],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2015],2)`) and `r round(coefs_spre_ordcii$median[coefs_spre_ordcii$yr==2020],2)`(`r round(coefs_spre_ordcii$lwr[coefs_spre_ordcii$yr==2020],2)`, `r round(coefs_spre_ordcii$upr[coefs_spre_ordcii$yr==2020],2)`) (Fig 5a).
 Controlling for the direct effect of exposure, the remaining effect of DBH *per se* was significantly positive in 2010, and not significant in 2015 or 2020 (Fig `r fig_microenvs`a, Fig `r fig_dag`a).
@@
 However, given the relative shorter duration of the timeseries, the effort required to collect these data, and the unpredictability of drought events, our analysis only included two specific droughts.-->
-The three droughts analyzed differed in multiple aspects, including ENSO strength, seasonal timing, magnitude and duration of meteorological anomalies, and severity as classified with SPEI (Table 1, Fig.
+The three droughts analyzed differed in multiple aspects, including seasonal timing, magnitude and duration of meteorological anomalies, and severity as classified with SPEI (Table 1, Fig.
 `r fig_droughts`).
@@
 <!----Limitations in generalisation/ species examined here may have greater diversity of drought adaptaitons than elsewhere in the tropics---->
-Periodic hot/dry conditions in tropical dry forests could create selective pressures for adaptative variation towards drought resistance potentially resulting in a greater diversity of drought responses here compared to other bioclimatic zones where adaptive pressures and thus ecological variability are different.
+Periodic hot/dry conditions in tropical dry forests could create selective pressures for adaptative variation towards drought resistance potentially resulting in a greater diversity of drought responses compared to other bioclimatic zones where adaptive pressures and thus ecological variability are different.
 We demonstrate that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.
+This aligns closely with recent global analyses using tree ring chronologies showing hetereogeneous responses of tree functional types to drought types [@zuidema_pantropical_2025].
 Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies resulting in diversification along the hydraulic safety-efficiency axis with diverse strategies to optimise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021].
@@
 For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the three events; it's growth nearly doubled in the 2010 and 2020 droughts but was severely reduced in the wet-season drought (Fig.
-S5).
+S5, Table S1).
 More generally, tree-ring records likely present a biased picture of drought responses, as ring formation is caused by dry-season dormancy (often associated with deciduous leaf habit) and sampling often targets large trees with exposed crowns and dry microenvironments [@speer_fundamentals_2010]-- all factors shown here to <!--directly and interactively -->influence drought sensitivity.
@@
 Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in three different droughts, showing that ecological and evolutionary strategies can make particular species and trees resistant to some droughts but vulnerable to others.
-These distinct responses of species and individuals to droughts suggests potential spatiotemporal insurance effects among tropical tree responses, in alignment with recent global analyses [@liu_drought_2022].
+These distinct responses of species and individuals to droughts suggests potential spatiotemporal insurance effects among tropical tree responses, in alignment with recent global analyses [@liu_drought_2022; @zuidema_pantropical_2025].
 
@@
 Solid lines represent relationships where 90% credible intervals do not overlap 0, blue and red lines represent positive and negative effects respectively and line thickness is scaled to the effect size.
-d) Modelled sensitivity across crown exposure classes in 2010, 2015 and 2020.
+d) Modelled sensitivity across CII classes in 2010, 2015 and 2020.

### 2025-09-15 doc/manuscript_hkk_drought_sensitivity.Rmd
update SI with model following backdoor criteria

@@
 <!-- Exposed trees showed more sensitivity to drought, but  larger sized trees or trees in wetter topographic positions seemed to buffer growth declines.  -->
-Heterogeneous drought responses of species and individuals indicate potential spatiotemporal insurance effects in diverse forests in the face of increased drought.
+Heterogeneous drought responses of species and individuals indicate potential biological insurance effects in diverse forests in the face of increased drought.
 
@@
 In the combined form, the model therefore used a multivariate distribution of Gaussian and logit.
+As an alternate approach, we applied backdoor criteria to the causal model and tested the effects of DBH, CII and TWI separately. 
+Details and results are reported in the Supplementary Material. 
 <!-- Finally, to test full mediation effects, or the hypothesis that DBH effects on sensitivity are fully explained by its effects on exposure, we tested the same models, removing the DBH as a predictor of sensitivity.
@@
 Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in three different droughts, showing that ecological and evolutionary strategies can make particular species and trees resistant to some droughts but vulnerable to others.
-These distinct responses of species and individuals to droughts suggests potential spatiotemporal insurance effects among tropical tree responses, in alignment with recent global analyses [@liu_drought_2022; @zuidema_pantropical_2025].
+These distinct responses of species and individuals to droughts suggests potential biological insurance effects among tropical tree responses, in alignment with recent global analyses [@liu_drought_2022; @zuidema_pantropical_2025].
 

### 2025-09-15 doc/supplementary_information.Rmd
updated reviewer comments and manuscript
update SI with model following backdoor criteria

@@
 For each species, median growth sensitivity +/- standard deviation is reported across the individuals analysed in 2010, 2015, and 2020 - the three drought years considered.
+IUCN category codes - "EN" = Endangered, "VU" = Vulnerable, "LC" = Least Concern, "DD" = Data Deficient.
+Empty columns indicate species which were not part of the IUCN assessment.
 
@@
 
+\newpage
+
 # Alternate model using species random effect only on intercept
@@
 ![**Figure S17:** Model results from Bayesian causal model of sensitivity as a function of DBH, CII and TWI with species random effect only on intercept.](display/FigSI_coefs_nospre.png)
+
+\newpage
+
+# Alternate models as causal inquiries following backdoor criteria
+
+For an alterate approach, we used a backdoor criteria approach as used in causal inquiry to test the effects of DBH, CII and TWI on sensitivity separately starting from a combined DAG. 
+First, we specificied the DAG on DAGitty [@textor_robust_2016] in the same strcuture as described in Fig 5 in the main text.
+We then set sensitivity as the response variable, or "outcome".
+For each variable - DBH, CII and TWI - whose causal effects we were interested in, we examined the minimum set of adjustment variables for each causal effect, by setting each of these as "exposure" one after the other and computing adjustment sets from DAGitty.
+This resulted in three different models to test the total causal effects:
+
+For TWI: $sensitivity \sim TWI$  
+For CII: $sensitivity \sim CII$  
+For DBH: $sensitivity \sim DBH + CII$
+
+We fit three different models on *brms*, with normal priors for sensitivity, TWI and DBH and a monotonic prior for CII.
+All models had a random effect of Species on both slopes and intercept. 
+Results from the causal inquiries following backdoor criteria align very closely with the SEM-like Bayesian causal model.
+
+![**Figure S18:** Model results from three independent causal inquiry models of sensitivity following backdoor criteria for total causal effect of DBH, CII and TWI with species random effect on intercept and slope.](display/causal_inquiry.png)
 

### 2025-09-15 scripts/06_causal.R

@@ -1,0 +1,214 @@
+# models considering cii as an ordered factor and using independent causal queries
+
+# Load required libraries---------------------
+library(tidyverse)
+library(brms)
+
+# load data--------------------------
+rm(list = ls())
+datasets <- readRDS("data/HKK-dendro/sensitivity_data.RData")
+
+tree.time <- datasets$tree.time
+tree_vars <- datasets$tree_vars
+sp_vars <- datasets$sp_vars
+
+colnames(sp_vars)[1] <- "Species"
+
+# merge tree_vars and sp_vars
+tree_vars <- merge(tree_vars, sp_vars, by = "Species", all.x = TRUE)
+# merge tree_vars and tree.time
+tree.time <- merge(tree.time, tree_vars, by = "Tag", all.x = TRUE)
+
+colnames(tree.time)
+
+colours <- c("#e15f41", "#546de5", "#f7b731")
+
+# standardise the variables within species and across all--------------------
+yrs <- c(2010, 2015, 2020)
+
+tree.time <- tree.time %>%
+    dplyr::mutate(
+        calcDBH_min1_scaled = scale(calcDBH_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled = scale(cii_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled = ifelse(cii_min1_scaled == "NaN", 0, cii_min1_scaled),
+        twi_scaled = scale(twi, center = TRUE, scale = TRUE),
+        tpi_scaled = scale(tpi, center = TRUE, scale = TRUE),
+        median_inc_scaled = scale(median_inc, center = TRUE, scale = TRUE),
+        avg_inc_tree_scaled = scale(avg_inc_tree, center = TRUE, scale = TRUE)
+    ) %>%
+    group_by(Species) %>%
+    # scale while retaining original values
+    dplyr::mutate(
+        calcDBH_min1_scaled_sp = scale(calcDBH_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled_sp = scale(cii_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled_sp = ifelse(cii_min1_scaled_sp == "NaN", 0, cii_min1_scaled_sp),
+        twi_scaled_sp = scale(twi, center = TRUE, scale = TRUE),
[…]

### 2025-09-17 doc/response_to_reviewer_R1.Rmd
update with edit to species response comment

@@
 csl: ecology-letters.csl 
+suppress-bibliography: TRUE
 ---
@@
 There are also "Vulnerable" and "Data Deficient" species in the list, some of which have consistent negative drought sensitivities.
-This table would therefore be useful for forest managers and conservation professionals and to guide species-specific research directions.**
+This table would therefore be useful for forest managers and conservation professionals and to guide species-specific research directions.
+We have also added a line highlighting the contrasts between species responses in the discussion, which now reads:**
+
+"An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings (n=4 at HKK; @vlam_temperature_2014) are unlikely to be representative of the drought responses of the entire community.
+For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the three events; it's growth nearly doubled in the 2010 and 2020 droughts but was severely reduced in the wet-season drought (Fig.
+S5, Table S1).
+Extrapolating from these could lead to erroneous inferences on species with conservation or commercial significance, e.g.
+*Dipterocarpus alatus* which presents a different drought response (Table S1)."
+
+**As to why different species might be responding differently, a key difference could be their evolutionary strategies along hydraulic safety vs. efficiency. 
+We have discussed our results in light of deciduousness as an example.
+However, we recognise that many other above and belowground traits may contribute to these differences and need to examined further.
+We have edited text in the discussion to read:**
+
+"Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different above- and below-ground traits along the hydraulic safety-efficiency spectrum respond to different types of drought, and how these responses may be modified by microenvironments."
 
@@
 
-RESPONSE:
+**RESPONSE: Thanks for this suggestion.
+Our conclusion here was based on the fact that both species and individuals within species (spatial) seem to have little similarity in their responses to different droughts.
+At the community/stand level, these could be either because of adaptive differences among species or environmental differences among sites that interact with different drought types.
+Upon further discussion, we think that the term "biological insurance" (*sensu* @loreau_biodiversity_2021) may be more appropriate here given its more general definition and inclusion of both species and spatial concepts of insurance.
+We have now revised the usage of this term in the manuscript and included a reference to this framework.**
 
@@
 We agree that there are several hydraulic strategies both above and belowground that might impact tree growth sensitivity to drought.
-We chose deciduousness as a simple, illustrative example of a strategy that might be associated with some of this variation.
+We chose degree of deciduousness as a simple, illustrative example of a strategy that might be associated with some of this variation.
 We agree that the timing of deciduousness may be an important factor to consider, as might leaf turgor loss point, wood density etc.

### 2025-09-17 doc/manuscript_hkk_drought_sensitivity.Rmd
update with edit to species response comment

@@
 temperate drylands reported in @vasey_intraspecific_2022), but may be distinct from the wet/everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different.
-Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different traits respond to different types of drought, and how these responses may be modified by microenvironments.
+Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different above- and below-ground traits along the hydraulic safety-efficiency spectrum respond to different types of drought, and how these responses may be modified by microenvironments.
 Again, continued expansion of dendrometer band records and ecophysiological studies will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.
@@
 S5, Table S1).
+Extrapolating from these could lead to erroneous inferences on species with conservation or commercial significance, e.g. *Dipterocarpus alatus* which presents a different drought response (Table S1).
 More generally, tree-ring records likely present a biased picture of drought responses, as ring formation is caused by dry-season dormancy (often associated with deciduous leaf habit) and sampling often targets large trees with exposed crowns and dry microenvironments [@speer_fundamentals_2010]-- all factors shown here to <!--directly and interactively -->influence drought sensitivity.
@@
 Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in three different droughts, showing that ecological and evolutionary strategies can make particular species and trees resistant to some droughts but vulnerable to others.
-These distinct responses of species and individuals to droughts suggests potential biological insurance effects among tropical tree responses, in alignment with recent global analyses [@liu_drought_2022; @zuidema_pantropical_2025].
+These distinct responses of species and individuals to droughts suggests potential biological insurance effects [@loreau_biodiversity_2021] among tropical tree responses, in alignment with recent global analyses [@liu_drought_2022; @zuidema_pantropical_2025].
 

### 2025-09-17 doc/response_to_reviewer_R1.Rmd
@teixeirak full draft of responses
- tentative answers in italics
- commit includes track changes version with _track in name

@@
 
-*RESPONSE:*
+*RESPONSE: Thanks for this suggestion.
+We were excited to see this global study published recently.
+We have now discussed our focused work in relation to this global study based on the alignment of the a key message - the diversity of drought responses with drought types and species traits (deciduousness).
+However, dendroband datasets can be key to filling gaps in spatial and climatic coverage of tree rings records in the tropics (reviewed by @groenendijk_importance_2025).
+Moreover, our results showcasing response diversity among individuals within a species demonstrates an important caveat to aggregate/species-chronology-based inferences from tree-ring records.*
 
@@
 While we agree that there is a lot more to discuss and analyse from this figure, we wanted to be careful not to overinterpret the growth patterns without adequate analysis of other covariates.
-For example, the pattern of suppressed growth after the 2015 drought is not necessarily because of lag times in recovery but more likely because drought conditions were prolonged into 2016 - this was a multi-year drought - and El Nino conditions switched to La Nina only late in 2016, leading to a late monsoon.
+For example, the pattern of suppressed growth after the 2015 drought is not necessarily because of lag times in recovery but more likely because drought conditions extended into 2016 - this was a multi-year drought - and El Nino conditions switched to La Nina only late in 2016, leading to a late monsoon.
 Our analysis was focussed on episodic droughts, and therefore we chose not to analyse and interpret the whole timeseries in this study.

### 2025-09-17 doc/manuscript_hkk_drought_sensitivity.Rmd
@teixeirak full draft of responses
- tentative answers in italics
- commit includes track changes version with _track in name

@@
 **Number of Words in Abstract:** 146  
-**Number of Words in Main Text:** 4999   
+**Number of Words in Main Text:** 5074   
 **Number of References:** 55  
@@
 temperate drylands reported in @vasey_intraspecific_2022), but may be distinct from the wet/everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different.
-Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different above- and below-ground traits along the hydraulic safety-efficiency spectrum respond to different types of drought, and how these responses may be modified by microenvironments.
+Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022, @groenendijk_importance_2025], we know exceedingly little about how tropical tree species with different above- and below-ground traits along the hydraulic safety-efficiency spectrum respond to different types of drought, and how these responses may be modified by microenvironments.
 Again, continued expansion of dendrometer band records and ecophysiological studies will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.

### 2025-10-07 doc/response_to_reviewer_R1.Rmd
updated response to reviewers from HCM comments

@@
 
-**Title:** Drought response is not a species trait: tropical tree drought sensitivity is shaped by drought characteristics, species adaptations, and individual microenvironments
+Dear Editor,
 
-Referees' comments to the author(s):
+We are pleased to submit a revised version of our manuscript, "Drought response is not a species trait: tropical tree drought sensitivity is shaped by drought characteristics, species adaptations, and individual microenvironments" () for consideration for publication in Ecology Letters.
+We have addressed all points raised by the reviewers as detailed in our responses to the reviewers. 
+The most significant changes were discussing global implications as well as species effects in more detail. 
+We have also revised the methodology and results to highlight the correspondence between the predictions and the results. 
+We have also updated Figures 1 and 2 as well as the relevant figures in the supplementary material as requested by the reviewers.
 
-Referee: 1
+Thank you for considering this revised version for publication. 
+We look forward to your response.
+
+Sincerely,
+On behalf of all coauthors,
+Krishna Anujan
+
+\newpage
+
+## Referees' comments to the author(s):
+
+### Referee: 1
 
@@
 
-RESPONSE:
+**RESPONSE: Thanks for this comment.
+This is a great idea, and we would like to include this as well to provide further context to readers.
+Given that we are at the limit of display items in the main text, we have included this as Figure S1 in the supplementary material.**
 
@@
 
-*RESPONSE: Thanks for this suggestion.
+**RESPONSE: Thanks for this suggestion.
 We were excited to see this global study published recently.
-We have now discussed our focused work in relation to this global study based on the alignment of the a key message - the diversity of drought responses with drought types and species traits (deciduousness).
+We have now discussed our focused work in relation to this global study based on the alignment of the key message - the diversity of drought responses with drought types and species traits (deciduousness).
 However, dendroband datasets can be key to filling gaps in spatial and climatic coverage of tree rings records in the tropics (reviewed by @groenendijk_importance_2025).
-Moreover, our results showcasing response diversity among individuals within a species demonstrates an important caveat to aggregate/species-chronology-based inferences from tree-ring records.*
+Moreover, our results showcasing response diversity among individuals within a species demonstrates an important caveat to aggregate/species-chronology-based inferences from tree-ring records.**
 
@@
 
-"An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings (n=4 at HKK; @vlam_temperature_2014) are unlikely to be representative of the drought responses of the entire community.
+**"An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings (n=4 at HKK; @vlam_temperature_2014) are unlikely to be representative of the drought responses of the entire community.
 For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the three events; it's growth nearly doubled in the 2010 and 2020 droughts but was severely reduced in the wet-season drought (Fig.
@@
 Extrapolating from these could lead to erroneous inferences on species with conservation or commercial significance, e.g.
-*Dipterocarpus alatus* which presents a different drought response (Table S1)."
+*Dipterocarpus alatus* which presents a different drought response (Table S1)."**
 
@@
 
-"Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different above- and below-ground traits along the hydraulic safety-efficiency spectrum respond to different types of drought, and how these responses may be modified by microenvironments."
+**"Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different above- and below-ground traits along the hydraulic safety-efficiency spectrum respond to different types of drought, and how these responses may be modified by microenvironments."**
 
@@
 
-**RESPONSE: Since all crown exposure categories had negative growth sensitivities in 2015 and 2020, crown exposure was not associated with large differences in the magnitude or direction of sensitivity.
+**RESPONSE: We agree the wording could be clearer.
+Our focus is on how growth sensitivity changes with crown exposure.
+Since all crown exposure categories had negative growth sensitivities in 2015 and 2020, crown exposure was not associated with large differences in the magnitude or direction of sensitivity.
 In other words, more exposed trees had more or less similar growth declines compared to less exposed/understory trees.
@@
 
-"Predicted effects of CII on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all CII categories in 2015 and 2020 has similar predictions, showing that exposure was not associated with altered sensitivity."
+**"Predicted effects of CII on sensitivity decreased monotonically, with strong decreases in 2010, weak decrease in 2015, and no effect in 2020."**
 
@@
 
-*RESPONSE: Thanks for this suggestion.
+**RESPONSE: Thanks for this suggestion.
 We have removed 2023 and 2024 from the x-axis.
@@
 Our analysis was focussed on episodic droughts, and therefore we chose not to analyse and interpret the whole timeseries in this study.
-We recognise however, that lagged responses and compensatory responses would be important in such timeseries and hope to answer these as the dataset accumulates.*
+We recognise however, that lagged responses and compensatory responses would be important in such timeseries and hope to answer these as the dataset accumulates.**
 
@@
 
-*RESPONSE: Thanks for this interesting comment.
+**RESPONSE: Thanks for this interesting comment.
 We agree that there are several hydraulic strategies both above and belowground that might impact tree growth sensitivity to drought.
@@
 We agree that the timing of deciduousness may be an important factor to consider, as might leaf turgor loss point, wood density etc.
-and considering these would be a key direction for future analyses, but beyond the scope of the current analysis.*
+and considering these would be a key direction for future analyses, but beyond the scope of the current analysis.**
 

### 2025-10-09 doc/manuscript_hkk_drought_sensitivity.Rmd
updated manuscript file and track changes

--- 2025-09-15 doc/manuscript_hkk_drought_sensitivity.Rmd
@@
 **Number of Words in Abstract:** 146  
-**Number of Words in Main Text:** 4999   
+**Number of Words in Main Text:** 5054   
 **Number of References:** 55  
@@
 KA, SMM, SJD and KAT conceptualised the analysis. 
-SB, SJD and NP have been leading long-term data collection at HKK, while HCM created and helped execute the dendrometer band protocol. 
+SB, SJD and NP have been leading long-term data collection at HKK, while HCM led the development of the dendrometer band protocol. 
 KA cleaned the data and conducted formal analysis with support from SMM, SJD, HCM and KAT. 
@@
 Increased frequency and severity of droughts threatens forest health worldwide<!--, with important implications for this critical carbon sink-->. 
-Tree species adaptations <!--to hot and dry conditions-->-- e.g., dry-season deciduousness in tropical seasonal forests -- and individual traits -- e.g., size, crown position -- shape drought resistance, but such resistance may be variable across species, microenvironments, and drought events.
+Tree species-specific adaptations <!--to hot and dry conditions-->-- e.g., dry-season deciduousness in tropical seasonal forests -- and individual traits -- e.g., size, crown position -- shape drought resistance, but such resistance may be variable across species, microenvironments, and drought events.
 Here, we assess growth responses of 1820 trees across 30 species to three climatically distinct droughts in a seasonally dry tropical forest in Western Thailand. 
@@
 
-Intense droughts are becoming more frequent [@ipcc_climate_2023], affecting the demography, diversity, and  carbon (C) cycling of forests globally. 
-However, drought impacts on tree woody growth and ecosystem C sequestration remain difficult to predict [@evans_scaling_2025], particularly for tropical forests, <!--whose critical C sink is threatened by drought [@brando_droughts_2019] but--> for which long-term records of annual woody tree growth are rare [@zuidema_tropical_2022].
+Intense droughts are becoming more frequent globally [@ipcc_climate_2023], affecting the demography, diversity, and  carbon (C) cycling of forests. 
+However, drought impacts on tree woody growth and ecosystem C sequestration remain difficult to predict [@evans_scaling_2025], particularly for tropical forests, <!--whose critical C sink is threatened by drought [@brando_droughts_2019] but--> for which long-term records of annual woody tree growth are limited [@zuidema_tropical_2022;@groenendijk_importance_2025].
 For forests worldwide, studies have separately shown that drought characteristics, species traits, tree size and microenvironment affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter understanding of forest responses to drought.
-Specifically, if these factors act independently, tree drought responses may be conserved, and linearly predictable; i.e., species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought.
-In contrast, if these factors interact, species and individuals that exhibit resistance to one drought may prove vulnerable to a drought with different characteristics. 
-The complexity and diversity of drought responses may act as a mechanism for maintaining species diversity and buffering ecosystem function even as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].
+If these factors act independently, tree drought responses may be conserved, and linearly predictable; i.e., species with low sensitivity to past droughts would have low sensitivity in any microhabitat and any drought.
+Alternatively, if these factors interact, then species and individuals that exhibit resistance to one drought may prove vulnerable to a drought with different characteristics. 
+These complex and diverse drought responses may buffer ecosystem function even as intense droughts become more common and may act as a mechanism for maintaining species diversity [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].
 
@@
 [Then go into some details with citations on ?] -->
-Drought is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the timing, severity and duration of climatic anomalies [@slette_how_2019].
+Drought is notoriously hard to define and can be characterized based on the timing, severity, or duration of climatic anomalies in a variety of meteorological variables[@slette_how_2019].
 <!-- ; we define it here as hot/dry conditions anomalously departing from climatic means during any time of year  -->
-Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], differences in seasonal timing, duration, or intensity of droughts may result in variable tree growth responses, especially in seasonal forests with diverse adaptive strategies to handle regular dry periods [@albert_cryptic_2019;@gao_dynamic_2018].
+<!-- Although forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], differences in seasonal timing, duration, or intensity of droughts may result in variable tree growth responses, especially in seasonal forests with diverse adaptive strategies to handle regular dry periods [@albert_cryptic_2019;@gao_dynamic_2018].
+-->
+This difficulty in definition, and subsequent predictions of tree growth responses, is pronounced in seasonal forests with diverse adaptive strategies that have evolved under consistent low-rainfall periods [@albert_cryptic_2019;@gao_dynamic_2018].
 <!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
 <!-- Sean::: don't think the above is true universally. Laura Boeschoten and Erick Esteban have separate papers well in prep showing some species demonstarating seasonal growth and others not -->
-In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020].
+In tropical forests with strong dry seasons, tree drought responses may depend less on the climatic extent of the drought and more on drought timing relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020].
 <!---- agreed but see above ref I put in to [@albert_ecology_letters_cryptic_phenology].
@@
 <!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
-Analyses to date have found that interannual variation in tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry-season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent a very limited set of tropical tree species and climates.
+For example, interannual variation in tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->has been shown to be most sensitive to dry-season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021].
+However, these analyses represent a limited set of tree species and climates. 
 It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those that intensify or conflict with regular seasonal cycles<!-- of wet and dry seasons-->.
@@
 <!-- Sean::: mention rooting depth? [@chitra-tarak: root of the drought and chitra-tarak: new_phyt 2022] -->
-Tree species adaptations to drought vary widely, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
-In tropical seasonal forests in particular, trees have evolved to optimize growth, reproduction, and survival (i.e., fitness) under dry seasons, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
+Species adaptations to drought vary widely, often leading to classification of tree species as "drought tolerant" or "drought sensitive" or "drought avoidant" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
+<!-- In tropical seasonal forests in particular, trees optimize growth, reproduction, and survival (i.e., fitness) under dry seasons, along a continuum spanning intensive allocation to hydraulic safety or to hydraulic efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
+-->
+In tropical seasonal forests, many tree species vary in their hydraulic strategies along a continuum from drought tolerant species having high hydraulic safety, low hydraulic efficiency and high tissue investment to drought "avoidant" species with high hydraulic efficiency, low hydraulic safety and low investment to tissues [@gonzález-m._diverging_2021].
+Traits associated with hydraulic safety and drought tolerance include deep roots (to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, to maintain gas exchange and photosynthesis under hot and dry conditions), and small vessels (to reduce embolism risk)<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021].
 Dry-season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss [@desouza_drought_2020]. 
-Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and large vessels (hydraulic efficiency)<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021].
-Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought vulnerability, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021].
+Under drought conditions, these strategies may confer more or less drought resistance depending on the specific trait combinations and drought characteristics [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021].
 <!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.
@@
 <!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
-Moreover, besides being adaptively conserved, these strategies can also have developmental or ecophysiological controls.
-For example, some species regulate $\pi_{tlp}$, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; @hulshof_variation_2010].
+These strategies can also have developmental or ecophysiological controls.
+For example, some species regulate $\pi_{tlp}$, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008;@hulshof_variation_2010].
 
@@
 -->
-Microenvironments defined by horizontal topographic and edaphic variation and vertical light and thermal variation further shape individual tree experience of, and response to, drought. 
-While habitats with relatively moist soils (e.g., low-lying topography) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted [e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021].
+Microenvironments defined by horizontal topographic and edaphic variation and vertical light and thermal variation shape individual tree experience of, and response to, drought. 
+Although habitats with relatively moist soils (e.g., low-lying topography) are more hospitable during drought, trees in these microenvironments may be less drought-adapted [e.g., due to less negative $\pi_{tlp}$, @kunert_leaf_2021] and thus more vulnerable under severe droughts.
 Indeed, during drought, trees near streams experienced greater growth declines in a temperate forest [@mcgregor_tree_2021] and increased mortality in a tropical forest [@zuleta_droughtinduced_2017]. 
 <!-- Across vertical axes, trees of different sizes have their crowns and root systems in sometimes dramatically divergent microenvironments. -->
-Tree crowns in canopy or emergent positions are exposed to higher evaporative demand and thermal stress driven by higher solar radiation, wind speed, and vapor pressure deficit [@vinod_thermal_2023], and such exposure makes trees more vulnerable to drought [@scharnweber_confessions_2019; @ma_tree_2023].
+Tree crowns in canopy or emergent positions are exposed to higher evaporative demand and thermal stress driven by higher solar radiation, wind speed, and vapor pressure deficit [@vinod_thermal_2023], and such exposure makes trees more vulnerable to drought [@scharnweber_confessions_2019;@ma_tree_2023].
 <!--In contrast, understory trees may be rarely exposed to direct solar radiation and extreme evaporative demand.-->
@@
 it doesn't really make sense unless coupled with other traits (e.g., deciduousness shifts accross ontogeny) as 'putting down a tap root' is one of the first things a seedling of a deep rooted species does (i.e., it's not allometry but strategy) -->
-However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
-While larger trees tend to suffer greater growth declines during drought [@bennett_larger_2015], this may be driven by crown exposure, by height <!--itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium --><!--, as described by Darcy's law -->and its pressures on stem water flow[@mcdowell_darcys_2015; @fernández-de-uña_role_2023], or by trait differences across vertical axis [@vinod_thermal_2023].
+However, as with wet-adapted trees, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021].
+Although larger trees tend to suffer greater growth declines during drought [@bennett_larger_2015], this may be driven by crown exposure, by height <!--itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium --><!--, as described by Darcy's law -->and its pressures on stem water flow[@mcdowell_darcys_2015;@fernández-de-uña_role_2023], or by systematic variation in leaf traits differences along the vertical axis [@vinod_thermal_2023].
 Thus, horizontal and vertical microenvironmental gradients and tree size modify tree drought responses, likely in interaction with drought characteristics and species traits.
@@
 We hypothesize that tree drought sensitivity is jointly and interactively shaped by drought characteristics, species adaptations, tree size and microenvironment, resulting in variable drought responses of species and individuals across droughts, and of species across microhabitats.
-Specifically, we test predictions that:
+We test predictions that:
 i) both species and individuals respond differently to different droughts;
@@
 iii) on the individual level, the influences of topographic location, crown exposure, and other size effects on drought responses are variable across droughts.
-We tested these hypotheses using  a 14-year record of dendrometer band measurements in a dry seasonal evergreen forest [@bunyavejchewin_seasonally_2011] in Thailand. 
+We tested these hypotheses using  a 14-year record of dendrometer band measurements in a dry seasonal evergreen forest in Thailand[@bunyavejchewin_seasonally_2011]. 
 During this period, the forest experienced three different dry periods, two associated with ENSO events, and all three differing with respect to timing and duration (Table 1). 
 With data on 1820 trees across 30 species, we built Bayesian regression models exploring how species characteristics and individual spatial context influenced growth responses to these droughts. 
-We used a causal modelling framework to account for potential confounding relationships among the variables. 
-This analysis adds important insight into the complexity of drought responses in species-rich seasonal dry tropical forests.
+We used a causal modelling framework to account for potential confounding relationships among the variables and to enable casual interpretation of estimated parameters. 
+This analysis demonstrates the complexity of drought responses in species-rich seasonal dry tropical forests.
 
@@
 In HKK, metal dendrometer bands were installed in 2008 on 2353 trees across 152 species following standard ForestGEO protocol [@muller-landau_metal_2008]. 
-Selected were a size-stratified and spatially-stratified random sample, with additional trees chosen to fill out sample sizes for selected species of interest.
-Thus relative abudances among the selected trees largely reflected abundance variation in the plot as a whole, with more individuals banded from the abundant species (n=1 to 230 trees per species). 
+Most selected trees were part of a size-stratified and spatially-stratified random sample, with additional trees chosen to fill out sample sizes for species of interest.
+Thus, relative abudances among the selected trees largely reflected abundance variation in the plot as a whole, with more individuals banded from the abundant species (n=1 to 230 trees per species). 
 On each tree, the dendrometer band was installed at a height of 10 cm above the regular point of measurement for the whole-plot census meaning installation at 1.4 m on most trees, and at higher heights on buttressed trees with higher measurement points.
@@
 We conducted further quality control on increments (see supplementary methods for details) to assemble the final dataset. 
-Diameter tape measurements $dia_{i, t}$ are not precise enough for calculating drought sensitivity. 
-Error for growth is twice measurement error = 2 mm; in comparison, median annual growth in HKK is ~1.4 mm and calculating deviation from annual growth needs sub-mm precision. 
-However, we used tape measurements to flag likely errors in the dendrometer dataset. 
+We used concurrent tape measurements to flag likely errors in the dendrometer dataset 
+(Diameter tape measurements alone are not precise enough for calculating drought sensitivity, as their measurement precision is ~2 mm compared with median annual growth in HKK of ~1.4 mm). 
+<!-- However, we used tape measurements to flag likely errors in the dendrometer dataset.  -->
 <!-- We calculated increments from annual tape measurements made on each dendrobanded tree at each census and removed dendroband measurements that had low agreement.  -->
-Finally, we excluded trees with negative increments over the whole timeseries, and included only species that had at least 10 individuals, resulting in a final dataset of 1820 individuals of 30 species.
+We also excluded trees with negative increments over the whole timeseries, and included only species that had at least 10 individuals, resulting in a final dataset of 1820 individuals of 30 species.
 
@@
 -->
-We identified 2010, 2015 and 2020 as drought years of interest in the dendroband timeseries based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses.
+We identified 2010, 2015 and 2020 as drought years of interest in the dendroband timeseries based on two drought indices, corroborated by meteorological data, local expertise, and plot-level growth responses.
 We calculated Standardised Precipitation-Evapotranspiration Index (SPEI) at 1-, 3-, 6- and 12- month scales using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
@@
 We used data from CHIRPS [@funk_climate_2015] for daily total precipitation and ERA5Land [@muñoz-sabater_era5land_2021] to calculate daily maximum temperature and vapour pressure deficit (VPD, @raesch_pvldcurve_2020).
-We then calculated rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature over 30-day periods across the whole timeseries.
-Finally, we calculated daily rolling anomalies from the long-term daily expectation for each variable for the drought years as $$anomaly_{day} = \frac{rolling.mean_{day} - \bar{rolling.mean_{day}}}{sd(rolling.mean_{day})}$$
+We then calculated 30-day rolling means of precipitation, number of dry days (precipitation = 0), VPD and maximum temperature across the whole timeseries.
+Finally, for each day in the drought years, we calculated anomalies of these rolling means from their long-term averages for the same day of year $$anomaly_{day} = \frac{rolling.mean_{day} - \overline{rolling.mean_{day}}}{sd(rolling.mean_{day})}$$
 
@@
 
-We first  calculated the mean annual increment $\bar{a.inc_{i}}$ for each tree from annualised increment $ainc_{i,t}$.
+We first  calculated the mean annual increment $\overline{a.inc_{i}}$ for each tree from annualised increment $ainc_{i,t}$.
 We then calculated growth sensitivity for each tree $i$ for each drought year $y$ as 
 $$
-sensitivity_{i, t=y} = \frac{ainc_{i, t=y} - \bar{a.inc_{i}}}{\bar{a.inc_{i}}}
+sensitivity_{i, t=y} = \frac{ainc_{i, t=y} - \overline{a.inc_{i}}}{\overline{a.inc_{i}}}
 $$
-By scaling the increment deviation to the mean annual growth of each tree, this metric allows us to compare across slow and fast growing individuals and species.
+By scaling the increment deviation to the mean annual growth of each tree, this metric allows us to compare across slow and fast-growing individuals and species.
 
@@
 Across all individuals, we calculated Topographic Wetness Index (TWI) for an expectation of the water availability across space based on the total upslope area and the slope at the location. 
-We used a Digital Elevation Model at 5 m spatial grain for the area of the plot and the *whitebox* package in R [@wu_whitebox_2022; @lindsay_whitebox_2016a] to calculate flow accumulation at each cell and then calculate TWI as $log_{e}\frac{As}{tan(slope)}$ where $As$ is the total upslope area within the plot and $slope$ is the absolute value of slope calculated at the 5m grain.
+We used a Digital Elevation Model at 5 m spatial grain for the area of the plot and the *whitebox* package in R [@wu_whitebox_2022;@lindsay_whitebox_2016a] to calculate flow accumulation at each cell and then calculate TWI as $log_{e}\frac{As}{tan(slope)}$ where $As$ is the total upslope area within the plot and $slope$ is the absolute value of slope calculated at the 5m grain.
 <!-- As an alternate measure of water availability, we also calculated Topographic Position Index (TPI) based on the concavity/convexity of a location.  -->
-This geolocated layer allows fine resolution and precise tree mapping with topography, but this omits upslope area beyond the plot.
+This geolocated layer allows fine resolution and precise tree mapping with topography, but the omission of upslope area beyond the plot means values are underestimates and variably so for different locations.
 For an alternate estimate of wetness independent of total upslope area (Topographic Position Index) and associated results, see supplementary material. 
@@
 For each species, these were calculated from observations of individual trees that were assigned categorical scores, and averaged, allowing species to take continuous, decimal values in the 0-4 range.
-Data were unavailable for _Alphonsea ventricosa_, an evergreen species, and so we manually assigned it a deciduousness score of 1. 
+Data were unavailable for _Alphonsea ventricosa_, an evergreen species; we assigned it a deciduousness score of 1. 
 <!-- We also considered an alternative measure of deciduousness - the duration of time spent at maximum crown loss. -->
@@
 <!-- The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a).  -->
-In summary, the three climatically distinct droughts can be characterised thus: 2010 was a moderate, dry-season drought, 2015 a severe wet-season drought and 2020, a moderate drought during the dry-to-wet transition (Table 1).
+The three climatically distinct droughts can be characterised as follows: 2010 was a moderate, dry-season drought, 2015 a severe wet-season drought and 2020, a moderate drought during the dry-to-wet transition (Figs `r fig_droughts`, S1, S2).
 
 Drought years were associated with lower plot-level growth and negative tree-level sensitivities, with the strongest negative response in 2015 (Fig `r fig_drought_responses`). 
-Median plot-level annual growth rate across years of measurement was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` $\pm$ `r round(sd(tree.time$inc_annual*10, na.rm=T),2)` mm. 
+Median growth rate across all trees and years was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` mm (range `r round(min(tree.time$inc_annual*10, na.rm=T),2)`, `r round(max(tree.time$inc_annual*10, na.rm=T),2)`).
 The drought year 2015 had the lowest median increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% lower than the median across all years.
@@
 <!-- On average, trees had negative drought sensitivities during drought years, but with considerable variation (Fig `r fig_drought_responses`b).  -->
-Median drought sensitivity across all trees in 2010, 2015 and 2020 were `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)`, `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)` and `r round(median(tree.time$sens.prop[tree.time$yr==2020]), 2)` $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2020]), 2)` respectively.
+Median drought sensitivity across all trees in 2010, 2015 and 2020 were `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)`<!-- $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)`-->, `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)`<!-- $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`--> and `r round(median(tree.time$sens.prop[tree.time$yr==2020]), 2)`<!-- $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2020]), 2)`--> respectively.
 <!-- *(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))* -->
@@
 
-Species median sensitivities and individual tree sensitivities to the three droughts were not correlated with each other, supporting prediction *i* (Fig S5, Fig S6). 
-Median annual increment across years varied among the `r nrow(tree.time.sp)` species analysed, with *`r tree.time.sp$spfull[1]`* the lowest (`r tree.time.sp$inc_annual_med[1]` $\pm$ `r tree.time.sp$inc_annual_sd[1]` mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]` $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]` mm).
+Species median sensitivities and individual tree sensitivities to the three droughts were not correlated with each other, supporting prediction (*i*) (Fig S5, Fig S6). 
+Median annual increment across years varied among the `r nrow(tree.time.sp)` species analysed, with *`r tree.time.sp$spfull[1]`* the lowest (`r tree.time.sp$inc_annual_med[1]`<!-- $\pm$ `r tree.time.sp$inc_annual_sd[1]`--> mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]`<!-- $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]`--> mm).
 In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/3]`*.
@@
 <!-- Deciduousness species responded more negatively than evergreen species in 2015, but not in 2010 (Fig `r fig_species`a).  -->
-Deciduous and evergreen species showed heterogenous responses to different droughts, supporting prediction *ii*.
-Deciduous species were more drought sensitive than evergreen species in 2015 <!--(Fig `r fig_species`a, Pearson's correlation coefficient r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`), but the opposite was true in 2020 (`r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2020], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2020], 2)`)-->(Fig `r fig_species`a).
-There was no difference in 2010 (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`). 
-Models of all trees with Topographic Wetness Index (TWI) and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significant and negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`,`r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`), positive in 2020 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2020],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2020],2)`,`r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2020],2)`), but no relationship in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).
+Deciduous and evergreen species showed heterogenous responses to different droughts, supporting prediction (*ii*).
+Deciduous species were more drought sensitive than evergreen species in 2015 <!--(Fig `r fig_species`a, Pearson's correlation coefficient r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2015], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2015], 2)`)-->, less sensitive in 2020 <!--(`r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2020], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2020], 2)`)-->, whereas there was no difference in 2010(Fig `r fig_species`a).
+<!-- There was no difference in 2010 (r = `r round(coefs_dec_lms$estimate[coefs_dec_lms$yr==2010], 2)`, p = `r round(coefs_dec_lms$p.value[coefs_dec_lms$yr==2010], 2)`.  -->
+Models of all trees with Topographic Wetness Index (TWI) and deciduousness as predictors showed that the conditional effect of deciduousness on sensitivity (controlling for topographic wetness) was significantly negative in 2015 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2015],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2015],2)`,`r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2015],2)`), significantly positive in 2020 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2020],2)`; 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2020],2)`,`r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2020],2)`), and non-existent in 2010 (`r round(isocline_coefs_dec$median[isocline_coefs_dec$yr==2010],2)`, 90% CI `r round(isocline_coefs_dec$lwr[isocline_coefs_dec$yr==2010],2)`, `r round(isocline_coefs_dec$upr[isocline_coefs_dec$yr==2010],2)`).
 
@@
 <!-- Individual trees did not respond similarly to the three drought events, supporting prediction *i* (Fig S6).  -->
-TWI had different effects across droughts, supporting prediction *iii*. 
-Across all trees, the model with TWI and deciduousness showed that TWI had no effect on drought sensitivity in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`) but a positive effect in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`) and 2020 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2020],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2020],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2020],2)`).
-The interaction between TWI and deciduousness was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`) but was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) or 2020, (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2020],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2020],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2020],2)`)
+TWI had different effects across droughts, supporting prediction (*iii*). 
+Across all trees, the model with wetness and deciduousness showed that TWI had a positive effect on drought sensitivity in 2015 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2015],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2015],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2015],2)`) and 2020 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2020],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2020],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2020],2)`), meaning that wetter areas showed less negative or more positive responses to drought, but no effect in 2010 (median effect =`r round(isocline_coefs_twi$median[isocline_coefs_twi$yr==2010],2)`, 90% CI `r round(isocline_coefs_twi$lwr[isocline_coefs_twi$yr==2010],2)`, `r round(isocline_coefs_twi$upr[isocline_coefs_twi$yr==2010],2)`).
+The interaction between wetness and deciduousness was significantly negative in 2015 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2015],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2015],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2015],2)`) but was not significant in 2010 (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2010],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2010],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2010],2)`) or 2020, (median effect =`r round(isocline_coefs_inter$median[isocline_coefs_inter$yr==2020],2)`, 90% CI `r round(isocline_coefs_inter$lwr[isocline_coefs_inter$yr==2020],2)`, `r round(isocline_coefs_inter$upr[isocline_coefs_inter$yr==2020],2)`), meaning that
 <!-- In 2015, evergreen trees responded more negatively in drier than in wetter areas, whereas deciduous trees - which were more negatively affected overall - responded more negatively in wetter than in drier areas.
 -->
-Predicted sensitivities of deciduous trees were more negative than evergreen trees in 2015, with this effect stronger in wetter sites (Fig `r fig_species`b). 
-In 2020, predicted sensitivities of deciduous trees were more positive than evergreen trees and this effect was observable in drier sites (Fig `r fig_species`b).
+the divergent sensitivities of deciduous and evergreen trees in 2015 and 2020 were more pronounced in wetter and drier sites respectively (Fig `r fig_species`b). 
+<!-- In 2020, predicted sensitivities of deciduous trees were more positive than evergreen trees and this effect was observable in drier sites (Fig `r fig_species`b). -->
 
@@
 Broadly, the effects of wetness and size were different across droughts while exposure tended to decrease growth with varying strengths across droughts (Fig `r fig_microenvs`a). 
-Wetness had a small negative effect in 2010 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but a positive effect in 2015 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`) and 2020 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2020],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2020],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2020],2)`).
+Accounting for DBH and exposure, wetness had a small negative effect in 2010 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2010],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2010],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2010],2)`), but a positive effect in 2015 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2015],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2015],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2015],2)`) and 2020 (median effect =`r round(coefs_spre_twi$median[coefs_spre_twi$yr==2020],2)`, 90% CI `r round(coefs_spre_twi$lwr[coefs_spre_twi$yr==2020],2)`, `r round(coefs_spre_twi$upr[coefs_spre_twi$yr==2020],2)`).
 <!-- The positive effect of TWI in 2015 was, however, mainly due to the response of evergreen species; deciduousness had a significant negative correlation with the slopes of the TWI effect in 2015 (Fig `r fig_microenvs`b).
@@
 <!-- Predictions across the range of observed variables also showed that deciduous species experienced greater growth declines in 2015 across the wetness gradient (Fig `r fig_microenvs`c).  -->
-Deciduous species experiences greater growth declines in  wetter sites in 2015, while these interactions were less clear in 2010 and 2020 (Fig `r fig_microenvs`c).
+<!-- Deciduous species experiences greater growth declines in  wetter sites in 2015, while these interactions were less clear in 2010 and 2020 (Fig `r fig_microenvs`c).   -->
 
@@
 
-Drought sensitivities of individuals with different crown exposures varied across droughts (Fig `r fig_dag`), supporting prediction *iii*.
+Drought sensitivities of individuals with different crown exposures varied across droughts (Fig `r fig_dag`), supporting prediction (*iii*).
 Crown Illumination index (CII) was associated with greater growth declines in 2010.
-In the same models described above, the simplex parameter for the ordered factor CII in 2010 was significant and negative (median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 and 2020 the negative effect was not as strong (2015 median effect =`r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`; 2020 median effect = `r round(coefs_spre_cii$median[coefs_spre_cii$yr==2020],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2020],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2020],2)`).
-Predicted effects of CII on sensitivity decreased monotonically, with a decrease to negative sensitivities in category 4 and 5 in 2010, but all CII categories in 2015 and 2020 has similar predictions, showing that exposure was not associated with altered sensitivity (Fig `r fig_dag`b).
+In the same models described above, the simplex parameter for the ordered factor CII in 2010 was significant and negative (median effect = `r round(coefs_spre_cii$median[coefs_spre_cii$yr==2010],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2010],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2010],2)`), while in 2015 and 2020 the negative effect was not as strong (2015 median effect = `r round(coefs_spre_cii$median[coefs_spre_cii$yr==2015],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2015],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2015],2)`; 2020 median effect = `r round(coefs_spre_cii$median[coefs_spre_cii$yr==2020],2)`, 90% CI `r round(coefs_spre_cii$lwr[coefs_spre_cii$yr==2020],2)`, `r round(coefs_spre_cii$upr[coefs_spre_cii$yr==2020],2)`).
+Predicted effects of CII on sensitivity decreased monotonically, with strong decreases in 2010, weak decrease in 2015, and no effect in 2020 (Fig `r fig_dag`b).
 
@@
 <!---CONCLUSIONS---->
-We show that, in a tropical seasonal forest, the drought sensitivity of tree growth is jointly and interactively shaped by drought characteristics, species adaptations, tree size and microenvironment (Table 1).
+We show that in a tropical seasonal forest, the drought sensitivity of tree growth is jointly and interactively shaped by drought characteristics, species adaptations, tree size and microenvironment (Table 1).
 Analysing three droughts in a 14-year dendrometer band time series, we found lower-than-average growth during drought years, resulting in lower plot-level growth (Fig `r fig_drought_responses`). 
-Across droughts, there was little consistency in species- or individual-level drought responses, reflecting dissimilar effects of a key species trait (deciduousness) and variables linked to tree drought experience (TWI, DBH, CII) under differing drought conditions.
+There was little consistency in species- or individual-level responses across droughts, reflecting dissimilar effects of a key species trait (deciduousness) and variables linked to tree drought experience (TWI, DBH, CII) under differing drought conditions.
 Despite lower extremes of temperature, vapor pressure deficit, and number of dry days (Fig.
 2a), the severe drought in 2015 that peaked in the wet season elicited stronger responses across trees, especially deciduous species, which are expected to have stronger seasonality in growth phenology (Fig `r fig_species`).
-Interestingly, in 2015 and 2020 (but not 2010), evergreen and deciduous species had divergent responses, especially along a wetness gradient; deciduous species showed greater growth declines in wetter sites in 2015 while evergreen species grew less in drier sites in 2020 (Fig `r fig_species`b, Fig `r fig_microenvs`b).
+Interestingly, deciduous species had divergent responses in 2015 and 2020, especially along a wetness gradient; they showed greater growth declines in wetter than drier sites in 2015 and the reverse in 2020, while evergreen species showed greater declines in drier sites in both 2015 and 2020 (Fig `r fig_species`b, Fig `r fig_microenvs`b).
 Although trees with more exposed crowns tended to grow less under drought, additional effects of DBH -- presumably including rooting volume and depth -- were dissimilar between the drought events (Fig `r fig_microenvs`a).
@@
 Taken together, these results indicate that combined effects of drought characteristics, species traits, and individual microenvironments produce important heterogeneity in individual and species drought responses of tropical trees.
-Divergent effects of species and individuals to droughts may contribute to the maintenance of species diversity and ecosystem stability.
+Divergent effects of species and individuals to droughts may contribute to stabilizing ecosystem function and  maintaining species diversity.
 
@@
 -->
-However, our finding that a severe wet-season drought in 2015 had greater impact on woody growth than dry-season drought in 2010 and dry-to-wet-season drought in 2020 questions the generality of previous findings.
+Our finding that a severe wet-season drought in 2015 had greater impact on woody growth than dry-season drought in 2010 and dry-to-wet-season drought in 2020 is somewhat contrary to previous findings.
 Tree ring studies from this site [@vlam_temperature_2014;@anderson-teixeira_joint_2022] and across the tropics [@zuidema_tropical_2022] have asserted that tree growth is most sensitive to dry-season precipitation.
-However, the strength of the ecological response to the 2015 wet-season drought (Fig. `r fig_droughts`) does not align with this assumption. 
-Generalization of tropical tree growth responses likely requires fine-scale monitoring of growth (i.e., dendrometer band measurements or dendrochronology when possible) on species that span strategies and over different timings and intensities of droughts.
+However, the strength of the ecological response to the 2015 wet-season drought (Fig. `r fig_droughts`) does not align with this conslcusion. 
+Generalization of tropical tree growth responses likely requires fine-scale monitoring of growth (i.e., dendrometer band measurements or dendrochronology when possible) on species that span strategies and over different timings and intensities of droughts, together with concurrent measures of soil water avaialability.
 Advancing efforts to collect these data is critical to assessment and prediction of tropical forest responses to changing precipitation patterns.
@@
 <!----Limitations in generalisation/ species examined here may have greater diversity of drought adaptaitons than elsewhere in the tropics---->
-Periodic hot/dry conditions in tropical dry forests could create selective pressures for adaptative variation towards drought resistance potentially resulting in a greater diversity of drought responses compared to other bioclimatic zones where adaptive pressures and thus ecological variability are different.
-We demonstrate that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.
+Periodic hot/dry conditions in tropical dry forests could create selective pressures for adaptive variation towards drought resistance potentially resulting in a greater diversity of drought responses compared to other bioclimatic zones where adaptive pressures and thus ecological variability are different.
+We demonstrate that species identity and a key species adaptation (deciduousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.
 This aligns closely with recent global analyses using tree ring chronologies showing hetereogeneous responses of tree functional types to drought types [@zuidema_pantropical_2025].
@@
 temperate drylands reported in @vasey_intraspecific_2022), but may be distinct from the wet/everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different.
-Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different traits respond to different types of drought, and how these responses may be modified by microenvironments.
-Again, continued expansion of dendrometer band records and ecophysiological studies will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.
+Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@groenendijk_importance_2025], we know exceedingly little about how tropical tree species with different above- and below-ground traits along the hydraulic safety-efficiency spectrum respond to different types of drought, and how these responses may be modified by microenvironments.
+Again, continued expansion of dendrometer band and ecophysiological studies will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.
 <!--Therefore, we caution that our results suggestive of community-level insurance effects under drought be interpreted carefully across forest types that may be structured primarily by nutrient limitation, fires, herbivory or pathogens.
@@
 S5, Table S1).
+Extrapolating from such dendrochronological data, therefore, could lead to important errors if attributed to other species, potentially misrepresenting drought impacts conservation or silviculture, e.g.
+*Dipterocarpus alatus*, a conservation and commercial priority species, presents a different drought response (Table S1).
 More generally, tree-ring records likely present a biased picture of drought responses, as ring formation is caused by dry-season dormancy (often associated with deciduous leaf habit) and sampling often targets large trees with exposed crowns and dry microenvironments [@speer_fundamentals_2010]-- all factors shown here to <!--directly and interactively -->influence drought sensitivity.
@@
 Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in three different droughts, showing that ecological and evolutionary strategies can make particular species and trees resistant to some droughts but vulnerable to others.
-These distinct responses of species and individuals to droughts suggests potential biological insurance effects among tropical tree responses, in alignment with recent global analyses [@liu_drought_2022; @zuidema_pantropical_2025].
+These distinct responses of species and individuals to droughts suggests potential biological insurance effects [@loreau_biodiversity_2021] among tropical tree responses, in alignment with recent global analyses [@liu_drought_2022; @zuidema_pantropical_2025; @langan_amazon_2025].
 
 <!----Science is better now---->
-Our study reveals that drought sensitivity is not a static trait associated to particular species, traits, or microenvironments, but rather a dynamic response shaped interactively by these factors and drought characteristics.
+Our study reveals that drought sensitivity is not a simple, static function of species identity, traits, or microenvironments, but rather a complex, dynamic response shaped interactively by these factors and drought characteristics.
 Despite net decreased growth at the plot-level (in agreement with landscape-scale in @aguirre-gutiérrez_functional_2022 and @bennett_sensitivity_2023), we show evidence that the same strategies are not uniformly successful across spatiotemporal climatic variation.
-Static traits are often used to explain and predict forest responses to drought, especially in the tropics where long-term records are few [@guillemot_small_2022; @vico_how_2017a]. 
+Simple, static traits are often used to explain and predict forest responses to drought, especially in the tropics where long-term records are few [@guillemot_small_2022;@vico_how_2017a]. 
 <!--aligns with remote sensing based inferences across global forests @liu_drought_2022-->
-More inclusive frameworks for drought responses encompass ecological complexity and covarying traits with environment [e.g., @trugman_why_2021] but still assume consistent sensitivity of individual trees across events.
+More comprehensive frameworks for drought responses encompass ecological complexity and covarying traits with environment [e.g., @trugman_why_2021] but still assume consistent sensitivity of individual trees across events.
 However, we demonstrate that drought responses of tropical trees are not linearly predictable across drought events, but rather are shaped by the unique characteristics of each drought based on interactions between species adaptations and local environmental conditions.
-Without overriding the net negative effects of drought on tropical tree growth, these complexities reduce the likelihood of any given species or tree suffering severe growth declines under multiple consecutive droughts, thereby promoting diversity and stabilizing functioning of diverse tropical forests under drought.
+Without overriding the net negative effects of drought on tropical tree growth, these complexities reduce the likelihood of any given species or tree suffering severe growth declines under multiple consecutive droughts, thereby stabilizing functioning of diverse tropical forests under drought.
 
@@
 José Medina-Vega, Valentine Herrmann and Eugenie Más provided input on improving analyses. 
+David Bauman provided helpful discussions on causal inference modelling. 
 KA was supported by the Smithsonian Institution Postdoctoral Fellowship and the ForestGEO Analytical Workshop, Asia 2024 (supported by NSF grant DEB-2020424 to SJD) in completing this work.
@@
 (c) Standardised Precipitation Evapotranspiration Index (SPEI) values for the drought years at four different time scales for each month for each drought year.
-A k-month SPEI integrates over k months ending in the month current month.
+A k-month SPEI integrates over k months ending in the current month.
 Across all panels, light blue background represents the average wet season, from May to October.
@@
 ***Figure `r fig_drought_responses`: Growth increment timeseries and drought responses*** across trees and species in the Huai Kha Khaeng ForestGEO plot.
-a) Median of annualised growth increments across ten dominant species (colours) and all trees (black) from 2009 to 2022.
+a) Median of annualised growth increments for thirty dominant species analysed (light grey), all trees (black), and median of species medians (dark grey) from 2009 to 2022.
 b) Distribution of drought sensitivity of tree growth across all individuals in the three drought years (see main text for equation).
@@
 ***Figure `r fig_species`: Variation in sensitivity among species*** in the Huai Kha Khaeng ForestGEO plot during droughts in 2010, 2015 and 2020.
-a) Relationship of the mean species sensitivities from model fits of an intercept-only model, with mean species deciduousness values.
+a) Relationship of the mean species sensitivities (from model fits of an intercept-only model), with mean species deciduousness values.
 Line shown for significant correlation.
@@
 ***Figure `r fig_microenvs`: Drivers of variation in sensitivity among individuals***.
-All panels represent model results from models (one for each year) predicting sensitivity with microenvironment conditions with species random effect on all slopes.
+All panels represent model results from models (one for each drought year) predicting sensitivity with microenvironment conditions with species random effect on all slopes.
 a) Coefficient plots showing median effects and 90% CI for wetness, exposure and DBH of tree.

### 2025-10-10 doc/manuscript_hkk_drought_sensitivity.Rmd
updated files with small cosmetic fixes

@@
 **Number of Words in Main Text:** 5054   
-**Number of References:** 55  
+**Number of References:** 58  
 **Number of Figures:** 5  
@@
 <!-- The increment timeseries using dendrometer band measurements revealed plot- and species-level patterns in interannual growth variation (Fig `r fig_drought_responses`a).  -->
-The three climatically distinct droughts can be characterised as follows: 2010 was a moderate, dry-season drought, 2015 a severe wet-season drought and 2020, a moderate drought during the dry-to-wet transition (Figs `r fig_droughts`, S1, S2).
+The three climatically distinct droughts can be characterised as follows: 2010 was a moderate, dry-season drought, 2015 a severe wet-season drought and 2020, a moderate drought during the dry-to-wet transition (Figs `r fig_droughts`, S2, S3).
 
@@
 
-Species median sensitivities and individual tree sensitivities to the three droughts were not correlated with each other, supporting prediction (*i*) (Fig S5, Fig S6). 
+Species median sensitivities and individual tree sensitivities to the three droughts were not correlated with each other, supporting prediction (*i*) (Fig S6, Fig S7). 
 Median annual increment across years varied among the `r nrow(tree.time.sp)` species analysed, with *`r tree.time.sp$spfull[1]`* the lowest (`r tree.time.sp$inc_annual_med[1]`<!-- $\pm$ `r tree.time.sp$inc_annual_sd[1]`--> mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]`<!-- $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]`--> mm).
@@
 In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/3]`*, and in 2020 from `r sens.sp[sens.sp$yr==2020,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2020,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[nrow(sens.sp)/3]`*.
-Species had more negative sensitivities to the 2015 drought than the 2010 or 2020 droughts, in alignment with median sensitivities across all trees <!--(for 2010 and 2015 Pearson's correlation coefficient r = `r round(sens.sp.cor_10_15[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_15[3]$p.value,2)`, for 2015 and 2020, r = `r round(sens.sp.cor_15_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_15_20[3]$p.value,2)`, and for 2010 and 2020, r = `r round(sens.sp.cor_10_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_20[3]$p.value,2)`, -->(Fig S5).
+Species had more negative sensitivities to the 2015 drought than the 2010 or 2020 droughts, in alignment with median sensitivities across all trees <!--(for 2010 and 2015 Pearson's correlation coefficient r = `r round(sens.sp.cor_10_15[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_15[3]$p.value,2)`, for 2015 and 2020, r = `r round(sens.sp.cor_15_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_15_20[3]$p.value,2)`, and for 2010 and 2020, r = `r round(sens.sp.cor_10_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_20[3]$p.value,2)`, -->(Fig S6).
 Median species sensitivities to the 2010 and 2020 droughts were variable, but no species had higher median increments than a typical year in 2015.
@@
 Tree ring studies from this site [@vlam_temperature_2014;@anderson-teixeira_joint_2022] and across the tropics [@zuidema_tropical_2022] have asserted that tree growth is most sensitive to dry-season precipitation.
-However, the strength of the ecological response to the 2015 wet-season drought (Fig. `r fig_droughts`) does not align with this conslcusion. 
-Generalization of tropical tree growth responses likely requires fine-scale monitoring of growth (i.e., dendrometer band measurements or dendrochronology when possible) on species that span strategies and over different timings and intensities of droughts, together with concurrent measures of soil water avaialability.
+However, the strength of the ecological response to the 2015 wet-season drought (Fig. `r fig_droughts`) does not align with this conclusion. 
+Generalization of tropical tree growth responses likely requires fine-scale monitoring of growth (i.e., dendrometer band measurements or dendrochronology when possible) on species that span strategies and over different timings and intensities of droughts, together with concurrent measures of soil water availability.
 Advancing efforts to collect these data is critical to assessment and prediction of tropical forest responses to changing precipitation patterns.
@@
 An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings (n=4 at HKK; @vlam_temperature_2014) are unlikely to be representative of the drought responses of the entire community.
-For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the three events; it's growth nearly doubled in the 2010 and 2020 droughts but was severely reduced in the wet-season drought (Fig.
-S5, Table S1).
+For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the three events; it's growth nearly doubled in the 2010 and 2020 droughts but was severely reduced in the wet-season drought (Fig S6, Table S1).
 Extrapolating from such dendrochronological data, therefore, could lead to important errors if attributed to other species, potentially misrepresenting drought impacts conservation or silviculture, e.g.

### 2025-10-10 doc/supplementary_information.Rmd
updated files with small cosmetic fixes

@@
 
+## Study site
+
+![**Figure S1:** Representative photographs of the 50-ha long-term monitoring site in Huai Kha Khaeng Thailand, with dendrometer banded trees.
+Photo credits: left image: Sarayudh Bunyavejchewin, right image: Krishna Anujan](display/site_photos.png)
+
+\newpage
+
 ## Climate data and correlations
 
-![**Figure S1:** Raw climate data timeseries from 2009 to 2023 from ERA5Land and CHIRPS.
+![**Figure S2:** Raw climate data timeseries from 2009 to 2023 from ERA5Land and CHIRPS.
 Measurements represent daily rolling means over a 30-day time window for precipitation, max temperature and VPD and rolling sum for number of dry days.
-The shaded areas indicate the three drought years analysed - 2010, 2015 and 2020.](display/climvars_fullseries_remote.png)
+The shaded areas indicate the three drought years analysed - 2010, 2015 and 2020.
+Dry-season in the region is typically from November to April.](display/climvars_fullseries_remote.png)
 
-![**Figure S2:** Climate anomalies for all variables from 2009 to 2020.
+![**Figure S3:** Climate anomalies for all variables from 2009 to 2020.
 Values represent number of standard deviations from the long-term mean for daily rolling mean values.
-The shaded areas indicate the three drought years analysed - 2010, 2015 and 2020.](display/climvars_anomalies_fullseries_remote.png)
+The shaded areas indicate the three drought years analysed - 2010, 2015 and 2020.Dry-season in the region is typically from November to April.](display/climvars_anomalies_fullseries_remote.png)
 
@@
 
-![**Figure S3:** Growth anomalies across the timeseries.
+![**Figure S4:** Growth anomalies across the timeseries.
 Growth anomalies for each year calculated as the number of standard deviations in growth each year from mean growth across all years and summarised for plot and species.](display/growth_anom_fullseries.png)
@@
 
-![**Figure S4:** Species sensitivities in drought years 2010, 2015 and 2020.
+![**Figure S5:** Species sensitivities in drought years 2010, 2015 and 2020.
 Boxplots represent 25th, 50th (median) and 75th percentile of sensitivity for each species in each year and whiskers represent 95% CIs.
@@
 
-![**Figure S5:** Correlations of species sensitivities across the two years.
-Panel a shows correlation of species median sensitivities in the two years and Pearson's correlation coefficient.
+![**Figure S6:** Correlations of species sensitivities across the two years.
+The top-left panel shows correlation of species median sensitivities and Pearson's correlation coefficient in the 2010 and 2015, top right shows 2010 and 2020, and bottom left shows 2015 and 2020.
 Colour gradient represents deciduousness scale and dotted line is the 1:1 line.
-Panel b shows difference in magnitude of species median sensitivites among the two years](display/Fig_SI_sensitivity_cor_dec.png)
+The bottom right panel shows how species drought sensitivites varies among the three years; the color scheme for droughts matches that in main text figure 1.](display/Fig_SI_sensitivity_cor_dec.png)
 
@@
 
-![**Figure S6:** Correlation of individual sensitivities across the two years along with Pearson's correlation coefficient and the 1:1 line.](display/sens_corr.png)
+![**Figure S7:** Correlation of individual sensitivities across the two years along with Pearson's correlation coefficient and the 1:1 line.](display/sens_corr.png)
 
@@
 
-![**Figure S7:** Map of trees with dendrobands in HKK, included in analyses in the 2010, 2015 and 2020 datasets](display/dendro_tree_map.png)
+![**Figure S8:** Map of trees with dendrobands in HKK, included in analyses in the 2010, 2015 and 2020 datasets](display/dendro_tree_map.png)
 
-![**Figure S8:** Calculated Topographic Wetness Index (TWI) values for the HKK plot](display/twi.png)
+![**Figure S9:** Calculated Topographic Wetness Index (TWI) values for the HKK plot](display/twi.png)
 
-![**Figure S9:** Distribution of DBH with Crown Illumination Index (CII) in 2014.](display/dbh_cii.png)
+![**Figure S10:** Distribution of DBH with Crown Illumination Index (CII) in 2014.](display/dbh_cii.png)
 
@@
 
-![**Figure S10:** Distribution of raw sensitivity values across years and their relationship with DBH, CII and TWI. Lines are linear model fits.](display/sens_var_rels.png)
+![**Figure S11:** Distribution of raw sensitivity values across years and their relationship with DBH, CII and TWI.
+Blue lines are linear model fits for statistically significant relationships at p<0.05.
+Red lines are Friedman's SuperSmoother fits for all.](display/sens_var_rels.png)
 
-![**Figure S11:** Distribution of residuals of species intercept models of sensitivity and their relationship with scaled DBH, TWI and CII and deciduousness.
+![**Figure S12:** Distribution of residuals of species intercept models of sensitivity and their relationship with scaled DBH, TWI and CII and deciduousness.
+Grey lines are linear model fits for statistically significant relationships at p<0.05.
 Species intercept model results are reported in main text.](display/residuals.png)
@@
 To analyse the influence of microenvironmental variables on growth sensitivity, we first created a Directed Acyclic Graph describing the relationships. 
-To be able to test the DAG with the dataset, we used *dagitty* to derive the conditional dependencies that need to be tested, namely CII $\perp\!\!\!\perp$ TWI and DBH $\perp\!\!\!\perp$ TWI.
+The consistency of the DAG with the dataset, and our ability build causally interpretable models, is conditional on the numerical independence among some specific variable groups, as implied by the DAG structure.
+We identified these "testable implications" or "conditional independences" for this DAG using *dagitty* as CII $\perp\!\!\!\perp$ TWI and DBH $\perp\!\!\!\perp$ TWI.
+We tested these correlations using the function *cor()* in R, with the simple criteria that a Pearson's correlation coefficient <0.3 implies independence among variable pairs and causal relevance of the DAG.
 
-![**Figure S12:** Testing correlations of variables across all individuals](display/cond_dep_alltrees.png)
+![**Figure S13:** Testing correlations of variables across all individuals](display/cond_dep_alltrees.png)
 
@@
 
-![**Figure S13:** Testing correlations for DBH $\perp\!\!\!\perp$ TWI by species](display/cond_dep_dbh_twi.png)
+![**Figure S14:** Testing correlations for DBH $\perp\!\!\!\perp$ TWI by species](display/cond_dep_dbh_twi.png)
 
-![**Figure S14:** Testing correlations for CII $\perp\!\!\!\perp$ TWI by species](display/cond_dep_cii_twi.png)
+![**Figure S15:** Testing correlations for CII $\perp\!\!\!\perp$ TWI by species](display/cond_dep_cii_twi.png)
 
@@
 
-![**Figure S15:** Correlation of modelled species sensitivities with Topographic Position Index](display/FigSI_TPI_preds.png)
+![**Figure S16:** Correlation of modelled species sensitivities with Topographic Position Index](display/FigSI_TPI_preds.png)
 
@@
 
-![**Figure S16:** Model results from Bayesian causal model of sensitivity as a function of DBH, CII and TPI.](display/Fig4_tpi_3yrs.png)
+![**Figure S17:** Model results from Bayesian causal model of sensitivity as a function of DBH, CII and TPI.](display/Fig4_tpi_3yrs.png)
 
@@
 
-![**Figure S17:** Model results from Bayesian causal model of sensitivity as a function of DBH, CII and TWI with species random effect only on intercept.](display/FigSI_coefs_nospre.png)
+![**Figure S18:** Model results from Bayesian causal model of sensitivity as a function of DBH, CII and TWI with species random effect only on intercept.](display/FigSI_coefs_nospre.png)
 
@@
 
-![**Figure S18:** Model results from three independent causal inquiry models of sensitivity following backdoor criteria for total causal effect of DBH, CII and TWI with species random effect on intercept and slope.](display/causal_inquiry.png)
+![**Figure S19:** Model results from three independent causal inquiry models of sensitivity following backdoor criteria for total causal effect of DBH, CII and TWI with species random effect on intercept and slope.](display/causal_inquiry.png)
 

### 2025-10-10 doc/manuscript_hkk_drought_sensitivity.Rmd
cosmetic change

@@
 ***Figure `r fig_drought_responses`: Growth increment timeseries and drought responses*** across trees and species in the Huai Kha Khaeng ForestGEO plot.
-a) Median of annualised growth increments for thirty dominant species analysed (light grey), all trees (black), and median of species medians (dark grey) from 2009 to 2022.
+a) Median of annualised growth increments for all trees of the thirty species analysed (black), and median of species medians (dark grey) from 2009 to 2022.
 b) Distribution of drought sensitivity of tree growth across all individuals in the three drought years (see main text for equation).

### 2025-10-10 doc/response_to_reviewer_R1.Rmd
updated files with small cosmetic fixes
cosmetic changes

@@
 
-We are pleased to submit a revised version of our manuscript, "Drought response is not a species trait: tropical tree drought sensitivity is shaped by drought characteristics, species adaptations, and individual microenvironments" () for consideration for publication in Ecology Letters.
+We are pleased to submit a revised version of our manuscript, "Drought response is not a species trait: tropical tree drought sensitivity is shaped by drought characteristics, species adaptations, and individual microenvironments" (ELE-00826-2025) for consideration for publication in Ecology Letters.
 We have addressed all points raised by the reviewers as detailed in our responses to the reviewers. 
@@
 We have also updated Figures 1 and 2 as well as the relevant figures in the supplementary material as requested by the reviewers.
+Finally, following a friendly review by a collaborator, we also present an alternate, analogous set of causal inquiry models in the supplementary information that supports the conclusions in the main text.
 
@@
 And so, reporting the number per deciduousness group would require arbitrarily discretising this continuous variable.
-Instead, we now report a supplementary figure showing the distribution of deciduousness among the selected species.**
+Instead, we now report a supplementary figure (Figure S9) showing the distribution of deciduousness among the selected species.**
 
@@
 
-**"An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings (n=4 at HKK; @vlam_temperature_2014) are unlikely to be representative of the drought responses of the entire community.
+**L421-430: "An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings (n=4 at HKK; @vlam_temperature_2014) are unlikely to be representative of the drought responses of the entire community.
 For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the three events; it's growth nearly doubled in the 2010 and 2020 droughts but was severely reduced in the wet-season drought (Fig.
 S5, Table S1).
-Extrapolating from these could lead to erroneous inferences on species with conservation or commercial significance, e.g.
-*Dipterocarpus alatus* which presents a different drought response (Table S1)."**
+Extrapolating from such dendrochronological data, therefore, could lead to important errors if attributed to other species, potentially misrepresenting drought impacts conservation or silviculture, e.g.
+*Dipterocarpus alatus*, a conservation and commercial priority species, presents a different drought response (Table S1)."**
 
@@
 
-**"Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different above- and below-ground traits along the hydraulic safety-efficiency spectrum respond to different types of drought, and how these responses may be modified by microenvironments."**
+**L400-404: "Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different above- and below-ground traits along the hydraulic safety-efficiency spectrum respond to different types of drought, and how these responses may be modified by microenvironments."**
 
@@
 
-**"Predicted effects of CII on sensitivity decreased monotonically, with strong decreases in 2010, weak decrease in 2015, and no effect in 2020."**
+**L340: "Predicted effects of CII on sensitivity decreased monotonically, with strong decreases in 2010, weak decrease in 2015, and no effect in 2020."**
 
@@
 We have now added dashed lines for the significant correlations.
-Following the same logic, for S11 and S12, we have now removed lines for the non-significant correlations.**
+Following the same logic, for S11 and S12, we have now removed lines for the non-significant correlations, but added Friedman's SuperSmoother fits for all plots to visually demonstrate the direction of the relationship.**
 

### 2025-10-10 doc/supplementary_information.Rmd
cosmetic changes

@@
 
-Map of trees
-
 ![**Figure S8:** Map of trees with dendrobands in HKK, included in analyses in the 2010, 2015 and 2020 datasets](display/dendro_tree_map.png)
 
-![**Figure S9:** Calculated Topographic Wetness Index (TWI) values for the HKK plot](display/twi.png)
+![**Figure S9:** Distribution of deciduousness values among the 30 species analysed](display/Fig_dec_dist.png)
 
-![**Figure S10:** Distribution of DBH with Crown Illumination Index (CII) in 2014.](display/dbh_cii.png)
+![**Figure S10:** Calculated Topographic Wetness Index (TWI) values for the HKK plot](display/twi.png)
+
+![**Figure S11:** Distribution of DBH with Crown Illumination Index (CII) in 2014.](display/dbh_cii.png)
 
@@
 
-![**Figure S11:** Distribution of raw sensitivity values across years and their relationship with DBH, CII and TWI.
+![**Figure S12:** Distribution of raw sensitivity values across years and their relationship with DBH, CII and TWI.
 Blue lines are linear model fits for statistically significant relationships at p<0.05.
@@
 
-![**Figure S12:** Distribution of residuals of species intercept models of sensitivity and their relationship with scaled DBH, TWI and CII and deciduousness.
+![**Figure S13:** Distribution of residuals of species intercept models of sensitivity and their relationship with scaled DBH, TWI and CII and deciduousness.
 Grey lines are linear model fits for statistically significant relationships at p<0.05.
@@
 
-![**Figure S13:** Testing correlations of variables across all individuals](display/cond_dep_alltrees.png)
+![**Figure S14:** Testing correlations of variables across all individuals](display/cond_dep_alltrees.png)
 
@@
 
-![**Figure S14:** Testing correlations for DBH $\perp\!\!\!\perp$ TWI by species](display/cond_dep_dbh_twi.png)
+![**Figure S15:** Testing correlations for DBH $\perp\!\!\!\perp$ TWI by species](display/cond_dep_dbh_twi.png)
 
-![**Figure S15:** Testing correlations for CII $\perp\!\!\!\perp$ TWI by species](display/cond_dep_cii_twi.png)
+![**Figure S16:** Testing correlations for CII $\perp\!\!\!\perp$ TWI by species](display/cond_dep_cii_twi.png)
 
@@
 
-![**Figure S16:** Correlation of modelled species sensitivities with Topographic Position Index](display/FigSI_TPI_preds.png)
+![**Figure S17:** Correlation of modelled species sensitivities with Topographic Position Index](display/FigSI_TPI_preds.png)
 
@@
 
-![**Figure S17:** Model results from Bayesian causal model of sensitivity as a function of DBH, CII and TPI.](display/Fig4_tpi_3yrs.png)
+![**Figure S18:** Model results from Bayesian causal model of sensitivity as a function of DBH, CII and TPI.](display/Fig4_tpi_3yrs.png)
 
@@
 
-![**Figure S18:** Model results from Bayesian causal model of sensitivity as a function of DBH, CII and TWI with species random effect only on intercept.](display/FigSI_coefs_nospre.png)
+![**Figure S19:** Model results from Bayesian causal model of sensitivity as a function of DBH, CII and TWI with species random effect only on intercept.](display/FigSI_coefs_nospre.png)
 
@@
 
-![**Figure S19:** Model results from three independent causal inquiry models of sensitivity following backdoor criteria for total causal effect of DBH, CII and TWI with species random effect on intercept and slope.](display/causal_inquiry.png)
+![**Figure S20:** Model results from three independent causal inquiry models of sensitivity following backdoor criteria for total causal effect of DBH, CII and TWI with species random effect on intercept and slope.](display/causal_inquiry.png)
 

### 2025-10-31 scripts/01_prepdata.R
reordered scripts to reduce code redundancy

@@ -495,272 +495,85 @@ sp_vars <- merge(sp.traits, dplyr::select(dec_williams, c("sp", "spfull", "willi
 # make an RData object with these dataframes
 saveRDS(list(tree_vars = tree_vars, tree.time = tree.time, sp_vars = sp_vars), "data/HKK-dendro/sensitivity_data.RData")
 
-# plot distribution for species with least and most spread
-max.spread <- sp_vars %>%
-  arrange(twi_sd) %>%
-  tail(1) %>%
-  pull(sp)
 
-min.spread <- sp_vars %>%
-  arrange(twi_sd) %>%
-  head(1) %>%
-  pull(sp)
+rm(list = ls())
+datasets <- readRDS("data/HKK-dendro/sensitivity_data.RData")
 
+tree.time <- datasets$tree.time
+tree_vars <- datasets$tree_vars
+sp_vars <- datasets$sp_vars
 
-png("doc/display/explore/twi_violin_extreme.png", width = 5, height = 5, units = "in", res = 300)
-ggplot(hkk.stem4 %>% filter(sp %in% c(min.spread, max.spread)), aes(x = twi, fill = sp)) +
-  # geom_violin() +
-  # geom_boxplot(width = 0.1) +
-  geom_density(alpha = 0.5) +
-  scale_fill_viridis_d() +
-  theme_minimal() +
-  labs(title = "Distribution of species with least and most spread \nin TWI space (HKK)")
-dev.off()
+colnames(sp_vars)[1] <- "Species"
 
-# plot distribution for species with least and most spread - dbh weighted
+# merge tree_vars and sp_vars
+tree_vars <- merge(tree_vars, sp_vars, by = "Species", all.x = TRUE)
+# merge tree_vars and tree.time
+tree.time <- merge(tree.time, tree_vars, by = "Tag", all.x = TRUE)
 
-max.spread <- sp_vars %>%
-  arrange(twi_dbh_sd) %>%
-  tail(1) %>%
-  pull(sp)
+colnames(tree.time)
 
-min.spread <- sp_vars %>%
-  arrange(twi_dbh_sd) %>%
-  head(1) %>%
-  pull(sp)
+colours <- c("#e15f41", "#546de5", "#f7b731")
 
-png("doc/display/explore/twi_violin_extreme_dbh.png", width = 5, height = 5, units = "in", res = 300)
-ggplot(hkk.stem4 %>% filter(sp %in% c(min.spread, max.spread)), aes(x = twi, fill = sp)) +
-  # geom_violin() +
[…]

### 2025-10-31 scripts/02_species.R

@@ -4,52 +4,9 @@ library(brms)
 
 # load data--------------------------
 rm(list = ls())
-datasets <- readRDS("data/HKK-dendro/sensitivity_data.RData")
+tree.time <- read.csv("data/HKK-dendro/to_publish/sensitivity_dataset_for_review.csv")
+median_incs <- read.csv("data/HKK-dendro/to_publish/summaries_dataset_for_review.csv")
 
-tree.time <- datasets$tree.time
-tree_vars <- datasets$tree_vars
-sp_vars <- datasets$sp_vars
-
-colnames(sp_vars)[1] <- "Species"
-
-# merge tree_vars and sp_vars
-tree_vars <- merge(tree_vars, sp_vars, by = "Species", all.x = TRUE)
-# merge tree_vars and tree.time
-tree.time <- merge(tree.time, tree_vars, by = "Tag", all.x = TRUE)
-
-colnames(tree.time)
-
-# colours <- c("#e15f41", "#546de5", "#f7b731")
-
-# standardise the variables within species and across all--------------------
-
-tree.time <- tree.time %>%
-    dplyr::mutate(
-        calcDBH_min1_scaled = scale(calcDBH_min1, center = TRUE, scale = TRUE),
-        cii_min1_scaled = scale(cii_min1, center = TRUE, scale = TRUE),
-        cii_min1_scaled = ifelse(cii_min1_scaled == "NaN", 0, cii_min1_scaled),
-        twi_scaled = scale(twi, center = TRUE, scale = TRUE),
-        tpi_scaled = scale(tpi, center = TRUE, scale = TRUE),
-        median_inc_scaled = scale(median_inc, center = TRUE, scale = TRUE),
-        avg_inc_tree_scaled = scale(avg_inc_tree, center = TRUE, scale = TRUE)
-    ) %>%
-    group_by(Species) %>%
-    # scale while retaining original values
-    dplyr::mutate(
-        calcDBH_min1_scaled_sp = scale(calcDBH_min1, center = TRUE, scale = TRUE),
-        cii_min1_scaled_sp = scale(cii_min1, center = TRUE, scale = TRUE),
-        cii_min1_scaled_sp = ifelse(cii_min1_scaled_sp == "NaN", 0, cii_min1_scaled_sp),
-        twi_scaled_sp = scale(twi, center = TRUE, scale = TRUE),
[…]

### 2025-10-31 scripts/03_orderedcii.R

@@ -6,105 +6,59 @@ library(brms)
 
 # load data--------------------------
 rm(list = ls())
-datasets <- readRDS("data/HKK-dendro/sensitivity_data.RData")
-
-tree.time <- datasets$tree.time
-tree_vars <- datasets$tree_vars
-sp_vars <- datasets$sp_vars
-
-colnames(sp_vars)[1] <- "Species"
-
-# merge tree_vars and sp_vars
-tree_vars <- merge(tree_vars, sp_vars, by = "Species", all.x = TRUE)
-# merge tree_vars and tree.time
-tree.time <- merge(tree.time, tree_vars, by = "Tag", all.x = TRUE)
-
-colnames(tree.time)
-
-colours <- c("#e15f41", "#546de5", "#f7b731")
-
-# standardise the variables within species and across all--------------------
-yrs <- c(2010, 2015, 2020)
-
-tree.time <- tree.time %>%
-    dplyr::mutate(
-        calcDBH_min1_scaled = scale(calcDBH_min1, center = TRUE, scale = TRUE),
-        cii_min1_scaled = scale(cii_min1, center = TRUE, scale = TRUE),
-        cii_min1_scaled = ifelse(cii_min1_scaled == "NaN", 0, cii_min1_scaled),
-        twi_scaled = scale(twi, center = TRUE, scale = TRUE),
-        tpi_scaled = scale(tpi, center = TRUE, scale = TRUE),
-        median_inc_scaled = scale(median_inc, center = TRUE, scale = TRUE),
-        avg_inc_tree_scaled = scale(avg_inc_tree, center = TRUE, scale = TRUE)
-    ) %>%
-    group_by(Species) %>%
-    # scale while retaining original values
-    dplyr::mutate(
-        calcDBH_min1_scaled_sp = scale(calcDBH_min1, center = TRUE, scale = TRUE),
-        cii_min1_scaled_sp = scale(cii_min1, center = TRUE, scale = TRUE),
-        cii_min1_scaled_sp = ifelse(cii_min1_scaled_sp == "NaN", 0, cii_min1_scaled_sp),
-        twi_scaled_sp = scale(twi, center = TRUE, scale = TRUE),
-        tpi_scaled_sp = scale(tpi, center = TRUE, scale = TRUE)
-    ) %>%
-    ungroup() %>%
-    # remove large outliers for each year
-    group_by(yr) %>%
[…]

### 2025-10-31 scripts/05_figures.R

@@ -14,164 +14,9 @@ library(patchwork)
 # read data
 
 spei <- read.csv("data/climate/SPEI_HKK_from_GEE.csv")
-ffstation <- read.csv("data/climate/WeatherData_ALL_forest_fire_research_station.csv")
-
-# enso <- read.table("data/climate/ENSO_index_meiv2.data", header = F, skip = 1)
-head(ffstation)
-# calculate vpd using ffstation data
-ffstation <- ffstation %>%
-    mutate(
-        vpdmax = 0.6108 * exp(17.27 * TempMax / (TempMax + 237.3)) * (1 - (Relative.Humidity / 100)),
-        vpdmin = 0.6108 * exp(17.27 * TempMin / (TempMin + 237.3)) * (1 - (Relative.Humidity / 100)),
-        vpdmean = 0.6108 * exp(17.27 * TempMean / (TempMean + 237.3)) * (1 - (Relative.Humidity / 100)),
-        dry_days = ifelse(Precipitation == 0, 1, 0)
-    )
-
-# make a figure with rolling monthly means
-ffstation_monthly_rl <- ffstation %>%
-    pivot_longer(cols = c(Precipitation, dry_days, TempMax, vpdmax, vpdmin, vpdmean), names_to = "climvar", values_to = "value") %>%
-    select(YearII, Month, Date, climvar, value) %>%
-    arrange(YearII, Month, Date, climvar) %>%
-    group_by(climvar) %>%
-    dplyr::mutate(
-        rlsum = zoo::rollsum(value, k = 30, fill = NA, align = "center"),
-        rlmean = zoo::rollmean(value, k = 30, fill = NA, align = "center"),
-        rlval = ifelse(climvar == c("Precipitation"), rlsum, rlmean)
-    ) %>%
-    rename(year = YearII, month = Month) %>%
-    dplyr::mutate(year = as.character(year)) %>%
-    dplyr::mutate(
-        Date2 = as.Date(paste0(year, "-", month, "-", Date)),
-        day_of_year = as.numeric(format(Date2, "%j"))
-    ) %>%
-    select(-c(rlsum, rlmean))
-
-# ffstation_lt_rl <- ffstation_monthly_rl %>%
-#     filter(year <= 2021, year >= 2009) %>%
-#     ungroup() %>%
-#     group_by(day_of_year, climvar) %>%
-#     dplyr::summarise(
-#         # rlse = sd(rlval, na.rm = T) / sqrt(sum(!is.na(rlval))),
[…]

### 2025-10-31 scripts/07_otherfigs.R

@@ -1,0 +1,438 @@
+# make a plot with anomalies instead of growth increments
+
+# first summarise anomalies by species
+tree.time.anom <- tree.time %>%
+    group_by(spname, yr) %>%
+    dplyr::summarise(
+        median_inc = median(inc_annual, na.rm = T)
+    ) %>%
+    ungroup() %>%
+    group_by(spname) %>%
+    dplyr::mutate(
+        mean = mean(median_inc, na.rm = T),
+        sd = sd(median_inc, na.rm = T),
+        anomaly = (median_inc - mean) / sd
+    ) %>%
+    ungroup()
+
+tree.time.anom.all <- tree.time %>%
+    group_by(yr) %>%
+    dplyr::summarise(
+        median_inc = median(inc_annual, na.rm = T)
+    ) %>%
+    dplyr::mutate(
+        mean = mean(median_inc, na.rm = T),
+        sd = sd(median_inc, na.rm = T),
+        anomaly = (median_inc - mean) / sd
+    ) %>%
+    ungroup() %>%
+    dplyr::mutate(spname = "all")
+
+tree.time.anom <- bind_rows(tree.time.anom, tree.time.anom.all)
+
+# bar plot of anomalies
+
+spagplot_top10_anom <- ggplot() +
+    # species plots
+    geom_bar(
+        data = tree.time.anom %>% filter(spname %in% top_10_sp$spfull),
+        aes(x = yr, y = anomaly, fill = spname), stat = "identity",
+        position = "dodge"
+    ) +
+    # mean of all trees
+    geom_bar(
+        data = tree.time.anom %>% filter(spname == "all"),
+        aes(x = yr, y = anomaly), fill = NA, col = "black", linewidth = 1, stat = "identity",
+        position = "dodge"
+    ) +
+    # make all years show on x-axis
+    scale_x_continuous(breaks = unique(tree.time$yr)) +
+    scale_fill_viridis_d() +
+    geom_hline(yintercept = c(-2, -1, 0, 1, 2), linetype = "dashed") +
+    # add text on these lines
+    geom_text(aes(
+        x = c(2010, 2015, 2020), y = 1,
+        label = c("ENSO drought", "ENSO drought", "drought")
+    ), hjust = 0.8, vjust = -0.2, angle = 90) +
+    guides(fill = guide_legend("species"), nrow = 3) +
[…]

### 2025-10-31 doc/manuscript_hkk_drought_sensitivity.Rmd
cosmetic edit to figure legend
small updates to code, README and statement

@@
 # Data Accessibility Statement {-}
-This manuscript does not use novel code. 
-The dendrometer band data used for this manuscript is archived in the ForestGEO data repository and will be made available upon request. 
-The climate data used in this work are publicly available. 
-The code used to analyse the data is archived in Zenodo with doi:10.5281/zenodo.15777974.
+<!-- This manuscript does not use novel code.  -->
+<!-- The dendrometer band data used for this manuscript is archived in the ForestGEO data repository and will be made available upon request.  -->
+<!-- The climate data used in this work are publicly available.  -->
+Data and code required to reproduce the results and figures are archived in Zenodo with doi:10.5281/zenodo.15777974.
 
@@
 Climatic characteristics of the focal drought years compared with the average across years*** in the Huai Kha Khaeng ForestGEO plot, for four daily climate variables: dry days, precipitation, maximum temperature and VPD.
-(a) 30-day running means.
+(a) 30-day rolling means (in colour for each year) along with their long-term mean and standard deviation (in grey).
 (b) Anomalies of the drought year running means from the 2008-2019 averages, in units of standard deviations.

### 2025-10-31 scripts/02_species.R

@@ -35,7 +35,7 @@ sens_sp <- tree.time %>%
 
 head(sens_sp)
 
-sens.sp <- merge(sens_sp, sp_vars, by = "Species", all.x = TRUE)
+sens.sp <- merge(sens_sp, median_incs, by = "Species", all.x = TRUE)
 
 # plot median sensitivity against deciduousness
 sens_sp_median <- ggplot(sens.sp, aes(x = williams_dec, y = sens_median)) +

### 2025-10-31 scripts/07_otherfigs.R

@@ -1,3 +1,58 @@
+# load libraries
+library(tidyverse)
+library(brms)
+library(ggplot2)
+
+# read data
+rm(list = ls())
+datasets <- readRDS("data/HKK-dendro/sensitivity_data.RData")
+
+tree.time <- datasets$tree.time
+tree_vars <- datasets$tree_vars
+sp_vars <- datasets$sp_vars
+
+colnames(sp_vars)[1] <- "Species"
+
+# merge tree_vars and sp_vars
+tree_vars <- merge(tree_vars, sp_vars, by = "Species", all.x = TRUE)
+# merge tree_vars and tree.time
+tree.time <- merge(tree.time, tree_vars, by = "Tag", all.x = TRUE)
+
+colnames(tree.time)
+
+colours <- c("#e15f41", "#546de5", "#f7b731")
+
+# standardise the variables within species and across all--------------------
+yrs <- c(2010, 2015, 2020)
+
+tree.time <- tree.time %>%
+    dplyr::mutate(
+        calcDBH_min1_scaled = scale(calcDBH_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled = scale(cii_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled = ifelse(cii_min1_scaled == "NaN", 0, cii_min1_scaled),
+        twi_scaled = scale(twi, center = TRUE, scale = TRUE),
+        tpi_scaled = scale(tpi, center = TRUE, scale = TRUE),
+        median_inc_scaled = scale(median_inc, center = TRUE, scale = TRUE),
+        avg_inc_tree_scaled = scale(avg_inc_tree, center = TRUE, scale = TRUE)
+    ) %>%
+    group_by(Species) %>%
+    # scale while retaining original values
+    dplyr::mutate(
+        calcDBH_min1_scaled_sp = scale(calcDBH_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled_sp = scale(cii_min1, center = TRUE, scale = TRUE),
+        cii_min1_scaled_sp = ifelse(cii_min1_scaled_sp == "NaN", 0, cii_min1_scaled_sp),
+        twi_scaled_sp = scale(twi, center = TRUE, scale = TRUE),
+        tpi_scaled_sp = scale(tpi, center = TRUE, scale = TRUE)
+    ) %>%
+    ungroup() %>%
[…]

### 2025-11-01 scripts/01_prepdata.R
data push and associated changes to scripts

@@ -555,7 +555,7 @@ tree.time.public <- tree.time %>%
 colnames(tree.time.public)
 unique(tree.time.public$yr)
 
-write.csv(tree.time.public, "data/HKK-dendro/to_publish/sensitivity_dataset_for_review.csv", row.names = F)
+write.csv(tree.time.public, "data/dendro/sensitivity_dataset.csv", row.names = F)
 
 median_incs <- tree.time %>%
   group_by(yr) %>%
@@ -576,4 +576,4 @@ median_incs$median_inc_sp_mm <- median_inc_sp$median_inc
 
 median_incs
 
-write.csv(median_incs, "data/HKK-dendro/to_publish/summaries_for_review.csv", row.names = F)
+write.csv(median_incs, "data/dendro/summaries_dataset.csv", row.names = F)

### 2025-11-01 scripts/02_species.R

@@ -4,8 +4,8 @@ library(brms)
 
 # load data--------------------------
 rm(list = ls())
-tree.time <- read.csv("data/HKK-dendro/to_publish/sensitivity_dataset_for_review.csv")
-median_incs <- read.csv("data/HKK-dendro/to_publish/summaries_dataset_for_review.csv")
+tree.time <- read.csv("data/dendro/sensitivity_dataset.csv")
+median_incs <- read.csv("data/dendro/summaries_dataset.csv")
 
 
 # plot the sensitivity values for all trees in 2010 and 2015

### 2025-11-01 scripts/03_orderedcii.R

@@ -6,7 +6,7 @@ library(brms)
 
 # load data--------------------------
 rm(list = ls())
-tree.time <- read.csv("data/HKK-dendro/to_publish/sensitivity_dataset_for_review.csv")
+tree.time <- read.csv("data/dendro/sensitivity_dataset.csv")
 
 
 # conditional dependencies---------------

### 2025-11-01 scripts/04_orderedcii_tpi.R

@@ -6,52 +6,7 @@ library(brms)
 
 # load data--------------------------
 rm(list = ls())
-datasets <- readRDS("data/HKK-dendro/sensitivity_data.RData")
-
-tree.time <- datasets$tree.time
-tree_vars <- datasets$tree_vars
-sp_vars <- datasets$sp_vars
-
-colnames(sp_vars)[1] <- "Species"
-
-# merge tree_vars and sp_vars
-tree_vars <- merge(tree_vars, sp_vars, by = "Species", all.x = TRUE)
-# merge tree_vars and tree.time
-tree.time <- merge(tree.time, tree_vars, by = "Tag", all.x = TRUE)
-
-colnames(tree.time)
-
-
-# standardise the variables within species and across all--------------------
-yrs <- c(2010, 2015, 2020)
-
-tree.time <- tree.time %>%
-    dplyr::mutate(
-        calcDBH_min1_scaled = scale(calcDBH_min1, center = TRUE, scale = TRUE),
-        cii_min1_scaled = scale(cii_min1, center = TRUE, scale = TRUE),
-        cii_min1_scaled = ifelse(cii_min1_scaled == "NaN", 0, cii_min1_scaled),
-        twi_scaled = scale(twi, center = TRUE, scale = TRUE),
-        tpi_scaled = scale(tpi, center = TRUE, scale = TRUE),
-        median_inc_scaled = scale(median_inc, center = TRUE, scale = TRUE),
-        avg_inc_tree_scaled = scale(avg_inc_tree, center = TRUE, scale = TRUE)
-    ) %>%
-    group_by(Species) %>%
-    # scale while retaining original values
-    dplyr::mutate(
-        calcDBH_min1_scaled_sp = scale(calcDBH_min1, center = TRUE, scale = TRUE),
-        cii_min1_scaled_sp = scale(cii_min1, center = TRUE, scale = TRUE),
-        cii_min1_scaled_sp = ifelse(cii_min1_scaled_sp == "NaN", 0, cii_min1_scaled_sp),
-        twi_scaled_sp = scale(twi, center = TRUE, scale = TRUE),
-        tpi_scaled_sp = scale(tpi, center = TRUE, scale = TRUE)
-    ) %>%
-    ungroup() %>%
-    # remove large outliers for each year
-    group_by(yr) %>%
[…]

### 2025-11-01 scripts/05_figures.R

@@ -357,8 +357,8 @@ dev.off()
 library(tidyverse)
 
 # load data--------------------------
-tree.time <- read.csv("data/HKK-dendro/to_publish/sensitivity_dataset_for_review.csv")
-median_incs <- read.csv("data/HKK-dendro/to_publish/summaries_dataset_for_review.csv")
+tree.time <- read.csv("data/dendro/sensitivity_dataset.csv")
+median_incs <- read.csv("data/dendro/summaries_dataset.csv")
 
 # plot the sensitivity values for all trees in 2010 and 2015
 yrs <- c(2010, 2015, 2020)

### 2025-11-01 scripts/06_causal.R

@@ -6,53 +6,8 @@ library(brms)
 
 # load data--------------------------
 rm(list = ls())
-datasets <- readRDS("data/HKK-dendro/sensitivity_data.RData")
-
-tree.time <- datasets$tree.time
-tree_vars <- datasets$tree_vars
-sp_vars <- datasets$sp_vars
-
-colnames(sp_vars)[1] <- "Species"
-
-# merge tree_vars and sp_vars
-tree_vars <- merge(tree_vars, sp_vars, by = "Species", all.x = TRUE)
-# merge tree_vars and tree.time
-tree.time <- merge(tree.time, tree_vars, by = "Tag", all.x = TRUE)
-
-colnames(tree.time)
-
-colours <- c("#e15f41", "#546de5", "#f7b731")
-
-# standardise the variables within species and across all--------------------
-yrs <- c(2010, 2015, 2020)
-
-tree.time <- tree.time %>%
-    dplyr::mutate(
-        calcDBH_min1_scaled = scale(calcDBH_min1, center = TRUE, scale = TRUE),
-        cii_min1_scaled = scale(cii_min1, center = TRUE, scale = TRUE),
-        cii_min1_scaled = ifelse(cii_min1_scaled == "NaN", 0, cii_min1_scaled),
-        twi_scaled = scale(twi, center = TRUE, scale = TRUE),
-        tpi_scaled = scale(tpi, center = TRUE, scale = TRUE),
-        median_inc_scaled = scale(median_inc, center = TRUE, scale = TRUE),
-        avg_inc_tree_scaled = scale(avg_inc_tree, center = TRUE, scale = TRUE)
-    ) %>%
-    group_by(Species) %>%
-    # scale while retaining original values
-    dplyr::mutate(
-        calcDBH_min1_scaled_sp = scale(calcDBH_min1, center = TRUE, scale = TRUE),
-        cii_min1_scaled_sp = scale(cii_min1, center = TRUE, scale = TRUE),
-        cii_min1_scaled_sp = ifelse(cii_min1_scaled_sp == "NaN", 0, cii_min1_scaled_sp),
-        twi_scaled_sp = scale(twi, center = TRUE, scale = TRUE),
-        tpi_scaled_sp = scale(tpi, center = TRUE, scale = TRUE)
-    ) %>%
-    ungroup() %>%
-    # remove large outliers for each year
[…]

### 2025-11-01 scripts/01_prepdata.R
minor script cleaning + session info

@@ -7,6 +7,7 @@ library(lubridate)
 library(httr)
 library(janitor)
 library(raster)
+library(ggplot2)
 
 # reading the GitHub PAT (stored privately and locally)
 # pat<-as.character(read.table("../data/HKK-dendro/git_pat.txt")[1,1])
@@ -305,7 +306,6 @@ tree.time <- sensitivity %>%
 trees <- trees %>% filter(Tag %in% tree.time$Tag)
 
 # twi values
-library(raster)
 twi <- raster::raster("data/HKK-other/TWI.tif")
 
 # pick twi values for each tree
@@ -408,7 +408,7 @@ hkk.stem4$tpi <- raster::extract(tpi, 0.1 * hkk.stem4[, c("gy", "gx")]) # multip
 head(hkk.stem4)
 
 # plot SACCLI trees on twi to make sure that the scale is right
-library(ggplot2)
+
 twi_df <- as.data.frame(twi, xy = TRUE)
 
 png("doc/display/explore/twi_SACCLI.png", width = 5, height = 8, units = "in", res = 300)

### 2025-11-01 scripts/02_species.R

@@ -1,6 +1,7 @@
 # Load required libraries---------------------
 library(tidyverse)
 library(brms)
+library(patchwork)
 
 # load data--------------------------
 rm(list = ls())
@@ -143,7 +144,6 @@ png("results/plots/non_negative/ranefs_intercept.png", width = 8, height = 8, un
 ranef_plot
 dev.off()
 
-library(patchwork)
 png("results/plots/non_negative/pp_intercept.png", width = 8, height = 4, res = 300, units = "in")
 pp_check(fits[[1]]) + pp_check(fits[[2]])
 dev.off()
@@ -300,9 +300,6 @@ coefs_plot <- ggplot(coefs_df %>% filter(param == "b_williams_dec"), aes(x = yr,
 coefs_plot
 
 # model with twi--------------------------------------------
-library(brms)
-# isocline_model <- bf(sens.prop ~ 1 + twi + williams_dec + (1 + twi | Species))
-
 isocline_model_twi <- bf(sens.prop ~ 1 + twi + williams_dec + twi:williams_dec)
 
 coefs <- list()
@@ -506,7 +503,6 @@ saveRDS(new_preds, "results/models/non_negative/new_preds_isoclines_twi_tpi.RDS"
 
 # model with size and deciduousness------------------------
 
-library(brms)
 isocline_model_dbh <- bf(sens.prop ~ 1 + calcDBH_min1 + williams_dec + calcDBH_min1:williams_dec)
 
 coefs <- list()
@@ -572,23 +568,6 @@ coefs_df$signif <- ifelse(coefs_df$lwr < 0 & coefs_df$upr > 0, "no",
 )
 
 coefs_df
-
-# library(ggplot2)
-# # plot for four species
-# iso_plot <- ggplot(
-#     new_preds_df,
-#     aes(x = williams_dec, y = twi, fill = Estimate)
-# ) +
-#     geom_tile() +
-#     geom_contour(aes(z = Estimate), colour = "black") +
-#     facet_grid(yr ~ Species) +
-#     theme_bw() +
-#     scale_fill_gradient2()
-
-# png("results/plots/non_negative/isocline_trial.png", width = 32, height = 4, units = "in", res = 300)
-# iso_plot
-# dev.off()
-
 
 iso_plot <- ggplot(
     new_preds_df,
[…]

### 2025-11-01 scripts/03_orderedcii.R

@@ -3,6 +3,7 @@
 # Load required libraries---------------------
 library(tidyverse)
 library(brms)
+library(patchwork)
 
 # load data--------------------------
 rm(list = ls())
@@ -275,7 +276,7 @@ pred_plot_cii <- ggplot(pred_cii_df, aes(x = cii_min1, y = median)) +
         y = "Predicted CII"
     )
 
-library(patchwork)
+
 png(paste0(plot_dir, "/pred_obs_fullmediation.png"), width = 8, height = 8, units = "in", res = 300)
 pred_plot / pred_plot_cii
 dev.off()

### 2025-11-01 scripts/04_orderedcii_tpi.R

@@ -3,10 +3,11 @@
 # Load required libraries---------------------
 library(tidyverse)
 library(brms)
+library(patchwork)
 
 # load data--------------------------
 rm(list = ls())
-tree.time <- read.csv("data/HKK-dendro/to_publish/sensitivity_dataset_for_review.csv")
+tree.time <- read.csv("data/dendro/sensitivity_dataset.csv")
 
 
 # directory
@@ -222,7 +223,6 @@ pred_plot_cii <- ggplot(pred_cii_df, aes(x = cii_min1, y = median)) +
         y = "Predicted CII"
     )
 
-library(patchwork)
 png(paste0(plot_dir, "/pred_obs_fullmediation.png"), width = 8, height = 8, units = "in", res = 300)
 pred_plot / pred_plot_cii
 dev.off()

### 2025-11-01 scripts/05_figures.R

@@ -5,6 +5,8 @@ library(tidyverse)
 library(lubridate)
 library(ggplot2)
 library(patchwork)
+library(ggpubr)
+library(tidybayes)
 
 # figure 1 ------------------------------------------------
 
@@ -32,8 +34,6 @@ spei_month <- spei %>%
     pivot_longer(cols = c(SPEI_01_month, SPEI_03_month, SPEI_06_month, SPEI_12_month), names_to = "spei_var", values_to = "spei_val") %>%
     select(year, month, spei_var, spei_val)
 
-
-library(patchwork)
 
 # figure 1 alternate with 3 years chirps and era5land-----------------------
 
@@ -233,7 +233,7 @@ climsat_anomaly_plot <- ggplot(climsat_rlmean %>% filter(year %in% c("2010", "20
 climsat_anomaly_plot
 speiplot_line
 
-library(patchwork)
+
 png("doc/display/climsat_plot.png", width = 6, height = 8, units = "in", res = 300)
 climsat_plot + climsat_anomaly_plot
 dev.off()
@@ -352,9 +352,6 @@ climplot_anomalies_SI
 dev.off()
 
 # figure 2 - growth increments ENSO plot + sensitivity raw distributions-------------------------
-
-# Load required libraries---------------------
-library(tidyverse)
 
 # load data--------------------------
 tree.time <- read.csv("data/dendro/sensitivity_dataset.csv")
@@ -504,8 +501,6 @@ layout <- "
 AAABB
 "
 
-library(patchwork)
-
 layout <- "
 AAA
 AAA
@@ -628,9 +623,6 @@ iso_plot_twi <- ggplot(
 
 iso_plot_twi
 
-
-library(patchwork)
-# png("doc/display/Fig3.png", width = 8, height = 8, units = "in", res = 300)
 png("doc/display/Fig3.png", width = 8, height = 8, units = "in", res = 300)
 dec_intercept_plot / iso_plot_twi + plot_annotation(tag_levels = "a") + plot_layout(heights = c(1.8, 1))
 dev.off()
@@ -785,8 +777,6 @@ dec_intercept_plot <- ggplot(ranef_df_long, aes(x = value, y = intercept)) +
         strip.text = element_text(size = 12)
     )
 
-
-library(patchwork)
[…]

### 2025-11-01 scripts/07_otherfigs.R

@@ -2,6 +2,8 @@
 library(tidyverse)
 library(brms)
 library(ggplot2)
+library(ggpubr)
+library(patchwork)
 
 # read data
 rm(list = ls())
@@ -213,7 +215,6 @@ cond_dep_all <- tree.time %>%
 
 # plot the conditional independencies
 
-library(ggpubr)
 cond_dep_all_plot <- ggscatter(
     data = cond_dep_all,
     x = "value", y = "twi",
@@ -434,7 +435,6 @@ cond_dep <- tree.time %>%
 
 # plot the conditional independencies
 
-library(ggpubr)
 cond_dep_dbh_twi <- ggscatter(
     data = tree.time %>% filter(Cno == 15),
     x = "calcDBH_min1", y = "twi",
@@ -487,7 +487,7 @@ cond_dep_dbh_twi_all <- ggscatter(
     facet_wrap(~yr)
 
 cond_dep_dbh_twi_all
-library(patchwork)
+
 png("doc/display/cond_dep_alltrees.png", width = 8, height = 8, units = "in", res = 300)
 cond_dep_dbh_twi_all + cond_dep_cii_twi_all + plot_layout(ncol = 1)
 dev.off()

### 2025-11-01 doc/response_to_reviewer_R2.Rmd
knit manuscript and responses

---
title: 'Responses to the Reviewer'
output: 
    bookdown::word_document2: 
    #   reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
      reference_docx: word-styles-reference-singlespace.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    #pdf_document:
    # engine: xelatex 
  #officedown::rdocx_document
bibliography: growth-precip-thailand.bib
# csl: new-phytologist.csl
csl: ecology-letters.csl 
suppress-bibliography: TRUE
---

Dear Editor,

We are pleased to submit a revised version of our manuscript, "Drought response is not a species trait: tropical tree drought sensitivity is shaped by drought characteristics, species adaptations, and individual microenvironments" (ELE-00826-2025) for consideration for publication in Ecology Letters. 
We have addressed all points raised by the Data Editor as detailed in our responses to the reviewers and have updated the files provided in the Zenodo library to ensure reproducibility of our results. 
We have also updated the Data Availability Statement in the manuscript to reflect this.

Thank you for considering this revised version for publication. 
We look forward to your response.

Sincerely,
On behalf of all coauthors,  
Krishna Anujan

\newpage

## Referees' comments to the author(s):

### Referee: 1

Comments for the Authors
Congratulations on sharing the code. However, the data are only shared upon request and are not open, contrary to the request of Ecology Letters, so the analyses are not reproducible. Briefly, the README is present and describes all the content. The scripts and manuscript codes are made available. The only thing lacking is metadata to understand the data's content. The only necessary step for reproducibility would be to share the data. To improve upon this, the author could consider defining package versions, or even better, using an environment manager such as renv. They could also take advantage of a forge such as GitHub to make things easier and improve reproducibility. Consequently, I rated the manuscript as '1) Inadequate (data or code not given or insufficient)' as the data are not provided.

**RESPONSE: Thanks for the favourable review of our manuscript code. We have now updated the Zenodo repository [doi: 10.5281/zenodo.15777974](https://zenodo.org/records/15777975) to include the datasets required to reproduce these analyses as csvs. We have also included metadata csvs describing the columns in the data csvs. Finally, as suggested here by the Data Editor, we have also included a session_info.txt file that describes the hardware and software versions (including packages) used to run these analyses. We believe that this revision allows for full reproducibility of our results and figures. Moreover, the entire analyses was managed through a GitHub repository - accessible through the Zenodo link - so the versions are traceable and reproducible.**

### 2025-11-09 doc/manuscript_hkk_drought_sensitivity.Rmd
data availability statement updated
updated manuscript Rmd with the public datasets

@@
 **Title:** Drought response is not a species trait: tropical tree drought sensitivity is shaped by drought characteristics, species adaptations, and individual microenvironments
-
-<!-- **Title 1:** Sensitivity of tree growth to drought in a seasonally dry tropical forest using long-term dendrometer band measurements
-
-**Title 2:** Contrasting effects of species and environmental predictors on tree growth sensitivity to different/two droughts in a seasonally dry tropical forest
-
-**Title 3:** Diversity of species- and environment-mediated drought responses in a seasonally dry tropical forest
-
-**Title 4:** Long-term dendrometer band measurements reveal diverse species- and environment-mediated drought responses in a seasonally dry tropical forest
-
-**Title 5:** Deciduousness may exacerbate/worsen/amplify tree growth sensitivity during wet season droughts in seasonally dry tropical forests -->
 
@@
 <!-- The climate data used in this work are publicly available.  -->
-Data and code required to reproduce the results and figures are archived in Zenodo with doi:10.5281/zenodo.15777974.
+Data and code required to reproduce the results and figures are archived in Zenodo with [doi:10.5281/zenodo.15777974](https://zenodo.org/records/15777975).
 
@@
 library(tidyverse)
-tree.time <- readRDS("../data/HKK-dendro/sensitivity_data_formodels.RData")
-# tree.time <- readRDS("data/HKK-dendro/sensitivity_data_formodels.RData")
-
-tree.time.sp <- tree.time %>%
-  group_by(spfull) %>%
-  dplyr::summarise( # change to mm
-    inc_annual_med = round(median(inc_annual * 10, na.rm = T), 2),
-    inc_annual_sd = round(sd(inc_annual * 10, na.rm = T), 2)
-  ) %>%
-  arrange(inc_annual_med)
+library(here)
+tree.time <- read.csv(here("data", "dendro", "sensitivity_dataset.csv"))
+ann.summary <- read.csv(here("data", "dendro", "summaries_dataset.csv"))
+# tree.time <- read.csv("data/dendro/sensitivity_dataset.csv")
+# ann.summary <- read.csv("data/dendro/summaries_dataset.csv")
 
 sens.sp <- tree.time %>%
-  filter(yr %in% c(2010, 2015, 2020)) %>%
   group_by(spfull, yr) %>%
@@
   arrange(sens_med)
-
-# spread this for each year
 
@@
 Drought years were associated with lower plot-level growth and negative tree-level sensitivities, with the strongest negative response in 2015 (Fig `r fig_drought_responses`). 
-Median growth rate across all trees and years was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` mm (range `r round(min(tree.time$inc_annual*10, na.rm=T),2)`, `r round(max(tree.time$inc_annual*10, na.rm=T),2)`).
-The drought year 2015 had the lowest median increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% lower than the median across all years.
-In the 2010 and 2020 droughts, the median increments were `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm and `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm respectively, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% and `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% lower than the median across all years.
-<!-- In the 2020 drought, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years.
--->
-<!-- On average, trees had negative drought sensitivities during drought years, but with considerable variation (Fig `r fig_drought_responses`b).  -->
+Median growth rate across all trees and years was `r round(ann.summary$median_inc_mm[ann.summary$yr=="all"],2)` mm (range -14.2, 29.32). 
+The drought year 2015 had the lowest median increment `r round(ann.summary$median_inc_mm[ann.summary$yr=="2015"],2)` mm, `r round(-100*((ann.summary$median_inc_mm[ann.summary$yr=="2015"]-ann.summary$median_inc_mm[ann.summary$yr=="all"])/ann.summary$median_inc_mm[ann.summary$yr=="all"]),0)`% lower than the median across all years.
+In the 2010 and 2020 droughts, the median increments were `r round(ann.summary$median_inc_mm[ann.summary$yr=="2010"],2)` mm and `r round(ann.summary$median_inc_mm[ann.summary$yr=="2020"],2)` mm respectively, `r round(-100*((ann.summary$median_inc_mm[ann.summary$yr=="2010"]-ann.summary$median_inc_mm[ann.summary$yr=="all"])/ann.summary$median_inc_mm[ann.summary$yr=="all"]),0)`% and `r round(-100*((ann.summary$median_inc_mm[ann.summary$yr=="2020"]-ann.summary$median_inc_mm[ann.summary$yr=="all"])/ann.summary$median_inc_mm[ann.summary$yr=="all"]),0)`% lower than the median across all years.
 Median drought sensitivity across all trees in 2010, 2015 and 2020 were `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)`<!-- $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)`-->, `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)`<!-- $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`--> and `r round(median(tree.time$sens.prop[tree.time$yr==2020]), 2)`<!-- $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2020]), 2)`--> respectively.
@@
 <!-- Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S4).  -->
-<!-- Across the years, on average, `r round(100*mean(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% of the trees had negative growth, with a minimum of `r round(100*min(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% in `r names(sort(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))))[1]`.
--->
-<!-- The drought year 2015 had the maximum proportion of trees with negative growth in the timeseries, `r round(100*sum(tree.time$inc_annual[tree.time$yr==2015]<0)/nrow(tree.time[tree.time$yr==2015,]),2)`%.
--->
 <!-- *(See [issue #28](https://github.com/forestgeo/growth-precip-thailand/issues/28))* -->
@@
 Species median sensitivities and individual tree sensitivities to the three droughts were not correlated with each other, supporting prediction (*i*) (Fig S6, Fig S7). 
-Median annual increment across years varied among the `r nrow(tree.time.sp)` species analysed, with *`r tree.time.sp$spfull[1]`* the lowest (`r tree.time.sp$inc_annual_med[1]`<!-- $\pm$ `r tree.time.sp$inc_annual_sd[1]`--> mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]`<!-- $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]`--> mm).
+Median annual increment across years varied among the 30 species analysed, with *Gluta obovata* the lowest (0.61 mm) and *Dipterocarpus alatus* the highest (3.58 mm).
 In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/3]`*.
@@
 ```{r model summaries, echo=F, warning=F, message=F}
-# load data--------------------------
-datasets <- readRDS("../data/HKK-dendro/sensitivity_data.RData")
-
-tree.time <- datasets$tree.time
-tree_vars <- datasets$tree_vars
-sp_vars <- datasets$sp_vars
-
-colnames(sp_vars)[1] <- "Species"
+# make sp_vars data
+sp_vars <- tree.time %>%
+  group_by(Species, spfull) %>%
+  dplyr::summarise(williams_dec = first(williams_dec))
 

### 2025-11-09 scripts/01_prepdata.R

@@ -411,35 +411,35 @@ head(hkk.stem4)
 
 twi_df <- as.data.frame(twi, xy = TRUE)
 
-png("doc/display/explore/twi_SACCLI.png", width = 5, height = 8, units = "in", res = 300)
-ggplot() +
-  geom_raster(data = twi_df, aes(x = x, y = y, fill = TWI)) +
-  scale_fill_distiller(palette = "Spectral", direction = 1) +
-  geom_point(data = hkk.stem4[hkk.stem4$sp == "SACCLI", ], aes(x = 0.1 * gy, y = 0.1 * gx), pch = 16, alpha = 0.5) +
-  coord_fixed()
-dev.off()
+# png("doc/display/explore/twi_SACCLI.png", width = 5, height = 8, units = "in", res = 300)
+# ggplot() +
+#   geom_raster(data = twi_df, aes(x = x, y = y, fill = TWI)) +
+#   scale_fill_distiller(palette = "Spectral", direction = 1) +
+#   geom_point(data = hkk.stem4[hkk.stem4$sp == "SACCLI", ], aes(x = 0.1 * gy, y = 0.1 * gx), pch = 16, alpha = 0.5) +
+#   coord_fixed()
+# dev.off()
 
-# plot distribution of twi across top 4 species
-png("doc/display/explore/twi_dist_top4.png", width = 5, height = 5, units = "in", res = 300)
-ggplot(hkk.stem4 %>% filter(sp %in% c("SACCLI", "HOPEOD", "POLYVI", "TETRNU")), aes(x = twi, fill = sp)) +
-  geom_density(alpha = 0.5) +
-  scale_fill_viridis_d() +
-  theme_minimal() +
-  labs(title = "Distribution of top 4 species in TWI space (HKK)")
-dev.off()
+# # plot distribution of twi across top 4 species
+# png("doc/display/explore/twi_dist_top4.png", width = 5, height = 5, units = "in", res = 300)
+# ggplot(hkk.stem4 %>% filter(sp %in% c("SACCLI", "HOPEOD", "POLYVI", "TETRNU")), aes(x = twi, fill = sp)) +
+#   geom_density(alpha = 0.5) +
+#   scale_fill_viridis_d() +
+#   theme_minimal() +
+#   labs(title = "Distribution of top 4 species in TWI space (HKK)")
+# dev.off()
 
 # plot distribution as violin plot for top 10 species
 dec_william_sort <- dec_williams %>%
   arrange(williams_dec)
 
[…]

### 2025-11-15 doc/response_to_reviewer_R3.Rmd
response to reviewer

---
title: 'Responses to the Reviewer'
output: 
    bookdown::word_document2: 
    #   reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
      reference_docx: word-styles-reference-singlespace.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    #pdf_document:
    # engine: xelatex 
  #officedown::rdocx_document
bibliography: growth-precip-thailand.bib
# csl: new-phytologist.csl
csl: ecology-letters.csl 
suppress-bibliography: TRUE
---

Dear Editor,

We are pleased to submit a revised version of our manuscript, "Drought response is not a species trait: tropical tree drought sensitivity is shaped by drought characteristics, species adaptations, and individual microenvironments" (ELE-00826-2025) for consideration for publication in Ecology Letters. 
We have addressed the main point raised by the Data Editor as detailed in our responses to the reviewers and have updated the files provided in the Zenodo library to improve reproducibility of our results. 
All our main analyses, figures and manuscript are now reproducible using the datasets and code currently made available.
As detailed in the response, one of the suggestions by the reviewer, sharing unprocessed data, would not be possible to accomplish within the policies of the organisation managing the long-term data collection in Thailand. 
Despite this, we have provided all processing steps for the use of these data and exporting to a processed dataset. 
All main analyses, as well as the manuscript can now be run using these files alone.
We hope that this revision demonstrates our commitment to reproducibility within institutional constraints.
We hope you can consider global differences in data policies, appreciate the enormous effort required in maintaining such a dataset, as well as the scientific merit of our work and consider this version for publication.

Thank you for considering this revised version for publication. 
We look forward to your response.

Sincerely,
On behalf of all coauthors,  
Krishna Anujan

\newpage

## Referees' comments to the author(s):

### Referee: 1

Comments for the Authors
Congratulations to the authors on improving the sharing of data and code. However, this revision does not enable full reproducibility of the results and figures as stated. For example, the file “../data/HKK-dendro/sensitivity_data_formodels.RData” is not available to run the manuscript file, and no line of code can produce it in the different scripts. Similarly, data is lacking to run the scripts. For example, the 'data/HKK-other/CTFSElev_hkk.rdata' file is missing for 'twi.R'. This is thus not reproducible. The 'band censuses' referred to in the materials from line 174 are also not available, despite constituting the raw base data of the analyses. As far as I am concerned, the intermediary data required to repeat the analyses is lacking, so I cannot grant a mark of '2) Adequate (link to data to repeat analyses, even if not full raw data, and link to code to repeat analysis)' and I am still marking it as '1) Inadequate (data not given or insufficient, code not given or insufficient)'.

**RESPONSE: Thanks for your comments. We apologise for the confusion - the previous manuscript version was inadvertently linked to a private copy of the same file. We have now corrected this and updated the Zenodo repository (DOI: 10.5281/zenodo.15777975) with a manuscript that relies exclusively on the shared data csv files. As shown in the track-changes version, no content in the manuscript itself has been modified.

However, institutional policies of the Thai Long-Term Forest Ecological Research project prevent us from sharing some raw data files - specifically CTFSelev_hkk.rdata and the raw dendroband census information. We acknowledge that this limits full data availability, but all processing scripts for these raw data are now openly accessible in the repository. In particular, scripts 000_twi.R and 01_prepdata.R contain complete codes to process the raw dendroband census data and elevation data. These scripts output data csvs that are currently made public under data/dendro. To improve transparency, we have reorganised the scripts as *processing* and *analysis* scripts in different folders.

These processed datasets are now minimally sufficient to reproduce all the results through analysis scripts (02_species.R, 03_orderedcii.R, 04_orderedcii_tpi.R, 06_causal.R) and produce figures in the main text and SI (05_figures.R). Two SI figures (S4 and S8) that require raw data are exported from the data prep file (01_prepdata.R).

We hope this clarification helps demonstrate that the shared data and code now allow full reproduction of the analyses from the processed data stage onward. We also ask that you consider the constraints of local data-sharing policies, the significant effort required to maintain such long-term datasets, and the scientific value of the work in your assessment of this revised version for publication.**

### 2025-11-19 doc/response_to_reviewer_R3.Rmd
small updates to dataset in SI file

@@
 All main analyses, as well as the manuscript can now be run using these files alone.
-We hope that this revision demonstrates our commitment to reproducibility within institutional constraints.
+
+While we are committed to reproducible science, the requirement of publishing raw data can disproportionately disadvantage researchers in the global south (*sensu* de Lima et al., 2022).
+We hope that this revision demonstrates our commitment to reproducibility within institutional and practical constraints.
 We hope you can consider global differences in data policies, appreciate the enormous effort required in maintaining such a dataset, as well as the scientific merit of our work and consider this version for publication.
@@
 We have now corrected this and updated the Zenodo repository (DOI: 10.5281/zenodo.15777975) with a manuscript that relies exclusively on the shared data csv files.
-As shown in the track-changes version, no content in the manuscript itself has been modified.
+As shown in the track-changes version, no content in the manuscript itself has been modified.**
 
-However, institutional policies of the Thai Long-Term Forest Ecological Research project prevent us from sharing some raw data files - specifically CTFSelev_hkk.rdata and the raw dendroband census information.
+**However, institutional policies of the Thai Long-Term Forest Ecological Research project prevent us from sharing some raw data files - specifically CTFSelev_hkk.rdata and the raw dendroband census information.
 We acknowledge that this limits full data availability, but all processing scripts for these raw data are now openly accessible in the repository.
@@
 These scripts output data csvs that are currently made public under data/dendro.
-To improve transparency, we have reorganised the scripts as *processing* and *analysis* scripts in different folders.
+To improve transparency, we have reorganised the scripts as *processing* and *analysis* scripts in different folders.**
 
-These processed datasets are now minimally sufficient to reproduce all the results through analysis scripts (02_species.R, 03_orderedcii.R, 04_orderedcii_tpi.R, 06_causal.R) and produce figures in the main text and SI (05_figures.R).
-Two SI figures (S4 and S8) that require raw data are exported from the data prep file (01_prepdata.R).
+**These processed datasets are now minimally sufficient to reproduce all the results through analysis scripts (02_species.R, 03_orderedcii.R, 04_orderedcii_tpi.R, 06_causal.R) and produce figures in the main text and SI (05_figures.R).
+Two SI figures (S4 and S8) that require raw data are exported from the data prep file (01_prepdata.R).**
 
-We hope this clarification helps demonstrate that the shared data and code now allow full reproduction of the analyses from the processed data stage onward.
+**We hope this clarification helps demonstrate that the shared data and code now allow full reproduction of the analyses from the processed data stage onward.
 We also ask that you consider the constraints of local data-sharing policies, the significant effort required to maintain such long-term datasets, and the scientific value of the work in your assessment of this revised version for publication.**

### 2025-11-19 doc/supplementary_information.Rmd
small updates to dataset in SI file

@@
 library(tidyverse)
-datasets <- readRDS("../data/HKK-dendro/sensitivity_data.RData")
-
-tree.time <- datasets$tree.time
-tree_vars <- datasets$tree_vars
-sp_vars <- datasets$sp_vars
-
-sp_new <- read.csv("../data/HKK-dendro/species.csv")
-
-colnames(sp_vars)[1] <- "Species"
-sp_vars$iucn <- sp_new$iucn[match(sp_vars$Species, sp_new$sp)]
-
-# merge tree_vars and sp_vars
-tree_vars <- merge(tree_vars, sp_vars, by = "Species", all.x = TRUE)
-# merge tree_vars and tree.time
-tree.time <- merge(tree.time, tree_vars, by = "Tag", all.x = TRUE)
-
-# colnames(tree.time)
-
-# standardise the variables within species and across all--------------------
-
-tree.time <- tree.time %>%
-  # remove large outliers for each year
-  group_by(yr) %>%
-  dplyr::mutate(sens.prop = ifelse(mean(sens.prop, na.rm = TRUE) + 4 * sd(sens.prop, na.rm = TRUE) > sens.prop & sens.prop > mean(sens.prop, na.rm = TRUE) - 4 * sd(sens.prop, na.rm = TRUE), sens.prop, NA)) %>%
-  filter(!is.na(sens.prop) & !is.na(cii_min1) & !is.na(calcDBH_min1) & !is.na(twi)) %>%
-  ungroup()
+library(here)
+tree.time <- read.csv(here("data", "dendro", "sensitivity_dataset.csv"))
+iucn <- read.csv(here("data", "dendro", "sp_iucn.csv"))
 
@@
   dplyr::summarise(
+    sp = dplyr::first(Species),
+    williams_dec = dplyr::first(williams_dec),
     n = n(),
@@
 
-sens_sp <- merge(sens_sp, dplyr::select(sp_vars, c(spfull, Species, williams_dec, iucn)), by = "spfull", all.x = T)
+sens_sp <- merge(sens_sp, iucn, by = "sp", all.x = T)
 
@@
     "Species name" = spfull,
-    "Species code" = Species
+    "Species code" = sp
   )

### 2025-11-25 doc/response_to_reviewer_R3.Rmd
response doc update

@@
 
-**These processed datasets are now minimally sufficient to reproduce all the results through analysis scripts (02_species.R, 03_orderedcii.R, 04_orderedcii_tpi.R, 06_causal.R) and produce figures in the main text and SI (05_figures.R).
+**These processed datasets are now minimally sufficient to reproduce all the results through analysis scripts (01_species.R, 02_orderedcii.R, 03_orderedcii_tpi.R, 04_causal.R) and produce figures in the main text and SI (05_figures.R).
 Two SI figures (S4 and S8) that require raw data are exported from the data prep file (01_prepdata.R).**

### 2025-11-25 scripts/analysis/01_species.R
updated species script to run independently

@@ -1,13 +1,32 @@
 # Load required libraries---------------------
-library(tidyverse)
-library(brms)
-library(patchwork)
+
+required_pkgs <- c(
+    "tidyverse", "brms", "patchwork"
+)
+
+# install any missing packages
+missing_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[, "Package"])]
+if (length(missing_pkgs) > 0) {
+    install.packages(missing_pkgs, dependencies = TRUE)
+}
+
+# load packages quietly
+invisible(lapply(required_pkgs, function(pkg) {
+    suppressPackageStartupMessages(require(pkg, character.only = TRUE))
+}))
 
 # load data--------------------------
 rm(list = ls())
 tree.time <- read.csv("data/dendro/sensitivity_dataset.csv")
 median_incs <- read.csv("data/dendro/summaries_dataset.csv")
 
+head(list.files(recursive = T))
+
+# ensure output directory exists
+if (!dir.exists("results/models/non_negative")) {
+    dir.create("results/models/non_negative", recursive = TRUE)
+    message("Created directory: results/models/non_negative")
+}
 
 # plot the sensitivity values for all trees in 2010 and 2015
 yrs <- c(2010, 2015, 2020)
@@ -30,13 +49,15 @@ sens_sp <- tree.time %>%
     ) %>%
     ungroup()
 
-# b <- boot::boot(tree.time$sens.prop, function(x, i) median(x[i]), R = 1000)
-# b$t0
-# boot::boot.ci(b)
-
 head(sens_sp)
+head(median_incs)
 
-sens.sp <- merge(sens_sp, median_incs, by = "Species", all.x = TRUE)
+# make sp_vars data
+sp_vars <- tree.time %>%
+    group_by(Species, spfull) %>%
+    dplyr::summarise(williams_dec = first(williams_dec))
+
+sens.sp <- merge(sens_sp, sp_vars, by = "Species", all.x = TRUE)
 
 # plot median sensitivity against deciduousness
 sens_sp_median <- ggplot(sens.sp, aes(x = williams_dec, y = sens_median)) +
@@ -54,24 +75,9 @@ sens_sp_median
 sens_sp_lms <- sens.sp %>%
     filter(Species != "ALPHVE") %>%
     nest_by(yr) %>%
[…]

### 2025-11-25 scripts/analysis/02_orderedcii.R
updated scripts for reusability

@@ -1,71 +1,42 @@
 # models considering cii as an ordered factor
 
 # Load required libraries---------------------
-library(tidyverse)
-library(brms)
-library(patchwork)
+
+required_pkgs <- c(
+    "tidyverse", "brms", "patchwork", "performance"
+)
+
+# install any missing packages
+missing_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[, "Package"])]
+if (length(missing_pkgs) > 0) {
+    install.packages(missing_pkgs, dependencies = TRUE)
+}
+
+# load packages quietly
+invisible(lapply(required_pkgs, function(pkg) {
+    suppressPackageStartupMessages(require(pkg, character.only = TRUE))
+}))
 
 # load data--------------------------
 rm(list = ls())
 tree.time <- read.csv("data/dendro/sensitivity_dataset.csv")
-
-
-# conditional dependencies---------------
-
-# ## test conditional independences for each species-----------------------------
-
-# # DBH | TWI
-# # CII | TWI
-
-# # 2015 data only
-
-# cond_dep <- tree.time %>%
-#     filter(yr == 2015) %>%
-#     group_by(Species) %>%
-#     dplyr::summarise(
-#         DBH_TWI = cor.test(calcDBH_min1, twi)[4]$estimate,
-#         DBH_TWI_p = cor.test(calcDBH_min1, twi)[3]$p.value,
-#         CII_TWI = cor.test(cii_min1, twi)[4]$estimate,
-#         CII_TWI_p = cor.test(cii_min1, twi)[3]$p.value
-#     )
-
-# # these variables are conditionally independent for the most part
-
-# # plot the conditional independencies
-
-# library(ggpubr)
-# cond_dep_dbh_twi <- ggscatter(
-#     data = tree.time %>% filter(Cno == 15),
-#     x = "calcDBH_min1", y = "twi",
-#     add = "reg.line", conf.int = TRUE,
-#     cor.coef = TRUE, cor.method = "spearman",
-#     xlab = "DBH", ylab = "TWI"
-# ) +
-#     facet_wrap(~ factor(Species, levels = names(sort(table(Species), decreasing = T))))
-
-# png("doc/display/cond_dep_dbh_twi.png", width = 12, height = 12, units = "in", res = 300)
-# cond_dep_dbh_twi
-# dev.off()
-
-# # CII | TWI
-# cond_dep_cii_twi <- ggscatter(
[…]

### 2025-11-25 scripts/analysis/03_orderedcii_tpi.R

@@ -1,19 +1,43 @@
 # models considering cii as an ordered factor
 
 # Load required libraries---------------------
-library(tidyverse)
-library(brms)
-library(patchwork)
+required_pkgs <- c(
+    "tidyverse", "brms", "patchwork", "performance"
+)
+
+# install any missing packages
+missing_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[, "Package"])]
+if (length(missing_pkgs) > 0) {
+    install.packages(missing_pkgs, dependencies = TRUE)
+}
+
+# load packages quietly
+invisible(lapply(required_pkgs, function(pkg) {
+    suppressPackageStartupMessages(require(pkg, character.only = TRUE))
+}))
+
 
 # load data--------------------------
 rm(list = ls())
 tree.time <- read.csv("data/dendro/sensitivity_dataset.csv")
+yrs <- c(2010, 2015, 2020)
 
-
-# directory
+# directories--------
 
 plot_dir <- "results/plots/orderedcii_tpi"
 models_dir <- "results/models/orderedcii_tpi"
+
+# ensure output directory exists
+if (!dir.exists(plot_dir)) {
+    dir.create(plots_dir, recursive = TRUE)
+    message(paste0("Created directory: ", plots_dir))
+}
+
+if (!dir.exists(models_dir)) {
+    dir.create(models_dir, recursive = TRUE)
+    message(paste0("Created directory: ", models_dir))
+}
+
 
 # define and run the model
 
@@ -26,9 +50,7 @@ pred <- list()
 fits <- list()
 
 for (i in 1:length(yrs)) {
-    # yrs<-c(2010, 2015, 2020)
     fit <- brm(tree_model,
-        # data = tree.time %>% filter(yr == yrs[i]), family = skew_normal(),
         data = tree.time %>% filter(yr == yrs[i]), family = gaussian(),
         chains = 4, iter = 4000, warmup = 2000, cores = 4
     )

### 2025-11-25 scripts/analysis/04_causal.R

@@ -1,19 +1,43 @@
 # models considering cii as an ordered factor and using independent causal queries
 
 # Load required libraries---------------------
-library(tidyverse)
-library(brms)
+
+required_pkgs <- c(
+    "tidyverse", "brms", "patchwork", "performance"
+)
+
+# install any missing packages
+missing_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[, "Package"])]
+if (length(missing_pkgs) > 0) {
+    install.packages(missing_pkgs, dependencies = TRUE)
+}
+
+# load packages quietly
+invisible(lapply(required_pkgs, function(pkg) {
+    suppressPackageStartupMessages(require(pkg, character.only = TRUE))
+}))
 
 # load data--------------------------
 rm(list = ls())
 tree.time <- read.csv("data/dendro/sensitivity_dataset.csv")
 median_incs <- read.csv("data/dendro/summaries_dataset.csv")
+yrs <- c(2010, 2015, 2020)
 
 # directory--------
 
 plot_dir <- "results/plots/causal"
 models_dir <- "results/models/causal"
 
+# ensure output directory exists
+if (!dir.exists(plot_dir)) {
+    dir.create(plots_dir, recursive = TRUE)
+    message(paste0("Created directory: ", plots_dir))
+}
+
+if (!dir.exists(models_dir)) {
+    dir.create(models_dir, recursive = TRUE)
+    message(paste0("Created directory: ", models_dir))
+}
 
 # define and run the models
 
@@ -29,7 +53,6 @@ pred <- list()
 fits <- list()
 i <- 1
 for (i in 1:length(yrs)) {
-    # yrs<-c(2010, 2015, 2020)
     fit <- brm(dbh_model,
         # data = tree.time %>% filter(yr == yrs[i]), family = skew_normal(),
         data = tree.time %>% filter(yr == yrs[i]), family = gaussian(),
@@ -63,7 +86,6 @@ pred <- list()
 fits <- list()
 
 for (i in 1:length(yrs)) {
-    # yrs<-c(2010, 2015, 2020)
     fit <- brm(twi_model,
         # data = tree.time %>% filter(yr == yrs[i]), family = skew_normal(),
         data = tree.time %>% filter(yr == yrs[i]), family = gaussian(),
[…]

### 2025-11-27 doc/response_to_reviewer_R4.Rmd
minor updates and responses file

---
title: 'Responses to the Reviewer'
output: 
    bookdown::word_document2: 
    #   reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
      reference_docx: word-styles-reference-singlespace.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    #pdf_document:
    # engine: xelatex 
  #officedown::rdocx_document
bibliography: growth-precip-thailand.bib
# csl: new-phytologist.csl
csl: ecology-letters.csl 
suppress-bibliography: TRUE
---

Dear Editor,

We are pleased to submit a revised version of our manuscript, "Drought response is not a species trait: tropical tree drought sensitivity is shaped by drought characteristics, species adaptations, and individual microenvironments" (ELE-00826-2025) for consideration for publication in Ecology Letters. 
We have now addressed all of the Data Editor's comments and updated the repository. 
Specifically, we have addressed issues with dependencies in the files and ensured that all analysis code and the manuscript are executable.
Thank you for your patience with this process and we hope that this revision sufficiently meets the journal guidelines. 
We appreciate the thoroughness and the attention to detail with the data review process at Ecology Letters and the focus on achieving reproducible science and reusable data under global collaboration.

Thank you for considering this revised version for publication. 
We look forward to your response.

Sincerely,  
On behalf of all coauthors,  
Krishna Anujan

\newpage

## Referees' comments to the author(s):

### Referee: 1

Comments for the Authors
I understand your argument regarding permanent plots data, particularly in the Global South, as I am working with similar networks myself. I was waiting for a decision from the editor because it falls within the limits of Ecology Letters' requirements. However, I completely agree with your suggestion to share reproducible analyses with intermediary files. That said, this is still not the case. I managed to reproduce analyses 2, 3 and 4 by manually editing and fixing the codes, but not analyses 1 and 5. The object 'yrs' is missing from scripts 2, 3, 4 and 5 for them to work (because the environment is purged when the scripts are started). Furthermore, subdirectories for the plots and model outputs must be created in scripts 2, 3 and 4 before the analyses can be launched. You can either add these to the repository, or add an R line of code to create them (dir.create). For analysis 1, the table's junction will not work because there is no “Species” column in “median_incs” from “summaries_dataset.csv”. For analysis 5, I will leave you to search for the issue, as running the rest of the Bayesian model was already very time consuming. Before the next submission, please ensure that your codes are reproducible by downloading your GitHub repository again to an empty folder and checking that it is running correctly. That said, congratulations on your efforts to ensure the reproducibility of your analyses, which I believe will be of deep interest to researchers and students. I apologise in advance if you feel my review is too harsh, but it is thorough and I believe you are almost there. I really believe you are almost there. The small improvement you need will make all the difference to whether it is reused in future or not.

**RESPONSE: Thank you for your detailed response and for your understanding relating to collaborative, global science. We are also grateful for the attention to detail and the focus on meaningful reproducible science. We appreciate your patience with the multiple rounds here. We have now updated the Zenodo repository (DOI: 10.5281/zenodo.15777975) with dir.create codes, better initialisation (including code to download any missing packages) and removed all remaining calls to previous datasets. All scripts under scripts/analysis are now sequentially executable. We have also confirmed this, as you suggested, by downloading the repo and running things.**

**As you pointed out, the Bayesian analyses are slow and the outputs files are large, so we have also removed a few deprecated/redundant models which we had originally left in the scripts for completeness.**

**We hope that this revision now meets the journal requirements and also accepted guidelines on open science.**

### 2025-11-27 scripts/analysis/03_orderedcii_tpi.R

@@ -85,45 +85,3 @@ saveRDS(fits, file = paste0(models_dir, "/fits_rel_spre.rds"))
 saveRDS(coefs, file = paste0(models_dir, "/coefs_rel_spre.rds"))
 saveRDS(pred_sens, file = paste0(models_dir, "/pred_sens_rel_spre.rds"))
 saveRDS(pred_cii, file = paste0(models_dir, "/pred_cii_rel_spre.rds"))
-
-
-# Model 2: varying slopes-----------------------------
-
-tree_model_spre <- bf(sens.prop ~ 1 + calcDBH_min1_scaled_sp + mo(cii_min1) + twi_scaled_sp + (1 + calcDBH_min1_scaled_sp + twi_scaled_sp + mo(cii_min1) | Species))
-cii_model <- bf(ordered(cii_min1) ~ calcDBH_min1_scaled_sp, family = cumulative(link = "logit"))
-
-# run the model
-coefs <- list()
-pred_sens <- list()
-pred_cii <- list()
-fits <- list()
-
-for (i in 1:length(yrs)) {
-    fit <- brm(tree_model_spre + cii_model + set_rescor(FALSE),
-        data = tree.time %>% filter(yr == yrs[i]), family = gaussian(),
-        chains = 4, iter = 4000, warmup = 2000, cores = 4,
-        control = list(adapt_delta = 0.9)
-    )
-    fits[[i]] <- fit
-    post <- posterior_samples(fit)
-    post_sum <- as.data.frame(t(apply(post, 2, quantile, probs = c(.5, .05, .95))))
-    colnames(post_sum) <- c("median", "lwr", "upr")
-    post_sum$param <- rownames(post_sum)
-    post_sum$yr <- rep(yrs[i], nrow(post_sum))
-    coefs[[i]] <- post_sum
-
-    # make predictions
-    preds <- posterior_predict(fit)
-    # this makes a dataframe with 4000 rows (chains* sampling iterations) and 1449 columns (number of trees)
-    pred_sens_df <- as.data.frame(t(apply(preds[, , 1], 2, quantile, probs = c(.5, .05, .95))))
-    pred_cii_df <- as.data.frame(t(apply(preds[, , 2], 2, quantile, probs = c(.5, .05, .95))))
-    colnames(pred_sens_df) <- colnames(pred_cii_df) <- c("median", "lwr", "upr")
-    # add this to the observation data
[…]

### 2025-11-27 scripts/analysis/05_figures.R

@@ -706,12 +706,6 @@ coefs_df_tpi <- do.call(rbind, coefs_df_tpi)
 coefs_df_twi$sig <- ifelse(coefs_df_twi$lwr > 0 | coefs_df_twi$upr < 0, "sig", "not sig")
 coefs_df_tpi$sig <- ifelse(coefs_df_tpi$lwr > 0 | coefs_df_tpi$upr < 0, "sig", "not sig")
 
-coefs_df_twi_tpi <- readRDS("results/models/non_negative/coefs_isoclines_twi_tpi.RDS")
-coefs_df_twi_tpi <- do.call(rbind, coefs_df_twi_tpi)
-coefs_df_twi_tpi$sig <- ifelse(coefs_df_twi_tpi$lwr > 0 | coefs_df_twi_tpi$upr < 0, "sig", "not sig")
-
-
-
 
 # supplementary plot of other values ------------------------------------
 

### 2025-11-27 scripts/analysis/01_species.R
minor update

@@ -295,47 +295,3 @@ saveRDS(fits, "results/models/non_negative/fits_isoclines_tpi.RDS")
 saveRDS(pred, "results/models/non_negative/pred_isoclines_tpi.RDS")
 saveRDS(coefs, "results/models/non_negative/coefs_isoclines_tpi.RDS")
 saveRDS(new_preds, "results/models/non_negative/new_preds_isoclines_tpi.RDS")
-
-# model with twi and tpi--------------------------
-isocline_model_twi_tpi <- bf(sens.prop ~ 1 + twi + tpi + williams_dec + twi:williams_dec + tpi:williams_dec)
-coefs <- list()
-pred <- list()
-fits <- list()
-new_preds <- list()
-
-newdata <- expand.grid(
-    twi = seq(min(tree.time$twi, na.rm = T), max(tree.time$twi, na.rm = T), length.out = 15),
-    tpi = seq(min(tree.time$tpi, na.rm = T), max(tree.time$tpi, na.rm = T), length.out = 15),
-    williams_dec = seq(min(tree.time$williams_dec, na.rm = T), max(tree.time$williams_dec, na.rm = T), length.out = 15)
-)
-
-for (i in 1:length(yrs)) {
-    # yrs<-c(2010, 2015, 2020)
-    fit <- brm(isocline_model_twi_tpi, data = tree.time %>% filter(yr == yrs[i]), family = gaussian(), iter = 3000, warmup = 1000, chains = 4, cores = 4)
-    fits[[i]] <- fit
-    post <- posterior_samples(fit)
-    post_sum <- as.data.frame(t(apply(post, 2, quantile, probs = c(.5, .05, .95))))
-    colnames(post_sum) <- c("median", "lwr", "upr")
-    post_sum$param <- rownames(post_sum)
-    post_sum$yr <- rep(yrs[i], nrow(post_sum))
-    coefs[[i]] <- post_sum
-
-    # make predictions
-    preds <- posterior_predict(fit)
-    # this makes a dataframe with 4000 rows (chains* sampling iterations) and 1449 columns (number of trees)
-    pred_sum <- as.data.frame(t(apply(preds, 2, quantile, probs = c(.5, .05, .95))))
-    colnames(pred_sum) <- c("median", "lwr", "upr")
-    # add this to the observation data
-    # pred[[i]] <- cbind(tree.time %>% filter(yr == yrs[i]), pred_sum)
[…]
